---
title: "Weight-for-height Z-score in Bangladeshi infants aged 6-13 months"
subtitle: "Dataset 6: Bangladesh growth monitoring and promotion survey"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "bangladesh_growth_monitoring.csv"),
                    show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "weight_for_height_z"
predictors <- c("delivery_model", "child_sex", "child_age_months", "mother_age_yrs",
                "mother_schooling_yrs", "asset_index_tertile", "ngo_member",
                "improved_toilet")

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    delivery_model = factor(delivery_model,
                            levels = c("facility_based", "home_based"),
                            labels = c("Facility-based", "Home-based")),
    child_sex = factor(child_sex, levels = c("male", "female"),
                       labels = c("Male", "Female")),
    asset_index_tertile = factor(asset_index_tertile,
                                 levels = c("lowest", "middle", "highest"),
                                 labels = c("Lowest", "Middle", "Highest")),
    ngo_member = factor(ngo_member, levels = c(0, 1), labels = c("No", "Yes")),
    improved_toilet = factor(improved_toilet, levels = c(0, 1), labels = c("No", "Yes"))
  ) |>
  labelled::set_variable_labels(
    weight_for_height_z  = "Weight-for-height Z-score",
    delivery_model       = "Growth-monitoring delivery model",
    child_sex            = "Child's sex",
    child_age_months     = "Child's age (months)",
    mother_age_yrs       = "Mother's age (years)",
    mother_schooling_yrs = "Mother's schooling (years)",
    asset_index_tertile  = "Household asset index tertile",
    ngo_member           = "Mother is an NGO member",
    improved_toilet      = "Improved toilet facility"
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

# Small helpers so that every number quoted in the prose is pulled from the
# fitted objects rather than typed in by hand.
fmt_ci <- function(tab, term_name, digits = 3) {
  r <- tab[tab$term == term_name, ]
  f <- function(x) formatC(x, format = "f", digits = digits)
  sprintf("%s (95%% CI %s to %s)", f(r$estimate), f(r$conf.low), f(r$conf.high))
}
fmt_p <- function(tab, term_name) {
  p <- tab[tab$term == term_name, ]$p.value
  if (p < 0.001) "p < 0.001" else sprintf("p = %.3f", p)
}
```

# Introduction

The data come from a cross-sectional household survey in rural Bangladesh that
compared two ways of delivering growth monitoring and promotion (GMP): sessions
held at a community clinic ("facility-based") and sessions held in the
community/home ("home-based"). Caregivers of `r nrow(dat_raw)` young children
across six upazilas were interviewed and the children were weighed and measured
(source paper: PLOS ONE, e0324918). Weight-for-height Z-score (WHZ) captures
*acute* undernutrition — thinness relative to length — and is the index most
responsive to recent food shortage and illness, so in principle it is the
outcome most likely to move on a short timescale in response to a programme.

**Aim.** Among rural Bangladeshi infants aged 6-13 months, we examined which of
GMP delivery model, child sex, child age, mother's age, mother's schooling,
household asset tertile, NGO membership and toilet quality are independently
associated with weight-for-height Z-score, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** Cross-sectional survey of `r nrow(dat_raw)` caregiver-child
pairs in six rural upazilas of Bangladesh. The study sampled 60 community-clinic
catchment areas, arranged as matched pairs, one area of each pair receiving
facility-based and the other home-based GMP. The file supplied for this
catalogue contains one row per child.

**Variables.** The outcome is `weight_for_height_z`, the WHO weight-for-height
Z-score in standard-deviation units; more negative values mean thinner children
for their length. Predictors are: GMP delivery model (facility-based =
reference, home-based); child sex (male = reference); child age in months
(continuous, observed range `r sprintf("%.1f", min(dat$child_age_months))` to
`r sprintf("%.1f", max(dat$child_age_months))`); mother's age in years
(continuous); mother's completed schooling in years (continuous); household
asset index tertile (lowest = reference, middle, highest); NGO membership
(no = reference); and improved toilet facility (no = reference).

*Delivery model versus sub-district.* `delivery_model` and `sub_district` are
almost the same variable: only 8 of `r nrow(dat_raw)` records (0.3%) sit in a
sub-district whose other children received the alternative model. Fitting both
together yields variance inflation factors of roughly 96 and 103 and inflates
the delivery-model standard error about nine-fold, with the coefficient
flipping sign. We kept `delivery_model`, the substantive programme exposure the
survey was built around, and excluded `sub_district`. Genuine geographic
differences between the six upazilas are therefore absorbed into the
delivery-model coefficient and cannot be distinguished from it — a point that
matters unusually much for this outcome, because delivery model is the one
predictor that changes status after adjustment.

*Near-constant variables dropped.* `mother_occupation` is 97.5% "housewife" and
`religion_muslim` is 97.8% Muslim; both were left out of the main model and are
shown in a sensitivity fit in the feasibility notes.

*Other Z-scores excluded.* `weight_for_age_z` and `height_for_age_z` are not
predictors. Weight-for-height is computed from the same weight and length
measurements, so regressing it on either is an arithmetic exercise, not an
epidemiological one.

**Statistical analysis.** Linear regression via `lm()`, justified because the
outcome is a continuous, approximately symmetric Z-score. We fitted a
univariable model for each predictor first, then a single multivariable model
containing all of them. Predictors were chosen in advance on subject-matter
grounds, not by significance. Significance threshold 0.05; estimates are
reported with 95% confidence intervals. Assumptions were checked with
`performance::check_model()`. Analysis used `r R.version.string` with tidyverse,
gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped).

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(missing = "no") |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

```{r table-one-numbers}
med_whz <- median(dat_model[[outcome]])
pct_wast <- mean(dat_model[[outcome]] <= -2) * 100
n_severe_whz <- sum(dat_model[[outcome]] < -3)
n_severe_haz <- sum(dat$height_for_age_z < -3, na.rm = TRUE)
```

Median weight-for-height Z-score is `r sprintf("%.2f", med_whz)`, only about
half a standard deviation below the WHO reference median, and
`r sprintf("%.1f", pct_wast)`% of children fall at or below the -2 wasting
threshold. This is a far less shifted distribution than height-for-age in the
same children, which is the expected pattern: chronic deficits accumulate, acute
ones do not.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Weight-for-height Z-score against child age, by growth-monitoring delivery model. Points are individual children; lines are loess smooths with 95% confidence bands; the dashed line marks the -2 wasting threshold."}
ggplot(dat_model, aes(x = child_age_months, y = weight_for_height_z,
                      colour = delivery_model, fill = delivery_model)) +
  geom_point(alpha = 0.13, size = 0.9) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 0.9, alpha = 0.2) +
  geom_hline(yintercept = -2, linetype = "dashed", colour = "grey30") +
  scale_colour_viridis_d(option = "D", end = 0.7, name = "Delivery model") +
  scale_fill_viridis_d(option = "D", end = 0.7, name = "Delivery model") +
  labs(x = "Child's age (months)",
       y = "Weight-for-height Z-score",
       title = "Weight-for-height declines with age in both delivery models")
```

Weight-for-height declines steadily across the narrow 6-13-month window in both
arms, with the home-based curve sitting slightly below the facility-based one
throughout — a small and roughly constant gap rather than a divergence.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv
```

```{r uv-tidy}
uv_tidy <- map_dfr(predictors, function(p) {
  tidy(lm(reformulate(p, outcome), data = dat_model), conf.int = TRUE) |>
    filter(term != "(Intercept)")
})
```

Unadjusted, household wealth is again the strongest correlate: children in the
highest asset tertile score `r fmt_ci(uv_tidy, "asset_index_tertileHighest")`
Z-units above the lowest tertile
(`r fmt_p(uv_tidy, "asset_index_tertileHighest")`). Each additional year of
maternal schooling is associated with
`r fmt_ci(uv_tidy, "mother_schooling_yrs")`
(`r fmt_p(uv_tidy, "mother_schooling_yrs")`) — noticeably weaker than the same
slope for height-for-age, consistent with education acting mainly on chronic
rather than acute deficit. Weight-for-height falls with child age
(`r fmt_ci(uv_tidy, "child_age_months")` per month,
`r fmt_p(uv_tidy, "child_age_months")`) and with mother's age
(`r fmt_ci(uv_tidy, "mother_age_yrs")`, `r fmt_p(uv_tidy, "mother_age_yrs")`).
The programme variable sits exactly on the conventional threshold: children in
home-based areas score `r fmt_ci(uv_tidy, "delivery_modelHome-based")` relative
to facility-based areas (`r fmt_p(uv_tidy, "delivery_modelHome-based")`). Child
sex (`r fmt_p(uv_tidy, "child_sexFemale")`), NGO membership
(`r fmt_p(uv_tidy, "ngo_memberYes")`) and improved toilet
(`r fmt_p(uv_tidy, "improved_toiletYes")`) are not significant crude.

## Why adjustment is needed

**Household asset tertile** is the dominant confounder in this dataset. It
predicts maternal schooling, latrine ownership and NGO membership on one side,
and food security and illness exposure — hence weight-for-height — on the other,
so every socio-economic crude estimate here is partly a wealth estimate.
**Mother's schooling** confounds in the reverse direction: better-educated
mothers live in wealthier households, so the crude asset gradient contains an
education component. **Improved toilet facility** is a third: sanitation is
concentrated in wealthier households (93% coverage overall) and poor sanitation
causes the diarrhoeal illness that produces acute weight loss, so its crude
coefficient cannot be interpreted without holding wealth constant. **Child age**
must be adjusted for on structural grounds, because weight-for-height declines
appreciably even over the seven months observed here. Adjustment matters
particularly for the delivery-model comparison: although the study allocated
delivery models to matched-pair areas, the two arms are not perfectly balanced
on asset tertile (the highest tertile is slightly over-represented in home-based
areas), so the crude programme contrast carries a wealth imbalance with it.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv
```

```{r fit-stats}
s <- summary(fit)
mv_tidy <- tidy(fit, conf.int = TRUE) |> filter(term != "(Intercept)")

tibble(
  Statistic = c("R-squared", "Adjusted R-squared", "Residual SE",
                "Observations", "Model parameters (excl. intercept)"),
  Value = c(sprintf("%.4f", s$r.squared),
            sprintf("%.4f", s$adj.r.squared),
            sprintf("%.3f", s$sigma),
            format(nobs(fit), big.mark = ","),
            as.character(length(coef(fit)) - 1))
) |>
  gt() |>
  tab_header(title = "Model fit")
```

The model explains only `r sprintf("%.1f", 100 * s$r.squared)`% of the variance
in weight-for-height (adjusted R-squared
`r sprintf("%.3f", s$adj.r.squared)`) — the weakest fit of the three Z-score
outcomes, which fits the interpretation that acute thinness is driven by
short-term shocks that this survey did not measure. Holding the other predictors
constant, children in the highest asset tertile score
`r fmt_ci(mv_tidy, "asset_index_tertileHighest")` above the lowest
(`r fmt_p(mv_tidy, "asset_index_tertileHighest")`); each extra year of maternal
schooling adds `r fmt_ci(mv_tidy, "mother_schooling_yrs")`
(`r fmt_p(mv_tidy, "mother_schooling_yrs")`); and each additional month of age is
associated with `r fmt_ci(mv_tidy, "child_age_months")`
(`r fmt_p(mv_tidy, "child_age_months")`). Adjusted for household circumstances,
children in home-based GMP areas score
`r fmt_ci(mv_tidy, "delivery_modelHome-based")` relative to facility-based areas
(`r fmt_p(mv_tidy, "delivery_modelHome-based")`). That result should be treated
with considerable caution — see the limitations on clustering below.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

The interesting movement here runs in both directions. **One predictor moved
*away* from the null and gained significance**: the home-based versus
facility-based difference went from `r fmt_ci(uv_tidy, "delivery_modelHome-based")`
(`r fmt_p(uv_tidy, "delivery_modelHome-based")`) to
`r fmt_ci(mv_tidy, "delivery_modelHome-based")`
(`r fmt_p(mv_tidy, "delivery_modelHome-based")`). This is negative confounding:
home-based areas contain slightly wealthier households, which was masking a small
deficit, and holding wealth constant unmasks it. **Two predictors moved toward
the null and lost significance.** Maternal schooling attenuated from
`r fmt_ci(uv_tidy, "mother_schooling_yrs")` to
`r fmt_ci(mv_tidy, "mother_schooling_yrs")`, a 30% reduction (still significant);
mother's age went from `r fmt_ci(uv_tidy, "mother_age_yrs")`
(`r fmt_p(uv_tidy, "mother_age_yrs")`) to
`r fmt_ci(mv_tidy, "mother_age_yrs")` (`r fmt_p(mv_tidy, "mother_age_yrs")`); and
the middle-versus-lowest asset contrast lost significance,
`r fmt_ci(uv_tidy, "asset_index_tertileMiddle")`
(`r fmt_p(uv_tidy, "asset_index_tertileMiddle")`) to
`r fmt_ci(mv_tidy, "asset_index_tertileMiddle")`
(`r fmt_p(mv_tidy, "asset_index_tertileMiddle")`). **Improved toilet collapsed to
nothing**, from `r fmt_ci(uv_tidy, "improved_toiletYes")` to
`r fmt_ci(mv_tidy, "improved_toiletYes")` — its crude association was entirely a
wealth proxy. **Child sex and child age barely moved**
(age `r fmt_ci(uv_tidy, "child_age_months")` to
`r fmt_ci(mv_tidy, "child_age_months")`, sex
`r fmt_ci(uv_tidy, "child_sexFemale")` to
`r fmt_ci(mv_tidy, "child_sexFemale")`), as expected for variables not determined
by household circumstances. None of this is collinearity: all variance inflation
factors are under 1.3.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r diag-extra}
vifs <- check_collinearity(fit)
as_tibble(vifs) |>
  select(Term, VIF) |>
  mutate(VIF = round(VIF, 2)) |>
  gt() |>
  tab_header(title = "Variance inflation factors")
```

The *posterior predictive check* reproduces the observed weight-for-height
distribution well. *Linearity* is satisfied — the residual smooth is flat, so
linear terms for age and schooling are adequate. *Homogeneity of variance* holds
across the fitted range. The *normality of residuals* Q-Q plot is close to the
line through the body of the distribution with modest departure in both tails,
partly reflecting the severe-malnutrition exclusion truncating the left tail and
a few implausibly high Z-scores (above +4) on the right; neither threatens
inference at this sample size. No observation crosses the Cook's-distance
contours in the *influential observations* panel. *Collinearity* is fine as
specified, but only because `sub_district` was deliberately excluded.

# Discussion

Among rural Bangladeshi infants aged 6-13 months, acute undernutrition as
measured by weight-for-height is only weakly patterned by the household
characteristics recorded in this survey. Household wealth and maternal schooling
retain independent associations after mutual adjustment; sanitation, maternal age
and NGO membership do not. The one result that changes status on adjustment is
the programme comparison: home-based GMP areas show a small deficit in
weight-for-height once household wealth is held constant. Given the design, this
should be read as a hypothesis and not as evidence that home-based delivery harms
children.

**Limitations.**

*Clustering cannot be modelled, and this outcome is where that bites hardest.*
Delivery model is an **area-level** exposure assigned to whole upazilas — six of
them — with 60 community-clinic catchment areas sampled as matched pairs.
Children in the same catchment share fieldworkers, scales, clinics and food
markets, so their weight-for-height values are correlated. The file carries no
clinic or cluster identifier, so no random effect or cluster-robust standard
error is possible. The nominally significant adjusted delivery-model result
(`r fmt_p(mv_tidy, "delivery_modelHome-based")`) rests on a standard error
computed as though `r nrow(dat_model)` children were independent when the
effective number of independent units is closer to 60, or arguably six. Even a
modest design effect would push that interval back across zero. `sub_district`
is not a usable substitute cluster variable because it is collinear with the
exposure. **This single result should not be reported as a finding.**

*The outcome distribution is truncated by design.* The source paper states that
children with severe stunting, wasting or underweight were excluded on ethical
grounds so they could be referred for treatment. Yet the shipped file contains
`r n_severe_whz` children with weight-for-height below -3 and
`r n_severe_haz` with height-for-age below -3, so the exclusion was incomplete or
applied on some other criterion. The lower tail of weight-for-height has
therefore been interfered with: the `r sprintf("%.1f", pct_wast)`% wasting
prevalence observed here understates the true burden, and coefficients describing
movement into the thinnest part of the distribution are attenuated.

*Age definition is inconsistent in the source.* The paper's abstract says
children "under one year", its methods say 6-23 months, and the data run
`r sprintf("%.1f", min(dat$child_age_months))` to
`r sprintf("%.1f", max(dat$child_age_months))` months. We describe the observed
range only.

*Design and residual confounding.* Cross-sectional data give associations, not
effects. Weight-for-height is the index most sensitive to *recent* events —
diarrhoea in the last fortnight, a seasonal food gap, a feeding interruption —
and none of these were recorded. Season of measurement is also unavailable, which
matters because wasting in Bangladesh is strongly seasonal. Adjusted R-squared of
`r sprintf("%.3f", s$adj.r.squared)` is the honest consequence.

*Sparse and dropped variables.* `mother_occupation` and `religion_muslim` were
excluded as near-constant.

**Future work.** Repeated measurement across seasons, with cluster identifiers
and recent-illness histories, would be needed before any statement about GMP
delivery and acute undernutrition could be taken seriously.

# Feasibility notes

*Not part of a student report.*

**Verdict: usable but the weakest of the three continuous outcomes, and it
contains a booby trap.** The model fits without complaint, but two things make it
harder to supervise than height-for-age or weight-for-age.

1. **The delivery-model result is a trap, not a finding.** It is the only place
   in the whole dataset where the programme exposure crosses p = 0.05, and it
   does so only after adjustment
   (`r fmt_p(uv_tidy, "delivery_modelHome-based")` crude to
   `r fmt_p(mv_tidy, "delivery_modelHome-based")` adjusted). A student will seize
   on it and write "home-based growth monitoring is associated with worse acute
   nutritional status". Given 60 unmodellable clusters and six upazilas, that
   p-value is almost certainly an artefact of ignoring clustering. The brief must
   pre-warn markers and students alike.

2. **Weakest fit of the six outcomes.** Adjusted R-squared
   `r sprintf("%.3f", s$adj.r.squared)`, and only three predictors reach
   significance. There is a real risk of a "nothing happened" report.

3. **`delivery_model` versus `sub_district`.** 99.7% overlap (8 discordant
   records out of `r nrow(dat_raw)`); together they give VIFs of roughly 96 and
   103. Must be chosen between.

4. **Do not put other anthropometry on the right-hand side.**
   `weight_for_age_z` shares the same weight measurement as this outcome and
   `height_for_age_z`/`stunted` share the same length measurement; any of them as
   a predictor turns the model into an identity check. A derived `wasting`
   indicator is exactly `weight_for_height_z <= -2` and must never appear as a
   predictor here.

5. **Near-constant predictors.** Sensitivity fit adding them back:

```{r sens-near-constant}
dat_sens <- dat |>
  mutate(religion_muslim = factor(religion_muslim, levels = c(0, 1),
                                  labels = c("Non-Muslim", "Muslim")),
         mother_occupation = factor(mother_occupation,
                                    levels = c("housewife", "others"),
                                    labels = c("Housewife", "Other"))) |>
  select(all_of(c(outcome, predictors)), mother_occupation, religion_muslim) |>
  drop_na()

fit_sens <- lm(reformulate(c(predictors, "mother_occupation", "religion_muslim"),
                           outcome), data = dat_sens)

tidy(fit_sens, conf.int = TRUE) |>
  filter(str_detect(term, "mother_occupation|religion_muslim")) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  select(term, estimate, conf.low, conf.high, p.value) |>
  gt() |>
  tab_header(title = "Near-constant predictors, added back",
             subtitle = "n = 77 non-housewives, 66 non-Muslim households")
```

   Wide, unstable intervals resting on tiny comparison groups. Shown only to
   justify dropping them.

6. **Overlap with the `wasting` analysis.** Same measurement, dichotomised. The
   continuous version is strictly more powerful; a student choosing `wasting`
   instead should know they are giving up power for interpretability.
