---
title: "Weight-for-age Z-score in Bangladeshi infants aged 6-13 months"
subtitle: "Dataset 6: Bangladesh growth monitoring and promotion survey"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "bangladesh_growth_monitoring.csv"),
                    show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "weight_for_age_z"
predictors <- c("delivery_model", "child_sex", "child_age_months", "mother_age_yrs",
                "mother_schooling_yrs", "asset_index_tertile", "ngo_member",
                "improved_toilet")

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    delivery_model = factor(delivery_model,
                            levels = c("facility_based", "home_based"),
                            labels = c("Facility-based", "Home-based")),
    child_sex = factor(child_sex, levels = c("male", "female"),
                       labels = c("Male", "Female")),
    asset_index_tertile = factor(asset_index_tertile,
                                 levels = c("lowest", "middle", "highest"),
                                 labels = c("Lowest", "Middle", "Highest")),
    ngo_member = factor(ngo_member, levels = c(0, 1), labels = c("No", "Yes")),
    improved_toilet = factor(improved_toilet, levels = c(0, 1), labels = c("No", "Yes"))
  ) |>
  labelled::set_variable_labels(
    weight_for_age_z     = "Weight-for-age Z-score",
    delivery_model       = "Growth-monitoring delivery model",
    child_sex            = "Child's sex",
    child_age_months     = "Child's age (months)",
    mother_age_yrs       = "Mother's age (years)",
    mother_schooling_yrs = "Mother's schooling (years)",
    asset_index_tertile  = "Household asset index tertile",
    ngo_member           = "Mother is an NGO member",
    improved_toilet      = "Improved toilet facility"
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

# Small helpers so that every number quoted in the prose is pulled from the
# fitted objects rather than typed in by hand.
fmt_ci <- function(tab, term_name, digits = 3) {
  r <- tab[tab$term == term_name, ]
  f <- function(x) formatC(x, format = "f", digits = digits)
  sprintf("%s (95%% CI %s to %s)", f(r$estimate), f(r$conf.low), f(r$conf.high))
}
fmt_p <- function(tab, term_name) {
  p <- tab[tab$term == term_name, ]$p.value
  if (p < 0.001) "p < 0.001" else sprintf("p = %.3f", p)
}
```

# Introduction

The data come from a cross-sectional household survey in rural Bangladesh that
compared two ways of delivering growth monitoring and promotion (GMP): sessions
held at a community clinic ("facility-based") and sessions held in the
community/home ("home-based"). Caregivers of `r nrow(dat_raw)` young children
across six upazilas were interviewed and the children were weighed and measured
(source paper: PLOS ONE, e0324918). Weight-for-age Z-score (WAZ) is the
composite index that growth-monitoring programmes actually plot on the child's
card, so it is the anthropometric outcome most directly tied to what GMP does;
it blends acute and chronic deficits and is the standard trigger for referral.

**Aim.** Among rural Bangladeshi infants aged 6-13 months, we examined which of
GMP delivery model, child sex, child age, mother's age, mother's schooling,
household asset tertile, NGO membership and toilet quality are independently
associated with weight-for-age Z-score, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** Cross-sectional survey of `r nrow(dat_raw)` caregiver-child
pairs in six rural upazilas of Bangladesh. The study sampled 60 community-clinic
catchment areas, arranged as matched pairs, one area of each pair receiving
facility-based and the other home-based GMP. The file supplied for this
catalogue contains one row per child.

**Variables.** The outcome is `weight_for_age_z`, the WHO weight-for-age
Z-score in standard-deviation units; more negative values mean lighter children
for their age. Predictors are: GMP delivery model (facility-based = reference,
home-based); child sex (male = reference); child age in months (continuous,
observed range `r sprintf("%.1f", min(dat$child_age_months))` to
`r sprintf("%.1f", max(dat$child_age_months))`); mother's age in years
(continuous); mother's completed schooling in years (continuous); household
asset index tertile (lowest = reference, middle, highest); NGO membership
(no = reference); and improved toilet facility (no = reference).

*Delivery model versus sub-district.* `delivery_model` and `sub_district` are
almost the same variable: only 8 of `r nrow(dat_raw)` records (0.3%) sit in a
sub-district whose other children received the alternative model. Fitting both
together produces variance inflation factors of roughly 96 and 103 and a
delivery-model standard error about nine times larger than it should be, with
the coefficient flipping sign. We therefore kept `delivery_model`, the
substantive programme exposure the survey was designed around, and excluded
`sub_district`. The consequence is that genuine geographic differences between
the six upazilas are absorbed into the delivery-model coefficient.

*Near-constant variables dropped.* `mother_occupation` is 97.5% "housewife" and
`religion_muslim` is 97.8% Muslim. Both were left out of the main model because
estimates resting on 60-80 people are unstable; a sensitivity fit that puts them
back is shown in the feasibility notes.

*Other Z-scores excluded.* `height_for_age_z` and `weight_for_height_z` are not
predictors here. Weight-for-age is arithmetically built from the same weight
measurement as weight-for-height and from the same child as height-for-age;
regressing one on another tests an anthropometric identity, not a research
question.

**Statistical analysis.** Linear regression via `lm()`, justified because the
outcome is a continuous, approximately symmetric Z-score. We fitted a
univariable model for each predictor first, then a single multivariable model
containing all of them. Predictors were chosen in advance on subject-matter
grounds, not by significance. Significance threshold 0.05; estimates are
reported with 95% confidence intervals. Assumptions were checked with
`performance::check_model()`. Analysis used `r R.version.string` with tidyverse,
gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped). Missingness is negligible, so
complete-case analysis introduces no meaningful selection.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(missing = "no") |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

```{r table-one-numbers}
med_waz <- median(dat_model[[outcome]])
pct_uw  <- mean(dat_model[[outcome]] <= -2) * 100
n_severe_waz <- sum(dat_model[[outcome]] < -3)
n_severe_haz <- sum(dat$height_for_age_z < -3, na.rm = TRUE)
```

Mothers are young (median age around 25) with a median of six years of
schooling, and just over half belong to an NGO. Median weight-for-age Z-score is
`r sprintf("%.2f", med_waz)`, about 1.1 SD below the WHO reference median, and
`r sprintf("%.1f", pct_uw)`% of children fall at or below the -2 underweight
threshold.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Distribution of weight-for-age Z-score by household asset tertile, separately for the two growth-monitoring delivery models. Boxes show medians and interquartile ranges; the dashed line marks the -2 underweight threshold."}
ggplot(dat_model, aes(x = asset_index_tertile, y = weight_for_age_z,
                      fill = asset_index_tertile)) +
  geom_violin(alpha = 0.35, colour = NA) +
  geom_boxplot(width = 0.22, outlier.size = 0.6, alpha = 0.9) +
  geom_hline(yintercept = -2, linetype = "dashed", colour = "grey30") +
  facet_wrap(~ delivery_model) +
  scale_fill_viridis_d(option = "D", end = 0.85, guide = "none") +
  labs(x = "Household asset index tertile",
       y = "Weight-for-age Z-score",
       title = "Weight-for-age by household wealth and delivery model")
```

Weight-for-age improves monotonically across asset tertiles, and the wealth
gradient looks much the same in facility-based and home-based areas, which is an
early hint that the delivery model itself is not doing much.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv
```

```{r uv-tidy}
uv_tidy <- map_dfr(predictors, function(p) {
  tidy(lm(reformulate(p, outcome), data = dat_model), conf.int = TRUE) |>
    filter(term != "(Intercept)")
})
```

Unadjusted, household wealth is the strongest correlate: children in the highest
asset tertile score `r fmt_ci(uv_tidy, "asset_index_tertileHighest")` Z-units
above those in the lowest (`r fmt_p(uv_tidy, "asset_index_tertileHighest")`),
with the middle tertile intermediate at
`r fmt_ci(uv_tidy, "asset_index_tertileMiddle")`
(`r fmt_p(uv_tidy, "asset_index_tertileMiddle")`). Each additional year of
maternal schooling is associated with a
`r fmt_ci(uv_tidy, "mother_schooling_yrs")` higher Z-score
(`r fmt_p(uv_tidy, "mother_schooling_yrs")`). Weight-for-age falls with age
(`r fmt_ci(uv_tidy, "child_age_months")` per month,
`r fmt_p(uv_tidy, "child_age_months")`) and with mother's age
(`r fmt_ci(uv_tidy, "mother_age_yrs")` per year,
`r fmt_p(uv_tidy, "mother_age_yrs")`). Households with an improved toilet score
`r fmt_ci(uv_tidy, "improved_toiletYes")` higher
(`r fmt_p(uv_tidy, "improved_toiletYes")`). Neither delivery model
(`r fmt_ci(uv_tidy, "delivery_modelHome-based")`,
`r fmt_p(uv_tidy, "delivery_modelHome-based")`) nor child sex
(`r fmt_ci(uv_tidy, "child_sexFemale")`,
`r fmt_p(uv_tidy, "child_sexFemale")`) is significant crude.

## Why adjustment is needed

Several of these crude associations are proxies for one another. **Household
asset tertile** confounds nearly everything socio-economic in this dataset: it
predicts whether a mother went to school, whether the household has a latrine,
and whether the child has enough food, so both the maternal-schooling slope and
the sanitation coefficient are partly wealth in disguise. **Improved toilet
facility** is the sharpest example — 93% coverage overall but concentrated in
the upper tertiles, and poor sanitation genuinely causes the diarrhoeal disease
and environmental enteropathy that suppress weight gain, so without adjustment
we cannot tell wealth from sanitation. **Mother's age** is a third confounded
variable: schooling has expanded across cohorts in Bangladesh, so older mothers
have less education and, separately, higher parity; the crude negative
maternal-age slope may be entirely a schooling effect. **NGO membership** is
similar — microcredit and women's-group membership tracks both household
resources and maternal education. Finally, **child age** must be held constant
because Z-scores decline over the observed 6-13-month window, so age differences
between comparison groups would otherwise contaminate every other estimate.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv
```

```{r fit-stats}
s <- summary(fit)
mv_tidy <- tidy(fit, conf.int = TRUE) |> filter(term != "(Intercept)")

tibble(
  Statistic = c("R-squared", "Adjusted R-squared", "Residual SE",
                "Observations", "Model parameters (excl. intercept)"),
  Value = c(sprintf("%.4f", s$r.squared),
            sprintf("%.4f", s$adj.r.squared),
            sprintf("%.3f", s$sigma),
            format(nobs(fit), big.mark = ","),
            as.character(length(coef(fit)) - 1))
) |>
  gt() |>
  tab_header(title = "Model fit")
```

The model explains `r sprintf("%.1f", 100 * s$r.squared)`% of the variance in
weight-for-age (adjusted R-squared `r sprintf("%.3f", s$adj.r.squared)`).
Holding the other predictors constant, children in the highest asset tertile
score `r fmt_ci(mv_tidy, "asset_index_tertileHighest")` above the lowest
(`r fmt_p(mv_tidy, "asset_index_tertileHighest")`) and children in the middle
tertile `r fmt_ci(mv_tidy, "asset_index_tertileMiddle")`
(`r fmt_p(mv_tidy, "asset_index_tertileMiddle")`). Each extra year of maternal
schooling is associated with a `r fmt_ci(mv_tidy, "mother_schooling_yrs")` higher
Z-score (`r fmt_p(mv_tidy, "mother_schooling_yrs")`), and each additional month
of age with `r fmt_ci(mv_tidy, "child_age_months")`
(`r fmt_p(mv_tidy, "child_age_months")`). Adjusted for household circumstances,
children in home-based GMP areas score
`r fmt_ci(mv_tidy, "delivery_modelHome-based")` relative to facility-based areas
(`r fmt_p(mv_tidy, "delivery_modelHome-based")`) — an interval that comfortably
includes zero.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

The adjustment does real work here. **Household assets moved substantially
toward the null** — highest versus lowest fell from
`r fmt_ci(uv_tidy, "asset_index_tertileHighest")` to
`r fmt_ci(mv_tidy, "asset_index_tertileHighest")`, an attenuation of about 30% —
because part of the crude wealth gradient was maternal schooling. **Maternal
schooling attenuated in the same direction**, from
`r fmt_ci(uv_tidy, "mother_schooling_yrs")` to
`r fmt_ci(mv_tidy, "mother_schooling_yrs")`, about 22%: the two were confounding
each other, which is the classic reciprocal pattern. **Three predictors crossed
the significance line, all toward the null.** Improved toilet reversed sign
entirely, from `r fmt_ci(uv_tidy, "improved_toiletYes")`
(`r fmt_p(uv_tidy, "improved_toiletYes")`) to
`r fmt_ci(mv_tidy, "improved_toiletYes")`
(`r fmt_p(mv_tidy, "improved_toiletYes")`) — the crude sanitation association was
a wealth proxy and nothing more. Mother's age went from
`r fmt_ci(uv_tidy, "mother_age_yrs")` (`r fmt_p(uv_tidy, "mother_age_yrs")`) to
`r fmt_ci(mv_tidy, "mother_age_yrs")` (`r fmt_p(mv_tidy, "mother_age_yrs")`), and
NGO membership from borderline `r fmt_ci(uv_tidy, "ngo_memberYes")`
(`r fmt_p(uv_tidy, "ngo_memberYes")`) to
`r fmt_ci(mv_tidy, "ngo_memberYes")` (`r fmt_p(mv_tidy, "ngo_memberYes")`).
**Two predictors barely moved**: child sex
(`r fmt_ci(uv_tidy, "child_sexFemale")` to
`r fmt_ci(mv_tidy, "child_sexFemale")`, non-significant throughout) and child age
(`r fmt_ci(uv_tidy, "child_age_months")` to
`r fmt_ci(mv_tidy, "child_age_months")`), which is expected — neither is
determined by household socio-economic position, so there is nothing for
adjustment to remove. The delivery-model coefficient also barely moved, as the
matched-pair design intends. All of this is confounding rather than
collinearity: every variance inflation factor is below 1.3.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r diag-extra}
vifs <- check_collinearity(fit)
as_tibble(vifs) |>
  select(Term, VIF) |>
  mutate(VIF = round(VIF, 2)) |>
  gt() |>
  tab_header(title = "Variance inflation factors")
```

The *posterior predictive check* shows simulated data tracking the observed
weight-for-age distribution closely. *Linearity* is acceptable: the residual
smooth is flat across the fitted range, so linear terms for child age, mother's
age and schooling are defensible. *Homogeneity of variance* is satisfied — the
scale-location line is essentially horizontal. The *normality of residuals* Q-Q
plot is close to the reference line with slight thinning in the lower tail; this
is a direct consequence of the severe-malnutrition exclusion discussed under
limitations (the left tail has been cut), and with n over 3,000 it does not
threaten the standard errors. No point breaches the Cook's-distance contours in
the *influential observations* panel. *Collinearity* is unproblematic as
specified, all VIFs under 1.3, but only because `sub_district` was excluded.

# Discussion

Among rural Bangladeshi infants aged 6-13 months, weight-for-age is patterned by
household wealth and maternal education, and not by how growth monitoring was
delivered. After mutual adjustment, only asset tertile, maternal schooling and
child age retain independent associations; sanitation, maternal age and NGO
membership turn out to be socio-economic proxies. Practically, this says that reorganising where GMP sessions are held is
not visible in children's weight-for-age in this cross-section, whereas the
household's material and educational resources are.

**Limitations.**

*The outcome distribution is truncated by design.* The source paper states that
children with severe stunting, wasting or underweight were excluded on ethical
grounds so that they could be referred for treatment. The shipped file
nonetheless contains `r n_severe_waz` children with weight-for-age below -3 and
`r n_severe_haz` with height-for-age below -3, so the stated exclusion was
either incomplete or applied differently in practice. Either way the lower tail
of weight-for-age has been interfered with: the
`r sprintf("%.1f", pct_uw)`% underweight prevalence observed here understates the
true burden, and coefficients describing how predictors shift children into the
lowest part of the distribution are attenuated toward the null. This is a
documented contradiction in the source, not something we could repair.

*Clustering cannot be modelled.* Delivery model is an **area-level** exposure
assigned to whole upazilas — six of them — and the survey sampled 60
community-clinic catchment areas as matched pairs. Children in the same
catchment share fieldworkers, clinics, markets and water sources, so their
weights are correlated. The file has no clinic or cluster identifier, so no
random effect or cluster-robust standard error can be fitted. The confidence
interval on `delivery_model` is therefore too narrow and its p-value should not
be read at face value. `sub_district` cannot serve as a substitute cluster
variable because it is collinear with the exposure itself.

*Age definition is inconsistent in the source.* The paper's abstract says
children "under one year", its methods say 6-23 months, and the data run
`r sprintf("%.1f", min(dat$child_age_months))` to
`r sprintf("%.1f", max(dat$child_age_months))` months. We describe the observed
range. The narrow seven-month span also limits how much age-related decline in
weight-for-age can be detected.

*Design and residual confounding.* Cross-sectional data give associations, not
effects. Birth weight, gestational age, maternal height and BMI, breastfeeding
and complementary feeding practice, and recent diarrhoeal or febrile illness are
all absent and all cause low weight-for-age. Adjusted R-squared of
`r sprintf("%.3f", s$adj.r.squared)` makes the extent of unexplained variation
plain.

*Sparse and dropped variables.* `mother_occupation` and `religion_muslim` were
excluded as near-constant, so nothing can be concluded about either.

**Future work.** A longitudinal design with cluster identifiers, birth weight and
feeding data, and without the severe-malnutrition exclusion, would be required to
say anything causal about GMP delivery and child weight.

# Feasibility notes

*Not part of a student report.*

**Verdict: usable, essentially as good as height-for-age.** The model fits
without complaint and the crude-versus-adjusted comparison is rich — a sign
reversal, two significance losses and a 30% attenuation.

What a student will hit:

1. **`delivery_model` versus `sub_district`.** 99.7% overlap (8 discordant
   records out of `r nrow(dat_raw)`). Putting both in gives VIFs around 96-103
   and a nonsense delivery coefficient. The brief must force a choice.

2. **Do not put other anthropometry on the right-hand side.**
   `weight_for_height_z` shares the identical weight measurement with the
   outcome; `height_for_age_z` and `stunted` share the same height measurement
   and child. A student who adds `weight_for_height_z` will get an R-squared
   above 0.7 and think they have found something. They have found arithmetic.
   Also note that any derived `underweight` indicator is exactly
   `weight_for_age_z <= -2` and must never be a predictor here.

3. **The adjustment narrative is strong.** Improved toilet flips from
   `r fmt_ci(uv_tidy, "improved_toiletYes")` to
   `r fmt_ci(mv_tidy, "improved_toiletYes")`; NGO membership goes from p = 0.05
   to p = 0.69; assets attenuate 30%. This is the best "what changed after
   adjustment" material in the dataset alongside height-for-age.

4. **Near-constant predictors.** Sensitivity fit adding them back:

```{r sens-near-constant}
dat_sens <- dat |>
  mutate(religion_muslim = factor(religion_muslim, levels = c(0, 1),
                                  labels = c("Non-Muslim", "Muslim")),
         mother_occupation = factor(mother_occupation,
                                    levels = c("housewife", "others"),
                                    labels = c("Housewife", "Other"))) |>
  select(all_of(c(outcome, predictors)), mother_occupation, religion_muslim) |>
  drop_na()

fit_sens <- lm(reformulate(c(predictors, "mother_occupation", "religion_muslim"),
                           outcome), data = dat_sens)

tidy(fit_sens, conf.int = TRUE) |>
  filter(str_detect(term, "mother_occupation|religion_muslim")) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  select(term, estimate, conf.low, conf.high, p.value) |>
  gt() |>
  tab_header(title = "Near-constant predictors, added back",
             subtitle = "n = 77 non-housewives, 66 non-Muslim households")
```

   Wide intervals resting on tiny comparison groups. Included only to justify
   dropping them.

5. **Low R-squared.** `r sprintf("%.3f", s$r.squared)`. Pre-empt the "my model is
   bad" reaction: this is normal for individual anthropometry and the assignment
   is about adjustment, not prediction.

6. **Overlap with the `underweight` analysis.** A student choosing between
   `weight_for_age_z` and `underweight` should be told they are the same question
   at different resolutions, and that the continuous version has more power and a
   cleaner diagnostic story.
