---
title: "Ultrasound-diagnosed fatty liver in a Japanese health-check cohort"
subtitle: "Dataset 9: Liver enzymes and arterial stiffness in a Japanese health-check cohort"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "ggt_arterial_stiffness.csv"),
                    show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names. The catalogue suggests
# 18 predictors; 12 are carried forward. The cut is justified in Methods.
outcome    <- "fatty_liver"
predictors <- c("age_yrs", "sex", "bmi", "systolic_bp", "smoking_status",
                "alcohol_g_per_week", "regular_exercise", "fasting_glucose",
                "triglycerides", "hdl_cholesterol", "egfr", "log2_ggt")

# The full catalogue list, used once to document the collinearity that motivated
# the cut. No inference is drawn from that model.
predictors_all <- c("age_yrs", "sex", "bmi", "systolic_bp", "diastolic_bp",
                    "smoking_status", "alcohol_g_per_week", "regular_exercise",
                    "fasting_glucose", "uric_acid", "total_cholesterol",
                    "hdl_cholesterol", "ldl_cholesterol", "triglycerides",
                    "egfr", "log2_ggt", "alt", "ast")

lab <- function(x, l) { attr(x, "label") <- l; x }

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex = lab(factor(sex, levels = c("male", "female"), labels = c("Male", "Female")),
              "Sex"),
    smoking_status = lab(factor(smoking_status,
                                levels = c("never", "former", "current"),
                                labels = c("Never", "Former", "Current")),
                         "Smoking status"),
    regular_exercise = lab(factor(regular_exercise, levels = c(0, 1),
                                  labels = c("No", "Yes")),
                           "Takes regular exercise"),
    age_yrs            = lab(age_yrs, "Age (years)"),
    bmi                = lab(bmi, "BMI (kg/m2)"),
    systolic_bp        = lab(systolic_bp, "Systolic BP (mmHg)"),
    diastolic_bp       = lab(diastolic_bp, "Diastolic BP (mmHg)"),
    alcohol_g_per_week = lab(alcohol_g_per_week, "Alcohol (g/week)"),
    fasting_glucose    = lab(fasting_glucose, "Fasting glucose (mg/dL)"),
    uric_acid          = lab(uric_acid, "Uric acid (mg/dL)"),
    total_cholesterol  = lab(total_cholesterol, "Total cholesterol (mg/dL)"),
    triglycerides      = lab(triglycerides, "Triglycerides (mg/dL)"),
    hdl_cholesterol    = lab(hdl_cholesterol, "HDL cholesterol (mg/dL)"),
    ldl_cholesterol    = lab(ldl_cholesterol, "LDL cholesterol (mg/dL)"),
    egfr               = lab(egfr, "eGFR (mL/min/1.73m2)"),
    log2_ggt           = lab(log2_ggt, "Gamma-GT (log2 U/L)"),
    alt                = lab(alt, "ALT (U/L)"),
    ast                = lab(ast, "AST (U/L)"),
    fatty_liver        = lab(fatty_liver, "Fatty liver on ultrasound")
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)
n_events  <- sum(dat_model[[outcome]])

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
pv  <- function(p) ifelse(p < 0.001, "<0.001", paste0("= ", fmt(p, 3)))
```

```{r collinearity-evidence}
# Two documented redundancies in the suggested predictor list, quantified once.
bp_r <- cor(dat$systolic_bp, dat$diastolic_bp)

# Friedewald: total cholesterol ~ LDL + HDL + triglycerides/5.
fw  <- lm(total_cholesterol ~ ldl_cholesterol + hdl_cholesterol + triglycerides,
          data = dat)
fw_r2   <- summary(fw)$r.squared
fw_coef <- coef(fw)

dat_all <- dat |> select(all_of(c(outcome, predictors_all))) |> drop_na()
fit_all <- glm(reformulate(predictors_all, outcome), data = dat_all,
               family = binomial)
cc_all  <- as.data.frame(check_collinearity(fit_all))
vall <- function(term, d = 1) fmt(cc_all$VIF[cc_all$Term == term], d)

enz_r <- cor(dat$alt, dat$ast)
```

# Introduction

A general health-check programme in Japan enrolled `r nrow(dat)` men and women
aged `r min(dat$age_yrs)` to `r max(dat$age_yrs)` years and combined abdominal
ultrasound, a fasting blood panel and automated vascular measurements in a single
visit; the cohort is reported in *BMJ Open* 2014;4:e005413. Fatty liver was
present on ultrasound in `r sum(dat_raw$fatty_liver, na.rm = TRUE)`
(`r fmt(100 * mean(dat_raw$fatty_liver, na.rm = TRUE), 1)`%) of those scanned.
Non-alcoholic fatty liver disease is now the commonest chronic liver condition in
high-income countries and is a marker of insulin resistance as much as of liver
injury, so knowing which routinely measured characteristics identify it at a
health check is directly useful for triage.

**Aim.** Among these health-check attendees, we examined which of age, sex, body
mass index, systolic blood pressure, smoking status, alcohol intake, regular
exercise, fasting glucose, triglycerides, HDL cholesterol, kidney function and
log2 gamma-GT are independently associated with ultrasound-diagnosed fatty liver,
mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** A cross-sectional Japanese health-check cohort of `r nrow(dat)`
adults attending a voluntary medical check-up programme that included abdominal
ultrasound, a fasting blood panel and automated measurement of pulse-wave
velocity and the ankle-brachial index.

**Variables.** The outcome is `fatty_liver`, a binary indicator where the **event
is fatty liver visible on abdominal ultrasound**
(`r sum(dat_raw$fatty_liver, na.rm = TRUE)` events among the
`r sum(!is.na(dat_raw$fatty_liver))` participants scanned). Predictors are age
(years), sex (reference Male, the larger group at `r sum(dat$sex == "Male")` of
`r nrow(dat)`), BMI (kg/m2), systolic blood pressure (mmHg), smoking status
(three levels, reference Never, the largest group at
`r sum(dat$smoking_status == "Never")`), alcohol intake (g/week), regular
exercise (reference No, the unexposed and larger group), fasting glucose (mg/dL),
triglycerides (mg/dL), HDL cholesterol (mg/dL), estimated glomerular filtration
rate (mL/min/1.73m2) and gamma-GT on a log2 scale, so that a one-unit increase is
a doubling of GGT.

**The predictor set was cut from 18 to `r length(predictors)` before any model
was fitted.** The events budget alone would not have forced this: with
`r n_events` events, the conventional ten-events-per-parameter rule supports
about `r floor(n_events / 10)` parameters, and all 18 predictors would cost only
`r length(coef(fit_all)) - 1L`. The cut was made instead for **redundancy** and
**interpretability**. The suggested list contains three sets of near-duplicates.
Systolic and diastolic pressure correlate at r = `r fmt(bp_r, 2)` (VIFs
`r vall("systolic_bp")` and `r vall("diastolic_bp")` with both in the model). The
four lipid measures satisfy the Friedewald identity
total = LDL + HDL + triglycerides/5 to within rounding: regressing total
cholesterol on the other three gives R-squared `r fmt(fw_r2, 3)` with
coefficients `r fmt(fw_coef["ldl_cholesterol"], 2)`,
`r fmt(fw_coef["hdl_cholesterol"], 2)` and `r fmt(fw_coef["triglycerides"], 2)`,
and with all four in the model the VIFs reach `r vall("total_cholesterol")`
(total cholesterol) and `r vall("ldl_cholesterol")` (LDL). ALT, AST and gamma-GT
are the third set (ALT-AST r = `r fmt(enz_r, 2)`). We therefore kept systolic
pressure in place of both pressures; HDL cholesterol and triglycerides, the two
metabolic-syndrome lipid components, in place of all four lipids; and log2
gamma-GT, the enzyme of interest in the source paper, in place of all three
enzymes. Uric acid was dropped because it duplicates information already carried
by adiposity and kidney function (r = `r fmt(cor(dat$uric_acid, dat$egfr), 2)`
with eGFR, r = `r fmt(cor(dat$uric_acid, dat$bmi), 2)` with BMI). Beyond the
statistics, a table of 18 mutually adjusted odds ratios cannot be read as an
answer to "which predictors survive adjustment", which is the question here.

**Statistical analysis.** `glm(family = binomial)`, justified because the outcome
is binary. We fitted a univariable model for each predictor first, then a single
multivariable model containing all of them. Predictors were chosen in advance on
subject-matter grounds, not by significance. Coefficients are exponentiated and
reported as odds ratios with 95% confidence intervals; significance threshold
0.05. Model adequacy was assessed with `performance::check_model()`, focusing on
binned residuals, collinearity, influential observations and events per
parameter. Analysis used `r R.version.string` with tidyverse, gtsummary and
performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped), leaving **`r n_events` events**. The
outcome itself is missing for `r sum(is.na(dat$fatty_liver))` participant; the
rest of the loss is `regular_exercise` (`r sum(is.na(dat$regular_exercise))`
missing) and `alcohol_g_per_week` (`r sum(is.na(dat$alcohol_g_per_week))`
missing), which do not overlap. Only `r sum(complete.cases(dat_raw))` of
`r nrow(dat)` rows are complete on *every* column in the file.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  mutate(fatty_liver = factor(fatty_liver, levels = c(0, 1),
                              labels = c("No fatty liver", "Fatty liver"))) |>
  tbl_summary(
    by = fatty_liver,
    missing = "no"
  ) |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics by fatty-liver status.**")
```

Participants with fatty liver are markedly heavier (median BMI
`r fmt(median(dat_model$bmi[dat_model$fatty_liver == 1]), 1)` versus
`r fmt(median(dat_model$bmi[dat_model$fatty_liver == 0]), 1)` kg/m2), have higher
gamma-GT (median `r fmt(median(dat_model$log2_ggt[dat_model$fatty_liver == 1]), 2)`
versus `r fmt(median(dat_model$log2_ggt[dat_model$fatty_liver == 0]), 2)` on the
log2 scale, i.e. about
`r fmt(2^(median(dat_model$log2_ggt[dat_model$fatty_liver == 1]) - median(dat_model$log2_ggt[dat_model$fatty_liver == 0])), 1)`-fold
higher on the raw scale), higher fasting glucose and triglycerides and lower HDL
cholesterol. They are overwhelmingly male
(`r fmt(100 * mean(dat_model$sex[dat_model$fatty_liver == 1] == "Male"), 0)`%
versus
`r fmt(100 * mean(dat_model$sex[dat_model$fatty_liver == 0] == "Male"), 0)`%), and
their median age is essentially the same as the unaffected group, which is
already a clue that this is a metabolic rather than a degenerative condition.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Proportion with ultrasound fatty liver by gamma-GT quartile and BMI category."}
dat_model |>
  mutate(
    ggt_q = cut(log2_ggt, breaks = quantile(log2_ggt, c(0, .25, .5, .75, 1)),
                include.lowest = TRUE,
                labels = c("Q1 (lowest)", "Q2", "Q3", "Q4 (highest)")),
    bmi_grp = cut(bmi, breaks = c(0, 22, 25, Inf), right = FALSE,
                  labels = c("<22", "22-24.9", "25+"))
  ) |>
  group_by(ggt_q, bmi_grp) |>
  summarise(p = mean(fatty_liver), n = n(), .groups = "drop") |>
  ggplot(aes(ggt_q, p, fill = bmi_grp)) +
  geom_col(position = position_dodge(0.85), width = 0.78) +
  geom_text(aes(label = paste0("n=", n)), position = position_dodge(0.85),
            vjust = -0.4, size = 2.8, colour = "grey30") +
  scale_y_continuous(labels = scales::percent, expand = expansion(c(0, 0.12))) +
  scale_fill_manual(values = c("<22" = "#7FA9C9", "22-24.9" = "#E0A458",
                               "25+" = "#B5495B")) +
  labs(x = "Gamma-GT quartile", y = "Proportion with fatty liver",
       fill = "BMI (kg/m2)")
```

Fatty-liver prevalence rises across gamma-GT quartiles within every BMI band, but
the BMI gradient is by far the steeper of the two: it is below 10% in the leanest
attendees at every level of GGT and above two-thirds in those with BMI 25 or more
in the top GGT quartile. (Both variables are categorised here for display only;
each enters the models as a continuous variable.)

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = glm,
    method.args = list(family = binomial),
    exponentiate = TRUE
  )
uv

utid <- map_dfr(predictors, function(v) {
  tidy(glm(reformulate(v, outcome), data = dat_model, family = binomial),
       conf.int = TRUE, exponentiate = TRUE) |>
    filter(term != "(Intercept)")
})
ug <- function(term, d = 2) {
  r <- utid[utid$term == term, ]
  sprintf("OR %s (95%% CI %s to %s)", fmt(r$estimate, d),
          fmt(r$conf.low, d), fmt(r$conf.high, d))
}
up <- function(term) pv(utid$p.value[utid$term == term])
```

BMI is the strongest crude predictor by a wide margin: each additional kg/m2
multiplies the odds of fatty liver by `r ug("bmi")` (p `r up("bmi")`), which
compounds to about
`r fmt(utid$estimate[utid$term == "bmi"]^5, 1)`-fold per 5 kg/m2. Each doubling
of gamma-GT multiplies the odds by `r ug("log2_ggt")` (p `r up("log2_ggt")`).
Women have `r ug("sexFemale")` the odds of men (p `r up("sexFemale")`), a nearly
four-fold male excess. The metabolic panel behaves as expected: fasting glucose
`r ug("fasting_glucose")` per mg/dL (p `r up("fasting_glucose")`), triglycerides
`r ug("triglycerides", 4)` per mg/dL (p `r up("triglycerides")`, i.e. about
`r fmt(utid$estimate[utid$term == "triglycerides"]^50, 2)`-fold per 50 mg/dL),
HDL cholesterol `r ug("hdl_cholesterol")` per mg/dL
(p `r up("hdl_cholesterol")`), and systolic pressure `r ug("systolic_bp")` per
mmHg (p `r up("systolic_bp")`). Regular exercise is associated with
`r ug("regular_exerciseYes")` (p `r up("regular_exerciseYes")`), and both former
(`r ug("smoking_statusFormer")`, p `r up("smoking_statusFormer")`) and current
smoking (`r ug("smoking_statusCurrent")`, p `r up("smoking_statusCurrent")`) with
raised odds. Two predictors show no crude association at all: age
(`r ug("age_yrs")`, p `r up("age_yrs")`) and - strikingly, for a liver outcome -
alcohol intake (`r ug("alcohol_g_per_week", 4)`, p `r up("alcohol_g_per_week")`).

## Why adjustment is needed

**Sex** confounds the gamma-GT association hard. Men have roughly
`r fmt(2^(median(dat_model$log2_ggt[dat_model$sex == "Male"]) - median(dat_model$log2_ggt[dat_model$sex == "Female"])), 1)`
times the GGT of women on the raw scale (medians
`r fmt(median(dat_model$log2_ggt[dat_model$sex == "Male"]), 2)` versus
`r fmt(median(dat_model$log2_ggt[dat_model$sex == "Female"]), 2)` on the log2
scale) and, independently of GGT, roughly four times the crude odds of fatty
liver, so part of any apparent GGT effect is simply a male effect. **BMI** is the
second, and it confounds nearly everything else in the list: heavier attendees
have higher GGT (r = `r fmt(cor(dat_model$bmi, dat_model$log2_ggt), 2)`), higher
triglycerides (r = `r fmt(cor(dat_model$bmi, dat_model$triglycerides), 2)`),
lower HDL (r = `r fmt(cor(dat_model$bmi, dat_model$hdl_cholesterol), 2)`), higher
blood pressure and higher glucose, and adiposity is the dominant cause of hepatic
steatosis. Any crude lipid, pressure or enzyme association is therefore partly a
body-mass association. **Alcohol** matters for the opposite reason. Drinking here
is almost a proxy for being male
(`r fmt(100 * mean(dat_model$alcohol_g_per_week[dat_model$sex == "Male"] > 0), 0)`%
of men report any alcohol against
`r fmt(100 * mean(dat_model$alcohol_g_per_week[dat_model$sex == "Female"] > 0), 0)`%
of women) and it is strongly associated with GGT
(r = `r fmt(cor(dat_model$alcohol_g_per_week, dat_model$log2_ggt), 2)`) and with
triglycerides
(r = `r fmt(cor(dat_model$alcohol_g_per_week, dat_model$triglycerides), 2)`).
Being male and having high GGT both raise the odds of fatty liver sharply, so
those two push the crude alcohol coefficient upward and cancel whatever the
drinking itself is doing. Within men alone the pattern is already visible:
`r fmt(100 * mean(dat_model$fatty_liver[dat_model$sex == "Male" & dat_model$alcohol_g_per_week == 0]), 0)`%
of male non-drinkers have fatty liver against
`r fmt(100 * mean(dat_model$fatty_liver[dat_model$sex == "Male" & dat_model$alcohol_g_per_week > 0]), 0)`%
of male drinkers. Mutual adjustment lets each odds ratio be read as a
comparison between attendees of the same age, sex, body mass, pressure, smoking,
drinking, activity and metabolic profile.

## Multivariable model

```{r multivariable}
fit <- glm(reformulate(predictors, outcome), data = dat_model, family = binomial)

mv <- tbl_regression(fit, exponentiate = TRUE)
mv

tid <- tidy(fit, conf.int = TRUE, exponentiate = TRUE)
g <- function(term, d = 2) {
  r <- tid[tid$term == term, ]
  sprintf("OR %s (95%% CI %s to %s)", fmt(r$estimate, d),
          fmt(r$conf.low, d), fmt(r$conf.high, d))
}
gp <- function(term) pv(tid$p.value[tid$term == term])
```

```{r fit-stats}
n_par    <- length(coef(fit)) - 1L
epp      <- n_events / n_par
null_fit <- glm(reformulate("1", outcome), data = dat_model, family = binomial)
tibble(
  Statistic = c("Complete-case N", "Events (fatty liver = 1)",
                "Parameters (excl. intercept)", "Events per parameter",
                "AIC (null model)", "AIC (full model)", "Tjur's R-squared"),
  Value = c(as.character(nrow(dat_model)), as.character(n_events),
            as.character(n_par), fmt(epp, 1),
            fmt(AIC(null_fit), 1), fmt(AIC(fit), 1),
            fmt(as.numeric(r2_tjur(fit)), 3))
) |>
  gt() |>
  tab_header(title = md("**Model fit: multivariable logistic model for fatty liver**"))
```

The model has `r n_events` events for `r n_par` estimated parameters, i.e.
**`r fmt(epp, 1)` events per parameter**, comfortably clear of the conventional
minimum of ten. AIC falls from `r fmt(AIC(null_fit), 1)` for the intercept-only
model to `r fmt(AIC(fit), 1)`, a drop of
`r fmt(AIC(null_fit) - AIC(fit), 1)`, and Tjur's R-squared is
`r fmt(as.numeric(r2_tjur(fit)), 3)` - a genuinely well-discriminating model by
the standards of observational epidemiology.

Among attendees of the same age, sex, blood pressure, smoking, drinking,
exercise, glucose, lipids, kidney function and gamma-GT, each additional kg/m2 of
BMI multiplies the odds of fatty liver by `r g("bmi")` (p `r gp("bmi")`), about
`r fmt(tid$estimate[tid$term == "bmi"]^5, 1)`-fold per 5 kg/m2. Among attendees
alike on everything else, each additional mg/dL of fasting glucose multiplies the
odds by `r g("fasting_glucose")` (p `r gp("fasting_glucose")`) and each
additional mg/dL of HDL cholesterol by `r g("hdl_cholesterol")`
(p `r gp("hdl_cholesterol")`), i.e. about
`r fmt(tid$estimate[tid$term == "hdl_cholesterol"]^10, 2)`-fold per 10 mg/dL of
HDL. Women retain `r g("sexFemale")` the odds of men (p `r gp("sexFemale")`).

Two coefficients are worth dwelling on. **Gamma-GT survives, but barely**: each
doubling gives `r g("log2_ggt")` (p `r gp("log2_ggt")`), against a crude
`r ug("log2_ggt")`. **Alcohol becomes protective**: `r g("alcohol_g_per_week", 4)`
per g/week (p `r gp("alcohol_g_per_week")`), which compounds to about
`r fmt(tid$estimate[tid$term == "alcohol_g_per_week"]^100, 2)`-fold per 100 g of
weekly alcohol, having been exactly null in the crude model. Age
(`r g("age_yrs")`), systolic pressure (`r g("systolic_bp")`), triglycerides
(`r g("triglycerides", 4)`), eGFR (`r g("egfr")`), regular exercise
(`r g("regular_exerciseYes")`) and both smoking contrasts are not independently
associated at the 0.05 level. These are associations among the attendees
observed, not effects of intervening.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted odds ratios.**")
```

**Gamma-GT moves furthest.** It falls from `r ug("log2_ggt")` crude
(p `r up("log2_ggt")`) to `r g("log2_ggt")` adjusted (p `r gp("log2_ggt")`), which
on the log-odds scale is about
`r fmt(100 * (1 - log(tid$estimate[tid$term == "log2_ggt"]) / log(utid$estimate[utid$term == "log2_ggt"])), 0)`%
of the way toward the null; the crude doubling-of-odds-per-doubling-of-GGT was
mostly adiposity and sex. It only just clears 0.05 after adjustment, so the
headline exposure of the source dataset is the most fragile coefficient in the
model. **Sex** moves the same way, from `r ug("sexFemale")` to `r g("sexFemale")`
(p `r up("sexFemale")` to p `r gp("sexFemale")`): roughly half of the crude
four-fold male excess is explained by men being heavier, drinking more and having
worse metabolic profiles, though a real excess remains. **Triglycerides** attenuate
from `r ug("triglycerides", 4)` to `r g("triglycerides", 4)` and lose significance
entirely; so does systolic pressure (`r ug("systolic_bp")` to `r g("systolic_bp")`,
p `r up("systolic_bp")` to p `r gp("systolic_bp")`). Both were largely BMI.

**Smoking reverses sign.** Crude odds ratios were above one for both former
(`r ug("smoking_statusFormer")`, p `r up("smoking_statusFormer")`) and current
(`r ug("smoking_statusCurrent")`, p `r up("smoking_statusCurrent")`) smoking;
adjusted they are below one and null (`r g("smoking_statusFormer")` and
`r g("smoking_statusCurrent")`). Smoking here is overwhelmingly a male behaviour -
`r sum(dat$sex == "Female" & dat$smoking_status == "Current")` of
`r sum(dat$sex == "Female")` women smoke currently, against
`r sum(dat$sex == "Male" & dat$smoking_status == "Current")` of
`r sum(dat$sex == "Male")` men - so the crude smoking association was a sex
association. **Alcohol goes the other way, from null to significant**:
`r ug("alcohol_g_per_week", 4)` crude (p `r up("alcohol_g_per_week")`) to
`r g("alcohol_g_per_week", 4)` adjusted (p `r gp("alcohol_g_per_week")`). This is
classic masking, and it can be traced variable by variable: adding sex alone to
the crude alcohol model moves the log-odds coefficient from
`r fmt(1000 * coef(glm(fatty_liver ~ alcohol_g_per_week, data = dat_model, family = binomial))["alcohol_g_per_week"], 2)`
to
`r fmt(1000 * coef(glm(fatty_liver ~ alcohol_g_per_week + sex, data = dat_model, family = binomial))["alcohol_g_per_week"], 2)`
per 1,000 g/week, and adding log2 gamma-GT alone moves it to
`r fmt(1000 * coef(glm(fatty_liver ~ alcohol_g_per_week + log2_ggt, data = dat_model, family = binomial))["alcohol_g_per_week"], 2)`.
Drinkers are male and have high GGT, and both of those raise the odds of fatty
liver; holding them constant unmasks a negative residual association.

**Two barely move.** BMI goes from `r ug("bmi")` to `r g("bmi")` and fasting
glucose from `r ug("fasting_glucose")` to `r g("fasting_glucose")` - modest
attenuation in both cases, but they remain the two strongest predictors in the
model and are not explained away by anything else measured here.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r binned-resid}
br <- binned_residuals(fit)
print(br)
br_pct <- attr(br, "resid_ok")
```

```{r collin}
cc <- as.data.frame(check_collinearity(fit))
check_collinearity(fit)
```

```{r sep-check}
# Smallest cross-classified cells a student might create, and the enzyme trap.
sparse_tab <- with(dat_model, table(sex, smoking_status, fatty_liver))
fit_enz <- glm(reformulate(c(predictors, "alt", "ast"), outcome),
               data = dat |> select(all_of(c(outcome, predictors, "alt", "ast"))) |>
                 drop_na(),
               family = binomial)
tid_enz <- tidy(fit_enz, conf.int = TRUE, exponentiate = TRUE)
ge <- function(term, d = 2) {
  r <- tid_enz[tid_enz$term == term, ]
  sprintf("OR %s (95%% CI %s to %s)", fmt(r$estimate, d),
          fmt(r$conf.low, d), fmt(r$conf.high, d))
}
gpe <- function(term) pv(tid_enz$p.value[tid_enz$term == term])
```

**Binned residuals are the check that fails.** Only
`r fmt(100 * br_pct, 0)`% of binned residuals lie inside their error bounds,
against a target of 95% or more, so the model is measurably miscalibrated
somewhere across the fitted-probability range. With `r n_events` events spread
over `r nrow(dat_model)` observations each bin holds around
`r round(nrow(dat_model) / nrow(br))` participants, so the check is noisy, but the
pattern is consistent with the linear-in-log-odds assumption failing for BMI: the
BMI-steatosis relationship is steeper in the middle of the range than at either
extreme, and a single linear term cannot capture that.

**Collinearity is not a problem in the reduced model.** The largest VIF is
`r fmt(max(cc$VIF), 2)` (`r cc$Term[which.max(cc$VIF)]`), well below the
conventional threshold of 5. In the 18-predictor version the largest VIF is
`r vall("total_cholesterol")`, so the a-priori cut is doing real work.

**There is no separation.** All coefficients are finite, the largest standard
error on the log-odds scale is
`r fmt(max(summary(fit)$coefficients[-1, 2]), 2)`, and every level of every
categorical predictor contains events (the smallest cell is
`r min(sparse_tab)` - female current smokers with fatty liver - which is thin but
not empty). That cell is worth noting because it is where a student adding a
`sex * smoking_status` interaction would run into trouble: with
`r sum(dat_model$sex == "Female" & dat_model$smoking_status == "Current")` female
current smokers in total, the interaction contrast is estimated from almost
nothing and its standard error explodes.

**Influence is unremarkable.** The largest Cook's distance is
`r fmt(max(cooks.distance(fit)), 3)` and the largest leverage
`r fmt(max(hatvalues(fit)), 3)`; no single participant drives the fit.

**Homogeneity of variance and normality of residuals do not apply** to a logistic
model and are not interpreted; the corresponding panels of `check_model()` should
be ignored here.

Finally, a targeted check of the liver-enzyme decision. Adding ALT and AST back
alongside gamma-GT does not produce alarming VIFs (largest
`r fmt(max(as.data.frame(check_collinearity(fit_enz))$VIF), 2)`), but it
**reverses the sign of the gamma-GT coefficient**: GGT goes from `r g("log2_ggt")`
in the model above to `r ge("log2_ggt")` (p `r gpe("log2_ggt")`) with the other
two enzymes present, while ALT takes over as the enzyme of record
(`r ge("alt")` per U/L, p `r gpe("alt")`). Three correlated markers of the same
organ competing for the same variance is exactly the situation in which a
conditional coefficient stops meaning what its name suggests, and it is the main
reason only one enzyme is carried in the primary model.

# Discussion

In this Japanese health-check cohort, ultrasound fatty liver is above all a
metabolic condition: body mass index and fasting glucose are the two dominant
independent predictors, with each 5 kg/m2 of BMI multiplying the odds by about
`r fmt(tid$estimate[tid$term == "bmi"]^5, 1)` and each 10 mg/dL of HDL
cholesterol dividing them by about
`r fmt(1 / tid$estimate[tid$term == "hdl_cholesterol"]^10, 2)`, holding everything
else constant. Gamma-GT, the marker this cohort was assembled to study, retains a
small independent association (`r g("log2_ggt")` per doubling) that only just
reaches significance, and roughly three-quarters of its crude association is
explained by adiposity and sex. Alcohol intake is *inversely* associated with
fatty liver once body mass and GGT are held constant, which is a reminder that
ultrasound steatosis in a screening population is largely non-alcoholic. In
practice, BMI and a fasting glucose identify most of the risk; the liver enzymes
add little beyond them.

**Limitations.** The design is cross-sectional, so temporal order is unknown and
reverse causation is entirely plausible - a fatty liver raises GGT, so the
GGT-steatosis association almost certainly runs partly backwards from outcome to
"predictor". The outcome is ultrasound-defined, which is operator-dependent, has
limited sensitivity for mild steatosis and is degraded by obesity itself, so
misclassification is probably differential with respect to BMI, the strongest
predictor. Health-check attendees are self-selected and healthier than the
general Japanese population, which compresses the range of every risk factor and
will attenuate the odds ratios. Alcohol is self-reported in g/week with
`r sum(dat_model$alcohol_g_per_week == 0)` attendees reporting exactly zero, and
under-reporting by heavy drinkers is the standard failure mode - the protective
alcohol coefficient should not be read as evidence that drinking prevents fatty
liver. No medication data are available, so statins, antihypertensives and
glucose-lowering drugs are unmeasured confounders. Waist circumference, insulin,
HOMA-IR and viral hepatitis status are absent, and the first two are more direct
measures of the visceral adiposity and insulin resistance that drive this
condition than BMI is. The binned-residual check fails
(`r fmt(100 * br_pct, 0)`% inside bounds), indicating miscalibration that a
non-linear term for BMI would probably improve. Complete-case analysis dropped
`r n_dropped` of `r nrow(dat)` rows. All estimates are associations, not causal
effects.

**Future work.** Refit with a spline or fractional polynomial for BMI to address
the calibration failure, and add waist circumference and a fasting insulin
measure so that the model separates visceral adiposity from body mass.

# Feasibility notes

*NOT part of a student report.*

- **Verdict: USABLE.** This is the strongest of the four outcomes in this
  dataset. `r n_events` events on `r nrow(dat_model)` complete cases,
  `r fmt(epp, 1)` events per parameter, Tjur's R-squared
  `r fmt(as.numeric(r2_tjur(fit)), 3)`, no separation, no collinearity problem
  and a clear, real signal. A competent student will get a clean model.
- **MAJOR: binned residuals fail** - only `r fmt(100 * br_pct, 0)`% of bins
  inside the error bounds against a 95% target. This is genuine miscalibration
  (the BMI-steatosis relationship is not linear in the log odds) rather than pure
  noise, and an honest student has to report a failed check. It is a good
  teaching case, but it will unsettle students who expect diagnostics to pass.
- **MAJOR: three correlated liver enzymes, and the choice between them changes
  the sign of the answer.** With only `log2_ggt` in the model, GGT is
  `r g("log2_ggt")` (p `r gp("log2_ggt")`). Add `alt` and `ast` and GGT becomes
  `r ge("log2_ggt")` (p `r gpe("log2_ggt")`) - a *protective* association - while
  ALT becomes the dominant enzyme (`r ge("alt")`, p `r gpe("alt")`). The VIFs
  never exceed `r fmt(max(as.data.frame(check_collinearity(fit_enz))$VIF), 1)`, so
  nothing warns the student that this has happened. This is the single most
  likely way a student produces a confidently wrong conclusion from this outcome,
  and it belongs in the `meta.json` cautions.
- **MAJOR collinearity if all four lipids are used.** VIFs in the 18-predictor
  logistic model: `r vall("total_cholesterol")` (total cholesterol),
  `r vall("ldl_cholesterol")` (LDL), `r vall("hdl_cholesterol")` (HDL),
  `r vall("triglycerides")` (triglycerides). Regressing total cholesterol on the
  other three gives R-squared `r fmt(fw_r2, 3)` with coefficients
  `r fmt(fw_coef["ldl_cholesterol"], 2)` / `r fmt(fw_coef["hdl_cholesterol"], 2)` /
  `r fmt(fw_coef["triglycerides"], 2)`, i.e. the Friedewald identity. It is
  *near*-deterministic, not exact - the residual
  `total_cholesterol - hdl_cholesterol - triglycerides/5 - ldl_cholesterol` has a
  median of about
  `r fmt(median(with(dat, total_cholesterol - hdl_cholesterol - triglycerides/5 - ldl_cholesterol), na.rm = TRUE), 1)`
  mg/dL - so the model fits without complaint and simply becomes
  uninterpretable.
- **MAJOR collinearity between the blood pressures** (r = `r fmt(bp_r, 2)`; VIFs
  `r vall("systolic_bp")` and `r vall("diastolic_bp")` with both in).
- **MAJOR: the exposure of interest is plausibly downstream of the outcome.**
  A steatotic liver leaks enzymes; GGT rises *because* of fatty liver at least as
  much as the reverse. Every GGT-fatty-liver statement in a cross-sectional file
  is uninterpretable causally, and the "predictor" framing invites students to
  get this backwards.
- **MINOR: near-separation risk only via interactions.** No main-effect term
  separates. The thinnest cell is `r min(sparse_tab)` (female current smokers
  with fatty liver, out of
  `r sum(dat_model$sex == "Female" & dat_model$smoking_status == "Current")` female
  current smokers). A `sex * smoking_status` interaction is therefore estimated
  from almost nothing; the coefficients stay finite but the standard errors are
  useless. Smoking is `r fmt(100 * mean(dat$sex[dat$smoking_status != "Never"] == "Male"), 0)`%
  male in this file.
- **MINOR: alcohol flips from exactly null to significantly protective on
  adjustment** (`r ug("alcohol_g_per_week", 4)` to
  `r g("alcohol_g_per_week", 4)`). Correct arithmetic, dangerous prose - a weak
  student will write that alcohol protects the liver. The mechanism is masking by
  BMI and sex, and the real explanation is under-reporting plus the fact that
  ultrasound steatosis here is mostly non-alcoholic.
- **MINOR: complete-case loss is small** - `r nrow(dat_model)` of `r nrow(dat)`
  (`r fmt(100 * n_dropped / nrow(dat), 1)`%), from `regular_exercise`
  (`r sum(is.na(dat$regular_exercise))` missing), `alcohol_g_per_week`
  (`r sum(is.na(dat$alcohol_g_per_week))` missing) and one missing outcome. Events fall
  from `r sum(dat_raw$fatty_liver, na.rm = TRUE)` to `r n_events`, which changes
  nothing.
- **MINOR: no medication, waist circumference, insulin or hepatitis serology.**
  Residual confounding is certain, and BMI is a blunt proxy for the visceral
  adiposity that actually drives steatosis.
- The outcome is **not** a deterministic function of any shipped column: the
  best single-variable discrimination is BMI, and even a BMI-only model reaches
  Tjur's R-squared only
  `r fmt(as.numeric(r2_tjur(glm(fatty_liver ~ bmi, data = dat_model, family = binomial))), 2)`.
