---
title: "Serum gamma-glutamyltranspeptidase in a Japanese health-check cohort"
subtitle: "Dataset 9: Liver enzymes and arterial stiffness in a Japanese health-check cohort"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "ggt_arterial_stiffness.csv"),
                    show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names. `ggt` and `log2_ggt`
# are the same variable on two scales and neither can appear on the right-hand
# side; ALT and AST can, and do.
outcome    <- "log2_ggt"
predictors <- c("age_yrs", "sex", "bmi", "systolic_bp", "smoking_status",
                "alcohol_g_per_week", "regular_exercise", "fasting_glucose",
                "triglycerides", "hdl_cholesterol", "egfr", "alt", "ast")

# The full catalogue list minus the outcome, used once to document the
# collinearity that motivated the cut. No inference is drawn from that model.
predictors_all <- c("age_yrs", "sex", "bmi", "systolic_bp", "diastolic_bp",
                    "smoking_status", "alcohol_g_per_week", "regular_exercise",
                    "fasting_glucose", "uric_acid", "total_cholesterol",
                    "hdl_cholesterol", "ldl_cholesterol", "triglycerides",
                    "egfr", "alt", "ast")

lab <- function(x, l) { attr(x, "label") <- l; x }

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex = lab(factor(sex, levels = c("male", "female"), labels = c("Male", "Female")),
              "Sex"),
    smoking_status = lab(factor(smoking_status,
                                levels = c("never", "former", "current"),
                                labels = c("Never", "Former", "Current")),
                         "Smoking status"),
    regular_exercise = lab(factor(regular_exercise, levels = c(0, 1),
                                  labels = c("No", "Yes")),
                           "Takes regular exercise"),
    age_yrs            = lab(age_yrs, "Age (years)"),
    bmi                = lab(bmi, "BMI (kg/m2)"),
    systolic_bp        = lab(systolic_bp, "Systolic BP (mmHg)"),
    diastolic_bp       = lab(diastolic_bp, "Diastolic BP (mmHg)"),
    alcohol_g_per_week = lab(alcohol_g_per_week, "Alcohol (g/week)"),
    fasting_glucose    = lab(fasting_glucose, "Fasting glucose (mg/dL)"),
    uric_acid          = lab(uric_acid, "Uric acid (mg/dL)"),
    total_cholesterol  = lab(total_cholesterol, "Total cholesterol (mg/dL)"),
    triglycerides      = lab(triglycerides, "Triglycerides (mg/dL)"),
    hdl_cholesterol    = lab(hdl_cholesterol, "HDL cholesterol (mg/dL)"),
    ldl_cholesterol    = lab(ldl_cholesterol, "LDL cholesterol (mg/dL)"),
    egfr               = lab(egfr, "eGFR (mL/min/1.73m2)"),
    alt                = lab(alt, "ALT (U/L)"),
    ast                = lab(ast, "AST (U/L)"),
    log2_ggt           = lab(log2_ggt, "Gamma-GT (log2 U/L)")
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
pv  <- function(p) ifelse(p < 0.001, "<0.001", paste0("= ", fmt(p, 3)))
# Convert a log2 coefficient into a multiplicative effect on raw GGT.
mult <- function(b, per = 1) fmt(2^(b * per), 3)
```

```{r collinearity-evidence}
# Documented redundancies in the suggested predictor list, quantified once.
bp_r  <- cor(dat$systolic_bp, dat$diastolic_bp)
enz_r <- cor(dat$alt, dat$ast)

fw  <- lm(total_cholesterol ~ ldl_cholesterol + hdl_cholesterol + triglycerides,
          data = dat)
fw_r2   <- summary(fw)$r.squared
fw_coef <- coef(fw)

dat_all <- dat |> select(all_of(c(outcome, predictors_all))) |> drop_na()
fit_all <- lm(reformulate(predictors_all, outcome), data = dat_all)
cc_all  <- as.data.frame(check_collinearity(fit_all))
vall <- function(term, d = 1) fmt(cc_all$VIF[cc_all$Term == term], d)

# The deterministic relationship that must never enter the model.
det_gap <- max(abs(log2(dat$ggt) - dat$log2_ggt))
skew <- function(x) mean((x - mean(x))^3) / sd(x)^3
```

# Introduction

A general health-check programme in Japan enrolled `r nrow(dat)` men and women
aged `r min(dat$age_yrs)` to `r max(dat$age_yrs)` years and measured a full
fasting biochemistry panel alongside abdominal ultrasound and automated vascular
measurements; the cohort is reported in *BMJ Open* 2014;4:e005413.
Gamma-glutamyltranspeptidase (GGT) is the liver enzyme most often used as a
marker of alcohol intake, but it also rises with hepatic steatosis, oxidative
stress and metabolic syndrome, and raised GGT predicts cardiovascular disease and
diabetes independently of drinking. Understanding what a raised GGT at a routine
health check actually reflects is therefore a question about the interpretation
of a very widely ordered test.

**Aim.** Among these health-check attendees, we examined which of age, sex, body
mass index, systolic blood pressure, smoking status, alcohol intake, regular
exercise, fasting glucose, triglycerides, HDL cholesterol, kidney function and
the two other liver enzymes (ALT and AST) are independently associated with serum
GGT, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** A cross-sectional Japanese health-check cohort of `r nrow(dat)`
adults attending a voluntary medical check-up programme that included a fasting
blood panel, abdominal ultrasound and automated measurement of pulse-wave
velocity and the ankle-brachial index.

**Variables.** The outcome is `log2_ggt`, serum gamma-GT on a base-2 logarithmic
scale. The raw enzyme is severely right-skewed (skewness
`r fmt(skew(dat$ggt), 2)`, median `r median(dat$ggt)` U/L, maximum
`r max(dat$ggt)` U/L); on the log2 scale the skewness falls to
`r fmt(skew(dat$log2_ggt), 2)`, which is why the catalogue ships the transformed
version and why we model it. A coefficient of *b* on this scale means that a
one-unit increase in the predictor multiplies GGT by 2^*b*. Predictors are age
(years), sex (reference Male, the larger group at `r sum(dat$sex == "Male")` of
`r nrow(dat)`), BMI (kg/m2), systolic blood pressure (mmHg), smoking status
(three levels, reference Never, the largest group at
`r sum(dat$smoking_status == "Never")`), alcohol intake (g/week), regular
exercise (reference No, the unexposed and larger group), fasting glucose
(mg/dL), triglycerides (mg/dL), HDL cholesterol (mg/dL), estimated glomerular
filtration rate (mL/min/1.73m2), ALT (U/L) and AST (U/L).

**`ggt` is excluded from the predictor set as a matter of arithmetic, not
judgement.** `log2_ggt` is exactly `log2(ggt)` in this file - the largest
absolute discrepancy across all `r nrow(dat)` rows is
`r format(det_gap, scientific = TRUE, digits = 2)`, i.e. floating-point noise -
so regressing the outcome on `ggt` is regressing a variable on a monotone
transformation of itself. Any model containing it is circular and its R-squared
is meaningless.

**The predictor set was cut from the 17 remaining suggested predictors to
`r length(predictors)`.** With `r nrow(dat_model)` complete cases and
`r length(coef(lm(reformulate(predictors, outcome), data = dat_model))) - 1L`
parameters this model already has about
`r fmt(nrow(dat_model) / (length(coef(lm(reformulate(predictors, outcome), data = dat_model))) - 1L), 0)`
observations per parameter, so the 10-15 observations-per-parameter budget would
have tolerated all 17; the cut is about **redundancy** and **interpretability**
instead. Systolic and diastolic pressure correlate at r = `r fmt(bp_r, 2)` and
carry VIFs of `r vall("systolic_bp")` and `r vall("diastolic_bp")` when both are
included, so only systolic pressure is kept. The four lipid measures satisfy the
Friedewald identity total = LDL + HDL + triglycerides/5 to within rounding
(regressing total cholesterol on the other three gives R-squared
`r fmt(fw_r2, 3)` with coefficients `r fmt(fw_coef["ldl_cholesterol"], 2)`,
`r fmt(fw_coef["hdl_cholesterol"], 2)` and `r fmt(fw_coef["triglycerides"], 2)`),
and with all four in the model the VIFs reach `r vall("total_cholesterol")` and
`r vall("ldl_cholesterol")`; only HDL cholesterol and triglycerides, the two
metabolic-syndrome components, are kept. Uric acid is dropped because it
duplicates information already carried by adiposity and kidney function
(r = `r fmt(cor(dat$uric_acid, dat$egfr), 2)` with eGFR,
r = `r fmt(cor(dat$uric_acid, dat$bmi), 2)` with BMI).

**ALT and AST are deliberately retained, despite correlating at
r = `r fmt(enz_r, 2)`.** With GGT as the outcome the other two enzymes stop being
nuisance duplicates and become the substantive question: does GGT carry
information about the liver beyond what ALT and AST already say? Their VIFs are
reported explicitly in the diagnostics, and the interpretation of each is
conditional on the other.

**Statistical analysis.** `lm()`, justified because the outcome is a continuous
(log-transformed) measurement. We fitted a univariable model for each predictor
first, then a single multivariable model containing all of them. Predictors were
chosen in advance on subject-matter grounds, not by significance. Significance
threshold 0.05; estimates are reported with 95% confidence intervals. Assumptions
were checked with `performance::check_model()`, `check_heteroscedasticity()` and
`check_collinearity()`, and because non-constant variance was flagged we
additionally report HC3 heteroscedasticity-consistent standard errors. Analysis
used `r R.version.string` with tidyverse, gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped). The outcome is complete; the entire
loss comes from `regular_exercise` (`r sum(is.na(dat$regular_exercise))` missing)
and `alcohol_g_per_week` (`r sum(is.na(dat$alcohol_g_per_week))` missing), which
do not overlap. Only `r sum(complete.cases(dat_raw))` of `r nrow(dat)` rows are
complete on *every* column in the file.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(missing = "no") |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

Median log2 GGT is `r fmt(median(dat_model$log2_ggt), 2)`, i.e.
`r fmt(2^median(dat_model$log2_ggt), 0)` U/L on the raw scale, with an
interquartile range spanning
`r fmt(2^quantile(dat_model$log2_ggt, 0.25), 0)` to
`r fmt(2^quantile(dat_model$log2_ggt, 0.75), 0)` U/L - a factor of about
`r fmt(2^(quantile(dat_model$log2_ggt, 0.75) - quantile(dat_model$log2_ggt, 0.25)), 1)`
between the quartiles, which is why a log scale is needed. The cohort is
`r fmt(100 * mean(dat_model$sex == "Male"), 0)`% male, median age
`r median(dat_model$age_yrs)` years, and reported alcohol intake has a median of
only `r fmt(median(dat_model$alcohol_g_per_week), 1)` g/week but an upper
quartile of `r fmt(quantile(dat_model$alcohol_g_per_week, 0.75), 0)` g/week: half
the cohort barely drinks and a quarter drinks heavily.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Log2 gamma-GT against ALT, by sex, with linear smooths."}
ggplot(dat_model, aes(alt, log2_ggt, colour = sex)) +
  geom_point(alpha = 0.35, size = 1.4) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.9) +
  scale_x_continuous(trans = "log2",
                     breaks = c(6, 12, 25, 50, 100, 150)) +
  scale_colour_manual(values = c(Male = "#2E6F95", Female = "#C46B4A")) +
  labs(x = "ALT (U/L, log2 axis)", y = "Gamma-GT (log2 U/L)", colour = "Sex")
```

GGT rises steadily with ALT in both sexes and the two slopes are close to
parallel, but the male line sits roughly half a log2 unit above the female line
across the whole range - at any given ALT, men have around
`r fmt(2^(median(dat_model$log2_ggt[dat_model$sex == "Male"]) - median(dat_model$log2_ggt[dat_model$sex == "Female"])), 1)`
times the GGT of women.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv

utid <- map_dfr(predictors, function(v) {
  tidy(lm(reformulate(v, outcome), data = dat_model), conf.int = TRUE) |>
    filter(term != "(Intercept)")
})
ug <- function(term, d = 4) {
  r <- utid[utid$term == term, ]
  sprintf("%s log2 units (95%% CI %s to %s)", fmt(r$estimate, d),
          fmt(r$conf.low, d), fmt(r$conf.high, d))
}
up <- function(term) pv(utid$p.value[utid$term == term])
ue <- function(term) utid$estimate[utid$term == term]
```

Almost everything is associated with GGT crudely. The two other liver enzymes are
the strongest: each additional U/L of AST is associated with `r ug("ast")`
(p `r up("ast")`), i.e. GGT multiplied by `r mult(ue("ast"), 10)` per 10 U/L, and
each additional U/L of ALT with `r ug("alt")` (p `r up("alt")`), a factor of
`r mult(ue("alt"), 10)` per 10 U/L. Women have `r ug("sexFemale", 3)` lower log2
GGT than men (p `r up("sexFemale")`), i.e. about
`r fmt(100 * (1 - 2^ue("sexFemale")), 0)`% lower GGT. Alcohol is associated with
`r ug("alcohol_g_per_week")` per g/week (p `r up("alcohol_g_per_week")`), a
factor of `r mult(ue("alcohol_g_per_week"), 100)` per 100 g/week. Triglycerides
(`r ug("triglycerides")` per mg/dL, p `r up("triglycerides")`), BMI
(`r ug("bmi", 3)` per kg/m2, p `r up("bmi")`), systolic pressure
(`r ug("systolic_bp")` per mmHg, p `r up("systolic_bp")`) and fasting glucose
(`r ug("fasting_glucose")` per mg/dL, p `r up("fasting_glucose")`) are all
positively associated; HDL cholesterol (`r ug("hdl_cholesterol")` per mg/dL,
p `r up("hdl_cholesterol")`), eGFR (`r ug("egfr")` per unit,
p `r up("egfr")`) and regular exercise (`r ug("regular_exerciseYes", 3)`,
p `r up("regular_exerciseYes")`) negatively. Both former
(`r ug("smoking_statusFormer", 3)`, p `r up("smoking_statusFormer")`) and current
smoking (`r ug("smoking_statusCurrent", 3)`, p `r up("smoking_statusCurrent")`)
are associated with GGT roughly `r fmt(100 * (2^ue("smoking_statusCurrent") - 1), 0)`%
higher than never smoking. The single exception is age, which shows no crude
association at all (`r ug("age_yrs")`, p `r up("age_yrs")`).

## Why adjustment is needed

**Sex** is the pivotal confounder for this outcome, and it confounds almost every
other association in the list. Men have median GGT
`r fmt(median(2^dat_model$log2_ggt[dat_model$sex == "Male"]), 0)` U/L against
`r fmt(median(2^dat_model$log2_ggt[dat_model$sex == "Female"]), 0)` U/L in women,
and men also drink far more (median
`r fmt(median(dat_model$alcohol_g_per_week[dat_model$sex == "Male"]), 0)` versus
`r fmt(median(dat_model$alcohol_g_per_week[dat_model$sex == "Female"]), 0)`
g/week), smoke far more
(`r fmt(100 * mean(dat_model$smoking_status[dat_model$sex == "Male"] != "Never"), 0)`%
ever-smokers against
`r fmt(100 * mean(dat_model$smoking_status[dat_model$sex == "Female"] != "Never"), 0)`%),
carry more weight and have higher blood pressure. Any crude behavioural or
metabolic association with GGT is therefore partly a sex association.

**ALT** is the second variable that must be handled explicitly, and it works on
the BMI association. Adiposity raises GGT largely by causing hepatic steatosis,
and steatosis raises ALT too (BMI-ALT r =
`r fmt(cor(dat_model$bmi, dat_model$alt), 2)`), so much of the crude BMI-GGT
association is really a shared-liver-injury association. This is confounding and
mediation at the same time: fat in the liver is a common cause of both enzymes
rising, but it is also the pathway by which body mass acts. **Alcohol** is the
third: it raises GGT directly, it is strongly patterned by sex, and drinkers here
are also smokers (median alcohol
`r fmt(median(dat_model$alcohol_g_per_week[dat_model$smoking_status == "Current"]), 0)`
g/week in current smokers against
`r fmt(median(dat_model$alcohol_g_per_week[dat_model$smoking_status == "Never"]), 0)`
in never smokers), so a crude smoking coefficient is partly a drinking
coefficient. Mutual adjustment lets each estimate be read as a comparison between
attendees alike on the other measured characteristics.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv

tid <- tidy(fit, conf.int = TRUE)
g <- function(term, d = 4) {
  r <- tid[tid$term == term, ]
  sprintf("%s log2 units (95%% CI %s to %s)", fmt(r$estimate, d),
          fmt(r$conf.low, d), fmt(r$conf.high, d))
}
gp <- function(term) pv(tid$p.value[tid$term == term])
ge <- function(term) tid$estimate[tid$term == term]
```

```{r fit-stats}
s <- summary(fit)
tibble(
  Statistic = c("Complete-case N", "Parameters (excl. intercept)",
                "Observations per parameter", "R-squared", "Adjusted R-squared",
                "Residual SE (log2 units)", "SD of outcome (log2 units)"),
  Value = c(as.character(nrow(dat_model)),
            as.character(length(coef(fit)) - 1L),
            fmt(nrow(dat_model) / (length(coef(fit)) - 1L), 1),
            fmt(s$r.squared, 3), fmt(s$adj.r.squared, 3),
            fmt(s$sigma, 3), fmt(sd(dat_model$log2_ggt), 3))
) |>
  gt() |>
  tab_header(title = md("**Model fit: multivariable linear model for log2 gamma-GT**"))
```

The model explains `r fmt(100 * s$r.squared, 1)`% of the variance in log2 GGT
(adjusted R-squared `r fmt(s$adj.r.squared, 3)`). Residual SE is
`r fmt(s$sigma, 3)` log2 units against an outcome SD of
`r fmt(sd(dat_model$log2_ggt), 3)`: in raw terms the model still leaves a typical
error of a factor of `r fmt(2^s$sigma, 2)` in either direction, which is a useful
reminder of how variable this enzyme is.

Among attendees of the same sex, body mass, blood pressure, smoking, drinking,
exercise, glucose, lipids, kidney function and other liver enzymes, each
additional U/L of ALT is associated with `r g("alt")` (p `r gp("alt")`),
i.e. GGT multiplied by `r mult(ge("alt"), 10)` per 10 U/L, and each additional
U/L of AST with `r g("ast")` (p `r gp("ast")`), a factor of
`r mult(ge("ast"), 10)` per 10 U/L. Among attendees alike on everything else,
each additional gram of weekly alcohol is associated with
`r g("alcohol_g_per_week")` (p `r gp("alcohol_g_per_week")`), a factor of
`r mult(ge("alcohol_g_per_week"), 100)` per 100 g/week, and women still have
`r g("sexFemale", 3)` lower log2 GGT than men (p `r gp("sexFemale")`), about
`r fmt(100 * (1 - 2^ge("sexFemale")), 0)`% lower. Triglycerides remain associated
(`r g("triglycerides")` per mg/dL, p `r gp("triglycerides")`), as do age
(`r g("age_yrs")` per year, p `r gp("age_yrs")`), systolic pressure
(`r g("systolic_bp")` per mmHg, p `r gp("systolic_bp")`) and regular exercise,
which is associated with `r g("regular_exerciseYes", 3)`
(p `r gp("regular_exerciseYes")`), i.e. GGT about
`r fmt(100 * (1 - 2^ge("regular_exerciseYes")), 0)`% lower.

Four predictors are *not* independently associated: BMI (`r g("bmi", 3)`,
p `r gp("bmi")`), fasting glucose (`r g("fasting_glucose")`,
p `r gp("fasting_glucose")`), eGFR (`r g("egfr")`, p `r gp("egfr")`) and both
smoking contrasts (`r g("smoking_statusFormer", 3)` and
`r g("smoking_statusCurrent", 3)`). HDL cholesterol is borderline and has changed
direction (`r g("hdl_cholesterol")`, p `r gp("hdl_cholesterol")`). These are
associations among the attendees observed, not effects of intervening.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

**BMI is fully explained away.** It falls from `r ug("bmi", 3)` crude
(p `r up("bmi")`) to `r g("bmi", 3)` adjusted (p `r gp("bmi")`) - a coefficient
that started clearly positive ends at essentially zero, and slightly negative.
ALT does most of that on its own: adding ALT alone to the crude BMI model moves
the coefficient from `r fmt(ue("bmi"), 4)` to
`r fmt(coef(lm(log2_ggt ~ bmi + alt, data = dat_model))["bmi"], 4)`. The reading
is that body mass raises GGT *through* liver fat, and once liver fat is proxied
by ALT there is nothing left for BMI to explain - which makes ALT a mediator here
rather than a confounder, and the adjusted BMI coefficient correspondingly hard
to interpret causally.

**Smoking reverses sign and loses significance entirely.** Current smoking goes
from `r ug("smoking_statusCurrent", 3)` (p `r up("smoking_statusCurrent")`) to
`r g("smoking_statusCurrent", 3)` (p `r gp("smoking_statusCurrent")`), and former
smoking likewise. Sex alone accounts for it: adding sex to the crude smoking
model moves the current-smoking coefficient from
`r fmt(ue("smoking_statusCurrent"), 3)` to
`r fmt(coef(lm(log2_ggt ~ smoking_status + sex, data = dat_model))["smoking_statusCurrent"], 3)`.
Smokers here are `r fmt(100 * mean(dat_model$sex[dat_model$smoking_status == "Current"] == "Male"), 0)`%
male, and men have much higher GGT, so the crude smoking association was a sex
association.

**HDL cholesterol changes direction.** Crudely it is `r ug("hdl_cholesterol")`
(p `r up("hdl_cholesterol")`) - higher HDL, lower GGT - and adjusted it is
`r g("hdl_cholesterol")` (p `r gp("hdl_cholesterol")`), pointing the other way.
Sex and triglycerides are jointly responsible: adding those two alone to the
crude model already moves the coefficient to
`r fmt(coef(lm(log2_ggt ~ hdl_cholesterol + sex + triglycerides, data = dat_model))["hdl_cholesterol"], 4)`.
Because HDL is inversely correlated with triglycerides
(r = `r fmt(cor(dat_model$hdl_cholesterol, dat_model$triglycerides), 2)`) and
women have both higher HDL and lower GGT, the crude negative association was
carrying two other variables. The adjusted estimate is borderline and should not
be over-read.

**Age goes the other way, from null to significant.** It moves *away* from the
null on adjustment, `r ug("age_yrs")` (p `r up("age_yrs")`) to `r g("age_yrs")`
(p `r gp("age_yrs")`), because sex was masking it: women are slightly older in
this cohort (median `r median(dat_model$age_yrs[dat_model$sex == "Female"])`
versus `r median(dat_model$age_yrs[dat_model$sex == "Male"])` years) and have much
lower GGT, so the crude age slope was flattened. **AST attenuates sharply**, from
`r ug("ast")` to `r g("ast")`, retaining significance but at only about
`r fmt(100 * ge("ast") / ue("ast"), 0)`% of its crude magnitude - that is ALT and
AST competing, since they correlate at r = `r fmt(cor(dat_model$alt, dat_model$ast), 2)`
and ALT wins. The surviving AST significance is marginal
(p `r gp("ast")`) and, as the diagnostics show, does not hold up under robust
standard errors.

**Two barely move in relative terms.** Alcohol falls only from
`r ug("alcohol_g_per_week")` to `r g("alcohol_g_per_week")`, retaining about
`r fmt(100 * ge("alcohol_g_per_week") / ue("alcohol_g_per_week"), 0)`% of its
crude size and remaining highly significant, and regular exercise from
`r ug("regular_exerciseYes", 3)` to `r g("regular_exerciseYes", 3)`. These two are
the behavioural findings that survive everything else in the file.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r het-check}
het <- check_heteroscedasticity(fit)
print(het)
nrm <- check_normality(fit)
print(nrm)
```

```{r collin}
cc <- as.data.frame(check_collinearity(fit))
check_collinearity(fit)
```

```{r robust-se}
rob   <- lmtest::coeftest(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
robci <- lmtest::coefci(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
tibble(
  Term = rownames(rob),
  Estimate = fmt(rob[, 1], 4),
  `SE (OLS)` = fmt(summary(fit)$coefficients[, 2], 4),
  `SE (HC3)` = fmt(rob[, 2], 4),
  `95% CI (HC3)` = paste0(fmt(robci[, 1], 3), " to ", fmt(robci[, 2], 3)),
  `p (HC3)` = fmt(rob[, 4], 3)
) |>
  filter(Term != "(Intercept)") |>
  gt() |>
  tab_header(title = md("**Table 3. HC3 heteroscedasticity-consistent standard errors.**"))
```

**Collinearity is real but tolerable, and it is concentrated exactly where
expected.** The two largest VIFs are `r fmt(cc$VIF[cc$Term == "alt"], 2)` for ALT
and `r fmt(cc$VIF[cc$Term == "ast"], 2)` for AST - the deliberately retained
correlated pair - followed by `r fmt(sort(cc$VIF, decreasing = TRUE)[3], 2)` for
`r cc$Term[order(cc$VIF, decreasing = TRUE)[3]]`. None exceeds the conventional
threshold of 5, so both enzyme coefficients are estimable, but they are competing:
the standard error on ALT is inflated by a factor of about
`r fmt(sqrt(cc$VIF[cc$Term == "alt"]), 2)` relative to an uncorrelated predictor,
and this is why AST's coefficient collapses on adjustment while ALT's survives.
Had all three enzymes been available as predictors the situation would be worse;
had all four lipids been included, worse still (largest VIF
`r vall("total_cholesterol")` in the 17-predictor model).

**Variance is not constant and residuals are not normal**, even on the log2
scale. `check_heteroscedasticity()` rejects (p `r pv(as.numeric(het))`) and
`check_normality()` rejects (p `r pv(as.numeric(nrm))`). The residual-versus-
fitted panel shows the spread narrowing at high fitted values, and the QQ plot
shows heavy tails at both ends. This is worth being clear about: the log2
transformation is doing a great deal of work - the same model on **raw** `ggt`
has R-squared
`r fmt(summary(lm(reformulate(predictors, "I(2^log2_ggt)"), data = dat_model))$r.squared, 3)`
against `r fmt(s$r.squared, 3)` here - but it does not fully normalise the
residuals. Table 3 therefore reports HC3 standard errors. They are close to the
OLS ones for most terms (ALT: OLS `r fmt(summary(fit)$coefficients["alt", 2], 4)`
versus HC3 `r fmt(rob["alt", 2], 4)`), and every substantive conclusion survives
with **one exception**: **AST loses significance under robust inference**, from
p `r pv(summary(fit)$coefficients["ast", 4])` under OLS to
p `r pv(rob["ast", 4])` under HC3, because its HC3 standard error
(`r fmt(rob["ast", 2], 4)`) is about
`r fmt(100 * (rob["ast", 2] / summary(fit)$coefficients["ast", 2] - 1), 0)`%
larger than the OLS one. Combined with the collinearity above, the honest reading
is that the marginal AST coefficient is not reliable and only ALT should be
reported as an independently associated enzyme.

**Influence.** The largest Cook's distance is
`r fmt(max(cooks.distance(fit)), 3)` and the largest leverage
`r fmt(max(hatvalues(fit)), 3)`. Both are well inside the usual contours; the
highest-leverage observations are the very heavy drinkers, whose reported intake
reaches `r max(dat_model$alcohol_g_per_week)` g/week, but no single participant
determines the fit.

# Discussion

In this Japanese health-check cohort, a raised gamma-GT is best explained by
three things: ALT, alcohol and sex. Holding everything else
constant, GGT is multiplied by `r mult(ge("alt"), 10)` per 10 U/L of ALT,
`r mult(ge("alcohol_g_per_week"), 100)` per 100 g of weekly alcohol, and is about
`r fmt(100 * (1 - 2^ge("sexFemale")), 0)`% lower in women than in men. Regular
exercise is associated with GGT about
`r fmt(100 * (1 - 2^ge("regular_exerciseYes")), 0)`% lower. The crude associations
with BMI, fasting glucose, kidney function and smoking all disappear once these
are accounted for; the BMI association in particular is fully absorbed by ALT,
which is consistent with body mass acting on GGT through hepatic fat rather than
through any separate route. Practically, this supports the conventional reading
of a raised GGT at a health check - ask about alcohol and look at the rest of the
liver panel - and argues against treating it as an independent metabolic marker
once ALT is known.

**Limitations.** The design is cross-sectional and the "predictors" here include
two other enzymes measured from the same blood draw, so nothing in the ALT and
AST coefficients can be read causally at all: they are contemporaneous markers of
the same organ, not exposures. Adjusting for ALT is adjusting for a mediator of
the BMI-GGT relationship, so the null BMI coefficient is an over-adjusted
estimate and should not be reported as "BMI does not affect GGT". Alcohol is
self-reported in g/week, with `r sum(dat_model$alcohol_g_per_week == 0)`
attendees reporting exactly zero and under-reporting concentrated among heavy
drinkers, which will bias the alcohol coefficient toward the null - the true
alcohol-GGT gradient is almost certainly steeper than the one estimated here.
Health-check attendees are self-selected and healthier than the general Japanese
population, and people with known liver disease may avoid or be excluded from
such programmes, further compressing the range of the outcome. Hepatitis B and C
serology, medication use (statins, anticonvulsants and several other drug classes
induce GGT) and body-composition measures beyond BMI are all absent. Residual
variance is non-constant and residuals are non-normal even after the log2
transformation; HC3 standard errors address inference but not fit.
`r n_dropped` of `r nrow(dat)` rows were dropped for missing exercise or alcohol.
All estimates are associations, not causal effects.

**Future work.** Repeat with a validated alcohol questionnaire and hepatitis
serology, and model GGT jointly with ALT and AST as correlated markers of a
single latent liver process rather than treating two of them as covariates for
the third.

# Feasibility notes

*NOT part of a student report.*

- **Verdict: USABLE.** `log2_ggt` is complete (0 missing), well behaved on the
  log2 scale, and produces a model with R-squared `r fmt(s$r.squared, 2)` and
  several clearly significant, substantively sensible coefficients. It is the
  most interpretable of the three linear outcomes in this dataset.
- **FATAL if `ggt` is used as a predictor.** `log2_ggt` is exactly `log2(ggt)`
  (maximum discrepancy across all `r nrow(dat)` rows:
  `r format(det_gap, scientific = TRUE, digits = 2)`). A model containing `ggt`
  is regressing the outcome on a monotone transformation of itself and returns
  an R-squared of about
  `r fmt(summary(lm(log2_ggt ~ ggt, data = dat))$r.squared, 3)` from that one
  term alone. `ggt` is not in the catalogue's suggested predictor list, but it is
  in the shipped file and it is the obvious thing for a careless student to grab.
  Worth an explicit caution.
- **MAJOR: `log2_ggt` must be removed from the suggested predictor list when it
  is the outcome.** `meta.json` currently lists `log2_ggt` both as an outcome and
  as a suggested predictor. A student copying the predictor vector verbatim will
  fit `log2_ggt ~ ... + log2_ggt`. R drops the term with the warning "the
  response appeared on the right-hand side and was dropped" - but the template's
  setup chunk sets `warning = FALSE`, so that warning never reaches the knitted
  document. The model runs, the table looks fine, and it is not the model the
  student described in their Methods.
- **MAJOR: ALT and AST are the interesting-but-collinear pair** (r =
  `r fmt(enz_r, 2)`; VIFs `r fmt(cc$VIF[cc$Term == "alt"], 2)` and
  `r fmt(cc$VIF[cc$Term == "ast"], 2)`). Both stay under 5, so no automated check
  will complain, but AST's coefficient drops from `r ug("ast")` crude to
  `r g("ast")` adjusted, i.e. to about
  `r fmt(100 * ge("ast") / ue("ast"), 0)`% of its crude value, purely because ALT
  is in the model. Students will report "AST is only weakly associated with GGT",
  which is a statement about ALT, not about AST.
- **MAJOR conceptual problem: the strongest predictors are not exposures.** ALT
  and AST come from the same blood sample as the outcome and index the same organ.
  Treating them as predictors of GGT is a prediction exercise dressed as an
  aetiological one, and adjusting for them turns the BMI coefficient from
  `r ug("bmi", 3)` into `r g("bmi", 3)` by conditioning on a mediator. A student
  who omits the enzymes gets a substantively different and arguably more
  defensible model (BMI coefficient
  `r fmt(coef(lm(reformulate(setdiff(predictors, c("alt", "ast")), outcome), data = dat_model))["bmi"], 4)`,
  R-squared
  `r fmt(summary(lm(reformulate(setdiff(predictors, c("alt", "ast")), outcome), data = dat_model))$r.squared, 3)`).
  Both are legitimate; the two lead to opposite conclusions about adiposity.
- **MAJOR: failed diagnostics even after transformation.**
  Heteroscedasticity p `r pv(as.numeric(het))` and non-normality
  p `r pv(as.numeric(nrm))`. The log2 scale is a large improvement over raw GGT
  (skewness `r fmt(skew(dat$ggt), 2)` to `r fmt(skew(dat$log2_ggt), 2)`) but is
  not sufficient. Under HC3 standard errors **AST loses significance**
  (p `r pv(summary(fit)$coefficients["ast", 4])` under OLS to
  p `r pv(rob["ast", 4])` under HC3); every other coefficient keeps its status.
  So the one borderline enzyme finding in the model is exactly the one that
  robust inference kills.
- **MINOR: HDL cholesterol reverses sign** on adjustment
  (`r ug("hdl_cholesterol")` to `r g("hdl_cholesterol")`, p `r gp("hdl_cholesterol")`),
  driven by sex and triglycerides. Borderline either way; a good discussion point
  but an easy one to get wrong.
- **MINOR: smoking looks like a strong crude predictor and is entirely a sex
  effect.** Current smokers are
  `r fmt(100 * mean(dat_model$sex[dat_model$smoking_status == "Current"] == "Male"), 0)`%
  male. Adding sex alone moves the current-smoking coefficient from
  `r fmt(ue("smoking_statusCurrent"), 3)` to
  `r fmt(coef(lm(log2_ggt ~ smoking_status + sex, data = dat_model))["smoking_statusCurrent"], 3)`.
- **MINOR: alcohol has a very heavy right tail** (median
  `r fmt(median(dat_model$alcohol_g_per_week), 1)` g/week, maximum
  `r max(dat_model$alcohol_g_per_week)`, with
  `r sum(dat_model$alcohol_g_per_week == 0)` exact zeros). It is the source of
  the highest-leverage observations (maximum hat value
  `r fmt(max(hatvalues(fit)), 3)`), though none is influential. A student might
  reasonably log or categorise it; nobody will be penalised for using it raw.
- **MINOR: complete-case loss is small** - `r nrow(dat_model)` of `r nrow(dat)`
  (`r fmt(100 * n_dropped / nrow(dat), 1)`%), from `regular_exercise` and
  `alcohol_g_per_week` only.
- No separation issues (linear model), no sparse-cell problems, and no predictor
  with heavy missingness.
