---
title: "Brachial-ankle pulse-wave velocity in a Japanese health-check cohort"
subtitle: "Dataset 9: Liver enzymes and arterial stiffness in a Japanese health-check cohort"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "ggt_arterial_stiffness.csv"),
                    show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names. The catalogue suggests
# 18 predictors; 11 are carried forward. The cut is justified in Methods.
outcome    <- "pwv_max"
predictors <- c("age_yrs", "sex", "bmi", "systolic_bp", "smoking_status",
                "alcohol_g_per_week", "regular_exercise", "fasting_glucose",
                "triglycerides", "hdl_cholesterol", "egfr", "log2_ggt")

# The full catalogue list, used once to document the collinearity that motivated
# the cut. No inference is drawn from this model.
predictors_all <- c("age_yrs", "sex", "bmi", "systolic_bp", "diastolic_bp",
                    "smoking_status", "alcohol_g_per_week", "regular_exercise",
                    "fasting_glucose", "uric_acid", "total_cholesterol",
                    "hdl_cholesterol", "ldl_cholesterol", "triglycerides",
                    "egfr", "log2_ggt", "alt", "ast")

lab <- function(x, l) { attr(x, "label") <- l; x }

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex = lab(factor(sex, levels = c("male", "female"), labels = c("Male", "Female")),
              "Sex"),
    smoking_status = lab(factor(smoking_status,
                                levels = c("never", "former", "current"),
                                labels = c("Never", "Former", "Current")),
                         "Smoking status"),
    regular_exercise = lab(factor(regular_exercise, levels = c(0, 1),
                                  labels = c("No", "Yes")),
                           "Takes regular exercise"),
    age_yrs            = lab(age_yrs, "Age (years)"),
    bmi                = lab(bmi, "BMI (kg/m2)"),
    systolic_bp        = lab(systolic_bp, "Systolic BP (mmHg)"),
    diastolic_bp       = lab(diastolic_bp, "Diastolic BP (mmHg)"),
    alcohol_g_per_week = lab(alcohol_g_per_week, "Alcohol (g/week)"),
    fasting_glucose    = lab(fasting_glucose, "Fasting glucose (mg/dL)"),
    uric_acid          = lab(uric_acid, "Uric acid (mg/dL)"),
    total_cholesterol  = lab(total_cholesterol, "Total cholesterol (mg/dL)"),
    triglycerides      = lab(triglycerides, "Triglycerides (mg/dL)"),
    hdl_cholesterol    = lab(hdl_cholesterol, "HDL cholesterol (mg/dL)"),
    ldl_cholesterol    = lab(ldl_cholesterol, "LDL cholesterol (mg/dL)"),
    egfr               = lab(egfr, "eGFR (mL/min/1.73m2)"),
    log2_ggt           = lab(log2_ggt, "Gamma-GT (log2 U/L)"),
    alt                = lab(alt, "ALT (U/L)"),
    ast                = lab(ast, "AST (U/L)"),
    pwv_max            = lab(pwv_max, "baPWV (cm/s)")
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
pv  <- function(p) ifelse(p < 0.001, "<0.001", paste0("= ", fmt(p, 3)))
```

```{r collinearity-evidence}
# Two documented redundancies in the suggested predictor list, quantified once.
bp_r <- cor(dat$systolic_bp, dat$diastolic_bp)

# Friedewald: total cholesterol ~ LDL + HDL + triglycerides/5.
fw  <- lm(total_cholesterol ~ ldl_cholesterol + hdl_cholesterol + triglycerides,
          data = dat)
fw_r2   <- summary(fw)$r.squared
fw_coef <- coef(fw)

dat_all <- dat |> select(all_of(c(outcome, predictors_all))) |> drop_na()
fit_all <- lm(reformulate(predictors_all, outcome), data = dat_all)
cc_all  <- as.data.frame(check_collinearity(fit_all))
vall <- function(term, d = 1) fmt(cc_all$VIF[cc_all$Term == term], d)

enz_r <- cor(dat$alt, dat$ast)
```

# Introduction

Between 2008 and 2010 a general health-check programme in Japan enrolled
`r nrow(dat)` men and women aged `r min(dat$age_yrs)` to `r max(dat$age_yrs)`
years and measured brachial-ankle pulse-wave velocity (baPWV) with an automatic
waveform analyser alongside a standard panel of anthropometric, haemodynamic and
biochemical measurements; the cohort was assembled to ask whether serum
gamma-glutamyltranspeptidase (GGT) marks arterial disease, and is reported in
*BMJ Open* 2014;4:e005413. Pulse-wave velocity is the reference non-invasive
measure of arterial stiffness and predicts cardiovascular events independently
of blood pressure, so identifying which routinely collected characteristics
track with it is a natural question for a screening programme.

**Aim.** Among these health-check attendees, we examined which of age, sex, body
mass index, systolic blood pressure, smoking status, alcohol intake, regular
exercise, fasting glucose, triglycerides, HDL cholesterol, kidney function and
log2 gamma-GT are independently associated with baPWV, mutually adjusting for the
others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** A cross-sectional Japanese health-check cohort of `r nrow(dat)`
adults attending a medical health check-up programme that included ultrasound of
the liver, a fasting blood panel and automated measurement of brachial-ankle
pulse-wave velocity and the ankle-brachial index. Attendance was voluntary, so
this is a self-selected screening population rather than a community sample.

**Variables.** The outcome is `pwv_max`, brachial-ankle pulse-wave velocity in
cm/s (the higher of the two legs), treated as continuous; higher values mean
stiffer arteries. Predictors are age (years), sex (reference Male, the larger
group at `r sum(dat$sex == "Male")` of `r nrow(dat)`), BMI (kg/m2), systolic
blood pressure (mmHg), smoking status (three levels, reference Never, the
largest group at `r sum(dat$smoking_status == "Never")`), alcohol intake
(g/week), regular exercise (reference No, the unexposed and larger group),
fasting glucose (mg/dL), triglycerides (mg/dL), HDL cholesterol (mg/dL),
estimated glomerular filtration rate (mL/min/1.73m2) and gamma-GT on a log2
scale (so a one-unit increase is a doubling of GGT).

**The predictor set was cut from 18 to `r length(predictors)` before any model
was fitted.** With `r nrow(dat_model)` complete cases and
`r length(coef(lm(reformulate(predictors, outcome), data = dat_model))) - 1L`
estimated parameters this model has about
`r fmt(nrow(dat_model) / (length(coef(lm(reformulate(predictors, outcome), data = dat_model))) - 1L), 0)`
observations per parameter, so the *arithmetic* budget of 10-15 observations per
parameter would in principle have tolerated all 18. The cut was made for two
other reasons. First, **redundancy**: the 18 suggested predictors contain three
sets of near-duplicates. Systolic and diastolic pressure correlate at
r = `r fmt(bp_r, 2)`, and in a model containing both their variance inflation
factors are `r vall("systolic_bp")` and `r vall("diastolic_bp")`. The four lipid
measures are worse: regressing total cholesterol on LDL, HDL and triglycerides
reproduces it almost exactly (R-squared `r fmt(fw_r2, 3)`, with coefficients
`r fmt(fw_coef["ldl_cholesterol"], 2)`,
`r fmt(fw_coef["hdl_cholesterol"], 2)` and
`r fmt(fw_coef["triglycerides"], 2)` - that is, the Friedewald identity
total = LDL + HDL + triglycerides/5), and with all four in the model the VIFs
reach `r vall("total_cholesterol")` for total cholesterol and
`r vall("ldl_cholesterol")` for LDL. The three liver enzymes are the third set
(ALT and AST correlate at r = `r fmt(enz_r, 2)`). Second, **interpretability**: a
table of 18 mutually adjusted coefficients cannot be read as an answer to "which
predictors survive adjustment", which is the question this analysis is asked to
answer. We therefore kept one variable per construct: systolic pressure (the
stronger correlate of baPWV, r = `r fmt(cor(dat$systolic_bp, dat$pwv_max), 2)`
versus `r fmt(cor(dat$diastolic_bp, dat$pwv_max), 2)` for diastolic) in place of
both pressures; HDL cholesterol and triglycerides (the two metabolic-syndrome
lipid components, VIFs `r vall("hdl_cholesterol")` and `r vall("triglycerides")`
even in the 18-variable model) in place of all four lipids; and log2 gamma-GT,
the enzyme of interest in the source paper, in place of all three enzymes. Uric
acid was dropped because it has no distinct a priori role once adiposity and
kidney function are in the model (it correlates
r = `r fmt(cor(dat$uric_acid, dat$egfr), 2)` with eGFR and
r = `r fmt(cor(dat$uric_acid, dat$triglycerides), 2)` with triglycerides).

**Systolic pressure is retained deliberately but is not innocent.** baPWV is
computed from a cuff-based pressure waveform, and arterial stiffness and blood
pressure are mutually causal: stiff arteries raise systolic pressure, and higher
distending pressure raises the measured velocity in an elastic tube. Adjusting
for systolic pressure is therefore partly adjustment for a common cause and
partly adjustment for a mediator or even for part of the measurement itself. We
report the main model with systolic pressure included, and a sensitivity model
omitting it, and interpret the two together.

**Statistical analysis.** `lm()`, justified because the outcome is a continuous
measurement in cm/s. We fitted a univariable model for each predictor first,
then a single multivariable model containing all of them. Predictors were chosen
in advance on subject-matter grounds, not by significance. Significance threshold
0.05; estimates are reported with 95% confidence intervals. Assumptions were
checked with `performance::check_model()`, `check_heteroscedasticity()` and
`check_collinearity()`, and because non-constant variance was flagged we
additionally report HC3 heteroscedasticity-consistent standard errors. Analysis
used `r R.version.string` with tidyverse, gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped). The outcome itself is complete; the
whole loss comes from `regular_exercise`
(`r sum(is.na(dat$regular_exercise))` missing) and `alcohol_g_per_week`
(`r sum(is.na(dat$alcohol_g_per_week))` missing), which do not overlap. Dropping
total cholesterol from the predictor set incidentally saved a further
`r sum(is.na(dat$total_cholesterol))` rows. Only
`r sum(complete.cases(dat_raw))` of `r nrow(dat)` rows are complete on *every*
column in the file.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(missing = "no") |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

The cohort is `r fmt(100 * mean(dat_model$sex == "Male"), 0)`% male with a median
age of `r median(dat_model$age_yrs)` years and a median BMI of
`r fmt(median(dat_model$bmi), 1)` kg/m2, which is lean by Western standards and
consistent with a Japanese screening population. Median baPWV is
`r format(median(dat_model$pwv_max), big.mark = ",")` cm/s with a long right tail
reaching `r format(max(dat_model$pwv_max), big.mark = ",")` cm/s, and blood
pressure, glucose and lipids all sit close to the middle of their normal ranges -
these are people who feel well enough to volunteer for a check-up.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Pulse-wave velocity against log2 gamma-GT, by sex, with linear smooths."}
ggplot(dat_model, aes(log2_ggt, pwv_max, colour = sex)) +
  geom_point(alpha = 0.35, size = 1.4) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.9) +
  scale_colour_manual(values = c(Male = "#2E6F95", Female = "#C46B4A")) +
  labs(x = "Gamma-GT (log2 U/L)", y = "baPWV (cm/s)", colour = "Sex")
```

baPWV rises gently with log2 gamma-GT in both sexes, but the scatter is very wide
and the slope is visibly shallow relative to the spread; the men sit higher than
the women across the whole range of GGT, which is the first hint that the crude
GGT-stiffness association may be carrying sex and everything correlated with it.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv

utid <- map_dfr(predictors, function(v) {
  tidy(lm(reformulate(v, outcome), data = dat_model), conf.int = TRUE) |>
    filter(term != "(Intercept)")
})
ug <- function(term, d = 1) {
  r <- utid[utid$term == term, ]
  sprintf("%s cm/s (95%% CI %s to %s)", fmt(r$estimate, d),
          fmt(r$conf.low, d), fmt(r$conf.high, d))
}
up <- function(term) pv(utid$p.value[utid$term == term])
```

Age and systolic pressure dominate the crude picture. Each additional year of age
is associated with `r ug("age_yrs")` higher baPWV (p `r up("age_yrs")`), roughly
`r fmt(10 * utid$estimate[utid$term == "age_yrs"], 0)` cm/s per decade, and each
additional mmHg of systolic pressure with `r ug("systolic_bp")`
(p `r up("systolic_bp")`). Each doubling of gamma-GT is associated with
`r ug("log2_ggt")` higher baPWV (p `r up("log2_ggt")`) - a substantial crude
association, and the one the source paper was built around. Higher fasting
glucose (`r ug("fasting_glucose")` per mg/dL, p `r up("fasting_glucose")`) and
higher triglycerides (`r ug("triglycerides")` per mg/dL,
p `r up("triglycerides")`) also mark stiffer arteries, while better kidney
function is associated with less stiffness (`r ug("egfr")` per unit of eGFR,
p `r up("egfr")`). Women average `r ug("sexFemale")` relative to men
(p `r up("sexFemale")`), and former smokers `r ug("smoking_statusFormer")`
relative to never-smokers (p `r up("smoking_statusFormer")`) - though current
smokers do not differ (`r ug("smoking_statusCurrent")`,
p `r up("smoking_statusCurrent")`), which is not the ordering one would expect
from a causal smoking effect. BMI (`r ug("bmi")`, p `r up("bmi")`), alcohol
(`r ug("alcohol_g_per_week")`, p `r up("alcohol_g_per_week")`) and regular
exercise (`r ug("regular_exerciseYes")`, p `r up("regular_exerciseYes")`) show no
crude association.

## Why adjustment is needed

**Age** is the single most important confounder in this dataset. It is
associated with baPWV more strongly than anything else except pressure
(r = `r fmt(cor(dat_model$age_yrs, dat_model$pwv_max), 2)`), and it is strongly
associated with kidney function
(r = `r fmt(cor(dat_model$age_yrs, dat_model$egfr), 2)` with eGFR) and more
weakly with fasting glucose
(r = `r fmt(cor(dat_model$age_yrs, dat_model$fasting_glucose), 2)`). The crude
eGFR association is therefore substantially an age association wearing a kidney's
clothes: older attendees have both lower eGFR and stiffer arteries for reasons
that have little to do with the kidney itself. Note that age is *not* a
confounder of the pressure association here - age and systolic pressure correlate
only r = `r fmt(cor(dat_model$age_yrs, dat_model$systolic_bp), 2)` in this
screening cohort - which is why the two of them together carry so much of the
model without competing.

**Sex** confounds the gamma-GT association. Men in this cohort have far higher
GGT than women (median `r fmt(median(dat_model$log2_ggt[dat_model$sex == "Male"]), 2)`
versus `r fmt(median(dat_model$log2_ggt[dat_model$sex == "Female"]), 2)` on the
log2 scale, i.e. about
`r fmt(2^(median(dat_model$log2_ggt[dat_model$sex == "Male"]) - median(dat_model$log2_ggt[dat_model$sex == "Female"])), 1)`-fold
higher on the raw scale) because they drink far more (median alcohol
`r fmt(median(dat_model$alcohol_g_per_week[dat_model$sex == "Male"]), 0)` versus
`r fmt(median(dat_model$alcohol_g_per_week[dat_model$sex == "Female"]), 0)` g per
week), and men also have higher baPWV. **Alcohol** is the third variable that
must be handled explicitly, for the same reason from the other direction: it
drives GGT (r = `r fmt(cor(dat_model$alcohol_g_per_week, dat_model$log2_ggt), 2)`
with log2 GGT) and is itself associated with cardiovascular risk, so a crude
GGT coefficient is partly a drinking coefficient. **Systolic pressure**
confounds - or mediates - almost everything, since adiposity, glucose and
smoking all raise pressure and pressure raises measured baPWV. Mutual adjustment
lets each coefficient be read as a comparison between attendees alike on the
other measured characteristics, with the pressure caveat kept in view.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv

tid <- tidy(fit, conf.int = TRUE)
g <- function(term, d = 1) {
  r <- tid[tid$term == term, ]
  sprintf("%s cm/s (95%% CI %s to %s)", fmt(r$estimate, d),
          fmt(r$conf.low, d), fmt(r$conf.high, d))
}
gp <- function(term) pv(tid$p.value[tid$term == term])
```

```{r fit-stats}
s <- summary(fit)
tibble(
  Statistic = c("Complete-case N", "Parameters (excl. intercept)",
                "Observations per parameter", "R-squared", "Adjusted R-squared",
                "Residual SE (cm/s)", "SD of outcome (cm/s)"),
  Value = c(as.character(nrow(dat_model)),
            as.character(length(coef(fit)) - 1L),
            fmt(nrow(dat_model) / (length(coef(fit)) - 1L), 1),
            fmt(s$r.squared, 3), fmt(s$adj.r.squared, 3),
            fmt(s$sigma, 1), fmt(sd(dat_model$pwv_max), 1))
) |>
  gt() |>
  tab_header(title = md("**Model fit: multivariable linear model for baPWV**"))
```

The model explains `r fmt(100 * s$r.squared, 1)`% of the variance in baPWV
(adjusted R-squared `r fmt(s$adj.r.squared, 3)`), which is high for an
observational model of this kind and is almost entirely the work of age and
systolic pressure. Residual SE is `r fmt(s$sigma, 1)` cm/s against an outcome SD
of `r fmt(sd(dat_model$pwv_max), 1)` cm/s.

Among attendees of the same sex, body mass, blood pressure, smoking, drinking,
exercise, glucose, lipids, kidney function and gamma-GT, each additional year of
age is associated with `r g("age_yrs")` higher baPWV (p `r gp("age_yrs")`), and
among attendees alike on everything else each additional mmHg of systolic
pressure with `r g("systolic_bp")` (p `r gp("systolic_bp")`). Two coefficients
change character on adjustment. BMI, which showed no crude association, becomes
`r g("bmi")` per kg/m2 (p `r gp("bmi")`) - *lower* stiffness at higher BMI once
pressure and the metabolic variables are held constant. Current smoking, which
was null crudely, becomes `r g("smoking_statusCurrent")` relative to never
smoking (p `r gp("smoking_statusCurrent")`), while former smoking is
`r g("smoking_statusFormer")` (p `r gp("smoking_statusFormer")`). Higher fasting
glucose remains associated with stiffer arteries (`r g("fasting_glucose")` per
mg/dL, p `r gp("fasting_glucose")`), higher alcohol intake with slightly less
stiffness (`r g("alcohol_g_per_week")` per g/week,
p `r gp("alcohol_g_per_week")`), and better kidney function with less stiffness
(`r g("egfr")`, p `r gp("egfr")`).

**Gamma-GT does not survive adjustment.** Each doubling of GGT is associated with
`r g("log2_ggt")` (p `r gp("log2_ggt")`), against a crude `r ug("log2_ggt")`: the
point estimate has fallen by about
`r fmt(100 * (1 - tid$estimate[tid$term == "log2_ggt"] / utid$estimate[utid$term == "log2_ggt"]), 0)`%
and the confidence interval now includes zero. Sex, triglycerides, HDL
cholesterol and regular exercise are also not independently associated. These are
associations among the attendees observed, not effects of intervening.

```{r sensitivity-nobp}
fit_nobp <- lm(reformulate(setdiff(predictors, "systolic_bp"), outcome),
               data = dat_model)
tid2 <- tidy(fit_nobp, conf.int = TRUE)
g2 <- function(term, d = 1) {
  r <- tid2[tid2$term == term, ]
  sprintf("%s cm/s (95%% CI %s to %s)", fmt(r$estimate, d),
          fmt(r$conf.low, d), fmt(r$conf.high, d))
}
gp2 <- function(term) pv(tid2$p.value[tid2$term == term])

tibble(
  Model = c("Main model (systolic BP included)",
            "Sensitivity model (systolic BP omitted)"),
  `log2 gamma-GT coefficient` = c(g("log2_ggt"), g2("log2_ggt")),
  `p` = c(gsub("^= ", "", gp("log2_ggt")), gsub("^= ", "", gp2("log2_ggt"))),
  `R-squared` = c(fmt(summary(fit)$r.squared, 3), fmt(summary(fit_nobp)$r.squared, 3))
) |>
  gt() |>
  tab_header(title = md("**Sensitivity: gamma-GT with and without systolic pressure in the model**"))
```

This is the honest version of the gamma-GT finding. With systolic pressure in the
model, GGT contributes `r g("log2_ggt")` (p `r gp("log2_ggt")`); with pressure
removed it contributes `r g2("log2_ggt")` (p `r gp2("log2_ggt")`), roughly
`r fmt(tid2$estimate[tid2$term == "log2_ggt"] / tid$estimate[tid$term == "log2_ggt"], 1)`
times larger and clearly significant. Because pressure and stiffness are mutually
causal - and because baPWV is measured *from* a pressure waveform - neither
number is the "true" GGT effect. The pressure-adjusted estimate is
conservative to the point of being an over-adjustment; the unadjusted-for-pressure
estimate contains whatever part of the GGT-stiffness link runs through blood
pressure. The defensible statement is that any independent contribution of GGT to
arterial stiffness in this cohort is small, and that most of the crude
association is explained by age, sex, drinking and pressure.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

Three predictors move a long way. **Sex** collapses from `r ug("sexFemale")`
crude (p `r up("sexFemale")`) to `r g("sexFemale")` adjusted
(p `r gp("sexFemale")`) - the entire apparent sex difference in arterial
stiffness is explained by the other measured characteristics, and the coefficient
crosses from clearly significant to null and even changes sign. Systolic pressure
does almost all of that work on its own: adding it alone to the crude model moves
the female coefficient from `r fmt(utid$estimate[utid$term == "sexFemale"], 1)`
to
`r fmt(coef(lm(pwv_max ~ sex + systolic_bp, data = dat_model))["sexFemale"], 1)` cm/s,
because the women in this cohort have median systolic pressure
`r fmt(median(dat_model$systolic_bp[dat_model$sex == "Female"]), 1)` mmHg against
`r fmt(median(dat_model$systolic_bp[dat_model$sex == "Male"]), 1)` mmHg in the
men. **Gamma-GT** falls from
`r ug("log2_ggt")` to `r g("log2_ggt")` and likewise loses significance
(p `r up("log2_ggt")` to p `r gp("log2_ggt")`), which is confounding by age, sex
and alcohol exactly as anticipated. **BMI** moves in the opposite direction, and
*through* the null: from `r ug("bmi")` crude (p `r up("bmi")`) to `r g("bmi")`
adjusted (p `r gp("bmi")`). This sign reversal is not a paradox but a classic
suppression pattern: heavier attendees have higher blood pressure, glucose and
triglycerides, all of which raise baPWV, so crudely BMI looks neutral or slightly
harmful; once those downstream consequences are held fixed, the residual
association of body mass with measured stiffness is negative, partly because
baPWV in a larger limb is measured over a longer, more muscular path.
**Triglycerides** and **eGFR** both attenuate strongly toward the null
(triglycerides `r ug("triglycerides")` to `r g("triglycerides")`, losing
significance; eGFR `r ug("egfr")` to `r g("egfr")`, retaining it but at about
`r fmt(100 * tid$estimate[tid$term == "egfr"] / utid$estimate[utid$term == "egfr"], 0)`%
of the crude magnitude) - both were largely age.

Two things barely move. **Systolic pressure** goes from `r ug("systolic_bp")` to
`r g("systolic_bp")`, essentially unchanged, and **age** from `r ug("age_yrs")` to
`r g("age_yrs")`: these two are not explained by anything else in the file, which
is why they carry the model. Finally, **smoking changes which level is
significant**: crudely it is former smoking that differs from never
(`r ug("smoking_statusFormer")`, p `r up("smoking_statusFormer")`) while current
smoking does not (`r ug("smoking_statusCurrent")`, p `r up("smoking_statusCurrent")`);
after adjustment the pattern inverts, with current smoking at
`r g("smoking_statusCurrent")` (p `r gp("smoking_statusCurrent")`) and former
smoking null (`r g("smoking_statusFormer")`, p `r gp("smoking_statusFormer")`).
The crude former-smoker signal was age and sex - former smokers are the oldest
group - and once that is removed the dose-response one expects from smoking
appears.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r het-check}
het <- check_heteroscedasticity(fit)
print(het)
nrm <- check_normality(fit)
print(nrm)
```

```{r collin}
cc <- as.data.frame(check_collinearity(fit))
check_collinearity(fit)
```

```{r dichot-check, include=FALSE}
# Feasibility only: what the >= 1400 cm/s dichotomisation would look like.
dat_dich <- dat_model |> mutate(pwv_high = as.integer(pwv_max >= 1400))
fit_dich <- glm(reformulate(predictors, "pwv_high"), data = dat_dich,
                family = binomial)
dich_events <- sum(dat_dich$pwv_high)
dich_par    <- length(coef(fit_dich)) - 1L
dich_tjur   <- as.numeric(r2_tjur(fit_dich))
dich_bins   <- attr(binned_residuals(fit_dich), "resid_ok")
```

```{r robust-se}
rob   <- lmtest::coeftest(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
robci <- lmtest::coefci(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
tibble(
  Term = rownames(rob),
  Estimate = fmt(rob[, 1], 2),
  `SE (OLS)` = fmt(summary(fit)$coefficients[, 2], 2),
  `SE (HC3)` = fmt(rob[, 2], 2),
  `95% CI (HC3)` = paste0(fmt(robci[, 1], 1), " to ", fmt(robci[, 2], 1)),
  `p (HC3)` = fmt(rob[, 4], 3)
) |>
  filter(Term != "(Intercept)") |>
  gt() |>
  tab_header(title = md("**Table 3. HC3 heteroscedasticity-consistent standard errors.**"))
```

**Collinearity is not a problem in the reduced model.** The largest VIF is
`r fmt(max(cc$VIF), 2)` (`r cc$Term[which.max(cc$VIF)]`), and every VIF is below
`r fmt(ceiling(max(cc$VIF) * 2) / 2, 1)`; nothing approaches the conventional
threshold of 5. This is the direct pay-off of the a-priori cut: in the
18-predictor version of this model the largest VIF is
`r vall("total_cholesterol")`.

**Variance is not constant and residuals are not normal.** The
residual-versus-fitted panel fans out clearly as fitted baPWV rises, and
`check_heteroscedasticity()` rejects (p `r pv(as.numeric(het))`); the normal-QQ
panel shows a heavy right tail and `check_normality()` also rejects
(p `r pv(as.numeric(nrm))`). Both are driven by the right skew of baPWV
(skewness `r fmt(mean((dat_model$pwv_max - mean(dat_model$pwv_max))^3) / sd(dat_model$pwv_max)^3, 2)`):
the stiffest arteries are also the most variable, so the spread of the outcome
grows with its mean. The posterior-predictive check shows the same thing - the
fitted symmetric distribution cannot reproduce the tail above about 2,000 cm/s.
Table 3 therefore reports HC3 robust standard errors. They are modestly wider for
the strongest predictors (age: OLS SE
`r fmt(summary(fit)$coefficients["age_yrs", 2], 2)` versus HC3
`r fmt(rob["age_yrs", 2], 2)`; systolic pressure
`r fmt(summary(fit)$coefficients["systolic_bp", 2], 2)` versus
`r fmt(rob["systolic_bp", 2], 2)`) but **no coefficient changes its significance
status**, so the substantive conclusions stand. Modelling `log(pwv_max)` improves
R-squared slightly (to
`r fmt(summary(lm(reformulate(predictors, "log(pwv_max)"), data = dat_model))$r.squared, 3)`)
but does **not** fix either check, so the skew is not a simple multiplicative
problem.

**No single observation drives the fit.** The largest Cook's distance is
`r fmt(max(cooks.distance(fit)), 3)` and the largest leverage
`r fmt(max(hatvalues(fit)), 3)`, both far inside the usual contours, and the
influence panel shows no point outside them.

# Discussion

In this Japanese health-check cohort, arterial stiffness measured as baPWV is
overwhelmingly a function of age and systolic blood pressure: each decade of age
is associated with about
`r fmt(10 * tid$estimate[tid$term == "age_yrs"], 0)` cm/s and each 10 mmHg of
systolic pressure with about
`r fmt(10 * tid$estimate[tid$term == "systolic_bp"], 0)` cm/s, holding everything
else constant, and together with the remaining predictors these account for
`r fmt(100 * s$r.squared, 0)`% of the variance. The crude association between
gamma-GT and stiffness - `r ug("log2_ggt")` per doubling of GGT - does not
survive mutual adjustment (`r g("log2_ggt")`), and the residual signal that does
appear when blood pressure is left out of the model (`r g2("log2_ggt")`) is
plausibly routed through pressure rather than independent of it. Practically,
this argues that GGT adds little to age and blood pressure for identifying stiff
arteries in a screening setting, and that a report claiming a liver-enzyme
contribution to arterial stiffness needs to be explicit about what it did with
blood pressure.

**Limitations.** The design is cross-sectional, so nothing here establishes
temporal order; stiff arteries and raised GGT could both be consequences of a
shared metabolic process, and reverse causation cannot be excluded. Adjusting for
systolic pressure is not clean confounder control: pressure is on the causal path
between most of these risk factors and stiffness, and baPWV is derived from a
pressure waveform, so the pressure-adjusted coefficients are probably
over-adjusted while the unadjusted-for-pressure ones are under-adjusted - the
sensitivity table brackets rather than resolves this. Health-check attendees are
a self-selected group who are healthier and better resourced than the general
Japanese population, which compresses the range of every risk factor and will
attenuate associations relative to a community sample. Residual variance is
non-constant and residuals are strongly right-skewed; HC3 standard errors repair
inference but not the model's inability to predict the stiffest attendees.
Alcohol is self-reported in g/week with `r sum(dat_model$alcohol_g_per_week == 0)`
attendees reporting exactly zero, so misreporting is likely and probably
differential by sex. Antihypertensive, lipid-lowering and glucose-lowering
medication is not recorded at all, which is a serious unmeasured confounder:
treated hypertensives appear in the data with near-normal pressures but stiff
arteries. Diet, socioeconomic position and diabetes duration are likewise absent.
`r n_dropped` of `r nrow(dat)` rows were dropped for missing exercise or alcohol.
All estimates are associations, not causal effects.

**Future work.** Refit on the log scale with medication use and a repeated baPWV
measurement, and treat blood pressure explicitly as a mediator with a formal
decomposition rather than as a covariate to be switched on and off.

# Feasibility notes

*NOT part of a student report.*

- **Verdict: USABLE.** This is one of the better linear outcomes in the
  catalogue. `pwv_max` is complete (0 missing), well spread, and has two
  genuinely strong predictors, so a student's model will "work" and produce an
  R-squared of about `r fmt(s$r.squared, 2)`. The interesting teaching content is
  that the headline exposure of the source paper (GGT) *does not* survive.
- **MAJOR: 18 suggested predictors is too many and the budget rule will not stop
  a student.** With `r nrow(dat_model)` complete cases, even 20 parameters gives
  ~44 observations per parameter, so the 10-15 observations-per-parameter rule
  gives no reason to cut. Students will therefore fit all 18 unless the brief
  tells them not to, and the resulting table is unreadable *and* collinear. The
  cut has to be justified on redundancy and interpretability, which is a harder
  argument for a student to construct unaided. Consider trimming the suggested
  list in `meta.json`.
- **MAJOR collinearity if the lipids are all included.** In the 18-predictor
  model the VIFs are `r vall("total_cholesterol")` (total cholesterol),
  `r vall("ldl_cholesterol")` (LDL), `r vall("hdl_cholesterol")` (HDL) and
  `r vall("triglycerides")` (triglycerides). Regressing total cholesterol on the
  other three gives R-squared `r fmt(fw_r2, 3)` with coefficients
  `r fmt(fw_coef["ldl_cholesterol"], 2)` / `r fmt(fw_coef["hdl_cholesterol"], 2)`
  / `r fmt(fw_coef["triglycerides"], 2)`, i.e. the Friedewald identity
  total = LDL + HDL + triglycerides/5 holds to within rounding. Note it is
  *near*-deterministic, not exact: LDL here appears to be directly measured, so
  `total_cholesterol - hdl_cholesterol - triglycerides/5` differs from
  `ldl_cholesterol` by a median of about
  `r fmt(median(with(dat, total_cholesterol - hdl_cholesterol - triglycerides/5 - ldl_cholesterol), na.rm = TRUE), 1)`
  mg/dL. The model does not become singular; it becomes uninterpretable, which is
  worse because nothing warns the student.
- **MAJOR collinearity between the blood pressures.** r = `r fmt(bp_r, 2)`;
  VIFs `r vall("systolic_bp")` and `r vall("diastolic_bp")` when both are
  included. With both in, the diastolic coefficient is essentially zero and the
  systolic one absorbs everything.
- **MAJOR conceptual trap: systolic pressure is partly the outcome.** baPWV is
  computed from a cuff pressure waveform and is pressure-dependent by
  construction. Including systolic pressure drives GGT from
  `r g2("log2_ggt")` (p `r gp2("log2_ggt")`) to `r g("log2_ggt")`
  (p `r gp("log2_ggt")`), which is the difference between "GGT is associated with
  arterial stiffness" and "it is not". Any student conclusion about GGT is
  therefore an artefact of a covariate choice they will not have thought about.
  This should be flagged in `meta.json` cautions.
- **MAJOR: failed diagnostics.** Heteroscedasticity p `r pv(as.numeric(het))` and
  non-normality of residuals p `r pv(as.numeric(nrm))`, both from the right skew
  of baPWV (skewness
  `r fmt(mean((dat_model$pwv_max - mean(dat_model$pwv_max))^3) / sd(dat_model$pwv_max)^3, 2)`).
  A log transformation raises R-squared only slightly and does *not* clear either
  test, so there is no easy fix; students must report a failed check and use
  robust SEs. Reassuringly, no coefficient changes significance under HC3.
- **MINOR: BMI reverses sign on adjustment** (`r ug("bmi")` crude to `r g("bmi")`
  adjusted, p `r gp("bmi")`). This is a suppression effect via blood pressure and
  the metabolic variables, and it is a good teaching moment, but a weak student
  will report "obesity protects the arteries".
- **MINOR: smoking swaps which level is significant** between the crude and
  adjusted models (former significant crude, current significant adjusted). With
  `r sum(dat_model$smoking_status == "Current")` current and
  `r sum(dat_model$smoking_status == "Former")` former smokers there is enough
  data for the contrast, but note that smoking is almost entirely a male
  behaviour here (`r sum(dat$sex == "Female" & dat$smoking_status == "Current")`
  female current smokers out of `r sum(dat$sex == "Female")` women), so smoking
  and sex are heavily entangled.
- **MINOR: complete-case loss is small** - `r nrow(dat_model)` of `r nrow(dat)`
  (`r fmt(100 * n_dropped / nrow(dat), 1)`%), driven entirely by
  `regular_exercise` (`r sum(is.na(dat$regular_exercise))`) and
  `alcohol_g_per_week` (`r sum(is.na(dat$alcohol_g_per_week))`). No predictor
  here silently costs a large share of the sample.
- **MINOR: no medication variables.** Antihypertensives, statins and
  glucose-lowering drugs are not in the file. For an outcome this dependent on
  blood pressure that is a substantial unmeasured confounder and should be in the
  cautions.
- **On the >= 1400 cm/s dichotomisation offered in the catalogue: it works, but
  it is strictly worse.** `pwv_max >= 1400` gives
  `r sum(dat_raw$pwv_max >= 1400)` / `r nrow(dat_raw)`
  (`r fmt(100 * mean(dat_raw$pwv_max >= 1400), 1)`%) events, and in the
  complete-case sample `r dich_events` events for `r dich_par` parameters, i.e.
  `r fmt(dich_events / dich_par, 1)` events per parameter - comfortably feasible,
  no separation, and the logistic model recovers the same ordering of predictors
  (age and systolic pressure dominant, Tjur's R-squared `r fmt(dich_tjur, 2)`).
  But the threshold is arbitrary rather than clinical, it throws away a
  well-measured continuous outcome, and its binned-residual check also fails
  (only `r fmt(100 * dich_bins, 0)`% of bins inside the error bounds). The
  catalogue is right to present `pwv_max` as linear; the "either" designation in
  `meta.json` should not be read as an invitation.
- No separation issues arise anywhere in the linear model, and there are no
  deterministic relationships between `pwv_max` and any shipped column
  (`abi_max` is the closest structurally related variable and correlates
  r = `r fmt(cor(dat$abi_max, dat$pwv_max, use = "complete.obs"), 2)` with it,
  i.e. not at all).
