---
title: "Measured hypertension in Kenyan outpatient attenders"
subtitle: "Dataset 5: Kenya hypertension screening survey"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "kenya_hypertension.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "hypertension"
predictors <- c("age_group", "sex", "education", "bmi_category",
                "current_smoker", "current_alcohol", "adequate_activity",
                "diabetes")

lab <- function(x, l) { attr(x, "label") <- l; x }

dat <- dat_raw |>
  mutate(
    # Event level is "Yes"; a two-level factor with "No" first makes glm() model
    # the probability of hypertension, and lets Table 1 be grouped by outcome.
    hypertension = lab(factor(hypertension, levels = c(0, 1), labels = c("No", "Yes")),
                       "Hypertension on measured BP"),
    # Deliberate reference levels: the largest band, or the clinically
    # "unexposed" group. 26-35 rather than 18-25 is the age reference because
    # 18-25 holds only 101 people, of whom just 6 are hypertensive; using it as
    # the reference would make every age odds ratio badly imprecise.
    age_group = lab(fct_relevel(factor(age_group),
                                "26-35", "18-25", "36-45", "46-55", "56-65", ">65"),
                    "Age band (years)"),
    sex = lab(factor(sex, levels = c("female", "male"), labels = c("Female", "Male")),
              "Sex"),
    # No schooling is the reference so higher categories read as "more education
    # than the reference".
    education = lab(fct_relevel(factor(education),
                                "No School", "Primary", "Secondary", "Tertiary"),
                    "Education"),
    # Healthy weight is both the largest category and the clinical comparator.
    bmi_category = lab(fct_relevel(factor(bmi_category),
                                   "Healthy", "Underweight", "Overweight", "Obese"),
                       "BMI category (WHO)"),
    current_smoker = lab(factor(current_smoker, levels = c(0, 1), labels = c("No", "Yes")),
                         "Current smoker"),
    current_alcohol = lab(factor(current_alcohol, levels = c(0, 1), labels = c("No", "Yes")),
                          "Current alcohol use"),
    adequate_activity = lab(factor(adequate_activity, levels = c(0, 1), labels = c("No", "Yes")),
                            "Meets activity recommendation"),
    # "unknown" is kept as its own level rather than set to missing: it is a
    # substantive answer (never tested) and dropping it would cost 261 rows.
    diabetes = lab(fct_relevel(factor(diabetes), "no", "yes", "unknown"),
                   "Self-reported diabetes"),
    cardiovascular_disease = lab(fct_relevel(factor(cardiovascular_disease),
                                             "no", "yes", "unknown"),
                                 "Self-reported cardiovascular disease"),
    study_site = lab(fct_relevel(factor(study_site), "public", "private", "faith"),
                     "Facility type"),
    monthly_income = lab(fct_relevel(factor(monthly_income),
                                     "0  - 15,000", "16,000 - 50,000",
                                     "51,000 - 100,000", "100,000 - 200,000",
                                     "Above 200,000"),
                         "Monthly household income (KSh)"),
    prior_htn_diagnosis = lab(factor(prior_htn_diagnosis, levels = c(0, 1),
                                     labels = c("No", "Yes")),
                              "Previously diagnosed with hypertension")
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)
n_events  <- sum(dat_model[[outcome]] == "Yes")

# The income-augmented sample is carried through the whole document so that the
# cost of adding a 31%-missing variable can be quantified rather than asserted.
dat_inc  <- dat |> select(all_of(c(outcome, predictors, "monthly_income"))) |> drop_na()
n_events_inc <- sum(dat_inc[[outcome]] == "Yes")

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
pv  <- function(p) ifelse(p < 0.001, "< 0.001", paste0("= ", fmt(p, 3)))
```

# Introduction

Hypertension is the leading modifiable cardiovascular risk factor in sub-Saharan
Africa, and Kenya's NCD strategy relies heavily on opportunistic screening of
people already attending health facilities. This dataset is a cross-sectional
screening survey of `r nrow(dat_raw)` adults recruited at three types of Kenyan
outpatient facility - a government (public) facility, a private facility and a
faith-based mission facility - in which every participant had blood pressure
measured three times and completed a WHO STEPS-style questionnaire on
demography, behaviour and self-reported morbidity. Knowing which attender
characteristics predict a raised measured pressure tells a clinic who to
prioritise when it cannot screen everybody.

**Aim.** Among these Kenyan outpatient attenders, we examined which of age band,
sex, education, BMI category, current smoking, current alcohol use, physical
activity and self-reported diabetes are independently associated with
hypertension on measured blood pressure, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** A cross-sectional facility-based screening survey of
`r nrow(dat_raw)` adults aged 18 and over attending outpatient services at three
Kenyan facilities (`r sum(dat$study_site == "public")` public,
`r sum(dat$study_site == "private")` private,
`r sum(dat$study_site == "faith")` faith-based). Blood pressure was measured
three times and the file carries the average of the three readings.

**Variables.** The outcome is `hypertension`, a binary indicator of raised
measured blood pressure derived from the average of the three readings; the
modelled event is **hypertension = Yes**
(`r sum(dat$hypertension == "Yes")` of `r nrow(dat)`,
`r fmt(100 * mean(dat$hypertension == "Yes"), 1)`%). Predictors are age band
(reference *26-35*); sex (reference *Female*); education (reference *No
School*); WHO BMI category (reference *Healthy*); current smoking (reference
*No*); current alcohol use (reference *No*); meeting the physical-activity
recommendation (reference *No*); and self-reported diabetes (reference *no*,
with *unknown* retained as a substantive third level rather than treated as
missing).

**`systolic_bp` and `diastolic_bp` are excluded by construction.** `hypertension`
in this file *is* the measured average blood pressure crossing the diagnostic
threshold - it is a coarsened version of the two continuous pressures, not a
separate fact about the participant. Putting either pressure on the right-hand
side would produce complete or near-complete separation, absurd odds ratios and
a model that says only "high blood pressure is associated with high blood
pressure". This is the single most likely way to destroy this analysis. The
relationship is not hidden by noise, it is definitional.

**`prior_htn_diagnosis` is excluded from the main model** and examined
separately below. A self-reported earlier diagnosis is a genuinely different
construct - it is about contact with care - but the two variables have a messy
bidirectional relationship: a diagnosis follows from previously high pressure,
and then leads to treatment that lowers today's measured pressure. Without a
medication variable (there is none in this file) that loop cannot be broken, so
it belongs in a sensitivity analysis, not the main model.

**BMI is categorical, not continuous.** Only the WHO `bmi_category` label is
supplied, so no per-unit BMI odds ratio can be reported. Indeed **no predictor
in this file is continuous** - age is banded too - so every estimate below is a
contrast between broad groups and within-band residual confounding is certain.

**Statistical analysis.** Logistic regression via `glm(family = binomial)`,
justified because the outcome is binary; estimates are reported as odds ratios.
We fitted a univariable model for each predictor first, then a single
multivariable model containing all eight. Predictors were chosen in advance on
subject-matter grounds, not by significance. Significance threshold 0.05;
estimates are reported with 95% confidence intervals. Assumptions were checked
with `performance::check_model()`. Analysis used `r R.version.string` with
tidyverse, gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped,
`r fmt(100 * n_dropped / nrow(dat), 1)`%), containing `r n_events` events. The
outcome and four predictors are 100% complete; the losses come from
`bmi_category` (`r sum(is.na(dat$bmi_category))` missing), `education`
(`r sum(is.na(dat$education))`) and `age_group` (`r sum(is.na(dat$age_group))`).
`monthly_income`, missing for `r sum(is.na(dat$monthly_income))` participants
(`r fmt(100 * mean(is.na(dat$monthly_income)), 1)`%), is not in the main model;
the section *The cost of `monthly_income`* below reports what including it would
do.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(
    by = hypertension,
    missing = "no"
  ) |>
  add_n() |>
  add_overall() |>
  modify_caption("**Table 1. Sample characteristics by measured hypertension status.**")
```

```{r table-one-numbers}
prev <- 100 * mean(dat_model$hypertension == "Yes")
```

`r fmt(prev, 1)`% of the analytic sample (`r n_events` of `r nrow(dat_model)`)
had hypertension on measured blood pressure. Hypertensive attenders are much
older (`r fmt(100 * mean(dat_model$age_group[dat_model$hypertension == "Yes"] %in% c("56-65", ">65")), 0)`%
are over 55 versus
`r fmt(100 * mean(dat_model$age_group[dat_model$hypertension == "No"] %in% c("56-65", ">65")), 0)`%
of the non-hypertensive), heavier
(`r fmt(100 * mean(dat_model$bmi_category[dat_model$hypertension == "Yes"] == "Obese"), 0)`%
obese versus
`r fmt(100 * mean(dat_model$bmi_category[dat_model$hypertension == "No"] == "Obese"), 0)`%),
more often male and less often tertiary-educated. Smoking and alcohol use are
too rare in both groups to show much in a table.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Prevalence of measured hypertension by age band and WHO BMI category."}
dat_model |>
  mutate(age_group = fct_relevel(age_group, "18-25", "26-35", "36-45",
                                 "46-55", "56-65", ">65")) |>
  group_by(age_group, bmi_category) |>
  summarise(prop = mean(hypertension == "Yes"), n = n(), .groups = "drop") |>
  ggplot(aes(age_group, prop, fill = bmi_category)) +
  geom_col(position = position_dodge(preserve = "single")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 1)) +
  scale_fill_viridis_d(option = "D", end = 0.9) +
  labs(x = "Age band (years)", y = "Hypertensive on measurement (%)",
       fill = "BMI category",
       caption = "Bars are observed proportions; several age-by-BMI cells contain fewer than 20 people, especially in the underweight group")
```

Prevalence climbs steeply across age bands within every BMI category, and within
most age bands the obese sit clearly above the healthy-weight group, so the two
gradients are partly independent of one another; several underweight cells
contain only a handful of people and their bars should not be read as estimates.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = glm,
    method.args = list(family = binomial),
    exponentiate = TRUE
  )
uv

utid <- map_dfr(predictors, function(v) {
  tidy(glm(reformulate(v, outcome), data = dat_model, family = binomial),
       conf.int = TRUE, exponentiate = TRUE) |>
    filter(term != "(Intercept)")
})
ug <- function(term, d = 2) {
  r <- utid[utid$term == term, ]
  sprintf("OR %s (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d), fmt(r$conf.high, d))
}
up <- function(term) pv(utid$p.value[utid$term == term])
```

Age dominates the crude picture: relative to the 26-35 band, the odds of
measured hypertension are `r ug("age_group46-55")` at 46-55,
`r ug("age_group56-65")` at 56-65 and `r ug("age_group>65")` above 65 (all
p `r up("age_group>65")`), a clean monotonic gradient. Obesity carries
`r ug("bmi_categoryObese")` relative to healthy weight
(p `r up("bmi_categoryObese")`), and underweight `r ug("bmi_categoryUnderweight")`
(p `r up("bmi_categoryUnderweight")`). Men have `r ug("sexMale")`
(p `r up("sexMale")`). Self-reported diabetes carries `r ug("diabetesyes")`
(p `r up("diabetesyes")`) and tertiary education `r ug("educationTertiary")`
(p `r up("educationTertiary")`), i.e. crudely protective. Alcohol use is
borderline (`r ug("current_alcoholYes")`, p `r up("current_alcoholYes")`) and
smoking (`r ug("current_smokerYes")`, p `r up("current_smokerYes")`) and
physical activity (`r ug("adequate_activityYes")`, p `r up("adequate_activityYes")`)
are not significant.

## Why adjustment is needed

The crude odds ratios are not comparable with one another because the predictors
are heavily interlocked in this sample. **Age band** is the dominant confounder:
it is by far the strongest determinant of measured hypertension, and it is also
associated with almost every other predictor. Diabetes is reported by
`r fmt(100 * mean(dat_model$diabetes[dat_model$age_group == ">65"] == "yes"), 0)`%
of the over-65s against
`r fmt(100 * mean(dat_model$diabetes[dat_model$age_group == "26-35"] == "yes"), 0)`%
of the 26-35 band, so the crude diabetes OR of
`r fmt(utid$estimate[utid$term == "diabetesyes"], 2)` is in large part an age
contrast. Education is even more strongly age-patterned:
`r fmt(100 * mean(dat_model$education[dat_model$age_group == ">65"] == "Tertiary"), 0)`%
of the over-65s have tertiary education against
`r fmt(100 * mean(dat_model$education[dat_model$age_group == "26-35"] == "Tertiary"), 0)`%
of the 26-35 band, so the apparently protective crude education OR is mostly the
youth of the educated - a textbook confounded contrast. **Sex** confounds both
behavioural exposures: `r sum(dat_model$current_smoker == "Yes" & dat_model$sex == "Male")`
of the `r sum(dat_model$current_smoker == "Yes")` current smokers and
`r sum(dat_model$current_alcohol == "Yes" & dat_model$sex == "Male")` of the
`r sum(dat_model$current_alcohol == "Yes")` current drinkers in the analytic
sample are men, and men have higher pressure at a given age, so the crude
smoking and alcohol odds ratios are partly sex odds ratios. **BMI category**
confounds the diabetes and activity contrasts, since adiposity causes both
insulin resistance and raised pressure. Mutual adjustment lets each odds ratio
be read as a comparison between attenders alike on the other seven measured
characteristics.

## Multivariable model

```{r multivariable}
fit <- glm(reformulate(predictors, outcome), data = dat_model, family = binomial)

mv <- tbl_regression(fit, exponentiate = TRUE)
mv

tid <- tidy(fit, conf.int = TRUE, exponentiate = TRUE)
g <- function(term, d = 2) {
  r <- tid[tid$term == term, ]
  sprintf("OR %s (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d), fmt(r$conf.high, d))
}
gp <- function(term) pv(tid$p.value[tid$term == term])
```

```{r fit-stats}
n_par <- length(coef(fit))
epp   <- n_events / (n_par - 1)
tibble(
  Statistic = c("Complete-case N", "Rows dropped by listwise deletion",
                "Events (hypertension = Yes)", "AIC", "Null deviance",
                "Residual deviance", "Model parameters (incl. intercept)",
                "Events per parameter"),
  Value = c(as.character(nrow(dat_model)), as.character(n_dropped),
            as.character(n_events), fmt(AIC(fit), 1),
            fmt(fit$null.deviance, 1), fmt(fit$deviance, 1),
            as.character(n_par), fmt(epp, 1))
) |>
  gt() |>
  tab_header(title = md("**Model fit: multivariable logistic model for measured hypertension**"))
```

The model was fitted on `r nrow(dat_model)` complete cases with `r n_events`
events and `r n_par` parameters, giving `r fmt(epp, 1)` **events per
parameter** - comfortably above the conventional minimum of 10, so the model is
well identified and there is no need to trim the predictor set. AIC is
`r fmt(AIC(fit), 1)` against a null deviance of `r fmt(fit$null.deviance, 1)`
and a residual deviance of `r fmt(fit$deviance, 1)`.

Conditional on the other seven predictors, the age gradient survives almost
intact: `r g("age_group46-55")` at 46-55, `r g("age_group56-65")` at 56-65 and
`r g("age_group>65")` above 65, all relative to the 26-35 band. Obesity retains
`r g("bmi_categoryObese")` (p `r gp("bmi_categoryObese")`) and underweight
`r g("bmi_categoryUnderweight")` (p `r gp("bmi_categoryUnderweight")`). Men have
`r g("sexMale")` the adjusted odds of women of the same age and body size
(p `r gp("sexMale")`), and current drinkers `r g("current_alcoholYes")`
(p `r gp("current_alcoholYes")`). Diabetes is null after adjustment
(`r g("diabetesyes")`, p `r gp("diabetesyes")`), as is smoking
(`r g("current_smokerYes")`, p `r gp("current_smokerYes")`) and physical
activity (`r g("adequate_activityYes")`, p `r gp("adequate_activityYes")`).
These are associations among the attenders observed, not effects of intervening
on any of these variables.

## The cost of `monthly_income`

```{r income-cost}
fit_inc      <- glm(reformulate(c(predictors, "monthly_income"), outcome),
                    data = dat_inc, family = binomial)
fit_noinc_cc <- glm(reformulate(predictors, outcome), data = dat_inc, family = binomial)

inc_tid <- tidy(fit_inc, conf.int = TRUE, exponentiate = TRUE)
epp_inc <- n_events_inc / (length(coef(fit_inc)) - 1)

tibble(
  Model = c("Main model (no income)",
            "Same model, income complete-cases only",
            "Model with monthly_income added"),
  N = c(nrow(dat_model), nrow(dat_inc), nrow(dat_inc)),
  `% of file` = fmt(100 * c(nrow(dat_model), nrow(dat_inc), nrow(dat_inc)) / nrow(dat), 1),
  Events = c(n_events, n_events_inc, n_events_inc),
  Parameters = c(n_par - 1, length(coef(fit_noinc_cc)) - 1,
                 length(coef(fit_inc)) - 1),
  `Events/parameter` = fmt(c(epp, n_events_inc / (length(coef(fit_noinc_cc)) - 1),
                             epp_inc), 1),
  AIC = fmt(c(AIC(fit), AIC(fit_noinc_cc), AIC(fit_inc)), 1)
) |>
  gt() |>
  tab_header(title = md("**Table 2. What adding `monthly_income` costs.**"),
             subtitle = md("`monthly_income` is missing for 31% of participants; listwise deletion removes them from the whole model, not just from that one coefficient. AICs from different samples are not comparable."))
```

```{r income-numbers}
inc_lost <- nrow(dat_model) - nrow(dat_inc)
inc_p    <- anova(fit_noinc_cc, fit_inc, test = "LRT")$`Pr(>Chi)`[2]
age65_main <- tid$estimate[tid$term == "age_group>65"]
age65_inc  <- inc_tid$estimate[inc_tid$term == "age_group>65"]
ob_main    <- tid$estimate[tid$term == "bmi_categoryObese"]
ob_inc     <- inc_tid$estimate[inc_tid$term == "bmi_categoryObese"]
al_main    <- tid$estimate[tid$term == "current_alcoholYes"]
al_inc     <- inc_tid$estimate[inc_tid$term == "current_alcoholYes"]
miss_by_site <- dat |> group_by(study_site) |>
  summarise(pct = 100 * mean(is.na(monthly_income)))
```

This is the most important methodological point in this dataset.
`monthly_income` is missing for
`r fmt(100 * mean(is.na(dat$monthly_income)), 1)`% of participants, and because
`glm()` deletes rows listwise, adding it removes those participants from *every*
coefficient, not just from the income coefficient. The complete-case N falls
from **`r nrow(dat_model)`** to **`r nrow(dat_inc)`**
(`r fmt(100 * nrow(dat_inc) / nrow(dat), 0)`% of the file), and the event count
from **`r n_events`** to **`r n_events_inc`**: `r inc_lost` participants and
`r n_events - n_events_inc` events discarded to buy
`r length(coef(fit_inc)) - n_par` extra parameters, which drops events per
parameter from `r fmt(epp, 1)` to `r fmt(epp_inc, 1)`. The likelihood-ratio test
comparing the two models on the *same* `r nrow(dat_inc)` rows gives
p `r pv(inc_p)`: income contributes nothing.

The missingness is also clearly not random. Non-response to the income question
ranges from `r fmt(miss_by_site$pct[miss_by_site$study_site == "faith"], 0)`% at
the faith-based facility to
`r fmt(miss_by_site$pct[miss_by_site$study_site == "private"], 0)`% at the
private facility, and rises with education (from
`r fmt(100 * mean(is.na(dat$monthly_income[dat$education == "No School"])), 0)`%
among those with no schooling to
`r fmt(100 * mean(is.na(dat$monthly_income[dat$education == "Tertiary"])), 0)`%
among the tertiary educated), so the reduced sample is not a random subsample
and complete-case estimates from it are not guaranteed unbiased. The
consequences show in the coefficients: the alcohol odds ratio moves from
`r fmt(al_main, 2)` to `r fmt(al_inc, 2)` and the over-65 contrast from
`r fmt(age65_main, 2)` to `r fmt(age65_inc, 2)`, while obesity is more stable
(`r fmt(ob_main, 2)` to `r fmt(ob_inc, 2)`). **The main model therefore excludes
`monthly_income`**; a student who wants the income gradient must either accept a
one-third smaller, non-random sample or use multiple imputation.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 3. Unadjusted and adjusted odds ratios.**")
```

Two odds ratios move a long way and change their conclusion. **Self-reported
diabetes** collapses from `r ug("diabetesyes")` crude (p `r up("diabetesyes")`)
to `r g("diabetesyes")` adjusted (p `r gp("diabetesyes")`) - essentially the
entire crude diabetes-hypertension association in these data is the age and
adiposity difference between people with and without diabetes. **Tertiary
education** moves toward the null from `r ug("educationTertiary")` crude
(p `r up("educationTertiary")`, apparently strongly protective) to
`r g("educationTertiary")` adjusted (p `r gp("educationTertiary")`, no longer
significant): the crude "protection" of education was the youth of the educated,
and once age is held constant it largely disappears. **Current smoking** also
attenuates to exactly the null, from `r ug("current_smokerYes")` to
`r g("current_smokerYes")`, because smokers here are almost all men. The **age
gradient** moves only modestly (over-65s `r ug("age_group>65")` crude to
`r g("age_group>65")` adjusted; 56-65 `r ug("age_group56-65")` to
`r g("age_group56-65")`), which is what one expects of the variable that
confounds the others while being itself confounded by little that is measured
here. **Obesity** barely moves at all (`r ug("bmi_categoryObese")` to
`r g("bmi_categoryObese")`), so the adiposity association is not an age artefact.
Two estimates cross into significance on adjustment: **current alcohol use**
(`r ug("current_alcoholYes")`, p `r up("current_alcoholYes")` crude, to
`r g("current_alcoholYes")`, p `r gp("current_alcoholYes")` adjusted) and, in
the opposite direction, **underweight** strengthens from
`r ug("bmi_categoryUnderweight")` to `r g("bmi_categoryUnderweight")`; both were
partly masked by the sex and age composition of the exposed groups, though the
underweight estimate rests on only
`r sum(dat_model$bmi_category == "Underweight" & dat_model$hypertension == "Yes")`
events and is best treated as suggestive.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r diag-numbers}
vifs <- check_collinearity(fit)
max_vif <- max(vifs$VIF)
br <- binned_residuals(fit)
br_pct <- 100 * attributes(br)$resid_ok
```

For a logistic model only some `check_model()` panels apply. **Binned
residuals**: `r fmt(br_pct, 1)`% of bins fall inside the error bounds against a
rule-of-thumb target of about 95%, so there is real lack of fit. The cause is
structural rather than a mis-specified curve: every predictor is a coarse
category, so the model can only produce a small number of distinct fitted
probabilities, and within a band (particularly the wide age bands and the broad
WHO BMI classes) risk varies a great deal that the model cannot follow. With
continuous age and continuous BMI this panel would almost certainly improve;
with this file it cannot be fixed. **Collinearity** is low throughout (maximum
VIF `r fmt(max_vif, 2)`, for `r vifs$Term[which.max(vifs$VIF)]`), so despite the
strong pairwise associations between age, education and diabetes the adjusted
odds ratios are estimated stably - the education attenuation above is genuine
confounding, not variance inflation. **Influential observations**: no point
exceeds the Cook's distance contours, as expected with `r nobs(fit)` rows.
**Events per parameter** is `r fmt(epp, 1)`, so overfitting and sparse-data bias
are not concerns, and there is **no separation**: every predictor level contains
both events and non-events (the thinnest cell is underweight-and-hypertensive,
with `r sum(dat_model$bmi_category == "Underweight" & dat_model$hypertension == "Yes")`
events). Homogeneity of variance and normality of residuals are printed for
completeness but do not apply to a binomial GLM.

## Sensitivity analysis: prior diagnosis, treatment and the awareness gap

```{r prior-sens}
dat_prior <- dat |>
  select(all_of(c(outcome, predictors, "prior_htn_diagnosis"))) |> drop_na()
fit_prior <- glm(reformulate(c(predictors, "prior_htn_diagnosis"), outcome),
                 data = dat_prior, family = binomial)
pt <- tidy(fit_prior, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term == "prior_htn_diagnosisYes")
tab <- table(measured = dat$hypertension, prior = dat$prior_htn_diagnosis)
aware <- 100 * tab["Yes", "Yes"] / sum(tab["Yes", ])
controlled <- 100 * tab["No", "Yes"] / sum(tab[, "Yes"])
```

```{r prior-tab}
as.data.frame.matrix(tab) |>
  rownames_to_column("Measured hypertension") |>
  rename(`Prior diagnosis: No` = No, `Prior diagnosis: Yes` = Yes) |>
  gt() |>
  tab_header(title = md("**Table 4. Measured hypertension against self-reported prior diagnosis.**"))
```

Only `r fmt(aware, 0)`% of participants who are hypertensive on measurement
report a previous diagnosis, and `r fmt(controlled, 0)`% of those who report a
previous diagnosis are *not* hypertensive today - the two variables agree on
only `r fmt(100 * (tab["No", "No"] + tab["Yes", "Yes"]) / sum(tab), 0)`% of
participants. Adding `prior_htn_diagnosis` to the model gives an adjusted
`r sprintf("OR %s (95%% CI %s to %s)", fmt(pt$estimate, 2), fmt(pt$conf.low, 2), fmt(pt$conf.high, 2))`,
p `r pv(pt$p.value)`. **This is not an effect of diagnosis on blood pressure and
should not be reported as one.** Two opposite mechanisms are tangled in it: a
diagnosis was given *because* pressure was high before (raising the OR), and a
diagnosis leads to antihypertensive treatment that lowers today's measured
pressure (lowering it). The file records **no medication at all**, so the two
cannot be separated. The observed OR above 1 mainly indicates that treatment in
this population is not returning pressure to the level of the never-diagnosed;
it is reported here as a sensitivity analysis and kept out of the main model.

# Discussion

Among these Kenyan outpatient attenders, age and body size are the
characteristics independently associated with hypertension on measured blood
pressure: adjusted for the other seven predictors, the over-65s have
`r g("age_group>65")` the odds of the 26-35 band and obese attenders
`r g("bmi_categoryObese")` the odds of healthy-weight attenders. Male sex and
current alcohol use add smaller independent associations. The crude association
with self-reported diabetes, which looked substantial, is fully explained by age
and adiposity, and the apparently protective effect of tertiary education is
almost entirely the youth of the educated. Practically, if a clinic can only
screen some attenders, age and visible body size are the two things worth acting
on, and no reassurance should be taken from a patient's education or from the
absence of a diabetes diagnosis. The awareness gap is stark:
`r fmt(100 - aware, 0)`% of those found hypertensive at this visit had never
been told they had hypertension.

**Limitations.** This is a **facility-based convenience sample** of outpatient
attenders and not a community sample - attenders are sicker and, here, far
better educated and more female than the Kenyan adult population - so the
`r fmt(prev, 0)`% prevalence must not be read as a population prevalence. The
design is **cross-sectional**, so nothing here establishes temporality; obesity,
alcohol use and blood pressure were all recorded at the same visit. The outcome
is a **dichotomised version of a continuous measurement** taken at a single
visit, which discards information, makes the results sensitive to the threshold
used, and misclassifies anyone whose pressure was atypical that day (white-coat
and masked hypertension both exist and neither can be detected here); the
continuous `systolic_bp` outcome is statistically the better analysis of the
same data. **Antihypertensive medication is not recorded**, which is a serious
unmeasured confounder in the wrong direction: treated participants appear
non-hypertensive despite higher underlying risk, which biases every coefficient
toward the null and makes the observed prevalence an underestimate of underlying
disease. **No predictor is continuous**: age is banded and BMI is a WHO
category, so residual confounding within band is certain, the binned-residual
check fails, and no per-unit statement is possible. Participants come from
**three facilities with no facility identifier**, so within-facility correlation
cannot be modelled and the standard errors are probably too small.
Complete-case analysis dropped `r n_dropped` rows; `monthly_income` was
excluded, so household economic position is unadjusted for. Smoking is rare
(`r sum(dat_model$current_smoker == "Yes")` smokers) and underweight is thin
(`r sum(dat_model$bmi_category == "Underweight")` participants), so those two
estimates are imprecise. Salt intake, family history and kidney function are
absent entirely.

**Future work.** Repeat with continuous age and measured BMI, repeat blood
pressure measurement on a second day, and record antihypertensive medication -
none of which this extract carries. Modelling `systolic_bp` continuously is a
better use of the same file.

# Feasibility notes

*NOT part of a student report.*

- **Works as a student logistic-regression project.** `r n_events` events in
  `r nrow(dat_model)` complete cases, `r fmt(epp, 1)` events per parameter, no
  separation, maximum VIF `r fmt(max_vif, 2)`, and two predictors (diabetes,
  tertiary education) that collapse convincingly on adjustment. That is exactly
  the Purpose 3 teaching shape.
- **FATAL trap: `systolic_bp` and `diastolic_bp` must never be predictors.**
  `hypertension` *is* the measured average pressure crossed against a threshold.
  A student who puts either pressure in the model gets separation or an
  effectively deterministic model. This is the single most likely way to destroy
  this analysis and must be stated in the brief.
- **Trap: `prior_htn_diagnosis` is not a clean predictor either.** The
  relationship is bidirectional - prior diagnosis follows past high pressure but
  brings treatment that lowers current pressure - and there is **no medication
  variable in the file** to break the loop. Adjusted
  `r sprintf("OR %s", fmt(pt$estimate, 2))` here; a student will interpret it as
  a risk factor unless warned.
- **THE missing-data teaching case.** `monthly_income` is
  `r fmt(100 * mean(is.na(dat$monthly_income)), 1)`% missing; including it drops
  the sample from `r nrow(dat_model)` to `r nrow(dat_inc)`
  (`r fmt(100 * nrow(dat_inc) / nrow(dat), 0)`% of the file) and events from
  `r n_events` to `r n_events_inc`, cutting events per parameter from
  `r fmt(epp, 1)` to `r fmt(epp_inc, 1)`, and it contributes nothing
  (LRT p `r pv(inc_p)`). Missingness is patterned by facility
  (`r fmt(miss_by_site$pct[miss_by_site$study_site == "private"], 0)`% private vs
  `r fmt(miss_by_site$pct[miss_by_site$study_site == "faith"], 0)`% faith) and
  education, so it is not MCAR. Make this a required part of the brief.
- **Binned residuals fail (`r fmt(br_pct, 1)`% of bins inside bounds).** This is
  structural - all predictors are coarse categories, so only a handful of
  distinct fitted probabilities exist - and cannot be fixed with this file.
  Students will otherwise write "diagnostics fine"; they need to be told what
  the panel means and that this one is not repairable here.
- **`current_smoker` is sparse**: `r sum(dat$current_smoker == "Yes")` smokers in
  the file (`r sum(dat_model$current_smoker == "Yes" & dat_model$hypertension == "Yes")`
  of them hypertensive in the analytic sample), and almost all male. Keep it to
  one degree of freedom; no interactions.
- **`bmi_category` "Underweight" is thin**
  (`r sum(dat_model$bmi_category == "Underweight")` people,
  `r sum(dat_model$bmi_category == "Underweight" & dat_model$hypertension == "Yes")`
  events) and the strongly protective coefficient is probably reverse causation
  (illness-related weight loss), not protection.
- **No continuous predictors exist in this file.** Students will try to report
  odds per year of age or per kg/m2 and cannot.
- **Redundant with `systolic_bp`.** Same question on a thrown-away version of the
  same measurement. Students should do one or the other; the continuous version
  is the better analysis, and this one is only preferable if the point of the
  project is logistic regression.
- **Clustering cannot be modelled** (three facilities, no facility identifier),
  so standard errors are optimistic.
