---
title: "Adequate antibiotic knowledge in Serbian primary-care attenders"
subtitle: "Dataset 1: Antibiotic knowledge and self-medication in Serbia"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()

# Helpers so every number in the prose is pulled from the fitted model rather
# than typed by hand.
fmt_p <- function(p) if (p < 0.001) "p < 0.001" else sprintf("p = %.3f", p)
row_of <- function(tab, trm) tab[tab$term == trm, ]
say_or <- function(tab, trm) {
  r <- row_of(tab, trm)
  sprintf("OR %.2f (95%% CI %.2f to %.2f, %s)",
          r$estimate, r$conf.low, r$conf.high, fmt_p(r$p.value))
}
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "serbia_antibiotics.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
#
# `knowledge_score` is deliberately NOT in this vector. `adequate_knowledge` is
# the indicator `knowledge_score >= 9` and nothing else, so putting the score in
# this model is complete separation, not a covariate.
outcome    <- "adequate_knowledge"
predictors <- c("age_group", "sex", "education_level", "employment_status",
                "gp_visits_past_year", "health_professional_in_family")

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    # Age: ">65" is both the largest band (n = 123) and the lowest-scoring group,
    # so every contrast reads as "younger adults versus the over-65s".
    age_group = factor(age_group,
                       levels = c(">65", "18-24", "25-34", "35-44", "45-54", "55-64"),
                       labels = c("Over 65", "18-24", "25-34", "35-44", "45-54", "55-64")),
    sex = factor(sex, levels = c("female", "male"), labels = c("Female", "Male")),
    # Education: "secondary" is the largest group (n = 259) and sits between the
    # other two, so both a lower and a higher contrast are available.
    education_level = factor(education_level,
                             levels = c("secondary", "primary", "tertiary"),
                             labels = c("Secondary", "Primary", "Tertiary")),
    employment_status = factor(employment_status,
                               levels = c("employed", "unemployed", "pensioner"),
                               labels = c("Employed", "Unemployed", "Pensioner")),
    # GP visits: "0" is the clinically unexposed group (no primary-care contact).
    gp_visits_past_year = factor(gp_visits_past_year,
                                 levels = c("0", "1-4", "5-10", "more than 10"),
                                 labels = c("None", "1-4", "5-10", "More than 10")),
    health_professional_in_family = factor(health_professional_in_family,
                                           levels = c(0, 1), labels = c("No", "Yes")),
    marital_status = factor(marital_status,
                            levels = c("married", "single", "divorced_widowed"),
                            labels = c("Married", "Single", "Divorced or widowed")),
    household_size = factor(household_size,
                            levels = c("1-3", "3-5", "more than 5"),
                            labels = c("1-3", "3-5", "More than 5"))
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

n_event    <- sum(dat_model[[outcome]] == 1)
n_nonevent <- sum(dat_model[[outcome]] == 0)
n_rarer    <- min(n_event, n_nonevent)

# Verification that the outcome is a deterministic recode of `knowledge_score`.
adeq_identical <- identical(dat_raw$adequate_knowledge,
                            as.numeric(dat_raw$knowledge_score >= 9))
rng_lo <- range(dat_raw$knowledge_score[dat_raw$adequate_knowledge == 0])
rng_hi <- range(dat_raw$knowledge_score[dat_raw$adequate_knowledge == 1])
xtab_check <- table(dat_raw$knowledge_score >= 9, dat_raw$adequate_knowledge)
n_offdiag  <- sum(xtab_check) - sum(diag(xtab_check))

# Confounding structure quoted in the text.
n_primary       <- sum(dat$education_level == "Primary")
n_primary_old   <- sum(dat$education_level == "Primary" & dat$age_group == "Over 65")
n_pensioner     <- sum(dat$employment_status == "Pensioner")
n_pensioner_old <- sum(dat$employment_status == "Pensioner" & dat$age_group == "Over 65")
cramers_v <- function(x, y) {
  tt <- suppressWarnings(table(x, y))
  sqrt(as.numeric(suppressWarnings(chisq.test(tt))$statistic) /
         (sum(tt) * (min(dim(tt)) - 1)))
}
v_age_emp <- cramers_v(dat$age_group, dat$employment_status)
```

# Introduction

Public campaigns against antibiotic misuse are usually evaluated against a pass
mark: what share of the population knows enough. These data come from a
cross-sectional questionnaire survey of `r nrow(dat_raw)` adults who consulted
general practitioners at four health centres in Serbia, reported by Horvat and
colleagues (*PLOS ONE* 2017;12(7):e0180799). Each respondent answered a 12-item
antibiotic quiz, and the source paper classified a score of 9 or more as
adequate knowledge; `r n_event` of `r nrow(dat_model)` respondents
(`r sprintf("%.0f", 100 * n_event / nrow(dat_model))`%) clear that bar. Knowing
which groups fall below it is what tells a campaign where to aim.

**Aim.** Among Serbian adults attending primary care, we examined which of age
group, sex, education, employment status, frequency of GP contact and having a
health professional in the family are independently associated with achieving
adequate antibiotic knowledge, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** Cross-sectional, interviewer-administered questionnaire survey
of adults consulting general practitioners at four Serbian health centres;
`r nrow(dat_raw)` respondents completed the entire questionnaire and all
`r nrow(dat_raw)` are analysed here. All measurements were taken on a single
occasion.

**Variables.** The outcome is `adequate_knowledge`, a binary indicator; the
modelled event is **adequate knowledge = yes**, defined by the source paper as a
score of 9 or more out of 12 (`r n_event` events, `r n_nonevent` non-events, so
the rarer class has `r n_rarer`). Predictors are `age_group` (six bands;
reference: over 65, the largest band at n = `r sum(dat$age_group == "Over 65")`),
`sex` (reference: female, n = `r sum(dat$sex == "Female")`), `education_level`
(primary / secondary / tertiary; reference: secondary, the largest group at n =
`r sum(dat$education_level == "Secondary")`), `employment_status` (reference:
employed, n = `r sum(dat$employment_status == "Employed")`),
`gp_visits_past_year` (four bands of visits in the previous 12 months;
reference: none, the group with no primary-care contact, n =
`r sum(dat$gp_visits_past_year == "None")`) and `health_professional_in_family`
(reference: no, n = `r sum(dat$health_professional_in_family == "No")`). All six
are categorical, so every estimate is an odds ratio against the stated reference
level; the file contains no continuous predictors.

**`knowledge_score` is not a predictor here, and cannot be.** The outcome is
exactly the indicator `knowledge_score >= 9` and nothing else. In this file the
respondents flagged 0 score `r rng_lo[1]` to `r rng_lo[2]` and those flagged 1
score `r rng_hi[1]` to `r rng_hi[2]`, with no overlap; re-deriving the flag from
the score reproduces the deposited column exactly (`identical()` returns
`r adeq_identical`, with `r n_offdiag` disagreements out of `r nrow(dat_raw)`).
Entering the score in this model is textbook complete separation: the fitted
coefficient diverges, the standard error explodes, and `glm()` returns a
meaningless answer with a convergence warning. The two columns are one outcome
measured two ways — continuous and dichotomised — so exactly one of them can be
the response in any given analysis and the other must be left out entirely. The
same file's continuous version of this analysis is in `knowledge_score.Rmd`; the
two documents answer the same question and should be read as alternatives, not
as complements.

Two of the eight background variables were also left out deliberately.
`marital_status` is largely a restatement of age here (Cramér's V =
`r sprintf("%.2f", cramers_v(dat$age_group, dat$marital_status))` with
`age_group`), and `household_size` has no plausible link to what a person knows
about antibiotics and none in the data
(univariable `r fmt_p(anova(glm(adequate_knowledge ~ household_size, dat, family = binomial), test = "LRT")[["Pr(>Chi)"]][2])`).
`self_medication` is a behaviour and a plausible *consequence* of knowledge
rather than a cause of it, so it is not used as a predictor. `id` is a row
number and is never modelled.

**Statistical analysis.** Logistic regression (`glm(family = binomial)`),
justified because the outcome is binary. We fitted a univariable model for each
predictor first, then a single multivariable model containing all of them.
Predictors were chosen in advance on subject-matter grounds, not by
significance. Estimates are odds ratios with 95% confidence intervals;
significance threshold 0.05. Assumptions were checked with
`performance::check_model()`, with attention to binned residuals, collinearity
and influential observations rather than to normality or constant variance,
which do not apply to a logistic model. Analysis used `r R.version.string` with
tidyverse, gtsummary and performance.

**Missing data.** There is none. Every one of the `r ncol(dat_raw)` columns is
`r sprintf("%.0f", 100 * mean(complete.cases(dat_raw)))`% complete, so the model
uses all `r nrow(dat_model)` of `r nrow(dat)` rows and `r n_dropped` were
dropped. There is no missing-data story to tell for this file, and it would be
wrong to manufacture one — though the survey retained only respondents who
completed the whole questionnaire, so partial responders were excluded upstream
and nothing is known about them.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  mutate(adequate_knowledge = factor(adequate_knowledge, levels = c(0, 1),
                                     labels = c("Score 2-8", "Score 9-12 (adequate)"))) |>
  tbl_summary(
    by = adequate_knowledge,
    missing = "no"
  ) |>
  add_n() |>
  add_overall() |>
  modify_caption("**Table 1. Sample characteristics, by adequate-knowledge status.**")
```

Education separates the two groups most sharply:
`r sprintf("%.0f", 100 * mean(dat_model$education_level[dat_model$adequate_knowledge == 1] == "Tertiary"))`%
of respondents with adequate knowledge hold a tertiary degree, against
`r sprintf("%.0f", 100 * mean(dat_model$education_level[dat_model$adequate_knowledge == 0] == "Tertiary"))`%
of those below the threshold. The inadequate group is also older
(`r sprintf("%.0f", 100 * mean(dat_model$age_group[dat_model$adequate_knowledge == 0] == "Over 65"))`%
over 65 versus
`r sprintf("%.0f", 100 * mean(dat_model$age_group[dat_model$adequate_knowledge == 1] == "Over 65"))`%)
and contains more pensioners. Sex and frequency of GP visits look similar in the
two groups.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Distribution of the underlying 12-item knowledge score within each education level, shaded by whether the respondent clears the adequate-knowledge threshold. The dashed line is the cut at 9; the outcome modelled here is nothing more than which side of that line a respondent falls."}
dat_fig <- dat |>
  mutate(adequate = factor(adequate_knowledge, levels = c(0, 1),
                           labels = c("Below threshold (2-8)", "Adequate (9-12)")))

ggplot(dat_fig, aes(x = knowledge_score, fill = adequate)) +
  geom_vline(xintercept = 8.5, linetype = "dashed", colour = "grey30") +
  geom_bar(width = 0.85) +
  facet_wrap(~ education_level, nrow = 1) +
  scale_fill_manual(values = c("Below threshold (2-8)" = "#c4694a",
                               "Adequate (9-12)" = "#2c7fb8")) +
  scale_x_continuous(breaks = seq(2, 12, 2)) +
  labs(x = "Antibiotic knowledge score (0-12)", y = "Respondents",
       fill = "Outcome")
```

The whole score distribution shifts right as education rises — the tertiary
panel is dominated by scores of 10 and above while the primary panel sits mostly
below the line — and the figure makes plain what the outcome is: a single cut
through a continuous score, which throws away the difference between a score of
2 and a score of 8, and between 9 and 12.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = glm,
    method.args = list(family = binomial),
    exponentiate = TRUE
  )
uv
```

```{r univariable-global}
# Global likelihood-ratio tests, so multi-level factors are not judged on a
# single dummy contrast.
uni_global <- map_dfr(predictors, function(p) {
  m <- glm(reformulate(p, outcome), data = dat_model, family = binomial)
  tibble(term = p, p.value = anova(m, test = "LRT")[["Pr(>Chi)"]][2])
})
gp <- function(v) fmt_p(uni_global$p.value[uni_global$term == v])
uv_tab <- map_dfr(predictors, ~ tidy(glm(reformulate(.x, outcome), data = dat_model,
                                         family = binomial),
                                     conf.int = TRUE, exponentiate = TRUE))
```

Education dominates the crude picture (`r gp("education_level")`): tertiary
education multiplies the odds of adequate knowledge nearly fourfold relative to
secondary (`r say_or(uv_tab, "education_levelTertiary")`), while primary
education cuts them by two-thirds (`r say_or(uv_tab, "education_levelPrimary")`)
— a spread of roughly
`r sprintf("%.0f", row_of(uv_tab, "education_levelTertiary")$estimate / row_of(uv_tab, "education_levelPrimary")$estimate)`-fold
in the odds from the bottom education band to the top. Age is next
(`r gp("age_group")`): every younger band clears the threshold more often than
the over-65s, most strongly the 25-34 band
(`r say_or(uv_tab, "age_group25-34")`). Employment status is also strongly
associated crudely (`r gp("employment_status")`, with pensioners at
`r say_or(uv_tab, "employment_statusPensioner")` against the employed), and
having a health professional in the family roughly doubles the odds
(`r say_or(uv_tab, "health_professional_in_familyYes")`). GP visits are weakly
associated
overall (`r gp("gp_visits_past_year")`) with no clear gradient, and sex is not
associated at all (`r gp("sex")`).

## Why adjustment is needed

The crude odds ratios above cannot be read individually because the background
variables in this survey are heavily confounded with one another.
`age_group` is the clearest confounder of the education association:
`r n_primary_old` of the `r n_primary` primary-educated respondents
(`r sprintf("%.0f", 100 * n_primary_old / n_primary)`%) are over 65, because
compulsory schooling expanded in the region within these respondents'
lifetimes, and age is itself associated with failing the knowledge threshold.
Unadjusted, part of what looks like an education effect is simply the age
composition of the education groups. `employment_status` is the second, and is
close to a relabelling of age at the top end: `r n_pensioner_old` of the
`r n_pensioner` pensioners (`r sprintf("%.0f", 100 * n_pensioner_old / n_pensioner)`%)
are over 65, Cramér's V between age and employment is
`r sprintf("%.2f", v_age_emp)`, and pensioners are also the least educated group
in the sample
(`r sprintf("%.0f", 100 * mean(dat$education_level[dat$employment_status == "Pensioner"] == "Tertiary"))`%
hold a tertiary degree, against
`r sprintf("%.0f", 100 * mean(dat$education_level[dat$employment_status == "Employed"] == "Tertiary"))`%
of the employed). The crude pensioner deficit therefore has at least two
alternative explanations, age and education, before pensioner status itself is
considered. A third route runs through `gp_visits_past_year`, which is strongly
age-patterned
(`r sprintf("%.0f", 100 * mean(dat$gp_visits_past_year[dat$age_group == "Over 65"] %in% c("5-10", "More than 10")))`%
of the over-65s report five or more visits against
`r sprintf("%.0f", 100 * mean(dat$gp_visits_past_year[dat$age_group != "Over 65"] %in% c("5-10", "More than 10")))`%
of everyone else), so any crude attendance association is partly an age
association. Only a model containing all six turns these into within-stratum
comparisons.

## Multivariable model

```{r multivariable}
fit <- glm(reformulate(predictors, outcome), data = dat_model, family = binomial)

mv <- tbl_regression(fit, exponentiate = TRUE)
mv
```

```{r fit-stats}
mv_tab  <- tidy(fit, conf.int = TRUE, exponentiate = TRUE)
n_param <- length(coef(fit)) - 1          # non-intercept parameters
epp     <- n_rarer / n_param
fit_null <- glm(reformulate("1", outcome), data = dat_model, family = binomial)
tjur     <- as.numeric(r2_tjur(fit))
lrt      <- anova(fit_null, fit, test = "LRT")

tibble(
  Statistic = c("AIC (fitted model)", "AIC (intercept-only model)",
                "Tjur's R-squared", "Likelihood-ratio test vs null (p)",
                "Events (adequate knowledge)", "Non-events",
                "Rarer class", "Non-intercept parameters",
                "Events per parameter"),
  Value = c(sprintf("%.1f", AIC(fit)), sprintf("%.1f", AIC(fit_null)),
            sprintf("%.3f", tjur),
            format.pval(lrt[["Pr(>Chi)"]][2], digits = 3, eps = 0.001),
            as.character(c(n_event, n_nonevent, n_rarer, n_param)),
            sprintf("%.1f", epp))
) |>
  gt() |>
  tab_header(title = "Model fit and events per parameter")
```

With `r n_rarer` in the rarer outcome class and `r n_param` non-intercept
parameters, there are `r sprintf("%.1f", epp)` events per parameter — above the
conventional minimum of ten, but not by a wide margin, so the six-level age
factor is close to the limit of what this sample supports. Adjusting for all six
predictors, only two associations survive. Education remains dominant:
conditional on age, sex, employment, attendance and family health-care exposure,
tertiary education carries `r say_or(mv_tab, "education_levelTertiary")` and
primary education `r say_or(mv_tab, "education_levelPrimary")` relative to
secondary. Having a health professional in the family retains
`r say_or(mv_tab, "health_professional_in_familyYes")`. Everything else falls
away: no age contrast is significant (the largest, 18-24, is
`r say_or(mv_tab, "age_group18-24")`), employment status is not
(pensioner `r say_or(mv_tab, "employment_statusPensioner")`), nor is sex
(`r say_or(mv_tab, "sexMale")`) or any attendance band (more than ten visits:
`r say_or(mv_tab, "gp_visits_past_yearMore than 10")`). The model is clearly
better than an intercept alone (AIC `r sprintf("%.1f", AIC(fit))` against
`r sprintf("%.1f", AIC(fit_null))`; likelihood-ratio
`r fmt_p(lrt[["Pr(>Chi)"]][2])`), and Tjur's R² is
`r sprintf("%.3f", tjur)` — modest in absolute terms but more than twice the
discrimination achieved for self-medication in the same file.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

The employment-status estimates move most and are the clearest confounding
demonstration in the table. Crudely, pensioners have
`r say_or(uv_tab, "employment_statusPensioner")` against the employed; adjusted,
that rises to `r say_or(mv_tab, "employment_statusPensioner")` and is no longer
significant. Pensioners fail the knowledge threshold because they are old and
less educated, not because they are pensioners. The age contrasts move the same
way and lose significance too: the 25-34 band falls from
`r say_or(uv_tab, "age_group25-34")` crude to
`r say_or(mv_tab, "age_group25-34")` adjusted, essentially all the way to the
null, because younger respondents in this sample are far better educated than
the over-65s and it is the education, not the birth year, that carries the
information. The 18-24 contrast is the exception in direction, moving from
`r say_or(uv_tab, "age_group18-24")` to `r say_or(mv_tab, "age_group18-24")`,
but on `r sum(dat_model$age_group == "18-24")` respondents its interval is far
too wide to interpret. In contrast, the two surviving predictors barely move at
all: tertiary education goes from `r say_or(uv_tab, "education_levelTertiary")`
to `r say_or(mv_tab, "education_levelTertiary")`, primary education from
`r say_or(uv_tab, "education_levelPrimary")` to
`r say_or(mv_tab, "education_levelPrimary")`, and having a health professional in
the family from `r say_or(uv_tab, "health_professional_in_familyYes")` to
`r say_or(mv_tab, "health_professional_in_familyYes")`. Their stability under
adjustment is the strongest evidence in the analysis that these two are not
proxies for age or employment. Sex is null and static in both columns
(`r say_or(uv_tab, "sexMale")` to `r say_or(mv_tab, "sexMale")`). Nothing gains
significance on adjustment; age and employment status both lose it, which is
what confounding removal looks like.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r diag-tests}
binned_residuals(fit)
check_collinearity(fit)
```

```{r diag-numbers}
br      <- binned_residuals(fit)
br_pct  <- 100 * attributes(br)$resid_ok
vif_tab <- check_collinearity(fit)
max_vif <- max(vif_tab$VIF)
max_vif_term <- vif_tab$Term[which.max(vif_tab$VIF)]
max_cook <- max(cooks.distance(fit))
pr <- range(fitted(fit))
min_cell <- min(table(dat_model$age_group, dat_model[[outcome]]))
```

The diagnostics are satisfactory. The binned-residuals check passes:
`r sprintf("%.0f", br_pct)`% of binned residuals fall inside their error bounds,
against the 95% wanted, so the fitted probabilities track the observed
proportions across the range — and that range is reasonably wide, from
`r sprintf("%.2f", pr[1])` to `r sprintf("%.2f", pr[2])`, which means the model
actually discriminates rather than predicting the base rate for everyone.
Collinearity is acceptable: the largest VIF is `r sprintf("%.2f", max_vif)`
(`r max_vif_term`), elevated by the age-employment overlap described above but
below the conventional threshold of 5 — enough to explain why the age and
employment coefficients are individually imprecise, not enough to invalidate
them. No observation is influential (largest Cook's distance
`r sprintf("%.4f", max_cook)`, far below 0.5). There is no separation: every
coefficient and standard error is finite, and the sparsest predictor-by-outcome
cell still contains `r min_cell` respondents (the 18-24 band below the
threshold). Homogeneity of variance and normality of residuals are not
assumptions of a logistic model, so those panels are not interpreted; the
linearity panel is also uninformative because every predictor is categorical.

# Discussion

Among Serbian adults attending primary care, formal education is overwhelmingly
the characteristic associated with reaching the adequate-knowledge threshold:
tertiary-educated respondents have roughly
`r sprintf("%.1f", row_of(mv_tab, "education_levelTertiary")$estimate)` times the
odds of secondary-educated ones, and primary-educated respondents about
`r sprintf("%.0f", 100 * (1 - row_of(mv_tab, "education_levelPrimary")$estimate))`%
lower odds, holding age, sex, employment and health-service contact constant.
Having a family member in health care nearly doubles the odds independently — a
plausible informal-information effect. Everything else, including age itself, is
explained away by education. Practically, that points campaigns towards
low-education audiences specifically, rather than towards the elderly as a group,
since the apparent age deficit is really an education deficit.

**Limitations.** The most important limitation is the outcome itself. Adequate
knowledge is not a naturally occurring category: it is the 12-item score cut at
9, an author-chosen threshold with no external validation, and dichotomising a
continuous score discards information and costs statistical power. The
consequence is visible in this file — in the linear analysis of the same
underlying score (`knowledge_score.Rmd`) several age contrasts remain
significant after adjustment, whereas here every one of them is lost, which is
the expected result of throwing away the distinction between a score of 2 and a
score of 8. Any conclusion about "adequate knowledge" is also hostage to the cut
point: moving it from 9 to 8 or 10 would change both the prevalence and the
coefficients. The design is cross-sectional and every predictor is a fixed
background characteristic, so nothing here is a causal effect. All predictors
are coarse categories, so no per-year or dose-response trend can be tested. The
primary-education group (n = `r n_primary`, of whom
`r sum(dat_model$education_level == "Primary" & dat_model$adequate_knowledge == 1)`
reach the threshold) and the 18-24 band (n =
`r sum(dat_model$age_group == "18-24")`, of whom
`r sum(dat_model$age_group == "18-24" & dat_model$adequate_knowledge == 0)` fall
below it) are both small, so those contrasts are much less precise than they
appear. Events per parameter is `r sprintf("%.1f", epp)`, adequate but not
generous. Health literacy, income and urban or rural residence are unmeasured
and are plausible common causes of both education and knowledge, so residual
confounding is likely. The quiz was administered in a clinic to people already
consulting a GP, so the sample is not representative of Serbian adults who avoid
primary care, and social desirability may inflate scores.

**Future work.** Reporting the continuous score alongside any threshold-based
result, and showing how the conclusions change as the threshold moves, would
make the dependence on an arbitrary cut point visible rather than hidden.

# Feasibility notes

*Not part of a student report.*

- **Workable, and statistically the stronger of the two logistic outcomes**
  (Tjur's R² `r sprintf("%.3f", tjur)` against 0.071 for `self_medication`; AIC
  `r sprintf("%.1f", AIC(fit))` against `r sprintf("%.1f", AIC(fit_null))` null;
  binned residuals pass at `r sprintf("%.0f", br_pct)`%). Education gives a
  large, unambiguous effect that students can interpret, and there are three
  clearly null predictors for contrast.
- **MAJOR conceptual problem: this is not really a separate outcome.**
  `adequate_knowledge` is exactly `knowledge_score >= 9` (verified:
  `identical()` = `r adeq_identical`, `r n_offdiag` disagreements in
  `r nrow(dat_raw)` rows; the 0-group spans `r rng_lo[1]`-`r rng_lo[2]`, the
  1-group `r rng_hi[1]`-`r rng_hi[2]`). It duplicates `knowledge_score.Rmd`
  with less information and less power. Offering both as final-project options
  invites two students to hand in the same analysis, and the linear version is
  strictly better — it keeps the full range and, as shown above, retains age
  associations that the dichotomised version loses. Worth listing them as
  alternatives rather than as two independent choices.
- **FATAL if the two are combined.** Using `knowledge_score` as a predictor of
  `adequate_knowledge` (or the reverse) is complete separation. `glm()` will
  return an astronomically large coefficient with an astronomically large
  standard error, plus a convergence warning that weaker students will ignore.
  This is the single most likely way for a student to wreck this dataset and
  needs an explicit warning in the catalogue.
- **Complete-case N = `r nrow(dat_model)`** of `r nrow(dat)`; `r n_dropped`
  dropped. The file is 100% complete, so there is no missing-data narrative and
  students should say so plainly rather than invent one.
- **Events per parameter `r sprintf("%.1f", epp)`** (`r n_rarer` in the rarer
  class, `r n_param` parameters). Fine for six predictors as specified, but
  there is no headroom: adding `marital_status` and `household_size` as well
  takes it to
  `r sprintf("%.1f", n_rarer / (n_param + 4))`, below the rule of ten.
- **Sparse cells**: the primary-education group is `r n_primary` people
  (`r sum(dat_model$education_level == "Primary" & dat_model$adequate_knowledge == 1)`
  events) and the 18-24 band `r sum(dat_model$age_group == "18-24")`
  (`r sum(dat_model$age_group == "18-24" & dat_model$adequate_knowledge == 0)`
  non-events). No separation, but a student who leaves 18-24 as the default
  reference level gets uninterpretably wide intervals on every age contrast;
  setting the reference to the largest band (over 65) fixes it.
- **Collinearity present but tolerable**: max VIF `r sprintf("%.2f", max_vif)`
  (`r max_vif_term`), from the age-employment overlap
  (`r sprintf("%.0f", 100 * n_pensioner_old / n_pensioner)`% of pensioners are
  over 65; Cramér's V `r sprintf("%.2f", v_age_emp)`). It is enough to strip
  significance from all six age contrasts, which is a good teaching example but
  frustrates students who expect age to "work".
- **All predictors are categorical**, so no per-unit odds ratio is available
  anywhere in this dataset and the `check_model()` linearity panel is
  meaningless.
