---
title: "Antibiotic knowledge score in Serbian primary-care attenders"
subtitle: "Dataset 1: Antibiotic knowledge and self-medication in Serbia"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()

# Helpers so every number in the prose is pulled from the fitted model rather
# than typed by hand.
fmt_p <- function(p) if (p < 0.001) "p < 0.001" else sprintf("p = %.3f", p)
row_of <- function(tab, trm) tab[tab$term == trm, ]
say_beta <- function(tab, trm, digits = 2) {
  r <- row_of(tab, trm)
  sprintf("%.*f (95%% CI %.*f to %.*f, %s)",
          digits, r$estimate, digits, r$conf.low, digits, r$conf.high, fmt_p(r$p.value))
}
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "serbia_antibiotics.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "knowledge_score"
predictors <- c("age_group", "sex", "education_level", "employment_status",
                "gp_visits_past_year", "health_professional_in_family")

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    # Age: ">65" is both the largest band (n = 123) and the group of substantive
    # interest, so every contrast reads as "younger adults versus the over-65s".
    age_group = factor(age_group,
                       levels = c(">65", "18-24", "25-34", "35-44", "45-54", "55-64"),
                       labels = c("Over 65", "18-24", "25-34", "35-44", "45-54", "55-64")),
    sex = factor(sex, levels = c("female", "male"), labels = c("Female", "Male")),
    # Education: "secondary" is the largest group (n = 259) and sits between the
    # other two, so both a lower and a higher contrast are available.
    education_level = factor(education_level,
                             levels = c("secondary", "primary", "tertiary"),
                             labels = c("Secondary", "Primary", "Tertiary")),
    employment_status = factor(employment_status,
                               levels = c("employed", "unemployed", "pensioner"),
                               labels = c("Employed", "Unemployed", "Pensioner")),
    # GP visits: "0" is the clinically unexposed group (no primary-care contact).
    gp_visits_past_year = factor(gp_visits_past_year,
                                 levels = c("0", "1-4", "5-10", "more than 10"),
                                 labels = c("None", "1-4", "5-10", "More than 10")),
    health_professional_in_family = factor(health_professional_in_family,
                                           levels = c(0, 1), labels = c("No", "Yes")),
    marital_status = factor(marital_status,
                            levels = c("married", "single", "divorced_widowed"),
                            labels = c("Married", "Single", "Divorced or widowed")),
    household_size = factor(household_size,
                            levels = c("1-3", "3-5", "more than 5"),
                            labels = c("1-3", "3-5", "More than 5"))
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

# Verification of the deterministic link between the two knowledge variables:
# `adequate_knowledge` is exactly the indicator `knowledge_score >= 9`.
adeq_identical <- identical(dat_raw$adequate_knowledge,
                            as.numeric(dat_raw$knowledge_score >= 9))
rng_lo <- range(dat_raw$knowledge_score[dat_raw$adequate_knowledge == 0])
rng_hi <- range(dat_raw$knowledge_score[dat_raw$adequate_knowledge == 1])

# Confounding structure quoted in the text.
n_primary        <- sum(dat$education_level == "Primary")
n_primary_old    <- sum(dat$education_level == "Primary" & dat$age_group == "Over 65")
n_pensioner      <- sum(dat$employment_status == "Pensioner")
n_pensioner_old  <- sum(dat$employment_status == "Pensioner" & dat$age_group == "Over 65")
cramers_v <- function(x, y) {
  tt <- suppressWarnings(table(x, y))
  sqrt(as.numeric(suppressWarnings(chisq.test(tt))$statistic) /
         (sum(tt) * (min(dim(tt)) - 1)))
}
v_age_emp <- cramers_v(dat$age_group, dat$employment_status)
v_age_edu <- cramers_v(dat$age_group, dat$education_level)
```

# Introduction

Antibiotic resistance in south-eastern Europe is driven in part by how the
general public uses antibiotics, and public knowledge is the usual target of
national stewardship campaigns. These data come from a cross-sectional
questionnaire survey of `r nrow(dat_raw)` adults who consulted general
practitioners at four health centres in Serbia, reported by Horvat and
colleagues (*PLOS ONE* 2017;12(7):e0180799). Each respondent answered a
12-item quiz about what antibiotics are for and how they should be taken,
scored 0 to 12. Knowing which groups score poorly tells campaign designers
where to aim.

**Aim.** Among Serbian adults attending primary care, we examined which of age
group, sex, education, employment status, frequency of GP contact and having a
health professional in the family are independently associated with the
antibiotic knowledge score, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** Cross-sectional, interviewer-administered questionnaire survey
of adults consulting general practitioners at four Serbian health centres;
`r nrow(dat_raw)` respondents completed the entire questionnaire and all
`r nrow(dat_raw)` are analysed here. All measurements were taken on a single
occasion, so this is a snapshot rather than a follow-up study.

**Variables.** The outcome is `knowledge_score`, the total antibiotic knowledge
score in points on a 0-12 scale, where higher means better knowledge; in this
sample it runs from `r min(dat$knowledge_score)` to
`r max(dat$knowledge_score)`. Predictors are `age_group` (six bands; reference:
over 65, the largest band at n = `r sum(dat$age_group == "Over 65")`), `sex`
(reference: female, the majority at n = `r sum(dat$sex == "Female")`),
`education_level` (primary / secondary / tertiary; reference: secondary, the
largest group at n = `r sum(dat$education_level == "Secondary")`),
`employment_status` (reference: employed, n =
`r sum(dat$employment_status == "Employed")`), `gp_visits_past_year` (four
bands of visits in the previous 12 months; reference: none, the group with no
primary-care contact, n = `r sum(dat$gp_visits_past_year == "None")`) and
`health_professional_in_family` (reference: no, n =
`r sum(dat$health_professional_in_family == "No")`). All six are categorical,
so every estimate is a difference in mean score against the stated reference
level; there are no continuous predictors and therefore no per-unit slopes.

Two of the eight available background variables were left out deliberately.
`marital_status` is largely a restatement of age in this sample (Cramér's
V = `r sprintf("%.2f", cramers_v(dat$age_group, dat$marital_status))` with
`age_group`: nearly all the divorced or widowed respondents are in the older
bands), so it would add two parameters and inflate the variance of the age
contrasts without adding a distinct construct. `household_size` is a
three-band count with no plausible mechanism linking it to what a person knows
about antibiotics, and it is unrelated to the outcome here
(univariable p = `r sprintf("%.2f", anova(lm(knowledge_score ~ household_size, dat))[["Pr(>F)"]][1])`).
`id` is a row number and is never modelled.

`adequate_knowledge` is **not** used, in either direction. It is the indicator
`knowledge_score >= 9` and nothing else: in this file the respondents flagged 0
score `r rng_lo[1]` to `r rng_lo[2]` and those flagged 1 score `r rng_hi[1]` to
`r rng_hi[2]`, and re-deriving the flag from the score reproduces the deposited
column exactly (`identical()` returns `r adeq_identical`). Putting it in a model
for `knowledge_score` would be circular, and the reverse — modelling
`adequate_knowledge` on `knowledge_score` — is complete separation. The two
columns are one outcome measured two ways, continuous and dichotomised, and only
one of them can be the response in any given analysis.

**Statistical analysis.** Linear regression (`lm()`), justified because the
outcome is a continuous count of correct answers spanning `r min(dat$knowledge_score)`
to `r max(dat$knowledge_score)` points with a roughly unimodal distribution. We
fitted a univariable model for each predictor first, then a single
multivariable model containing all of them. Predictors were chosen in advance
on subject-matter grounds, not by significance. Significance threshold 0.05;
estimates are reported with 95% confidence intervals. Assumptions were checked
with `performance::check_model()`. Analysis used `r R.version.string` with
tidyverse, gtsummary and performance.

**Missing data.** There is none. Every one of the `r ncol(dat_raw)` columns is
`r sprintf("%.0f", 100 * mean(complete.cases(dat_raw)))`% complete, so the
complete-case model uses all `r nrow(dat_model)` of `r nrow(dat)` rows and
`r n_dropped` were dropped. The survey retained only respondents who completed
the entire questionnaire, so there is no missing-data story to tell here — but
note that this is itself a form of selection, since partial responders were
excluded before the file was created and nothing is known about them.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(missing = "no") |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

The sample skews older and female: about a quarter are over 65 and
`r sprintf("%.0f", 100 * mean(dat_model$sex == "Female"))`% are women. Education
is concentrated in the secondary and tertiary bands, with only
`r n_primary` respondents educated to primary level. The median knowledge score
is `r median(dat_model$knowledge_score)` of 12 (IQR
`r quantile(dat_model$knowledge_score, 0.25)` to
`r quantile(dat_model$knowledge_score, 0.75)`), so the typical respondent
answers about three-quarters of the items correctly and the scale is used
mainly in its upper half.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Antibiotic knowledge score by education level, with respondents split by whether a family member works in health care. Boxes show medians and interquartile ranges; points are individual respondents."}
ggplot(dat_model, aes(x = education_level, y = knowledge_score)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, fill = "grey93", colour = "grey40") +
  geom_jitter(aes(colour = health_professional_in_family),
              width = 0.18, height = 0.18, alpha = 0.6, size = 1.4) +
  scale_colour_manual(values = c("No" = "#7f8c9b", "Yes" = "#2c7fb8")) +
  scale_y_continuous(breaks = seq(2, 12, 2), limits = c(1, 12.6)) +
  labs(x = "Highest education level", y = "Antibiotic knowledge score (0-12)",
       colour = "Health professional\nin the family")
```

Knowledge rises steadily across the three education bands, and within every
band the respondents with a health professional in the family sit slightly
higher; the primary-education group is both the lowest-scoring and much the
smallest, so its box is based on only `r n_primary` people.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv
```

```{r univariable-global}
# Global (Wald F) tests for each multi-level predictor, so that factors are not
# judged on a single dummy contrast.
uni_global <- map_dfr(predictors, function(p) {
  m <- lm(reformulate(p, outcome), data = dat_model)
  tibble(term = p, p.value = anova(m)[["Pr(>F)"]][1],
         r2 = summary(m)$r.squared)
})
gp <- function(v) fmt_p(uni_global$p.value[uni_global$term == v])
gr2 <- function(v) sprintf("%.3f", uni_global$r2[uni_global$term == v])
uv_tab <- map_dfr(predictors, ~ tidy(lm(reformulate(.x, outcome), data = dat_model),
                                     conf.int = TRUE))
```

Education is by far the strongest crude predictor. Relative to secondary
education, tertiary education is associated with a
`r say_beta(uv_tab, "education_levelTertiary")` point higher score, while
primary education is associated with a
`r say_beta(uv_tab, "education_levelPrimary")` point difference — a spread of
close to two points on a 12-point scale, and on its own education accounts for
R² = `r gr2("education_level")` of the variance (`r gp("education_level")`).
Age is the next strongest (`r gp("age_group")`, R² = `r gr2("age_group")`):
every younger band scores above the over-65s, for example the 18-24 group by
`r say_beta(uv_tab, "age_group18-24")` points. Having a health professional in
the family is associated with a `r say_beta(uv_tab, "health_professional_in_familyYes")`
point advantage (`r gp("health_professional_in_family")`), and employment
status is also associated crudely (`r gp("employment_status")`, driven by
pensioners scoring low). GP visits show a weak crude gradient in the
unexpected direction (`r gp("gp_visits_past_year")` overall; respondents
attending more than ten times score `r say_beta(uv_tab, "gp_visits_past_yearMore than 10")`
points *below* those who never attended), and sex is not associated at all
(`r gp("sex")`).

## Why adjustment is needed

The crude estimates above are not independent of each other, because the
background variables in this survey travel together in a very specific way.
`age_group` is the clearest confounder of the education association:
`r n_primary_old` of the `r n_primary` primary-educated respondents
(`r sprintf("%.0f", 100 * n_primary_old / n_primary)`%) are over 65, a legacy of
when compulsory schooling expanded in the region, and age is itself associated
with lower knowledge scores. Without adjustment, part of what looks like an
education effect is simply the age composition of the education groups.
`employment_status` confounds in the same direction and is almost a
restatement of age at the top end: `r n_pensioner_old` of the
`r n_pensioner` pensioners (`r sprintf("%.0f", 100 * n_pensioner_old / n_pensioner)`%)
are over 65, and Cramér's V between age and employment is
`r sprintf("%.2f", v_age_emp)`, so the crude pensioner deficit in knowledge may
be nothing more than the age effect wearing a different label. A third,
weaker route runs through `gp_visits_past_year`: older respondents attend far
more often (`r sprintf("%.0f", 100 * mean(dat$gp_visits_past_year[dat$age_group == "Over 65"] %in% c("5-10", "More than 10")))`%
of the over-65s report five or more visits, against
`r sprintf("%.0f", 100 * mean(dat$gp_visits_past_year[dat$age_group != "Over 65"] %in% c("5-10", "More than 10")))`%
of everyone else), so any crude association between GP contact and knowledge is
partly an age association. Adjusting for all six simultaneously is what turns
these into within-stratum comparisons — the education contrast becomes
"comparable respondents of the same age and employment status", which is the
question stewardship campaigns actually need answered.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv
```

```{r fit-stats}
gl <- glance(fit)
mv_tab <- tidy(fit, conf.int = TRUE)
tibble(
  Statistic = c("R-squared", "Adjusted R-squared", "Residual SE", "N",
                "Model df", "Observations per parameter"),
  Value = c(round(gl$r.squared, 3), round(gl$adj.r.squared, 3),
            round(gl$sigma, 2), nrow(dat_model), gl$df,
            round(nrow(dat_model) / gl$df, 1))
) |>
  gt() |>
  tab_header(title = "Model fit")
```

Adjusting for all six predictors, education remains the dominant association:
conditional on age, sex, employment, GP contact and family health-care
exposure, tertiary education is associated with a
`r say_beta(mv_tab, "education_levelTertiary")` point higher score than
secondary education, and primary education with a
`r say_beta(mv_tab, "education_levelPrimary")` point difference. Having a health
professional in the family retains an independent
`r say_beta(mv_tab, "health_professional_in_familyYes")` point advantage. The
age contrasts survive adjustment in attenuated form — the 18-24 band scores
`r say_beta(mv_tab, "age_group18-24")` points above the over-65s and the 35-44
band `r say_beta(mv_tab, "age_group35-44")` — but employment status no longer
carries any information once age is in the model (pensioner versus employed:
`r say_beta(mv_tab, "employment_statusPensioner")`). Neither sex
(`r say_beta(mv_tab, "sexMale")`) nor any GP-visit contrast is associated. The
model explains R² = `r round(gl$r.squared, 3)` of the variance in knowledge
score (adjusted R² = `r round(gl$adj.r.squared, 3)`), with a residual SD of
`r round(gl$sigma, 2)` points against a raw SD of
`r round(sd(dat_model$knowledge_score), 2)`: real, but most of what makes one
respondent better informed than another is not captured by demography.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

The employment-status estimates move the most and are the clearest case of
confounding by age. Crudely, pensioners score
`r say_beta(uv_tab, "employment_statusPensioner")` points below the employed;
adjusted, that collapses to `r say_beta(mv_tab, "employment_statusPensioner")`
— essentially all the way to the null and no longer significant. Pensioners
score badly because they are old, not because they are pensioners. The age
contrasts move in the opposite way but for the same reason: the crude 18-24
advantage of `r say_beta(uv_tab, "age_group18-24")` points *shrinks* to
`r say_beta(mv_tab, "age_group18-24")` once education is held constant,
because young respondents in this sample are also much better educated, so part
of the crude youth advantage was education. Education itself barely moves
(tertiary `r say_beta(uv_tab, "education_levelTertiary")` crude versus
`r say_beta(mv_tab, "education_levelTertiary")` adjusted; primary
`r say_beta(uv_tab, "education_levelPrimary")` versus
`r say_beta(mv_tab, "education_levelPrimary")`), which is the key result: the
education gradient is not a disguised age gradient, whereas the pensioner
gradient is. Having a health professional in the family also barely moves
(`r say_beta(uv_tab, "health_professional_in_familyYes")` to
`r say_beta(mv_tab, "health_professional_in_familyYes")`), so it is not standing
in for education or age. The GP-visit gradient is the third estimate to move materially and it moves
all the way through the null: the most frequent attenders score
`r say_beta(uv_tab, "gp_visits_past_yearMore than 10")` points below
non-attenders crudely, but `r say_beta(mv_tab, "gp_visits_past_yearMore than 10")`
after adjustment — the crude deficit was age and education, since the people
attending most often are the oldest and least educated, and once those are held
constant frequent attendance carries no information about knowledge at all. Sex
is null in both columns and moves hardly at all
(`r say_beta(uv_tab, "sexMale")` to `r say_beta(mv_tab, "sexMale")`). No
predictor crosses from non-significant to significant; the two significance
changes are employment status and frequent GP attendance both losing it, which
is the expected direction when the crude estimates were confounded by age.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r diag-tests}
check_heteroscedasticity(fit)
check_normality(fit)
check_collinearity(fit)
```

```{r diag-numbers}
het_p  <- as.numeric(check_heteroscedasticity(fit))
norm_p <- as.numeric(check_normality(fit))
vif_tab <- check_collinearity(fit)
max_vif <- max(vif_tab$VIF)
max_vif_term <- vif_tab$Term[which.max(vif_tab$VIF)]
max_cook <- max(cooks.distance(fit))
```

Two assumptions fail formally and both trace back to the same feature of the
outcome. Homogeneity of variance is rejected
(`check_heteroscedasticity()`, `r fmt_p(het_p)`): because the score is capped at
12 and most respondents sit near the top, groups with high fitted means have
less room to vary than groups with low fitted means, which shows up as a funnel
in the fitted-versus-residual panel. Normality of residuals is also rejected
(`r fmt_p(norm_p)`); the Q-Q plot bends at the lower tail, driven by the handful
of respondents scoring 2 to 5. Neither is fatal at n = `r nrow(dat_model)` — the
central limit theorem protects the coefficient estimates and the intervals are
only mildly affected — but a robust (sandwich) standard error would be the
honest fix, and this is the single most important caveat on the model. The
linearity panel is uninformative here because every predictor is categorical,
so the fitted values take only a few dozen distinct levels. Collinearity is
acceptable: the largest VIF is `r sprintf("%.2f", max_vif)`
(`r max_vif_term`), which is elevated by the age-employment overlap described
above but stays well below the conventional threshold of 5. No observation is
influential (largest Cook's distance `r sprintf("%.3f", max_cook)`, far below
0.5). The posterior-predictive panel shows the model's simulated data are
smoother than the observed score distribution, which is expected: the outcome
is an integer count of correct answers and the model treats it as continuous.

# Discussion

Among Serbian adults attending primary care, formal education is the
characteristic most strongly and independently associated with antibiotic
knowledge: tertiary-educated respondents score about a point higher than
secondary-educated ones and primary-educated respondents nearly a point lower,
and this gradient is not explained away by age, employment or health-service
contact. Having a family member who works in health care is worth roughly half
a point independently, which is a plausible informal-information effect. Age
matters somewhat, sex and frequency of GP visits do not, and the apparent
knowledge deficit among pensioners disappears entirely once age is accounted
for. Practically, this argues for stewardship messaging pitched at low-literacy
audiences and delivered outside the consultation, since attending the GP more
often does not appear to make people better informed.

**Limitations.** The design is cross-sectional and every predictor is a fixed
background characteristic, so nothing here is a causal effect; "education
raises knowledge" and "the sort of person who completes tertiary education also
absorbs health information" are indistinguishable in these data. Formal
diagnostics fail: residual variance is not constant (`r fmt_p(het_p)`) and
residuals are not normal (`r fmt_p(norm_p)`), both because the 0-12 score is
bounded and ceiling-heavy, so the reported confidence intervals are slightly
optimistic and robust standard errors would be preferable. All predictors are
coarse categories — age in six bands rather than years, GP visits in four bands
rather than a count — which loses information and prevents any test of a linear
trend. The `r n_primary`-person primary-education group is small, so that
contrast is the least precise in the model. There are no measures of health
literacy, income, rural or urban residence, or prior antibiotic use in this
file, and each is a plausible common cause of both education and knowledge, so
residual confounding is likely. The survey enrolled only people
who were already consulting a GP and who completed the whole questionnaire, so
it does not represent Serbian adults who avoid primary care. Finally, the
outcome is a self-completed quiz administered in a clinic, where social
desirability may inflate scores.

**Future work.** A campaign-evaluation design measuring the same 12-item score
before and after a stewardship intervention, in the same respondents, would
separate stable educational differences from modifiable knowledge gaps.

# Feasibility notes

*Not part of a student report.*

- **Works well, and is the best of the three outcomes in this file.**
  `knowledge_score` is complete for all `r nrow(dat_raw)` rows, spans
  `r min(dat$knowledge_score)`-`r max(dat$knowledge_score)` of a possible 0-12,
  and produces a clean mix of strongly significant (education, family health
  professional), partly significant (age), fully confounded-away (employment
  status, frequent GP attendance) and flatly null (sex) predictors — exactly
  the spread a Purpose 3 exercise needs, including two textbook confounding
  demonstrations.
- **Complete-case N = `r nrow(dat_model)`** of `r nrow(dat)`; `r n_dropped`
  dropped. The file is 100% complete, so there is genuinely no missing-data
  narrative available. Students who are expected to write a missing-data
  paragraph should be told to say so plainly rather than invent one.
- **The trap in this file is `adequate_knowledge`.** It is exactly
  `knowledge_score >= 9` (verified: `identical()` = `r adeq_identical`; the
  0-group spans `r rng_lo[1]`-`r rng_lo[2]` and the 1-group
  `r rng_hi[1]`-`r rng_hi[2]`). Used as a predictor here it gives R² =
  `r sprintf("%.2f", summary(lm(knowledge_score ~ adequate_knowledge, dat_raw))$r.squared)`
  on its own; used as the outcome with `knowledge_score` as predictor it is
  perfect separation. Every student working on this dataset will need warning.
- **All predictors are categorical.** There is not one continuous predictor in
  the file, so no analysis of this dataset can demonstrate a per-unit slope,
  and the `check_model()` linearity panel is meaningless. That is a real
  teaching limitation, not a defect of this particular model.
- **Diagnostics fail formally**: heteroscedasticity `r fmt_p(het_p)`, normality
  `r fmt_p(norm_p)`, both driven by the ceiling at 12. Visually mild and
  harmless at this n, but a student running `check_model()` will get two red
  warnings and needs to know what to do with them.
- **Collinearity is present but tolerable**: `age_group` and
  `employment_status` overlap heavily (Cramér's V `r sprintf("%.2f", v_age_emp)`;
  `r sprintf("%.0f", 100 * n_pensioner_old / n_pensioner)`% of pensioners are
  over 65), max VIF `r sprintf("%.2f", max_vif)`. Adding `marital_status` as
  well (V = `r sprintf("%.2f", cramers_v(dat$age_group, dat$marital_status))`
  with age) pushes this further and is best avoided.
- **Modest R² (`r round(gl$r.squared, 3)`)**, which is normal for a
  demographics-only survey model but will disappoint students expecting a
  "strong" model.
