---
title: "Self-medication with antibiotics in Serbian primary-care attenders"
subtitle: "Dataset 1: Antibiotic knowledge and self-medication in Serbia"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()

# Helpers so every number in the prose is pulled from the fitted model rather
# than typed by hand.
fmt_p <- function(p) if (p < 0.001) "p < 0.001" else sprintf("p = %.3f", p)
row_of <- function(tab, trm) tab[tab$term == trm, ]
say_or <- function(tab, trm) {
  r <- row_of(tab, trm)
  sprintf("OR %.2f (95%% CI %.2f to %.2f, %s)",
          r$estimate, r$conf.low, r$conf.high, fmt_p(r$p.value))
}
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "serbia_antibiotics.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "self_medication"
predictors <- c("age_group", "sex", "education_level", "employment_status",
                "gp_visits_past_year", "health_professional_in_family")

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    # Age: ">65" is both the largest band (n = 123) and the group with the
    # highest self-medication prevalence, so every contrast reads as "younger
    # adults versus the over-65s".
    age_group = factor(age_group,
                       levels = c(">65", "18-24", "25-34", "35-44", "45-54", "55-64"),
                       labels = c("Over 65", "18-24", "25-34", "35-44", "45-54", "55-64")),
    sex = factor(sex, levels = c("female", "male"), labels = c("Female", "Male")),
    # Education: "secondary" is the largest group (n = 259) and sits between the
    # other two, so both a lower and a higher contrast are available.
    education_level = factor(education_level,
                             levels = c("secondary", "primary", "tertiary"),
                             labels = c("Secondary", "Primary", "Tertiary")),
    employment_status = factor(employment_status,
                               levels = c("employed", "unemployed", "pensioner"),
                               labels = c("Employed", "Unemployed", "Pensioner")),
    # GP visits: "0" is the clinically unexposed group (no primary-care contact).
    gp_visits_past_year = factor(gp_visits_past_year,
                                 levels = c("0", "1-4", "5-10", "more than 10"),
                                 labels = c("None", "1-4", "5-10", "More than 10")),
    health_professional_in_family = factor(health_professional_in_family,
                                           levels = c(0, 1), labels = c("No", "Yes")),
    marital_status = factor(marital_status,
                            levels = c("married", "single", "divorced_widowed"),
                            labels = c("Married", "Single", "Divorced or widowed")),
    household_size = factor(household_size,
                            levels = c("1-3", "3-5", "more than 5"),
                            labels = c("1-3", "3-5", "More than 5"))
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

n_event    <- sum(dat_model[[outcome]] == 1)
n_nonevent <- sum(dat_model[[outcome]] == 0)
n_rarer    <- min(n_event, n_nonevent)

# Confounding structure quoted in the text.
n_primary       <- sum(dat$education_level == "Primary")
n_primary_old   <- sum(dat$education_level == "Primary" & dat$age_group == "Over 65")
n_pensioner     <- sum(dat$employment_status == "Pensioner")
n_pensioner_old <- sum(dat$employment_status == "Pensioner" & dat$age_group == "Over 65")
cramers_v <- function(x, y) {
  tt <- suppressWarnings(table(x, y))
  sqrt(as.numeric(suppressWarnings(chisq.test(tt))$statistic) /
         (sum(tt) * (min(dim(tt)) - 1)))
}
v_age_emp <- cramers_v(dat$age_group, dat$employment_status)
```

# Introduction

Self-medication with antibiotics — taking them without a prescription, keeping
leftovers for the next illness, or asking a pharmacist directly — is one of the
main routes by which antibiotics reach people who do not need them, and it is
common across south-eastern Europe. These data come from a cross-sectional
questionnaire survey of `r nrow(dat_raw)` adults who consulted general
practitioners at four health centres in Serbia, reported by Horvat and
colleagues (*PLOS ONE* 2017;12(7):e0180799). Just over half the sample
(`r n_event`/`r nrow(dat_model)`,
`r sprintf("%.0f", 100 * n_event / nrow(dat_model))`%) met the study's
definition of self-medication. Identifying who does it is the first step in
deciding whom stewardship programmes should target.

**Aim.** Among Serbian adults attending primary care, we examined which of age
group, sex, education, employment status, frequency of GP contact and having a
health professional in the family are independently associated with
self-medication with antibiotics, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** Cross-sectional, interviewer-administered questionnaire survey
of adults consulting general practitioners at four Serbian health centres;
`r nrow(dat_raw)` respondents completed the entire questionnaire and all
`r nrow(dat_raw)` are analysed here. All measurements were taken on a single
occasion, so exposure and outcome were recorded at the same moment.

**Variables.** The outcome is `self_medication`, a binary indicator; the modelled
event is **self-medication = yes** under the source paper's definition
(`r n_event` events, `r n_nonevent` non-events, so the rarer class has
`r n_rarer`). Predictors are `age_group` (six bands; reference: over 65, the
largest band at n = `r sum(dat$age_group == "Over 65")`), `sex` (reference:
female, n = `r sum(dat$sex == "Female")`), `education_level` (primary /
secondary / tertiary; reference: secondary, the largest group at n =
`r sum(dat$education_level == "Secondary")`), `employment_status` (reference:
employed, n = `r sum(dat$employment_status == "Employed")`),
`gp_visits_past_year` (four bands of visits in the previous 12 months;
reference: none, the group with no primary-care contact, n =
`r sum(dat$gp_visits_past_year == "None")`) and
`health_professional_in_family` (reference: no, n =
`r sum(dat$health_professional_in_family == "No")`). All six are categorical, so
every estimate is an odds ratio against the stated reference level; the file
contains no continuous predictors at all.

Two of the eight available background variables were left out deliberately.
`marital_status` is largely a restatement of age here (Cramér's V =
`r sprintf("%.2f", cramers_v(dat$age_group, dat$marital_status))` with
`age_group`), so it would consume two parameters and inflate the variance of the
age contrasts without adding a distinct construct. `household_size` is
unrelated to the outcome
(univariable `r fmt_p(anova(glm(self_medication ~ household_size, dat, family = binomial), test = "LRT")[["Pr(>Chi)"]][2])`)
and has no plausible mechanism. `id` is a row number and is never modelled.

The two knowledge variables in this file are also excluded, on different
grounds. `knowledge_score` and `adequate_knowledge` are the same quantity
measured twice — the latter is exactly the indicator `knowledge_score >= 9`,
which re-deriving from the score reproduces exactly
(`identical()` returns
`r identical(dat_raw$adequate_knowledge, as.numeric(dat_raw$knowledge_score >= 9))`)
— so at most one of them could ever enter a model, and never both. More
importantly, neither is a *background characteristic*: knowledge is on the
causal pathway between education and behaviour, and its association with
self-medication is exactly what the source paper set out to test. Including it
here would change the question from "who self-medicates?" to "does knowledge
explain self-medication over and above demography?", which is a different
analysis with a mediation structure. This document keeps to the demographic
question.

**Statistical analysis.** Logistic regression (`glm(family = binomial)`),
justified because the outcome is binary. We fitted a univariable model for each
predictor first, then a single multivariable model containing all of them.
Predictors were chosen in advance on subject-matter grounds, not by
significance. Estimates are odds ratios with 95% confidence intervals;
significance threshold 0.05. Assumptions were checked with
`performance::check_model()`, with attention to binned residuals, collinearity
and influential observations rather than to normality or constant variance,
which do not apply to a logistic model. Analysis used `r R.version.string` with
tidyverse, gtsummary and performance.

**Missing data.** There is none. Every one of the `r ncol(dat_raw)` columns is
`r sprintf("%.0f", 100 * mean(complete.cases(dat_raw)))`% complete, so the model
uses all `r nrow(dat_model)` of `r nrow(dat)` rows and `r n_dropped` were
dropped. There is no missing-data story to tell for this file — though the
survey kept only respondents who completed the whole questionnaire, so partial
responders were excluded before the data reached us and nothing is known about
them.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  mutate(self_medication = factor(self_medication, levels = c(0, 1),
                                  labels = c("No self-medication", "Self-medicated"))) |>
  tbl_summary(
    by = self_medication,
    missing = "no"
  ) |>
  add_n() |>
  add_overall() |>
  modify_caption("**Table 1. Sample characteristics, by self-medication status.**")
```

The two groups are close to equal in size and differ mainly on age and
education. Self-medicators are older
(`r sprintf("%.0f", 100 * mean(dat_model$age_group[dat_model$self_medication == 1] == "Over 65"))`%
are over 65, against
`r sprintf("%.0f", 100 * mean(dat_model$age_group[dat_model$self_medication == 0] == "Over 65"))`%
of non-self-medicators) and less likely to hold a tertiary degree
(`r sprintf("%.0f", 100 * mean(dat_model$education_level[dat_model$self_medication == 1] == "Tertiary"))`%
versus
`r sprintf("%.0f", 100 * mean(dat_model$education_level[dat_model$self_medication == 0] == "Tertiary"))`%).
Sex, employment, GP attendance and having a health professional in the family
look almost identical in the two groups.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Proportion of respondents reporting self-medication with antibiotics, by age group and education level. Error bars are 95% Wilson intervals for the proportion; the number above each bar is the group size. The dashed line is the overall prevalence."}
prop_dat <- dat_model |>
  count(age_group, education_level, wt = NULL, name = "n_tot") |>
  left_join(dat_model |> group_by(age_group, education_level) |>
              summarise(k = sum(self_medication), .groups = "drop"),
            by = c("age_group", "education_level")) |>
  mutate(p = k / n_tot,
         # Wilson score interval, stable for the small cells in this table.
         z = 1.96,
         centre = (p + z^2 / (2 * n_tot)) / (1 + z^2 / n_tot),
         halfw  = z * sqrt(p * (1 - p) / n_tot + z^2 / (4 * n_tot^2)) / (1 + z^2 / n_tot),
         lo = pmax(0, centre - halfw), hi = pmin(1, centre + halfw))

ggplot(prop_dat, aes(x = age_group, y = p, fill = education_level)) +
  geom_hline(yintercept = mean(dat_model$self_medication),
             linetype = "dashed", colour = "grey40") +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  geom_errorbar(aes(ymin = lo, ymax = hi), position = position_dodge(width = 0.8),
                width = 0.2, colour = "grey30") +
  geom_text(aes(y = hi + 0.04, label = n_tot),
            position = position_dodge(width = 0.8), size = 2.6, colour = "grey30") +
  scale_fill_manual(values = c("Secondary" = "#7f8c9b", "Primary" = "#c4694a",
                               "Tertiary" = "#2c7fb8")) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1.15),
                     breaks = seq(0, 1, 0.25)) +
  labs(x = "Age group", y = "Self-medicating (%)", fill = "Education")
```

Self-medication rises with age and falls with education, but the two work
against each other and several age-by-education cells contain only a handful of
people (the primary-education bars below age 65 rest on one to four
respondents each), so the crude picture is unstable and the confidence
intervals in those cells are uninformative.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = glm,
    method.args = list(family = binomial),
    exponentiate = TRUE
  )
uv
```

```{r univariable-global}
# Global likelihood-ratio tests, so multi-level factors are not judged on a
# single dummy contrast.
uni_global <- map_dfr(predictors, function(p) {
  m <- glm(reformulate(p, outcome), data = dat_model, family = binomial)
  tibble(term = p, p.value = anova(m, test = "LRT")[["Pr(>Chi)"]][2])
})
gp <- function(v) fmt_p(uni_global$p.value[uni_global$term == v])
uv_tab <- map_dfr(predictors, ~ tidy(glm(reformulate(.x, outcome), data = dat_model,
                                         family = binomial),
                                     conf.int = TRUE, exponentiate = TRUE))
```

Only two predictors are associated with self-medication at all. Education is the
strongest (`r gp("education_level")`): compared with secondary education,
tertiary education halves the odds of self-medicating
(`r say_or(uv_tab, "education_levelTertiary")`), while the small
primary-education group runs the other way
(`r say_or(uv_tab, "education_levelPrimary")`). Age is the other
(`r gp("age_group")`): every younger band self-medicates less than the over-65s,
most clearly the 18-24 group (`r say_or(uv_tab, "age_group18-24")`) and the
25-34 group (`r say_or(uv_tab, "age_group25-34")`), though the youngest band
contains only `r sum(dat_model$age_group == "18-24")` people. The remaining four
are null as whole variables: sex (`r gp("sex")`), employment status
(`r gp("employment_status")`), GP visits (`r gp("gp_visits_past_year")`) and
having a health professional in the family
(`r gp("health_professional_in_family")`), the last of these almost perfectly
flat (`r say_or(uv_tab, "health_professional_in_familyYes")`). One caveat on GP
visits: although the variable as a whole is not associated, the single contrast
of 1-4 visits against none is nominally significant
(`r say_or(uv_tab, "gp_visits_past_year1-4")`) with no gradient across the
higher bands, which is the pattern one expects from multiple testing across four
levels rather than a real dose-response.

## Why adjustment is needed

The two predictors that do show a crude association are entangled with each
other and with a third variable, so neither crude estimate can be read at face
value. `age_group` confounds the education association: `r n_primary_old` of the
`r n_primary` primary-educated respondents
(`r sprintf("%.0f", 100 * n_primary_old / n_primary)`%) are over 65, because
schooling expanded in the region within these respondents' lifetimes, and age is
independently associated with self-medicating; so part of what looks like an
education effect is the age composition of the education groups, and part of the
crude age effect is really the lower education of the older bands.
`employment_status` is the second: `r n_pensioner_old` of the `r n_pensioner`
pensioners (`r sprintf("%.0f", 100 * n_pensioner_old / n_pensioner)`%) are over
65 (Cramér's V with age = `r sprintf("%.2f", v_age_emp)`), and pensioner status
is also tied to income and to what people can afford to do about an infection,
so leaving it out risks attributing an income effect to age. `gp_visits_past_year`
is a third and more awkward case: older and less educated respondents attend
far more often
(`r sprintf("%.0f", 100 * mean(dat$gp_visits_past_year[dat$age_group == "Over 65"] %in% c("5-10", "More than 10")))`%
of the over-65s report five or more visits, against
`r sprintf("%.0f", 100 * mean(dat$gp_visits_past_year[dat$age_group != "Over 65"] %in% c("5-10", "More than 10")))`%
of everyone else), so any crude association between attendance and
self-medication would be partly an age association. Fitting all six together
turns each estimate into a within-stratum comparison — the education contrast
becomes "respondents of the same age, employment status and attendance
frequency", which is the comparison a targeting decision actually requires.

## Multivariable model

```{r multivariable}
fit <- glm(reformulate(predictors, outcome), data = dat_model, family = binomial)

mv <- tbl_regression(fit, exponentiate = TRUE)
mv
```

```{r fit-stats}
mv_tab  <- tidy(fit, conf.int = TRUE, exponentiate = TRUE)
n_param <- length(coef(fit)) - 1          # non-intercept parameters
epp     <- n_rarer / n_param
fit_null <- glm(reformulate("1", outcome), data = dat_model, family = binomial)
tjur     <- as.numeric(r2_tjur(fit))
lrt      <- anova(fit_null, fit, test = "LRT")

tibble(
  Statistic = c("AIC (fitted model)", "AIC (intercept-only model)",
                "Tjur's R-squared", "Likelihood-ratio test vs null (p)",
                "Events (self-medication = yes)", "Non-events",
                "Rarer class", "Non-intercept parameters",
                "Events per parameter"),
  Value = c(sprintf("%.1f", AIC(fit)), sprintf("%.1f", AIC(fit_null)),
            sprintf("%.3f", tjur), format.pval(lrt[["Pr(>Chi)"]][2], digits = 3, eps = 0.001),
            as.character(c(n_event, n_nonevent, n_rarer, n_param)),
            sprintf("%.1f", epp))
) |>
  gt() |>
  tab_header(title = "Model fit and events per parameter")
```

With `r n_rarer` in the rarer outcome class and `r n_param` non-intercept
parameters, there are `r round(epp, 1)` events per parameter, comfortably above
the conventional minimum of ten, so the model is not overfitted despite the
number of factor levels. Adjusting for all six predictors, tertiary education
remains independently associated with lower odds of self-medicating
(`r say_or(mv_tab, "education_levelTertiary")`), and the age gradient persists at
the young end: conditional on education, employment, attendance and sex, the
18-24 band has `r say_or(mv_tab, "age_group18-24")` and the 25-34 band
`r say_or(mv_tab, "age_group25-34")` relative to the over-65s, while the middle
bands are no longer distinguishable from the oldest
(45-54: `r say_or(mv_tab, "age_group45-54")`). Primary education is not
independently associated (`r say_or(mv_tab, "education_levelPrimary")`), nor is
sex (`r say_or(mv_tab, "sexMale")`), employment (pensioner:
`r say_or(mv_tab, "employment_statusPensioner")`), attendance (more than ten
visits: `r say_or(mv_tab, "gp_visits_past_yearMore than 10")`) or family
health-care exposure (`r say_or(mv_tab, "health_professional_in_familyYes")`).
The whole model is only marginally better than an intercept alone
(AIC `r round(AIC(fit), 1)` versus `r round(AIC(fit_null), 1)`; likelihood-ratio
`r fmt_p(lrt[["Pr(>Chi)"]][2])`) and Tjur's R² is
`r sprintf("%.3f", tjur)` — that is, these six background characteristics
discriminate self-medicators from non-self-medicators barely at all.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

Three sets of estimates move meaningfully, and they move in different
directions. The age contrasts move *away* from the null and strengthen: the
18-24 band goes from `r say_or(uv_tab, "age_group18-24")` crude to
`r say_or(mv_tab, "age_group18-24")` adjusted, and 25-34 from
`r say_or(uv_tab, "age_group25-34")` to `r say_or(mv_tab, "age_group25-34")`.
This is negative confounding by education: young respondents are much better
educated than the over-65s and education protects, so the crude age comparison
*understated* how much lower self-medication is in the young. Employment status
moves the same way and further: the crude pensioner estimate is essentially null
(`r say_or(uv_tab, "employment_statusPensioner")`) but adjusted it becomes
`r say_or(mv_tab, "employment_statusPensioner")` — pensioners look like everyone
else until age is held constant, at which point they self-medicate *less* than
employed people of the same age, a borderline suppression effect that should not
be over-read given its width and that it never crosses 0.05. The GP-visit
estimates move the opposite way, *towards* the null, and produce the only loss
of significance in the table: 1-4 visits versus none falls from
`r say_or(uv_tab, "gp_visits_past_year1-4")` to
`r say_or(mv_tab, "gp_visits_past_year1-4")`. That crude excess was age and
education — frequent attenders are older and less educated — and once those are
held constant, attendance carries nothing.

Three estimates barely move at all, and they are the informative ones. Tertiary
education is almost unchanged (`r say_or(uv_tab, "education_levelTertiary")` to
`r say_or(mv_tab, "education_levelTertiary")`), which says the education effect
is not an age effect in disguise; primary education is likewise essentially
static (`r say_or(uv_tab, "education_levelPrimary")` to
`r say_or(mv_tab, "education_levelPrimary")`) and non-significant in both
columns, despite three-quarters of that group being over 65 — the two
confounding routes evidently cancel. Sex (`r say_or(uv_tab, "sexMale")` to
`r say_or(mv_tab, "sexMale")`) and family health-care exposure
(`r say_or(uv_tab, "health_professional_in_familyYes")` to
`r say_or(mv_tab, "health_professional_in_familyYes")`) are flat and null
throughout. In summary: two age contrasts gain strength, one GP-visit contrast
loses significance, and the education result is robust to adjustment.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r diag-tests}
binned_residuals(fit)
check_collinearity(fit)
```

```{r diag-numbers}
br      <- binned_residuals(fit)
br_pct  <- 100 * attributes(br)$resid_ok
vif_tab <- check_collinearity(fit)
max_vif <- max(vif_tab$VIF)
max_vif_term <- vif_tab$Term[which.max(vif_tab$VIF)]
max_cook <- max(cooks.distance(fit))
pr <- range(fitted(fit))
```

The binned-residuals check **fails**: only
`r sprintf("%.0f", br_pct)`% of the binned residuals fall inside their error
bounds, against the 95% that would indicate an adequately specified model. In
practice this means the model's fitted probabilities do not track the observed
proportions well across the range — unsurprising given that the predicted
probabilities span only `r sprintf("%.2f", pr[1])` to `r sprintf("%.2f", pr[2])`
around a base rate of `r sprintf("%.2f", mean(dat_model$self_medication))`, so
there is almost no spread to track. This is the numerical expression of the very
weak signal reported above, not a coding error, and it is the single most
important diagnostic finding for this outcome. Collinearity is acceptable: the
largest VIF is `r sprintf("%.2f", max_vif)` (`r max_vif_term`), inflated by the
age-employment overlap but well below the threshold of 5, and it is not what is
limiting the model. No observation is influential (largest Cook's distance
`r sprintf("%.4f", max_cook)`, far below 0.5), and there is no separation: every
coefficient is finite with a finite standard error, which is guaranteed here
because the sparsest predictor cell still contains
`r min(table(dat_model$age_group, dat_model$self_medication))` respondents.
Homogeneity of variance and normality of residuals are not assumptions of a
logistic model and the corresponding panels are not interpreted. The
posterior-predictive panel is uninformative for a binary outcome that the model
predicts near 0.5 throughout.

# Discussion

Among Serbian adults attending primary care, the only characteristics
independently associated with self-medicating with antibiotics are education and
age: tertiary-educated respondents have roughly `r sprintf("%.0f", 100 * (1 - row_of(mv_tab, "education_levelTertiary")$estimate))`%
lower odds than secondary-educated ones, and adults under 35 have markedly lower
odds than the over-65s, once the other five characteristics are held constant.
Sex, how often a person sees their GP and whether a relative works in health
care carry no information at all; employment status is borderline at best
(pensioner versus employed `r say_or(mv_tab, "employment_statusPensioner")`) and
only after age is adjusted for. The honest headline, though,
is how little any of this explains: Tjur's R² is `r sprintf("%.3f", tjur)` and
the fitted probabilities never move far from the sample prevalence of
`r sprintf("%.0f", 100 * mean(dat_model$self_medication))`%, so demography does
not identify who self-medicates in any useful way. Practically, that argues
against targeting stewardship campaigns by demographic group and in favour of
universal measures — pharmacy dispensing controls, prescription enforcement —
which do not require knowing in advance who the self-medicators are.

**Limitations.** The single most serious limitation is that
`gp_visits_past_year` is bidirectional and cannot be treated as a predictor in
the ordinary sense. Someone who obtains antibiotics without a prescription has
less reason to consult a doctor, which would push attendance down among
self-medicators; equally, someone who consults often may accumulate leftover
prescriptions and become more able to self-medicate later, which would push it
up. Both stories are consistent with the flat association observed here, and
because attendance was measured over the same recall window as the behaviour,
the data cannot distinguish them; the two mechanisms may even be cancelling
each other out and hiding a real effect in each direction. Any causal reading of
that coefficient is unwarranted, and the same caution applies more weakly to
employment status. The design is cross-sectional throughout, so nothing here is
an effect. The outcome is self-reported in a clinical setting, where admitting
to a behaviour that doctors disapprove of is under-reported, and if
under-reporting is worse among the better educated then the education effect is
overstated. The binned-residual check fails, indicating the model does not fit
the data well anywhere along its very narrow predicted range. Every predictor is
a coarse category, so no dose-response or per-year trend can be tested. The
primary-education group (n = `r n_primary`) and the 18-24 band
(n = `r sum(dat_model$age_group == "18-24")`, of whom
`r sum(dat_model$age_group == "18-24" & dat_model$self_medication == 1)` self-medicated)
are small, so those two contrasts are much less precise than the table's odds
ratios suggest. The file contains no measure of income, insurance status,
distance to a pharmacy, or whether antibiotics can be bought over the counter
locally — all plausible drivers of self-medication and all unmeasured. Finally,
the sample is drawn from people who were consulting a GP, which is precisely the
population least likely to bypass one.

**Future work.** A study that records the *source* of the antibiotic (leftover
stock, pharmacy without prescription, family member) alongside attendance would
separate the two competing explanations for the null GP-visit association.

# Feasibility notes

*Not part of a student report.*

- **Usable but weak, and the weakest of the three outcomes in this file.**
  The outcome is beautifully balanced (`r n_event` versus `r n_nonevent`,
  `r sprintf("%.0f", 100 * mean(dat_model$self_medication))`% prevalence) and
  100% complete, so there is no separation, no sparse-outcome problem and
  `r round(epp, 1)` events per parameter. What it lacks is signal: Tjur's R² =
  `r sprintf("%.3f", tjur)`, AIC `r round(AIC(fit), 1)` against
  `r round(AIC(fit_null), 1)` for the null model, and only two of six predictors
  reach significance.
- **MAJOR: the binned-residuals check fails** (`r sprintf("%.0f", br_pct)`%
  within bounds, want 95%+). Students who run `check_model()` will get a red
  flag and there is no fix available to them — the model genuinely does not fit,
  because the predictors do not discriminate. Worth deciding in advance whether
  that is a teachable moment or an unfair trap.
- **Complete-case N = `r nrow(dat_model)`** of `r nrow(dat)`; `r n_dropped`
  dropped. The file is 100% complete, so there is no missing-data narrative to
  write and students should be told to say so plainly rather than invent one.
- **`gp_visits_past_year` is bidirectional** with respect to this outcome
  (self-medicating may reduce GP contact, or frequent GP contact may supply the
  leftover antibiotics). It is null here, so nothing turns on it statistically,
  but any student who writes a causal sentence about it is wrong. This makes
  `self_medication` the better of the three outcomes for teaching reverse
  causation and the worse one for teaching effect estimation.
- **The knowledge trap applies here too, in a subtler form.** Neither
  `knowledge_score` nor `adequate_knowledge` can be used as a predictor without
  turning this into a mediation analysis (knowledge sits between education and
  behaviour), and they can never both be used because
  `adequate_knowledge` is exactly `knowledge_score >= 9`. Students will be
  tempted, because the crude difference is real: self-medicators average
  `r sprintf("%.2f", mean(dat$knowledge_score[dat$self_medication == 1]))`
  points against
  `r sprintf("%.2f", mean(dat$knowledge_score[dat$self_medication == 0]))` for
  the rest.
- **Sparse cells in two contrasts.** The 18-24 band has only
  `r sum(dat_model$age_group == "18-24")` respondents
  (`r sum(dat_model$age_group == "18-24" & dat_model$self_medication == 1)`
  events) and the primary-education group only `r n_primary`
  (`r sum(dat_model$education_level == "Primary" & dat_model$self_medication == 1)`
  events). Not sparse enough to break the fit, but enough that a student using
  18-24 as the age reference level — the obvious choice — gets confidence
  intervals up to twenty-fold wide. Setting the reference to the largest band
  (over 65) fixes this and is worth flagging explicitly.
- **All predictors are categorical**, so no analysis of this file can
  demonstrate a per-unit odds ratio, and the `check_model()` linearity panel is
  meaningless.
- **Collinearity is present but tolerable**: max VIF `r sprintf("%.2f", max_vif)`
  (`r max_vif_term`), from the age-employment overlap
  (`r sprintf("%.0f", 100 * n_pensioner_old / n_pensioner)`% of pensioners are
  over 65; Cramér's V `r sprintf("%.2f", v_age_emp)`). Adding `marital_status`
  makes it worse for no gain.
