---
title: "Simpson's paradox: percent body fat vs height (AIS)"
subtitle: "`DAAG::ais` — pooled slope can reverse after adjusting for sex"
author: "Research Statistics with R — Health Sciences"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    fig-cap-location: top
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(DAAG)
library(dplyr)
library(ggplot2)
library(broom)
library(patchwork)

theme_set(
  theme_minimal(base_size = 13) +
    theme(
      plot.title    = element_text(face = "bold"),
      plot.subtitle = element_text(colour = "grey35", size = rel(0.92)),
      legend.position = "top"
    )
)
```

## Question

In **`DAAG::ais`**, **elite athletes**:

> Does **height** (`ht`, cm) predict **percent body fat** (`pcBfat`)?

**Pooled** OLS often yields a **negative** slope: taller people in the combined sample tend to **lower** % body fat. That clashes with what you see **within each sex**: among women (and more weakly among men), taller athletes tend to carry **more** % fat along the spectrum of the cohort. Men are taller on average but much leaner on average, so pooling creates a classic **sign reversal** once you fit **`pcBfat ~ ht + sex`**.

::: {.callout-note}
## Companion example (attenuation, not flip)

For **no sign reversal**—a strong pooled slope that **shrinks toward zero** after `+ sex`—use **`confounding_ais_ferritin_wt.qmd`** (`ferr ~ wt`). Same dataset, different geometry.
:::

```{r}
#| label: load
data(ais, package = "DAAG")
d <- ais |>
  filter(!is.na(pcBfat), !is.na(ht), !is.na(sex)) |>
  mutate(
    sex = factor(sex, levels = c("f", "m"), labels = c("Female", "Male"))
  )
```

::: {.callout-note}
## Data

`data(ais, package = "DAAG")`. Complete cases: *N* = **`r nrow(d)`**. See `?ais` for units (**`ht`** is cm).

**Reference:** Maindonald & Braun, *Data Analysis and Graphics Using R* (`DAAG`).
:::

## Descriptive pattern

```{r}
#| label: summarise
d |>
  group_by(sex) |>
  summarise(
    n = n(),
    mean_ht_cm = mean(ht),
    sd_ht = sd(ht),
    mean_pcBfat = mean(pcBfat),
    sd_pcBfat = sd(pcBfat)
  ) |>
  knitr::kable(digits = 2, caption = "Height and % body fat by sex (AIS)")
```

Men are **taller** and **leaner** (% fat) on average—sex lines up with both `ht` and `pcBfat` in opposite directions across groups, which is exactly the setup for a **marginal** association that disagrees with **within-sex** trends.

## Regression: bivariate vs adjusted

```{r}
#| label: models
mod_bi  <- lm(pcBfat ~ ht, data = d)
mod_adj <- lm(pcBfat ~ ht + sex, data = d)

bind_rows(
  tidy(mod_bi)  |> mutate(model = "Bivariate:  pcBfat ~"),
  tidy(mod_adj) |> mutate(model = "Adjusted:     pcBfat ~")
) |>
  filter(term %in% c("(Intercept)", "ht", "sexMale")) |>
  transmute(model, term, estimate, std.error, statistic, p.value) |>
  knitr::kable(digits = 4, caption = "Focus on the slope for `ht` (cm → % body fat)")
```

After `+ sex`, the slope on height is **positive** and substantial: holding sex fixed, taller athletes tend to higher % fat. The **pooled** line was **negative** because it absorbed the sex composition.

```{r}
#| label: within-sex
#| tbl-cap: "Separate OLS within each sex (exploratory)."
bind_rows(
  tidy(lm(pcBfat ~ ht, data = filter(d, sex == "Female"))) |>
    mutate(sex = "Female"),
  tidy(lm(pcBfat ~ ht, data = filter(d, sex == "Male"))) |>
    mutate(sex = "Male")
) |>
  filter(term == "ht") |>
  select(sex, estimate, std.error, statistic, p.value) |>
  knitr::kable(digits = 4)
```

## Scatterplots: pooled vs stratified

```{r}
#| label: fig-compare
#| fig-height: 5
#| fig-cap: "Left: pooled line slopes down. Right: within-sex lines slope up; dashed = misleading pooled fit."
xlab <- "Height (cm)"
ylab <- "Percent body fat (%)"

p_pooled <- ggplot(d, aes(ht, pcBfat)) +
  geom_point(alpha = 0.35, colour = "grey35", size = 2.2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#6a040f", linewidth = 1) +
  labs(
    title = "Ignoring sex",
    subtitle = "pcBfat ~ ht — pooled (negative slope)",
    x = xlab, y = ylab
  )

pal <- c(Female = "#9b2226", Male = "#005f73")

p_strat <- ggplot(d, aes(ht, pcBfat, colour = sex)) +
  geom_point(alpha = 0.45, size = 2.2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1, aes(colour = sex)) +
  geom_smooth(
    method = "lm", se = FALSE,
    aes(group = 1), colour = "grey20",
    linetype = "dashed", linewidth = 0.75, inherit.aes = TRUE
  ) +
  scale_colour_manual(values = pal, name = NULL) +
  labs(
    title = "By sex",
    subtitle = "Solid = within-sex OLS; dashed = pooled line",
    x = xlab, y = ylab
  )

p_pooled + p_strat + plot_layout(widths = c(1, 1))
```

```{r}
#| label: fig-one-panel
#| fig-height: 4.8
#| fig-cap: "Slide-friendly: upward slopes within sex vs downward dashed pooled line."
ggplot(d, aes(ht, pcBfat, colour = sex)) +
  geom_point(alpha = 0.45, size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1, aes(colour = sex)) +
  geom_smooth(
    method = "lm", se = FALSE,
    aes(group = 1), colour = "grey25",
    linetype = "dashed", linewidth = 0.9, inherit.aes = TRUE
  ) +
  scale_colour_manual(values = pal, name = "Sex") +
  labs(
    title = "Percent body fat vs height (AIS)",
    subtitle = "Dashed = pooled; coloured = within-sex fits",
    x = xlab, y = ylab
  )
```

## Adjusted model: parallel slopes

```{r}
#| label: fig-adjusted-model
#| fig-height: 4.8
#| fig-cap: "Predictions from lm(pcBfat ~ ht + sex): positive common slope for height, intercept shift by sex."
ht_seq <- seq(min(d$ht), max(d$ht), length.out = 80)
d_pred <- expand.grid(ht = ht_seq, sex = levels(d$sex), stringsAsFactors = FALSE)
d_pred$sex <- factor(d_pred$sex, levels = levels(d$sex))
d_pred$fit <- predict(mod_adj, newdata = d_pred)

ggplot(d, aes(ht, pcBfat, colour = sex)) +
  geom_point(alpha = 0.35, size = 2.2) +
  geom_line(data = d_pred, aes(y = fit, colour = sex), linewidth = 1) +
  scale_colour_manual(values = pal, name = "Sex") +
  labs(
    title = "Adjusted model (parallel slopes)",
    subtitle = "Lines = predictions from lm(pcBfat ~ ht + sex)",
    x = xlab, y = ylab
  )
```

## Take-home

1. **Sign flips** in real data mean the **pooled** association is a **bad summary** when a binary like **sex** shifts both `X` and `Y`.
2. **`lm(Y ~ X + sex)`** estimates the slope of `X` **at fixed sex**—here it recovers a **positive** height effect that was **hidden** in the marginal scatter.
3. Athletes are a **selected** sample; this is for **methods** and visual intuition.

::: {.callout-important}
## Install

If needed: `install.packages("DAAG")` and `install.packages("patchwork")`.
:::
