---
title: "Confounding: plasma ferritin vs body weight (AIS)"
subtitle: "`DAAG::ais` — when sex explains a strong bivariate association"
author: "Research Statistics with R — Health Sciences"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    fig-cap-location: top
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(DAAG)
library(dplyr)
library(ggplot2)
library(broom)
library(patchwork)

theme_set(
  theme_minimal(base_size = 13) +
    theme(
      plot.title    = element_text(face = "bold"),
      plot.subtitle = element_text(colour = "grey35", size = rel(0.92)),
      legend.position = "top"
    )
)
```

## Question

The **Australian Institute of Sport (AIS)** dataset contains blood and body-composition measures on athletes. Consider:

> Does **body weight** (`wt`, kg) predict **plasma ferritin** (`ferr`, iron-storage protein; ng dl⁻¹ in the `DAAG` codebook)?

In the **pooled** sample, higher weight strongly predicts higher ferritin. That pattern largely **vanishes** after adding **sex**: men are heavier on average and have higher ferritin on average, so the pooled slope mixes **between-sex** differences with **within-sex** trends. This is a straightforward **confounding** story — same direction bivariate and within-sex (no sign flip), but the naive effect is **inflated** and becomes negligible once `sex` is controlled.

::: {.callout-note}
## Want a **sign reversal** instead?

**`pcBfat ~ height`** in the same **`DAAG::ais`** data is a textbook **Simpson’s paradox**: pooled slope **negative**, **`+ sex`** flips it **positive**. See **`confounding_ais_bodyfat_height.qmd`**.  
This ferritin example stays on **attenuation** (strong pooled slope → **nonsignificant** after `+ sex`, same direction).
:::

```{r}
#| label: load
data(ais, package = "DAAG")
d <- ais |>
  filter(!is.na(ferr), !is.na(wt), !is.na(sex)) |>
  mutate(
    sex = factor(sex, levels = c("f", "m"), labels = c("Female", "Male"))
  )
```

::: {.callout-note}
## Data

`data(ais, package = "DAAG")`.  
*AIS* has **202** athletes (here *N* = **`r nrow(d)`** complete on `ferr`, `wt`, `sex`). See `?ais` for variable definitions (`rcc`, `wcc`, `hc`, `hg`, `ferr`, `bmi`, `ssf`, `pcBfat`, `lbm`, `ht`, `wt`, `sport`, …).

**Reference:** Maindonald & Braun, *Data Analysis and Graphics Using R* (`DAAG`).
:::

## Descriptive pattern

Men have higher mean **weight** and higher mean **ferritin** than women. Both conditions for **confounding** are visible: sex is associated with the predictor (`wt`) and with the outcome (`ferr`).

```{r}
#| label: summarise
d |>
  group_by(sex) |>
  summarise(
    n = n(),
    mean_wt_kg = mean(wt),
    sd_wt = sd(wt),
    mean_ferr = mean(ferr),
    sd_ferr = sd(ferr)
  ) |>
  knitr::kable(digits = 2, caption = "Body weight and plasma ferritin by sex (AIS)")
```

## Regression: bivariate vs adjusted

```{r}
#| label: models
mod_bi  <- lm(ferr ~ wt, data = d)
mod_adj <- lm(ferr ~ wt + sex, data = d)

bind_rows(
  tidy(mod_bi)  |> mutate(model = "Bivariate:  ferr ~"),
  tidy(mod_adj) |> mutate(model = "Adjusted:     ferr ~")
) |>
  filter(term %in% c("(Intercept)", "wt", "sexMale")) |>
  transmute(model, term, estimate, std.error, statistic, p.value) |>
  knitr::kable(digits = 3, caption = "Focus on the slope for `wt` (kg → ferritin)")
```

After **adjusting for sex**, the slope on `wt` is **small** and **not** statistically distinguishable from zero, while the **sex** term captures much of the between-group difference in ferritin.

```{r}
#| label: within-sex
#| tbl-cap: "Separate OLS fits within each sex (exploratory)."
bind_rows(
  tidy(lm(ferr ~ wt, data = filter(d, sex == "Female"))) |>
    mutate(sex = "Female"),
  tidy(lm(ferr ~ wt, data = filter(d, sex == "Male"))) |>
    mutate(sex = "Male")
) |>
  filter(term == "wt") |>
  select(sex, estimate, std.error, statistic, p.value) |>
  knitr::kable(digits = 3)
```

## Scatterplots: pooled vs stratified fits

- **Left:** one OLS line for everyone — the **strong positive** pooled `ferr ~ wt` slope.
- **Right:** points by sex; **solid** lines = OLS within each sex (**near-flat**); **dashed** = pooled line (same as left).

```{r}
#| label: fig-compare
#| fig-height: 5
#| fig-cap: "Left: naive pooled association. Right: within-sex slopes are shallow; dashed shows the misleading pooled line."
xlab <- "Body weight (kg)"
ylab <- "Plasma ferritin (ng/dL)"

p_pooled <- ggplot(d, aes(wt, ferr)) +
  geom_point(alpha = 0.35, colour = "grey35", size = 2.2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#6a040f", linewidth = 1) +
  labs(
    title = "Ignoring sex",
    subtitle = "ferr ~ wt — pooled sample",
    x = xlab, y = ylab
  )

pal <- c(Female = "#9b2226", Male = "#005f73")

p_strat <- ggplot(d, aes(wt, ferr, colour = sex)) +
  geom_point(alpha = 0.45, size = 2.2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1, aes(colour = sex)) +
  geom_smooth(
    method = "lm", se = FALSE,
    aes(group = 1), colour = "grey20",
    linetype = "dashed", linewidth = 0.75, inherit.aes = TRUE
  ) +
  scale_colour_manual(values = pal, name = NULL) +
  labs(
    title = "By sex",
    subtitle = "Solid = within-sex OLS; dashed = pooled line",
    x = xlab, y = ylab
  )

p_pooled + p_strat + plot_layout(widths = c(1, 1))
```

```{r}
#| label: fig-one-panel
#| fig-height: 4.8
#| fig-cap: "Slide-friendly: shallow within-sex fits vs steep pooled (dashed) line."
ggplot(d, aes(wt, ferr, colour = sex)) +
  geom_point(alpha = 0.45, size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1, aes(colour = sex)) +
  geom_smooth(
    method = "lm", se = FALSE,
    aes(group = 1), colour = "grey25",
    linetype = "dashed", linewidth = 0.9, inherit.aes = TRUE
  ) +
  scale_colour_manual(values = pal, name = "Sex") +
  labs(
    title = "Plasma ferritin vs body weight (AIS)",
    subtitle = "Dashed = fit ignoring sex; coloured = within-sex fits",
    x = xlab, y = ylab
  )
```

## Adjusted model: parallel slopes

`lm(ferr ~ wt + sex)` uses **one slope for `wt`** and **different intercepts** by sex.

```{r}
#| label: fig-adjusted-model
#| fig-height: 4.8
#| fig-cap: "Predictions from `lm(ferr ~ wt + sex)`: parallel lines (shallow slope for weight once sex is held constant)."
wt_seq <- seq(min(d$wt), max(d$wt), length.out = 80)
d_pred <- expand.grid(wt = wt_seq, sex = levels(d$sex), stringsAsFactors = FALSE)
d_pred$sex <- factor(d_pred$sex, levels = levels(d$sex))
d_pred$fit <- predict(mod_adj, newdata = d_pred)

ggplot(d, aes(wt, ferr, colour = sex)) +
  geom_point(alpha = 0.35, size = 2.2) +
  geom_line(data = d_pred, aes(y = fit, colour = sex), linewidth = 1) +
  scale_colour_manual(values = pal, name = "Sex") +
  labs(
    title = "Adjusted model (parallel slopes)",
    subtitle = "Lines = predictions from lm(ferr ~ wt + sex)",
    x = xlab, y = ylab
  )
```

## Other AIS pairs (same lesson, screened on `ais`)

These also show **strong pooled associations** driven largely by **sex**, with **much weaker** slopes after `+ sex` — useful for exercises or alternate examples (see `?ais` for units):

| Outcome `Y` | Predictor `X` | Pooled story (brief) |
|-------------|----------------|----------------------|
| `hg` (hemoglobin) | `ssf` (sum skinfolds) | “Leaner blood” vs adiposity — largely sex-composition |
| `hg` | `pcBfat` | Same idea with % body fat |
| `ferr` | `lbm` (lean body mass) | More lean mass ↔ more iron stores; men drive both |
| `lbm` | `hc` (hematocrit) | Strong pooled link, weakens with sex |

**Counterexample for class:** **`wt ~ ht`** stays strongly positive after `+ sex` — a reminder that **not every** predictor is badly confounded by sex.

## Take-home

1. **`lm(ferr ~ wt)`** can show a strong positive slope even when **`lm(ferr ~ wt + sex)`** does not support a within-sex weight effect on ferritin — because **sex** shifts both the *distribution* of `wt` and mean `ferr`.
2. Stratified plots (solid within-sex lines vs dashed pooled line) make **composition** visible.
3. These are **highly selected athletes**; use for **methods**, not population iron policy.

::: {.callout-important}
## Install

If needed: `install.packages("DAAG")` and `install.packages("patchwork")`.
:::
