---
title: "Confounding: reported energy intake vs DXA body fat % (NHANES)"
subtitle: "Bivariate vs multivariate regression with sex in the model"
author: "Research Statistics with R — Health Sciences"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    fig-cap-location: top
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
if (!requireNamespace("nhanesA", quietly = TRUE)) {
  install.packages("nhanesA", repos = "https://cloud.r-project.org")
}
library(nhanesA)
library(dplyr)
library(ggplot2)
library(broom)
library(patchwork)

theme_set(
  theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold"),
      plot.subtitle = element_text(colour = "grey35", size = rel(0.92)),
      legend.position = "top"
    )
)
```

## Question

**Does higher self-reported calorie intake predict higher total body fat percentage?**

NHANES provides **24-hour dietary recall** energy (`DR1TKCAL`, kcal, day 1) and **DXA** total body fat percent (`DXDTOPF`) on overlapping participants. **Sex** relates to both average intake and average adiposity. A **pooled** scatter can therefore show a **Simpson-type pattern**: the marginal association between intake and fat % may look **more negative** (or more “wrong”) than the **same predictor after adjusting for sex** in a linear model — even though the within-person measurement story is messy (single-day recall is noisy; cross-sections mix body size and behaviour).

This note mirrors the FAMuSS lesson: **one panel = fit ignoring sex**, **stratified panels = by sex**, plus an explicit **multivariate (parallel-lines)** picture for `pbf ~ kcal + sex`.

::: {.callout-warning}
## What to expect empirically
In recent cycles (e.g. 2017–2018), **adding sex usually moves the `kcal` coefficient a lot toward zero** (big confounding by sex), but **the adjusted slope on `kcal` may still be negative** because a one-day recall and unmeasured factors do not give a “clean positive dose–response” for adiposity. Use this document for **geometry of adjustment**, not for causal claims about diet.
:::

## Data (NHANES 2017–2018, cycle `J`)

Files (same letter suffix) merged on `SEQN`:

- **DEMO_J** — `RIAGENDR` (sex), `RIDAGEYR` (age)
- **DXX_J** — `DXAEXSTS` (DXA exam status), `DXDTOPF` (total % fat, whole-body DXA)
- **DR1TOT_J** — `DR1TKCAL` (day-1 total energy, kcal)

We keep **adults aged 20–59** (DXA % fat in these files targets ages 8–59; adults are the usual teaching slice), participants with a **completed whole-body DXA scan** (`DXAEXSTS` first level in the labelled factor: scan completed), and **plausible** day-1 recalls (`DR1TKCAL` between 500 and 8000 kcal).

```{r}
#| label: load-nhanes
demo <- nhanes("DEMO_J")
dxx  <- nhanes("DXX_J")
dr1  <- nhanes("DR1TOT_J")

d_raw <- demo |>
  select(SEQN, RIAGENDR, RIDAGEYR) |>
  inner_join(dxx |> select(SEQN, DXAEXSTS, DXDTOPF), by = "SEQN") |>
  inner_join(dr1 |> select(SEQN, DR1TKCAL), by = "SEQN") |>
  mutate(
    age = as.numeric(as.character(RIDAGEYR)),
    kcal = as.numeric(as.character(DR1TKCAL)),
    pbf = as.numeric(as.character(DXDTOPF)),
    sex = RIAGENDR
  )

d <- d_raw |>
  filter(
    !is.na(age), age >= 20, age <= 59,
    as.numeric(DXAEXSTS) == 1L,
    !is.na(kcal), !is.na(pbf),
    kcal >= 500, kcal <= 8000
  ) |>
  mutate(sex = factor(sex, levels = c("Female", "Male"))) |>
  select(SEQN, age, sex, kcal, pbf)
```

::: {.callout-note}
## Sample size
Complete cases in this document: *N* = **`r nrow(d)`** (Female **`r sum(d$sex == "Female")`**, Male **`r sum(d$sex == "Male")`**).
:::

## Descriptive pattern

```{r}
#| label: summarise
d |>
  group_by(sex) |>
  summarise(
    n = n(),
    mean_age = mean(age),
    mean_kcal = mean(kcal),
    sd_kcal = sd(kcal),
    mean_pbf = mean(pbf),
    sd_pbf = sd(pbf),
    .groups = "drop"
  ) |>
  knitr::kable(
    digits = 2,
    caption = "Age, reported kcal (day 1), and DXA total % fat by sex"
  )
```

## Regression: bivariate vs adjusted

```{r}
#| label: models
mod_bi  <- lm(pbf ~ kcal, data = d)
mod_adj <- lm(pbf ~ kcal + sex, data = d)

bind_rows(
  tidy(mod_bi)  |> mutate(model = "Bivariate:   pbf ~ kcal"),
  tidy(mod_adj) |> mutate(model = "Adjusted:      pbf ~ kcal + sex")
) |>
  filter(term %in% c("(Intercept)", "kcal", "sexMale")) |>
  select(model, term, estimate, std.error, statistic, p.value) |>
  knitr::kable(
    digits = 4,
    caption = "Coefficient tables — focus on `kcal` (change in fat % per 1 kcal)"
  )
```

Slopes in **percentage points per 1000 kcal** (easier to read):

```{r}
#| label: slopes-1000
tibble(
  model = c("Bivariate (pooled)", "Adjusted (+ sex)"),
  per_1000_kcal = c(coef(mod_bi)["kcal"], coef(mod_adj)["kcal"]) * 1000
) |>
  knitr::kable(digits = 3, caption = "Fat % change per +1000 kcal (linear approximation)")
```

**Within-sex** OLS slopes (exploratory; not the same estimand as the adjusted model unless assumptions line up):

```{r}
#| label: within-sex
d |>
  group_by(sex) |>
  group_map(~ broom::tidy(lm(pbf ~ kcal, data = .x)) |>
    filter(term == "kcal") |>
    mutate(sex = .y$sex, .before = 1)) |>
  bind_rows() |>
  transmute(
    sex,
    estimate_per_kcal = estimate,
    per_1000_kcal = estimate * 1000,
    p.value
  ) |>
  knitr::kable(digits = 4, caption = "Within-sex bivariate slopes on kcal")
```

## Figures

### Pooled vs stratified (same layout as FAMuSS)

```{r}
#| label: fig-pooled-strat
#| fig-height: 5
#| fig-cap: "Left: single OLS line (bivariate). Right: within-sex OLS (solid) and pooled line repeated (dashed)."
pal <- c(Female = "#9b2226", Male = "#005f73")
xlab <- "Day-1 reported energy intake (kcal)"
ylab <- "DXA total body fat (%)"

p_pooled <- ggplot(d, aes(kcal, pbf)) +
  geom_point(alpha = 0.32, colour = "grey35", size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#1b4332", linewidth = 1) +
  labs(
    title = "Ignoring sex",
    subtitle = "One line • pbf ~ kcal",
    x = xlab, y = ylab
  )

p_strat <- ggplot(d, aes(kcal, pbf, colour = sex)) +
  geom_point(alpha = 0.38, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1, aes(colour = sex)) +
  geom_smooth(
    method = "lm", se = FALSE,
    aes(group = 1),
    colour = "grey20",
    linetype = "dashed",
    linewidth = 0.75,
    inherit.aes = TRUE
  ) +
  scale_colour_manual(values = pal, name = NULL) +
  labs(
    title = "By sex",
    subtitle = "Solid = within-sex OLS; dashed = pooled line",
    x = xlab, y = ylab
  )

p_pooled + p_strat + plot_layout(widths = c(1, 1))
```

### Multivariate model: parallel lines (`pbf ~ kcal + sex`)

The **adjusted** model constrains one **shared** slope for `kcal` with **different intercepts** by sex (no interaction). Overlaying those fits shows what “holding sex constant” looks like on the scatter.

```{r}
#| label: fig-multivariate-parallel
#| fig-height: 4.8
#| fig-cap: "Fitted lines from the multivariate additive model pbf ~ kcal + sex (same slope for kcal, intercept shift by sex)."
kcal_grid <- seq(min(d$kcal), max(d$kcal), length.out = 150)
pred_line <- expand.grid(
  kcal = kcal_grid,
  sex = levels(d$sex),
  stringsAsFactors = FALSE
) |>
  mutate(sex = factor(sex, levels = levels(d$sex)))
pred_line$pbf_hat <- predict(mod_adj, newdata = pred_line)

ggplot(d, aes(kcal, pbf, colour = sex)) +
  geom_point(alpha = 0.38, size = 2) +
  geom_line(
    data = pred_line,
    aes(x = kcal, y = pbf_hat, colour = sex),
    linewidth = 1.05
  ) +
  scale_colour_manual(values = pal, name = "Sex") +
  labs(
    title = "Multivariate (additive): parallel slopes for kcal",
    subtitle = "Lines = predicted from lm(pbf ~ kcal + sex); one kcal slope + sex-specific intercepts",
    x = xlab, y = ylab
  )
```

### Slide-style single panel: stratified + pooled dashed

```{r}
#| label: fig-one-panel
#| fig-height: 4.8
#| fig-cap: "Stratified OLS lines with the pooled bivariate line (dashed) for comparison."
ggplot(d, aes(kcal, pbf, colour = sex)) +
  geom_point(alpha = 0.38, size = 2.2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1, aes(colour = sex)) +
  geom_smooth(
    method = "lm", se = FALSE,
    aes(group = 1),
    colour = "grey25",
    linetype = "dashed",
    linewidth = 0.9,
    inherit.aes = TRUE
  ) +
  scale_colour_manual(values = pal, name = "Sex") +
  labs(
    title = "DXA % fat vs reported intake",
    subtitle = "Dashed = fit ignoring sex; solid = within-sex bivariate fits",
    x = xlab, y = ylab
  )
```

## Take-home

1. **`Y ~ X` and `Y ~ X + Z` can tell very different stories** when `Z` is associated with both `X` and `Y`. Here `Z` is **sex**.
2. The **additive multivariate model** `pbf ~ kcal + sex` is drawn as **parallel lines** (same `kcal` slope, sex-specific intercept). Compare that geometry to the **dashed pooled** line from `pbf ~ kcal` alone.
3. For **inference** on NHANES you would normally account for the **survey design** (weights, PSUs); these plots use unweighted OLS for teaching visuals only.
4. Prefer **two-day averaging** of recalls (`DR1TKCAL`, `DR2TKCAL`) in serious analyses to reduce random day-to-day noise.

## Source

CDC/NCHS National Health and Nutrition Examination Survey. Example cycle: **NHANES 2017–2018**. Module documentation: search variable names `DR1TKCAL`, `DXDTOPF` on the NHANES website.
