---
title: "Confounding: height vs strength-training response (FAMuSS)"
subtitle: "Why `Y ~ height` misleads until you add `sex`"
author: "Research Statistics with R — Health Sciences"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    fig-cap-location: top
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(readr)
library(dplyr)
library(ggplot2)
library(broom)
library(patchwork)

theme_set(
  theme_minimal(base_size = 13) +
    theme(
      plot.title    = element_text(face = "bold"),
      plot.subtitle = element_text(colour = "grey35", size = rel(0.92)),
      legend.position = "top"
    )
)
```

## Question

In the **Functional ANS and Muscle Strength Study (FAMuSS)**, adults completed a standardised upper-body strength training programme. The outcome **`ndrm_change_pct`** is the percent change in non-dominant-arm strength. Does **baseline height** predict how much stronger people get?

A naïve regression might suggest it does. Once **sex** is in the model, that association largely disappears — a clean individual-level example of **confounding** by a variable that is related to both height and the outcome.

```{r}
#| label: load
d <- read_csv("datasets_staging/cleaned/19_famuss.csv", show_col_types = FALSE) |>
  mutate(sex = factor(sex, levels = c("female", "male"))) |>
  filter(!is.na(ndrm_change_pct), !is.na(height_in), !is.na(sex))
```

::: {.callout-note}
## Data
`datasets_staging/cleaned/19_famuss.csv` (also `oibiostat::famuss`). Complete cases in this document: *N* = **`r nrow(d)`**.
:::

## Descriptive pattern

Men are taller on average; they also show **smaller mean % strength gains** than women in this cohort. Height and “response to training” line up in the full sample partly because **sex** lines up with both.

```{r}
#| label: summarise
d |>
  group_by(sex) |>
  summarise(
    n = n(),
    mean_height_in = mean(height_in),
    sd_height_in = sd(height_in),
    mean_pct_gain = mean(ndrm_change_pct),
    sd_pct_gain = sd(ndrm_change_pct)
  ) |>
  knitr::kable(digits = 2, caption = "Height and % strength change by sex")
```

## Regression: bivariate vs adjusted

```{r}
#| label: models
mod_bi  <- lm(ndrm_change_pct ~ height_in, data = d)
mod_adj <- lm(ndrm_change_pct ~ height_in + sex, data = d)

bind_rows(
  tidy(mod_bi)  |> mutate(model = "Bivariate:  ndrm_change_pct ~"),
  tidy(mod_adj) |> mutate(model = "Adjusted:     ndrm_change_pct ~")
) |>
  filter(term %in% c("(Intercept)", "height_in", "sexmale")) |>
  transmute(
    model,
    term,
    estimate,
    std.error,
    statistic,
    p.value
  ) |>
  knitr::kable(digits = 3, caption = "Coefficient tables — focus on `height_in`")
```

The **pooled** slope on `height_in` is strongly negative; after **adjusting for sex**, the slope is small and no longer significant. Within each sex separately, height contributes little — the pooled effect was largely a **sex composition** artifact.

## Scatterplots: how the fitted slope changes

The two-panel figure is built to mirror how you teach this in class.

- **Left:** what you get if you ignore sex — one ordinary least-squares line through everyone (the **bivariate** model).
- **Right:** points coloured by sex; **solid** lines = separate OLS fits within each sex (what you *see* after you stratify); **dashed** line = the same pooled line as the left panel, for comparison on the same axes.

```{r}
#| label: fig-compare
#| fig-height: 5
#| fig-cap: "Left: a single slope for the full sample. Right: within-sex slopes (solid) vs the misleading pooled slope (dashed)."
xlab <- "Height (inches)"
ylab <- "Non-dominant arm % strength change"

p_pooled <- ggplot(d, aes(height_in, ndrm_change_pct)) +
  geom_point(alpha = 0.35, colour = "grey35", size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#1b4332", linewidth = 1) +
  labs(
    title = "Ignoring sex",
    subtitle = "One regression line • ndrm_change_pct ~ height_in",
    x = xlab, y = ylab
  )

pal <- c(female = "#9b2226", male = "#005f73")

p_strat <- ggplot(d, aes(height_in, ndrm_change_pct, colour = sex)) +
  geom_point(alpha = 0.4, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1, aes(colour = sex)) +
  geom_smooth(
    method = "lm", se = FALSE,
    aes(group = 1), colour = "grey20",
    linetype = "dashed", linewidth = 0.75, inherit.aes = TRUE
  ) +
  scale_colour_manual(values = pal, name = NULL) +
  labs(
    title = "By sex",
    subtitle = "Solid = within-sex OLS; dashed = pooled (misleading) line",
    x = xlab, y = ylab
  )

p_pooled + p_strat + plot_layout(widths = c(1, 1))
```

```{r}
#| label: fig-one-panel
#| fig-height: 4.8
#| fig-cap: "Single plot for slides: stratified fits with the pooled line overlaid as a dashed curve."
ggplot(d, aes(height_in, ndrm_change_pct, colour = sex)) +
  geom_point(alpha = 0.4, size = 2.2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1, aes(colour = sex)) +
  geom_smooth(
    method = "lm", se = FALSE,
    aes(group = 1), colour = "grey25",
    linetype = "dashed", linewidth = 0.9, inherit.aes = TRUE
  ) +
  scale_colour_manual(values = pal, name = "Sex") +
  labs(
    title = "Strength gain vs height",
    subtitle = "Dashed line = fit ignoring sex; coloured lines = fits within each sex",
     x = xlab, y = ylab
  )
```

## Take-home

1. A significant slope in **`Y ~ X`** does not mean **`X` causes `Y`**, or even that **`X` is a useful predictor of `Y`** once other major factors are included.
2. **Sex** is a textbook example of a variable associated with **height** and with many **performance / training** outcomes. Forgetting it can create a strong but spurious pooled association.
3. **Adjusting for sex** (`lm(Y ~ X + sex)`) estimates the slope of `X` **holding sex constant** — equivalent to combining within-sex information when the model is additive and parallel (one shared slope for `X` across sexes in the adjusted model; the exploratory within-sex lines help intuition).
