---
title: "Confounding, Multiple Regression, and Interaction"
subtitle: "Testosterone and body weight in NHANES adults"
author: "RSR Resources"
format:
  html:
    toc: false
    number-sections: false
    theme: cosmo
    fig-width: 9
    fig-height: 5.5
    css: |
      section.level1 {
        min-height: 92vh;
        padding: 2.5rem 0 3rem;
        border-bottom: 1px solid #dee2e6;
        display: flex;
        flex-direction: column;
        justify-content: center;
      }
      section.level1:last-of-type { border-bottom: none; }
      .quarto-title-block { margin-bottom: 0; }
      h1 { color: #006D77; font-weight: 700; }
      h2 { color: #006D77; font-size: 1.35rem; margin-top: 1.25rem; }
      .callout { margin-top: 1rem; }
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(broom)

nhanes <- read_csv(
  "datasets_staging/cleaned/openintro_nhanes_samp_adult_500.csv",
  show_col_types = FALSE
) |>
  filter(
    !is.na(Testosterone), !is.na(Weight), !is.na(Gender), !is.na(Age)
  ) |>
  mutate(
    log_testo = log(Testosterone),
    Gender = factor(Gender, levels = c("female", "male"))
  )

theme_rsr <- theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", color = "#006D77"),
    legend.position = "bottom"
  )

pal_sex <- c(female = "#E07A5F", male = "#006D77")
naive_col <- "#333333"
```

# The data

We ask whether **body weight** is associated with **testosterone** in US adults.

The OpenIntro NHANES sample has lab values for 500 adults. After dropping missing testosterone, weight, sex, or age, we keep **`r nrow(nhanes)`** complete cases.

::: {.callout-note}
Testosterone is right-skewed, so we model **`log(Testosterone)`** rather than the raw lab value.
:::

# What is confounding?

A **confounder** is a third variable that:

1. Is associated with the **exposure** (weight)
2. Is associated with the **outcome** (log testosterone)
3. Is **not** on the causal pathway from exposure to outcome

If we ignore it, the simple relationship we see can be **misleading**.

```{r confounding-diagram, echo=FALSE, fig.cap="Sex sits on back-door paths between weight and testosterone."}
#| fig-width: 8
#| fig-height: 3
confound_df <- data.frame(
  x = c(1, 2, 3),
  y = c(2, 2, 2),
  label = c("Weight", "Sex", "log(Testosterone)")
)

ggplot(confound_df, aes(x, y, label = label)) +
  geom_label(fill = "white", color = "#006D77", size = 5, label.size = 0.4) +
  annotate(
    "segment", x = 1.35, xend = 1.65, y = 2.15, yend = 2.35,
    arrow = arrow(length = unit(0.18, "cm")), linewidth = 0.6, color = "#555"
  ) +
  annotate(
    "segment", x = 2.35, xend = 2.65, y = 2.15, yend = 2.35,
    arrow = arrow(length = unit(0.18, "cm")), linewidth = 0.6, color = "#555"
  ) +
  annotate(
    "segment", x = 1.25, xend = 2.75, y = 1.85, yend = 1.85,
    arrow = arrow(length = unit(0.18, "cm")), linewidth = 0.6, color = "#888", linetype = "dashed"
  ) +
  annotate("text", x = 2, y = 1.55, label = "spurious pooled slope?", size = 4, color = "#888") +
  coord_equal(xlim = c(0.6, 3.4), ylim = c(1.2, 2.7)) +
  theme_void()
```

**Sex** is a classic confounder here: males are heavier on average and have much higher testosterone.

# First look — scatter

Before fitting anything, plot the outcome against the main predictor. Points only — no lines yet.

```{r scatter-raw, fig.cap="Higher weight goes with higher log testosterone mainly because males dominate the upper-left cloud."}
ggplot(nhanes, aes(Weight, log_testo)) +
  geom_point(alpha = 0.65, size = 2.2, color = "#555") +
  labs(x = "Weight (kg)", y = "log(Testosterone)") +
  theme_rsr
```

Does more weight go with higher or lower testosterone? The pooled cloud slopes upward.

# Naive model — weight only

```{r model-simple}
m_simple <- lm(log_testo ~ Weight, data = nhanes)
summary(m_simple)
```

```{r scatter-simple, fig.cap="The pooled slope is positive and highly significant."}
ggplot(nhanes, aes(Weight, log_testo)) +
  geom_point(alpha = 0.55, size = 2, color = "#555") +
  geom_smooth(method = "lm", se = TRUE, color = naive_col, linewidth = 1) +
  labs(
    x = "Weight (kg)",
    y = "log(Testosterone)",
    title = "Pooled regression ignores sex"
  ) +
  theme_rsr
```

Each extra **kg** is associated with about **`r round(coef(m_simple)["Weight"], 3)`** units on the log scale. That sounds like a real weight effect — but we have not accounted for sex.

# Same plot — colour reveals the pattern

```{r scatter-colour, fig.cap="Within each sex, weight and log testosterone are nearly flat; pooling groups creates an upward slope."}
ggplot(nhanes, aes(Weight, log_testo, color = Gender)) +
  geom_point(alpha = 0.55, size = 2) +
  geom_smooth(
    aes(group = 1),
    method = "lm", se = FALSE, color = naive_col, linewidth = 1
  ) +
  scale_color_manual(values = pal_sex) +
  labs(
    x = "Weight (kg)",
    y = "log(Testosterone)",
    color = "Sex",
    title = "The naive line still comes from the pooled model"
  ) +
  theme_rsr
```

::: {.callout-important}
## Simpson's paradox

The pooled trend says weight goes with higher testosterone. Within each sex, the trend is flat or even slightly negative. **Sex** connects to both weight and testosterone — a back-door path.
:::

# Confounding does not always flip the sign

A sign reversal is the dramatic version. Confounding can also **shrink**, **inflate**, or **hide** an association without reversing it.

The real question is always: **what comparison is this coefficient making?**

# Multiple linear regression — adjust for sex

Add sex on the right-hand side. Same weight slope in every group, different intercepts — **parallel lines**.

$$
\log(\text{Testosterone}) = \beta_0 + \beta_1\,\text{Weight} + \beta_2\,\text{Male}
$$

```{r model-additive}
m_add <- lm(log_testo ~ Weight + Gender, data = nhanes)
tidy(m_add, conf.int = TRUE) |>
  mutate(across(where(is.numeric), \(x) round(x, 4)))
```

```{r scatter-additive, fig.cap="Parallel lines: one weight slope, shifted intercepts by sex."}
pred_add <- expand.grid(
  Weight = seq(min(nhanes$Weight), max(nhanes$Weight), length.out = 100),
  Gender = levels(nhanes$Gender)
)
pred_add$fitted <- predict(m_add, newdata = pred_add)

ggplot(nhanes, aes(Weight, log_testo, color = Gender)) +
  geom_point(alpha = 0.55, size = 2) +
  geom_line(data = pred_add, aes(y = fitted), linewidth = 1) +
  scale_color_manual(values = pal_sex) +
  labs(
    x = "Weight (kg)",
    y = "log(Testosterone)",
    color = "Sex",
    title = "Additive model: Weight + Sex"
  ) +
  theme_rsr
```

After adjustment, the weight coefficient is near **zero** (p = `r format.pval(summary(m_add)$coefficients["Weight", "Pr(>|t|)"], digits = 2, eps = 0.001)`). Sex carries most of the difference in testosterone levels.

# Coefficients at a glance

```{r coef-table}
bind_rows(
  naive    = tidy(m_simple, conf.int = TRUE),
  adjusted = tidy(m_add, conf.int = TRUE),
  .id = "model"
) |>
  filter(term %in% c("Weight", "(Intercept)")) |>
  mutate(across(where(is.numeric), \(x) round(x, 4)))
```

The weight slope shrinks from **`r round(coef(m_simple)["Weight"], 3)`** to **`r round(coef(m_add)["Weight"], 3)`** once we hold sex constant.

# Add another predictor — age (additive only)

Real analyses often adjust for more than one factor. **Age** also relates to weight and to hormones, so we add it **additively** alongside sex:

$$
\log(\text{Testosterone}) = \beta_0 + \beta_1\,\text{Weight} + \beta_2\,\text{Male} + \beta_3\,\text{Age}
$$

```{r model-age}
m_age <- lm(log_testo ~ Weight + Gender + Age, data = nhanes)
tidy(m_age, conf.int = TRUE) |>
  mutate(across(where(is.numeric), \(x) round(x, 4)))
```

Each term is a **separate adjustment**: the weight coefficient is now the association between weight and log testosterone **among people of the same sex and age**.

We do **not** add `Weight * Age` here. Age is a confounder we want to control, not a group where we expect the weight slope to change.

::: {.callout-tip}
## When to use `+` (addition)

Use **`+`** when a variable is a **confounder** or **precision covariate** and you are happy assuming the weight effect is the **same** across its levels. That gives parallel adjustment: one slope for weight, intercept shifts for sex, a linear shift per year of age.
:::

# When a common weight slope is too strict

The additive model **forces the same weight slope** for males and females. That may be wrong if biology makes the weight–testosterone link **different by sex**.

Use an **interaction** instead of only adding sex:

$$
\log(\text{Testosterone}) = \beta_0 + \beta_1\,\text{Weight} + \beta_2\,\text{Male} + \beta_3\,(\text{Weight} \times \text{Male})
$$

In R: `Weight * Gender` expands to `Weight + Gender + Weight:Gender`.

```{r model-interaction}
m_int <- lm(log_testo ~ Weight * Gender, data = nhanes)
tidy(m_int, conf.int = TRUE) |>
  mutate(across(where(is.numeric), \(x) round(x, 4)))
```

The **`Weight:Gendermale`** row tests whether the weight slope **differs** between sexes (p = `r format.pval(summary(m_int)$coefficients["Weight:Gendermale", "Pr(>|t|)"], digits = 2, eps = 0.001)`).

```{r scatter-interaction, fig.cap="Interaction model: separate weight slopes for males and females."}
pred_int <- expand.grid(
  Weight = seq(min(nhanes$Weight), max(nhanes$Weight), length.out = 100),
  Gender = levels(nhanes$Gender)
)
pred_int$fitted <- predict(m_int, newdata = pred_int)

ggplot(nhanes, aes(Weight, log_testo, color = Gender)) +
  geom_point(alpha = 0.55, size = 2) +
  geom_line(data = pred_int, aes(y = fitted), linewidth = 1) +
  scale_color_manual(values = pal_sex) +
  labs(
    x = "Weight (kg)",
    y = "log(Testosterone)",
    color = "Sex",
    title = "Interaction model: Weight * Sex"
  ) +
  theme_rsr
```

::: {.callout-tip}
## Rules of thumb: `+` versus `*`

| Situation | Typical choice |
|-----------|----------------|
| Variable is a confounder; same exposure effect in all groups | **`+`** (e.g. `Weight + Gender + Age`) |
| You have a scientific reason the exposure effect **differs by group** | **`*`** (e.g. `Weight * Gender`) |
| Stratified plots show **parallel** lines | Prefer **`+`** |
| Stratified plots show **clearly different slopes** | Consider **`*`** |
| Interaction p-value is large and lines look parallel | **`+`** is often enough |
| Many predictors; interaction not pre-specified | Start with **`+`**; add `*` only when justified |

**Addition** answers: "What is the weight association **holding** sex (and age) constant?" **Multiplication** answers: "Does the weight association **depend on** sex?"
:::

# Compare weight slopes across models

```{r coef-compare, fig.cap="The pooled weight effect shrinks after adjustment; interaction allows different slopes by sex."}
coef_df <- bind_rows(
  tidy(m_simple, conf.int = TRUE) |>
    filter(term == "Weight") |>
    mutate(model = "1. Pooled"),
  tidy(m_add, conf.int = TRUE) |>
    filter(term == "Weight") |>
    mutate(model = "2. + Sex"),
  tidy(m_age, conf.int = TRUE) |>
    filter(term == "Weight") |>
    mutate(model = "3. + Sex + Age"),
  tidy(m_int, conf.int = TRUE) |>
    filter(term == "Weight") |>
    mutate(model = "4. Interaction (female slope)")
) |>
  mutate(model = factor(model, levels = unique(model)))

ggplot(coef_df, aes(estimate, model, xmin = conf.low, xmax = conf.high)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#888") +
  geom_errorbarh(height = 0.15, linewidth = 0.8, color = "#006D77") +
  geom_point(size = 3, color = "#006D77") +
  labs(
    x = "Weight slope on log(Testosterone) scale",
    y = NULL,
    title = "Weight slopes across models"
  ) +
  theme_rsr
```

# Takeaways

1. **Confounding** can make a pooled association look real when it is mostly compositional (here, mixing males and females).
2. **Multiple linear regression** changes the comparison: from pooled association to association **holding selected variables constant**.
3. Add confounders with **`+`** when you want shared slopes; use **`*`** when the exposure effect may **differ by group**.
4. Always **plot by group** before trusting one-number summaries.
5. Causal language still needs more than adjustment — but the right controls clarify **which** association you are estimating.

::: {.callout-note}
Data: OpenIntro NHANES adult sample (`openintro_nhanes_samp_adult_500.csv`), CC-BY-SA 4.0; underlying NHANES is US-government public domain.
:::
