---
title: "Regression tables with the PREVEND data"
output: html_document
---

```{r setup, include=FALSE}
library(openintro)
library(dplyr)
library(modelsummary)
library(broom)
library(ggplot2)
library(gt)

knitr::opts_chunk$set(message = FALSE, warning = FALSE)

prevend <- openintro::prevend |>
  mutate(
    Statin = factor(Statin, 0:1, c("No statin", "Statin")),
    Gender = factor(Gender, 0:1, c("Men", "Women")),
    Education = factor(
      Education,
      0:3,
      c("Primary school", "Lower secondary", "Higher secondary", "University")
    )
  )

models <- list(
  "Crude" = lm(RFFT ~ Statin, data = prevend),
  "Adjusted" = lm(RFFT ~ Statin + Age + Gender + Education, data = prevend)
)

coef_names <- c(
  "StatinStatin" = "Statin use: yes",
  "Age" = "Age, years",
  "GenderWomen" = "Gender: women",
  "EducationLower secondary" = "Education: lower secondary",
  "EducationHigher secondary" = "Education: higher secondary",
  "EducationUniversity" = "Education: university"
)

paper_table <- function(x) {
  x |>
    tab_options(
      table.font.names = "Times New Roman",
      table.font.size = px(13),
      heading.title.font.size = px(15),
      column_labels.font.weight = "bold",
      table.border.top.width = px(2),
      table.border.bottom.width = px(2)
    )
}

category_means <- bind_rows(
  prevend |> summarize(term = paste("Mean RFFT:", Statin), value = sprintf("%.1f (n = %s)", mean(RFFT), n()), .by = Statin),
  prevend |> summarize(term = paste("Mean RFFT:", Gender), value = sprintf("%.1f (n = %s)", mean(RFFT), n()), .by = Gender),
  prevend |> summarize(term = paste("Mean RFFT:", Education), value = sprintf("%.1f (n = %s)", mean(RFFT), n()), .by = Education)
) |>
  transmute(term, Crude = value, Adjusted = value)
```

## 1. Standard regression table

```{r}
modelsummary(
  models,
  coef_map = coef_names,
  estimate = "{estimate}{stars}",
  statistic = "({std.error})",
  gof_map = c("nobs", "r.squared"),
  output = "gt"
) |>
  tab_header(title = "OLS regression of RFFT score") |>
  tab_source_note("Coefficients are changes in RFFT score. Standard errors are shown in parentheses.") |>
  tab_source_note("Reference groups: no statin use, men, and primary school education.") |>
  paper_table()
```

**Table 1. Standard regression table.** The adjusted model estimates the association between statin use and RFFT score after controlling for age, gender, and education.

## 2. Regression table with category descriptives

```{r}
modelsummary(
  models,
  coef_map = coef_names,
  estimate = "{estimate}{stars}",
  statistic = "({std.error})",
  gof_map = c("nobs", "r.squared"),
  add_rows = category_means,
  output = "gt"
) |>
  tab_header(title = "OLS regression with outcome means by category") |>
  tab_source_note("Added rows show the mean RFFT score within each category, with group size in parentheses.") |>
  tab_source_note("Reference groups: no statin use, men, and primary school education.") |>
  paper_table()
```

**Table 2. Regression table with descriptives.** The lower panel gives the mean RFFT score for each statin, gender, and education category used in the regression model.

## 3. Visual regression display

```{r, fig.width=7, fig.height=4}
tidy(models$Adjusted, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(
    term = recode(term, !!!coef_names),
    term = factor(term, rev(coef_names)),
    label = sprintf("%.1f [%.1f, %.1f]", estimate, conf.low, conf.high)
  ) |>
  ggplot(aes(x = estimate, y = term)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), width = 0.15) +
  geom_point(size = 2.4) +
  geom_text(aes(label = label), hjust = 0, nudge_x = 1.5, size = 3.4) +
  scale_x_continuous(limits = c(-8, 26)) +
  labs(x = "Change in RFFT score", y = NULL) +
  theme_classic(base_family = "Times New Roman")
```

**Figure 1. Regression estimates with confidence intervals.** Dots show adjusted regression coefficients and horizontal lines show 95% confidence intervals; the numeric estimate and interval are printed inside the figure.
