---
title: "Psychological quality of life in Brazilian medical students"
subtitle: "Dataset 7: Resilience, learning environment and quality of life in Brazilian medical students"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "medstudent_qol.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "whoqol_psychological"
predictors <- c("sex", "year_group", "age_yrs", "resilience_score",
                "dreem_global", "state_anxiety", "trait_anxiety")

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex = factor(sex, levels = c("female", "male"),
                 labels = c("Female", "Male")),
    year_group = factor(year_group,
                        levels = c("basic_sciences", "clinical_sciences", "clerkship"),
                        labels = c("Basic sciences", "Clinical sciences", "Clerkship")),
    school_legal_status = factor(school_legal_status, levels = c("public", "private"),
                                 labels = c("Public", "Private")),
    school_location = factor(school_location, levels = c("state_capital", "other_city"),
                             labels = c("State capital", "Other city"))
  ) |>
  labelled::set_variable_labels(
    whoqol_psychological = "WHOQOL psychological domain (0-100)",
    sex                  = "Sex",
    year_group           = "Stage of course",
    age_yrs              = "Age (years)",
    resilience_score     = "Resilience score (16-98)",
    dreem_global         = "DREEM global score (0-200)",
    state_anxiety        = "State anxiety (20-80)",
    trait_anxiety        = "Trait anxiety (20-80)"
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)
```

# Introduction

Tempski and colleagues surveyed a random sample of 1,350 students at 22 Brazilian
medical schools in 2011-2012, measuring quality of life (WHOQOL-BREF), the
perceived educational environment (DREEM), resilience, depressive symptoms and
anxiety in a single cross-sectional questionnaire (Tempski et al., *PLOS ONE*
2015; doi:10.1371/journal.pone.0131535). The psychological domain of the
WHOQOL-BREF - positive and negative feelings, self-esteem, concentration, body
image and meaning in life - is the domain most consistently depressed in medical
trainees, and was the paper's primary quality-of-life outcome.

**Aim.** Among Brazilian medical students, we examined which of sex, stage of the
course, age, resilience, the perceived educational environment, state anxiety and
trait anxiety are independently associated with psychological quality of life,
mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** Multicentre cross-sectional survey of a random sample of 1,350
students drawn from 22 medical schools across Brazil, stratified by stage of the
course (basic sciences, clinical sciences, clerkship). All variables used here
are complete.

**Variables.** The outcome is `whoqol_psychological`, the WHOQOL-BREF
psychological domain transformed to a 0-100 scale where higher is better
quality of life. Predictors are sex (reference: female, the larger group),
stage of the course (reference: basic sciences, the earliest stage), age in
years, the resilience scale score (16-98, higher = more resilient),
the DREEM global score (sum of the five DREEM subscales, higher = a more
positive perceived educational environment), and the State-Trait Anxiety
Inventory state and trait scales (each 20-80, higher = more anxious). All
continuous predictors are modelled per one-unit increase.

Two deliberate choices about construct overlap. First, we used `dreem_global`
and **not** the five DREEM subscales, because the global score is their exact
sum and putting both in one model is a perfect linear dependency. Second, we did
**not** use any other WHOQOL domain as a predictor: physical, social and
environment are subscales of the same instrument as the outcome, and a model of
one domain on another describes the instrument's internal structure rather than
anything about students. State and trait anxiety are retained together because
they are conceptually distinct (anxiety now versus anxiety as a disposition) and
the paper treats them separately, but they correlate at *r* = 0.70 and their
variance inflation factors are reported below.

**Statistical analysis.** Linear regression (`lm()`), justified because the
outcome is continuous and roughly symmetric. We fitted a univariable model for
each predictor first, then a single multivariable model containing all of them.
Predictors were chosen in advance on subject-matter grounds, not by significance.
Significance threshold 0.05; estimates are reported with 95% confidence
intervals. Assumptions were checked with `performance::check_model()`. Analysis
used `r R.version.string` with tidyverse, gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped).

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(
    missing = "no"
  ) |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

Students were young (median 22 years), slightly more often female, and roughly
evenly spread across the three stages of the course. The median psychological
quality-of-life score was 62.5 out of 100, well below the physical domain median
in the same students, and median trait anxiety (45) was above the state anxiety
median (43).

## Exploratory figure

```{r explore, fig.cap="Figure 1. Psychological quality of life against trait anxiety, by stage of the course. Points are individual students; lines are linear smoothers."}
ggplot(dat_model, aes(x = trait_anxiety, y = whoqol_psychological,
                      colour = year_group)) +
  geom_point(alpha = 0.3, size = 1.2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.9) +
  scale_colour_manual(values = c("#3B7EA1", "#D1741F", "#5B8C5A")) +
  labs(x = "Trait anxiety (20-80)",
       y = "WHOQOL psychological domain (0-100)",
       colour = "Stage of course") +
  theme(legend.position = "top")
```

Psychological quality of life falls steeply and near-linearly as trait anxiety
rises, and the three stages of the course lie almost on top of one another,
suggesting the stage of the course matters little once anxiety is accounted for.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    hide_n = TRUE
  ) |>
  modify_caption("**Univariable linear regression.**")
uv
```

```{r uv-numbers}
uv_trait <- lm(whoqol_psychological ~ trait_anxiety, data = dat_model) |>
  tidy(conf.int = TRUE) |> filter(term == "trait_anxiety")
uv_res <- lm(whoqol_psychological ~ resilience_score, data = dat_model) |>
  tidy(conf.int = TRUE) |> filter(term == "resilience_score")
```

Trait anxiety is the strongest crude predictor: each one-point increase on the
trait anxiety scale is associated with a
`r sprintf("%.2f", abs(uv_trait$estimate))`-point *lower* WHOQOL psychological
score (95% CI `r sprintf("%.2f to %.2f", uv_trait$conf.low, uv_trait$conf.high)`,
p < 0.001). Over the observed trait-anxiety range of 20 to 80 that is roughly
`r sprintf("%.0f", abs(uv_trait$estimate) * 60)` points of the 0-100 scale, an
enormous crude association. Resilience runs the other way
(`r sprintf("%.2f", uv_res$estimate)` points per point of resilience,
95% CI `r sprintf("%.2f to %.2f", uv_res$conf.low, uv_res$conf.high)`), and the
DREEM global score is positively associated with the outcome. Age and sex show
no crude association, and the stage of the course shows only a small one.

## Why adjustment is needed

The crude estimates cannot be read as independent effects because the
psychosocial predictors are entangled with each other. **Trait anxiety** is a
confounder of the resilience association: dispositionally anxious students score
lower on the resilience scale (*r* = -0.61) and also report worse psychological
quality of life, so part of the crude resilience effect is really trait anxiety
in disguise. **State anxiety** plays the same role for trait anxiety - the two
scales share 49% of their variance - so the crude trait coefficient absorbs
situational anxiety as well. The **DREEM global score** is a third: students who
rate their school environment poorly are both more anxious and report worse
quality of life, so the crude anxiety coefficients partly reflect a bad learning
environment rather than the student's temperament. Finally **age** and **stage of
the course** move together almost deterministically (clerkship students are older
by construction), so any crude age effect must be read as partly a
stage-of-course effect. A single model containing all seven predictors lets each
estimate be interpreted holding the others fixed.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit) |>
  modify_caption("**Multivariable linear regression.**")
mv
```

```{r fit-stats}
gl <- glance(fit)
tibble(
  Statistic = c("R-squared", "Adjusted R-squared", "Residual SE", "n", "Model df"),
  Value = c(sprintf("%.3f", gl$r.squared), sprintf("%.3f", gl$adj.r.squared),
            sprintf("%.2f", gl$sigma), as.character(gl$nobs), as.character(gl$df))
) |> gt() |> tab_header(title = md("**Model fit**"))
```

```{r mv-numbers}
tid <- tidy(fit, conf.int = TRUE)
g <- function(x) tid |> filter(term == x)
b_trait <- g("trait_anxiety"); b_state <- g("state_anxiety")
b_res <- g("resilience_score"); b_dreem <- g("dreem_global"); b_age <- g("age_yrs")
fmt <- function(r) sprintf("%.2f (95%% CI %.2f to %.2f)", r$estimate, r$conf.low, r$conf.high)
```

The model explains `r sprintf("%.1f%%", gl$r.squared * 100)` of the variance in
psychological quality of life (adjusted R-squared
`r sprintf("%.3f", gl$adj.r.squared)`), which is very high for cross-sectional
questionnaire data and is a warning in itself (see Discussion). Holding the other
six predictors fixed, each additional point of trait anxiety is associated with a
`r fmt(b_trait)`-point change in the WHOQOL psychological score, and each point of
state anxiety with `r fmt(b_state)`. Resilience remains independently associated
in the expected direction (`r fmt(b_res)` points per point), as does the DREEM
global score (`r fmt(b_dreem)` points per point, so roughly
`r sprintf("%.1f", b_dreem$estimate * 20)` points across a 20-point difference in
perceived environment). Age is associated with a small decrement
(`r fmt(b_age)` points per year) once stage of the course is held fixed, while
sex and stage of the course show no independent association. These are
conditional associations in a cross-sectional survey, not effects of
intervening on anxiety or resilience.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

```{r compare-numbers}
uv_state <- lm(whoqol_psychological ~ state_anxiety, data = dat_model) |>
  tidy(conf.int = TRUE) |> filter(term == "state_anxiety")
uv_dreem <- lm(whoqol_psychological ~ dreem_global, data = dat_model) |>
  tidy(conf.int = TRUE) |> filter(term == "dreem_global")
uv_age <- lm(whoqol_psychological ~ age_yrs, data = dat_model) |>
  tidy(conf.int = TRUE) |> filter(term == "age_yrs")
uv_sex <- lm(whoqol_psychological ~ sex, data = dat_model) |>
  tidy(conf.int = TRUE) |> filter(term == "sexMale")
b_sex <- g("sexMale")
uv_yr <- lm(whoqol_psychological ~ year_group, data = dat_model) |>
  tidy(conf.int = TRUE) |> filter(term == "year_groupClerkship")
b_yr <- g("year_groupClerkship")
```

Almost every estimate shrinks toward the null, and the size of the shrinkage
ranks the predictors. **Trait anxiety** retains the most, falling from
`r sprintf("%.2f", uv_trait$estimate)` crude to
`r sprintf("%.2f", b_trait$estimate)` adjusted
(`r sprintf("%.0f%%", 100 * abs(b_trait$estimate / uv_trait$estimate))` of its
crude size) - the signature of the dominant predictor in the set. **Resilience**
keeps only `r sprintf("%.0f%%", 100 * b_res$estimate / uv_res$estimate)`
(`r sprintf("%.2f", uv_res$estimate)` to `r sprintf("%.2f", b_res$estimate)`),
because much of what the resilience scale captures is simply low trait anxiety.
**State anxiety** keeps
`r sprintf("%.0f%%", 100 * b_state$estimate / uv_state$estimate)`
(`r sprintf("%.2f", uv_state$estimate)` to `r sprintf("%.2f", b_state$estimate)`)
for the same reason: it and trait anxiety are largely measuring one
negative-affect disposition. The **DREEM global score** collapses hardest, from
`r sprintf("%.3f", uv_dreem$estimate)` to `r sprintf("%.3f", b_dreem$estimate)`
per point (`r sprintf("%.0f%%", 100 * b_dreem$estimate / uv_dreem$estimate)`
retained); its crude association was mostly a proxy for the anxiety and
resilience of the student doing the rating.

Two coefficients change significance, in opposite directions. **Sex** is a strong
crude predictor - men score `r sprintf("%.2f", uv_sex$estimate)` points higher,
p < 0.001 - but the adjusted difference is `r sprintf("%.2f", b_sex$estimate)`
points and non-significant (p = `r sprintf("%.2f", b_sex$p.value)`): the crude
male advantage in psychological quality of life is entirely explained by women in
this sample scoring higher on both anxiety scales. **Age** moves the other way -
crude `r sprintf("%.2f", uv_age$estimate)` and non-significant
(p = `r sprintf("%.2f", uv_age$p.value)`), adjusted
`r sprintf("%.2f", b_age$estimate)` and significant
(p = `r sprintf("%.3f", b_age$p.value)`) - a suppression effect that appears once
stage of the course is held fixed, because within a given stage the older
students are those who entered late or are repeating. **Stage of the course**
barely moves and stays non-significant throughout (clerkship
`r sprintf("%.2f", uv_yr$estimate)` crude versus
`r sprintf("%.2f", b_yr$estimate)` adjusted).

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r vif-table}
performance::check_collinearity(fit) |>
  as_tibble() |>
  transmute(Term, VIF = round(VIF, 2), `Tolerance` = round(Tolerance, 3)) |>
  gt() |> tab_header(title = md("**Variance inflation factors**"))
```

```{r het-test}
het <- performance::check_heteroscedasticity(fit)
```

**Linearity.** The fitted-versus-residual panel is flat across most of the range
with a mild upward bend at the lowest fitted values: students predicted to have
the worst psychological quality of life score slightly better than the model
expects, because the WHOQOL scale is floored at 0.

**Homoscedasticity.** Breusch-Pagan rejects constant variance
(p = `r sprintf("%.3f", attr(het, "p"))`); residual spread narrows at high fitted
values, again a scale-ceiling artefact. Because of this we refitted with
heteroscedasticity-consistent (HC3) standard errors as a sensitivity check:

```{r robust}
rob <- lmtest::coeftest(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
broom::tidy(rob) |>
  transmute(term,
            estimate = round(estimate, 3),
            `robust SE` = round(std.error, 3),
            statistic = round(statistic, 2),
            p.value = format.pval(p.value, digits = 3, eps = 0.001)) |>
  gt() |> tab_header(title = md("**HC3 robust standard errors (sensitivity check)**"))
```

The robust standard errors are within a few percent of the model-based ones and
no conclusion changes, so the heteroscedasticity is not consequential here.

**Normality of residuals.** The Q-Q plot is close to the diagonal with a short
left tail; the formal test flags non-normality
(p = `r sprintf("%.3f", attr(performance::check_normality(fit), "p"))`), but with
n = `r nrow(dat_model)` this is a power artefact rather than a threat, and the
central limit theorem covers the coefficient standard errors.

**Influential points.** No observation exceeds the Cook's distance threshold; the
largest Cook's distance is
`r sprintf("%.3f", max(cooks.distance(fit)))`, far below 0.5. No single student
drives the results.

**Collinearity.** The highest VIF is
`r sprintf("%.2f", max(performance::check_collinearity(fit)$VIF))`
(trait anxiety), with state anxiety at
`r sprintf("%.2f", performance::check_collinearity(fit) |> as_tibble() |> filter(Term == "state_anxiety") |> pull(VIF))`.
Both sit in the "moderate" band and well below the conventional threshold of 5,
so the two anxiety scales can be carried in the same model - but their standard
errors are roughly 1.5 times what they would be if the scales were independent,
and the split of the joint anxiety effect between them should not be
over-interpreted.

# Discussion

Among Brazilian medical students, psychological quality of life is dominated by
the student's own affective disposition. Trait anxiety alone accounts for most of
the explained variance, and state anxiety, resilience and the perceived
educational environment each add a smaller independent contribution. The
educational-environment association survives adjustment but is modest: a 20-point
better DREEM score corresponds to about
`r sprintf("%.1f", b_dreem$estimate * 20)` points of WHOQOL psychological score,
roughly a fifth of the effect of a 20-point difference in trait anxiety.
Practically, this says that school-level improvements to the learning environment
are worth pursuing but will not by themselves move a highly anxious student's
psychological wellbeing very far.

**Limitations.** *Construct overlap is the central problem.* The WHOQOL
psychological domain contains items on negative feelings and self-esteem, and the
trait anxiety scale contains items on the same feelings; an R-squared of
`r sprintf("%.2f", gl$r.squared)` reflects partly shared item content, not only a
substantive relationship. The trait-anxiety coefficient should be read as
"students who describe themselves as anxious also describe themselves as having
poor psychological wellbeing", which is close to a tautology. *Clustering is not
modelled.* Students were sampled from 22 medical schools, but the released file
carries no school identifier, only `school_legal_status` and `school_location`;
students within a school share teachers, curriculum and city, so the true
standard errors are wider than those reported and no random intercept can be
fitted. *Cross-sectional design.* Anxiety, resilience and quality of life were
measured on the same day by the same instrument; the direction of the arrow
between them cannot be established, and reverse causation (poor quality of life
raising anxiety) is at least as plausible as the direction implied by the model.
*Common-method bias.* Every variable except sex, age and the two school
characteristics is self-reported on one questionnaire, so a student's general
negativity inflates all associations at once. *No missing data* is a
genuine strength here - all `r nrow(dat)` rows were analysed.

**Future work.** A longitudinal design that measures trait anxiety at entry and
quality of life a year later, with school identifiers retained so that
between-school variation can be separated from between-student variation.

# Feasibility notes

*Not part of a student report.*

- **Knits cleanly, no numerical trouble.** n = 1,350 complete cases, seven
  predictors, no separation, no sparse cells, no influential points
  (max Cook's D `r sprintf("%.3f", max(cooks.distance(fit)))`).
- **The real hazard is construct overlap, not collinearity.** VIFs are all
  under `r sprintf("%.1f", max(performance::check_collinearity(fit)$VIF))`, so
  the model is statistically stable. But `trait_anxiety` correlates with the
  outcome at *r* = -0.75 and `bdi_depression` at *r* = -0.74; a student who puts
  BDI in this model will get R-squared near 0.7 and will be, in effect,
  regressing one self-reported mood measure on another. `bdi_depression` was
  deliberately excluded for this reason. Students should be told explicitly that
  a high R-squared here is a bad sign, not a good one.
- **Do not allow other WHOQOL domains as predictors.** `whoqol_physical`
  correlates 0.67 and `whoqol_social` 0.57 with this outcome; they are subscales
  of the same instrument and a model containing them is uninterpretable.
- **`dreem_global` and the five DREEM subscales must never appear together** -
  the global score is their exact arithmetic sum.
- **Heteroscedasticity is flagged** (BP p = `r sprintf("%.3f", attr(het, "p"))`)
  in this and every other outcome in this dataset; it is a ceiling/floor artefact
  of bounded scales and HC3 robust SEs change nothing. Worth one sentence in a
  student report, not a re-analysis.
- **Verdict: workable, and the best-powered of the six outcomes**, but only if
  the predictor set is policed. Left unpoliced it is the outcome most likely to
  produce a spuriously impressive model.
