---
title: "Asymptomatic malaria parasitaemia in household members of confirmed malaria patients"
subtitle: "Dataset 13: Asymptomatic malaria in households of confirmed cases in Abuja"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "nigeria_malaria_households.csv"),
                    show_col_types = FALSE)

lab <- function(x, l) { attr(x, "label") <- l; x }

# --------------------------------------------------------------------------
# Derived variables. Three of the fourteen shipped predictors are reshaped
# before modelling; the reasoning is set out in Methods.
#   1. age_group    6 levels -> 4 (the 20-24 stratum holds only 50 people)
#   2. owns_llin + used_llin_previous_night -> one 3-level net variable, because
#      the two are exactly nested (you cannot use a net you do not own)
#   3. the five house_* variables -> one 3-level housing-quality index
# --------------------------------------------------------------------------
dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest group, or the "unexposed" one.
    age_grp = lab(factor(
      case_when(age_group == "under_5"  ~ "Under 5",
                age_group == "5_to_9"   ~ "5-9",
                age_group == "10_to_19" ~ "10-19",
                TRUE                    ~ "20 and over"),
      levels = c("20 and over", "10-19", "5-9", "Under 5")),
      "Age group (years)"),

    sex = lab(factor(sex, levels = c("female", "male"), labels = c("Female", "Male")),
              "Sex"),

    educ = lab(factor(education, levels = c("none", "primary", "secondary", "tertiary"),
                      labels = c("None", "Primary", "Secondary", "Tertiary")),
               "Highest education level"),

    council = lab(factor(area_council,
                         levels = c("abuja_municipal", "kuje", "kwali"),
                         labels = c("Abuja Municipal", "Kuje", "Kwali")),
                  "Area council"),

    bushes = lab(factor(bushes_near_home, levels = c(0, 1), labels = c("No", "Yes")),
                 "Bushes around the home"),

    llin = lab(factor(
      case_when(owns_llin == 0                 ~ "No net",
                used_llin_previous_night == 0  ~ "Owns, not used last night",
                TRUE                           ~ "Owns, used last night"),
      levels = c("No net", "Owns, not used last night", "Owns, used last night")),
      "Insecticidal net"),

    # Housing-quality index: one point for each of five "improved" features.
    house_score = as.integer(house_floor == "ceramictiles") +
                  as.integer(house_wall  == "cement_plastered") +
                  as.integer(house_roof  %in% c("metal", "roofshingles")) +
                  as.integer(house_window == "screenorglasswindows") +
                  as.integer(house_type  != "hut"),

    housing = lab(factor(
      cut(house_score, breaks = c(-1, 2, 4, 5),
          labels = c("Basic (0-2 features)", "Intermediate (3-4)", "Improved (all 5)")),
      levels = c("Improved (all 5)", "Intermediate (3-4)", "Basic (0-2 features)")),
      "Housing quality index"),

    same_room = lab(factor(same_room_as_index_case, levels = c(0, 1),
                           labels = c("No", "Yes")),
                    "Sleeps in same room as index case"),

    water = lab(factor(uncovered_water_receptacles, levels = c(0, 1),
                       labels = c("No", "Yes")),
                "Uncovered water receptacles"),

    malaria_parasitaemia = lab(malaria_parasitaemia, "Malaria parasitaemia")
  )

outcome    <- "malaria_parasitaemia"
predictors <- c("age_grp", "sex", "educ", "council", "bushes", "llin",
                "housing", "same_room", "water")

dat_model  <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped  <- nrow(dat) - nrow(dat_model)
n_events   <- sum(dat_model[[outcome]])
n_nonevent <- nrow(dat_model) - n_events

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
pv  <- function(p) ifelse(p < 0.001, "<0.001", paste0("= ", fmt(p, 3)))
pct <- function(x) fmt(100 * x, 1)

# Parameter arithmetic for the full shipped predictor set versus the reduced one.
shipped_preds <- c("age_group", "sex", "education", "area_council", "bushes_near_home",
                   "owns_llin", "used_llin_previous_night", "house_floor", "house_roof",
                   "house_type", "house_wall", "house_window", "same_room_as_index_case",
                   "uncovered_water_receptacles")
n_par_shipped <- dat_raw |>
  mutate(across(all_of(shipped_preds), as.factor)) |>
  select(all_of(shipped_preds)) |>
  summarise(across(everything(), ~ nlevels(.x) - 1L)) |>
  rowSums() |>
  as.integer()

n_par_final <- length(coef(glm(reformulate(predictors, outcome),
                               data = dat_model, family = binomial))) - 1L

# Correlations among the five binary "improved housing" indicators, quoted in
# Methods as the justification for collapsing them into one index.
house_cor <- dat |>
  transmute(floor  = as.integer(house_floor == "ceramictiles"),
            wall   = as.integer(house_wall  == "cement_plastered"),
            roof   = as.integer(house_roof  %in% c("metal", "roofshingles")),
            window = as.integer(house_window == "screenorglasswindows"),
            type   = as.integer(house_type  != "hut")) |>
  cor()
house_cor_lo <- min(house_cor[lower.tri(house_cor)])
house_cor_hi <- max(house_cor[lower.tri(house_cor)])
```

# Introduction

Onyiah and colleagues carried out a cross-sectional household survey in the
Federal Capital Territory, Abuja, Nigeria between March and August 2016
(*PLOS ONE* 2018, doi:10.1371/journal.pone.0203686). Laboratory-confirmed
malaria patients were recruited consecutively from twelve health facilities in
three randomly selected area councils; their households were then visited within
seven days and all consenting members were enrolled and screened, giving
`r nrow(dat)` participants from 107 households. Asymptomatic parasitaemia
matters because people carrying parasites without symptoms are not treated, and
so act as a silent reservoir that sustains transmission around a treated index
case.

**Aim.** Among household members of confirmed malaria patients in Abuja, we
examined which of age, sex, education, area council, bushes around the home,
insecticidal net use, housing quality, sleeping in the same room as the index
case and uncovered water receptacles are independently associated with
asymptomatic malaria parasitaemia, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** A cross-sectional household survey in three of the six Abuja
area councils (Abuja Municipal, Kuje, Kwali), March to August 2016. Index cases
were patients with microscopy- or RDT-confirmed malaria attending twelve
selected health facilities; the sampling unit for this analysis is the household
member, and this teaching extract holds `r nrow(dat)` household members with
`r ncol(dat_raw)` variables. Because entry into the sample required living with
a confirmed malaria patient, **the sample is selected on exposure to a household
infection** and nothing here is a population prevalence estimate.

**Variables.** The outcome is `malaria_parasitaemia`, a binary indicator where
the **event is a positive thick-film microscopy result**. There are
`r n_events` events and `r n_nonevent` non-events in the file, i.e.
`r pct(mean(dat_raw$malaria_parasitaemia))`% positive. This is worth pausing on:
the event is the *majority* class, which is unusual for a disease outcome and
has a practical consequence noted under the parameter budget below. Predictors
are age group (reference 20 and over), sex (reference female), highest education
(reference none), area council (reference Abuja Municipal, the largest),
bushes around the home (reference no), insecticidal net status (reference no
net), a housing-quality index (reference improved, the largest group), sleeping
in the same room as the index case (reference no) and uncovered water
receptacles near the home (reference no).

**Three predictors were reshaped before any model was fitted.** *(i)* The five
housing variables `house_floor`, `house_roof`, `house_type`, `house_wall` and
`house_window` are five correlated measurements of one underlying construct,
housing quality; pairwise correlations between the "improved" indicators derived
from them run from `r fmt(house_cor_lo, 2)` to `r fmt(house_cor_hi, 2)`,
and the cross-tabulations contain structural zeros (every one of the
`r sum(dat$house_type == "hut")` huts has mud walls and none has screened or
glass windows; all `r sum(dat$house_floor == "ceramictiles")` ceramic-tiled
dwellings have screened or glass windows). Entered together they cost
`r 2 + 2 + 3 + 1 + 3` parameters and generate a variance inflation factor of
about 27 for `house_window`. They were therefore replaced by a single index
counting how many of five "improved" features a dwelling has (ceramic-tiled
floor, cement-plastered walls, metal or shingle roof, screened or glass windows,
a dwelling other than a hut), collapsed to three levels. *(ii)* `owns_llin` and
`used_llin_previous_night` are **exactly nested**: all
`r sum(dat_raw$used_llin_previous_night)` participants who used a net owned one,
and all `r sum(dat_raw$owns_llin == 0)` non-owners are coded 0 for use, so the
"did not use" category silently mixes people with no net and people with an
unused net. They were combined into one three-level variable. *(iii)*
`age_group` was collapsed from six levels to four, because the 20-24 stratum
holds only `r sum(dat_raw$age_group == "20_to_24")` participants and the three
adult strata have nearly identical parasitaemia
(`r pct(mean(dat_raw$malaria_parasitaemia[dat_raw$age_group == "20_to_24"]))`%,
`r pct(mean(dat_raw$malaria_parasitaemia[dat_raw$age_group == "25_to_34"]))`% and
`r pct(mean(dat_raw$malaria_parasitaemia[dat_raw$age_group == "35_plus"]))`%);
the childhood strata, which are the clinically interesting ones, are kept
separate.

**Parameter budget.** The fourteen predictors as shipped cost
`r n_par_shipped` parameters. The relevant denominator for the ten-events-per-
parameter rule is not the `r n_events` positives but the `r n_nonevent`
**negatives**, because a logistic model is constrained by the smaller of the two
outcome classes; the shipped set therefore runs at
`r fmt(n_nonevent / n_par_shipped, 1)` per parameter, below the conventional
minimum. The reduced set above costs `r n_par_final` parameters, i.e.
`r fmt(n_nonevent / n_par_final, 1)` negatives per parameter, which clears it.
The budget was fixed before any model was fitted.

**Statistical analysis.** `glm(family = binomial)`, justified because the
outcome is binary. We fitted a univariable model for each predictor first, then
a single multivariable model containing all of them. Predictors were chosen in
advance on subject-matter grounds, not by significance. Coefficients are
exponentiated and reported as odds ratios with 95% confidence intervals;
significance threshold 0.05. Model adequacy was assessed with
`performance::check_model()`, focusing on binned residuals, collinearity,
influential observations and events per parameter, and every categorical
predictor was cross-tabulated against the outcome to check for separation before
fitting. **The models treat the `r nrow(dat)` participants as independent, which
they are not**: they sit inside 107 households, and no household identifier is
shipped in this extract, so the clustering cannot be adjusted for here. The
consequences are set out under Limitations. Analysis used `r R.version.string`
with tidyverse, gtsummary and performance.

**Missing data.** There is none. All `r ncol(dat_raw)` columns are 100%
complete, so the complete-case analysis retains `r nrow(dat_model)` of
`r nrow(dat)` rows (`r n_dropped` dropped) and all `r n_events` events. This is
the one respect in which the dataset is unusually easy to work with.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  mutate(malaria_parasitaemia = factor(malaria_parasitaemia, levels = c(0, 1),
                                       labels = c("Negative", "Positive"))) |>
  tbl_summary(
    by = malaria_parasitaemia,
    missing = "no"
  ) |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics by microscopy result.**")
```

Positives outnumber negatives in every stratum of every predictor, which is the
direct consequence of a `r pct(mean(dat_model[[outcome]]))`% overall positivity
rate; the informative comparisons are therefore between columns rather than
within them. The clearest contrast is bushes around the home, present for
`r pct(mean(dat_model$bushes[dat_model[[outcome]] == 1] == "Yes"))`% of positives
against `r pct(mean(dat_model$bushes[dat_model[[outcome]] == 0] == "Yes"))`% of
negatives, and education, where
`r pct(mean(dat_model$educ[dat_model[[outcome]] == 1] == "None"))`% of positives
have no formal education against
`r pct(mean(dat_model$educ[dat_model[[outcome]] == 0] == "None"))`% of negatives.
Net status and sleeping in the same room as the index case look almost identical
in the two columns.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Percentage of household members with microscopy-positive malaria parasitaemia, by age group and whether there are bushes around the home."}
dat_model |>
  mutate(age_grp = factor(age_grp, levels = c("Under 5", "5-9", "10-19", "20 and over"))) |>
  group_by(age_grp, bushes) |>
  summarise(p = mean(.data[[outcome]]), n = n(), .groups = "drop") |>
  ggplot(aes(age_grp, p, fill = bushes)) +
  geom_col(position = position_dodge(0.85), width = 0.78) +
  geom_text(aes(label = paste0("n=", n)), position = position_dodge(0.85),
            vjust = -0.4, size = 2.9, colour = "grey30") +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1),
                     expand = expansion(c(0, 0.08))) +
  scale_fill_manual(values = c("No" = "#7FA9C9", "Yes" = "#B5495B")) +
  labs(x = "Age group (years)", y = "Positive by microscopy",
       fill = "Bushes around the home")
```

Parasitaemia is higher where there are bushes around the home in every age
group, and the age pattern is a childhood peak at 5-9 years rather than a
gradient: the school-age children in bushy compounds are the highest-prevalence
cell in the figure at
`r {t <- dat_model |> filter(age_grp == "5-9", bushes == "Yes"); pct(mean(t[[outcome]]))}`%.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = glm,
    method.args = list(family = binomial),
    exponentiate = TRUE
  )
uv

utid <- map_dfr(predictors, function(v) {
  tidy(glm(reformulate(v, outcome), data = dat_model, family = binomial),
       conf.int = TRUE, exponentiate = TRUE) |>
    filter(term != "(Intercept)")
})
ug <- function(term, d = 2) {
  r <- utid[utid$term == term, ]
  sprintf("OR %s (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d), fmt(r$conf.high, d))
}
up <- function(term) pv(utid$p.value[utid$term == term])
```

Bushes around the home carry the strongest crude association: their presence
multiplies the odds of parasitaemia by `r ug("bushesYes")`
(p `r up("bushesYes")`). Education shows a monotone crude gradient against the
no-education reference — primary `r ug("educPrimary")` (p `r up("educPrimary")`),
secondary `r ug("educSecondary")` (p `r up("educSecondary")`) and tertiary
`r ug("educTertiary")` (p `r up("educTertiary")`) — and age shows a peak rather
than a gradient, with children aged 5-9 at `r ug("age_grp5-9")`
(p `r up("age_grp5-9")`) relative to adults while under-5s
(`r ug("age_grpUnder 5")`, p `r up("age_grpUnder 5")`) and 10-19 year-olds
(`r ug("age_grp10-19")`, p `r up("age_grp10-19")`) are not distinguishable from
them. Men have `r ug("sexMale")` (p `r up("sexMale")`), just inside the
threshold. Intermediate housing sits at `r ug("housingIntermediate (3-4)")`
(p `r up("housingIntermediate (3-4)")`) but basic housing does not
(`r ug("housingBasic (0-2 features)")`, p `r up("housingBasic (0-2 features)")`),
so the housing "gradient" is not monotone and should not be read as one.

Three predictors that the study design puts front and centre are flatly null in
crude analysis. Sleeping in the same room as the index case gives
`r ug("same_roomYes")` (p `r up("same_roomYes")`) — the point estimate is if
anything *protective*. Owning and using a net the previous night gives
`r ug("llinOwns, used last night")` (p `r up("llinOwns, used last night")`), and
owning a net but not using it gives
`r ug("llinOwns, not used last night")`
(p `r up("llinOwns, not used last night")`), so neither ownership nor use marks
lower parasitaemia. Uncovered water receptacles give `r ug("waterYes")`
(p `r up("waterYes")`), and neither of the two non-reference area councils
differs from Abuja Municipal (Kuje `r ug("councilKuje")`,
p `r up("councilKuje")`; Kwali `r ug("councilKwali")`, p `r up("councilKwali")`).

## Why adjustment is needed

**Area council** confounds the housing and bushes associations. The three
councils differ sharply in dwelling stock: only
`r {t <- dat_model |> filter(council == "Abuja Municipal"); pct(mean(t$housing == "Basic (0-2 features)"))}`%
of Abuja Municipal participants live in basic housing against
`r {t <- dat_model |> filter(council == "Kuje"); pct(mean(t$housing == "Basic (0-2 features)"))}`%
in Kuje, and bushes around the home are reported by
`r {t <- dat_model |> filter(council == "Abuja Municipal"); pct(mean(t$bushes == "Yes"))}`%
of Abuja Municipal households against
`r {t <- dat_model |> filter(council == "Kuje"); pct(mean(t$bushes == "Yes"))}`%
in Kuje. Area council is also plausibly related to the outcome independently of
either, because the councils differ in transmission intensity, vector density
and distance to care. Without adjustment, a housing coefficient is partly a
geography coefficient.

**Age** confounds the education association, and severely. Education in a
household survey is largely an age marker for children: of the
`r sum(dat_model$age_grp %in% c("Under 5", "5-9"))` participants under ten,
`r sum(dat_model$age_grp %in% c("Under 5", "5-9") & dat_model$educ %in% c("None", "Primary"))`
are recorded as having no education or primary education only and the remaining
`r sum(dat_model$age_grp %in% c("Under 5", "5-9") & dat_model$educ == "Secondary")`
as secondary; not one is recorded as tertiary, whereas
`r pct(mean(dat_model$educ[dat_model$age_grp == "20 and over"] == "Tertiary"))`%
of the adults are. The low-education strata are therefore
dominated by exactly the young children who have the highest parasitaemia for
immunological reasons. Age is independently related to the outcome through
acquired immunity, and it is related to recorded education by construction. The
crude education gradient therefore cannot be read as an effect of schooling or
of the socioeconomic position it proxies until age is held constant.

## Multivariable model

```{r multivariable}
fit <- glm(reformulate(predictors, outcome), data = dat_model, family = binomial)

mv <- tbl_regression(fit, exponentiate = TRUE)
mv

tid <- tidy(fit, conf.int = TRUE, exponentiate = TRUE)
g <- function(term, d = 2) {
  r <- tid[tid$term == term, ]
  sprintf("OR %s (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d), fmt(r$conf.high, d))
}
gp <- function(term) pv(tid$p.value[tid$term == term])
```

```{r fit-stats}
n_par    <- length(coef(fit)) - 1L
epp      <- n_events / n_par
enpp     <- n_nonevent / n_par
null_fit <- glm(reformulate("1", outcome), data = dat_model, family = binomial)
tibble(
  Statistic = c("Complete-case N", "Events (parasitaemia = 1)",
                "Non-events (parasitaemia = 0)", "Parameters (excl. intercept)",
                "Events per parameter", "Non-events per parameter (binding)",
                "AIC (null model)", "AIC (full model)", "Tjur's R-squared"),
  Value = c(as.character(nrow(dat_model)), as.character(n_events),
            as.character(n_nonevent), as.character(n_par),
            fmt(epp, 1), fmt(enpp, 1),
            fmt(AIC(null_fit), 1), fmt(AIC(fit), 1),
            fmt(as.numeric(r2_tjur(fit)), 3))
) |>
  gt() |>
  tab_header(title = md("**Model fit: multivariable logistic model for malaria parasitaemia**"))
```

The model carries `r n_par` parameters. On the usual count that is
`r fmt(epp, 1)` events per parameter, comfortably clear of ten, but that number
flatters the model: because the event is the majority class the binding
constraint is the `r n_nonevent` negatives, giving
`r fmt(enpp, 1)` per parameter. AIC falls only from `r fmt(AIC(null_fit), 1)`
for the intercept-only model to `r fmt(AIC(fit), 1)`, a drop of
`r fmt(AIC(null_fit) - AIC(fit), 1)`, and Tjur's R-squared is
`r fmt(as.numeric(r2_tjur(fit)), 3)`. The nine predictors together explain very
little: with `r pct(mean(dat_model[[outcome]]))`% of everybody positive, the
model has limited room to discriminate.

Among participants of the same sex, education, area council, housing, net
status, room-sharing and water storage, bushes around the home remain associated
with `r g("bushesYes")` (p `r gp("bushesYes")`) — essentially unchanged from the
crude estimate and the only predictor significant at 0.05 with a wide margin.
Among participants alike on everything else, children aged 5-9 have
`r g("age_grp5-9")` (p `r gp("age_grp5-9")`) relative to adults, while under-5s
(`r g("age_grpUnder 5")`) and 10-19 year-olds (`r g("age_grp10-19")`) are
indistinguishable from adults. Education retains a gradient after adjustment:
tertiary `r g("educTertiary")` (p `r gp("educTertiary")`) and primary
`r g("educPrimary")` (p `r gp("educPrimary")`) against no education, though
secondary (`r g("educSecondary")`, p `r gp("educSecondary")`) is not significant
and the levels are not ordered as a clean dose-response.

The three design-relevant exposures stay null. Sleeping in the same room as the
index case gives `r g("same_roomYes")` (p `r gp("same_roomYes")`); using a net
the previous night gives `r g("llinOwns, used last night")`
(p `r gp("llinOwns, used last night")`), and owning an unused net
`r g("llinOwns, not used last night")`
(p `r gp("llinOwns, not used last night")`). Housing quality
(`r g("housingIntermediate (3-4)")` for intermediate,
`r g("housingBasic (0-2 features)")` for basic) and uncovered water receptacles
(`r g("waterYes")`) are also null. These are associations among the participants
observed, not effects of intervening.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted odds ratios.**")
```

The coefficient that moves furthest is **age under 5**, from
`r ug("age_grpUnder 5")` crude to `r g("age_grpUnder 5")` adjusted — the whole
of the apparent excess in the youngest children disappears, moving all the way
to the null. This is the education entanglement described above operating in
reverse: under-5s are almost all in the "no education" reference stratum, so
once education is in the model the "no education" coefficient absorbs the
childhood excess and nothing is left for age to carry. Fitting the same model
without education returns the under-5 estimate to
`r {f_noed <- glm(reformulate(setdiff(predictors, "educ"), outcome), data = dat_model, family = binomial); fmt(exp(coef(f_noed)["age_grpUnder 5"]), 2)}`,
close to its crude value, which confirms the mechanism. **Area council** moves
in the opposite direction, away from the null: Kuje goes from
`r ug("councilKuje")` to `r g("councilKuje")`. Crudely Kuje looks like Abuja
Municipal, but Kuje households have far poorer housing and lower education, both
of which are associated with more parasitaemia; hold those constant and a lower
Kuje odds emerges. This is classic masking rather than confounding away a real
signal. **Uncovered water receptacles** move the same way, from
`r ug("waterYes")` to `r g("waterYes")`, though null in both columns.

Two coefficients barely move. **Bushes** goes from `r ug("bushesYes")` to
`r g("bushesYes")`, a shift of a few percent, so its association is essentially
independent of everything else measured; and **tertiary education** goes from
`r ug("educTertiary")` to `r g("educTertiary")`.

Three contrasts change significance, all in the same direction — losing it.
**Sex** goes from p `r up("sexMale")` to p `r gp("sexMale")`, crossing 0.05 with
almost no change in the point estimate (`r ug("sexMale")` to `r g("sexMale")`),
which is a loss of precision rather than confounding. **Secondary education**
goes from `r ug("educSecondary")` (p `r up("educSecondary")`) to
`r g("educSecondary")` (p `r gp("educSecondary")`), attenuating toward the null
because secondary-educated participants are disproportionately adolescents and
young adults, so age adjustment removes part of the crude contrast. **Intermediate
housing** goes from `r ug("housingIntermediate (3-4)")`
(p `r up("housingIntermediate (3-4)")`) to `r g("housingIntermediate (3-4)")`
(p `r gp("housingIntermediate (3-4)")`), confounded by area council as
anticipated. Nothing becomes significant on adjustment.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r binned-resid}
br <- binned_residuals(fit)
print(br)
```

```{r collin}
collin <- check_collinearity(fit)
collin
resid_ok <- attr(br, "resid_ok")
```

```{r separation}
# Explicit separation audit: every shipped categorical predictor crossed with the
# outcome, plus the estimated coefficients and standard errors of the final model.
sep_tab <- map_dfr(shipped_preds, function(v) {
  t <- table(dat_raw[[v]], dat_raw[[outcome]])
  tibble(Variable = v, Levels = nrow(t),
         `Zero cells` = sum(t == 0),
         `Smallest cell` = min(t),
         `Level with smallest cell` = rownames(t)[which.min(apply(t, 1, min))])
})
sep_tab |>
  gt() |>
  tab_header(title = md("**Separation audit: every shipped categorical predictor against the outcome**"))
```

**Separation.** No shipped predictor separates the outcome on its own: every one
of the `r nrow(sep_tab)` variables has both a positive and a negative in every
level, so `Zero cells` is `r max(sep_tab[["Zero cells"]])` throughout. The
smallest cell anywhere is `r min(sep_tab[["Smallest cell"]])` observations,
in `r sep_tab$Variable[which.min(sep_tab[["Smallest cell"]])]` at level
"`r sep_tab[["Level with smallest cell"]][which.min(sep_tab[["Smallest cell"]])]`",
which has only `r sum(dat_raw$house_type == "bungalow" & dat_raw[[outcome]] == 0)`
negatives among `r sum(dat_raw$house_type == "bungalow")` participants. In the
fitted model no coefficient is extreme: the largest absolute log-odds is
`r fmt(max(abs(coef(fit)[-1])), 2)`, the largest standard error is
`r fmt(max(summary(fit)$coefficients[-1, 2]), 2)`, and fitted probabilities span
`r fmt(min(fitted(fit)), 2)` to `r fmt(max(fitted(fit)), 2)`, nowhere near 0 or
1. There is no separation in this model. Sparse *combinations* are a different
matter and are documented in the feasibility notes.

**Collinearity.** All variance inflation factors are modest once the housing
variables have been reduced to an index: the largest is
`r fmt(max(collin$VIF), 2)` for `r collin$Term[which.max(collin$VIF)]`, and
because most of these are multi-level factors the fairer figure is the standard-
error inflation factor, whose maximum across all nine terms is
`r fmt(max(collin$SE_factor), 2)`. The five raw housing variables entered
together would have given a VIF near 27 for `house_window`; the index removes
that entirely.

**Binned residuals.** Only `r fmt(100 * resid_ok, 0)`% of binned
residuals lie inside their error bounds, against a target of 95% or more, so the
model is imperfectly calibrated. With so few negatives spread across
`r nrow(dat_model)` observations each bin contains only a handful of them, which
makes the check noisy, but the shortfall is real and should be reported rather
than waved away.

**Influence.** No observation dominates: the largest Cook's distance is
`r fmt(max(cooks.distance(fit)), 3)` and the largest leverage is
`r fmt(max(hatvalues(fit)), 3)`, both far below any conventional threshold. This
is expected in a dataset with no continuous predictors and no rare categories.

**What does not apply.** Homogeneity of variance and normality of residuals are
assumptions of the linear model, not of logistic regression, and the panels
`check_model()` produces for them are not interpreted here.

# Discussion

Among household members of confirmed malaria patients in Abuja,
`r pct(mean(dat_model[[outcome]]))`% carried malaria parasites without symptoms,
and almost nothing measured in this survey distinguishes those who did from
those who did not. The one robust finding is that having bushes around the home
is associated with roughly `r fmt(tid$estimate[tid$term == "bushesYes"], 1)`-fold
higher odds of parasitaemia, an association that is essentially unchanged by
adjustment for age, sex, education, geography, housing quality, net use and room
sharing. Children aged 5-9 also stand out, at `r g("age_grp5-9")` relative to
adults. Practically, this points at peridomestic vector breeding habitat and at
school-age children as the targets for reactive case detection around a treated
index case.

The negative findings are as important as the positive one. Sleeping in the same
room as the index case — the exposure the study design exists to measure — shows
no association at all (`r g("same_roomYes")`), and neither owning nor using an
insecticidal net the previous night is associated with lower parasitaemia
(`r g("llinOwns, used last night")` for use). With a positivity rate of
`r pct(mean(dat_model[[outcome]]))`% across the whole sample, infection is close
to universal in these compounds, and a within-household exposure contrast has
very little variation left to explain.

**Limitations.** The most serious problem is **clustering**. The
`r nrow(dat_model)` participants come from 107 households, an average of about
`r fmt(nrow(dat_model) / 107, 1)` people each, and household members share the
same compound, the same vector exposure, the same bushes, the same roof and the
same index case. Every standard error and p-value above assumes they are
independent, which they plainly are not, so the confidence intervals are too
narrow and the p-values too small — most acutely for the household-level
variables (bushes, housing, area council, water receptacles), which really have
at most 107 independent values, not `r nrow(dat_model)`. **No household
identifier is shipped in this extract**, so this cannot be corrected here; the
published analysis used generalised estimating equations with the household code
as the clustering variable, which is not reproducible from these columns. The
quantitative size of the problem is estimated in the feasibility notes.

The sample is **selected on household exposure**: everybody here lives with a
laboratory-confirmed malaria patient, so `r pct(mean(dat_model[[outcome]]))`%
must not be quoted as a prevalence for Abuja or for Nigeria, and the null net
and room-sharing associations may reflect a ceiling effect in a very
high-transmission subset rather than genuine absence of effect. The design is
**cross-sectional**, so reverse causation is live for the modifiable exposures:
households that have recently had a malaria case may have acquired or started
using nets in response, which would bias the net association toward or past the
null. Exposure measurement is crude — bushes around the home, uncovered water
receptacles and net use the previous night are single self-reported or
observed-once items, and net *use* says nothing about how the net was used or
whether it was still insecticidal. The **housing index is a construction**, not
a validated instrument; it weights five features equally and its middle category
is not intermediate in risk, so its coefficients should be read as descriptive.
**Education is substantively invalid for a large part of the sample**:
`r sum(dat_model$age_grp %in% c("Under 5", "5-9"))` participants are under ten
and `r sum(dat_raw$age_group == "under_5" & dat_raw$education == "primary")`
under-fives are recorded as having primary education, so the variable is partly
an age marker and partly noise, and its coefficients cannot be given a schooling
interpretation. The binned-residual check fails. Unmeasured confounders that
matter here — indoor residual spraying, recent antimalarial treatment, altitude
and proximity to standing water, whether the index case was still parasitaemic —
are simply not in the file. All estimates are associations, not causal effects.

**Future work.** Refit with the household identifier as a clustering variable,
using either generalised estimating equations or a household random intercept,
and add parasite density so that low-density carriage can be distinguished from
clinically meaningful infection.

# Feasibility notes

*NOT part of a student report.*

```{r feas-cluster}
# --------------------------------------------------------------------------
# The shipped CSV carries no household identifier. To quantify how much the
# clustering matters, the household code was recovered from the PLOS ONE
# supporting-information file (row order is identical: sex, age group and
# room-sharing match row for row) and is embedded here in run-length form.
# STUDENTS CANNOT DO THIS. It exists only to size the problem for the brief.
# --------------------------------------------------------------------------
suppressMessages({library(sandwich); library(lme4)})

rl <- c(1,8, 2,5, 3,3, 4,7, 5,2, 6,4, 107,1, 7,3, 8,1, 9,4, 10,6, 11,3, 12,5,
        13,6, 14,2, 15,1, 16,8, 17,4, 18,1, 19,9, 21,4, 22,12, 23,7, 24,4, 25,4,
        26,3, 27,6, 28,10, 29,3, 30,5, 31,4, 32,3, 33,4, 34,3, 35,5, 36,2, 38,4,
        39,2, 40,1, 41,1, 42,9, 43,3, 44,1, 46,4, 47,6, 48,3, 49,1, 50,2, 49,4,
        50,3, 52,4, 53,3, 54,2, 55,5, 56,6, 57,4, 58,8, 59,5, 60,11, 61,5, 62,3,
        63,8, 64,3, 65,3, 66,15, 67,2, 68,3, 69,10, 70,3, 71,5, 72,7, 73,8, 74,1,
        75,4, 76,3, 77,4, 78,3, 79,4, 80,4, 81,1, 82,4, 83,20, 84,4, 85,16, 86,4,
        87,3, 88,2, 89,3, 90,3, 91,8, 92,6, 93,5, 94,10, 96,4, 97,3, 98,5, 99,3,
        100,4, 101,4, 102,4, 104,7, 105,12, 106,8, 12,2, 15,4, 18,2, 20,5, 23,1,
        29,1, 34,5, 36,5, 37,2, 41,6, 44,2, 45,3, 46,2, 51,4, 59,1, 61,6, 63,5,
        67,3, 69,5, 70,1, 74,4, 76,5, 78,1, 86,3, 89,2, 90,7, 91,2, 93,1, 95,3,
        96,1, 97,5, 98,1, 99,1, 103,4, 105,2)
hh <- factor(rep(rl[c(TRUE, FALSE)], rl[c(FALSE, TRUE)]))

dat_hh   <- dat_model |> mutate(hh = hh)
hh_sizes <- as.numeric(table(hh))
hh_pos   <- tapply(dat_hh[[outcome]], hh, mean)

# Naive versus cluster-robust standard errors on the very same model.
se_naive <- summary(fit)$coefficients[, 2]
se_clust <- sqrt(diag(vcovCL(fit, cluster = hh)))
ratio    <- se_clust / se_naive
p_naive  <- 2 * pnorm(-abs(coef(fit) / se_naive))
p_clust  <- 2 * pnorm(-abs(coef(fit) / se_clust))

is_hh_level  <- grepl("^council|^bushes|^housing|^water", names(se_naive))
is_ind_level <- grepl("^age_grp|^sexM|^educ|^same_room", names(se_naive))

# Household random intercept, for the intraclass correlation.
ctrl <- glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
m_null <- glmer(malaria_parasitaemia ~ 1 + (1 | hh), data = dat_hh,
                family = binomial, control = ctrl)
m_full <- glmer(reformulate(c(predictors, "(1 | hh)"), outcome), data = dat_hh,
                family = binomial, control = ctrl)
v_null <- as.numeric(VarCorr(m_null)$hh)
v_full <- as.numeric(VarCorr(m_full)$hh)
icc_null <- v_null / (v_null + pi^2 / 3)
icc_full <- v_full / (v_full + pi^2 / 3)
deff <- 1 + (mean(hh_sizes) - 1) * icc_null

tibble(
  Term = names(se_naive),
  `SE (naive glm)` = fmt(se_naive, 3),
  `SE (cluster-robust)` = fmt(se_clust, 3),
  `Inflation` = fmt(ratio, 2),
  `p (naive)` = ifelse(p_naive < 0.001, "<0.001", fmt(p_naive, 3)),
  `p (clustered)` = ifelse(p_clust < 0.001, "<0.001", fmt(p_clust, 3))
) |>
  gt() |>
  tab_header(title = md("**How much the clustering costs: naive versus cluster-robust standard errors**"))
```

```{r feas-separation}
# Deliberate attempts to break the outcome. Warnings are captured rather than
# printed so that their presence can be reported in the table.
dat_fac <- dat_raw |> mutate(across(all_of(shipped_preds), as.factor))

sep_check <- function(label, form, data = dat_fac) {
  w <- character()
  f <- withCallingHandlers(
    glm(form, data = data, family = binomial),
    warning = function(cond) { w <<- c(w, conditionMessage(cond)); invokeRestart("muffleWarning") }
  )
  co <- summary(f)$coefficients
  tibble(Model = label,
         Parameters = length(coef(f)) - 1L,
         `Aliased (NA) terms` = sum(is.na(coef(f))),
         `Max |coefficient|` = fmt(max(abs(coef(f)[-1]), na.rm = TRUE), 1),
         `Max SE` = fmt(max(co[-1, 2]), 1),
         `Max fitted prob.` = fmt(max(fitted(f)), 7),
         `0/1 fitted warning` = if (any(grepl("numerically 0 or 1", w))) "YES" else "no")
}

stress <- bind_rows(
  sep_check("All 14 shipped predictors",
            reformulate(shipped_preds, outcome)),
  sep_check("All five house_* variables + area_council",
            reformulate(c("house_floor", "house_roof", "house_type", "house_wall",
                          "house_window", "area_council"), outcome)),
  sep_check("house_window * house_wall",
            as.formula(paste(outcome, "~ house_window * house_wall"))),
  sep_check("age_group * education",
            as.formula(paste(outcome, "~ age_group * education"))),
  sep_check("house_type * area_council",
            as.formula(paste(outcome, "~ house_type * area_council"))),
  sep_check("Final reduced model (this document)",
            reformulate(predictors, outcome), data = dat_model)
)
stress |>
  gt() |>
  tab_header(title = md("**Separation stress tests**"))

sv <- function(label, col) stress[[col]][stress$Model == label]

n_cw_open <- sum(dat_raw$house_wall == "cement_plastered" & dat_raw$house_window == "openwindows")
n_cw_open_pos <- sum(dat_raw$house_wall == "cement_plastered" &
                     dat_raw$house_window == "openwindows" &
                     dat_raw[[outcome]] == 1)

fit_shipped <- glm(reformulate(shipped_preds, outcome), data = dat_fac, family = binomial)
fit_noeduc  <- glm(reformulate(setdiff(predictors, "educ"), outcome),
                   data = dat_model, family = binomial)
```

- **MAJOR (structural, and the dominant problem): the 602 participants sit in
  107 households and no household identifier is shipped.** Household sizes run
  from `r min(hh_sizes)` to `r max(hh_sizes)` with a mean of
  `r fmt(mean(hh_sizes), 1)`. Recovering the household code from the source
  supporting file (see the chunk above; students cannot do this) gives an
  intraclass correlation of `r fmt(icc_null, 2)` on the latent scale from an
  empty random-intercept model — household variance `r fmt(v_null, 2)` against
  the logistic residual variance of `r fmt(pi^2 / 3, 2)` — and it stays at
  `r fmt(icc_full, 2)` after adjustment for all nine predictors, so the
  covariates explain almost none of the household-level variation. At a mean
  cluster size of `r fmt(mean(hh_sizes), 1)` that is a design effect of about
  **`r fmt(deff, 1)`**: the effective sample size is roughly
  `r round(nrow(dat_model) / deff)` participants, not `r nrow(dat_model)`, and
  the effective number of negatives is about `r round(n_nonevent / deff)`, which
  turns the model's `r fmt(enpp, 1)` negatives per parameter into about
  `r fmt((n_nonevent / deff) / n_par, 1)`. On the same fitted model,
  cluster-robust standard errors are a median of
  `r fmt(median(ratio), 2)` times the naive ones overall, but the split is the
  instructive part: a median of `r fmt(median(ratio[is_hh_level]), 2)` for the
  household-level predictors (worst:
  `r names(se_naive)[which.max(ratio)]` at `r fmt(max(ratio), 2)`) against
  `r fmt(median(ratio[is_ind_level]), 2)` for individual-level ones, and one
  individual-level term (`sexMale`) actually gets *smaller*. In all,
  `r sum(xor(p_naive < 0.05, p_clust < 0.05))` coefficients cross 0.05 when the
  clustering is respected:
  `r paste0("\x60", names(se_naive)[p_naive < 0.05 & p_clust >= 0.05], "\x60", collapse = " and ")`
  lose significance and
  `r paste0("\x60", names(se_naive)[p_naive >= 0.05 & p_clust < 0.05], "\x60", collapse = " and ")`
  gains it. A household random intercept inflates the standard errors further
  still (`bushesYes` from `r fmt(se_naive["bushesYes"], 3)` to
  `r fmt(sqrt(diag(vcov(m_full)))[names(fixef(m_full)) == "bushesYes"], 3)`).
  **Recommendation: ship `household_id` in the CSV.** It is one integer column,
  it is present in the source file, the published paper's own analysis depends
  on it, and without it a student cannot do anything about the single largest
  methodological problem in the dataset. If it stays out, the caution needs to
  say plainly that the clustering cannot be corrected, not merely that a simple
  model "does not account for" it.
- **MAJOR: selection on exposure.** Every participant lives with a
  laboratory-confirmed malaria patient. The `r pct(mean(dat_raw[[outcome]]))`%
  positivity is a within-outbreak-household figure and must never be quoted as a
  population prevalence. It also compresses the exposure contrasts: with seven in
  ten positive, `same_room_as_index_case`, `owns_llin` and
  `used_llin_previous_night` have almost no discriminating room left, which is
  the likeliest explanation for all three being null.
- **MAJOR: the outcome is the majority class**, `r n_events` positive versus
  `r n_nonevent` negative. Students who count "events" as the 1s will compute
  `r fmt(epp, 1)` events per parameter and conclude they can afford a large
  model; the binding constraint is the `r n_nonevent` negatives
  (`r fmt(enpp, 1)` per parameter here, and only about
  `r fmt((n_nonevent / deff) / n_par, 1)` once clustering is allowed for). The
  full fourteen-predictor set costs `r n_par_shipped` parameters, i.e.
  `r fmt(n_nonevent / n_par_shipped, 1)` negatives per parameter, which is
  over-fitting. Worth flagging in the brief because every other logistic dataset
  in this catalogue has a minority-class event.
- **MAJOR: the five `house_*` variables are one construct with structural
  zeros.** `house_type` and `house_wall` are nearly aliased (all
  `r sum(dat_raw$house_type == "hut")` huts have mud walls; all
  `r sum(dat_raw$house_type %in% c("bungalow", "duplex"))` bungalows and duplexes
  have cement-plastered walls), and all
  `r sum(dat_raw$house_floor == "ceramictiles")` ceramic-tiled dwellings have
  screened or glass windows. Entered together they give `house_window` a VIF of
  about 27 and `house_type` about 9 while adding almost nothing: the crude
  positivity rates across the three levels of the composite index are
  `r paste(pct(tapply(dat$malaria_parasitaemia, dat$housing, mean)[c("Basic (0-2 features)", "Intermediate (3-4)", "Improved (all 5)")]), collapse = "%, ")`%
  respectively, i.e. non-monotone and essentially flat. Housing quality simply
  does not predict this outcome, so the choice between the five raw variables
  and a composite changes nothing substantive — but the five raw variables will
  produce nonsense-looking coefficients that a student then has to explain.
- **MAJOR: `age_group * education` separates completely.** The interaction is
  the obvious thing to reach for once someone notices the education gradient is
  age-driven, and it fails hard:
  `r sv("age_group * education", "Aliased (NA) terms")` terms are aliased (no
  under-5 has secondary or tertiary education, no 5-9 year-old has tertiary),
  the largest surviving interaction coefficient is
  `r sv("age_group * education", "Max |coefficient|")` on the log-odds scale
  with a standard error of `r sv("age_group * education", "Max SE")`, and the
  largest fitted probability is
  `r sv("age_group * education", "Max fitted prob.")`. An interaction between
  two of the housing variables does the same: `house_window * house_wall`
  produces a coefficient of `r sv("house_window * house_wall", "Max |coefficient|")`
  with a standard error of `r sv("house_window * house_wall", "Max SE")`,
  because only `r n_cw_open` participants have cement-plastered walls and open
  windows and **all `r n_cw_open_pos` of them are positive**. That single
  four-person cell is the specific culprit. `house_type * area_council` does not
  diverge but aliases `r sv("house_type * area_council", "Aliased (NA) terms")`
  terms, because duplexes occur only in Abuja Municipal and huts only in Kuje and
  Kwali.
- **MAJOR, and easy to miss: R issues no separation warning in any of these
  models.** The `0/1 fitted warning` column in the stress-test table reads "no"
  throughout, including for `age_group * education`, because `glm()` only warns
  when a fitted probability exceeds `1 - 1e-8` and the largest here is
  `r sv("age_group * education", "Max fitted prob.")` — just under the trigger.
  A student who treats "no warning" as "no separation" will report a coefficient
  of `r sv("age_group * education", "Max |coefficient|")` with a standard error
  of `r sv("age_group * education", "Max SE")` as a finding. The reliable
  detectors here are the standard errors, the aliased `NA` coefficients and the
  cross-tabulations, not the warning.
- **MAJOR: `owns_llin` and `used_llin_previous_night` are exactly nested.** All
  `r sum(dat_raw$used_llin_previous_night == 1)` users are owners; all
  `r sum(dat_raw$owns_llin == 0)` non-owners are coded 0 for use. Putting both in
  a model does not separate — the model runs — but the "did not use" reference
  then pools `r sum(dat_raw$owns_llin == 0)` people with no net and
  `r sum(dat_raw$owns_llin == 1 & dat_raw$used_llin_previous_night == 0)` people
  with an unused net, the two coefficients become uninterpretable mirror images
  (in the full shipped model they are
  `r fmt(coef(fit_shipped)["owns_llin1"], 2)` and
  `r fmt(coef(fit_shipped)["used_llin_previous_night1"], 2)` on the log-odds
  scale, cancelling almost exactly), and each carries a variance inflation factor
  above 4.5. The three-level recode used here is the only coherent
  parameterisation.
- **MINOR but worth knowing: `education` is fully observed and substantively
  meaningless for a third of the sample.** There is no missingness to warn
  about — which is itself the trap, because nothing flags the problem.
  `r sum(dat_raw$age_group == "under_5")` participants are under five, of whom
  `r sum(dat_raw$age_group == "under_5" & dat_raw$education == "primary")` are
  recorded as having *primary* education, and
  `r sum(dat_raw$age_group %in% c("under_5", "5_to_9") & dat_raw$education == "secondary")`
  children under ten are recorded as *secondary*. The variable is an age proxy
  below age ten and an SES proxy above it, and the two age groups it entangles
  are the ones with the highest parasitaemia. Concretely, adding `educ` to the
  model moves the under-5 odds ratio from
  `r fmt(exp(coef(fit_noeduc)["age_grpUnder 5"]), 2)` to
  `r fmt(tid$estimate[tid$term == "age_grpUnder 5"], 2)` — the childhood signal
  is transferred wholesale to the education term.
- **MINOR: `area_council` is thin at one level.** Kwali holds only
  `r sum(dat_raw$area_council == "kwali")` participants
  (`r sum(dat_raw$area_council == "kwali" & dat_raw[[outcome]] == 0)` negatives),
  so its confidence interval is wide (`r g("councilKwali")`), and Kwali contains
  no roof-shingle dwellings at all, which is part of why the raw housing
  variables behave badly alongside it. Three levels is still workable; two would
  not be an unreasonable collapse.
- **MINOR: `house_type = bungalow` is the sparsest single cell**, with
  `r sum(dat_raw$house_type == "bungalow" & dat_raw[[outcome]] == 0)` negatives
  out of `r sum(dat_raw$house_type == "bungalow")` participants
  (`r pct(mean(dat_raw[[outcome]][dat_raw$house_type == "bungalow"]))`% positive),
  and `house_window = metal_shutters` has only
  `r sum(dat_raw$house_window == "metal_shutters")` participants in total. Neither
  produces separation on its own, but both are one dropped observation away from
  it.
- **MINOR: binned residuals fail** (`r fmt(100 * resid_ok, 0)`% inside the
  bounds against a 95% target). With `r n_nonevent` negatives spread over
  `r nrow(dat_model)` rows the check is genuinely noisy, but a student following
  the diagnostics honestly will have to report a failed check and should not be
  penalised for it.
- **MINOR: the outcome is test-dependent, and the source file shows the tests
  disagree.** The supporting file carries both a microscopy result and an RDT
  result for every participant. Microscopy gives
  `r sum(dat_raw[[outcome]])` positives; the RDT gives 83, the two agree on only
  39.9% of participants, and Cohen's kappa is 0.07 — no better than chance. Only
  the microscopy result is shipped, which is defensible (it is the result the
  paper's headline figure uses), but it means the
  `r pct(mean(dat_raw[[outcome]]))`% figure is a property of one assay, not a
  robust fact about the households.
- **Not a problem: missing data.** All `r ncol(dat_raw)` columns are 100%
  complete, so the complete-case sample is the full `r nrow(dat_model)` rows with
  all `r n_events` events. There is no listwise-deletion story to tell here at
  all, which makes this a good dataset for teaching everything *except* missing
  data.
- **Not a problem: model stability.** The final nine-predictor model has a
  maximum standard error of
  `r fmt(max(summary(fit)$coefficients[-1, 2]), 2)`, a maximum Cook's distance of
  `r fmt(max(cooks.distance(fit)), 3)`, a maximum leverage of
  `r fmt(max(hatvalues(fit)), 3)` and fitted probabilities confined to
  `r fmt(min(fitted(fit)), 2)`-`r fmt(max(fitted(fit)), 2)`. Nothing is unstable;
  the trouble with this dataset is conceptual, not numerical.
- **Overall verdict: usable, with warnings.** The outcome supports a real
  Purpose 3 analysis with a defensible reduced predictor set, and it produces one
  clear finding (bushes), one clear age pattern and a set of instructive nulls.
  But Tjur's R-squared is only `r fmt(as.numeric(r2_tjur(fit)), 3)` and the AIC
  gain over the null model is `r fmt(AIC(null_fit) - AIC(fit), 1)`, so students
  should be warned in advance that a well-executed analysis here will find very
  little, and that the interesting write-up is about *why*.
