---
title: "Depressive symptoms (HADS-D) in adults with chronic disease"
subtitle: "Dataset 9: Patient activation for self-management in chronic disease"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "patient_activation.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "hads_depression"
predictors <- c("age_yrs", "sex", "education_level", "financial_distress",
                "disease", "charlson_index", "social_support_total")

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex = factor(sex, levels = c("male", "female"), labels = c("Male", "Female")),
    education_level = factor(education_level, levels = c("lower", "middle", "higher"),
                             labels = c("Lower", "Middle", "Higher")),
    financial_distress = factor(financial_distress, levels = c("none", "low", "high"),
                                labels = c("None", "Low", "High")),
    smoking_status = factor(smoking_status, levels = c("never", "former", "current"),
                            labels = c("Never", "Former", "Current")),
    disease = factor(disease,
                     levels = c("diabetes_t2", "copd", "heart_failure", "renal_disease"),
                     labels = c("Type 2 diabetes", "COPD", "Heart failure", "Renal disease")),
    disease_duration = factor(disease_duration,
                              levels = c("under_2_yrs", "2_to_5_yrs", "over_5_yrs"),
                              labels = c("<2 years", "2-5 years", ">5 years")),
    living_alone = factor(living_alone, levels = c(0, 1), labels = c("No", "Yes"))
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

# Missing questionnaire answers were coded 99 in the original deposit; the
# conversion to NA was already applied upstream, so no re-coding is done here.
# The only 99 anywhere in the file is a participant identifier (id == 99).
n99 <- dat_raw |>
  select(-id) |>
  select(where(is.numeric)) |>
  summarise(across(everything(), ~ sum(.x == 99, na.rm = TRUE))) |>
  rowSums()
```

# Introduction

Depressive symptoms are common in long-term physical illness and are associated
with worse self-care, more hospital use and higher mortality. These data come
from a cross-sectional survey of 1,154 Dutch adults treated in primary and
secondary care for type 2 diabetes, COPD, chronic heart failure or chronic renal
disease (Bos-Touwen et al., *PLoS ONE* 2015;10(5):e0126400), in which the
Hospital Anxiety and Depression Scale was administered alongside measures of
social and economic circumstances. Knowing which patients carry the heaviest
symptom burden helps target psychological support in chronic-disease clinics.

**Aim.** Among Dutch adults with one of four chronic conditions, we examined
which of age, sex, education, financial distress, index disease, comorbidity
burden and perceived social support are independently associated with HADS
depression score, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** Cross-sectional survey of adults with type 2 diabetes, COPD,
chronic heart failure or chronic renal disease recruited from Dutch general
practices and hospital outpatient clinics; `r nrow(dat_raw)` participants, all
measured on a single occasion.

**Variables.** The outcome is `hads_depression`, the HADS depression subscale
scored 0-21, where higher means more depressive symptoms and a score of 8 or
more is conventionally taken to suggest probable caseness. Predictors are
`age_yrs` (years), `sex` (reference: male), `education_level` (reference: lower,
the largest group), `financial_distress` (none / low / high, reference: none —
the clinically "unexposed" group and the largest), `disease` (reference: type 2
diabetes, the largest group), `charlson_index` (points; higher = more
comorbidity) and `social_support_total` (MSPSS total, 12-84; higher = more
perceived support).

**Sibling scale deliberately excluded.** `hads_anxiety` is the other subscale of
the same 14-item instrument, completed at the same sitting, and correlates 0.66
with the outcome. Entering it would give a variance-inflation factor of only
1.19 — collinearity is *not* the reason to leave it out — but adjusted R² would
jump from 0.19 to 0.49, and essentially all of that gain is the HADS predicting
itself. That is a restatement of the outcome, not an explanation of it, so
`hads_anxiety` is excluded. `sf12_mental` is excluded for the same reason: it
is a mental-health summary score that overlaps conceptually with HADS-D
(r = -0.71). `bmi` in this file is exactly `weight_kg / (height_cm/100)^2`, so
those three would never appear together; none is used here.
`n_comorbidities` correlates 0.78 with `charlson_index`, so only the Charlson
index is used.

**Statistical analysis.** Linear regression (`lm()`), justified because the
outcome is a continuous 0-21 score. We fitted a univariable model for each
predictor first, then a single multivariable model containing all of them.
Predictors were chosen in advance on subject-matter grounds, not by significance.
Significance threshold 0.05; estimates are reported with 95% confidence
intervals. Assumptions were checked with `performance::check_model()`; because
non-constant error variance was detected, heteroscedasticity-consistent (HC3)
standard errors were also computed with `lmtest::coeftest()` and
`sandwich::vcovHC()`. Analysis used `r R.version.string` with tidyverse,
gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped). Missing questionnaire answers were
coded 99 in the original deposit, but that conversion to `NA` has already been
applied in this CSV; a check across all numeric columns other than `id` found
`r n99` remaining values equal to 99, so no further re-coding was needed.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(missing = "no") |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

The HADS depression distribution is strongly right-skewed: the median is 4 with
an interquartile range of roughly 2 to 8, so most of this sample is well below
the caseness threshold while a minority score in the teens. Financial distress
is uncommon at the "high" level (fewer than one in ten), which makes that
comparison the least precise in the model.

## Exploratory figure

```{r explore, fig.cap="Figure 1. HADS depression score by self-reported financial distress, within index disease group. Boxes show medians and interquartile ranges; the dashed line is the caseness threshold of 8."}
ggplot(dat_model, aes(x = financial_distress, y = hads_depression)) +
  geom_hline(yintercept = 8, linetype = "dashed", colour = "grey50") +
  geom_boxplot(outlier.shape = NA, fill = "#a6bddb") +
  geom_jitter(width = 0.18, height = 0.15, alpha = 0.2, size = 0.8) +
  facet_wrap(~ disease) +
  labs(x = "Financial distress", y = "HADS depression score (0-21)")
```

Depressive symptoms rise steadily across the three financial-distress
categories, and the gradient is visible in all four disease groups, though the
"high" group is small everywhere.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv
```

Financial distress is the strongest crude predictor: compared with patients
reporting no financial distress, those reporting high distress score 3.69 HADS
points higher (95% CI 2.84 to 4.54, p < 0.001) and those reporting low distress
1.54 points higher (p < 0.001) — a difference large enough to move a typical
patient from well below the caseness threshold to at or above it. Perceived
social support runs the other way (-0.06 points per support point, 95% CI -0.075
to -0.046, p < 0.001; about -3.3 points across the interquartile range of
support). Comorbidity burden (+0.43 per Charlson point, 95% CI 0.30 to 0.56),
higher education (-2.09 versus lower education), age (+0.037 per year,
p < 0.001) and female sex (+0.55, p = 0.025) are all crudely associated, as are
COPD (+1.06, p < 0.001) and heart failure (+0.99, p = 0.004) relative to
diabetes.

## Why adjustment is needed

The crude associations here are heavily entangled. `education_level` is an
obvious confounder of the financial-distress association: less-educated patients
are both more likely to report financial distress and, independently, report
more depressive symptoms, so part of the crude distress gradient is really
educational disadvantage. `charlson_index` is a second: comorbidity burden rises
with age (r = 0.34), differs sharply between the four disease groups, and is
itself associated with depressive symptoms, so both the crude age association
and the crude COPD/heart-failure associations partly reflect how sick these
patients are overall rather than anything specific to age or diagnosis.
`social_support_total` is a third candidate, since patients who live in
financially precarious circumstances also report thinner support networks. And
because the four disease groups were pooled by design and differ in age, sex mix
and comorbidity, `disease` must be in the model for any other coefficient to be
read as a within-disease comparison.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv
```

```{r fit-stats}
gl <- glance(fit)
tibble(
  Statistic = c("R-squared", "Adjusted R-squared", "Residual SE", "N", "Model df"),
  Value = c(round(gl$r.squared, 3), round(gl$adj.r.squared, 3),
            round(gl$sigma, 2), nrow(dat_model), gl$df)
) |>
  gt() |>
  tab_header(title = "Model fit")
```

Conditional on all other predictors, high financial distress remains associated
with a 2.94-point higher HADS depression score than no distress (95% CI 2.11 to
3.78, p < 0.001) and low distress with a 1.10-point increase (95% CI 0.64 to
1.55, p < 0.001). Each additional point of perceived social support is
associated with a 0.054-point lower score (95% CI -0.068 to -0.040, p < 0.001),
and each Charlson point with a 0.39-point higher score (95% CI 0.25 to 0.53,
p < 0.001). Higher education retains a 1.39-point advantage over lower education
(95% CI -2.04 to -0.73, p < 0.001). Age (p = 0.070) and sex (p = 0.097) are no
longer associated once socioeconomic circumstances and comorbidity are held
constant. The model explains R² = `r round(gl$r.squared, 3)` of the variance
(adjusted R² = `r round(gl$adj.r.squared, 3)`).

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

Two estimates move a long way towards the null. High financial distress falls
from +3.69 to +2.94 and low distress from +1.54 to +1.10, so roughly a quarter
of the crude socioeconomic gradient in depressive symptoms is accounted for by
education, comorbidity and social support — but three-quarters of it survives.
The disease coefficients also shrink: COPD from +1.06 to +0.77 (still
significant, p = 0.006), heart failure from +0.99 to +0.46 (now
non-significant, p = 0.16) and renal disease from +0.47 to -0.24, i.e. the
heart-failure excess is largely explained by comorbidity burden. Two predictors
change significance: age (p < 0.001 crude to p = 0.070 adjusted) and sex
(p = 0.025 to p = 0.097), both attenuated by the Charlson index and by
socioeconomic variables respectively. Social support barely moves (-0.061 to
-0.054) and the Charlson index only modestly (0.43 to 0.39), which indicates
these two carry information not duplicated by the rest of the model. The overall
picture is classic confounding by socioeconomic position and comorbidity rather
than mediation or collinearity — all VIFs are below 1.5.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r diag-tests}
check_heteroscedasticity(fit)
check_normality(fit)
check_collinearity(fit)
```

The linearity panel is acceptable but the homogeneity-of-variance panel is not:
`check_heteroscedasticity()` rejects constant error variance (p < 0.001). This
is expected for a floor-bounded symptom score — patients predicted to have low
scores cannot vary much below zero, while patients predicted to have high scores
vary widely — and it is visible as a fan opening to the right. Normality of
residuals is also rejected (p < 0.001) because of the right skew and the floor
at zero. Collinearity is not an issue (all VIFs below 1.5, largest
`charlson_index` 1.43), and no observation exceeds the Cook's distance
threshold.

Because heteroscedasticity was flagged, the model was re-tested with HC3
heteroscedasticity-consistent standard errors:

```{r robust-se}
robust <- lmtest::coeftest(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
tibble(
  Term = rownames(robust),
  Estimate = round(robust[, 1], 3),
  `Model SE` = round(sqrt(diag(vcov(fit))), 3),
  `HC3 SE` = round(robust[, 2], 3),
  `HC3 p` = format.pval(robust[, 4], digits = 2, eps = 0.001)
) |>
  gt() |>
  tab_header(title = "HC3 robust standard errors",
             subtitle = "Compared with the model-based standard errors")
```

The robust standard errors are close to the model-based ones for every term: the
largest inflation is for `social_support_total` (HC3 SE 12% larger than the
model-based SE) and for the heart-failure contrast (8% larger), while the
education contrasts actually become 7-9% *tighter*. The financial-distress
standard errors are essentially unchanged (0.425 versus 0.425). No conclusion
changes: every predictor significant under model-based inference remains
significant under HC3, and age and sex remain non-significant. The
heteroscedasticity is therefore real but benign for these estimates.

# Discussion

Among Dutch adults with one of four chronic conditions, self-reported financial
distress is the characteristic most strongly and independently associated with
depressive symptoms, followed by low perceived social support, higher
comorbidity burden and lower education. Index disease adds little once these
are accounted for, and age and sex add nothing. Practically, this argues for
asking about money worries and social isolation in chronic-disease review
appointments, since these mark out patients at psychological risk far better
than the diagnosis does.

**Limitations.** The design is cross-sectional, so the direction of the
financial-distress and social-support associations cannot be established:
depression could plausibly cause a person to perceive their finances and their
support network more negatively, which is exactly the reverse of the causal
story above. Financial distress and social support are self-reported on the same
questionnaire occasion as the HADS, so common-method bias will inflate those
associations. The file has no disease-specific severity markers — no eGFR or
dialysis status, no NYHA class, no GOLD stage — so severity is proxied only by
the generic Charlson index and residual confounding by severity is likely.
The "high" financial-distress cell contains fewer than 100 people, so that
estimate is the least precise. Error variance is non-constant and residuals are
non-normal; HC3 standard errors were reported and changed nothing, but a count
or ordinal model would be a more principled fit for a floor-bounded 0-21 score.
Complete-case analysis dropped `r n_dropped` participants
(`r round(100*n_dropped/nrow(dat), 1)`%), and non-response to mood items is
plausibly related to mood itself. All estimates are associations, not effects.

**Future work.** A longitudinal design measuring financial circumstances and
mood repeatedly would separate the "poverty causes distress" and "distress
colours reporting" explanations.

# Feasibility notes

*Not part of a student report.*

- **Works well, with one assumption caveat.** Real socioeconomic gradients, a
  clean dose-response in `financial_distress`, and a good mix of predictors that
  survive and predictors that disappear after adjustment.
- **Complete-case N = `r nrow(dat_model)`** of `r nrow(dat)` (`r n_dropped`
  dropped, `r round(100*n_dropped/nrow(dat), 1)`% — the largest loss of the five
  outcomes here, driven mainly by `financial_distress` and the HADS items
  themselves).
- **Heteroscedasticity is genuinely flagged** (p < 0.001) and normality is
  rejected. This is a feature, not a bug, for teaching: it forces the student
  through `check_heteroscedasticity()` and HC3 robust SEs. The robust SEs are
  almost identical to the model-based ones, so the conclusions are stable and
  the student does not end up with a broken analysis.
- **The sibling-scale trap is live.** `hads_anxiety` gives VIF 1.19 — students
  will look at the VIF, see no collinearity, and keep it. Adjusted R² then goes
  from 0.186 to 0.485 and the "finding" is that the HADS predicts the HADS.
  `sf12_mental` (r = -0.71) is the same trap in a different costume. Both need
  to be explicitly ruled out in the brief.
- **No collinearity problem** in the chosen set: max VIF 1.43.
- **Sparse cell**: `financial_distress = high` has fewer than 100 participants;
  fine for a main effect, hopeless for an interaction.
