---
title: "Low patient activation (PAM level 1-2) in adults with chronic disease"
subtitle: "Dataset 9: Patient activation for self-management in chronic disease"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "patient_activation.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "low_activation"
predictors <- c("age_yrs", "sex", "education_level", "disease",
                "charlson_index", "ipq_total", "social_support_total")

dat <- dat_raw |>
  mutate(
    # Outcome: PAM activation levels 1 and 2 are "low activation" (the event).
    low_activation = factor(if_else(pam_level %in% c(1, 2), 1L, 0L),
                            levels = c(0, 1), labels = c("No", "Yes")),
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex = factor(sex, levels = c("male", "female"), labels = c("Male", "Female")),
    education_level = factor(education_level, levels = c("lower", "middle", "higher"),
                             labels = c("Lower", "Middle", "Higher")),
    financial_distress = factor(financial_distress, levels = c("none", "low", "high"),
                                labels = c("None", "Low", "High")),
    smoking_status = factor(smoking_status, levels = c("never", "former", "current"),
                            labels = c("Never", "Former", "Current")),
    disease = factor(disease,
                     levels = c("diabetes_t2", "copd", "heart_failure", "renal_disease"),
                     labels = c("Type 2 diabetes", "COPD", "Heart failure", "Renal disease")),
    disease_duration = factor(disease_duration,
                              levels = c("under_2_yrs", "2_to_5_yrs", "over_5_yrs"),
                              labels = c("<2 years", "2-5 years", ">5 years")),
    living_alone = factor(living_alone, levels = c(0, 1), labels = c("No", "Yes"))
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

n_events    <- sum(dat_model[[outcome]] == "Yes")
n_nonevents <- sum(dat_model[[outcome]] == "No")

# Missing questionnaire answers were coded 99 in the original deposit; the
# conversion to NA was already applied upstream, so no re-coding is done here.
# The only 99 anywhere in the file is a participant identifier (id == 99).
n99 <- dat_raw |>
  select(-id) |>
  select(where(is.numeric)) |>
  summarise(across(everything(), ~ sum(.x == 99, na.rm = TRUE))) |>
  rowSums()
```

# Introduction

The Patient Activation Measure sorts patients into four activation levels, and
levels 1 and 2 — patients who are disengaged or who lack confidence to act —
are the group that self-management support programmes are usually designed to
reach. These data come from a cross-sectional survey of 1,154 Dutch adults
treated in primary and secondary care for type 2 diabetes, COPD, chronic heart
failure or chronic renal disease (Bos-Touwen et al., *PLoS ONE*
2015;10(5):e0126400). Being able to recognise low-activation patients from
characteristics already known to the clinic would make that targeting practical.

**Aim.** Among Dutch adults with one of four chronic conditions, we examined
which of age, sex, education, index disease, comorbidity burden, illness
perceptions and perceived social support are independently associated with the
odds of low activation (PAM level 1 or 2), mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** Cross-sectional survey of adults with type 2 diabetes, COPD,
chronic heart failure or chronic renal disease recruited from Dutch general
practices and hospital outpatient clinics; `r nrow(dat_raw)` participants, all
measured on a single occasion.

**Variables.** The outcome `low_activation` was derived from `pam_level`:
activation levels 1 and 2 were coded as the event (`Yes`) and levels 3 and 4 as
the non-event (`No`). In the full file this gives 631 events out of
`r nrow(dat_raw)` (54.7%). Predictors are `age_yrs` (years), `sex` (reference:
male), `education_level` (reference: lower, the largest group), `disease`
(reference: type 2 diabetes, the largest group), `charlson_index` (points;
higher = more comorbidity), `ipq_total` (Brief Illness Perception Questionnaire
total, 0-80; higher = a more threatening view of one's illness) and
`social_support_total` (MSPSS total, 12-84; higher = more perceived support).

**Circularity avoided.** `pam_level` is a deterministic banding of `pam_score`,
and `low_activation` is a deterministic collapse of `pam_level`. Neither
`pam_score` nor `pam_level` can appear as a predictor here: either would
reproduce the outcome exactly and separate the data completely. `bmi` in this
file is exactly `weight_kg / (height_cm/100)^2`, so those three could never
appear together; none is used, since body size has no clear prior link to
activation. `n_comorbidities` correlates 0.78 with `charlson_index`, so only the
Charlson index is used.

**Statistical analysis.** Logistic regression (`glm(family = binomial)`),
justified because the outcome is binary. We fitted a univariable model for each
predictor first, then a single multivariable model containing all of them.
Predictors were chosen in advance on subject-matter grounds, not by significance.
Results are reported as odds ratios with 95% profile-likelihood confidence
intervals; significance threshold 0.05. Assumptions were checked with
`performance::check_model()`. Analysis used `r R.version.string` with tidyverse,
gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped). Missing questionnaire answers were
coded 99 in the original deposit, but that conversion to `NA` has already been
applied in this CSV; a check across all numeric columns other than `id` found
`r n99` remaining values equal to 99, so no further re-coding was needed.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(
    by = low_activation,
    missing = "no"
  ) |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics by activation status (Yes = PAM level 1-2).**")
```

`r n_events` of `r nrow(dat_model)` analysed participants
(`r round(100 * n_events / nrow(dat_model), 1)`%) had low activation, so the two
groups are close to balanced. Low-activation patients report more threatening
illness perceptions and less social support; the demographic and comorbidity
differences between the groups are small.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Proportion with low activation (PAM level 1-2) by index disease. Error bars are 95% Wilson confidence intervals; the dashed line is the overall proportion."}
prop_dat <- dat_model |>
  count(disease, low_activation) |>
  group_by(disease) |>
  mutate(total = sum(n)) |>
  filter(low_activation == "Yes") |>
  mutate(p = n / total,
         se = sqrt(p * (1 - p) / total),
         lo = pmax(0, p - 1.96 * se),
         hi = pmin(1, p + 1.96 * se)) |>
  ungroup()

ggplot(prop_dat, aes(x = disease, y = p)) +
  geom_hline(yintercept = n_events / nrow(dat_model),
             linetype = "dashed", colour = "grey50") +
  geom_col(fill = "#a6bddb", width = 0.6) +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.15) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.8)) +
  labs(x = "Index chronic condition", y = "Low activation (PAM level 1-2)")
```

The proportion with low activation rises from 48% in the diabetes group to 65%
in the renal disease group, with COPD and heart failure in between — a spread of
roughly 17 percentage points across the four pooled conditions.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = glm,
    method.args = list(family = binomial),
    exponentiate = TRUE
  )
uv
```

Illness perception is the strongest crude predictor: each additional IPQ point
multiplies the odds of low activation by 1.045 (95% CI 1.034 to 1.056,
p < 0.001), so a 20-point difference in illness perception corresponds to about
a 2.4-fold difference in odds. Perceived social support runs the other way
(OR 0.982 per point, 95% CI 0.975 to 0.990, p < 0.001). Comorbidity burden is
crudely associated (OR 1.14 per Charlson point, 95% CI 1.07 to 1.22, p < 0.001),
as is higher education, which halves the odds relative to lower education
(OR 0.65, 95% CI 0.45 to 0.92, p = 0.016). Crudely, renal disease patients have
about twice the odds of the diabetes group (OR 1.98, 95% CI 1.40 to 2.82,
p < 0.001) and heart failure patients about 1.6 times (95% CI 1.14 to 2.25,
p = 0.006). Age (p = 0.09) and sex (p = 0.39) show no crude association.

## Why adjustment is needed

The crude disease contrasts are the clearest case for adjustment.
`charlson_index` confounds them: the renal and heart-failure groups carry more
comorbidity than the diabetes group, and comorbidity is itself crudely
associated with low activation, so part of the doubled odds in renal patients is
simply that they are sicker overall. `ipq_total` confounds them in the same
direction: patients with the more visibly threatening conditions hold more
threatening illness beliefs, and illness beliefs are the strongest single
correlate of activation, so the disease contrasts partly measure differences in
illness construal. `education_level` is a third confounder, of the illness
perception and social support associations: better-educated patients report both
less threatening illness beliefs and higher activation, so the crude IPQ
association is partly educational. Because the four disease groups were pooled
by design, `disease` has to be in the model for anything else to be read as a
within-disease comparison.

## Multivariable model

```{r multivariable}
fit <- glm(reformulate(predictors, outcome), data = dat_model, family = binomial)

mv <- tbl_regression(fit, exponentiate = TRUE)
mv
```

```{r fit-stats}
n_params <- length(coef(fit))
tibble(
  Statistic = c("AIC", "Null deviance", "Residual deviance", "Tjur's R2",
                "N", "Events (low activation)", "Non-events",
                "Model parameters (incl. intercept)",
                "Events per parameter"),
  Value = c(round(AIC(fit), 1), round(fit$null.deviance, 1),
            round(fit$deviance, 1), round(performance::r2_tjur(fit), 3),
            nrow(dat_model), n_events, n_nonevents, n_params,
            round(min(n_events, n_nonevents) / (n_params - 1), 1))
) |>
  gt() |>
  tab_header(title = "Model fit")
```

The model has `r n_params - 1` estimated coefficients besides the intercept and
the smaller outcome class contains `r min(n_events, n_nonevents)` observations,
giving about `r round(min(n_events, n_nonevents) / (n_params - 1), 1)` events per
parameter — far above the conventional minimum of 10, so overfitting and sparse-
data bias are not concerns here.

Conditional on all other predictors, only two associations survive. Each extra
IPQ point multiplies the odds of low activation by 1.043 (95% CI 1.032 to 1.055,
p < 0.001), and each extra point of perceived social support multiplies them by
0.984 (95% CI 0.976 to 0.992, p < 0.001). Comorbidity burden is no longer
associated (OR 1.07 per point, 95% CI 0.98 to 1.16, p = 0.13), nor is higher
education (OR 0.79, 95% CI 0.54 to 1.16, p = 0.24), nor any disease contrast
(renal disease OR 1.37, 95% CI 0.92 to 2.05, p = 0.12), nor age or sex.
Discrimination is weak: Tjur's R² is
`r round(performance::r2_tjur(fit), 3)`, meaning the average predicted
probability differs between the two outcome groups by only about 10 percentage
points.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted odds ratios.**")
```

Three estimates move substantially towards the null and lose significance.
Renal disease falls from OR 1.98 to 1.37 (p < 0.001 to p = 0.12) and heart
failure from 1.60 to 1.01 (p = 0.006 to p = 0.96): the apparent activation
deficit in those groups is explained almost entirely by their higher comorbidity
and more threatening illness perceptions, not by the diagnosis. The Charlson
index itself moves from 1.14 to 1.07 (p < 0.001 to p = 0.13) — most of the crude
comorbidity association was really illness perception. Higher education moves
from 0.65 to 0.79 (p = 0.016 to p = 0.24), again attenuated by illness
perception and social support, which lie on the pathway between education and
activation as much as they confound it. Two estimates barely move: `ipq_total`
(1.045 to 1.043) and `social_support_total` (0.982 to 0.984), confirming that
these two are not proxies for the demographic or clinical variables. Age and sex
were null crudely and remain null. Nothing here is collinearity — all VIFs are
below 1.5.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r diag-tests}
check_collinearity(fit)
check_overdispersion(fit)
```

For a logistic model the relevant panels are binned residuals, collinearity and
influential observations; homogeneity of variance and normality of residuals do
not apply and are not produced. The binned residual plot has most points inside
the error bounds with no systematic curvature, so the linear-in-the-logit
assumption for `age_yrs`, `charlson_index`, `ipq_total` and
`social_support_total` is acceptable. Collinearity is low: the largest VIF is
`disease` at 1.45, then `charlson_index` at 1.37. No observation exceeds the
Cook's distance threshold, so no single participant drives the results.
`check_overdispersion()` gives a dispersion ratio of 1.00 (p = 0.93), as
expected for a binary outcome. Events per parameter is about
`r round(min(n_events, n_nonevents) / (n_params - 1), 1)`, comfortably adequate.
There is no separation: every predictor has a finite coefficient with a finite
confidence interval.

# Discussion

Among Dutch adults with one of four chronic conditions, only two characteristics
are independently associated with low patient activation: how threatening the
patient perceives their illness to be, and how much social support they feel
they have. Neither the diagnosis, nor comorbidity burden, nor education, nor age
or sex adds anything once those two are known. Practically, this means a clinic
cannot identify low-activation patients from routinely recorded demographics and
diagnoses — it has to ask them about their beliefs and their support. Even then
discrimination is weak (Tjur's R² = `r round(performance::r2_tjur(fit), 3)`), so
this is a description of associations rather than a usable screening rule.

**Limitations.** The design is cross-sectional, so nothing here shows that
changing illness beliefs would change activation; a patient who feels unable to
manage may simply come to see their illness as more threatening. IPQ, MSPSS and
PAM are all self-reported at the same sitting, so common-method bias inflates
the two surviving associations. Dichotomising the PAM at the level 2/3 boundary
discards the continuous information in `pam_score` and costs statistical power;
the cut point is conventional, not empirical. The file has no disease-specific
severity markers — no eGFR or dialysis status, no NYHA class, no GOLD stage — so
severity is proxied only by the generic Charlson index, and the attenuation of
the renal-disease and heart-failure contrasts should be read with that in mind.
Complete-case analysis dropped `r n_dropped` participants
(`r round(100*n_dropped/nrow(dat), 1)`%). All estimates are associations, not
effects.

**Future work.** A prospective study measuring illness perceptions before and
after a self-management intervention would test whether the association with
activation is modifiable.

# Feasibility notes

*Not part of a student report.*

- **Statistically the safest logistic outcome in this dataset.**
  `r n_events` events and `r n_nonevents` non-events out of
  `r nrow(dat_model)` complete cases — almost a 50/50 split, no sparse cells, no
  separation, `r round(min(n_events, n_nonevents) / (n_params - 1), 1)` events
  per parameter, max VIF 1.45, dispersion ratio 1.00, no influential points.
  Nothing breaks.
- **But it is a dichotomised copy of `pam_score`.** `low_activation` is derived
  from `pam_level`, which is a deterministic banding of `pam_score`. A student
  who analyses both outcomes is answering the same research question twice, once
  with the information thrown away. The logistic model finds *fewer* significant
  predictors than the linear one on the same predictor set — education and the
  Charlson index survive in the `pam_score` model and are lost here — which is
  exactly the power cost of dichotomising. If a student must do a logistic
  analysis, this is a workable outcome; if they have a free choice, `pam_score`
  is strictly better.
- **Circularity trap is severe.** Putting `pam_score` or `pam_level` on the
  right-hand side gives complete separation and an infinite coefficient. The
  brief must forbid this explicitly, because both variables sit in the file
  right next to the outcome and are the obvious things to grab.
- **Complete-case N = `r nrow(dat_model)`** of `r nrow(dat)` (`r n_dropped`
  dropped, `r round(100*n_dropped/nrow(dat), 1)`%).
- **Weak discrimination.** Tjur's R² = `r round(performance::r2_tjur(fit), 3)`.
  Only two of seven predictors are significant after adjustment, and the whole
  set of demographic and clinical variables contributes essentially nothing.
  Students expecting a strong predictive model should be warned.
- **Good confounding story**, though: three crude associations (renal disease,
  heart failure, Charlson index) are wiped out by adjustment, which makes the
  unadjusted-versus-adjusted comparison genuinely instructive.
