---
title: "Patient activation (PAM-13) in adults with chronic disease"
subtitle: "Dataset 9: Patient activation for self-management in chronic disease"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "patient_activation.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "pam_score"
predictors <- c("age_yrs", "sex", "education_level", "disease",
                "charlson_index", "ipq_total", "social_support_total")

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex = factor(sex, levels = c("male", "female"), labels = c("Male", "Female")),
    education_level = factor(education_level, levels = c("lower", "middle", "higher"),
                             labels = c("Lower", "Middle", "Higher")),
    financial_distress = factor(financial_distress, levels = c("none", "low", "high"),
                                labels = c("None", "Low", "High")),
    smoking_status = factor(smoking_status, levels = c("never", "former", "current"),
                            labels = c("Never", "Former", "Current")),
    disease = factor(disease,
                     levels = c("diabetes_t2", "copd", "heart_failure", "renal_disease"),
                     labels = c("Type 2 diabetes", "COPD", "Heart failure", "Renal disease")),
    disease_duration = factor(disease_duration,
                              levels = c("under_2_yrs", "2_to_5_yrs", "over_5_yrs"),
                              labels = c("<2 years", "2-5 years", ">5 years")),
    living_alone = factor(living_alone, levels = c(0, 1), labels = c("No", "Yes"))
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

# Missing questionnaire answers were coded 99 in the original deposit; the
# conversion to NA was already applied upstream, so no re-coding is done here.
# The only 99 anywhere in the file is a participant identifier (id == 99).
n99 <- dat_raw |>
  select(-id) |>
  select(where(is.numeric)) |>
  summarise(across(everything(), ~ sum(.x == 99, na.rm = TRUE))) |>
  rowSums()
```

# Introduction

The Patient Activation Measure (PAM-13) captures how far a patient has the
knowledge, skills and confidence to manage their own long-term condition.
These data come from a cross-sectional survey of 1,154 Dutch adults treated in
primary and secondary care for type 2 diabetes, COPD, chronic heart failure or
chronic renal disease (Bos-Touwen et al., *PLoS ONE* 2015;10(5):e0126400).
Activation matters because self-management support programmes are increasingly
targeted at patients on the basis of how activated they already are, so it is
useful to know which patient characteristics mark out the least activated.

**Aim.** Among Dutch adults with one of four chronic conditions, we examined
which of age, sex, education, index disease, comorbidity burden, illness
perceptions and perceived social support are independently associated with
PAM-13 score, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** Cross-sectional postal and online survey of adults with type 2
diabetes, COPD, chronic heart failure or chronic renal disease recruited from
Dutch general practices and hospital outpatient clinics; `r nrow(dat_raw)`
participants with complete PAM-13 data. All measurements were taken at a single
occasion.

**Variables.** The outcome is `pam_score`, the PAM-13 total score on a 0-100
scale, where higher scores mean greater ability and confidence to self-manage.
Predictors are `age_yrs` (years), `sex` (reference: male), `education_level`
(lower / middle / higher, reference: lower, the largest group),
`disease` (reference: type 2 diabetes, the largest group), `charlson_index`
(Charlson comorbidity index, points; higher = more comorbidity), `ipq_total`
(Brief Illness Perception Questionnaire total, 0-80; higher = a more
threatening view of one's illness) and `social_support_total` (Multidimensional
Scale of Perceived Social Support total, 12-84; higher = more perceived
support).

`pam_level` is a deterministic banding of `pam_score` itself and is therefore
never used as a predictor here. `bmi` in this file is exactly
`weight_kg / (height_cm/100)^2`, so at most one of those three would ever enter
a model; none is used here because body size has no clear prior link to
activation. `n_comorbidities` correlates 0.78 with `charlson_index`, so only the
Charlson index is used.

**Statistical analysis.** Linear regression (`lm()`), justified because the
outcome is continuous and roughly symmetric. We fitted a univariable model for
each predictor first, then a single multivariable model containing all of them.
Predictors were chosen in advance on subject-matter grounds, not by significance.
Significance threshold 0.05; estimates are reported with 95% confidence
intervals. Assumptions were checked with `performance::check_model()`. Analysis
used `r R.version.string` with tidyverse, gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped). Missing questionnaire answers were
coded 99 in the original deposit, but that conversion to `NA` has already been
applied in this CSV; a check across all numeric columns other than `id`
found `r n99` remaining values equal to 99, so no further re-coding was needed.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(missing = "no") |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

The analysed sample is elderly (median age around 70) and predominantly male,
with a median Charlson index of 3 and mostly lower or middle education. The
median PAM-13 score sits near 53, which is at the boundary between activation
levels 2 and 3, so a large share of this sample is in the lower half of the
activation range.

## Exploratory figure

```{r explore, fig.cap="Figure 1. PAM-13 score against illness-perception score, by index disease. Lines are linear smooths with 95% confidence bands."}
ggplot(dat_model, aes(x = ipq_total, y = pam_score)) +
  geom_point(alpha = 0.25, size = 1) +
  geom_smooth(method = "lm", formula = y ~ x, colour = "#2c7fb8") +
  facet_wrap(~ disease) +
  labs(x = "Illness perception (IPQ total, higher = more threatening)",
       y = "PAM-13 score (0-100)")
```

Patients who see their illness as more threatening report lower activation, and
the slope looks broadly similar in all four disease groups, so a single
common slope for `ipq_total` is a reasonable simplification.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv
```

Illness perception is the strongest crude predictor: each additional point on
the IPQ total is associated with a 0.24-point lower PAM-13 score
(95% CI -0.29 to -0.19, p < 0.001), so a 20-point difference in illness
perception corresponds to roughly 4.8 PAM points. Perceived social support runs
the other way (+0.13 PAM points per point of support, p < 0.001), higher
education is associated with a 4.2-point advantage over lower education
(p < 0.001), and both age (-0.07 per year, p = 0.024) and the Charlson index
(-0.79 per point, p < 0.001) are associated with lower activation. Crudely,
renal disease patients score 3.8 points below the diabetes group (p < 0.001).

## Why adjustment is needed

The crude associations mix several patient characteristics that travel
together in this sample. `charlson_index` is one clear confounder: it rises
with `age_yrs` (r = 0.34) and is itself associated with lower activation, so
part of the crude age effect is really comorbidity burden, and part of the
crude Charlson effect is really age. `disease` is a second: the four groups
differ substantially in age, sex mix and comorbidity — the renal disease group
in particular is older and more comorbid — so the crude renal-disease deficit
in activation partly reflects who ends up in that group rather than the
condition itself. `education_level` is a third: education is associated with
both illness perceptions (better-educated patients report less threatening
views of their illness) and with activation, so the crude IPQ slope is
partly an education effect. Because the four disease groups were pooled by
design, adjusting for `disease` is not optional here — it is what makes the
other coefficients interpretable as within-disease comparisons.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv
```

```{r fit-stats}
gl <- glance(fit)
tibble(
  Statistic = c("R-squared", "Adjusted R-squared", "Residual SE", "N", "Model df"),
  Value = c(round(gl$r.squared, 3), round(gl$adj.r.squared, 3),
            round(gl$sigma, 2), nrow(dat_model), gl$df)
) |>
  gt() |>
  tab_header(title = "Model fit")
```

Adjusting for all seven predictors, illness perception remains the dominant
association: conditional on age, sex, education, disease, comorbidity and social
support, each extra IPQ point is associated with a 0.22-point lower PAM score
(95% CI -0.27 to -0.17, p < 0.001). Perceived social support remains
independently associated in the opposite direction (+0.11 per point,
95% CI 0.07 to 0.15, p < 0.001), and higher education retains a 2.8-point
advantage over lower education (95% CI 0.97 to 4.64, p = 0.003). The Charlson
index is now only weakly associated (-0.41 per point, 95% CI -0.80 to -0.02,
p = 0.037) and age is no longer associated at all (p = 0.29). The model explains
R² = `r round(gl$r.squared, 3)` of the variance in PAM score
(adjusted R² = `r round(gl$adj.r.squared, 3)`), which is modest: most of what
drives activation in this sample is not captured by these variables.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

Three estimates move meaningfully. Age halves and loses significance (-0.067 to
-0.032, p = 0.024 to p = 0.29): the crude age association was largely
comorbidity, which age brings with it. The Charlson index also shrinks towards
the null by about half (-0.79 to -0.41) but stays just significant, for the
mirror-image reason plus overlap with disease group. Renal disease moves
strongly towards the null (-3.77 to -1.33, p < 0.001 to p = 0.17), so the
apparent activation deficit in renal patients is mostly explained by their
older age, higher comorbidity and more threatening illness perceptions rather
than by the diagnosis. COPD moves in the other direction, from a null crude
estimate (-0.43, p = 0.60) to a positive adjusted one (+1.98, p = 0.017): once
the heavier illness-perception burden of COPD patients is held constant, they
look slightly *more* activated than diabetes patients — a suppression effect
worth flagging rather than over-interpreting. Social support and illness
perception barely move (0.127 to 0.110; -0.241 to -0.223), which tells us those
two are not simply standing in for the demographic and clinical variables.
Education attenuates from +4.2 to +2.8 but keeps its significance.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r diag-tests}
check_heteroscedasticity(fit)
check_normality(fit)
check_collinearity(fit)
```

Linearity and the posterior-predictive panel look acceptable: the fitted-versus-
residual smooth is close to flat, and simulated data track the observed PAM
distribution. Homogeneity of variance passes (`check_heteroscedasticity()`
p = 0.093), so ordinary standard errors are reported and no robust-SE
re-estimation was needed. Normality of residuals is formally rejected
(p < 0.001) — the PAM distribution is slightly lumpy because the PAM-13 has a
finite set of achievable scores — but the departure is mild in the Q-Q plot and,
with n over 1,000, it has little practical effect on the confidence intervals.
Collinearity is not a problem: all VIFs are below 1.5 (largest: `disease` 1.46,
`charlson_index` 1.39). No observation exceeds the Cook's distance threshold.

# Discussion

Among Dutch adults with one of four chronic conditions, how threatening a
patient perceives their illness to be and how much social support they feel
they have are the characteristics most strongly and independently associated
with patient activation; higher education adds a further modest advantage.
Age, sex and index disease add very little once these are accounted for.
Practically, this points self-management support towards illness beliefs and
the patient's support network rather than towards demographic targeting — but
the model explains only about 13% of the variance, so activation is far from
predictable from routinely collected characteristics.

**Limitations.** The design is cross-sectional, so nothing here establishes
direction: a patient who feels unable to manage their illness may well come to
see it as more threatening, rather than the reverse, and the same circularity
applies to social support. The file contains no disease-specific severity
markers — no eGFR or dialysis status for renal disease, no NYHA class for heart
failure, no GOLD stage for COPD — so "severity" is represented only by the
generic Charlson index, and residual confounding by severity is likely,
especially for the `disease` coefficients. `ipq_total` and `pam_score` are both
self-reported on the same questionnaire occasion, so common-method bias will
inflate their association. Residual normality is formally violated. Complete-
case analysis dropped `r n_dropped` participants (`r round(100*n_dropped/nrow(dat), 1)`%),
and if those with missing questionnaire items are less activated the estimates
are mildly optimistic. All estimates are associations, not effects.

**Future work.** A longitudinal follow-up measuring illness perceptions and
activation at two time points would separate the two directions of the
IPQ-PAM association.

# Feasibility notes

*Not part of a student report.*

- **Works well.** This is the cleanest outcome in the dataset. `pam_score` has
  zero missing values, is continuous over a wide range (24.4-91.6), and has
  several genuinely different predictors with a mix of significant and
  non-significant results — good teaching material for Purpose 3.
- **Complete-case N = `r nrow(dat_model)`** of `r nrow(dat)` (`r n_dropped`
  dropped, `r round(100*n_dropped/nrow(dat), 1)`%). Losses come from the
  questionnaire scales, not the outcome.
- **No collinearity problem**: max VIF 1.46. But two traps students will fall
  into are live in this file: (i) `pam_level` is a deterministic banding of
  `pam_score` and gives an R² of 0.87 on its own (0.89 as a factor) if used as
  a predictor;
  (ii) `charlson_index` and `n_comorbidities` correlate at r = 0.78 and should
  not both be entered.
- **Diagnostics**: homoscedasticity OK (p = 0.09), residual normality rejected
  (p < 0.001) but visually mild; no influential points. No robust SEs needed.
- **Weak R² (0.137)**: acceptable for a survey outcome, but students expecting a
  "strong model" will be disappointed. Worth telling them in advance.
- **Suppression effect on COPD** (crude null to positive adjusted) is a nice
  discussion point but confuses weaker students.
