---
title: "Mental health-related quality of life (SF-12 MCS) in adults with chronic disease"
subtitle: "Dataset 9: Patient activation for self-management in chronic disease"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "patient_activation.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "sf12_mental"
predictors <- c("age_yrs", "sex", "living_alone", "financial_distress",
                "disease", "charlson_index", "social_support_total")

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex = factor(sex, levels = c("male", "female"), labels = c("Male", "Female")),
    education_level = factor(education_level, levels = c("lower", "middle", "higher"),
                             labels = c("Lower", "Middle", "Higher")),
    financial_distress = factor(financial_distress, levels = c("none", "low", "high"),
                                labels = c("None", "Low", "High")),
    smoking_status = factor(smoking_status, levels = c("never", "former", "current"),
                            labels = c("Never", "Former", "Current")),
    disease = factor(disease,
                     levels = c("diabetes_t2", "copd", "heart_failure", "renal_disease"),
                     labels = c("Type 2 diabetes", "COPD", "Heart failure", "Renal disease")),
    disease_duration = factor(disease_duration,
                              levels = c("under_2_yrs", "2_to_5_yrs", "over_5_yrs"),
                              labels = c("<2 years", "2-5 years", ">5 years")),
    living_alone = factor(living_alone, levels = c(0, 1), labels = c("No", "Yes"))
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

# Missing questionnaire answers were coded 99 in the original deposit; the
# conversion to NA was already applied upstream, so no re-coding is done here.
# The only 99 anywhere in the file is a participant identifier (id == 99).
n99 <- dat_raw |>
  select(-id) |>
  select(where(is.numeric)) |>
  summarise(across(everything(), ~ sum(.x == 99, na.rm = TRUE))) |>
  rowSums()
```

# Introduction

Mental health-related quality of life in long-term physical illness is shaped at
least as much by a patient's social and economic circumstances as by their
diagnosis. These data come from a cross-sectional survey of 1,154 Dutch adults
treated in primary and secondary care for type 2 diabetes, COPD, chronic heart
failure or chronic renal disease (Bos-Touwen et al., *PLoS ONE*
2015;10(5):e0126400), which recorded living arrangements, financial distress and
perceived social support alongside the SF-12. Identifying the social
circumstances most strongly associated with poor mental wellbeing helps decide
what a chronic-disease service should ask about beyond the illness itself.

**Aim.** Among Dutch adults with one of four chronic conditions, we examined
which of age, sex, living alone, financial distress, index disease, comorbidity
burden and perceived social support are independently associated with the SF-12
mental component score, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** Cross-sectional survey of adults with type 2 diabetes, COPD,
chronic heart failure or chronic renal disease recruited from Dutch general
practices and hospital outpatient clinics; `r nrow(dat_raw)` participants, all
measured on a single occasion.

**Variables.** The outcome is `sf12_mental`, the SF-12 mental component score on
a 0-100 scale, where higher means better mental wellbeing. Predictors are
`age_yrs` (years), `sex` (reference: male), `living_alone` (reference: no —
the clinically "unexposed" group and the larger one), `financial_distress`
(none / low / high, reference: none), `disease` (reference: type 2 diabetes,
the largest group), `charlson_index` (points; higher = more comorbidity) and
`social_support_total` (MSPSS total, 12-84; higher = more perceived support).

**Sibling and overlapping scales deliberately excluded.** `sf12_physical` is the
other component of the same 12-item instrument (r = 0.75 with the outcome). Its
variance-inflation factor if entered would be only 1.38, so collinearity is not
the argument against it — the argument is that adjusted R² would rise from 0.22
to 0.59 and the model would largely be the SF-12 predicting itself, which
restates the outcome rather than explaining it. `hads_depression` (r = -0.71) and
`hads_anxiety` (r = -0.68) are excluded for the same reason: adding
`hads_depression` alone takes adjusted R² to 0.56, and "mental health score
predicts mental health score" is not a research question. `bmi` in this file is
exactly `weight_kg / (height_cm/100)^2`, so those three could never appear
together; none is used here. `n_comorbidities` correlates 0.78 with
`charlson_index`, so only the Charlson index is used.

**Statistical analysis.** Linear regression (`lm()`), justified because the
outcome is a continuous 0-100 score. We fitted a univariable model for each
predictor first, then a single multivariable model containing all of them.
Predictors were chosen in advance on subject-matter grounds, not by significance.
Significance threshold 0.05; estimates are reported with 95% confidence
intervals. Assumptions were checked with `performance::check_model()`; because
non-constant error variance was detected, heteroscedasticity-consistent (HC3)
standard errors were also computed with `lmtest::coeftest()` and
`sandwich::vcovHC()`. Analysis used `r R.version.string` with tidyverse,
gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped). Missing questionnaire answers were
coded 99 in the original deposit, but that conversion to `NA` has already been
applied in this CSV; a check across all numeric columns other than `id` found
`r n99` remaining values equal to 99, so no further re-coding was needed.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(missing = "no") |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

Median SF-12 mental score is 70 with an interquartile range of roughly 48 to 85,
so mental wellbeing is on average better than physical wellbeing in this sample
but with a long left tail. About one in four participants lives alone, and
fewer than one in ten reports high financial distress.

## Exploratory figure

```{r explore, fig.cap="Figure 1. SF-12 mental component score against perceived social support, coloured by financial distress. Lines are linear smooths within financial-distress category."}
ggplot(dat_model, aes(x = social_support_total, y = sf12_mental,
                      colour = financial_distress)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  scale_colour_manual(values = c("None" = "#2c7fb8", "Low" = "#e6a000",
                                 "High" = "#c0392b")) +
  labs(x = "Perceived social support (MSPSS total)",
       y = "SF-12 mental component score (0-100)",
       colour = "Financial distress")
```

Mental wellbeing rises with perceived social support, and the three
financial-distress groups are separated by a large vertical gap at every level
of support — the two social variables appear to act largely independently rather
than one explaining the other.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv
```

Financial distress is the strongest crude predictor: compared with patients
reporting none, those with high distress score 20.96 SF-12 mental points lower
(95% CI -26.01 to -15.92, p < 0.001) and those with low distress 8.75 points
lower (p < 0.001) — an enormous gap on a 0-100 scale. Female sex is associated
with a 10.66-point deficit (95% CI -13.48 to -7.83, p < 0.001). Each point of
perceived social support is associated with a 0.31-point higher score (95% CI
0.22 to 0.40, p < 0.001), and each Charlson point with a 2.76-point lower score
(95% CI -3.52 to -2.00). Living alone is crudely associated with a 6.89-point
deficit (95% CI -9.98 to -3.79, p < 0.001). Crudely, heart failure (-9.70) and
COPD (-7.99) sit below diabetes, while renal disease does not differ (p = 0.16).

## Why adjustment is needed

`social_support_total` is the key confounder of the living-alone association:
people who live alone report substantially less perceived social support, and
perceived support is itself strongly associated with mental wellbeing, so the
crude living-alone deficit may be entirely a support effect wearing a different
label. `financial_distress` is a second confounder, of both the disease
contrasts and the sex contrast: financial strain is unevenly distributed across
the four disease groups and between men and women in this sample, and it is by
far the strongest correlate of the outcome. `charlson_index` is a third: it rises
with age (r = 0.34), differs between disease groups, and independently predicts
worse mental wellbeing, so both the crude age association and the crude
COPD/heart-failure deficits partly reflect overall illness burden. Because the
four disease groups were pooled by design and differ in age, sex and comorbidity,
`disease` must be in the model for the other coefficients to be read as
within-disease comparisons.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv
```

```{r fit-stats}
gl <- glance(fit)
tibble(
  Statistic = c("R-squared", "Adjusted R-squared", "Residual SE", "N", "Model df"),
  Value = c(round(gl$r.squared, 3), round(gl$adj.r.squared, 3),
            round(gl$sigma, 2), nrow(dat_model), gl$df)
) |>
  gt() |>
  tab_header(title = "Model fit")
```

Conditional on all other predictors, high financial distress remains associated
with an 18.32-point lower SF-12 mental score than no distress (95% CI -23.12 to
-13.53, p < 0.001) and low distress with a 6.66-point deficit (95% CI -9.30 to
-4.02, p < 0.001). Female sex retains a 10.78-point deficit (95% CI -13.43 to
-8.14, p < 0.001), each point of social support a 0.26-point advantage (95% CI
0.17 to 0.34, p < 0.001) and each Charlson point a 2.68-point deficit (95% CI
-3.50 to -1.85, p < 0.001). COPD (-6.55, 95% CI -9.79 to -3.30) and heart
failure (-6.04, 95% CI -9.86 to -2.23) remain below diabetes; renal disease does
not differ (p = 0.43). Living alone is now essentially null (-0.28, 95% CI -3.23
to 2.67, p = 0.85), and age is only borderline (-0.13 per year, p = 0.048). The
model explains R² = `r round(gl$r.squared, 3)` of the variance
(adjusted R² = `r round(gl$adj.r.squared, 3)`).

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

The most dramatic movement is `living_alone`, which collapses from a clearly
significant -6.89 to a flat -0.28 (p < 0.001 to p = 0.85). Living alone carries
no independent association with mental wellbeing once perceived social support
is in the model: the crude effect was confounding by support, not a housing
effect. Age also loses most of its association (-0.235 to -0.129, from p < 0.001
to borderline p = 0.048), attenuated by the Charlson index. High financial
distress attenuates from -20.96 to -18.32 — a shift, but only about an eighth of
the crude gap, so the socioeconomic gradient is largely irreducible by these
covariates. The disease coefficients move modestly and in opposite directions:
heart failure towards the null (-9.70 to -6.04) as comorbidity is accounted for,
COPD only slightly (-7.99 to -6.55), and renal disease away from the null but
still non-significant (-2.80 to +1.57). Two predictors barely move at all: female
sex (-10.66 to -10.78) and social support (0.307 to 0.256), which is the same
message Figure 1 gave — sex and support are not standing in for the economic
variables. Only `living_alone` and (marginally) `age_yrs` change significance
status. All VIFs are below 1.5, so none of this is collinearity.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r diag-tests}
check_heteroscedasticity(fit)
check_normality(fit)
check_collinearity(fit)
```

Linearity is acceptable. Homogeneity of variance fails:
`check_heteroscedasticity()` rejects constant error variance (p = 0.009). This
is expected from a ceiling-bounded score — patients predicted to be near 90 have
little room to vary upwards, while those predicted around 50 vary in both
directions — and it shows as a funnel narrowing to the right. Normality of
residuals is also rejected (p < 0.001), reflecting the left skew and the fact
that `sf12_mental` takes only 65 distinct values. Collinearity is not a problem
(all VIFs below 1.5, largest `charlson_index` 1.42), and no observation exceeds
the Cook's distance threshold.

Because heteroscedasticity was flagged, the model was re-tested with HC3
heteroscedasticity-consistent standard errors:

```{r robust-se}
robust <- lmtest::coeftest(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
tibble(
  Term = rownames(robust),
  Estimate = round(robust[, 1], 3),
  `Model SE` = round(sqrt(diag(vcov(fit))), 3),
  `HC3 SE` = round(robust[, 2], 3),
  `HC3 p` = format.pval(robust[, 4], digits = 2, eps = 0.001)
) |>
  gt() |>
  tab_header(title = "HC3 robust standard errors",
             subtitle = "Compared with the model-based standard errors")
```

The robust standard errors are within about 10% of the model-based ones for
every term (largest inflation: `social_support_total`, 9.5%; `age_yrs`, 5.4%),
and only one conclusion is affected: `age_yrs` moves from marginally significant
under model-based inference (p = 0.048) to non-significant under HC3
(p = 0.061). Since that estimate sits on the threshold either way, the honest
reading is that age has no reliable independent association here. Every other
conclusion is unchanged.

# Discussion

Among Dutch adults with one of four chronic conditions, financial distress and
female sex are the characteristics most strongly and independently associated
with poor mental health-related quality of life, followed by low perceived
social support and higher comorbidity burden. Living alone, which looks
important on its own, adds nothing once perceived support is accounted for —
what matters is whether patients feel supported, not whether there is someone in
the house. Practically, this argues for asking chronic-disease patients about
money and about felt support rather than recording household composition.

**Limitations.** The design is cross-sectional, and the direction of the
social-support and financial-distress associations is genuinely ambiguous:
poor mental wellbeing plausibly leads people to perceive both their finances and
their support networks more negatively. Support, distress and the SF-12 are all
self-reported at the same sitting, so common-method bias inflates those
associations. The file has no disease-specific severity markers — no eGFR or
dialysis status, no NYHA class, no GOLD stage — so severity is represented only
by the generic Charlson index and the `disease` coefficients carry residual
confounding. The high financial-distress cell has fewer than 100 participants,
so that large estimate is also the least precise. Error variance is non-constant
and residuals are non-normal; HC3 standard errors were reported and changed only
the borderline age result. Complete-case analysis dropped `r n_dropped`
participants (`r round(100*n_dropped/nrow(dat), 1)`%). All estimates are
associations, not effects.

**Future work.** Measuring support and mood repeatedly over a year would show
whether changes in perceived support precede changes in wellbeing or follow
them.

# Feasibility notes

*Not part of a student report.*

- **Works well.** Large, clearly interpretable effects; a memorable confounding
  story (`living_alone` collapsing once social support is adjusted for); and a
  legitimate reason to reach for robust standard errors.
- **Complete-case N = `r nrow(dat_model)`** of `r nrow(dat)` (`r n_dropped`
  dropped, `r round(100*n_dropped/nrow(dat), 1)`%).
- **Heteroscedasticity flagged** (p = 0.009) and normality rejected. HC3 SEs
  flip only the borderline `age_yrs` result (p = 0.048 to p = 0.061), which is a
  good, non-catastrophic illustration of why robust SEs matter.
- **The sibling / same-construct trap is worst for this outcome.** Three
  variables in the file are near-restatements of `sf12_mental`: `sf12_physical`
  (VIF only 1.38 but takes adjusted R² from 0.22 to 0.59), `hads_depression`
  (r = -0.71; adjusted R² to 0.56) and `hads_anxiety` (r = -0.68). None of them
  trips a VIF alarm, so students who screen only on VIF will keep them and
  report a meaningless "finding". This outcome needs the most explicit
  predictor guardrails of the five.
- **No collinearity problem** in the chosen set: max VIF 1.42.
- **Sparse cell**: `financial_distress = high` has fewer than 100 participants
  and produces the largest coefficient in the model; fine for a main effect,
  unusable for interactions.
