---
title: "Body mass index in adults aged 30 and over in Bharatpur, Nepal"
subtitle: "Dataset 4: Nepal hypertension and cardiovascular risk-factor survey"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)
library(sandwich)
library(lmtest)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "nepal_hypertension.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
#
# Deliberately EXCLUDED from the predictor set:
#   systolic_bp, diastolic_bp, hypertension, bp_awareness,
#   antihypertensive_drug - all of these sit DOWNSTREAM of body mass index.
#                           Adiposity raises blood pressure, not the reverse, so
#                           putting blood pressure on the right-hand side would
#                           be adjusting for a consequence of the outcome.
#   mets_min_per_week     - physical_activity is a deterministic cut of it at
#                           600 MET-min/week, so only one of the pair may enter.
#   cigarettes_per_week   - zero for every never-smoker, so it is smoking_status
#                           re-expressed on a different scale.
#   marital_status        - only 37 unmarried participants and largely a function
#                           of age and sex.
outcome    <- "bmi"
predictors <- c("sex", "age_group", "education", "occupation",
                "below_poverty_line", "smoking_status", "smokeless_tobacco",
                "alcohol_last_month", "fruit_veg_servings", "salt_g_per_day",
                "physical_activity", "family_history_htn")

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex = factor(sex, levels = c("female", "male"), labels = c("Female", "Male")),
    # Youngest band is the reference so the age contrasts read upwards.
    age_group = factor(age_group,
                       levels = c("30_39", "40_49", "50_59", "60_plus"),
                       labels = c("30-39", "40-49", "50-59", "60 or more")),
    # No or informal schooling is both the largest group and the most deprived.
    education = factor(education,
                       levels = c("none_or_informal", "primary", "secondary", "higher"),
                       labels = c("None or informal", "Primary", "Secondary", "Higher")),
    # Household work is the largest occupational group.
    occupation = factor(occupation,
                        levels = c("household_work", "agriculture_or_labour",
                                   "job_or_self_employed", "unemployed"),
                        labels = c("Household work", "Agriculture or labour",
                                   "Job or self-employed", "Unemployed")),
    # Never-smoker is the clinically unexposed reference.
    smoking_status = factor(smoking_status,
                            levels = c("never", "former", "current"),
                            labels = c("Never", "Former", "Current")),
    # The more active group is the reference, so "Low" reads as the risk category.
    physical_activity = factor(physical_activity,
                               levels = c("moderate_or_high", "low"),
                               labels = c("Moderate or high", "Low")),
    across(c(below_poverty_line, smokeless_tobacco, alcohol_last_month,
             family_history_htn, hypertension, bp_awareness, antihypertensive_drug),
           \(x) factor(x, levels = c(0, 1), labels = c("No", "Yes")))
  ) |>
  labelled::set_variable_labels(
    bmi                   = "Body mass index (kg/m2)",
    sex                   = "Sex",
    age_group             = "Age band (years)",
    education             = "Education",
    occupation            = "Main occupation",
    below_poverty_line    = "Below poverty line",
    smoking_status        = "Smoking status",
    smokeless_tobacco     = "Smokeless tobacco use",
    alcohol_last_month    = "Alcohol in the last month",
    fruit_veg_servings    = "Fruit and vegetable servings per day",
    salt_g_per_day        = "Salt intake (g/day)",
    physical_activity     = "Physical activity level",
    family_history_htn    = "Family history of hypertension"
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)
```

# Introduction

This is a community-based cross-sectional survey of adults aged 30 years and
over living in Bharatpur, Chitwan district, Nepal, run to estimate the
prevalence of hypertension and its cardiovascular risk factors (Dhungana and
colleagues, *PLOS ONE* 2016, e0185806). Height and weight were measured directly
by trained enumerators rather than self-reported, and the WHO STEPS instrument
supplied the behavioural and socio-economic information. Body mass index is
worth analysing in its own right in this setting: Nepal is in the middle of a
nutrition transition in which under- and over-nutrition coexist in the same
communities, and the classical Western socio-economic gradient in adiposity
(poorer people heavier) is often reversed in South Asia.

**Aim.** Among adults aged 30 and over in Bharatpur, we examined which of sex,
age band, education, occupation, poverty, smoking, smokeless tobacco, alcohol,
fruit and vegetable intake, salt intake, physical activity and family history of
hypertension are independently associated with body mass index, mutually
adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** A cross-sectional survey of `r nrow(dat_raw)` adults aged 30
years and over resident in Bharatpur Municipality, Chitwan, Nepal. The source
study used multistage cluster sampling: four wards were selected at random and
roughly 300 participants enrolled in each, with the sample size calculation
assuming a design effect of 1.5. **The ward identifier is not present in this
extract** (see Limitations). Each row is one participant.

**Variables.** The outcome is `bmi`, measured body mass index in kg/m2.
Predictors are sex (reference *Female*); age band in years (reference *30-39*;
age was recorded only in four bands, so a per-year effect cannot be estimated);
education (reference *None or informal*); main occupation (reference *Household
work*); family income per head below the poverty line (reference *No*); smoking
status (reference *Never*); current smokeless tobacco use (reference *No*); any
alcohol in the last month (reference *No*); fruit and vegetable servings per
day; estimated salt intake in g/day; physical activity level (reference
*Moderate or high*); and family history of hypertension (reference *No*).

Because body mass index is the outcome here it is of course not also a
predictor, and -- less obviously -- neither are `systolic_bp`, `diastolic_bp`,
`hypertension`, `bp_awareness` or `antihypertensive_drug`. Adiposity raises
blood pressure, so blood pressure is a *consequence* of the outcome; conditioning
on it would be adjusting for a mediator of everything upstream and would bias
every remaining coefficient. This is the mirror image of the systolic and
diastolic analyses in this catalogue, where `bmi` is a legitimate predictor.
Two further pairs of variables measure the same construct twice and only one of
each may enter a model: `physical_activity` is a deterministic cut of
`mets_min_per_week` at 600 MET-minutes per week (the "Low" group spans 0-600 and
"Moderate or high" spans 620-29,260), and the categorical version is used here
because it is the WHO STEPS classification the source paper reports and because
the continuous version is extremely right-skewed; and `cigarettes_per_week` is
zero for every never-smoker, so `smoking_status` is used instead because it also
separates former from never smokers.

**Statistical analysis.** Linear regression via `lm()`, justified because the
outcome is continuous. We fitted a univariable model for each predictor first,
then a single multivariable model containing all of them. Predictors were chosen
in advance on subject-matter grounds, not by significance. Significance
threshold 0.05; estimates are reported with 95% confidence intervals.
Assumptions were checked with `performance::check_model()` and
`performance::check_heteroscedasticity()`. Analysis used `r R.version.string`
with tidyverse, gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped). Every variable in this file is 100%
complete, so listwise deletion removes
`r if (n_dropped == 0) "no rows at all" else paste(n_dropped, "rows")` and there
is no missing-data story: no imputation, no complete-case selection bias.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(
    missing = "no"
  ) |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

```{r table-one-numbers}
bmi_mean <- mean(dat_model$bmi)
bmi_sd   <- sd(dat_model$bmi)
pct_over23 <- round(100 * mean(dat_model$bmi >= 23), 1)
pct_over275 <- round(100 * mean(dat_model$bmi >= 27.5), 1)
pct_under185 <- round(100 * mean(dat_model$bmi < 18.5), 1)
pct_female <- round(100 * mean(dat_model$sex == "Female"), 1)
```

Mean body mass index was `r round(bmi_mean, 1)` kg/m2 (SD `r round(bmi_sd, 1)`),
with a mildly right-skewed distribution running from 13.4 to 41.7 kg/m2. Using
the WHO Asian action points, `r pct_over23`% of the sample is at or above 23
kg/m2 and `r pct_over275`% at or above 27.5 kg/m2, while `r pct_under185`% are
below 18.5 kg/m2 -- the coexistence of under- and over-nutrition that
characterises the Nepali nutrition transition. The sample is `r pct_female`%
female, a consequence of daytime household recruitment in the source survey.

## Exploratory figure

```{r explore, fig.cap="Figure 3. Body mass index by age band and sex. Dashed lines mark the WHO Asian action points of 23 and 27.5 kg/m2."}
dat_model |>
  ggplot(aes(age_group, bmi, fill = sex)) +
  geom_violin(alpha = 0.5, position = position_dodge(width = 0.85),
              trim = TRUE, colour = NA) +
  geom_boxplot(width = 0.18, position = position_dodge(width = 0.85),
               outlier.alpha = 0.3, alpha = 0.9) +
  geom_hline(yintercept = c(23, 27.5), linetype = "dashed") +
  scale_fill_viridis_d(option = "D", end = 0.75) +
  labs(x = "Age band (years)", y = "Body mass index (kg/m2)", fill = "Sex",
       caption = "Dashed lines: WHO Asian action points at 23 and 27.5 kg/m2")
```

Body mass index is remarkably flat across the age bands until the oldest group,
where it drops by about a kilogram per square metre in both sexes, and women sit
slightly above men at every age -- so unlike blood pressure, adiposity in this
population is not primarily an age story.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv
```

```{r uv-numbers}
uv_fit <- function(v) lm(reformulate(v, outcome), data = dat_model) |> tidy(conf.int = TRUE)
u_sex   <- uv_fit("sex") |> filter(term == "sexMale")
u_age60 <- uv_fit("age_group") |> filter(term == "age_group60 or more")
u_pov   <- uv_fit("below_poverty_line") |> filter(term == "below_poverty_lineYes")
u_cur   <- uv_fit("smoking_status") |> filter(term == "smoking_statusCurrent")
u_form  <- uv_fit("smoking_status") |> filter(term == "smoking_statusFormer")
u_smok  <- uv_fit("smokeless_tobacco") |> filter(term == "smokeless_tobaccoYes")
u_fam   <- uv_fit("family_history_htn") |> filter(term == "family_history_htnYes")
u_fv    <- uv_fit("fruit_veg_servings") |> filter(term == "fruit_veg_servings")
u_alc   <- uv_fit("alcohol_last_month") |> filter(term == "alcohol_last_monthYes")
u_agri  <- uv_fit("occupation") |> filter(term == "occupationAgriculture or labour")
u_job   <- uv_fit("occupation") |> filter(term == "occupationJob or self-employed")
u_pa    <- uv_fit("physical_activity") |> filter(term == "physical_activityLow")
u_salt  <- uv_fit("salt_g_per_day") |> filter(term == "salt_g_per_day")
```

The crude associations are all small. The largest is **family history of
hypertension**, whose presence is associated with a `r round(u_fam$estimate, 2)`
kg/m2 higher body mass index (95% CI `r round(u_fam$conf.low, 2)` to
`r round(u_fam$conf.high, 2)`, p `r format.pval(u_fam$p.value, digits = 2, eps = 0.001)`)
-- plausible, since familial hypertension and familial adiposity share both
genes and household food environment. **Current smoking** is associated with a
`r round(u_cur$estimate, 2)` kg/m2 *lower* BMI (95% CI
`r round(u_cur$conf.low, 2)` to `r round(u_cur$conf.high, 2)`), and **smokeless
tobacco** with `r round(u_smok$estimate, 2)` kg/m2 (95% CI
`r round(u_smok$conf.low, 2)` to `r round(u_smok$conf.high, 2)`), both familiar
findings. **Poverty** is associated with `r round(u_pov$estimate, 2)` kg/m2
(95% CI `r round(u_pov$conf.low, 2)` to `r round(u_pov$conf.high, 2)`) -- the
South Asian direction, in which the poor are thinner, the opposite of the
high-income-country gradient. Men average `r round(u_sex$estimate, 2)` kg/m2
below women, and the oldest band averages `r round(u_age60$estimate, 2)` kg/m2
below the 30-39 band. Salt intake is flat (`r round(u_salt$estimate, 3)` kg/m2
per g/day) and, notably, so is **low physical activity**
(`r round(u_pa$estimate, 3)` kg/m2, 95% CI `r round(u_pa$conf.low, 2)` to
`r round(u_pa$conf.high, 2)`, p = `r round(u_pa$p.value, 2)`), which is not what
one would predict and is the first sign that the behavioural variables in this
file carry little information about adiposity.

## Why adjustment is needed

The crude estimates above are entangled through two variables in particular.
**Sex** is the main one, because every tobacco and alcohol variable in this
dataset is strongly sex-patterned while sex is itself associated with BMI:
`r round(100 * mean(dat_model$smoking_status[dat_model$sex == "Male"] == "Current"), 0)`%
of men are current smokers against
`r round(100 * mean(dat_model$smoking_status[dat_model$sex == "Female"] == "Current"), 0)`%
of women, and
`r round(100 * mean(dat_model$alcohol_last_month[dat_model$sex == "Male"] == "Yes"), 0)`%
of men drank in the last month against
`r round(100 * mean(dat_model$alcohol_last_month[dat_model$sex == "Female"] == "Yes"), 0)`%
of women, while men average `r abs(round(u_sex$estimate, 2))` kg/m2 less than
women. A crude "smoking lowers BMI" coefficient therefore contains a large "being
male lowers BMI" component. **Occupation** is the second, and it is confounded
with sex in an extreme way in this sample: 595 of the 659 people doing household
work are women, whereas paid employment is `r round(100 * mean(dat_model$sex[dat_model$occupation == "Job or self-employed"] == "Male"), 0)`%
male. Occupation is also a proxy for both physical exertion at work and
household income, both of which act on BMI. **Age band** plays a third, smaller
role: BMI falls in the oldest band while smoking cessation, low education and
low physical activity all rise with age, so crude coefficients for those
variables absorb some of the age-related weight loss. Mutual adjustment is
needed to tell these apart.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv
```

```{r fit-stats}
gl <- glance(fit)
tibble(
  Statistic = c("R-squared", "Adjusted R-squared", "Residual SD (kg/m2)",
                "Observations", "Model parameters (incl. intercept)",
                "Observations per parameter"),
  Value = c(round(gl$r.squared, 3), round(gl$adj.r.squared, 3),
            round(sigma(fit), 2), nobs(fit), length(coef(fit)),
            round(nobs(fit) / length(coef(fit)), 1))
) |>
  gt() |>
  tab_header(title = "Model fit")
```

```{r mv-numbers}
tid <- tidy(fit, conf.int = TRUE)
g <- function(t) tid |> filter(term == t)
m_sex   <- g("sexMale")
m_age60 <- g("age_group60 or more")
m_pov   <- g("below_poverty_lineYes")
m_cur   <- g("smoking_statusCurrent")
m_form  <- g("smoking_statusFormer")
m_smok  <- g("smokeless_tobaccoYes")
m_fam   <- g("family_history_htnYes")
m_fv    <- g("fruit_veg_servings")
m_alc   <- g("alcohol_last_monthYes")
m_agri  <- g("occupationAgriculture or labour")
m_job   <- g("occupationJob or self-employed")
m_pa    <- g("physical_activityLow")
m_salt  <- g("salt_g_per_day")
```

The model explains `r round(100 * gl$r.squared, 1)`% of the variance in body
mass index (adjusted R-squared `r round(gl$adj.r.squared, 3)`), so more than 90%
of the variation between these adults is unexplained by twelve
socio-demographic and behavioural characteristics. That is the headline result
and it is a real one, not a technical failure: this predictor set simply does
not contain the determinants of adiposity. There are `r nobs(fit)` observations
for `r length(coef(fit))` parameters
(`r round(nobs(fit) / length(coef(fit)), 1)` per parameter), so the weak fit is
not an overfitting artefact.

Conditional on the other eleven predictors, **family history of hypertension**
remains the largest effect at `r round(m_fam$estimate, 2)` kg/m2 (95% CI
`r round(m_fam$conf.low, 2)` to `r round(m_fam$conf.high, 2)`, p
`r format.pval(m_fam$p.value, digits = 2, eps = 0.001)`). **Current smokers**
average `r round(m_cur$estimate, 2)` kg/m2 below never-smokers (95% CI
`r round(m_cur$conf.low, 2)` to `r round(m_cur$conf.high, 2)`), **smokeless
tobacco users** `r round(m_smok$estimate, 2)` kg/m2 below non-users (95% CI
`r round(m_smok$conf.low, 2)` to `r round(m_smok$conf.high, 2)`), and people
**below the poverty line** `r round(m_pov$estimate, 2)` kg/m2 below those above
it (95% CI `r round(m_pov$conf.low, 2)` to `r round(m_pov$conf.high, 2)`).
Participants in **paid employment or self-employment** average
`r round(m_job$estimate, 2)` kg/m2 above those doing household work (95% CI
`r round(m_job$conf.low, 2)` to `r round(m_job$conf.high, 2)`), and the **oldest
band** `r round(m_age60$estimate, 2)` kg/m2 below the youngest (95% CI
`r round(m_age60$conf.low, 2)` to `r round(m_age60$conf.high, 2)`). Every one of
these is under one and a half kg/m2 -- roughly 4 kg for an adult of average
height -- so even the "significant" findings are modest.

Physical activity is the striking null: low activity is associated with
`r round(m_pa$estimate, 2)` kg/m2 (95% CI `r round(m_pa$conf.low, 2)` to
`r round(m_pa$conf.high, 2)`, p = `r round(m_pa$p.value, 2)`), pointing in the
wrong direction of nothing at all. Salt (`r round(m_salt$estimate, 3)` kg/m2 per
g/day) and education (all categories p > 0.2) are equally flat.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

Three predictors move meaningfully and all three move toward the null. **Sex**
goes from `r round(u_sex$estimate, 2)` kg/m2 crude (95% CI
`r round(u_sex$conf.low, 2)` to `r round(u_sex$conf.high, 2)`, p =
`r round(u_sex$p.value, 3)`) to `r round(m_sex$estimate, 2)` kg/m2 adjusted
(95% CI `r round(m_sex$conf.low, 2)` to `r round(m_sex$conf.high, 2)`, p =
`r round(m_sex$p.value, 2)`) and loses significance: about a third of the crude
male--female difference in BMI is really the fact that men smoke, chew tobacco
and work outside the home. **Smokeless tobacco** shrinks from
`r round(u_smok$estimate, 2)` to `r round(m_smok$estimate, 2)` kg/m2, a
`r round(100 * (1 - m_smok$estimate / u_smok$estimate))`% reduction, and only
just keeps significance (p = `r round(m_smok$p.value, 3)`), for the same reason.
**Fruit and vegetable servings** moves from `r round(u_fv$estimate, 3)` kg/m2
per serving crude (p = `r round(u_fv$p.value, 3)`) to `r round(m_fv$estimate, 3)`
adjusted (p = `r round(m_fv$p.value, 2)`), crossing from significant to not.
**Occupation in agriculture or labour** likewise falls from
`r round(u_agri$estimate, 2)` kg/m2 (p = `r round(u_agri$p.value, 3)`) to
`r round(m_agri$estimate, 2)` (p = `r round(m_agri$p.value, 2)`), confirming that
the crude occupational gradient was mostly sex.

One predictor flips sign. **Alcohol in the last month** is
`r round(u_alc$estimate, 2)` kg/m2 crude (95% CI `r round(u_alc$conf.low, 2)` to
`r round(u_alc$conf.high, 2)`, p = `r round(u_alc$p.value, 2)`) but
`r round(m_alc$estimate, 2)` kg/m2 adjusted (95% CI
`r round(m_alc$conf.low, 2)` to `r round(m_alc$conf.high, 2)`, p =
`r round(m_alc$p.value, 2)`) -- neither estimate significant, but the direction
reverses because drinkers are overwhelmingly male and men are lighter; once sex
is held fixed, alcohol's own calorie contribution shows through. This is textbook
negative confounding, and worth pointing out precisely because neither estimate
is significant: sign changes are not confined to significant coefficients.

Two predictors barely move. **Family history of hypertension** goes from
`r round(u_fam$estimate, 2)` to `r round(m_fam$estimate, 2)` kg/m2 (a shift of
under 5%) and stays highly significant, which is what one expects of a variable
essentially unrelated to the others in the model. **Low physical activity** goes
from `r round(u_pa$estimate, 3)` to `r round(m_pa$estimate, 3)` kg/m2 and is null
throughout. The maximum variance inflation factor is
`r round(max(check_collinearity(fit)$VIF), 2)`, so none of these movements
reflect collinearity; they are confounding by sex, occupation and age.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r het}
het <- check_heteroscedasticity(fit)
het_p <- as.numeric(het)
het_ok <- het_p >= 0.05
rob <- coeftest(fit, vcov = vcovHC(fit, type = "HC3"))
keep <- c("sexMale", "age_group60 or more", "below_poverty_lineYes",
          "smoking_statusCurrent", "smokeless_tobaccoYes", "family_history_htnYes")
tibble(
  Term = keep,
  Estimate = round(rob[keep, "Estimate"], 3),
  `SE (OLS)` = round(summary(fit)$coefficients[keep, "Std. Error"], 3),
  `SE (HC3)` = round(rob[keep, "Std. Error"], 3),
  `p (OLS)` = format.pval(summary(fit)$coefficients[keep, "Pr(>|t|)"], digits = 2, eps = 0.001),
  `p (HC3)` = format.pval(rob[keep, "Pr(>|t|)"], digits = 2, eps = 0.001)
) |>
  gt() |>
  tab_header(title = "HC3 standard errors, shown for completeness",
             subtitle = "Heteroscedasticity was NOT flagged for this model")
```

The **homogeneity of variance** panel is the one that matters most for a linear
model and here it passes: `check_heteroscedasticity()` returns p =
`r round(het_p, 3)`, so constant error variance is a defensible assumption and
robust standard errors are not required. They are shown above anyway for
comparison and, as expected when the assumption holds, they are almost identical
to the ordinary ones. The **posterior-predictive check** shows the model
tracking the centre of the BMI distribution while under-predicting the right
tail, because BMI is right-skewed and a linear model with additive terms
generates a symmetric predictive distribution; the **normality of residuals**
panel says the same thing, with a clear upward departure in the upper QQ tail
(`check_normality()` flags it at p < 0.001). Neither is a serious threat with
`r nobs(fit)` observations, but it does mean prediction intervals for individuals
would be wrong even though the coefficient estimates are fine. The **linearity**
panel is flat. **Collinearity** is unproblematic despite the strong sex--occupation
association: maximum VIF `r round(max(check_collinearity(fit)$VIF), 2)`.
**Influential observations** shows the handful of participants with BMI above 40
approaching but not crossing the Cook's distance contour. The real diagnostic
problem is not in these panels at all: it is that adjusted R-squared is
`r round(gl$adj.r.squared, 3)`, so the model is a poor description of adiposity
whatever its assumptions.

# Discussion

Body mass index in these Bharatpur adults is barely predicted by the
socio-demographic and behavioural variables collected in a STEPS survey. The
associations that survive adjustment are all small and mostly point the way
South Asian nutrition-transition literature would predict: people below the
poverty line are lighter, not heavier (`r round(m_pov$estimate, 2)` kg/m2);
tobacco users of both kinds are lighter; people in paid employment are heavier
than those doing household work; and BMI declines in the oldest band. Family
history of hypertension carries the largest single association at
`r round(m_fam$estimate, 2)` kg/m2, consistent with shared familial and
household determinants of cardiometabolic risk. Physical activity, salt, fruit
and vegetables and education are all null. In practical terms, the message is
that adiposity in this population is not readily explained by, or targeted
through, the behaviours this instrument measures.

**Limitations.** The dominant limitation is the **weak model**: adjusted
R-squared `r round(gl$adj.r.squared, 3)` means that essentially nothing here
explains BMI, and the most important determinants -- total energy intake, dietary
composition beyond fruit and vegetable servings, sedentary time, parity and
menopausal status in women, genetics -- are simply not measured. Second, the
**cluster sampling cannot be accounted for**: only four wards were randomly
selected with around 300 participants each and a design effect of 1.5 assumed in
the sample size calculation, but this extract contains no ward identifier, so all
confidence intervals above are too narrow (plausibly by around 20%) and cannot be
corrected. Estimates from four clusters cannot be generalised to Nepal or even to
all of Bharatpur. Third, **age is available only in four bands**, so no per-year
change in BMI can be estimated and the top band is open-ended. Fourth, this is
cross-sectional, so the smoking and tobacco coefficients cannot be read causally:
low body weight may lead to continued smoking as much as the reverse, and people
who gain weight may quit. Fifth, `physical_activity` is a crude binary derived
from a self-reported MET score with well-known over-reporting problems, which may
explain its complete lack of association. Sixth, BMI itself does not distinguish
fat from lean mass and performs poorly in South Asian populations, where
cardiometabolic risk begins at lower BMI values -- waist circumference would have
been a better outcome but is not in this extract. On the credit side, **the file
has no missing data at all** (all `r nrow(dat_raw)` rows complete on all
`r ncol(dat_raw)` variables), and height and weight were measured rather than
self-reported, so the outcome itself is of good quality.

**Future work.** Repeat with waist circumference or waist-to-height ratio as the
outcome, add dietary energy intake and sedentary time, and refit with the ward
identifier as a random effect.

# Feasibility notes

*Not part of a student report.*

- **Technically clean but substantively thin.** Complete data, a well-behaved
  continuous outcome, homoscedastic residuals (p = `r round(het_p, 3)`), maximum
  VIF `r round(max(check_collinearity(fit)$VIF), 2)`, `r nobs(fit)` rows over
  `r length(coef(fit))` parameters. The problem is that adjusted R-squared is
  `r round(gl$adj.r.squared, 3)` -- this predictor set does not explain BMI.
- **Set expectations in the brief.** A student who equates a good report with a
  high R-squared will conclude they have done something wrong. The teaching
  value is the opposite: this is what an honest, well-conducted, largely null
  Purpose 3 analysis looks like.
- **BMI is a predictor in the other files in this dataset and the outcome here.**
  It must be dropped from the predictor set in this file, and blood pressure
  (`systolic_bp`, `diastolic_bp`, `hypertension`, `bp_awareness`,
  `antihypertensive_drug`) must *not* be added as predictors: BP is downstream
  of adiposity, so conditioning on it adjusts away the very pathway of interest.
  This is the single most likely student error in this file, and it is seductive
  because adding systolic BP would roughly triple R-squared.
- **Deterministic twins.** `physical_activity` is exactly
  `mets_min_per_week <= 600`; `cigarettes_per_week` is zero for all
  never-smokers. Never both.
- **The interesting stories are counter-intuitive.** Poverty is associated with
  *lower* BMI, and physical activity with nothing at all. Both are correct for
  this setting but will read as errors to students trained on Western data, so
  they need framing.
- **One genuine sign flip** (alcohol, `r round(u_alc$estimate, 2)` crude to
  `r round(m_alc$estimate, 2)` adjusted) makes this a good file for teaching
  negative confounding, though neither estimate is significant.
- **No ward identifier**, so the assumed design effect of 1.5 cannot be
  honoured; standard errors are too narrow throughout.
- **Age in bands only**, so no per-year BMI trajectory is estimable.
- **Residual non-normality** from the right tail of BMI is flagged and will
  worry students unnecessarily; it does not threaten the coefficients at this
  sample size.
