---
title: "Leg-pain intensity in primary-care patients with sciatica"
subtitle: "Dataset 3: Sciatica in primary care"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "sciatica_primary_care.csv"),
                    show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "leg_pain_intensity"
predictors <- c("age_yrs", "sex", "back_pain_intensity", "leg_pain_duration",
                "below_knee_pain", "leg_pain_worse", "neural_tension_positive",
                "mri_nerve_root_compression")

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex = factor(sex, levels = c("female", "male"), labels = c("Female", "Male")),
    back_pain_duration = factor(back_pain_duration,
                                levels = c("under_6_weeks", "6_to_12_weeks", "over_12_weeks"),
                                labels = c("<6 weeks", "6-12 weeks", ">12 weeks")),
    leg_pain_duration = factor(leg_pain_duration,
                               levels = c("under_6_weeks", "6_to_12_weeks", "over_12_weeks"),
                               labels = c("<6 weeks", "6-12 weeks", ">12 weeks")),
    comorbidities = factor(comorbidities,
                           levels = c("none", "one", "two_or_more"),
                           labels = c("None", "One", "Two or more")),
    general_health = factor(general_health,
                            levels = c("good", "fair", "poor"),
                            labels = c("Good", "Fair", "Poor")),
    across(c(leg_pain_worse, cough_sneeze_positive, subjective_sensory_changes,
             below_knee_pain, neural_tension_positive, neurological_deficit,
             mri_nerve_root_compression, clinician_sciatica),
           ~ factor(.x, levels = c(0, 1), labels = c("No", "Yes")))
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

# `id` is a row label with 395 distinct values and is never entered into a model.
miss_tbl <- dat |>
  summarise(across(everything(), ~ sum(is.na(.x)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  filter(n_missing > 0) |>
  arrange(desc(n_missing))

# `leg_pain_worse` is a self-reported comparison, not an arithmetic function of the
# two numeric pain scores; this cross-tabulation documents that it is not
# deterministic and is therefore safe to enter as a predictor.
agree <- mean(
  (dat$leg_pain_worse == "Yes") ==
    (dat$leg_pain_intensity > dat$back_pain_intensity)
)
```

# Introduction

In sciatica it is leg pain, not back pain, that defines the syndrome and that
patients most want relieved; leg-pain intensity is the primary outcome in most
sciatica trials and one of the few things a GP can act on at the first visit.
These data come from a cross-sectional diagnostic study of `r nrow(dat_raw)`
adults presenting to Dutch general practices with a new episode of sciatica, each
of whom had a standardised history, a physical examination and a lumbar MRI
(Verwoerd et al.; sciatica in general practice, MRI substudy). Knowing which
history items and examination findings mark out the patients in most leg pain
tells us whether severity is something the consultation can identify or whether
it is essentially unexplained by the routine assessment.

**Aim.** Among primary-care patients presenting with sciatica, we examined which
of age, sex, concurrent back-pain intensity, duration of the leg-pain episode,
pain radiating below the knee, leg pain being worse than back pain, a positive
neural-tension test and nerve-root compression on MRI are independently
associated with leg-pain intensity, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** Cross-sectional baseline data from a primary-care sciatica
cohort: `r nrow(dat_raw)` adults consulting Dutch general practitioners with a
new episode of sciatica, all assessed at a single occasion. Age ranged from
`r round(min(dat$age_yrs), 0)` to `r round(max(dat$age_yrs), 0)` years.

**Variables.** The outcome is `leg_pain_intensity`, average leg-pain intensity
on a 0-10 numerical rating scale, where higher means worse pain. Predictors are
`age_yrs` (years), `sex` (reference: female, the larger group),
`back_pain_intensity` (0-10 numerical rating scale), `leg_pain_duration`
(reference: under 6 weeks, the largest group), `below_knee_pain` (pain extends
below the knee; reference: no), `leg_pain_worse` (patient reports leg pain worse
than back pain; reference: no), `neural_tension_positive` (at least one positive
neural-tension test; reference: no) and `mri_nerve_root_compression` (nerve-root
compression visible on MRI; reference: no).

**Two predictor choices need justification.** First, `back_pain_intensity` is
entered even though it is a pain measure like the outcome. They are related but
distinct: back pain and leg pain are separately elicited on separate scales, they
correlate r = `r round(cor(dat$back_pain_intensity, dat$leg_pain_intensity), 2)`
rather than 1, and `r round(100 * mean(dat$back_pain_intensity > dat$leg_pain_intensity), 0)`%
of patients report *more* back pain than leg pain. Including it answers a
sharper question -- which findings mark out severe leg pain over and above a
patient's general pain level -- and excluding it would leave every other
coefficient confounded by overall pain reporting. Second, `leg_pain_worse` is a
patient's own comparison of the two sites, not an arithmetic recoding of the two
scores: it agrees with the sign of
`leg_pain_intensity - back_pain_intensity` in only
`r round(100 * agree, 0)`% of patients, so it carries independent information
and is not a deterministic function of the outcome.

`id` is a participant label and is excluded from every model.
`clinician_sciatica` -- the GP's high-confidence diagnosis -- is deliberately
**not** used. It is a summary judgement built from the same examination findings
that appear as predictors here, correlating 0.61 with `neural_tension_positive`
and 0.54 with `below_knee_pain`; putting it in the same model as those signs
would make their coefficients uninterpretable.

**Statistical analysis.** Linear regression (`lm()`), justified because the
outcome is a numerical rating scale spread across its full 0-10 range and close
to symmetric. We fitted a univariable model for each predictor first, then a
single multivariable model containing all of them. Predictors were chosen in
advance on subject-matter grounds, not by significance. Significance threshold
0.05; estimates are reported with 95% confidence intervals. Assumptions were
checked with `performance::check_model()` plus formal tests of
heteroscedasticity, residual normality and collinearity; HC3 robust standard
errors via `lmtest::coeftest()` with `sandwich::vcovHC()` are reported as a
sensitivity check because the outcome is bounded. Analysis used
`r R.version.string` with tidyverse, gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped,
`r round(100 * n_dropped / nrow(dat), 1)`%). Missingness in the file is confined
to
`r paste(sprintf("%s (%d)", miss_tbl$variable, miss_tbl$n_missing), collapse = ", ")`.
The outcome is complete; essentially all the loss here comes from
`leg_pain_duration`, the single most incomplete column in the dataset. This is
the largest complete-case loss of the three outcomes analysed for this dataset,
and it is entirely a consequence of including that one predictor.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(missing = "no") |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

The analysed sample has a median leg-pain intensity of
`r round(median(dat_model[[outcome]]), 1)`/10 with the interquartile range
spanning `r round(quantile(dat_model[[outcome]], 0.25), 1)` to
`r round(quantile(dat_model[[outcome]], 0.75), 1)`, so leg pain is moderate to
severe in most patients but far from uniformly maximal. Clinical signs are
common: `r round(100 * mean(dat_model$below_knee_pain == "Yes"), 0)`% report pain
below the knee, `r round(100 * mean(dat_model$neural_tension_positive == "Yes"), 0)`%
have a positive neural-tension test and
`r round(100 * mean(dat_model$mri_nerve_root_compression == "Yes"), 0)`% have
nerve-root compression on MRI. Only about half say their leg pain is worse than
their back pain, which is a reminder that "sciatica" in primary care is often a
mixed back-and-leg presentation.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Leg-pain intensity against back-pain intensity, split by whether the patient reports leg pain as worse than back pain. The dashed line is equality; the solid line is a linear smooth with a 95% confidence band. Points are jittered because both scales are recorded in thirds of a point."}
ggplot(dat_model, aes(x = back_pain_intensity, y = .data[[outcome]])) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "grey55") +
  geom_jitter(width = 0.12, height = 0.12, alpha = 0.3, size = 1.1) +
  geom_smooth(method = "lm", formula = y ~ x, colour = "#2c7fb8") +
  facet_wrap(~ leg_pain_worse,
             labeller = labeller(leg_pain_worse = c(
               "No" = "Leg pain NOT worse than back pain",
               "Yes" = "Leg pain worse than back pain"))) +
  labs(x = "Back-pain intensity (0-10 NRS)",
       y = "Leg-pain intensity (0-10 NRS)")
```

Leg pain and back pain rise together with a slope well below the line of
equality, and the whole cloud shifts upward by roughly a point when the patient
says their leg pain is the worse of the two -- but the two panels overlap
substantially, confirming that `leg_pain_worse` is a self-report rather than a
recoding of which number is larger.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv
```

```{r uv-numbers}
uv_est <- function(term, vars) {
  tidy(lm(reformulate(vars, outcome), data = dat_model), conf.int = TRUE) |>
    filter(.data$term == !!term)
}
fmtp <- function(p) ifelse(p < 0.001, "< 0.001", paste("=", round(p, 3)))
u_bpi <- uv_est("back_pain_intensity", "back_pain_intensity")
u_lpw <- uv_est("leg_pain_worseYes", "leg_pain_worse")
u_bkp <- uv_est("below_knee_painYes", "below_knee_pain")
u_nt  <- uv_est("neural_tension_positiveYes", "neural_tension_positive")
u_mri <- uv_est("mri_nerve_root_compressionYes", "mri_nerve_root_compression")
u_age <- uv_est("age_yrs", "age_yrs")
```

Back-pain intensity is by far the strongest crude predictor: each additional
point of back pain is associated with a `r round(u_bpi$estimate, 2)`-point higher
leg-pain score (95% CI `r round(u_bpi$conf.low, 2)` to
`r round(u_bpi$conf.high, 2)`, p `r fmtp(u_bpi$p.value)`), and on its own it
accounts for R² = `r round(summary(lm(leg_pain_intensity ~ back_pain_intensity, dat_model))$r.squared, 2)`
of the variance in leg pain. Among the clinical variables, patients who report
their leg pain as worse than their back pain score
`r round(u_lpw$estimate, 2)` points higher (95% CI `r round(u_lpw$conf.low, 2)`
to `r round(u_lpw$conf.high, 2)`, p `r fmtp(u_lpw$p.value)`), pain below the knee
carries `r round(u_bkp$estimate, 2)` points (p `r fmtp(u_bkp$p.value)`), a
positive neural-tension test `r round(u_nt$estimate, 2)` points
(p `r fmtp(u_nt$p.value)`) and nerve-root compression on MRI
`r round(u_mri$estimate, 2)` points (p `r fmtp(u_mri$p.value)`). Age, sex and
leg-pain duration show no crude association: the age slope is
`r round(u_age$estimate, 3)` points per year
(p `r fmtp(u_age$p.value)`), and patients in their first six weeks report much
the same leg pain as those beyond twelve.

## Why adjustment is needed

The crude estimates conflate a patient's specific sciatic pain with how much
pain they report in general. `back_pain_intensity` is the clearest confounder of
every clinical sign: patients with a positive neural-tension test have mean back
pain
`r round(mean(dat_model$back_pain_intensity[dat_model$neural_tension_positive == "Yes"]), 1)`
versus
`r round(mean(dat_model$back_pain_intensity[dat_model$neural_tension_positive == "No"]), 1)`
in those without, and back pain is strongly associated with leg pain
(r = `r round(cor(dat_model$back_pain_intensity, dat_model[[outcome]]), 2)`), so
part of the crude neural-tension association is simply that patients with more
pain everywhere are more likely to flinch on examination. `leg_pain_worse` is a
second confounder, this time of the anatomical signs: it correlates with
`below_knee_pain`
(`r round(100 * mean(dat_model$leg_pain_worse[dat_model$below_knee_pain == "Yes"] == "Yes"), 0)`%
of patients with below-knee pain say leg pain is worse, versus
`r round(100 * mean(dat_model$leg_pain_worse[dat_model$below_knee_pain == "No"] == "Yes"), 0)`%
of those without) and is itself strongly associated with leg-pain intensity, so
the crude below-knee estimate partly reflects the patient's own ranking of their
two pains. `age_yrs` is a third, milder one: older patients are more likely to
have nerve-root compression on MRI
(r = `r round(cor(dat_model$age_yrs, as.numeric(dat_model$mri_nerve_root_compression) - 1), 2)`)
and report slightly more leg pain, so a little of the crude MRI association is
age. Adjustment converts "patients with a positive neural-tension test have worse
leg pain" into the more useful "at the same back-pain level, age and pain
ranking, a positive neural-tension test still marks out worse leg pain".

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv
```

```{r fit-stats}
gl <- glance(fit)
n_params <- length(coef(fit))
tibble(
  Statistic = c("R-squared", "Adjusted R-squared", "Residual SE",
                "N (complete cases)", "Estimated parameters",
                "Observations per parameter", "AIC"),
  Value = c(round(gl$r.squared, 3), round(gl$adj.r.squared, 3),
            round(gl$sigma, 2), nrow(dat_model), n_params,
            round(nrow(dat_model) / n_params, 1), round(AIC(fit), 1))
) |>
  gt() |>
  tab_header(title = "Model fit")
```

```{r mv-numbers}
tm <- tidy(fit, conf.int = TRUE)
g <- function(t) tm |> filter(term == t)
# What the model looks like without the back-pain predictor, for context.
r2_noback <- summary(lm(reformulate(setdiff(predictors, "back_pain_intensity"), outcome),
                        data = dat_model))$r.squared
```

Adjusting for all eight predictors, back-pain intensity remains overwhelmingly
the strongest association: conditional on everything else, each extra point of
back pain is associated with a `r round(g("back_pain_intensity")$estimate, 2)`-point
higher leg-pain score (95% CI `r round(g("back_pain_intensity")$conf.low, 2)` to
`r round(g("back_pain_intensity")$conf.high, 2)`,
p `r fmtp(g("back_pain_intensity")$p.value)`). Three clinical variables remain
independently associated. Reporting leg pain as worse than back pain adds
`r round(g("leg_pain_worseYes")$estimate, 2)` points (95% CI
`r round(g("leg_pain_worseYes")$conf.low, 2)` to
`r round(g("leg_pain_worseYes")$conf.high, 2)`,
p `r fmtp(g("leg_pain_worseYes")$p.value)`), pain below the knee adds
`r round(g("below_knee_painYes")$estimate, 2)` points
(p `r fmtp(g("below_knee_painYes")$p.value)`), and nerve-root compression on MRI
adds `r round(g("mri_nerve_root_compressionYes")$estimate, 2)` points
(p `r fmtp(g("mri_nerve_root_compressionYes")$p.value)`). The neural-tension test
is now only marginally associated
(`r round(g("neural_tension_positiveYes")$estimate, 2)` points,
p `r fmtp(g("neural_tension_positiveYes")$p.value)`). Age retains a small
association (`r round(g("age_yrs")$estimate, 3)` points per year,
p `r fmtp(g("age_yrs")$p.value)`); sex and leg-pain duration contribute nothing.

The model explains R² = `r round(gl$r.squared, 3)`
(adjusted R² = `r round(gl$adj.r.squared, 3)`) with
`r round(nrow(dat_model) / n_params, 1)` observations per estimated parameter.
That R² looks impressive but is mostly borrowed: dropping `back_pain_intensity`
alone takes it from `r round(gl$r.squared, 2)` to `r round(r2_noback, 2)`, so the
seven clinical and demographic predictors together explain only about
`r round(100 * r2_noback, 0)`% of the variance in leg-pain intensity. The honest
reading is that the consultation explains leg-pain severity poorly unless you are
allowed to ask how much back pain the patient also has.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

Three estimates move substantially, all toward the null. The **neural-tension
test** collapses from `r round(u_nt$estimate, 2)` points crude to
`r round(g("neural_tension_positiveYes")$estimate, 2)` adjusted -- a
`r round(100 * (1 - g("neural_tension_positiveYes")$estimate / u_nt$estimate), 0)`%
attenuation -- and drops from p `r fmtp(u_nt$p.value)` to
p `r fmtp(g("neural_tension_positiveYes")$p.value)`; its crude association was
largely the general-pain confounding described above, and it is the one predictor
whose significance is genuinely fragile. **Pain below the knee** halves from
`r round(u_bkp$estimate, 2)` to `r round(g("below_knee_painYes")$estimate, 2)`
points but stays firmly significant, so radiation below the knee marks out worse
leg pain even at the same overall pain level and the same self-ranked pain
pattern. **MRI nerve-root compression** attenuates from
`r round(u_mri$estimate, 2)` to
`r round(g("mri_nerve_root_compressionYes")$estimate, 2)` points and stays just
significant -- notable because it means the scan adds a little information about
symptom severity beyond the history, though less than a single point of pain.
**Leg pain worse than back pain** barely moves
(`r round(u_lpw$estimate, 2)` to `r round(g("leg_pain_worseYes")$estimate, 2)`)
and neither does **age** (`r round(u_age$estimate, 3)` to
`r round(g("age_yrs")$estimate, 3)`), confirming that neither is standing in for
the rest of the set. **Back-pain intensity** actually moves slightly *away* from
the null (`r round(u_bpi$estimate, 2)` to
`r round(g("back_pain_intensity")$estimate, 2)`): the clinical signs were
suppressing it a little, because patients with strong sciatic signs report
relatively more leg pain for a given amount of back pain. No predictor gains
significance on adjustment; one (neural tension) nearly loses it.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r diag-tests}
het <- check_heteroscedasticity(fit)
nor <- check_normality(fit)
col <- check_collinearity(fit)
out <- check_outliers(fit)
het; nor; col; out
```

```{r diag-numbers}
vif_tab <- as.data.frame(col) |> arrange(desc(VIF))
max_vif <- round(max(vif_tab$VIF), 2)
max_vif_term <- vif_tab$Term[which.max(vif_tab$VIF)]
p_het <- as.numeric(het); p_nor <- as.numeric(nor)
n_out <- sum(out)
n_pred_out <- sum(predict(fit) < 0 | predict(fit) > 10)
```

**Linearity.** The fitted-versus-residual smooth is essentially flat and the
posterior-predictive check tracks the observed distribution reasonably, though
simulated values spill below 0 and above 10 where the real scale cannot go.
In practice `r n_pred_out` of the `r nrow(dat_model)` fitted values fall outside
the 0-10 range, so the bound is not actually being violated at observed covariate
patterns; the linear model is adequate here even though it is formally defined on
the wrong support.

**Homoscedasticity.** Constant variance is **not** rejected
(`check_heteroscedasticity()` p = `r round(p_het, 3)`), so ordinary standard
errors are valid and HC3 re-estimation is not required. Because the outcome is
bounded and residual spread narrows near the ceiling, robust standard errors are
nevertheless reported below as a sensitivity check.

```{r hc3, results="asis"}
rob <- lmtest::coeftest(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
tibble(
  Term = rownames(rob),
  `SE (OLS)` = round(summary(fit)$coefficients[, 2], 3),
  `SE (HC3)` = round(rob[, 2], 3),
  `p (OLS)` = round(summary(fit)$coefficients[, 4], 3),
  `p (HC3)` = round(rob[, 4], 3)
) |>
  gt() |> tab_header(title = "HC3 robust standard errors (sensitivity)") |>
  as_raw_html() |> cat()
```

```{r hc3-note}
p_nt_ols <- summary(fit)$coefficients["neural_tension_positiveYes", 4]
p_nt_hc3 <- rob["neural_tension_positiveYes", 4]
```

Under HC3 the neural-tension coefficient moves from p = `r round(p_nt_ols, 3)` to
p = `r round(p_nt_hc3, 3)`, i.e. it sits on the 0.05 boundary either way; every
other conclusion is unchanged. This is the one substantive result in the model
that a reader should treat as provisional.

**Normality of residuals.** Formally **rejected**
(`check_normality()` p = `r round(p_nor, 3)`). The Q-Q plot shows mild departure
in the tails, driven by the coarse recording of the 0-10 scale in thirds of a
point and by the `r sum(dat_model[[outcome]] == 10)` patients piled at the
ceiling of 10. With n = `r nrow(dat_model)` the
central limit theorem covers the coefficient standard errors, so the practical
consequence is small, but this is a genuine assumption failure and should be
reported as one rather than waved away.

**Influential points.** `check_outliers()` flags `r n_out` observation(s) at the
Cook's-distance threshold.

**Collinearity.** No problem: the largest variance inflation factor is
`r max_vif_term` at `r max_vif`, well under 5. `leg_pain_worse` and
`below_knee_pain` are correlated but not collinear
(VIF `r round(vif_tab$VIF[vif_tab$Term == "leg_pain_worse"], 2)` and
`r round(vif_tab$VIF[vif_tab$Term == "below_knee_pain"], 2)`), and
`back_pain_intensity` has a VIF of only
`r round(vif_tab$VIF[vif_tab$Term == "back_pain_intensity"], 2)` despite
correlating 0.6 with the outcome -- a useful reminder that VIF concerns
predictor-predictor, not predictor-outcome, correlation. Excluding
`clinician_sciatica` avoided the one genuine collinearity threat in this file.

# Discussion

Among primary-care patients presenting with sciatica, leg-pain intensity is
predicted overwhelmingly by how much back pain the patient also reports. Beyond
that, three things independently mark out worse leg pain: the patient's own
judgement that the leg is the worse of the two sites (about
`r round(g("leg_pain_worseYes")$estimate, 1)` points), radiation below the knee
(about `r round(g("below_knee_painYes")$estimate, 1)` points) and nerve-root
compression on MRI (about
`r round(g("mri_nerve_root_compressionYes")$estimate, 1)` points on the 0-10
scale). A positive neural-tension test, which looks impressive crudely, is
almost entirely explained away by overall pain level. Practically, the history
carries more information about leg-pain severity than the examination does, and
the scan adds little: the MRI coefficient is smaller than the minimal clinically
important difference on a 0-10 pain scale, which is usually put at 1.5 to 2
points. Duration of the leg-pain episode is irrelevant to severity, which argues
against "how long has it been going on" as a triage question for pain intensity.

**Limitations.** The design is cross-sectional and all variables come from the
same consultation, so the back-pain-leg-pain association cannot be given a
direction and is inflated by common-method bias: the same patient rated both
scales minutes apart. The R² of `r round(gl$r.squared, 2)` is therefore not a
measure of clinical predictability -- strip out back pain and it falls to
`r round(r2_noback, 2)`. Residual normality is formally violated, and the
outcome is bounded at 0 and 10 with `r sum(dat_model[[outcome]] == 10)` patients
at the ceiling, so fitted values can and do stray outside the achievable range.
Complete-case analysis dropped `r n_dropped` patients
(`r round(100 * n_dropped / nrow(dat), 1)`%), almost all of them missing
`leg_pain_duration`; if patients who could not date their episode differ
systematically in pain, the estimates are mildly biased. The file has no
analgesic use, no psychological measures and no work status, all of which
plausibly affect reported pain, so residual confounding is likely. All estimates
are associations, not causal effects.

**Future work.** A repeat pain rating at three months would show whether the
below-knee and MRI findings predict persistence of leg pain, which is the
clinically actionable question.

# Feasibility notes

*Not part of a student report.*

- **Workable but weaker than `disability_score`.** The headline R² of
  `r round(gl$r.squared, 2)` is dominated by one predictor;
  without `back_pain_intensity` the model explains only
  `r round(r2_noback, 2)`. Students who include back pain will report a
  gratifying model and learn the wrong lesson about how predictable pain is;
  students who exclude it get an honest but flat result. Either way the
  supervisor has to raise the point.
- **Complete-case N = `r nrow(dat_model)`** of `r nrow(dat)`
  (`r n_dropped` dropped, `r round(100 * n_dropped / nrow(dat), 1)`%) -- the
  largest loss of the three outcomes, and entirely attributable to
  `leg_pain_duration` (19 missing), which then turns out to be non-significant.
  A good live demonstration that listwise deletion charges you for predictors
  before you know whether they are worth it.
- **Residual normality is rejected** (p = `r round(p_nor, 3)`) and the outcome is
  bounded and discrete: values are recorded in thirds of a point,
  `r sum(dat_model[[outcome]] == 10)` patients sit at the 10 ceiling and
  `r sum(dat_model[[outcome]] == 0)` at the 0 floor, and `r n_pred_out` fitted
  values fall outside 0-10. Homoscedasticity passes
  (p = `r round(p_het, 3)`), so HC3 is not required; it is shown anyway and
  moves the neural-tension p-value across the 0.05 line, which is worth knowing
  before a student stakes a conclusion on it.
- **No collinearity problem**: max VIF `r max_vif`. The `clinician_sciatica`
  trap applies here exactly as elsewhere -- it correlates 0.61 with
  `neural_tension_positive` and 0.54 with `below_knee_pain` and must stay out of
  any model containing those signs.
- **Watch `leg_pain_worse`.** It looks like it might be a deterministic
  recoding of `leg_pain_intensity > back_pain_intensity`, but it agrees with that
  comparison in only `r round(100 * agree, 0)`% of patients, so it is legitimate.
  Verify this before assuming either way -- a genuinely deterministic version of
  this variable would have made the outcome unusable.
- **Fragile result to flag**: `neural_tension_positive` sits at
  p `r fmtp(g("neural_tension_positiveYes")$p.value)` adjusted and crosses 0.05
  under robust SEs. Do not let a student build a discussion around it.
