---
title: "Body mass index in US adults"
subtitle: "Dataset 10: NHANES adults (2009-10 and 2011-12 extract)"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "nhanes_adults.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "bmi"
predictors <- c("age_yrs", "sex", "race", "education", "poverty_ratio",
                "phys_active", "sleep_hours", "smoked_100_cigs", "general_health")

lab <- function(x, l) { attr(x, "label") <- l; x }

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex   = lab(factor(sex, levels = c("female", "male"), labels = c("Female", "Male")),
                "Sex"),
    race  = lab(fct_relevel(factor(race), "White"), "Race/ethnicity"),
    education = lab(factor(education,
                           levels = c("College Grad", "Some College", "High School",
                                      "9 - 11th Grade", "8th Grade")),
                    "Education"),
    # Excellent (n = 47) and Poor (n = 11) are collapsed with their neighbours so no
    # cell is too thin to estimate; the reference is the healthiest group.
    general_health = lab(factor(case_when(
      general_health %in% c("Excellent", "Vgood") ~ "Excellent/Very good",
      general_health == "Good"                    ~ "Good",
      general_health %in% c("Fair", "Poor")       ~ "Fair/Poor"),
      levels = c("Excellent/Very good", "Good", "Fair/Poor")),
      "Self-rated general health"),
    smoked_100_cigs = lab(factor(smoked_100_cigs, levels = c(0, 1), labels = c("No", "Yes")),
                          "Ever smoked 100+ cigarettes"),
    phys_active = lab(factor(phys_active, levels = c(0, 1), labels = c("No", "Yes")),
                      "Recreationally physically active"),
    age_yrs = lab(age_yrs, "Age (years)"),
    poverty_ratio = lab(poverty_ratio, "Income-to-poverty ratio"),
    sleep_hours = lab(sleep_hours, "Sleep (hours/night)"),
    bmi = lab(bmi, "BMI (kg/m2)")
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
pv  <- function(p) ifelse(p < 0.001, "<0.001", paste0("= ", fmt(p, 3)))
```

# Introduction

The National Health and Nutrition Examination Survey (NHANES) is a continuous
cross-sectional survey of the civilian non-institutionalised US population that
pairs a household interview with a standardised examination, including measured
height and weight. This teaching extract holds `r nrow(dat)` adults aged 21 to 80
from the 2009-10 and 2011-12 cycles. Body mass index is the standard population
measure of adiposity and the proximal risk factor for type 2 diabetes,
hypertension and several cancers, so identifying which social and behavioural
characteristics track with it is a standard descriptive epidemiology question.

**Aim.** Among these US adults, we examined which of age, sex, race/ethnicity,
education, income-to-poverty ratio, recreational physical activity, sleep
duration, smoking history and self-rated general health are independently
associated with BMI, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** NHANES 2009-10 and 2011-12, a complex multistage probability
sample of the US civilian non-institutionalised population. This extract contains
`r nrow(dat)` adults with 31 variables. Height and weight were measured in the
mobile examination centre, not self-reported.

**Variables.** The outcome is `bmi`, body mass index in kg/m2, treated as
continuous. Predictors are age (years), sex (reference female), race/ethnicity
(reference White, the largest group), education (reference College Grad, the
largest group), income-to-poverty ratio (household income as a multiple of the
federal poverty line), recreational moderate or vigorous physical activity
(reference no), usual sleep hours per night, ever having smoked 100 cigarettes
(reference no) and self-rated general health collapsed to three levels
(reference Excellent/Very good).

Three variables are deliberately excluded. `weight_kg` and `height_cm` are
excluded because `bmi` is exactly `weight_kg / (height_cm/100)^2` in this file -
the largest absolute discrepancy across all rows is 0.05 kg/m2, which is pure
rounding - so including any two of the three produces a deterministic
relationship and a meaningless model. `obese` is excluded because it is exactly
`bmi >= 30` on every row where BMI is observed; regressing BMI on a dichotomised
copy of itself is circular.

**Statistical analysis.** `lm()`, justified because the outcome is continuous. We
fitted a univariable model for each predictor first, then a single multivariable
model containing all of them. Predictors were chosen in advance on subject-matter
grounds, not by significance. Significance threshold 0.05; estimates are reported
with 95% confidence intervals. Assumptions were checked with
`performance::check_model()`, and because non-constant variance was flagged we
additionally report HC3 heteroscedasticity-consistent standard errors. Models are
**unweighted**: the NHANES survey weights, strata and primary sampling units are
not carried in this extract, so no result here is a nationally representative
estimate. Analysis used `r R.version.string` with tidyverse, gtsummary and
performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped). The main cost is `general_health`
(`r sum(is.na(dat$general_health))` missing) and `poverty_ratio`
(`r sum(is.na(dat$poverty_ratio))` missing). Only
`r sum(complete.cases(dat_raw))` of `r nrow(dat)` rows are complete on *every*
variable in the file, so listwise deletion is a live threat and the complete-case
N is reported for every model.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(missing = "no") |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

Median BMI sits in the overweight range with a long right tail reaching beyond
60 kg/m2. The sample is predominantly White and skews toward the more educated
categories, with roughly half reporting recreational physical activity and about
a third rating their health as no better than Good.

## Exploratory figure

```{r explore, fig.cap="Figure 1. BMI by self-rated general health and physical activity."}
ggplot(dat_model, aes(general_health, bmi, fill = phys_active)) +
  geom_boxplot(outlier.alpha = 0.4, width = 0.7) +
  scale_fill_manual(values = c(No = "#C46B4A", Yes = "#3F8F7A")) +
  labs(x = "Self-rated general health", y = "BMI (kg/m2)",
       fill = "Physically active")
```

BMI rises monotonically as self-rated health worsens, and within every
health category the physically active group is shifted slightly lower, though
the distributions overlap heavily.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv

utid <- map_dfr(predictors, function(v) {
  tidy(lm(reformulate(v, outcome), data = dat_model), conf.int = TRUE) |>
    filter(term != "(Intercept)")
})
ug <- function(term, d = 2) {
  r <- utid[utid$term == term, ]
  sprintf("%s kg/m2 (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d), fmt(r$conf.high, d))
}
up <- function(term) pv(utid$p.value[utid$term == term])
```

Self-rated health is the strongest crude predictor: compared with participants
rating their health Excellent or Very good, those rating it Good average
`r ug("general_healthGood")` higher BMI (p `r up("general_healthGood")`) and
those rating it Fair or Poor average `r ug("general_healthFair/Poor")` higher
(p `r up("general_healthFair/Poor")`). Recreational physical activity is
associated with `r ug("phys_activeYes")` (p `r up("phys_activeYes")`) lower BMI,
each unit of income-to-poverty ratio with `r ug("poverty_ratio")`
(p `r up("poverty_ratio")`), and men average `r ug("sexMale")`
(p `r up("sexMale")`) relative to women. Participants in the "Other"
race category average `r ug("raceOther")` (p `r up("raceOther")`) lower BMI than
White participants. Age, education, sleep hours and smoking show no crude
association.

## Why adjustment is needed

**Income-to-poverty ratio** is a clear confounder of the education and
race/ethnicity associations: education and income are strongly correlated in
this sample, and higher income is associated with lower BMI, so part of any
crude education gradient is really an income gradient. **Self-rated general
health** is the other variable that must be handled explicitly - it is
associated with income, education and physical activity (people who are poorer,
less educated and inactive rate their health worse) and it is strongly associated
with BMI, so it will absorb part of every socioeconomic and behavioural
coefficient. Note that the causal direction here is ambiguous: high BMI plausibly
*causes* worse self-rated health, in which case adjusting for it is adjusting for
a mediator rather than a confounder, and the adjusted activity and income
coefficients are correspondingly hard to interpret. **Age** confounds the sleep
and activity associations, since older participants sleep and exercise
differently and BMI varies with age. Mutual adjustment lets each coefficient be
read as a comparison between participants alike on the other measured
characteristics, with the mediation caveat kept in view.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv

tid <- tidy(fit, conf.int = TRUE)
g <- function(term, d = 2) {
  r <- tid[tid$term == term, ]
  sprintf("%s kg/m2 (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d), fmt(r$conf.high, d))
}
gp <- function(term) pv(tid$p.value[tid$term == term])
```

```{r fit-stats}
s <- summary(fit)
tibble(
  Statistic = c("Complete-case N", "Parameters (excl. intercept)",
                "R-squared", "Adjusted R-squared", "Residual SE (kg/m2)"),
  Value = c(as.character(nrow(dat_model)),
            as.character(length(coef(fit)) - 1L),
            fmt(s$r.squared, 3), fmt(s$adj.r.squared, 3), fmt(s$sigma, 2))
) |>
  gt() |>
  tab_header(title = md("**Model fit: multivariable linear model for BMI**"))
```

The model explains `r fmt(100 * s$r.squared, 1)`% of the variance in BMI
(adjusted R-squared `r fmt(s$adj.r.squared, 3)`). The gap between R-squared and
adjusted R-squared is noticeable because the model spends
`r length(coef(fit)) - 1L` parameters on `r nrow(dat_model)` observations, many of
them on category contrasts that contribute little. Among participants of the same
age, sex, race, education, income, activity, sleep and smoking status, rating
one's health as Good rather than Excellent/Very good is associated with
`r g("general_healthGood")` higher BMI (p `r gp("general_healthGood")`), and
Fair/Poor with `r g("general_healthFair/Poor")` higher
(p `r gp("general_healthFair/Poor")`). Physical activity remains associated with
`r g("phys_activeYes")` (p `r gp("phys_activeYes")`). Participants in the "Other"
race group average `r g("raceOther")` (p `r gp("raceOther")`) and Hispanic
participants `r g("raceHispanic")` (p `r gp("raceHispanic")`) relative to White
participants. Age, sex, education, income, sleep and smoking are not
independently associated. These are associations among the participants observed,
not effects of intervening.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

Two predictors move meaningfully toward the null and lose significance. Sex falls
from `r ug("sexMale")` crude to `r g("sexMale")` adjusted, and income-to-poverty
ratio from `r ug("poverty_ratio")` to `r g("poverty_ratio")` - in both cases the
crude estimate was carrying self-rated health and education, which are correlated
with both. Physical activity attenuates but survives, from `r ug("phys_activeYes")`
to `r g("phys_activeYes")`, so part but not all of the crude activity difference
was socioeconomic and health-status confounding. By contrast, the two general-
health contrasts barely move at all (Good: `r ug("general_healthGood")` to
`r g("general_healthGood")`; Fair/Poor: `r ug("general_healthFair/Poor")` to
`r g("general_healthFair/Poor")`), and race "Other" is essentially unchanged
(`r ug("raceOther")` to `r g("raceOther")`) - these are not explained by the other
measured characteristics. One estimate crosses *into* significance: Hispanic
ethnicity goes from `r ug("raceHispanic")` (p `r up("raceHispanic")`) to
`r g("raceHispanic")` (p `r gp("raceHispanic")`), which with n =
`r sum(dat_model$race == "Hispanic")` Hispanic participants is a fragile finding
rather than a discovery.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r het-check}
het <- check_heteroscedasticity(fit)
print(het)
```

```{r robust-se}
rob <- lmtest::coeftest(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
robci <- lmtest::coefci(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
tibble(
  Term = rownames(rob),
  Estimate = fmt(rob[, 1], 3),
  `SE (OLS)` = fmt(summary(fit)$coefficients[, 2], 3),
  `SE (HC3)` = fmt(rob[, 2], 3),
  `95% CI (HC3)` = paste0(fmt(robci[, 1], 2), " to ", fmt(robci[, 2], 2)),
  `p (HC3)` = fmt(rob[, 4], 3)
) |>
  filter(Term != "(Intercept)") |>
  gt() |>
  tab_header(title = md("**Table 3. HC3 heteroscedasticity-consistent standard errors.**"))
```

The diagnostics are the weakest part of this analysis. The posterior-predictive
check shows the model cannot reproduce the long right tail of the BMI
distribution: observed BMI is strongly right-skewed and the fitted normal
predictive distribution is symmetric. The residual-versus-fitted panel fans out
markedly, and `check_heteroscedasticity()` confirms non-constant error variance
(p < 0.001). The normal-QQ panel shows a pronounced heavy right tail, and
`check_normality()` also rejects. Collinearity is the highest of any model in
this dataset but still acceptable: the largest VIF is 2.09 for `education`,
followed by 1.76 for race/ethnicity and 1.40 for income-to-poverty ratio. The
first two are four-degree-of-freedom terms, and the correlation between education
and income is the substantive reason for the inflation; nothing here approaches
the conventional VIF threshold of 5. A small number of very obese participants sit
near the Cook's distance contours but none dominates the fit.

Because variance is non-constant, Table 3 reports HC3 robust standard errors.
The HC3 SE for the Fair/Poor health contrast is
`r fmt(rob["general_healthFair/Poor", 2], 3)` against an OLS SE of
`r fmt(summary(fit)$coefficients["general_healthFair/Poor", 2], 3)`, i.e. wider,
which is the expected direction when the worse-health group also has more
variable BMI. The substantive conclusions do not change: the general-health
contrasts, race "Other" and physical activity keep their significance status
under robust inference. A log or square-root transformation of BMI would address
both the skew and much of the heteroscedasticity, at the cost of interpreting
coefficients on a multiplicative scale.

# Discussion

In this NHANES extract, worse self-rated general health is the characteristic
most strongly and independently associated with higher BMI - roughly
`r fmt(tid$estimate[tid$term == "general_healthGood"], 1)` kg/m2 for Good and
`r fmt(tid$estimate[tid$term == "general_healthFair/Poor"], 1)` kg/m2 for
Fair/Poor relative to Excellent/Very good - followed by lower BMI among the
physically active. The crude sex and income differences are fully explained by
the other measured characteristics. Practically, self-rated health is a single
question that identifies a group averaging several BMI units higher, which is
useful for targeting, but the association is very likely bidirectional and should
not be read as health perception causing weight.

**Limitations.** The models are unweighted, and the NHANES sampling weights,
strata and PSUs are not in this extract, so the sample **must not be described as
nationally representative**. The design is cross-sectional, so temporality is
unknown; in particular, adjusting for self-rated general health is probably
adjusting for a *consequence* of high BMI, which biases the remaining
coefficients in an unpredictable direction - this is over-adjustment, not
confounder control. Physical activity is a crude yes/no self-report with no
intensity or duration, and reverse causation is likely (high BMI reduces
exercise). Diet, which is the dominant determinant of BMI, is not measured at
all. Complete-case analysis dropped `r n_dropped` of `r nrow(dat)` rows
(`r fmt(100 * n_dropped / nrow(dat), 0)`%). Residual variance is non-constant and
residuals are strongly non-normal; HC3 SEs address inference but not the model's
inability to predict the obese tail. Hispanic (n =
`r sum(dat_model$race == "Hispanic")`) and Other (n =
`r sum(dat_model$race == "Other")`) groups are small, so those contrasts are
unstable. All estimates are associations, not causal effects.

**Future work.** Refit on log(BMI) with the NHANES survey design and dietary
recall variables from the full public files, and drop self-rated health from the
adjustment set to avoid conditioning on a mediator.

# Feasibility notes

*NOT part of a student report.*

- **Usable but not the strongest choice.** The signal is real but almost entirely
  carried by `general_health`, which is arguably a consequence of BMI rather than
  a cause. A student who does not spot the mediation issue will write a
  confidently wrong discussion.
- **Trap: `obese` must never be a predictor here.** `obese` is exactly
  `bmi >= 30` on all 485 rows where BMI is observed (the 3 apparent exceptions are
  rows where BMI itself is missing, so `obese` is missing too). Including it
  yields an R-squared near 0.6 that means nothing.
- **Trap: `weight_kg` and `height_cm` must be excluded.** `bmi` is exactly
  `weight_kg/(height_cm/100)^2` (max discrepancy 0.05 kg/m2). Any two of the
  three in one model is a deterministic relationship.
- **Failed diagnostics.** Heteroscedasticity p < 0.001 *and* non-normality
  p < 0.001, both driven by the long right tail of BMI. HC3 fixes inference;
  it does not fix the fit. This is a good teaching case for transformation, but a
  student who reports untransformed BMI with a straight face has an assumption
  violation on the record.
- **Low explained variance (R-squared `r fmt(s$r.squared, 3)`, adjusted
  `r fmt(s$adj.r.squared, 3)`).** The `r length(coef(fit)) - 1L` parameters buy
  very little; the adjusted-versus-raw gap is worth pointing out.
- **Complete-case N `r nrow(dat_model)` of `r nrow(dat)`**
  (`r fmt(100 * n_dropped / nrow(dat), 0)`% lost), driven by `general_health`
  (10% missing) and `poverty_ratio` (8% missing).
- **Small race cells.** Hispanic `r sum(dat_model$race == "Hispanic")`, Other
  `r sum(dat_model$race == "Other")` in the analytic sample; the Hispanic
  coefficient flips into significance on adjustment and should not be trusted.
- **Unweighted survey analysis.** Must be stated in Methods and Limitations.
- No separation (linear model). Highest collinearity of the four linear models
  (max VIF 2.09, `education`; race 1.76; `poverty_ratio` 1.40) but well short of
  a threshold that would require action.
