---
title: "Diagnosed diabetes in US adults"
subtitle: "Dataset 10: NHANES adults (2009-10 and 2011-12 extract)"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "nhanes_adults.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names. The predictor set is
# capped at five parameters because there are only 57 events in the file.
outcome    <- "diabetes"
predictors <- c("age_yrs", "sex", "bmi", "phys_active", "smoked_100_cigs")

lab <- function(x, l) { attr(x, "label") <- l; x }

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex   = lab(factor(sex, levels = c("female", "male"), labels = c("Female", "Male")),
                "Sex"),
    smoked_100_cigs = lab(factor(smoked_100_cigs, levels = c(0, 1), labels = c("No", "Yes")),
                          "Ever smoked 100+ cigarettes"),
    phys_active = lab(factor(phys_active, levels = c(0, 1), labels = c("No", "Yes")),
                      "Recreationally physically active"),
    age_yrs = lab(age_yrs, "Age (years)"),
    bmi = lab(bmi, "BMI (kg/m2)"),
    diabetes = lab(diabetes, "Diagnosed diabetes")
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)
n_events  <- sum(dat_model[[outcome]])

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
pv  <- function(p) ifelse(p < 0.001, "<0.001", paste0("= ", fmt(p, 3)))
```

# Introduction

The National Health and Nutrition Examination Survey (NHANES) is a continuous
cross-sectional survey of the civilian non-institutionalised US population,
pairing a household interview with a standardised examination. This teaching
extract holds `r nrow(dat)` adults aged 21 to 80 from the 2009-10 and 2011-12
cycles, of whom `r sum(dat_raw$diabetes)` (`r fmt(100 * mean(dat_raw$diabetes), 1)`%)
report a doctor's diagnosis of diabetes. Type 2 diabetes is one of the most
consequential chronic conditions in adult populations, and identifying which
routinely collected characteristics mark higher prevalence supports targeted
screening.

**Aim.** Among these US adults, we examined which of age, sex, BMI, recreational
physical activity and smoking history are independently associated with having a
diagnosis of diabetes, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** NHANES 2009-10 and 2011-12, a complex multistage probability
sample of the US civilian non-institutionalised population. This extract contains
`r nrow(dat)` adults with 31 variables. Height and weight were measured in the
mobile examination centre; diabetes status is self-reported diagnosis by a
doctor, collected in the household interview.

**Variables.** The outcome is `diabetes`, a binary indicator where the **event is
a reported doctor's diagnosis of diabetes** (`r sum(dat_raw$diabetes)` events in
the full file). Predictors are age (years), sex (reference female), BMI (kg/m2),
recreational moderate or vigorous physical activity (reference no) and ever
having smoked 100 cigarettes (reference no).

**The predictor set is deliberately small.** With `r sum(dat_raw$diabetes)`
events in the file, the conventional rule of ten events per estimated parameter
supports about **five parameters**. The five predictors above cost exactly five
parameters (all are continuous or binary, so none consumes more than one degree
of freedom). Adding race/ethnicity, education or marital status - each of which
would cost four or five further parameters - would take the model well past what
the data can support and produce unstable, over-optimistic estimates. In the case
of race/ethnicity it would do worse than that: the "Other" category contains
`r sum(dat_raw$race == "Other")` participants and **none** of them reports
diabetes, so adding race produces complete separation and a coefficient that
diverges. This budget was fixed before any model was fitted.

`weight_kg` and `height_cm` are excluded because `bmi` is exactly
`weight_kg / (height_cm/100)^2` in this file. `obese` is excluded because it is
exactly `bmi >= 30` and would be collinear with BMI while discarding information.

**Statistical analysis.** `glm(family = binomial)`, justified because the outcome
is binary. We fitted a univariable model for each predictor first, then a single
multivariable model containing all of them. Predictors were chosen in advance on
subject-matter grounds, not by significance. Coefficients are exponentiated and
reported as odds ratios with 95% confidence intervals; significance threshold
0.05. Model adequacy was assessed with `performance::check_model()`, focusing on
binned residuals, collinearity, influential observations and events per
parameter. Models are **unweighted**: the NHANES survey weights, strata and
primary sampling units are not carried in this extract, so no result here is a
nationally representative estimate. Analysis used `r R.version.string` with
tidyverse, gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped, all for missing BMI), leaving
`r n_events` events. This is the least attrition of any model in this dataset
because none of the five predictors has meaningful missingness. Only
`r sum(complete.cases(dat_raw))` of `r nrow(dat)` rows are complete on *every*
variable in the file, so a larger predictor set would have cost far more than
`r n_dropped` rows - another reason to keep the model small.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  mutate(diabetes = factor(diabetes, levels = c(0, 1), labels = c("No diabetes", "Diabetes"))) |>
  tbl_summary(
    by = diabetes,
    missing = "no"
  ) |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics by diabetes status.**")
```

Participants with diagnosed diabetes are markedly older (median 63 versus 47
years) and heavier (median BMI 32 versus 28 kg/m2) than those without, and only
about a quarter report recreational physical activity compared with about half of
those without diabetes. Sex and smoking history look similar in the two groups.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Proportion with diagnosed diabetes by age group and BMI category."}
dat_model |>
  mutate(
    age_grp = cut(age_yrs, breaks = c(20, 40, 55, 70, 81), right = FALSE,
                  labels = c("21-39", "40-54", "55-69", "70-80")),
    bmi_grp = cut(bmi, breaks = c(0, 25, 30, Inf), right = FALSE,
                  labels = c("<25", "25-29.9", "30+"))
  ) |>
  group_by(age_grp, bmi_grp) |>
  summarise(p = mean(diabetes), n = n(), .groups = "drop") |>
  ggplot(aes(age_grp, p, fill = bmi_grp)) +
  geom_col(position = position_dodge(0.85), width = 0.78) +
  geom_text(aes(label = paste0("n=", n)), position = position_dodge(0.85),
            vjust = -0.4, size = 2.9, colour = "grey30") +
  scale_y_continuous(labels = scales::percent, expand = expansion(c(0, 0.12))) +
  scale_fill_manual(values = c("<25" = "#7FA9C9", "25-29.9" = "#E0A458", "30+" = "#B5495B")) +
  labs(x = "Age group (years)", y = "Proportion with diagnosed diabetes",
       fill = "BMI (kg/m2)")
```

Diabetes prevalence rises steeply with both age and BMI and the two effects
compound: it is near zero in normal-weight adults under 40 and approaches a third
in obese adults aged 55 and over. (BMI category is used here for display only;
BMI enters the models as a continuous variable.)

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = glm,
    method.args = list(family = binomial),
    exponentiate = TRUE
  )
uv

utid <- map_dfr(predictors, function(v) {
  tidy(glm(reformulate(v, outcome), data = dat_model, family = binomial),
       conf.int = TRUE, exponentiate = TRUE) |>
    filter(term != "(Intercept)")
})
ug <- function(term, d = 2) {
  r <- utid[utid$term == term, ]
  sprintf("OR %s (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d), fmt(r$conf.high, d))
}
up <- function(term) pv(utid$p.value[utid$term == term])
```

Three predictors show strong crude associations. Each additional year of age
multiplies the odds of diagnosed diabetes by `r ug("age_yrs")`
(p `r up("age_yrs")`), which compounds to roughly
`r fmt(utid$estimate[utid$term == "age_yrs"]^10, 1)`-fold per decade. Each
additional kg/m2 of BMI multiplies the odds by `r ug("bmi")`
(p `r up("bmi")`), i.e. about
`r fmt(utid$estimate[utid$term == "bmi"]^5, 1)`-fold per 5 kg/m2. Recreational
physical activity is associated with `r ug("phys_activeYes")`
(p `r up("phys_activeYes")`), a two-thirds lower odds. Sex
(`r ug("sexMale")`, p `r up("sexMale")`) and smoking (`r ug("smoked_100_cigsYes")`,
p `r up("smoked_100_cigsYes")`) show no crude association.

## Why adjustment is needed

**Age** is the primary confounder of the physical-activity association. Older
adults exercise less and, independently, have far higher diabetes prevalence
because incidence accumulates over the life course; the crude two-thirds lower
odds among active participants is therefore partly just the youth of the active
group. **BMI** is the second confounder, and it works on the activity association
in the same direction: physically active people are leaner, and adiposity is the
strongest modifiable driver of type 2 diabetes, so the crude activity odds ratio
also carries a body-composition difference. BMI additionally confounds the sex
comparison, since BMI distributions differ between men and women in this sample.
Mutual adjustment lets each odds ratio be read as a comparison between
participants of the same age, sex, body mass, activity level and smoking history.

## Multivariable model

```{r multivariable}
fit <- glm(reformulate(predictors, outcome), data = dat_model, family = binomial)

mv <- tbl_regression(fit, exponentiate = TRUE)
mv

tid <- tidy(fit, conf.int = TRUE, exponentiate = TRUE)
g <- function(term, d = 2) {
  r <- tid[tid$term == term, ]
  sprintf("OR %s (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d), fmt(r$conf.high, d))
}
gp <- function(term) pv(tid$p.value[tid$term == term])
```

```{r fit-stats}
n_par  <- length(coef(fit)) - 1L
epp    <- n_events / n_par
null_fit <- glm(reformulate("1", outcome), data = dat_model, family = binomial)
tibble(
  Statistic = c("Complete-case N", "Events (diabetes = 1)", "Parameters (excl. intercept)",
                "Events per parameter", "AIC (null model)", "AIC (full model)",
                "Tjur's R-squared"),
  Value = c(as.character(nrow(dat_model)), as.character(n_events), as.character(n_par),
            fmt(epp, 1), fmt(AIC(null_fit), 1), fmt(AIC(fit), 1),
            fmt(as.numeric(r2_tjur(fit)), 3))
) |>
  gt() |>
  tab_header(title = md("**Model fit: multivariable logistic model for diabetes**"))
```

The model has `r n_events` events for `r n_par` estimated parameters, i.e.
**`r fmt(epp, 1)` events per parameter**, which clears the conventional minimum
of ten - but only just, and only because the predictor set was capped at five in
advance. AIC falls from `r fmt(AIC(null_fit), 1)` for the intercept-only model to
`r fmt(AIC(fit), 1)`, a drop of `r fmt(AIC(null_fit) - AIC(fit), 1)`, so the
predictors collectively carry real information; Tjur's R-squared is
`r fmt(as.numeric(r2_tjur(fit)), 3)`.

Among participants of the same sex, BMI, activity level and smoking history, each
additional year of age multiplies the odds of diagnosed diabetes by
`r g("age_yrs")` (p `r gp("age_yrs")`). Among participants alike on everything
else, each additional kg/m2 of BMI multiplies the odds by `r g("bmi")`
(p `r gp("bmi")`) - about `r fmt(tid$estimate[tid$term == "bmi"]^5, 1)`-fold per
5 kg/m2. Physical activity is associated with `r g("phys_activeYes")`
(p `r gp("phys_activeYes")`), no longer significant at the 0.05 level. Sex
(`r g("sexMale")`) and smoking (`r g("smoked_100_cigsYes")`) remain null. These
are associations among the participants observed, not effects of intervening.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted odds ratios.**")
```

Physical activity is the coefficient that moves most: from `r ug("phys_activeYes")`
crude to `r g("phys_activeYes")` adjusted, a shift of about 60% of the way toward
the null on the log-odds scale, and it crosses from clearly significant
(p `r up("phys_activeYes")`) to non-significant (p `r gp("phys_activeYes")`). This
is confounding by age and BMI exactly as anticipated: active participants are
younger and leaner, and once those are held constant most of the apparent
protection disappears. Sex moves in the other direction, from
`r ug("sexMale")` to `r g("sexMale")`, i.e. from a slight apparent male advantage
to a slight male excess - masking by BMI, since women in this sample have higher
BMI than men, though the estimate is null in both columns. The two dominant
predictors barely move: age from `r ug("age_yrs")` to `r g("age_yrs")` and BMI
from `r ug("bmi")` to `r g("bmi")`, both if anything slightly *stronger* after
adjustment, because age and BMI partially masked each other (older participants
in this sample are not the heaviest). Smoking is unchanged and null throughout.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r binned-resid}
br <- binned_residuals(fit)
print(br)
```

```{r collin}
check_collinearity(fit)
```

The binned-residual check is the one that matters here, and it is **not
reassuring**: only about 64% of binned residuals fall inside the error bounds,
against a target of 95% or more. With `r n_events` events spread across
`r nrow(dat_model)` observations, each bin holds very few events, so the check is
itself noisy - but the pattern is consistent with the linear-in-log-odds
assumption for age and BMI being imperfect at the extremes, where predicted
probabilities are very small. Collinearity is not a problem: all VIFs are below
1.2, so the age, BMI and activity coefficients are not competing for the same
variance. No observation exceeds the influence contours (maximum Cook's distance
`r fmt(max(cooks.distance(fit)), 3)`), and the largest leverage is
`r fmt(max(hatvalues(fit)), 3)`, so no single participant drives the fit. There is
no separation: every predictor has finite coefficients and standard errors below
`r fmt(max(summary(fit)$coefficients[, 2]), 2)`, and both levels of each binary
predictor contain events. Homogeneity of variance and normality of residuals do
not apply to a logistic model and are not interpreted.

# Discussion

In this NHANES extract, age and BMI are independently associated with having a
diabetes diagnosis, and their associations are strong: roughly
`r fmt(tid$estimate[tid$term == "age_yrs"]^10, 1)`-fold higher odds per decade of
age and `r fmt(tid$estimate[tid$term == "bmi"]^5, 1)`-fold per 5 kg/m2 of BMI,
each holding the other constant. The crude protective association with
recreational physical activity is largely explained by activity being commoner in
younger and leaner participants and does not survive adjustment. Practically,
this supports age and BMI as the primary screening axes and cautions against
reading a crude activity-diabetes contrast as evidence that exercise prevents
diabetes.

**Limitations.** The models are unweighted, and the NHANES sampling weights,
strata and PSUs are not in this extract, so the sample **must not be described as
nationally representative**. The outcome is *self-reported diagnosis*, not
measured glycaemia: undiagnosed diabetes is misclassified as non-diabetic, and
because undiagnosed diabetes is commoner in younger people and those with less
contact with health services, this misclassification is differential with respect
to age and income - a serious bias in the direction of exaggerating the age
association. The design is cross-sectional, so reverse causation is entirely
plausible for BMI and activity: people diagnosed with diabetes are advised to
lose weight and exercise, which would attenuate both associations. With only
`r n_events` events the model could carry no more than five parameters, so
race/ethnicity, education and family history - all relevant - are omitted and
residual confounding is certain. The binned-residual check flags poor calibration.
Complete-case analysis dropped only `r n_dropped` rows here, but that is a
consequence of the deliberately small predictor set, not of the file being
complete. All estimates are associations, not causal effects.

**Future work.** Refit with the NHANES survey design and with measured HbA1c or
fasting glucose to capture undiagnosed diabetes, which would remove the dominant
misclassification problem.

# Feasibility notes

*NOT part of a student report.*

- **Viable, but tightly constrained by the event count.** `r sum(dat_raw$diabetes)`
  events in the file, `r n_events` in the analytic sample, supporting about
  **five parameters at ten events per parameter**. The model here uses exactly
  five (EPP = `r fmt(epp, 1)`). A student who throws in `race` (4 df) and
  `education` (4 df) will be at 13 parameters and EPP ~4.3, which is over-fitting
  regardless of what the p-values say. This must be stated explicitly in the
  brief.
- **No separation at this predictor set** - all coefficients finite, max SE
  `r fmt(max(summary(fit)$coefficients[, 2]), 2)`, max VIF 1.13.
- **FATAL if `race` is added: complete separation.** The "Other" stratum has
  **0 diabetes events in `r sum(dat_raw$race == "Other")` participants** (Hispanic
  has 3, Mexican 4). Adding `race` to this model produces a `raceOther`
  coefficient of about -14 with a standard error of about 729 - textbook
  separation - alongside a "fitted probabilities numerically 0 or 1" warning. This
  is the single most likely way a student wrecks this outcome, and it is also a
  second, independent reason for the five-parameter cap.
- **MAJOR: binned residuals fail** (~64% inside bounds vs 95% target). Partly an
  artefact of few events per bin, partly genuine miscalibration. A student
  following the diagnostics honestly will have to report a failed check.
- **MAJOR conceptual problem: the outcome is self-reported *diagnosis*, not
  diabetes.** Undiagnosed cases are counted as non-cases, and this is
  differential by age and healthcare access. Strong students will spot it; weak
  ones will write "age causes diabetes".
- **Reverse causation is severe.** BMI and physical activity are both plausibly
  *consequences* of a diabetes diagnosis in a cross-sectional file. The activity
  finding is uninterpretable causally.
- **Complete-case N `r nrow(dat_model)` of `r nrow(dat)`** - the mildest
  attrition of any outcome here, but only because the predictor set had to be
  small. Any predictor with real missingness (`poverty_ratio`, `general_health`,
  `alcohol_days_year`) costs both rows *and* events.
- **Trap: `obese` and `bmi` are the same variable.** `obese` is exactly
  `bmi >= 30`; never put both in this model.
- **Unweighted survey analysis.** Must be stated.
