---
title: "HDL cholesterol in US adults"
subtitle: "Dataset 10: NHANES adults (2009-10 and 2011-12 extract)"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "nhanes_adults.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "hdl_cholesterol"
predictors <- c("age_yrs", "sex", "bmi", "phys_active", "smoked_100_cigs",
                "alcohol_days_year", "diabetes")

lab <- function(x, l) { attr(x, "label") <- l; x }

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex   = lab(factor(sex, levels = c("female", "male"), labels = c("Female", "Male")),
                "Sex"),
    smoked_100_cigs = lab(factor(smoked_100_cigs, levels = c(0, 1), labels = c("No", "Yes")),
                          "Ever smoked 100+ cigarettes"),
    diabetes = lab(factor(diabetes, levels = c(0, 1), labels = c("No", "Yes")),
                   "Diagnosed diabetes"),
    phys_active = lab(factor(phys_active, levels = c(0, 1), labels = c("No", "Yes")),
                      "Recreationally physically active"),
    age_yrs = lab(age_yrs, "Age (years)"),
    bmi = lab(bmi, "BMI (kg/m2)"),
    alcohol_days_year = lab(alcohol_days_year, "Drinking days in past year"),
    hdl_cholesterol = lab(hdl_cholesterol, "HDL cholesterol (mmol/L)")
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

fmt <- function(x, d = 3) formatC(x, format = "f", digits = d)
pv  <- function(p) ifelse(p < 0.001, "<0.001", paste0("= ", formatC(p, format = "f", digits = 3)))
```

# Introduction

The National Health and Nutrition Examination Survey (NHANES) is a continuous
cross-sectional survey of the civilian non-institutionalised US population,
pairing a household interview with a standardised examination and venous blood
draw. This teaching extract holds `r nrow(dat)` adults aged 21 to 80 from the
2009-10 and 2011-12 cycles. HDL cholesterol is the "good" lipid fraction: low HDL
is a component of the metabolic syndrome definition and an independent marker of
cardiovascular risk, and unlike total cholesterol it is only weakly affected by
statin therapy, so its determinants can be studied with less treatment
contamination.

**Aim.** Among these US adults, we examined which of age, sex, BMI, recreational
physical activity, smoking history, alcohol frequency and diagnosed diabetes are
independently associated with HDL cholesterol, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** NHANES 2009-10 and 2011-12, a complex multistage probability
sample of the US civilian non-institutionalised population. This extract contains
`r nrow(dat)` adults with 31 variables. Lipids were measured on a venous sample
taken in the mobile examination centre.

**Variables.** The outcome is `hdl_cholesterol`, HDL cholesterol in mmol/L,
treated as continuous. Predictors are age (years), sex (reference female), BMI
(kg/m2), recreational moderate or vigorous physical activity (reference no), ever
having smoked 100 cigarettes (reference no), number of drinking days in the past
year and doctor-diagnosed diabetes (reference no). These seven were chosen
because each has an established biological pathway to HDL: adiposity and insulin
resistance lower it, aerobic activity and moderate alcohol raise it, and smoking
lowers it.

`total_cholesterol` is deliberately excluded: HDL is by laboratory definition a
component of total cholesterol, so the two are related partly by construction and
a coefficient for total cholesterol here has no substantive meaning.
`weight_kg` and `height_cm` are excluded because `bmi` is exactly
`weight_kg / (height_cm/100)^2` in this file.

**Statistical analysis.** `lm()`, justified because the outcome is continuous. We
fitted a univariable model for each predictor first, then a single multivariable
model containing all of them. Predictors were chosen in advance on subject-matter
grounds, not by significance. Significance threshold 0.05; estimates are reported
with 95% confidence intervals. Assumptions were checked with
`performance::check_model()`, and because non-constant variance was flagged we
additionally report HC3 heteroscedasticity-consistent standard errors. Models are
**unweighted**: the NHANES survey weights, strata and primary sampling units are
not carried in this extract, so no result here is a nationally representative
estimate. Analysis used `r R.version.string` with tidyverse, gtsummary and
performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped). Most of the loss comes from
`alcohol_days_year` (`r sum(is.na(dat$alcohol_days_year))` missing, 18%) plus
`hdl_cholesterol` itself (`r sum(is.na(dat$hdl_cholesterol))` missing). Only
`r sum(complete.cases(dat_raw))` of `r nrow(dat)` rows are complete on *every*
variable in the file, so listwise deletion is a live threat and the complete-case
N is reported for every model.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(missing = "no") |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

Median HDL is about 1.3 mmol/L, with a mild right skew. Reported drinking days
per year are heavily right-skewed - a large group reports very few days and a
minority reports near-daily drinking - and the median BMI is in the overweight
range.

## Exploratory figure

```{r explore, fig.cap="Figure 1. HDL cholesterol against BMI, by sex, with LOESS smoothers."}
ggplot(dat_model, aes(bmi, hdl_cholesterol, colour = sex)) +
  geom_point(alpha = 0.45, size = 1.6) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 0.9) +
  scale_colour_manual(values = c(Female = "#B25D9C", Male = "#2E6F9E")) +
  labs(x = "BMI (kg/m2)", y = "HDL cholesterol (mmol/L)", colour = "Sex")
```

HDL falls steadily as BMI rises in both sexes, and the two sex curves are
separated by a near-constant vertical gap of roughly 0.3 mmol/L - women sit
higher at every level of BMI - which is exactly the additive pattern a linear
model without an interaction assumes.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv

utid <- map_dfr(predictors, function(v) {
  tidy(lm(reformulate(v, outcome), data = dat_model), conf.int = TRUE) |>
    filter(term != "(Intercept)")
})
ug <- function(term, d = 3) {
  r <- utid[utid$term == term, ]
  sprintf("%s mmol/L (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d), fmt(r$conf.high, d))
}
up <- function(term) pv(utid$p.value[utid$term == term])
```

Three predictors show strong crude associations. Men average `r ug("sexMale")`
lower HDL than women (p `r up("sexMale")`), which is the largest single
difference in the dataset. Each additional kg/m2 of BMI is associated with
`r ug("bmi")` lower HDL (p `r up("bmi")`), so a 10-unit BMI difference
corresponds to about `r fmt(abs(10 * utid$estimate[utid$term == "bmi"]), 2)`
mmol/L. Each additional drinking day per year is associated with
`r ug("alcohol_days_year", 4)` higher HDL (p `r up("alcohol_days_year")`); scaled
up, 100 extra drinking days a year corresponds to about
`r fmt(100 * utid$estimate[utid$term == "alcohol_days_year"], 2)` mmol/L. Age is
weakly positive (`r ug("age_yrs")`, p `r up("age_yrs")`). Physical activity,
smoking and diabetes show no clear crude association.

## Why adjustment is needed

**Sex** is the key confounder here. Men have markedly lower HDL than women and
men also drink on many more days per year and have different BMI distributions,
so the crude alcohol coefficient is pulled toward the null by sex (higher
drinking travels with lower-HDL male sex) and the crude BMI and activity
coefficients are similarly contaminated. **BMI** is the second confounder:
adiposity lowers HDL directly, and it is also associated with physical inactivity
and with diagnosed diabetes, so the crude activity and diabetes coefficients
partly reflect body size rather than the exposure of interest. **Age** confounds
the alcohol and activity associations, since drinking frequency and exercise both
change with age while HDL drifts upward. Mutual adjustment lets each coefficient
be read as a comparison between participants alike on the other measured
characteristics.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv

tid <- tidy(fit, conf.int = TRUE)
g <- function(term, d = 3) {
  r <- tid[tid$term == term, ]
  sprintf("%s mmol/L (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d), fmt(r$conf.high, d))
}
gp <- function(term) pv(tid$p.value[tid$term == term])
```

```{r fit-stats}
s <- summary(fit)
tibble(
  Statistic = c("Complete-case N", "Parameters (excl. intercept)",
                "R-squared", "Adjusted R-squared", "Residual SE (mmol/L)"),
  Value = c(as.character(nrow(dat_model)),
            as.character(length(coef(fit)) - 1L),
            fmt(s$r.squared, 3), fmt(s$adj.r.squared, 3), fmt(s$sigma, 3))
) |>
  gt() |>
  tab_header(title = md("**Model fit: multivariable linear model for HDL cholesterol**"))
```

The model explains `r fmt(100 * s$r.squared, 1)`% of the variance in HDL
(adjusted R-squared `r fmt(s$adj.r.squared, 3)`), by far the best fit of the four
continuous outcomes in this dataset. Among participants of the same age, BMI,
activity, smoking, drinking and diabetes status, men average `r g("sexMale")`
lower HDL than women (p `r gp("sexMale")`). Among participants alike on
everything else, each additional kg/m2 of BMI is associated with `r g("bmi")`
lower HDL (p `r gp("bmi")`), each extra drinking day per year with
`r g("alcohol_days_year", 4)` higher HDL (p `r gp("alcohol_days_year")`), and each
year of age with `r g("age_yrs")` higher HDL (p `r gp("age_yrs")`). Physical
activity (`r g("phys_activeYes")`, p `r gp("phys_activeYes")`), smoking
(`r g("smoked_100_cigsYes")`, p `r gp("smoked_100_cigsYes")`) and diabetes
(`r g("diabetesYes")`, p `r gp("diabetesYes")`) are not independently associated.
These are associations among the participants observed, not effects of
intervening.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

Sex moves *away* from the null, from `r ug("sexMale")` crude to `r g("sexMale")`
adjusted - an 18% increase in magnitude. This is masking rather than confounding:
men drink on more days, and drinking raises HDL, so the crude male deficit was
partly offset by the male drinking advantage; removing it makes the sex gap
larger. Diabetes moves substantially toward the null, from `r ug("diabetesYes")`
to `r g("diabetesYes")`, a two-thirds attenuation, because the crude diabetes
deficit was mostly the high BMI of participants with diabetes. Physical activity
also attenuates toward the null (`r ug("phys_activeYes")` to
`r g("phys_activeYes")`) for the same reason - active people are leaner, and once
BMI is held constant the residual activity association nearly disappears, which
is arguably over-adjustment since lower BMI is one of the mechanisms by which
exercise raises HDL. Two predictors barely move: BMI (`r ug("bmi")` to
`r g("bmi")`) and alcohol (`r ug("alcohol_days_year", 4)` to
`r g("alcohol_days_year", 4)`), both remaining highly significant. No predictor
changes significance status in either direction.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r het-check}
het <- check_heteroscedasticity(fit)
print(het)
```

```{r robust-se}
rob <- lmtest::coeftest(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
robci <- lmtest::coefci(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
tibble(
  Term = rownames(rob),
  Estimate = fmt(rob[, 1], 4),
  `SE (OLS)` = fmt(summary(fit)$coefficients[, 2], 4),
  `SE (HC3)` = fmt(rob[, 2], 4),
  `95% CI (HC3)` = paste0(fmt(robci[, 1], 3), " to ", fmt(robci[, 2], 3)),
  `p (HC3)` = formatC(rob[, 4], format = "f", digits = 3)
) |>
  filter(Term != "(Intercept)") |>
  gt() |>
  tab_header(title = md("**Table 3. HC3 heteroscedasticity-consistent standard errors.**"))
```

The residual-versus-fitted panel widens as fitted HDL rises, and
`check_heteroscedasticity()` rejects constant variance (p = 0.008): participants
predicted to have high HDL vary far more than those predicted to have low HDL,
which is the natural consequence of HDL having a floor near 0.5 mmol/L and no
ceiling. The normal-QQ panel shows a heavy right tail from a handful of
participants with HDL above 2.5 mmol/L, and `check_normality()` rejects.
Linearity of the BMI and age terms looks acceptable. Collinearity is a non-issue:
the highest VIF is 1.14. No observation crosses the Cook's distance contours.

Because variance is non-constant, Table 3 reports HC3 robust standard errors. For
BMI the HC3 SE is `r fmt(rob["bmi", 2], 4)` against an OLS SE of
`r fmt(summary(fit)$coefficients["bmi", 2], 4)`, and for alcohol
`r fmt(rob["alcohol_days_year", 2], 4)` against
`r fmt(summary(fit)$coefficients["alcohol_days_year", 2], 4)`. The differences are
small and every predictor keeps its significance status, so the conclusions are
robust to the variance assumption. A log transformation of HDL would remove both
the skew and most of the heteroscedasticity if a student wanted a cleaner model.

# Discussion

In this NHANES extract, sex, BMI, alcohol frequency and age are each
independently associated with HDL cholesterol, and together they explain
`r fmt(100 * s$r.squared, 1)`% of its variance. The sex difference is large -
men average about `r fmt(abs(tid$estimate[tid$term == "sexMale"]), 2)` mmol/L
lower than otherwise-similar women - and adiposity is the main modifiable
correlate, with roughly `r fmt(abs(10 * tid$estimate[tid$term == "bmi"]), 2)`
mmol/L lower HDL per 10 kg/m2 of BMI. The positive alcohol association is
consistent with the well-documented HDL-raising effect of regular moderate
drinking, but it is not a reason to recommend drinking. Practically, the results
reinforce that HDL should be interpreted against sex-specific thresholds, as
clinical guidelines already do.

**Limitations.** The models are unweighted, and the NHANES sampling weights,
strata and PSUs are not in this extract, so the sample **must not be described as
nationally representative**. The design is cross-sectional, so no temporal
ordering is established. `alcohol_days_year` counts *days*, not volume, so it
cannot separate a daily light drinker from a daily heavy drinker, and heavy
drinking has a different lipid profile from moderate drinking; it is also
self-reported and under-reporting is systematic. Physical activity is a bare
yes/no with no intensity or duration. Adjusting for BMI probably removes part of
the genuine effect of activity on HDL, since weight is on the causal path.
Complete-case analysis dropped `r n_dropped` of `r nrow(dat)` rows
(`r fmt(100 * n_dropped / nrow(dat), 0)`%), most of them for missing alcohol
data, and drinkers who decline to report are unlikely to be a random subset.
Residual variance is non-constant and residuals are right-skewed; HC3 SEs address
inference but the model still cannot predict the high-HDL tail. All estimates are
associations, not causal effects.

**Future work.** Refit on log(HDL) with the NHANES survey design, and use the
full public files' quantity-frequency drinking variables in place of drinking
days.

# Feasibility notes

*NOT part of a student report.*

- **The best of the four linear outcomes, and probably the best outcome in the
  dataset.** R-squared `r fmt(s$r.squared, 3)`, four clearly significant
  predictors with textbook biological interpretations, and a genuinely
  instructive unadjusted-versus-adjusted table containing both a masking effect
  (sex moves away from the null) and confounding (diabetes attenuates by two
  thirds).
- **Complete-case N `r nrow(dat_model)` of `r nrow(dat)`**
  (`r fmt(100 * n_dropped / nrow(dat), 0)`% lost), almost entirely because of
  `alcohol_days_year` (18% missing). Dropping alcohol raises N to about 445 but
  removes one of the two most interesting findings. This is a good, concrete
  illustration of the listwise-deletion trade-off.
- **Heteroscedasticity flagged (p = 0.008)** and non-normality of residuals. HC3
  changes nothing substantive. Clean case for the "check, then report robust SEs"
  workflow.
- **Trap: `total_cholesterol` should not be a predictor** - HDL is a component of
  it by definition. Nothing will error, but the coefficient is partly an identity.
  Less obvious to students than the `bmi`/`obese` trap.
- **Trap: `weight_kg` / `height_cm` / `bmi`** - `bmi` is exactly
  `weight_kg/(height_cm/100)^2`; do not use all three.
- **Adjusting for BMI blocks part of the activity pathway.** Worth flagging as an
  over-adjustment discussion point rather than an error.
- **Unweighted survey analysis.** Must be stated.
- No separation, no collinearity (max VIF 1.14), no influential observations.
