---
title: "Recreational physical activity in US adults"
subtitle: "Dataset 10: NHANES adults (2009-10 and 2011-12 extract)"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "nhanes_adults.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names. phys_active is the
# OUTCOME in this file, so it is deliberately absent from the predictor set even
# though it is a suggested predictor for other outcomes in this catalogue.
outcome    <- "phys_active"
predictors <- c("age_yrs", "sex", "race", "education", "poverty_ratio",
                "bmi", "general_health", "smoked_100_cigs")

lab <- function(x, l) { attr(x, "label") <- l; x }

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex   = lab(factor(sex, levels = c("female", "male"), labels = c("Female", "Male")),
                "Sex"),
    race  = lab(fct_relevel(factor(race), "White"), "Race/ethnicity"),
    education = lab(factor(education,
                           levels = c("College Grad", "Some College", "High School",
                                      "9 - 11th Grade", "8th Grade")),
                    "Education"),
    # Excellent (n = 47) and Poor (n = 11) are collapsed with their neighbours so no
    # cell is too thin to estimate; the reference is the healthiest group.
    general_health = lab(factor(case_when(
      general_health %in% c("Excellent", "Vgood") ~ "Excellent/Very good",
      general_health == "Good"                    ~ "Good",
      general_health %in% c("Fair", "Poor")       ~ "Fair/Poor"),
      levels = c("Excellent/Very good", "Good", "Fair/Poor")),
      "Self-rated general health"),
    smoked_100_cigs = lab(factor(smoked_100_cigs, levels = c(0, 1), labels = c("No", "Yes")),
                          "Ever smoked 100+ cigarettes"),
    age_yrs = lab(age_yrs, "Age (years)"),
    bmi = lab(bmi, "BMI (kg/m2)"),
    poverty_ratio = lab(poverty_ratio, "Income-to-poverty ratio"),
    phys_active = lab(phys_active, "Recreationally physically active")
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)
n_events  <- sum(dat_model[[outcome]])

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
pv  <- function(p) ifelse(p < 0.001, "<0.001", paste0("= ", fmt(p, 3)))
```

# Introduction

The National Health and Nutrition Examination Survey (NHANES) is a continuous
cross-sectional survey of the civilian non-institutionalised US population,
pairing a household interview with a standardised examination. This teaching
extract holds `r nrow(dat)` adults aged 21 to 80 from the 2009-10 and 2011-12
cycles, of whom `r sum(dat_raw$phys_active)`
(`r fmt(100 * mean(dat_raw$phys_active), 1)`%) report doing moderate or vigorous
recreational physical activity. Physical inactivity is a leading modifiable risk
factor for cardiovascular disease and diabetes, and describing *who* is inactive
is the standard first step in targeting activity promotion programmes.

**Aim.** Among these US adults, we examined which of age, sex, race/ethnicity,
education, income-to-poverty ratio, BMI, self-rated general health and smoking
history are independently associated with reporting recreational physical
activity, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.
Note that this is a **behavioural surveillance question, not a clinical one**: the
outcome is a self-reported behaviour, and the sensible reading of the results is
descriptive (which groups report low activity) rather than causal.

# Methods

**Data source.** NHANES 2009-10 and 2011-12, a complex multistage probability
sample of the US civilian non-institutionalised population. This extract contains
`r nrow(dat)` adults with 31 variables. The activity question was asked in the
household interview.

**Variables.** The outcome is `phys_active`, a binary indicator where the **event
is reporting moderate or vigorous recreational physical activity**
(`r sum(dat_raw$phys_active)` events in the full file). Predictors are age
(years), sex (reference female), race/ethnicity (reference White, the largest
group), education (reference College Grad, the largest group), income-to-poverty
ratio, BMI (kg/m2), self-rated general health collapsed to three levels
(reference Excellent/Very good) and ever having smoked 100 cigarettes (reference
no).

**`phys_active` is the outcome here and is therefore not a predictor.** It is
listed as a suggested predictor elsewhere in this catalogue; in this file it has
been removed from the predictor set entirely.

**Parameter budget.** There are `r sum(dat_raw$phys_active)` events in the file
(and `r nrow(dat_raw) - sum(dat_raw$phys_active)` non-events, which is the
binding number when the split is near 50/50), supporting roughly 24 parameters at
ten per parameter. This model uses 15 (race costs four, education four, general
health two, everything else one), well inside the budget - the most generous of
any logistic outcome in this dataset.

`weight_kg` and `height_cm` are excluded because `bmi` is exactly
`weight_kg / (height_cm/100)^2` in this file; `obese` is excluded because it is
exactly `bmi >= 30`.

**Statistical analysis.** `glm(family = binomial)`, justified because the outcome
is binary. We fitted a univariable model for each predictor first, then a single
multivariable model containing all of them. Predictors were chosen in advance on
subject-matter grounds, not by significance. Coefficients are exponentiated and
reported as odds ratios with 95% confidence intervals; significance threshold
0.05. Model adequacy was assessed with `performance::check_model()`, focusing on
binned residuals, collinearity, influential observations and events per
parameter. Models are **unweighted**: the NHANES survey weights, strata and
primary sampling units are not carried in this extract, so no result here is a
nationally representative estimate. Analysis used `r R.version.string` with
tidyverse, gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped), leaving `r n_events` events. The loss
comes from `general_health` (`r sum(is.na(dat$general_health))` missing) and
`poverty_ratio` (`r sum(is.na(dat$poverty_ratio))` missing). Only
`r sum(complete.cases(dat_raw))` of `r nrow(dat)` rows are complete on *every*
variable in the file, so listwise deletion is a live threat and the complete-case
N is reported for every model.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  mutate(phys_active = factor(phys_active, levels = c(0, 1),
                              labels = c("Not active", "Active"))) |>
  tbl_summary(
    by = phys_active,
    missing = "no"
  ) |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics by physical activity.**")
```

The clearest difference is educational: college graduates make up a much larger
share of the active group than of the inactive group. Active participants are
also younger, leaner, better off on the income-to-poverty ratio and more likely
to rate their health as Excellent or Very good. Smoking history is similar in the
two groups.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Proportion reporting recreational physical activity by education and sex."}
dat_model |>
  group_by(education, sex) |>
  summarise(p = mean(phys_active), n = n(), .groups = "drop") |>
  ggplot(aes(education, p, fill = sex)) +
  geom_col(position = position_dodge(0.8), width = 0.72) +
  geom_text(aes(label = paste0("n=", n)), position = position_dodge(0.8),
            vjust = -0.4, size = 2.8, colour = "grey30") +
  scale_y_continuous(labels = scales::percent, expand = expansion(c(0, 0.12))) +
  scale_fill_manual(values = c(Female = "#B25D9C", Male = "#2E6F9E")) +
  labs(x = "Education", y = "Proportion physically active", fill = "Sex") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

There is a steep educational gradient: roughly two thirds of college graduates
report recreational activity against about a quarter of those with 8th-grade
education, in both sexes, and the lowest education cells are small (n as low as
`r dat_model |> count(education, sex) |> pull(n) |> min()`).

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = glm,
    method.args = list(family = binomial),
    exponentiate = TRUE
  )
uv

utid <- map_dfr(predictors, function(v) {
  tidy(glm(reformulate(v, outcome), data = dat_model, family = binomial),
       conf.int = TRUE, exponentiate = TRUE) |>
    filter(term != "(Intercept)")
})
ug <- function(term, d = 2) {
  r <- utid[utid$term == term, ]
  sprintf("OR %s (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d), fmt(r$conf.high, d))
}
up <- function(term) pv(utid$p.value[utid$term == term])
```

Education dominates the crude picture. Relative to college graduates, the odds of
reporting activity are `r ug("educationSome College")` for some college
(p `r up("educationSome College")`), `r ug("educationHigh School")` for high
school (p `r up("educationHigh School")`),
`r ug("education9 - 11th Grade")` for 9th-11th grade
(p `r up("education9 - 11th Grade")`) and `r ug("education8th Grade")` for
8th grade (p `r up("education8th Grade")`) - a monotone gradient spanning more
than a fivefold difference in odds. Each unit of income-to-poverty ratio is
associated with `r ug("poverty_ratio")` (p `r up("poverty_ratio")`), each year of
age with `r ug("age_yrs")` (p `r up("age_yrs")`) and each kg/m2 of BMI with
`r ug("bmi")` (p `r up("bmi")`). Worse self-rated health is associated with lower
odds (Good `r ug("general_healthGood")`, p `r up("general_healthGood")`;
Fair/Poor `r ug("general_healthFair/Poor")`, p `r up("general_healthFair/Poor")`).
Sex and smoking show no crude association.

## Why adjustment is needed

**Education** is the confounder that matters most here, because it is associated
with almost every other predictor. Better-educated participants have higher
income-to-poverty ratios by construction of the labour market, lower BMI, better
self-rated health and a different age and ethnic distribution; since education is
also strongly associated with the outcome, the crude income, BMI and general-health
coefficients are all partly educational gradients in disguise. **Income-to-poverty
ratio** confounds the race/ethnicity contrasts for the same reason: minority
groups in this sample have lower income-to-poverty ratios, and low income is
associated with low reported activity, so a crude Mexican-versus-White comparison
partly measures income. Mutual adjustment lets each odds ratio be read as a
comparison between participants alike on the other measured characteristics.

A caveat that adjustment cannot fix: **BMI and self-rated general health are
plausibly consequences of physical activity rather than causes of it.** In a
cross-sectional file we cannot tell which came first, and conditioning on a
downstream variable biases the remaining coefficients unpredictably. They are
retained here because the question is descriptive - who reports low activity? -
rather than causal, but no coefficient in this model should be read as an effect.

## Multivariable model

```{r multivariable}
fit <- glm(reformulate(predictors, outcome), data = dat_model, family = binomial)

mv <- tbl_regression(fit, exponentiate = TRUE)
mv

tid <- tidy(fit, conf.int = TRUE, exponentiate = TRUE)
g <- function(term, d = 2) {
  r <- tid[tid$term == term, ]
  sprintf("OR %s (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d), fmt(r$conf.high, d))
}
gp <- function(term) pv(tid$p.value[tid$term == term])
```

```{r fit-stats}
n_par  <- length(coef(fit)) - 1L
epp    <- min(n_events, nrow(dat_model) - n_events) / n_par
null_fit <- glm(reformulate("1", outcome), data = dat_model, family = binomial)
tibble(
  Statistic = c("Complete-case N", "Events (active = 1)", "Non-events",
                "Parameters (excl. intercept)",
                "Events per parameter (smaller class)", "AIC (null model)",
                "AIC (full model)", "Tjur's R-squared"),
  Value = c(as.character(nrow(dat_model)), as.character(n_events),
            as.character(nrow(dat_model) - n_events), as.character(n_par),
            fmt(epp, 1), fmt(AIC(null_fit), 1), fmt(AIC(fit), 1),
            fmt(as.numeric(r2_tjur(fit)), 3))
) |>
  gt() |>
  tab_header(title = md("**Model fit: multivariable logistic model for physical activity**"))
```

The outcome is close to balanced (`r n_events` active,
`r nrow(dat_model) - n_events` not), so the binding count for the
events-per-parameter rule is the smaller class:
`r min(n_events, nrow(dat_model) - n_events)` for `r n_par` parameters, i.e.
**`r fmt(epp, 1)` per parameter**, above the conventional minimum of ten. AIC
falls from `r fmt(AIC(null_fit), 1)` for the intercept-only model to
`r fmt(AIC(fit), 1)`, a drop of `r fmt(AIC(null_fit) - AIC(fit), 1)`; Tjur's
R-squared is `r fmt(as.numeric(r2_tjur(fit)), 3)`.

Among participants of the same age, sex, race, income, BMI, health rating and
smoking history, the educational gradient persists almost undiminished: relative
to college graduates, the odds of reporting activity are
`r g("educationSome College")` for some college (p `r gp("educationSome College")`),
`r g("educationHigh School")` for high school (p `r gp("educationHigh School")`)
and `r g("education9 - 11th Grade")` for 9th-11th grade
(p `r gp("education9 - 11th Grade")`). Each year of age is associated with
`r g("age_yrs")` (p `r gp("age_yrs")`) and each kg/m2 of BMI with `r g("bmi")`
(p `r gp("bmi")`). Income-to-poverty ratio (`r g("poverty_ratio")`), self-rated
health (Good `r g("general_healthGood")`; Fair/Poor
`r g("general_healthFair/Poor")`), sex (`r g("sexMale")`), smoking
(`r g("smoked_100_cigsYes")`) and all race/ethnicity contrasts are no longer
independently associated. These are associations among the participants observed,
not effects of intervening.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted odds ratios.**")
```

Two predictors are wiped out by adjustment. Self-rated health Fair/Poor moves
from `r ug("general_healthFair/Poor")` (p `r up("general_healthFair/Poor")`) to
`r g("general_healthFair/Poor")` (p `r gp("general_healthFair/Poor")`) - more than
three quarters of the way to the null - and income-to-poverty ratio moves from
`r ug("poverty_ratio")` (p `r up("poverty_ratio")`) to `r g("poverty_ratio")`
(p `r gp("poverty_ratio")`). Both are explained by education: the crude health and
income associations were mostly the education gradient showing through. The
Mexican-versus-White contrast tells the same story, moving from
`r ug("raceMexican")` (p `r up("raceMexican")`) to `r g("raceMexican")`
(p `r gp("raceMexican")`) and losing significance, which is confounding by
education and income. By contrast the education contrasts themselves barely move
(some college `r ug("educationSome College")` to `r g("educationSome College")`;
9th-11th grade `r ug("education9 - 11th Grade")` to
`r g("education9 - 11th Grade")`), and age is almost unchanged
(`r ug("age_yrs")` to `r g("age_yrs")`). BMI attenuates only slightly, from
`r ug("bmi")` to `r g("bmi")`, and stays significant. Nothing crosses into
significance on adjustment.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r binned-resid}
br <- binned_residuals(fit)
print(br)
```

```{r collin}
check_collinearity(fit)
```

The binned-residual check passes: about 95% of binned residuals fall inside the
error bounds, so predicted probabilities are reasonably well calibrated across
the range - helped by the near-50/50 outcome split, which spreads predictions
across the middle of the probability scale where the model is best behaved.
Collinearity is acceptable but not negligible: the highest VIF is about 1.8, for
`education`, followed by 1.7 for race and 1.3 for income-to-poverty ratio. Those
first two are four-degree-of-freedom terms, and their *adjusted* VIFs (which
rescale for the number of parameters) are both about 1.1, so the inflation is
mild. Still, education and income are correlated enough that the income standard
error is inflated, which contributes to that coefficient losing significance: the
null income result should be read as "not separable from education" rather than
"no association". No observation crosses the influence
contours (maximum Cook's distance `r fmt(max(cooks.distance(fit)), 3)`, maximum
leverage `r fmt(max(hatvalues(fit)), 3)`). There is no separation: all
coefficients are finite with standard errors below
`r fmt(max(summary(fit)$coefficients[-1, 2]), 2)`, and every category contains
both active and inactive participants - although the 8th-grade stratum has only
`r sum(dat_model$education == "8th Grade" & dat_model$phys_active == 1)` active
participants out of `r sum(dat_model$education == "8th Grade")`, which is why its
confidence interval is by far the widest in the table. Homogeneity of variance
and normality of residuals do not apply to a logistic model and are not
interpreted.

# Discussion

In this NHANES extract, education is the characteristic most strongly and
independently associated with reporting recreational physical activity: adults
who did not complete high school have roughly a quarter of the odds of
college graduates, and the gradient is monotone across the four categories.
Younger age and lower BMI are also independently associated. The crude income and
self-rated-health associations are almost entirely explained by education.
Practically, if the aim is to raise population activity, education level is a
sharper targeting variable than income or self-reported health - though it is
better understood as a marker for the time, money, safety and facilities that
make recreational exercise possible than as something that acts on its own.

**Limitations.** The models are unweighted, and the NHANES sampling weights,
strata and PSUs are not in this extract, so the sample **must not be described as
nationally representative**. The outcome is a bare self-reported yes/no with no
frequency, duration or intensity, so it cannot distinguish a weekly walk from
daily training, and it excludes occupational and transport activity entirely -
which biases the comparison against manual workers, who are concentrated in the
low-education groups, and therefore probably exaggerates the education gradient.
Social-desirability bias in self-reported exercise is well documented and may
differ by education. The design is cross-sectional and two of the predictors
(BMI, self-rated health) are as plausibly *consequences* of activity as causes of
it, so the model conditions on downstream variables and no coefficient can be
read causally. Complete-case analysis dropped `r n_dropped` of `r nrow(dat)` rows
(`r fmt(100 * n_dropped / nrow(dat), 0)`%). The 8th-grade and Other/Hispanic
strata are small, so those contrasts are imprecise. Income-to-poverty ratio and
education are collinear enough (VIF ~1.8) that their separate contributions
cannot really be distinguished. All estimates are associations, not causal
effects.

**Future work.** Refit with the NHANES survey design and with the accelerometer
or activity-minutes variables from the full public files, which would replace the
yes/no outcome with a graded measure.

# Feasibility notes

*NOT part of a student report.*

- **Statistically the most comfortable logistic outcome in this dataset, but the
  weakest as a research question.** The outcome is near-balanced
  (`r n_events`/`r nrow(dat_model) - n_events`), which gives the largest parameter
  budget (EPP `r fmt(epp, 1)` with `r n_par` parameters), binned residuals pass
  at ~95%, and there is a clean, monotone, highly significant education gradient
  that barely moves on adjustment. Everything runs.
- **BE CRITICAL: this is not really a health-outcome question.** "What predicts
  whether someone exercises?" is a behavioural-surveillance question. It is
  legitimate for targeting, but a student expecting to say something clinical
  will find there is nothing to say beyond "richer, better-educated, younger,
  leaner people report more exercise", which is already well known.
- **MAJOR: the model conditions on downstream variables.** BMI and
  `general_health` are as plausibly consequences of activity as causes.
  Adjusting for them is over-adjustment, and the "adjusted" BMI odds ratio has no
  coherent causal reading. A student will almost certainly write it up as though
  high BMI reduces exercise. This is the main reason to steer students away.
- **`phys_active` is a suggested *predictor* elsewhere in the catalogue.** When it
  is the outcome it must be dropped from the predictor set. Easy to get wrong if
  a student copies a predictor list between analyses.
- **Outcome measurement is poor.** Bare yes/no, recreational only, self-reported.
  Occupational activity is excluded, which systematically misclassifies manual
  workers as inactive and inflates the education gradient - the headline finding
  is partly an artefact of the question wording.
- **Collinearity of note:** highest VIF ~1.8 (`education`, adjusted VIF 1.08),
  with `poverty_ratio` at 1.3. Not fatal, but the education-income correlation is
  why income goes null.
- **Sparse-ish cell:** 8th grade has only
  `r sum(dat_model$education == "8th Grade" & dat_model$phys_active == 1)`
  active of `r sum(dat_model$education == "8th Grade")`; its CI is very wide. No
  separation.
- **Complete-case N `r nrow(dat_model)` of `r nrow(dat)`**
  (`r fmt(100 * n_dropped / nrow(dat), 0)`% lost), from `general_health` and
  `poverty_ratio`.
- **Unweighted survey analysis.** Must be stated.
