---
title: "Reported trouble sleeping in US adults"
subtitle: "Dataset 10: NHANES adults (2009-10 and 2011-12 extract)"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "nhanes_adults.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "sleep_trouble"
predictors <- c("age_yrs", "sex", "bmi", "days_ment_health_bad", "general_health",
                "phys_active", "smoked_100_cigs", "sleep_hours")

lab <- function(x, l) { attr(x, "label") <- l; x }

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex   = lab(factor(sex, levels = c("female", "male"), labels = c("Female", "Male")),
                "Sex"),
    # Excellent (n = 47) and Poor (n = 11) are collapsed with their neighbours so no
    # cell is too thin to estimate; the reference is the healthiest group.
    general_health = lab(factor(case_when(
      general_health %in% c("Excellent", "Vgood") ~ "Excellent/Very good",
      general_health == "Good"                    ~ "Good",
      general_health %in% c("Fair", "Poor")       ~ "Fair/Poor"),
      levels = c("Excellent/Very good", "Good", "Fair/Poor")),
      "Self-rated general health"),
    smoked_100_cigs = lab(factor(smoked_100_cigs, levels = c(0, 1), labels = c("No", "Yes")),
                          "Ever smoked 100+ cigarettes"),
    phys_active = lab(factor(phys_active, levels = c(0, 1), labels = c("No", "Yes")),
                      "Recreationally physically active"),
    age_yrs = lab(age_yrs, "Age (years)"),
    bmi = lab(bmi, "BMI (kg/m2)"),
    days_ment_health_bad = lab(days_ment_health_bad, "Poor mental health days (of 30)"),
    sleep_hours = lab(sleep_hours, "Sleep (hours/night)"),
    sleep_trouble = lab(sleep_trouble, "Told a doctor about trouble sleeping")
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)
n_events  <- sum(dat_model[[outcome]])

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
pv  <- function(p) ifelse(p < 0.001, "<0.001", paste0("= ", fmt(p, 3)))
```

# Introduction

The National Health and Nutrition Examination Survey (NHANES) is a continuous
cross-sectional survey of the civilian non-institutionalised US population,
pairing a household interview with a standardised examination. This teaching
extract holds `r nrow(dat)` adults aged 21 to 80 from the 2009-10 and 2011-12
cycles, of whom `r sum(dat_raw$sleep_trouble)`
(`r fmt(100 * mean(dat_raw$sleep_trouble), 1)`%) report having told a doctor
about trouble sleeping. Sleep problems are common, under-treated, and tightly
bound up with mental health and chronic disease, so knowing which characteristics
mark higher prevalence is useful for primary-care case-finding.

**Aim.** Among these US adults, we examined which of age, sex, BMI, days of poor
mental health, self-rated general health, recreational physical activity, smoking
history and usual sleep duration are independently associated with having
reported trouble sleeping to a doctor, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** NHANES 2009-10 and 2011-12, a complex multistage probability
sample of the US civilian non-institutionalised population. This extract contains
`r nrow(dat)` adults with 31 variables. Sleep questions were asked in the
household interview.

**Variables.** The outcome is `sleep_trouble`, a binary indicator where the
**event is having told a doctor about trouble sleeping**
(`r sum(dat_raw$sleep_trouble)` events in the full file). Predictors are age
(years), sex (reference female), BMI (kg/m2), number of days of poor mental
health in the past 30, self-rated general health collapsed to three levels
(reference Excellent/Very good), recreational moderate or vigorous physical
activity (reference no), ever having smoked 100 cigarettes (reference no) and
usual sleep hours per night.

**Parameter budget.** There are `r sum(dat_raw$sleep_trouble)` events in the
file, which at ten events per parameter supports about 13 parameters. This model
uses `r length(predictors) + 1L` (general health costs two, everything else one),
comfortably inside the budget.

`weight_kg` and `height_cm` are excluded because `bmi` is exactly
`weight_kg / (height_cm/100)^2` in this file; `obese` is excluded because it is
exactly `bmi >= 30` and would duplicate BMI.

**Statistical analysis.** `glm(family = binomial)`, justified because the outcome
is binary. We fitted a univariable model for each predictor first, then a single
multivariable model containing all of them. Predictors were chosen in advance on
subject-matter grounds, not by significance. Coefficients are exponentiated and
reported as odds ratios with 95% confidence intervals; significance threshold
0.05. Model adequacy was assessed with `performance::check_model()`, focusing on
binned residuals, collinearity, influential observations and events per
parameter. Models are **unweighted**: the NHANES survey weights, strata and
primary sampling units are not carried in this extract, so no result here is a
nationally representative estimate. Analysis used `r R.version.string` with
tidyverse, gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped), leaving `r n_events` events. The loss
comes almost entirely from `general_health` and `days_ment_health_bad`, which are
missing for the same `r sum(is.na(dat$general_health))` participants. Only
`r sum(complete.cases(dat_raw))` of `r nrow(dat)` rows are complete on *every*
variable in the file, so listwise deletion is a live threat and the complete-case
N is reported for every model.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  mutate(sleep_trouble = factor(sleep_trouble, levels = c(0, 1),
                                labels = c("No sleep trouble", "Sleep trouble"))) |>
  tbl_summary(
    by = sleep_trouble,
    missing = "no"
  ) |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics by reported sleep trouble.**")
```

Participants reporting sleep trouble are more often female, report many more days
of poor mental health, rate their general health worse, sleep fewer hours per
night and are more likely to have smoked. Age and BMI differ much less between
the two groups.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Proportion reporting sleep trouble by self-rated general health and sex."}
dat_model |>
  group_by(general_health, sex) |>
  summarise(p = mean(sleep_trouble), n = n(), .groups = "drop") |>
  ggplot(aes(general_health, p, fill = sex)) +
  geom_col(position = position_dodge(0.8), width = 0.72) +
  geom_text(aes(label = paste0("n=", n)), position = position_dodge(0.8),
            vjust = -0.4, size = 2.9, colour = "grey30") +
  scale_y_continuous(labels = scales::percent, expand = expansion(c(0, 0.12))) +
  scale_fill_manual(values = c(Female = "#B25D9C", Male = "#2E6F9E")) +
  labs(x = "Self-rated general health", y = "Proportion reporting sleep trouble",
       fill = "Sex")
```

Reported sleep trouble becomes more common as self-rated health worsens, and
within every health category women report it more often than men - the two
gradients appear roughly additive, which is what the model without an interaction
assumes.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = glm,
    method.args = list(family = binomial),
    exponentiate = TRUE
  )
uv

utid <- map_dfr(predictors, function(v) {
  tidy(glm(reformulate(v, outcome), data = dat_model, family = binomial),
       conf.int = TRUE, exponentiate = TRUE) |>
    filter(term != "(Intercept)")
})
ug <- function(term, d = 2) {
  r <- utid[utid$term == term, ]
  sprintf("OR %s (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d), fmt(r$conf.high, d))
}
up <- function(term) pv(utid$p.value[utid$term == term])
```

Most predictors are associated with the outcome crudely. Each additional day of
poor mental health in the past 30 multiplies the odds of reported sleep trouble
by `r ug("days_ment_health_bad")` (p `r up("days_ment_health_bad")`), which over
a 10-day difference compounds to about
`r fmt(utid$estimate[utid$term == "days_ment_health_bad"]^10, 1)`-fold. Each
additional hour of usual sleep is associated with `r ug("sleep_hours")`
(p `r up("sleep_hours")`), a 30% lower odds per hour. Men have
`r ug("sexMale")` (p `r up("sexMale")`) the odds of women. Worse self-rated
health is associated with higher odds (Good `r ug("general_healthGood")`,
p `r up("general_healthGood")`; Fair/Poor `r ug("general_healthFair/Poor")`,
p `r up("general_healthFair/Poor")`), and ever-smokers have
`r ug("smoked_100_cigsYes")` (p `r up("smoked_100_cigsYes")`). Age
(`r ug("age_yrs")`) and BMI (`r ug("bmi")`) show weak but significant crude
associations; physical activity shows none.

## Why adjustment is needed

**Days of poor mental health** is the dominant confounder of the general-health
and smoking associations. Psychological distress is strongly associated with
rating one's overall health as Fair or Poor, is commoner among smokers, and is
one of the principal causes of insomnia; a crude comparison across self-rated
health categories therefore largely reflects the mental-health difference between
them rather than anything about physical health. **Sex** is the second
confounder: women in this sample report both more poor mental-health days and
more sleep trouble, so the crude mental-health coefficient carries a sex
component, and women's higher BMI also contaminates the crude BMI association.
**Sleep hours** confounds the general-health association in the same direction,
because short sleepers rate their health worse. Mutual adjustment lets each odds
ratio be read as a comparison between participants alike on the other measured
characteristics - with the caveat that mental health and sleep trouble are so
entangled that adjusting for one when studying the other is close to adjusting
for a different manifestation of the same syndrome.

## Multivariable model

```{r multivariable}
fit <- glm(reformulate(predictors, outcome), data = dat_model, family = binomial)

mv <- tbl_regression(fit, exponentiate = TRUE)
mv

tid <- tidy(fit, conf.int = TRUE, exponentiate = TRUE)
g <- function(term, d = 2) {
  r <- tid[tid$term == term, ]
  sprintf("OR %s (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d), fmt(r$conf.high, d))
}
gp <- function(term) pv(tid$p.value[tid$term == term])
```

```{r fit-stats}
n_par  <- length(coef(fit)) - 1L
epp    <- n_events / n_par
null_fit <- glm(reformulate("1", outcome), data = dat_model, family = binomial)
tibble(
  Statistic = c("Complete-case N", "Events (sleep trouble = 1)", "Parameters (excl. intercept)",
                "Events per parameter", "AIC (null model)", "AIC (full model)",
                "Tjur's R-squared"),
  Value = c(as.character(nrow(dat_model)), as.character(n_events), as.character(n_par),
            fmt(epp, 1), fmt(AIC(null_fit), 1), fmt(AIC(fit), 1),
            fmt(as.numeric(r2_tjur(fit)), 3))
) |>
  gt() |>
  tab_header(title = md("**Model fit: multivariable logistic model for sleep trouble**"))
```

The model has `r n_events` events for `r n_par` estimated parameters, i.e.
**`r fmt(epp, 1)` events per parameter**, comfortably above the conventional
minimum of ten. AIC falls from `r fmt(AIC(null_fit), 1)` for the intercept-only
model to `r fmt(AIC(fit), 1)`, a drop of `r fmt(AIC(null_fit) - AIC(fit), 1)`;
Tjur's R-squared is `r fmt(as.numeric(r2_tjur(fit)), 3)`.

Among participants alike on everything else, each additional day of poor mental
health multiplies the odds of reported sleep trouble by
`r g("days_ment_health_bad")` (p `r gp("days_ment_health_bad")`). Men have
`r g("sexMale")` the odds of otherwise-similar women (p `r gp("sexMale")`). Each
additional hour of usual sleep is associated with `r g("sleep_hours")`
(p `r gp("sleep_hours")`), ever-smoking with `r g("smoked_100_cigsYes")`
(p `r gp("smoked_100_cigsYes")`), each year of age with `r g("age_yrs")`
(p `r gp("age_yrs")`), and rating one's health Good rather than Excellent/Very
good with `r g("general_healthGood")` (p `r gp("general_healthGood")`). BMI
(`r g("bmi")`) and physical activity (`r g("phys_activeYes")`) are not
independently associated. These are associations among the participants observed,
not effects of intervening.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted odds ratios.**")
```

The two general-health contrasts move furthest toward the null. Fair/Poor falls
from `r ug("general_healthFair/Poor")` crude to `r g("general_healthFair/Poor")`
adjusted and loses significance entirely (p `r up("general_healthFair/Poor")` to
p `r gp("general_healthFair/Poor")`), and Good falls from
`r ug("general_healthGood")` to `r g("general_healthGood")`. This is exactly the
anticipated confounding by poor mental-health days: once distress is held
constant, self-rated physical health adds little. BMI also attenuates to the null
and loses significance, from `r ug("bmi")` to `r g("bmi")` - the crude BMI
association was carrying sex and general health. Two predictors barely move: sex
is identical in both columns (`r ug("sexMale")` to `r g("sexMale")`) and smoking
moves only slightly (`r ug("smoked_100_cigsYes")` to
`r g("smoked_100_cigsYes")`), so neither is explained by the other measured
characteristics. Age moves *away* from the null (`r ug("age_yrs")` to
`r g("age_yrs")`), a masking effect: older participants report fewer poor
mental-health days, which was hiding a genuine age gradient. Physical activity
moves from `r ug("phys_activeYes")` to `r g("phys_activeYes")` but is null in
both columns.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r binned-resid}
br <- binned_residuals(fit)
print(br)
```

```{r collin}
check_collinearity(fit)
```

The binned-residual check is borderline: about 81% of binned residuals fall
inside the error bounds, against the 95% target. The misses are concentrated in
the highest predicted-probability bins, where the model somewhat under-predicts
observed sleep trouble - the linear-in-log-odds assumption for
`days_ment_health_bad` is the likely culprit, since that variable is heavily
zero-inflated (a large majority report zero bad days) and its effect probably
saturates rather than compounding indefinitely. Collinearity is not a problem:
all VIFs are below 1.2 despite mental-health days and general health being
conceptually related. No observation crosses the influence contours (maximum
Cook's distance `r fmt(max(cooks.distance(fit)), 3)`, maximum leverage
`r fmt(max(hatvalues(fit)), 3)`). There is no separation: all coefficients are
finite with standard errors below
`r fmt(max(summary(fit)$coefficients[-1, 2]), 2)`, and every category of general
health contains both events and non-events. Homogeneity of variance and normality
of residuals do not apply to a logistic model and are not interpreted.

# Discussion

In this NHANES extract, poor mental health, female sex, short sleep, smoking and
older age are each independently associated with having reported trouble sleeping
to a doctor. Poor mental-health days is the strongest signal: ten extra bad days
in a month corresponds to roughly
`r fmt(tid$estimate[tid$term == "days_ment_health_bad"]^10, 1)`-fold higher odds
among otherwise-similar participants. The crude association with worse self-rated
general health is almost entirely explained by mental health. Practically, this
supports asking about mood whenever a patient raises sleep, and vice versa,
rather than treating sleep complaints as a marker of general physical ill health.

**Limitations.** The models are unweighted, and the NHANES sampling weights,
strata and PSUs are not in this extract, so the sample **must not be described as
nationally representative**. The outcome is not "has a sleep disorder" but "has
*told a doctor* about trouble sleeping", which mixes the underlying condition
with help-seeking behaviour and access to care; women and people with regular
healthcare contact are more likely to have had that conversation, so the sex
coefficient in particular is partly a care-seeking effect rather than a
prevalence difference. The design is cross-sectional, and the mental-health
association is almost certainly bidirectional - insomnia causes distress as
readily as distress causes insomnia - so it cannot be given a direction here.
`days_ment_health_bad` is heavily zero-inflated and is modelled as linear on the
log-odds scale, which the binned-residual check suggests is imperfect.
Complete-case analysis dropped `r n_dropped` of `r nrow(dat)` rows
(`r fmt(100 * n_dropped / nrow(dat), 0)`%), almost all because self-rated health
and mental-health days are missing together for the same participants, and people
who skip health-perception questions are unlikely to be a random subset. All
estimates are associations, not causal effects.

**Future work.** Refit with the NHANES survey design, and use a validated
insomnia instrument rather than a help-seeking question as the outcome.

# Feasibility notes

*NOT part of a student report.*

- **One of the better logistic outcomes here.** `r n_events` events in the
  analytic sample, `r n_par` parameters, EPP `r fmt(epp, 1)`, no separation, max
  VIF < 1.2, and several coefficients that move interestingly on adjustment
  (general health collapses, age is unmasked). Good material for a Purpose 3
  write-up.
- **Binned residuals borderline (~81% inside bounds vs 95%).** Traceable to the
  zero-inflated `days_ment_health_bad`. Worth flagging; a student can address it
  by categorising bad-days into 0 / 1-13 / 14+, which is a legitimate extension.
- **MAJOR conceptual caveat: the outcome measures help-seeking, not the
  condition.** "Told a doctor about trouble sleeping" conflates prevalence with
  access. The strong female excess is partly a care-seeking artefact. Students who
  do not notice this will over-claim.
- **Mental health and sleep trouble are near-tautologically related.** Adjusting
  for `days_ment_health_bad` when the outcome is sleep trouble is close to
  adjusting for a co-manifestation, not a confounder. It dominates the model and
  it hollows out the general-health coefficients. Defensible, but the discussion
  has to acknowledge it.
- **Complete-case N `r nrow(dat_model)` of `r nrow(dat)`**
  (`r fmt(100 * n_dropped / nrow(dat), 0)`% lost), entirely because
  `general_health` and `days_ment_health_bad` are missing for the same 48 people.
  Dropping either alone saves nothing; dropping both recovers all 48.
- **`general_health` was collapsed from 5 levels to 3.** As shipped, Poor has
  n = 11 and Excellent n = 47; with 5 levels the Poor contrast would be estimated
  off a handful of events. Students should be told to collapse it.
- **Trap: `obese` / `bmi` / `weight_kg` / `height_cm`** are deterministically
  related; use only `bmi`.
- **Unweighted survey analysis.** Must be stated.
