---
title: "Systolic blood pressure in US adults"
subtitle: "Dataset 10: NHANES adults (2009-10 and 2011-12 extract)"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "nhanes_adults.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "systolic_bp"
predictors <- c("age_yrs", "sex", "race", "bmi", "poverty_ratio",
                "smoked_100_cigs", "alcohol_days_year", "diabetes", "phys_active")

lab <- function(x, l) { attr(x, "label") <- l; x }

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex   = lab(factor(sex, levels = c("female", "male"), labels = c("Female", "Male")),
                "Sex"),
    race  = lab(fct_relevel(factor(race), "White"), "Race/ethnicity"),
    smoked_100_cigs = lab(factor(smoked_100_cigs, levels = c(0, 1), labels = c("No", "Yes")),
                          "Ever smoked 100+ cigarettes"),
    diabetes = lab(factor(diabetes, levels = c(0, 1), labels = c("No", "Yes")),
                   "Diagnosed diabetes"),
    phys_active = lab(factor(phys_active, levels = c(0, 1), labels = c("No", "Yes")),
                      "Recreationally physically active"),
    age_yrs = lab(age_yrs, "Age (years)"),
    bmi = lab(bmi, "BMI (kg/m2)"),
    poverty_ratio = lab(poverty_ratio, "Income-to-poverty ratio"),
    alcohol_days_year = lab(alcohol_days_year, "Drinking days in past year"),
    systolic_bp = lab(systolic_bp, "Systolic BP (mmHg)")
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
pv  <- function(p) ifelse(p < 0.001, "<0.001", paste0("= ", fmt(p, 3)))
```

# Introduction

The National Health and Nutrition Examination Survey (NHANES) is a continuous
cross-sectional survey of the civilian non-institutionalised US population,
combining a household interview with a standardised physical examination in a
mobile examination centre. This teaching extract holds `r nrow(dat)` adults aged
21 to 80 from the 2009-10 and 2011-12 cycles. Systolic blood pressure is the
single strongest modifiable determinant of stroke and ischaemic heart disease, so
knowing which measured characteristics track with it is directly relevant to
screening.

**Aim.** Among these US adults, we examined which of age, sex, race/ethnicity,
BMI, income-to-poverty ratio, smoking history, alcohol frequency, diagnosed
diabetes and recreational physical activity are independently associated with
systolic blood pressure, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** NHANES 2009-10 and 2011-12, a complex multistage probability
sample of the US civilian non-institutionalised population. This extract contains
`r nrow(dat)` adults (`r sum(dat$survey_cycle == "2009_10")` from 2009-10,
`r sum(dat$survey_cycle == "2011_12")` from 2011-12) with 31 variables. Blood
pressure was measured in the mobile examination centre; the file carries the mean
of the recorded readings, rounded to whole mmHg.

**Variables.** The outcome is `systolic_bp`, systolic blood pressure in mmHg,
treated as continuous. Predictors are age (years), sex (reference female),
race/ethnicity (reference White, the largest group), BMI (kg/m2),
income-to-poverty ratio (household income as a multiple of the federal poverty
line, higher = better off), ever having smoked 100 cigarettes (reference no),
number of drinking days in the past year, doctor-diagnosed diabetes (reference
no) and recreational moderate or vigorous physical activity (reference no).

`diastolic_bp` and `hypertensive` are deliberately excluded. `hypertensive` is
not an independent variable at all: in this file it reproduces
`systolic_bp >= 130 | diastolic_bp >= 80` exactly on every row where both
pressures are observed, so putting it in the model would be regressing systolic
pressure on a function of itself. `weight_kg` and `height_cm` are also excluded
because `bmi` is exactly `weight_kg / (height_cm/100)^2` in this file (largest
discrepancy 0.05 kg/m2, attributable to rounding), and the three cannot sit in
one model without near-perfect collinearity.

**Statistical analysis.** `lm()`, justified because the outcome is continuous and
approximately unimodal. We fitted a univariable model for each predictor first,
then a single multivariable model containing all of them. Predictors were chosen
in advance on subject-matter grounds, not by significance. Significance threshold
0.05; estimates are reported with 95% confidence intervals. Assumptions were
checked with `performance::check_model()`, and because non-constant variance was
flagged we additionally report HC3 heteroscedasticity-consistent standard errors.
Models are **unweighted**: the NHANES survey weights, strata and primary sampling
units are not carried in this extract, so no result here is a nationally
representative estimate. Analysis used `r R.version.string` with tidyverse,
gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped). The two costly predictors are
`alcohol_days_year` (`r sum(is.na(dat$alcohol_days_year))` missing) and
`poverty_ratio` (`r sum(is.na(dat$poverty_ratio))` missing); systolic pressure
itself is missing for `r sum(is.na(dat$systolic_bp))` participants. Only
`r sum(complete.cases(dat_raw))` of `r nrow(dat)` rows are complete on *every*
variable in the file, so listwise deletion is a live threat here and the
complete-case N is reported for every model.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(missing = "no") |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

The analytic sample has a median systolic pressure in the low 120s mmHg with a
long right tail, a median BMI in the overweight range, and roughly balanced sex.
Diagnosed diabetes is present in a small minority, and just under half report
recreational physical activity.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Systolic blood pressure against age, by sex, with LOESS smoothers."}
ggplot(dat_model, aes(age_yrs, systolic_bp, colour = sex)) +
  geom_point(alpha = 0.45, size = 1.6) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 0.9) +
  scale_colour_manual(values = c(Female = "#B25D9C", Male = "#2E6F9E")) +
  labs(x = "Age (years)", y = "Systolic blood pressure (mmHg)", colour = "Sex")
```

Systolic pressure rises steadily and close to linearly across the adult age range
in both sexes, with substantial scatter at every age and a visibly wider spread
among older participants.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv

utid <- map_dfr(predictors, function(v) {
  tidy(lm(reformulate(v, outcome), data = dat_model), conf.int = TRUE) |>
    filter(term != "(Intercept)")
})
ug <- function(term, d = 2) {
  r <- utid[utid$term == term, ]
  sprintf("%s mmHg (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d), fmt(r$conf.high, d))
}
up <- function(term) pv(utid$p.value[utid$term == term])
```

Age is the dominant crude predictor: each additional year of age is associated
with `r ug("age_yrs")` higher systolic pressure (p `r up("age_yrs")`), so a
30-year age difference corresponds to roughly
`r fmt(30 * utid$estimate[utid$term == "age_yrs"], 0)` mmHg. Diagnosed diabetes
is associated with `r ug("diabetesYes")` (p `r up("diabetesYes")`) higher
pressure crudely, and
participants in the "Other" race category average `r ug("raceOther")`
(p `r up("raceOther")`) lower pressure than White participants. BMI, sex,
income-to-poverty ratio, smoking, alcohol frequency and physical activity show no
clear crude association.

## Why adjustment is needed

The crude estimates are confounded by age above all. **Age** is associated with
almost every predictor here and with the outcome: older participants are far more
likely to have diagnosed diabetes (diabetes prevalence rises steeply with age)
and less likely to be recreationally active, and age independently raises
systolic pressure through arterial stiffening. The crude diabetes coefficient of
about 10 mmHg is therefore partly just the age difference between people with and
without diabetes. **BMI** is the second obvious confounder: it is strongly
associated with diabetes and with physical inactivity, and adiposity raises blood
pressure through sodium retention and sympathetic activation, so the crude
diabetes and activity coefficients also carry adiposity. **Sex** confounds the
alcohol association, because men drink on more days per year and men and women
differ in blood pressure at a given age. Mutual adjustment lets each coefficient
be read as a comparison between participants who are similar on the other
measured characteristics.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv

tid <- tidy(fit, conf.int = TRUE)
g <- function(term, d = 2) {
  r <- tid[tid$term == term, ]
  sprintf("%s mmHg (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d), fmt(r$conf.high, d))
}
gp <- function(term) pv(tid$p.value[tid$term == term])
```

```{r fit-stats}
s <- summary(fit)
tibble(
  Statistic = c("Complete-case N", "Parameters (excl. intercept)",
                "R-squared", "Adjusted R-squared", "Residual SE (mmHg)"),
  Value = c(as.character(nrow(dat_model)),
            as.character(length(coef(fit)) - 1L),
            fmt(s$r.squared, 3), fmt(s$adj.r.squared, 3), fmt(s$sigma, 2))
) |>
  gt() |>
  tab_header(title = md("**Model fit: multivariable linear model for systolic BP**"))
```

The model explains `r fmt(100 * s$r.squared, 1)`% of the variance in systolic
pressure (adjusted R-squared `r fmt(s$adj.r.squared, 3)`), which is modest but
typical for blood pressure with demographic and behavioural predictors only.
Among participants of the same sex, race, BMI, income level, smoking, drinking,
diabetes and activity status, each additional year of age is associated with
`r g("age_yrs")` higher systolic pressure (p `r gp("age_yrs")`). No other
predictor reaches significance: diabetes is associated with `r g("diabetesYes")`
(p `r gp("diabetesYes")`), BMI with `r g("bmi")` per kg/m2 (p `r gp("bmi")`), and
being physically active with `r g("phys_activeYes")` (p `r gp("phys_activeYes")`).
These are associations among the participants observed, not effects of
intervening on any of these variables.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

Two coefficients move a long way toward the null. Diabetes falls from
`r ug("diabetesYes")` crude to `r g("diabetesYes")` adjusted and loses
significance, a drop of about two thirds; this is confounding by age and BMI,
exactly as anticipated, and the residual estimate is too imprecise to interpret.
Race "Other" moves from `r ug("raceOther")` (significant) to `r g("raceOther")`
(not significant), a shift explained by the much younger age (mean 40 vs 52
years) and lower BMI (mean 24 vs 29 kg/m2) of that group, which is very small
here (n = `r sum(dat_model$race == "Other")`). Physical activity moves in
the opposite direction, from `r ug("phys_activeYes")` crude to
`r g("phys_activeYes")` adjusted: once age and BMI are held constant, active
participants have slightly *higher* rather than slightly lower pressure, a sign
that the crude estimate was masked by the fact that active people are younger and
leaner. Age itself barely moves (`r ug("age_yrs")` to `r g("age_yrs")`), which is
what you expect for a variable that confounds the others but is itself confounded
by nothing measured here. Nothing crosses into significance on adjustment.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r het-check}
het <- check_heteroscedasticity(fit)
print(het)
```

```{r robust-se}
rob <- lmtest::coeftest(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
robci <- lmtest::coefci(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
tibble(
  Term = rownames(rob),
  Estimate = fmt(rob[, 1], 3),
  `SE (OLS)` = fmt(summary(fit)$coefficients[, 2], 3),
  `SE (HC3)` = fmt(rob[, 2], 3),
  `95% CI (HC3)` = paste0(fmt(robci[, 1], 2), " to ", fmt(robci[, 2], 2)),
  `p (HC3)` = fmt(rob[, 4], 3)
) |>
  filter(Term != "(Intercept)") |>
  gt() |>
  tab_header(title = md("**Table 3. HC3 heteroscedasticity-consistent standard errors.**"))
```

The posterior-predictive and residual panels show the model captures the centre
of the distribution but under-predicts the upper tail: a handful of participants
with systolic pressures above 170 mmHg are poorly fitted. The residual-versus-
fitted panel fans out toward higher fitted values, and `check_heteroscedasticity()`
confirms non-constant error variance (p < 0.001) - the spread of systolic
pressure genuinely grows with its mean. The normal-QQ panel shows a heavy right
tail, consistent with the same few high-pressure participants. Collinearity is a
non-issue: the highest VIF is 1.31, for race/ethnicity. No point exceeds the Cook's distance contours,
so no single observation drives the fit.

Because variance is non-constant, Table 3 reports HC3 robust standard errors.
The HC3 SE for age is `r fmt(rob["age_yrs", 2], 3)` against an OLS SE of
`r fmt(summary(fit)$coefficients["age_yrs", 2], 3)`, so the age conclusion is
unchanged; no predictor changes its significance status under robust inference.
The point estimates are unaffected - only the uncertainty around them is
re-estimated.

# Discussion

In this NHANES extract, age is the only characteristic independently associated
with systolic blood pressure once the others are held constant, at roughly
`r fmt(tid$estimate[tid$term == "age_yrs"], 2)` mmHg per year. The crude diabetes
association, which looked large and highly significant, is substantially
explained by the age and BMI differences between participants with and without
diabetes. Practically, this reinforces that age is the dominant stratifier for
blood-pressure screening, and that a crude comparison of pressure across
diabetes status will badly overstate the difference attributable to diabetes
itself.

**Limitations.** The models are unweighted, and the NHANES sampling weights,
strata and PSUs are not in this extract, so the sample **must not be described as
nationally representative**: minority groups are oversampled by design and
standard errors ignore the clustered multistage design, which typically makes
them too small. The design is cross-sectional, so nothing here establishes
temporality - diabetes, activity and blood pressure were measured at the same
visit. Complete-case analysis dropped `r n_dropped` of `r nrow(dat)` rows
(`r fmt(100 * n_dropped / nrow(dat), 0)`%), mostly for missing alcohol and
income data, and if those data are not missing at random the estimates are
biased. Non-constant residual variance was detected and handled with HC3 SEs, but
the heavy right tail means predictions for hypertensive participants are poor.
Antihypertensive medication is not recorded in this extract, which is a serious
unmeasured confounder: treated participants have lower measured pressure
*because* of their higher underlying risk, biasing the diabetes and BMI
coefficients toward the null. Race/ethnicity categories are small outside White
(n = `r sum(dat_model$race != "White")` in total across four groups) so those
contrasts are imprecise. All estimates are associations, not causal effects.

**Future work.** Refit with the NHANES survey design using the `survey` package
and add antihypertensive medication use, which is available in the full public
NHANES files.

# Feasibility notes

*NOT part of a student report.*

- **Works well as a student project.** Continuous, well-measured outcome; one
  large, unambiguous, easily interpreted signal (age); several predictors that
  attenuate on adjustment, which is exactly the teaching point of Purpose 3.
- **Complete-case N drops to `r nrow(dat_model)` of `r nrow(dat)`
  (`r fmt(100 * n_dropped / nrow(dat), 0)`% lost).** Driven by
  `alcohol_days_year` (18% missing) and `poverty_ratio` (8% missing) plus the 20
  rows missing blood pressure. Dropping `alcohol_days_year` recovers about 60
  rows. Students will not notice this unless told to check.
- **Trap: `hypertensive` and `diastolic_bp` must be excluded.** `hypertensive` is
  an exact recomputation of `systolic_bp >= 130 | diastolic_bp >= 80` on all 468
  rows with both pressures observed (100% agreement; the 20 apparent exceptions
  are simply rows where BP is missing, so `hypertensive` is missing too). A
  student who includes it will get a huge R-squared and an incoherent model.
- **Trap: `bmi`, `weight_kg`, `height_cm`.** `bmi` is exactly
  `weight_kg/(height_cm/100)^2` (max absolute discrepancy 0.05, pure rounding).
  All three in one model gives severe collinearity.
- **Heteroscedasticity is real and flagged (p < 0.001)**, plus non-normal
  residuals with a heavy right tail. This is a feature, not a bug: it gives the
  student a genuine reason to run `check_heteroscedasticity()` and HC3, and the
  conclusions survive.
- **Low R-squared (`r fmt(s$r.squared, 3)`)** may disappoint students expecting a
  "good" model; worth pre-empting.
- **Unweighted survey analysis.** Must be stated. The extract carries no weights,
  strata or PSU, so a design-based analysis is impossible with this file alone.
- No separation, no collinearity (max VIF 1.31), no influential-point problems.
