---
title: "Regression diagnostics: SHIN checks for simple linear regression"
subtitle: "Reading check_model() diagnostic plots for shape, homoscedasticity, influential points, and normality"
author: "RSR Resources"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 7,
  fig.align = "center"
)
```

# Setup

```{r libraries}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, performance, qqplotr)
```

## SUNSHINE and SHIN

We created an acronym (SUNSHINE) to help you remember the eight checks you should keep in mind when fitting a linear regression model.

- **S** (**Sample size**): do you have enough observations for the model?
- **U** (**Uncorrelated errors**): are the residuals uncorrelated across observations (no patterns by time or space order)?
- **N** (**No multicollinearity**): are the predictors free of strong correlations with each other?
- **S** (**Shape / linearity**): is the relationship actually a straight line?
- **H** (**Homogeneity of variance**): does the spread of residuals stay constant?
- **I** (**Influential points**): is the fit stable if you drop any single point?
- **N** (**Normality of residuals**): are the prediction errors roughly bell-shaped?
- **E** (**Exogeneity**): are the predictors unrelated to anything the model left out?

This script focuses on **SHIN** (Shape, Homoscedasticity, Influential points, and Normality), because those are the four assumptions you can check directly in R with `check_model()` on a simple linear regression.

# Continuous predictor

Each section below walks through one SHIN check with a continuous predictor: a good example, a bad example, then a practice question.

## S: Shape (Linearity)

**The question:** is the relationship actually a straight line?

### Exercise hours and VO₂ max (linear relationship)

Weekly exercise hours vs. VO₂ max: a linear relationship.

```{r shape-good-data}
shape_good <- data.frame(
  exercise_hours = c(1, 1.5, 2, 2.5, 3, 3.5, 4, 4.5, 5, 5.5, 6, 6.5, 7, 7.5, 8, 8.5, 9, 9.5, 10, 10.5, 2, 3, 4.5, 6, 7, 8, 9, 1.5, 5, 8.5),
  vo2_max        = c(31.6, 29.4, 34.2, 30, 33.8, 30.6, 36.4, 34.2, 40, 34.8, 38.6, 35.4, 41.2, 36, 43.8, 40.6, 44.4, 41.2, 45, 41.8, 31.2, 34.8, 34.2, 40.6, 37.2, 41.8, 41.4, 32.4, 33, 42.6)
)
```

```{r shape-good-scatter}
ggplot(shape_good, aes(exercise_hours, vo2_max)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#C83B3B") +
  labs(title = "Exercise hours vs VO₂ max: linear relationship") +
  theme_minimal(base_size = 14)
```

```{r shape-good-fit}
shape_good_model <- lm(vo2_max ~ exercise_hours, data = shape_good)
summary(shape_good_model)
```

```{r shape-good-check}
check_model(shape_good_model)
```

The *Linearity* green line is flat and near zero, showing no leftover pattern.

### Study hours and quiz score (non-linear relationship)

Weekly study hours vs. quiz score.

Going from no studying to moderate studying helps a lot; extra hours past that add little because scores are near the maximum of the scale.

```{r shape-bad-data}
shape_bad <- data.frame(
  study_hours = c(2, 4, 6, 9, 11, 13, 15, 17, 18, 20, 21, 23, 26, 29, 31, 34, 37, 39, 41, 43, 46, 48, 51, 53, 54, 56, 58, 60, 61, 62),
  quiz_score  = c(19.5, 30, 46.2, 55.9, 65.9, 68.3, 76.5, 76.4, 81.6, 81.3, 86.3, 84.4, 91.2, 88.8, 93, 90.4, 95.6, 92.2, 95.9, 92.6, 96.2, 92.9, 97.5, 92.9, 97.4, 93.5, 97.1, 93.4, 97.8, 93.2)
)
```

Look at the raw scatter first; you can often already see the bend.

```{r shape-bad-scatter}
ggplot(shape_bad, aes(study_hours, quiz_score)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#C83B3B") +
  labs(title = "Study hours vs quiz score: non-linear relationship fitted with a straight line") +
  theme_minimal(base_size = 14)
```

Next, inspect the diagnostic panel, especially the Linearity panel.

```{r shape-bad-fit}
shape_bad_model <- lm(quiz_score ~ study_hours, data = shape_bad)
summary(shape_bad_model)
```

```{r shape-bad-check}
check_model(shape_bad_model)
```

The green line in the *Linearity* panel arches, indicating a non-linear relationship.

So your model overestimates the quiz score for students who study for a few hours, and underestimates it for students who study for a lot of hours.

In such cases, you may need to transform your data or use a different kind of model.

### Practice

Fertilizer applied (kg per plot) vs. crop yield (kg).

Fit the model, look at the scatter plot, then read the **Linearity** panel of the diagnostics. Does the relationship satisfy the linearity assumption?

```{r practice-1-data}
practice_1 <- data.frame(
  fertilizer = c(2, 4, 6, 8, 10, 12, 14, 16, 18, 20, 22, 24, 26, 28, 30, 32, 34, 36, 38, 40, 5, 11, 17, 23, 29, 35, 7, 15, 25, 33),
  yield      = c(13.9, 20.2, 32.7, 35, 42.6, 43.2, 50.9, 50.5, 56.6, 53.3, 58.1, 54.7, 60.6, 57.7, 62.3, 58.2, 61.6, 58.4, 63.1, 59.8, 26.9, 44.7, 51.1, 59.2, 57.5, 62.2, 32.5, 52.3, 56.6, 61.4)
)
```

```{r practice-1-scatter}
ggplot(practice_1, aes(fertilizer, yield)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#C83B3B") +
  labs(title = "Fertilizer vs crop yield") +
  theme_minimal(base_size = 14)
```

```{r practice-1-fit}
practice_1_model <- lm(yield ~ fertilizer, data = practice_1)
summary(practice_1_model)
```

```{r practice-1-check}
check_model(practice_1_model)
```

**Your answer here:**

------------------------------------------------------------------------

------------------------------------------------------------------------

## H: Homoscedasticity (Equal Variance)

The question here is: does the *spread* of the residuals stay constant across the range?

For this, we read the **Homogeneity of variance** panel (scale-location).

A flat green line means even spread. If the line slopes upward, then the spread of the residuals increases with the fitted value.

### Exercise duration and calories (homoscedastic variance)

Exercise duration vs. calories burned.

The spread looks the same for short and long workouts.

```{r hom-good-data}
hom_good <- data.frame(
  exercise_min = c(15, 20, 22, 25, 28, 30, 33, 35, 37, 40, 42, 45, 48, 50, 52, 55, 58, 60, 63, 65, 18, 27, 34, 39, 44, 49, 54, 57, 62, 36),
  calories     = c(149, 158, 193, 185, 218, 210, 250, 246, 281, 271, 299, 298, 338, 334, 366, 360, 393, 386, 426, 419, 167, 197, 255, 261, 311, 323, 377, 374, 419, 247)
)
```

```{r hom-good-scatter}
ggplot(hom_good, aes(exercise_min, calories)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#C83B3B") +
  labs(title = "Exercise duration vs calories: even spread across the range") +
  theme_minimal(base_size = 14)
```

```{r hom-good-fit}
hom_good_model <- lm(calories ~ exercise_min, data = hom_good)
summary(hom_good_model)
```

```{r hom-good-check}
check_model(hom_good_model)
```

Here the *Homogeneity* green line is roughly flat.

This means that the spread of the residuals is roughly constant across the range of the predictor.

### Income and medical spending (heteroscedasticity)

Household income vs. out-of-pocket medical spending.

Wealthier households have far more discretionary spending, so the spread gets larger at the high end, giving you a fan-shaped scatter.

```{r hom-bad-data}
hom_bad <- data.frame(
  income_k      = c(20, 25, 28, 32, 35, 38, 42, 45, 48, 52, 55, 58, 62, 65, 68, 72, 75, 80, 85, 90, 95, 100, 110, 120, 22, 40, 56, 70, 88, 105),
  medical_spend = c(464, 475, 573, 622, 495, 786, 593, 943, 744, 828, 682, 1052, 1000, 919, 842, 1294, 851, 1526, 961, 994, 1573, 1052, 2024, 2030, 555, 560, 1152, 718, 1729, 855)
)
```

```{r hom-bad-scatter}
ggplot(hom_bad, aes(income_k, medical_spend)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#C83B3B") +
  labs(title = "Income vs medical spending: residual spread increases with income") +
  theme_minimal(base_size = 14)
```

```{r hom-bad-fit}
hom_bad_model <- lm(medical_spend ~ income_k, data = hom_bad)
summary(hom_bad_model)
```

```{r hom-bad-check}
check_model(hom_bad_model)
```

The *Homogeneity of variance* green line slopes upward, indicating that the spread of the residuals increases with the fitted value.

This is a sign of heteroscedasticity.

It means that your model's standard errors are too large for high-income households, and too small for low-income households.

### Practice

Daily patient volume at a clinic vs. patient waiting time (minutes).

Fit the model, look at the scatter plot, then read the **Homogeneity of variance** panel. Does the relationship satisfy the equal-variance assumption?

```{r practice-2-data}
practice_2 <- data.frame(
  patient_volume = c(10, 14, 18, 22, 26, 30, 34, 38, 42, 46, 50, 54, 58, 62, 66, 70, 74, 78, 82, 90, 12, 28, 44, 60, 76, 86, 16, 36, 56, 72),
  wait_time      = c(23, 31, 28, 41, 33, 47, 35, 63, 41, 71, 40, 78, 93, 48, 42, 116, 52, 131, 52, 39, 28, 31, 74, 39, 131, 46, 33, 33, 96, 38)
)
```

```{r practice-2-scatter}
ggplot(practice_2, aes(patient_volume, wait_time)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#C83B3B") +
  labs(title = "Patient volume vs waiting time") +
  theme_minimal(base_size = 14)
```

```{r practice-2-fit}
practice_2_model <- lm(wait_time ~ patient_volume, data = practice_2)
summary(practice_2_model)
```

```{r practice-2-check}
check_model(practice_2_model)
```

**Your answer here:**

------------------------------------------------------------------------

------------------------------------------------------------------------

## I: Influential Points

The question here is: is any single observation pulling the line around?

For this, we read the **Influential observations** panel (residuals vs leverage).

Points that sit *outside the dashed contour* combine an unusual x-value (leverage) with a large residual (off the line).

### BMI and cholesterol without outliers (no influential points)

BMI vs. cholesterol from a routine screening, with no data-entry errors.

```{r inf-good-data}
inf_bad <- data.frame(
  bmi         = c(19, 20, 21, 22, 23, 23.5, 24, 24.5, 25, 25.5, 26, 26.5, 27, 27.5, 28, 28.5, 29, 29.5, 30, 30.5, 31, 20.5, 23, 25.5, 27, 29, 22, 24.5, 26.5, 28.5, 42),
  cholesterol = c(216, 204, 229, 211, 228, 214, 236, 226, 249, 225, 243, 225, 252, 238, 258, 240, 265, 243, 261, 245, 276, 220, 214, 241, 234, 265, 213, 242, 229, 256, 130)
)
inf_good <- inf_bad %>% filter(bmi != 42)
```

```{r inf-good-scatter}
ggplot(inf_good, aes(bmi, cholesterol)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#C83B3B") +
  labs(title = "BMI vs cholesterol: no influential observations") +
  theme_minimal(base_size = 14)
```

```{r inf-good-fit}
inf_good_model <- lm(cholesterol ~ bmi, data = inf_good)
summary(inf_good_model)
```

```{r inf-good-check}
check_model(inf_good_model)
```

Every point stays well inside the dashed contour in the *Influential observations* panel.

No single observation stands out as influential.

### BMI and cholesterol with data-entry error (influential point)

The same screening, but with a likely data-entry error: BMI = 42 paired with cholesterol = 130, which is unusually low for that BMI.

```{r inf-bad-scatter}
ggplot(inf_bad, aes(bmi, cholesterol)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#C83B3B") +
  labs(title = "BMI vs cholesterol: influential observation at BMI 42") +
  theme_minimal(base_size = 14)
```

To diagnose, we look at the *Influential observations* panel in the check_model() output.

```{r inf-bad-check}
inf_bad_model  <- lm(cholesterol ~ bmi, data = inf_bad)
check_model(inf_bad_model)
```

In the *Influential observations* panel, one point lies far to the right, outside the dashed contour.

It has both high leverage (BMI 42 is well above the rest of the data) and a large residual (far from the fitted line).

Removing it would noticeably change the fit.

### Practice

Blood lead level (µg/dL) vs. reaction time on a cognitive test (milliseconds).

Fit the model, look at the scatter plot, then read the **Influential observations** panel. Is any single observation pulling the line around?

```{r practice-3-data}
practice_3 <- data.frame(
  lead_level    = c(2, 4, 6, 8, 10, 12, 14, 16, 18, 20, 22, 24, 26, 28, 30, 3, 9, 15, 21, 27, 5, 11, 17, 23, 29, 7, 13, 19, 25, 45),
  reaction_time = c(208, 208, 219, 217, 227, 226, 238, 239, 250, 247, 257, 256, 268, 268, 279, 205, 225, 234, 256, 267, 218, 225, 245, 254, 276, 216, 237, 245, 265, 215)
)
```

```{r practice-3-scatter}
ggplot(practice_3, aes(lead_level, reaction_time)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#C83B3B") +
  labs(title = "Blood lead level vs reaction time") +
  theme_minimal(base_size = 14)
```

```{r practice-3-fit}
practice_3_model <- lm(reaction_time ~ lead_level, data = practice_3)
summary(practice_3_model)
```

```{r practice-3-check}
check_model(practice_3_model)
```

**Your answer here:**

------------------------------------------------------------------------

------------------------------------------------------------------------

## N: Normality of Residuals

**The question:** are the prediction errors roughly bell-shaped?

We read the **Normality of residuals** panel (the Q-Q plot).

Points close to the reference line are good; points diverging at the ends indicate skew or heavy tails.

### Height and lung capacity (normal residuals)

Height vs. lung capacity.

Errors scatter symmetrically around the line.

```{r norm-good-data}
norm_good <- data.frame(
  height_cm     = c(154.7, 155.6, 155.8, 156.3, 157.6, 158, 158.6, 159.4, 159.6, 160.6, 160.4, 161, 162, 162.4, 163.3, 164.2, 164.8, 164.7, 165.8, 166.4, 166.6, 167.1, 167.8, 169, 168.9, 169.7, 169.9, 171.4, 171.4, 171.7, 173, 173.5, 173.7, 174.7, 174.9, 175.4, 175.9, 176.8, 177.4, 177.8, 179.1, 179, 180.1, 180.5, 181.2, 182.2, 181.9, 182.9, 183.8, 184, 185.1, 185.8, 185.8, 185.9, 187.1, 188, 188.4, 188.3, 189, 190.1),
  lung_capacity = c(3.54, 3.88, 3.61, 3.92, 3.98, 3.75, 3.43, 3.89, 3.59, 3.67, 3.47, 3.7, 3.98, 3.41, 3.78, 4.08, 3.85, 3.91, 3.77, 3.93, 3.82, 3.72, 3.99, 4.2, 4.27, 4.1, 3.78, 4.13, 4.15, 4.4, 4.08, 4.16, 4.2, 4.21, 4.54, 4.67, 4.55, 4.54, 4.54, 4.34, 4.62, 4.53, 4.49, 4.57, 4.67, 4.41, 4.25, 4.92, 4.64, 4.8, 4.63, 4.8, 5.03, 4.88, 4.94, 4.92, 5.01, 4.93, 4.8, 5.17)
)
```

```{r norm-good-scatter}
ggplot(norm_good, aes(height_cm, lung_capacity)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#C83B3B") +
  labs(title = "Height vs lung capacity: symmetric scatter around the line") +
  theme_minimal(base_size = 14)
```

```{r norm-good-fit}
norm_good_model <- lm(lung_capacity ~ height_cm, data = norm_good)
summary(norm_good_model)
```

```{r norm-good-check}
check_model(norm_good_model)
```

The Q-Q points sit on (or very near) the diagonal across the whole range, and inside the shaded band.

That is what approximately normal residuals look like.

### Poverty rate and ER visits (non-normal residuals)

Neighbourhood poverty rate vs. ER visits per 1,000.

Most neighbourhoods follow the trend, but several have unusually high visit rates, producing strongly right-skewed residuals.

```{r norm-bad-data}
norm_bad <- data.frame(
  poverty_rate = c(5, 7, 8, 10, 12, 13, 15, 16, 18, 19, 20, 22, 23, 25, 27, 28, 30, 32, 34, 35, 37, 38, 40, 42, 6, 14, 21, 29, 36, 41),
  er_visits    = c(113, 105, 107, 112, 213, 122, 119, 124, 205, 154, 150, 159, 243, 155, 176, 178, 280, 191, 207, 199, 280, 195, 222, 220, 99, 129, 219, 181, 211, 296)
)
```

```{r norm-bad-scatter}
ggplot(norm_bad, aes(poverty_rate, er_visits)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#C83B3B") +
  labs(title = "Poverty rate vs ER visits: right-skewed residuals") +
  theme_minimal(base_size = 14)
```

```{r norm-bad-fit}
norm_bad_model <- lm(er_visits ~ poverty_rate, data = norm_bad)
summary(norm_bad_model)
```

```{r norm-bad-check}
check_model(norm_bad_model)
```

In the *Normality* Q-Q panel, the points bend away from the reference line at the upper end.

That is the long right tail of the skewed residuals.

### Practice

Gestational age at delivery (weeks) vs. birth weight (grams).

Fit the model, look at the scatter plot, then read the **Normality of residuals** panel. Are the residuals roughly bell-shaped?

```{r practice-4-data}
practice_4 <- data.frame(
  gestational_age = c(35, 35.5, 36, 36.5, 37, 37.5, 38, 38.5, 39, 39.5, 40, 40.5, 41, 41.5, 42, 35.2, 36.8, 37.6, 38.4, 39.2, 40.2, 41.2, 36.2, 37.2, 38.2, 39.6, 40.6, 37.8, 39.8, 41.6),
  birth_weight    = c(2730, 2506, 2639, 2903, 2674, 3008, 3016, 3212, 3240, 3326, 3469, 3667, 3713, 3836, 3789, 2633, 2775, 2915, 3121, 3525, 3463, 3791, 2765, 2975, 3157, 3531, 3676, 3009, 3538, 3816)
)
```

```{r practice-4-scatter}
ggplot(practice_4, aes(gestational_age, birth_weight)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#C83B3B") +
  labs(title = "Gestational age vs birth weight") +
  theme_minimal(base_size = 14)
```

```{r practice-4-fit}
practice_4_model <- lm(birth_weight ~ gestational_age, data = practice_4)
summary(practice_4_model)
```

```{r practice-4-check}
check_model(practice_4_model)
```

**Your answer here:**

------------------------------------------------------------------------

------------------------------------------------------------------------

# Categorical polytomous predictor

A polytomous predictor has three or more levels (for example, three treatment groups).

The fitted line connects the group means. Within each group, every observation has the same x-value (its group code).

**Shape (linearity)** is not a meaningful check here. The line passes through the group means, so there is no curvature to assess.

**Influential points** are not meaningful either. Every observation in a group shares the same x-value, so no point can have unusual leverage. A large outcome value is an outlier, not an influential point in the sense used above.

Only **H** (homogeneity of variance) and **N** (normality of residuals) apply.

With three or more groups, the homogeneity panel in `check_model()` behaves as expected.

The Linearity and Influential observations panels still appear in the default output, but you can hide them with the `check` argument. For example, `check = c("homogeneity", "normality")` shows only the two panels you need.

## H: Homoscedasticity (Equal Variance)

The question here is: is the spread of the outcome similar in each group?

For this, we read the **Homogeneity of variance** panel.

A flat green line means even spread across groups.

### Knee rehabilitation program (equal spread)

Knee range-of-motion gain (degrees) for patients on one of three rehabilitation programs.

```{r cat-multi-data}
cat_multi <- data.frame(
  program = factor(rep(c("standard", "intensive", "aquatic"), each = 18),
                   levels = c("standard", "intensive", "aquatic")),
  rom_gain = c(40.2, 28.6, 34.2, 35.8, 34.4, 31.4, 41.1, 31.4, 44.1, 31.6, 39.8, 45.7, 23.7, 30.3, 31.2, 35.8, 30.3, 16.1,
               30.4, 52.9, 43.2, 34.3, 44, 52.3, 56.4, 42.4, 43.5, 34.4, 47.8, 41.2, 47.7, 49.2, 51.2, 41.3, 48, 34.7,
               34.3, 33.9, 24.5, 39.2, 40.2, 36.8, 43.5, 34.6, 30.8, 41.6, 34.1, 47.7, 36.4, 42.9, 40.9, 34.3, 48.5, 42.9)
)
```

```{r cat-multi-scatter}
ggplot(cat_multi, aes(program, rom_gain)) +
  geom_boxplot(outlier.shape = NA, fill = "grey90") +
  geom_jitter(width = 0.12, alpha = 0.4, size = 2) +
  labs(title = "Range-of-motion gain by therapy program", y = "ROM gain (degrees)") +
  theme_minimal(base_size = 14)
```

```{r cat-multi-fit}
cat_multi_model <- lm(rom_gain ~ program, data = cat_multi)
summary(cat_multi_model)
```

```{r cat-multi-check}
check_model(cat_multi_model)
```

The default output includes all four SHIN panels. Linearity and Influential observations do not apply here.

To show only the homogeneity panel:

```{r cat-multi-check-h}
check_model(cat_multi_model, check = "homogeneity")
```

The green line is reasonably flat, indicating similar spread across groups.

### Length of stay by ward type (unequal spread)

Hospital length of stay (days) for patients on general, surgical, or ICU wards.

General and surgical stays cluster tightly. ICU stays vary much more widely.

```{r cat-poly-hom-bad-data}
cat_poly_hom_bad <- data.frame(
  ward = factor(rep(c("general", "surgical", "icu"), each = 18),
                levels = c("general", "surgical", "icu")),
  los_days = c(3.4, 6.7, 5.1, 4.3, 3.8, 6, 3.9, 3.8, 3.8, 4.3, 3.2, 4.1, 4.8, 3.9, 3.1, 3, 4.9, 6.2,
               6.2, 8.1, 7.3, 5.8, 9.2, 6.5, 7.8, 10.1, 6.9, 8.4, 7.1, 9.5, 6.3, 8.8, 7.6, 10.3, 6.8, 9.1,
               11.2, 0.5, 4.4, 7, 4, 2.3, 4.6, 9.5, 8.1, 2.9, 10.7, 0.5, 9.4, 1.3, 9.9, 4.6, 2.4, 3.4)
)
```

```{r cat-poly-hom-bad-scatter}
ggplot(cat_poly_hom_bad, aes(ward, los_days)) +
  geom_boxplot(outlier.shape = NA, fill = "grey90") +
  geom_jitter(width = 0.12, alpha = 0.4, size = 2) +
  labs(title = "Length of stay by ward type", y = "Length of stay (days)") +
  theme_minimal(base_size = 14)
```

```{r cat-poly-hom-bad-fit}
cat_poly_hom_bad_model <- lm(los_days ~ ward, data = cat_poly_hom_bad)
summary(cat_poly_hom_bad_model)
```

```{r cat-poly-hom-bad-check}
check_model(cat_poly_hom_bad_model)
```

To show only the homogeneity panel:

```{r cat-poly-hom-bad-check-h}
check_model(cat_poly_hom_bad_model, check = "homogeneity")
```

The homogeneity green line slopes upward, indicating that spread increases from general to ICU.

```{r cat-poly-hom-bad-sds}
cat_poly_hom_bad %>%
  group_by(ward) %>%
  summarise(mean = mean(los_days), sd = sd(los_days), n = n())
```

The ICU standard deviation is much larger than the others.

### Practice

Pain score reduction (points) for patients on one of three analgesic regimens.

Fit the model, look at the boxplot, then read the **Homogeneity of variance** panel. Use `check_model(model, check = "homogeneity")` if you want to hide the other panels. Is the spread similar across groups?

```{r practice-6-data}
practice_6 <- data.frame(
  regimen = factor(rep(c("standard", "extended", "combination"), each = 18),
                   levels = c("standard", "extended", "combination")),
  pain_reduction = c(3.2, 4.1, 3.8, 4.5, 3.5, 4.2, 3.9, 4.0, 3.6, 4.3, 3.7, 4.4, 3.4, 4.1, 3.8, 4.2, 3.9, 4.0,
                   5.1, 6.2, 5.8, 6.5, 5.4, 6.0, 5.6, 6.3, 5.2, 6.1, 5.7, 6.4, 5.5, 6.0, 5.9, 6.2, 5.3, 6.1,
                   4.8, 12.5, 5.2, 11.8, 4.5, 13.1, 5.0, 12.2, 4.9, 11.5, 5.3, 12.8, 4.7, 13.4, 5.1, 11.9, 4.6, 12.6)
)
```

```{r practice-6-scatter}
ggplot(practice_6, aes(regimen, pain_reduction)) +
  geom_boxplot(outlier.shape = NA, fill = "grey90") +
  geom_jitter(width = 0.12, alpha = 0.4, size = 2) +
  labs(title = "Pain reduction by analgesic regimen", y = "Pain reduction (points)") +
  theme_minimal(base_size = 14)
```

```{r practice-6-fit}
practice_6_model <- lm(pain_reduction ~ regimen, data = practice_6)
summary(practice_6_model)
```

```{r practice-6-check}
check_model(practice_6_model)
```

**Your answer here:**

------------------------------------------------------------------------

------------------------------------------------------------------------

## N: Normality of Residuals

**The question:** are the residuals roughly bell-shaped?

We read the **Normality of residuals** panel (the Q-Q plot), as with a continuous predictor.

### Knee rehabilitation program (normal residuals)

Using the three-group rehabilitation data from above.

```{r cat-multi-norm-check}
check_model(cat_multi_model)
```

To show only the normality panel:

```{r cat-multi-norm-check-n}
check_model(cat_multi_model, check = "normality")
```

The Q-Q points lie close to the reference line across the range.

### CRP level by clinic site (non-normal residuals)

C-reactive protein (mg/L) measured at three clinic sites.

Most values are moderate, but a few patients at each site have very high CRP, producing right-skewed residuals.

```{r cat-poly-norm-bad-data}
cat_poly_norm_bad <- data.frame(
  site = factor(rep(c("clinic_a", "clinic_b", "clinic_c"), each = 18),
                levels = c("clinic_a", "clinic_b", "clinic_c")),
  crp_mg = c(2.1, 3.4, 2.8, 4.2, 3.1, 2.5, 3.8, 2.9, 3.5, 4.0, 2.7, 3.3, 2.4, 3.6, 2.8, 3.2, 2.6, 18.5,
             2.3, 3.1, 2.9, 3.7, 2.6, 3.4, 3.0, 2.8, 3.5, 2.5, 3.2, 2.7, 3.9, 2.4, 3.1, 2.8, 3.6, 16.2,
             2.5, 3.0, 2.7, 3.3, 2.9, 3.8, 2.6, 3.4, 2.8, 3.1, 2.4, 3.5, 2.7, 3.2, 2.5, 3.6, 2.9, 19.8)
)
```

```{r cat-poly-norm-bad-scatter}
ggplot(cat_poly_norm_bad, aes(site, crp_mg)) +
  geom_boxplot(outlier.shape = NA, fill = "grey90") +
  geom_jitter(width = 0.12, alpha = 0.4, size = 2) +
  labs(title = "CRP level by clinic site", y = "CRP (mg/L)") +
  theme_minimal(base_size = 14)
```

```{r cat-poly-norm-bad-fit}
cat_poly_norm_bad_model <- lm(crp_mg ~ site, data = cat_poly_norm_bad)
summary(cat_poly_norm_bad_model)
```

```{r cat-poly-norm-bad-check}
check_model(cat_poly_norm_bad_model)
```

To show only the normality panel:

```{r cat-poly-norm-bad-check-n}
check_model(cat_poly_norm_bad_model, check = "normality")
```

In the *Normality* Q-Q panel, the points bend away from the reference line at the upper end.

That is the long right tail from the high CRP values.

### Practice

Serum albumin (g/dL) for patients in one of three nutritional support programs.

Fit the model, look at the boxplot, then read the **Normality of residuals** panel. Use `check_model(model, check = "normality")` if you want to hide the other panels. Are the residuals roughly bell-shaped?

```{r practice-7-data}
practice_7 <- data.frame(
  program = factor(rep(c("standard", "high_protein", "enteral"), each = 18),
                   levels = c("standard", "high_protein", "enteral")),
  albumin = c(3.2, 3.5, 3.4, 3.6, 3.3, 3.5, 3.4, 3.7, 3.3, 3.6, 3.5, 3.4, 3.3, 3.6, 3.5, 3.4, 3.7, 3.3,
              3.4, 3.6, 3.5, 3.7, 3.4, 3.6, 3.5, 3.8, 3.4, 3.6, 3.5, 3.7, 3.4, 3.6, 3.5, 3.7, 3.4, 3.6,
              3.3, 3.5, 3.4, 3.6, 3.5, 3.4, 3.7, 3.3, 3.5, 3.6, 3.4, 3.5, 3.3, 3.6, 3.4, 3.5, 3.7, 3.3)
)
```

```{r practice-7-scatter}
ggplot(practice_7, aes(program, albumin)) +
  geom_boxplot(outlier.shape = NA, fill = "grey90") +
  geom_jitter(width = 0.12, alpha = 0.4, size = 2) +
  labs(title = "Serum albumin by nutritional support program", y = "Albumin (g/dL)") +
  theme_minimal(base_size = 14)
```

```{r practice-7-fit}
practice_7_model <- lm(albumin ~ program, data = practice_7)
summary(practice_7_model)
```

```{r practice-7-check}
check_model(practice_7_model)
```

**Your answer here:**

------------------------------------------------------------------------

------------------------------------------------------------------------

# Categorical binary predictor

A binary predictor has exactly two levels (for example, exposed vs. unexposed).

The same reasoning as above applies: **Shape** and **Influential points** are not meaningful checks.

Only **H** and **N** apply.

The default `check_model()` output still includes all four SHIN panels. Use `check = c("homogeneity", "normality")` to show only the two that apply.

## H: Homoscedasticity (Equal Variance)

The question here is: is the spread of the outcome similar in each group?

Read the **Homogeneity of variance** panel in `check_model()` first. For a binary predictor, the green line in that panel is not reliable; use the base R scale-location plot (`plot(model, which = 3)`) to compare spread across groups.

### Birth weight by smoking status (equal spread)

Birth weight (grams) for babies of non-smokers vs. smokers.

```{r cat-binary-data}
cat_binary <- data.frame(
  smoke = factor(rep(c("nonsmoker", "smoker"), c(19, 17)),
                 levels = c("nonsmoker", "smoker")),
  birth_weight = c(3059, 3040, 3296, 3245, 3516, 2899, 3803, 3343, 3256, 3212, 3835, 3172, 3377, 3249, 4051, 3031, 3321, 2355, 3687,
                   2681, 3618, 3019, 2849, 3302, 2687, 3969, 3035, 2681, 3109, 3090, 2603, 2830, 3349, 3014, 2962, 2566)
)
```

```{r cat-binary-scatter}
ggplot(cat_binary, aes(smoke, birth_weight)) +
  geom_boxplot(outlier.shape = NA, fill = "grey90") +
  geom_jitter(width = 0.12, alpha = 0.4, size = 2) +
  labs(title = "Birth weight by smoking status", y = "Birth weight (g)") +
  theme_minimal(base_size = 14)
```

```{r cat-binary-fit}
cat_binary_model <- lm(birth_weight ~ smoke, data = cat_binary)
summary(cat_binary_model)
```

```{r cat-binary-check}
check_model(cat_binary_model)
```

To show only the panels that apply:

```{r cat-binary-check-hn}
check_model(cat_binary_model, check = c("homogeneity", "normality"))
```

Look at the **Homogeneity of variance** panel:

```{r cat-binary-check-h}
check_model(cat_binary_model, check = "homogeneity")
```

There are only two fitted values (one per group), but `check_model()` still draws a smooth curve through them. Interpolating between two points is unstable, so the green line can shoot up to extreme values. That does not mean the model is wrong; the line just does not make sense here. The dots are still the residuals you would compare across groups.

For the actual check, use base R instead:

```{r cat-binary-baseR}
plot(cat_binary_model, which = 3)
```

The plot shows two vertical strips of points, one per group.

The two strips are similar in height.

```{r cat-binary-sds-good}
cat_binary %>%
  group_by(smoke) %>%
  summarise(mean = mean(birth_weight), sd = sd(birth_weight), n = n())
```

The standard deviations are similar, so the equal-variance assumption is satisfied.

### ICU length of stay by admission type (unequal spread)

ICU length of stay (days) for elective vs. emergency admissions.

Elective admissions have shorter, more predictable stays. Emergency admissions vary more widely.

```{r cat-hom-bad-data}
cat_hom_bad <- data.frame(
  admission = factor(rep(c("elective", "emergency"), c(18, 18)),
                     levels = c("elective", "emergency")),
  los_days = c(3.4, 6.7, 5.1, 4.3, 3.8, 6, 3.9, 3.8, 3.8, 4.3, 3.2, 4.1, 4.8, 3.9, 3.1, 3, 4.9, 6.2,
               11.2, 0.5, 4.4, 7, 4, 2.3, 4.6, 9.5, 8.1, 2.9, 10.7, 0.5, 9.4, 1.3, 9.9, 4.6, 2.4, 3.4)
)
```

```{r cat-hom-bad-scatter}
ggplot(cat_hom_bad, aes(admission, los_days)) +
  geom_boxplot(outlier.shape = NA, fill = "grey90") +
  geom_jitter(width = 0.12, alpha = 0.4, size = 2) +
  labs(title = "ICU length of stay by admission type", y = "Length of stay (days)") +
  theme_minimal(base_size = 14)
```

```{r cat-hom-bad-fit}
cat_hom_bad_model <- lm(los_days ~ admission, data = cat_hom_bad)
summary(cat_hom_bad_model)
```

Use the base R scale-location plot to compare spread across groups:

```{r cat-hom-bad-baseR}
plot(cat_hom_bad_model, which = 3)
```

```{r cat-hom-bad-sds}
cat_hom_bad %>%
  group_by(admission) %>%
  summarise(mean = mean(los_days), sd = sd(los_days), n = n())
```

The emergency strip is taller than the elective strip, and the standard deviations confirm unequal spread.

The homogeneity assumption is violated.

### Practice

A clinical trial of a blood-pressure drug. Systolic blood pressure (mmHg) for patients on placebo vs. the active drug.

Fit the model, look at the boxplot, then read the **Homogeneity of variance** panel in `check_model()`. Use `check_model(model, check = "homogeneity")` to hide the other panels, and `plot(model, which = 3)` to compare spread across groups. Is the spread similar in each group?

```{r practice-5-data}
practice_5 <- data.frame(
  treatment = factor(rep(c("placebo", "drug"), c(18, 18)),
                     levels = c("placebo", "drug")),
  systolic_bp = c(141, 142, 151, 141, 123, 146, 135, 149, 149, 132, 144, 142, 159, 144, 148, 142, 141, 145,
                  132, 126, 127, 134, 138, 135, 133, 133, 136, 149, 114, 123, 145, 132, 129, 138, 129, 144)
)
```

```{r practice-5-scatter}
ggplot(practice_5, aes(treatment, systolic_bp)) +
  geom_boxplot(outlier.shape = NA, fill = "grey90") +
  geom_jitter(width = 0.12, alpha = 0.4, size = 2) +
  labs(title = "Systolic blood pressure by treatment", y = "Systolic BP (mmHg)") +
  theme_minimal(base_size = 14)
```

```{r practice-5-fit}
practice_5_model <- lm(systolic_bp ~ treatment, data = practice_5)
summary(practice_5_model)
```

```{r practice-5-check}
check_model(practice_5_model)
```

```{r practice-5-baseR}
plot(practice_5_model, which = 3)
```

**Your answer here:**

------------------------------------------------------------------------

------------------------------------------------------------------------

## N: Normality of Residuals

**The question:** are the residuals roughly bell-shaped?

We read the **Normality of residuals** panel in `check_model()`.

### Birth weight by smoking status (normal residuals)

Using the birth weight data from above.

```{r cat-binary-check-n}
check_model(cat_binary_model, check = "normality")
```

The Q-Q points lie close to the reference line.

### Hospital charges by insurance type (non-normal residuals)

Total hospital charges (USD) for privately insured vs. publicly insured patients.

A few very high charges among privately insured patients produce right-skewed residuals.

```{r cat-bin-norm-bad-data}
cat_bin_norm_bad <- data.frame(
  insurance = factor(rep(c("public", "private"), c(20, 20)),
                     levels = c("public", "private")),
  charges = c(4200, 5100, 4800, 5300, 4600, 4900, 5200, 4500, 5000, 4700, 5100, 4400, 4900, 4600, 5200, 4800, 5000, 4700, 5100, 4900,
              8200, 9100, 7800, 9500, 8600, 8800, 9200, 8400, 8900, 8700, 9000, 8300, 8800, 8500, 9100, 8600, 8900, 8400, 8700, 28500)
)
```

```{r cat-bin-norm-bad-scatter}
ggplot(cat_bin_norm_bad, aes(insurance, charges)) +
  geom_boxplot(outlier.shape = NA, fill = "grey90") +
  geom_jitter(width = 0.12, alpha = 0.4, size = 2) +
  labs(title = "Hospital charges by insurance type", y = "Charges (USD)") +
  theme_minimal(base_size = 14)
```

```{r cat-bin-norm-bad-fit}
cat_bin_norm_bad_model <- lm(charges ~ insurance, data = cat_bin_norm_bad)
summary(cat_bin_norm_bad_model)
```

```{r cat-bin-norm-bad-check}
check_model(cat_bin_norm_bad_model)
```

To show only the normality panel:

```{r cat-bin-norm-bad-check-n}
check_model(cat_bin_norm_bad_model, check = "normality")
```

In the *Normality* Q-Q panel, the points bend away from the reference line at the upper end.

That is the long right tail from the high charge values.

### Practice

Peak expiratory flow (L/min) for patients with vs. without a chronic respiratory condition.

Fit the model, look at the boxplot, then read the **Normality of residuals** panel in `check_model()`. Use `check_model(model, check = "normality")` if you want to hide the other panels. Are the residuals roughly bell-shaped?

```{r practice-8-data}
practice_8 <- data.frame(
  condition = factor(rep(c("none", "chronic"), c(18, 18)),
                     levels = c("none", "chronic")),
  peak_flow = c(420, 435, 410, 445, 430, 425, 440, 415, 450, 428, 438, 422, 442, 418, 448, 432, 426, 444,
                310, 325, 305, 330, 315, 320, 328, 308, 332, 312, 318, 322, 326, 310, 324, 316, 320, 314)
)
```

```{r practice-8-scatter}
ggplot(practice_8, aes(condition, peak_flow)) +
  geom_boxplot(outlier.shape = NA, fill = "grey90") +
  geom_jitter(width = 0.12, alpha = 0.4, size = 2) +
  labs(title = "Peak expiratory flow by respiratory condition", y = "Peak flow (L/min)") +
  theme_minimal(base_size = 14)
```

```{r practice-8-fit}
practice_8_model <- lm(peak_flow ~ condition, data = practice_8)
summary(practice_8_model)
```

```{r practice-8-check}
check_model(practice_8_model)
```

**Your answer here:**

------------------------------------------------------------------------

------------------------------------------------------------------------
