---
title: "Depressive symptoms in Brazilian medical students"
subtitle: "Dataset 8: Resilience, learning environment and quality of life in Brazilian medical students"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "medstudent_qol.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "bdi_depression"
predictors <- c("sex", "age_yrs", "year_group", "school_legal_status",
                "resilience_score", "dreem_global")

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex = factor(sex, levels = c("female", "male"),
                 labels = c("Female", "Male")),
    year_group = factor(year_group,
                        levels = c("basic_sciences", "clinical_sciences", "clerkship"),
                        labels = c("Basic sciences", "Clinical sciences", "Clerkship")),
    school_legal_status = factor(school_legal_status, levels = c("public", "private"),
                                 labels = c("Public", "Private")),
    school_location = factor(school_location, levels = c("state_capital", "other_city"),
                             labels = c("State capital", "Other city"))
  ) |>
  labelled::set_variable_labels(
    bdi_depression      = "BDI total score (0-63)",
    sex                 = "Sex",
    age_yrs             = "Age (years)",
    year_group          = "Stage of course",
    school_legal_status = "School legal status",
    resilience_score    = "Resilience score (16-98)",
    dreem_global        = "DREEM global score (0-200)"
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)
```

# Introduction

Tempski and colleagues surveyed a random sample of 1,350 students at 22 Brazilian
medical schools in 2011-2012, measuring quality of life (WHOQOL-BREF), the
perceived educational environment (DREEM), resilience, depressive symptoms and
anxiety in a single cross-sectional questionnaire (Tempski et al., *PLOS ONE*
2015; doi:10.1371/journal.pone.0131535). Depression in medical students is
consistently more common than in age-matched peers and is the outcome with the
clearest link to attrition and to later physician burnout, so identifying which
student and school characteristics travel with it is directly useful to
curriculum committees and student support services.

**Aim.** Among Brazilian medical students, we examined which of sex, age, stage
of the course, school legal status, resilience and the perceived educational
environment are independently associated with the Beck Depression Inventory
score, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** Multicentre cross-sectional survey of a random sample of 1,350
students drawn from 22 medical schools across Brazil, stratified by stage of the
course (basic sciences, clinical sciences, clerkship). All variables used here
are complete.

**Variables.** The outcome is `bdi_depression`, the Beck Depression Inventory
total score, an integer from 0 to 63 (observed range 0 to 42) where higher means
more depressive symptoms; a score of 14 or more conventionally indicates moderate
or worse depression. Predictors are sex (reference: female, the larger group),
age in years, stage of the course (reference: basic sciences, the earliest
stage), school legal status (reference: public, the larger group and the
traditional Brazilian medical school), the resilience scale score (16-98, higher
= more resilient) and the DREEM global score (sum of the five DREEM subscales,
higher = a more positive perceived educational environment). Continuous
predictors are modelled per one-unit increase.

Deliberate exclusions. `dreem_global` is used **instead of** its five subscales,
which sum to it exactly. Neither `state_anxiety` nor `trait_anxiety` is used as
a predictor, even though both are strongly associated with the BDI (*r* = 0.59
and 0.73): all three are self-report scales of the same negative-affect
construct, measured on the same day, and a model of depression on trait anxiety
would be an instrument-validation exercise rather than a study of students. The
four WHOQOL domains are excluded for the same reason - `whoqol_psychological`
correlates -0.74 with the BDI. What is left is a model of depressive symptoms on
demography, stage of training and the educational environment, which is the
question a medical school would actually ask.

**Statistical analysis.** Linear regression (`lm()`), justified because the
outcome is a continuous summed score with 38 distinct observed values.
We fitted a univariable model for each predictor first, then a single
multivariable model containing all of them. Predictors were chosen in advance on
subject-matter grounds, not by significance. Significance threshold 0.05;
estimates are reported with 95% confidence intervals. Assumptions were checked
with `performance::check_model()`. Analysis used `r R.version.string` with
tidyverse, gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped).

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(
    missing = "no"
  ) |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

```{r prev}
n_mod <- sum(dat_model$bdi_depression >= 14)
```

The median BDI score was 8 (IQR 4 to 13), and
`r n_mod` students (`r sprintf("%.1f%%", 100 * n_mod / nrow(dat_model))`) scored
14 or more, the conventional threshold for moderate or worse depression. The
distribution is strongly right-skewed with a floor at 0, which matters for the
diagnostics below.

## Exploratory figure

```{r explore, fig.cap="Figure 1. BDI score against resilience score, by sex. Points are individual students; lines are linear smoothers. The dashed line marks the moderate-depression threshold of 14."}
ggplot(dat_model, aes(x = resilience_score, y = bdi_depression, colour = sex)) +
  geom_point(alpha = 0.3, size = 1.2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.9) +
  geom_hline(yintercept = 14, linetype = "dashed", colour = "grey40") +
  scale_colour_manual(values = c("#B3589A", "#3B7EA1")) +
  labs(x = "Resilience score (16-98)", y = "BDI total score (0-63)",
       colour = "Sex") +
  theme(legend.position = "top")
```

Depressive symptoms fall steeply as resilience rises, the two sexes have nearly
parallel lines with women sitting a couple of points higher throughout, and the
funnel shape - wide scatter at low resilience, tight against the floor at high
resilience - is the heteroscedasticity that the diagnostics will pick up.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    hide_n = TRUE
  ) |>
  modify_caption("**Univariable linear regression.**")
uv
```

```{r uv-numbers}
u <- function(v, tm = v) lm(reformulate(v, outcome), data = dat_model) |>
  tidy(conf.int = TRUE) |> filter(term == tm)
uv_res   <- u("resilience_score")
uv_dreem <- u("dreem_global")
uv_sex   <- u("sex", "sexMale")
uv_priv  <- u("school_legal_status", "school_legal_statusPrivate")
uv_clerk <- u("year_group", "year_groupClerkship")
uv_clin  <- u("year_group", "year_groupClinical sciences")
uv_age   <- u("age_yrs")
```

Resilience is the strongest crude predictor: each additional resilience point is
associated with a `r sprintf("%.3f", abs(uv_res$estimate))`-point *lower* BDI
score (95% CI
`r sprintf("%.3f to %.3f", uv_res$conf.low, uv_res$conf.high)`, p < 0.001), so
across the observed 16-to-98 resilience range that is about
`r sprintf("%.0f", abs(uv_res$estimate) * 82)` BDI points - more than the whole
observed spread of the outcome, which already signals that resilience and
depression are close to being the same measurement. The DREEM global score is
also inversely associated (`r sprintf("%.3f", uv_dreem$estimate)` per point,
95% CI `r sprintf("%.3f to %.3f", uv_dreem$conf.low, uv_dreem$conf.high)`). Men
score `r sprintf("%.2f", uv_sex$estimate)` points lower than women (95% CI
`r sprintf("%.2f to %.2f", uv_sex$conf.low, uv_sex$conf.high)`, p < 0.001).
Neither age (p = `r sprintf("%.2f", uv_age$p.value)`), stage of the course
(clerkship p = `r sprintf("%.2f", uv_clerk$p.value)`) nor school legal status
(p = `r sprintf("%.2f", uv_priv$p.value)`) shows any crude association.

## Why adjustment is needed

Two variables in this file plausibly confound the crude estimates. **The DREEM
global score** confounds the school legal status contrast: private-school
students rate their educational environment much more favourably (mean DREEM 128
versus 115 at public schools) and a better-rated environment goes with fewer
depressive symptoms, so the null crude private-school coefficient is a mixture of
two opposing forces and cannot be read at face value. **Stage of the course**
confounds the crude age association, because the stages are ordered in time and
clerkship students are on average four years older than basic-sciences students;
an unadjusted age coefficient is really asking "do later-stage students differ",
not "do older students differ". **Sex** is a third, though a weak one: women in this
sample report slightly lower resilience scores (mean 78.1 versus 79.2) as well as
clearly higher BDI scores, so a small part of the crude resilience association is
a sex difference. Finally, **stage of the course** also
confounds the DREEM association in the opposite direction - clerkship students
rate their environment worst of the three stages (mean DREEM 114 versus 124) -
so holding stage fixed is necessary before the DREEM coefficient can be
interpreted. A single model with all six predictors makes each estimate
conditional on the rest.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit) |>
  modify_caption("**Multivariable linear regression.**")
mv
```

```{r fit-stats}
gl <- glance(fit)
tibble(
  Statistic = c("R-squared", "Adjusted R-squared", "Residual SE", "n", "Model df"),
  Value = c(sprintf("%.3f", gl$r.squared), sprintf("%.3f", gl$adj.r.squared),
            sprintf("%.2f", gl$sigma), as.character(gl$nobs), as.character(gl$df))
) |> gt() |> tab_header(title = md("**Model fit**"))
```

```{r mv-numbers}
tid <- tidy(fit, conf.int = TRUE)
g <- function(x) tid |> filter(term == x)
b_res <- g("resilience_score"); b_dreem <- g("dreem_global"); b_sex <- g("sexMale")
b_priv <- g("school_legal_statusPrivate"); b_clerk <- g("year_groupClerkship")
b_clin <- g("year_groupClinical sciences"); b_age <- g("age_yrs")
fmt <- function(r) sprintf("%.2f (95%% CI %.2f to %.2f)", r$estimate, r$conf.low, r$conf.high)
fmt3 <- function(r) sprintf("%.3f (95%% CI %.3f to %.3f)", r$estimate, r$conf.low, r$conf.high)
```

The model explains `r sprintf("%.1f%%", gl$r.squared * 100)` of the variance in
BDI score (adjusted R-squared `r sprintf("%.3f", gl$adj.r.squared)`). Holding the
other five predictors fixed, each resilience point is associated with
`r fmt3(b_res)` BDI points and each DREEM point with `r fmt3(b_dreem)` BDI
points, so a 20-point better perceived environment corresponds to about
`r sprintf("%.1f", abs(b_dreem$estimate) * 20)` fewer BDI points. Men score
`r fmt(b_sex)` points lower than women. Two coefficients that were null crudely
become significant: clerkship students score `r fmt(b_clerk)` points lower than
basic-sciences students (clinical sciences `r fmt(b_clin)`), and private-school
students score `r fmt(b_priv)` points *higher* than public-school students. Age
has no independent association (`r fmt(b_age)`). All estimates are conditional
associations from one cross-sectional survey and carry no causal claim.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

The two headline predictors move only modestly toward the null. **Resilience**
goes from `r sprintf("%.3f", uv_res$estimate)` to
`r sprintf("%.3f", b_res$estimate)`, retaining
`r sprintf("%.0f%%", 100 * b_res$estimate / uv_res$estimate)` of its crude size -
essentially nothing in this model explains it away. **Sex** retains
`r sprintf("%.0f%%", 100 * b_sex$estimate / uv_sex$estimate)`
(`r sprintf("%.2f", uv_sex$estimate)` to `r sprintf("%.2f", b_sex$estimate)`),
so the female excess in depressive symptoms is largely not accounted for by
anything else in the model. **The DREEM global score** is the one that
halves, from `r sprintf("%.3f", uv_dreem$estimate)` to
`r sprintf("%.3f", b_dreem$estimate)`
(`r sprintf("%.0f%%", 100 * b_dreem$estimate / uv_dreem$estimate)` retained):
much of the crude "bad environment goes with depression" association is really
"low-resilience students both rate their school badly and score high on the BDI".

Two coefficients change significance, both through suppression. **Private
schooling** flips sign, from `r sprintf("%.2f", uv_priv$estimate)` crude and
clearly non-significant (p = `r sprintf("%.2f", uv_priv$p.value)`) to
`r sprintf("%.2f", b_priv$estimate)` adjusted and significant
(p = `r sprintf("%.3f", b_priv$p.value)`). Private-school students rate their
educational environment 13 DREEM points better on average, which pushes their
crude BDI scores down; once DREEM is held fixed, a genuine private-school
*excess* of depressive symptoms appears. **Clerkship** does the same:
`r sprintf("%.2f", uv_clerk$estimate)` crude (p =
`r sprintf("%.2f", uv_clerk$p.value)`) becomes
`r sprintf("%.2f", b_clerk$estimate)` (p = `r sprintf("%.3f", b_clerk$p.value)`),
because clerkship students rate their environment worst of the three stages and
that was masking their lower depressive-symptom scores. **Age** barely moves in
absolute terms (`r sprintf("%.3f", uv_age$estimate)` to
`r sprintf("%.3f", b_age$estimate)`) and is non-significant throughout.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r vif-table}
performance::check_collinearity(fit) |>
  as_tibble() |>
  transmute(Term, VIF = round(VIF, 2), Tolerance = round(Tolerance, 3)) |>
  gt() |> tab_header(title = md("**Variance inflation factors**"))
```

```{r het-test}
het <- performance::check_heteroscedasticity(fit)
```

**Linearity.** The fitted-versus-residual reference line bends upward at the low
end: for students the model predicts to have very low BDI scores, the residuals
are systematically positive, because the BDI cannot go below 0 and the model can.
This is the main structural failure of a linear model for this outcome.

**Homoscedasticity.** Breusch-Pagan strongly rejects constant variance
(p `r ifelse(attr(het, "p") < 0.001, "< 0.001", sprintf("= %.3f", attr(het, "p")))`)
and the scale-location panel rises steeply from left to right - residual spread
is small among students predicted to score near 0 and large among those predicted
to score high, exactly the fan seen in Figure 1. This is the expected pattern for
a bounded, right-skewed count-like score and it is more pronounced here than for
any of the WHOQOL outcomes. As a sensitivity check we refitted with
heteroscedasticity-consistent (HC3) standard errors:

```{r robust}
rob <- lmtest::coeftest(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
broom::tidy(rob) |>
  transmute(term,
            estimate = round(estimate, 3),
            `robust SE` = round(std.error, 3),
            statistic = round(statistic, 2),
            p.value = format.pval(p.value, digits = 3, eps = 0.001)) |>
  gt() |> tab_header(title = md("**HC3 robust standard errors (sensitivity check)**"))
```

The robust standard errors are close to the model-based ones and every
coefficient keeps its significance status, so the inference survives despite the
clear heteroscedasticity.

**Normality of residuals.** This is the weakest assumption here. The Q-Q plot
shows a pronounced right tail: the BDI distribution is skewed
(skewness `r sprintf("%.2f", (function(x) mean((x - mean(x))^3) / sd(x)^3)(dat_model$bdi_depression))`)
and the residuals inherit it, with the formal test rejecting normality
(p `r ifelse(attr(performance::check_normality(fit), "p") < 0.001, "< 0.001", sprintf("= %.3f", attr(performance::check_normality(fit), "p")))`).
With n = `r nrow(dat_model)` the coefficient standard errors are still protected
by the central limit theorem, but any prediction interval from this model is
untrustworthy and would include impossible negative BDI scores.

**Influential points.** No observation exceeds the Cook's distance threshold; the
largest is `r sprintf("%.3f", max(cooks.distance(fit)))`, so despite the skew no
single student drives the fit.

**Collinearity.** The highest VIF is
`r sprintf("%.2f", max(performance::check_collinearity(fit)$VIF))` (age and stage
of the course, which are strongly related by design), and every value is below 2.
This is the direct result of excluding `state_anxiety` and `trait_anxiety`: had
trait anxiety been added, it would have correlated 0.73 with the outcome and 0.61
with resilience, and the resilience coefficient would have been halved and its
interpretation destroyed.

# Discussion

Depressive symptoms among Brazilian medical students are most strongly associated
with low resilience, with women scoring about `r sprintf("%.1f", abs(b_sex$estimate))`
BDI points higher than men, and with a poorly rated educational environment
(about `r sprintf("%.1f", abs(b_dreem$estimate) * 20)` BDI points per 20 DREEM
points). Two findings emerge only after adjustment and are worth attention:
clerkship students score lower than basic-sciences students once the educational
environment is held fixed, and private-school students score higher than
public-school students on the same basis. Practically, the DREEM association is
the one a school can act on, but its size is modest - moving a school from the
25th to the 75th percentile of DREEM would be expected to shift the average BDI
score by only a couple of points, well short of the 14-point clinical threshold.

**Limitations.** *Distributional mismatch.* The BDI is a bounded, right-skewed,
integer score with a hard floor at 0 and 23% of students above the clinical
threshold; a linear model gives interpretable coefficients but fits the shape of
the outcome badly, produces impossible negative predictions at the low end, and
shows marked heteroscedasticity. Dichotomising at 14 and fitting logistic
regression, or modelling on the log scale, would respect the outcome's shape
better - at the cost of throwing away information or of interpretability.
*Clustering is not modelled.* Students came from 22 medical schools, but the
released file carries no school identifier, only `school_legal_status` and
`school_location`; students in one school share teachers, curriculum and
assessment load, so the reported standard errors - especially for the
private-versus-public contrast, which is a school-level variable - are too
narrow, and no random intercept can be fitted. *Cross-sectional design.*
Resilience and depressive symptoms were measured at the same sitting, and a
depressed student will answer a resilience questionnaire more negatively; the
resilience association is therefore not evidence that resilience training would
reduce depression. *No clinical assessment.* The BDI is a screening
questionnaire, not a diagnosis, and self-report of mood on a survey administered
by the medical school may be understated. *Unmeasured confounders* include
family income, personal or family psychiatric history, and academic standing,
none of which is in the file.

**Future work.** A cohort design measuring BDI at entry and at the end of each
stage, with school identifiers retained, would separate selection into schools
from the effect of the training environment.

# Feasibility notes

*Not part of a student report.*

- **Knits cleanly; no numerical trouble.** n = 1,350 complete cases, six
  predictors, max VIF
  `r sprintf("%.2f", max(performance::check_collinearity(fit)$VIF))`, max Cook's
  distance `r sprintf("%.3f", max(cooks.distance(fit)))`.
- **This outcome is a collinearity trap.** `trait_anxiety` correlates 0.73 with
  the BDI, `state_anxiety` 0.59, and `whoqol_psychological` -0.74. A student who
  drops all of those into the model gets R-squared above 0.6 and coefficients
  that are meaningless - they will have regressed one mood questionnaire on
  another. Excluding them is not optional; it is the whole design decision. If
  supervisors let students include `trait_anxiety` here, this outcome should be
  marked unusable.
- **Worst diagnostics of the six outcomes.** Skewness
  `r sprintf("%.2f", (function(x) mean((x - mean(x))^3) / sd(x)^3)(dat_model$bdi_depression))`,
  a hard floor at 0, strongly rejected normality and the steepest scale-location
  slope in the dataset. HC3 rescues the inference, but a student who takes
  `check_model()` seriously will have a lot to write about, and honest
  prediction from this model is not possible.
- **Best suppression examples in the dataset.** Both `school_legal_status` and
  `year_group` flip from null to significant once `dreem_global` is held fixed,
  driven by the 13-point DREEM gap between private and public schools. This is a
  genuinely instructive unadjusted-versus-adjusted table.
- **The catalogue also offers this outcome as binary** (BDI >= 14, 23.3% of
  students, `r n_mod`/`r nrow(dat_model)` here). For a student who is
  uncomfortable defending a linear model on a skewed bounded score, the logistic
  version is the safer route and loses little.
- **Guardrails.** `dreem_global` must never appear with its five subscales; no
  WHOQOL domain and no anxiety scale as a predictor.
- **Verdict: workable but demanding.** Good effect sizes and a rich adjustment
  story, undermined by an outcome distribution that a linear model does not fit
  well and by an unusually tempting set of forbidden predictors.
