---
title: "Self-rated medical-school quality of life in Brazilian medical students"
subtitle: "Dataset 8: Resilience, learning environment and quality of life in Brazilian medical students"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "medstudent_qol.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "medical_school_qol"
predictors <- c("sex", "year_group", "dreem_learning", "dreem_teachers",
                "dreem_academic_self", "dreem_atmosphere", "dreem_social_self")

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex = factor(sex, levels = c("female", "male"),
                 labels = c("Female", "Male")),
    year_group = factor(year_group,
                        levels = c("basic_sciences", "clinical_sciences", "clerkship"),
                        labels = c("Basic sciences", "Clinical sciences", "Clerkship")),
    school_legal_status = factor(school_legal_status, levels = c("public", "private"),
                                 labels = c("Public", "Private")),
    school_location = factor(school_location, levels = c("state_capital", "other_city"),
                             labels = c("State capital", "Other city"))
  ) |>
  labelled::set_variable_labels(
    medical_school_qol  = "Medical-school quality of life (0-10)",
    sex                 = "Sex",
    year_group          = "Stage of course",
    dreem_learning      = "DREEM: perception of learning (0-48)",
    dreem_teachers      = "DREEM: perception of teachers (0-44)",
    dreem_academic_self = "DREEM: academic self-perception (0-32)",
    dreem_atmosphere    = "DREEM: perception of atmosphere (0-48)",
    dreem_social_self   = "DREEM: social self-perception (0-28)"
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)
```

# Introduction

Tempski and colleagues surveyed a random sample of 1,350 students at 22 Brazilian
medical schools in 2011-2012, measuring quality of life (WHOQOL-BREF), the
perceived educational environment (DREEM), resilience, depressive symptoms and
anxiety in a single cross-sectional questionnaire (Tempski et al., *PLOS ONE*
2015; doi:10.1371/journal.pone.0131535). Alongside the standard instruments,
students rated their *medical-school* quality of life on a single 0-10 item.
That single item is the outcome a dean can most directly act on, and the DREEM
instrument - which breaks the educational environment into five distinct
components - is designed precisely to say *which part* of the school experience
needs attention.

**Aim.** Among Brazilian medical students, we examined which of the five DREEM
domains - perception of learning, perception of teachers, academic
self-perception, perception of atmosphere and social self-perception - are
independently associated with self-rated medical-school quality of life,
adjusting also for sex and stage of the course.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** Multicentre cross-sectional survey of a random sample of 1,350
students drawn from 22 medical schools across Brazil, stratified by stage of the
course (basic sciences, clinical sciences, clerkship). All variables used here
are complete.

**Variables.** The outcome is `medical_school_qol`, the student's self-rating of
medical-school quality of life on a 0-10 scale (observed range 0 to 10, higher is
better). Predictors are sex (reference: female, the larger group), stage of the
course (reference: basic sciences, the earliest stage) and the five DREEM
subscales, each modelled per one-point increase: perception of learning (0-48),
perception of teachers (0-44), academic self-perception (0-32), perception of
atmosphere (0-48) and social self-perception (0-28).

The single most important design decision is that `dreem_global` is **excluded**.
It is the exact arithmetic sum of the five subscales, so including it alongside
them would make the design matrix singular. Using the subscales rather than the
global score is the whole point of this analysis - the question is which
*component* of the environment carries the association - but it comes at a price:
the five subscales correlate with one another at *r* = 0.45 to 0.77, so their
variance inflation factors are elevated and are reported and discussed below.
No WHOQOL domain, BDI or anxiety scale is used as a predictor: all of them are
alternative self-reported wellbeing measures collected on the same form, and
adding them would turn the analysis into one questionnaire predicting another.

**Statistical analysis.** Linear regression (`lm()`), justified because the
outcome is an ordered 0-10 rating with eleven levels that is conventionally
treated as continuous. We fitted a univariable model for each predictor first,
then a single multivariable model containing all of them. Predictors were chosen
in advance on subject-matter grounds, not by significance. Significance threshold
0.05; estimates are reported with 95% confidence intervals. Assumptions were
checked with `performance::check_model()`. Analysis used `r R.version.string`
with tidyverse, gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped).

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(
    missing = "no"
  ) |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

The median medical-school quality-of-life rating was 7 out of 10, one point below
the median overall quality-of-life rating in the same students. Each DREEM
subscale sits near the middle of its possible range, with wide spread, and the
three stages of the course are represented in similar numbers.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Medical-school quality of life against each DREEM subscale. Points are individual students (jittered vertically); lines are linear smoothers. Note the shared y axis and the differing x ranges."}
dat_model |>
  select(medical_school_qol, starts_with("dreem_")) |>
  pivot_longer(starts_with("dreem_"), names_to = "subscale", values_to = "score") |>
  mutate(subscale = recode(subscale,
                           dreem_learning      = "Learning (0-48)",
                           dreem_teachers      = "Teachers (0-44)",
                           dreem_academic_self = "Academic self (0-32)",
                           dreem_atmosphere    = "Atmosphere (0-48)",
                           dreem_social_self   = "Social self (0-28)")) |>
  ggplot(aes(x = score, y = medical_school_qol)) +
  geom_jitter(height = 0.25, width = 0, alpha = 0.15, size = 0.8,
              colour = "#3B7EA1") +
  geom_smooth(method = "lm", se = TRUE, colour = "#D1741F", linewidth = 0.9) +
  facet_wrap(~ subscale, scales = "free_x", nrow = 2) +
  labs(x = "DREEM subscale score", y = "Medical-school quality of life (0-10)")
```

All five subscales slope upward, and the social self-perception panel is the
steepest relative to its narrow 0-28 range - a first hint that how a student
feels about their social life at the school, rather than how they rate the
teaching, is what tracks their overall verdict on medical school.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    hide_n = TRUE
  ) |>
  modify_caption("**Univariable linear regression.**")
uv
```

```{r uv-numbers}
u <- function(v, tm = v) lm(reformulate(v, outcome), data = dat_model) |>
  tidy(conf.int = TRUE) |> filter(term == tm)
uv_learn  <- u("dreem_learning")
uv_teach  <- u("dreem_teachers")
uv_acad   <- u("dreem_academic_self")
uv_atmos  <- u("dreem_atmosphere")
uv_social <- u("dreem_social_self")
uv_sex    <- u("sex", "sexMale")
uv_clerk  <- u("year_group", "year_groupClerkship")
uv_clin   <- u("year_group", "year_groupClinical sciences")
```

Every DREEM subscale is strongly associated with medical-school quality of life
on its own (all p < 0.001). Social self-perception is the strongest per point:
each additional point is associated with a
`r sprintf("%.3f", uv_social$estimate)`-point higher rating (95% CI
`r sprintf("%.3f to %.3f", uv_social$conf.low, uv_social$conf.high)`), so across
its full 0-28 range that is about `r sprintf("%.1f", uv_social$estimate * 28)`
points of the 0-10 outcome. Academic self-perception follows
(`r sprintf("%.3f", uv_acad$estimate)` per point, 95% CI
`r sprintf("%.3f to %.3f", uv_acad$conf.low, uv_acad$conf.high)`), then
atmosphere (`r sprintf("%.3f", uv_atmos$estimate)`), then learning and teachers
(both `r sprintf("%.3f", uv_learn$estimate)` per point). Clerkship students rate
their medical-school quality of life `r sprintf("%.2f", uv_clerk$estimate)`
points lower than basic-sciences students (p =
`r sprintf("%.3f", uv_clerk$p.value)`), and sex shows no crude association
(p = `r sprintf("%.2f", uv_sex$p.value)`).

## Why adjustment is needed

The crude subscale estimates are almost uninterpretable on their own because the
five subscales confound each other. **Perception of atmosphere** is the clearest
example: it correlates 0.77 with perception of learning and 0.76 with academic
self-perception, and it is also associated with the outcome, so the crude
learning coefficient is substantially a measure of atmosphere and vice versa. A
student who is generally positive about their school marks every DREEM item
higher, which is exactly the structure that produces five separately significant
crude coefficients when only one or two components matter. **Social
self-perception** confounds the other four in the same way: it correlates 0.45 to
0.63 with them and it is by far the strongest crude predictor, so a large share
of the crude learning, teachers, academic-self and atmosphere associations is
really social self-perception. Beyond the DREEM subscales, **stage of the
course** is a genuine confounder in the ordinary sense: clerkship students rate
both the educational environment and their medical-school quality of life lower
than basic-sciences students, so it is associated with predictor and outcome
alike and must be held fixed. Fitting all seven predictors together answers the
question the crude models cannot: which *component* of the environment carries
the association once general positivity is accounted for.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit) |>
  modify_caption("**Multivariable linear regression.**")
mv
```

```{r fit-stats}
gl <- glance(fit)
tibble(
  Statistic = c("R-squared", "Adjusted R-squared", "Residual SE", "n", "Model df"),
  Value = c(sprintf("%.3f", gl$r.squared), sprintf("%.3f", gl$adj.r.squared),
            sprintf("%.2f", gl$sigma), as.character(gl$nobs), as.character(gl$df))
) |> gt() |> tab_header(title = md("**Model fit**"))
```

```{r mv-numbers}
tid <- tidy(fit, conf.int = TRUE)
g <- function(x) tid |> filter(term == x)
b_learn <- g("dreem_learning"); b_teach <- g("dreem_teachers")
b_acad <- g("dreem_academic_self"); b_atmos <- g("dreem_atmosphere")
b_social <- g("dreem_social_self"); b_sex <- g("sexMale")
b_clerk <- g("year_groupClerkship"); b_clin <- g("year_groupClinical sciences")
fmt3 <- function(r) sprintf("%.3f (95%% CI %.3f to %.3f)", r$estimate, r$conf.low, r$conf.high)
fmt <- function(r) sprintf("%.2f (95%% CI %.2f to %.2f)", r$estimate, r$conf.low, r$conf.high)
```

The model explains `r sprintf("%.1f%%", gl$r.squared * 100)` of the variance in
self-rated medical-school quality of life (adjusted R-squared
`r sprintf("%.3f", gl$adj.r.squared)`). Only two of the seven predictors survive
mutual adjustment. **Social self-perception** dominates: each additional point is
associated with `r fmt3(b_social)` points of medical-school quality of life
holding the other four DREEM domains, sex and stage fixed - about
`r sprintf("%.1f", b_social$estimate * 28)` points of the 0-10 outcome across the
full subscale range. **Perception of atmosphere** retains a smaller independent
association (`r fmt3(b_atmos)` per point). Perception of learning
(`r fmt3(b_learn)`), perception of teachers (`r fmt3(b_teach)`) and academic
self-perception (`r fmt3(b_acad)`) all have confidence intervals spanning zero,
and the first two have point estimates on the wrong side of it. Sex
(`r fmt(b_sex)`) and stage of the course (clerkship `r fmt(b_clerk)`, clinical
sciences `r fmt(b_clin)`) show no independent association. These are conditional
associations from a cross-sectional survey and describe covariation, not the
effect of changing any DREEM domain.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

This is the most dramatic unadjusted-versus-adjusted table in the dataset, and
almost all of the movement is collinearity rather than confounding in the usual
sense. **Perception of learning** and **perception of teachers** each go from a
highly significant `r sprintf("%.3f", uv_learn$estimate)` and
`r sprintf("%.3f", uv_teach$estimate)` (both p < 0.001) to
`r sprintf("%.3f", b_learn$estimate)` and `r sprintf("%.3f", b_teach$estimate)` -
both cross zero and become non-significant (p =
`r sprintf("%.2f", b_learn$p.value)` and
`r sprintf("%.2f", b_teach$p.value)`). Their entire crude association was shared
with the other three subscales. **Academic self-perception** loses almost as
much, from `r sprintf("%.3f", uv_acad$estimate)` to
`r sprintf("%.3f", b_acad$estimate)`
(`r sprintf("%.0f%%", 100 * b_acad$estimate / uv_acad$estimate)` retained) and
also becomes non-significant. **Perception of atmosphere** keeps
`r sprintf("%.0f%%", 100 * b_atmos$estimate / uv_atmos$estimate)`
(`r sprintf("%.3f", uv_atmos$estimate)` to
`r sprintf("%.3f", b_atmos$estimate)`) and stays significant.

By contrast **social self-perception** barely moves, from
`r sprintf("%.3f", uv_social$estimate)` to
`r sprintf("%.3f", b_social$estimate)` -
`r sprintf("%.0f%%", 100 * b_social$estimate / uv_social$estimate)` retained -
and is the only predictor whose crude story survives intact. Among the
non-DREEM terms, **clerkship** loses its significance
(`r sprintf("%.2f", uv_clerk$estimate)`, p =
`r sprintf("%.3f", uv_clerk$p.value)`, to
`r sprintf("%.2f", b_clerk$estimate)`, p =
`r sprintf("%.2f", b_clerk$p.value)`): clerkship students' lower ratings of
medical school are explained by their lower DREEM scores. **Sex** moves slightly
away from the null in the negative direction
(`r sprintf("%.2f", uv_sex$estimate)` to `r sprintf("%.2f", b_sex$estimate)`) but
remains non-significant in both models. The right reading of this table is not
that teaching and learning do not matter, but that these five subscales are too
correlated for the model to apportion credit among them; only the two most
distinctive - social self-perception and atmosphere - emerge with separable
signal.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r vif-table}
vt <- performance::check_collinearity(fit) |> as_tibble()
vt |>
  transmute(Term, VIF = round(VIF, 2), Tolerance = round(Tolerance, 3)) |>
  arrange(desc(VIF)) |>
  gt() |> tab_header(title = md("**Variance inflation factors**"))
```

```{r het-test}
het <- performance::check_heteroscedasticity(fit)
```

**Collinearity - the main diagnostic finding here.** The five DREEM subscales
have VIFs of
`r paste(sprintf("%.2f", vt |> filter(str_detect(Term, "dreem")) |> arrange(desc(VIF)) |> pull(VIF)), collapse = ", ")`,
the largest being perception of atmosphere at
`r sprintf("%.2f", max(vt$VIF))`. All are below the conventional cut-off of 5, so
the model is estimable and the coefficients are not numerically unstable - but a
VIF of `r sprintf("%.1f", max(vt$VIF))` means the standard error for atmosphere
is `r sprintf("%.1f", sqrt(max(vt$VIF)))` times what it would be if the subscales
were uncorrelated. That inflation is the direct cause of the three
non-significant subscale coefficients above, and it should be reported alongside
them rather than left to the reader to infer.

**Linearity.** The fitted-versus-residual reference line is close to flat with a
mild downward drift at the top of the fitted range, where the 0-10 rating hits
its ceiling.

**Homoscedasticity.** Breusch-Pagan strongly rejects constant variance
(p `r ifelse(attr(het, "p") < 0.001, "< 0.001", sprintf("= %.3f", attr(het, "p")))`):
students predicted to rate their school highly are bunched against the ceiling of
10 and have small residuals, while those predicted low have much wider scatter.
As a sensitivity check we refitted with heteroscedasticity-consistent (HC3)
standard errors:

```{r robust}
rob <- lmtest::coeftest(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
broom::tidy(rob) |>
  transmute(term,
            estimate = round(estimate, 3),
            `robust SE` = round(std.error, 3),
            statistic = round(statistic, 2),
            p.value = format.pval(p.value, digits = 3, eps = 0.001)) |>
  gt() |> tab_header(title = md("**HC3 robust standard errors (sensitivity check)**"))
```

No coefficient changes its significance status under HC3, so the conclusions do
not depend on the constant-variance assumption.

**Normality of residuals.** The Q-Q plot shows a clear left tail: a minority of
students rate medical school 0 to 3 despite favourable DREEM scores, and the
model cannot accommodate them. The formal test rejects normality
(p `r ifelse(attr(performance::check_normality(fit), "p") < 0.001, "< 0.001", sprintf("= %.3f", attr(performance::check_normality(fit), "p")))`).
The residuals are also visibly discrete, because an 11-point rating scale can
only produce residuals on a grid. With n = `r nrow(dat_model)` the coefficient
standard errors remain usable, but prediction intervals from this model are not
meaningful.

**Influential points.** No observation exceeds the Cook's distance threshold; the
largest is `r sprintf("%.3f", max(cooks.distance(fit)))`.

# Discussion

How Brazilian medical students rate their medical-school quality of life is
tracked overwhelmingly by one component of the educational environment: their
social self-perception - whether they have good friends at the school, whether
they are lonely, whether they are too tired to enjoy the course. Once that is
held fixed, perception of the teachers and perception of the learning have no
detectable independent association at all, and perception of the atmosphere
retains only a small one. Practically, this points a dean towards the social
fabric of the school - peer support, timetabling that leaves room for a life,
accommodation, cohort size - rather than towards teaching quality, which
dominates most quality-improvement conversations. The caveat is large, however:
the five DREEM subscales are so intercorrelated that "teaching does not matter"
is not a safe conclusion, only "teaching adds nothing once social experience is
known".

**Limitations.** *Collinearity among the exposures.* With subscale VIFs up to
`r sprintf("%.1f", max(vt$VIF))` the model cannot cleanly separate the five DREEM
components, and the three null coefficients reflect shared variance as much as
absent association. A model using `dreem_global` alone would be cleaner but could
not answer the question posed. *Single-item outcome.* Medical-school quality of
life is one 0-10 question with no reliability estimate; measurement error in it
biases every coefficient toward the null and puts a ceiling on R-squared.
*Clustering is not modelled.* Students came from 22 medical schools, but the
released file carries no school identifier - only `school_legal_status` and
`school_location` - and DREEM scores are substantially a property of the school
rather than the student. The effective sample size for anything school-level is
therefore far below `r nrow(dat_model)`, standard errors are too narrow, and no
random intercept can be fitted. *Cross-sectional design and common-method bias.*
DREEM and the outcome were answered on the same form minutes apart; a student in
a bad mood marks both down, which inflates every association reported here.
*Reverse direction.* A student who is unhappy at medical school will describe
their social self-perception more negatively, so the arrow may point from the
outcome to the predictor.

**Future work.** Measure DREEM at the start of the year and the quality-of-life
rating at the end, retain the school identifier, and fit a mixed model so that
school-level environment can be separated from individual perception of it.

# Feasibility notes

*Not part of a student report.*

- **Knits cleanly.** n = 1,350 complete cases, seven predictors, no sparse cells,
  max Cook's distance `r sprintf("%.3f", max(cooks.distance(fit)))`.
- **The DREEM subscales are collinear but usable.** VIFs: atmosphere
  `r sprintf("%.2f", vt |> filter(Term == "dreem_atmosphere") |> pull(VIF))`,
  learning `r sprintf("%.2f", vt |> filter(Term == "dreem_learning") |> pull(VIF))`,
  academic self `r sprintf("%.2f", vt |> filter(Term == "dreem_academic_self") |> pull(VIF))`,
  teachers `r sprintf("%.2f", vt |> filter(Term == "dreem_teachers") |> pull(VIF))`,
  social self `r sprintf("%.2f", vt |> filter(Term == "dreem_social_self") |> pull(VIF))`.
  All below 5, so nothing breaks - but three of the five coefficients are
  non-significant purely because of the inflation, and a student must be
  prepared to say so rather than reporting "no association".
- **`dreem_global` must never be added.** It is the exact sum of these five
  columns; `lm()` would drop it with an NA coefficient and a student who does not
  notice will report a broken model.
- **This is the best VIF teaching example in the whole dataset.** Two subscales
  flip sign after adjustment, one keeps 79% of its crude coefficient, and the
  contrast between them is entirely explained by the correlation structure.
- **Weakest outcome measurement of the six.** A single 0-10 item with 11 discrete
  levels: residuals are visibly gridded, normality is firmly rejected, and
  heteroscedasticity is strong because of ceiling bunching at 8-10. HC3 rescues
  the inference. The catalogue also offers a binary version (<= 5, 21.8%) if a
  student prefers logistic regression.
- **Verdict: workable and pedagogically valuable**, but only for a student who
  will engage honestly with the collinearity. A student who reports "only social
  self-perception matters" without mentioning VIF has drawn the wrong conclusion.
