---
title: "Self-rated overall quality of life in Brazilian medical students"
subtitle: "Dataset 8: Resilience, learning environment and quality of life in Brazilian medical students"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "medstudent_qol.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "overall_qol"
predictors <- c("sex", "age_yrs", "year_group", "school_legal_status",
                "resilience_score", "dreem_global", "trait_anxiety")

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex = factor(sex, levels = c("female", "male"),
                 labels = c("Female", "Male")),
    year_group = factor(year_group,
                        levels = c("basic_sciences", "clinical_sciences", "clerkship"),
                        labels = c("Basic sciences", "Clinical sciences", "Clerkship")),
    school_legal_status = factor(school_legal_status, levels = c("public", "private"),
                                 labels = c("Public", "Private")),
    school_location = factor(school_location, levels = c("state_capital", "other_city"),
                             labels = c("State capital", "Other city"))
  ) |>
  labelled::set_variable_labels(
    overall_qol         = "Overall quality of life (1-10)",
    sex                 = "Sex",
    age_yrs             = "Age (years)",
    year_group          = "Stage of course",
    school_legal_status = "School legal status",
    resilience_score    = "Resilience score (16-98)",
    dreem_global        = "DREEM global score (0-200)",
    trait_anxiety       = "Trait anxiety (20-80)"
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)
```

# Introduction

Tempski and colleagues surveyed a random sample of 1,350 students at 22 Brazilian
medical schools in 2011-2012, measuring quality of life (WHOQOL-BREF), the
perceived educational environment (DREEM), resilience, depressive symptoms and
anxiety in a single cross-sectional questionnaire (Tempski et al., *PLOS ONE*
2015; doi:10.1371/journal.pone.0131535). Alongside the multi-item instruments,
students rated their *overall* quality of life - life as a whole, not just
medical school - on a single 1-10 item. This global self-rating is the outcome
that matters most to students themselves, and asking how much of it is explained
by anything a medical school controls is a fair test of how far training shapes a
student's life.

**Aim.** Among Brazilian medical students, we examined which of sex, age, stage
of the course, school legal status, resilience, the perceived educational
environment and trait anxiety are independently associated with self-rated
overall quality of life, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** Multicentre cross-sectional survey of a random sample of 1,350
students drawn from 22 medical schools across Brazil, stratified by stage of the
course (basic sciences, clinical sciences, clerkship). All variables used here
are complete.

**Variables.** The outcome is `overall_qol`, the student's self-rating of overall
quality of life on a 1-10 scale (observed range 2 to 10, higher is better).
Predictors are sex (reference: female, the larger group), age in years, stage of
the course (reference: basic sciences, the earliest stage), school legal status
(reference: public, the larger group and the traditional Brazilian medical
school), the resilience scale score (16-98, higher = more resilient), the DREEM
global score (sum of the five DREEM subscales, higher = a more positive perceived
educational environment) and the trait anxiety scale (20-80, higher = more
anxious). Continuous predictors are modelled per one-unit increase.

Deliberate choices about construct overlap. `dreem_global` is used **instead of**
its five subscales, which sum to it exactly. Only one of the three negative-affect
scales is included: `trait_anxiety` correlates 0.73 with `bdi_depression` and
0.70 with `state_anxiety`, so entering all three would split one construct across
three collinear terms; trait anxiety was chosen as the dispositional measure,
least likely to be a same-day consequence of how life is going. The four WHOQOL
domains are excluded as predictors because they are alternative quality-of-life
measurements rather than explanations of quality of life - `whoqol_psychological`
correlates 0.45 with this outcome and would simply restate it.

**Statistical analysis.** Linear regression (`lm()`), justified because the
outcome is an ordered 1-10 rating conventionally treated as continuous. We fitted
a univariable model for each predictor first, then a single multivariable model
containing all of them. Predictors were chosen in advance on subject-matter
grounds, not by significance. Significance threshold 0.05; estimates are reported
with 95% confidence intervals. Assumptions were checked with
`performance::check_model()`. Analysis used `r R.version.string` with tidyverse,
gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped).

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(
    missing = "no"
  ) |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

```{r skew-numbers}
tab_o <- table(dat_model$overall_qol)
pct_high <- 100 * mean(dat_model$overall_qol >= 8)
skew <- (function(x) mean((x - mean(x))^3) / sd(x)^3)(dat_model$overall_qol)
```

The median overall quality-of-life rating was 8 out of 10 (IQR 7 to 9), a point
higher than the same students' rating of their medical-school quality of life.
The distribution is compressed against the top of the scale:
`r sprintf("%.0f%%", pct_high)` of students chose 8, 9 or 10, and only
`r sum(dat_model$overall_qol <= 4)` students in
`r nrow(dat_model)` chose 4 or less. This ceiling effect (skewness
`r sprintf("%.2f", skew)`) limits how much variance any model can explain and
recurs throughout the diagnostics.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Distribution of trait anxiety by overall quality-of-life rating. Boxes show the median and interquartile range; the width of each box is proportional to the square root of the number of students giving that rating."}
dat_model |>
  ggplot(aes(x = factor(overall_qol), y = trait_anxiety)) +
  geom_boxplot(varwidth = TRUE, fill = "#3B7EA1", alpha = 0.75,
               outlier.alpha = 0.3) +
  labs(x = "Overall quality of life (1-10)", y = "Trait anxiety (20-80)")
```

Trait anxiety falls steadily as the overall quality-of-life rating rises, but the
box widths show how unbalanced the outcome is: the great majority of students sit
at 7, 8 or 9, and the ratings at the bottom of the scale rest on a handful of
students each.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    hide_n = TRUE
  ) |>
  modify_caption("**Univariable linear regression.**")
uv
```

```{r uv-numbers}
u <- function(v, tm = v) lm(reformulate(v, outcome), data = dat_model) |>
  tidy(conf.int = TRUE) |> filter(term == tm)
uv_trait <- u("trait_anxiety")
uv_res   <- u("resilience_score")
uv_dreem <- u("dreem_global")
uv_age   <- u("age_yrs")
uv_sex   <- u("sex", "sexMale")
uv_priv  <- u("school_legal_status", "school_legal_statusPrivate")
uv_clerk <- u("year_group", "year_groupClerkship")
uv_clin  <- u("year_group", "year_groupClinical sciences")
```

Trait anxiety is the strongest crude predictor: each additional point is
associated with a `r sprintf("%.3f", abs(uv_trait$estimate))`-point *lower*
overall quality-of-life rating (95% CI
`r sprintf("%.3f to %.3f", uv_trait$conf.low, uv_trait$conf.high)`, p < 0.001),
which across the observed 20-to-80 range is about
`r sprintf("%.1f", abs(uv_trait$estimate) * 60)` points of the 1-10 scale.
Resilience is positively associated (`r sprintf("%.3f", uv_res$estimate)` per
point, 95% CI
`r sprintf("%.3f to %.3f", uv_res$conf.low, uv_res$conf.high)`), as is the DREEM
global score (`r sprintf("%.3f", uv_dreem$estimate)` per point). Each additional
year of age goes with a `r sprintf("%.3f", abs(uv_age$estimate))`-point lower
rating (p < 0.001), and private-school students rate their overall quality of
life `r sprintf("%.3f", uv_priv$estimate)` points higher than public-school
students (p = `r sprintf("%.3f", uv_priv$p.value)`). Sex
(p = `r sprintf("%.2f", uv_sex$p.value)`) and stage of the course
(clerkship p = `r sprintf("%.2f", uv_clerk$p.value)`) show no crude association.

## Why adjustment is needed

Several crude estimates here are clearly contaminated by other variables in the
file. **Trait anxiety** confounds the resilience association: dispositionally
anxious students score much lower on the resilience scale (*r* = -0.61) and rate
their lives worse, so a large part of the crude resilience coefficient is really
anxiety. **Trait anxiety** does the same to the DREEM association, since anxious
students rate their educational environment more harshly (*r* = -0.43); without
adjustment, "a better learning environment goes with a better life" is partly
just "less anxious students say better things about everything". **Stage of the
course** confounds the crude age association, and severely: clerkship students
are on average four years older than basic-sciences students, so the crude age
coefficient mixes ageing with progression through the course, and the two must be
separated. **The DREEM global score** confounds the school legal status contrast:
private-school students rate their educational environment about 13 points higher
on average (mean 128 versus 115), so their apparent overall quality-of-life
advantage may be nothing more than that. Fitting all seven predictors together
makes each estimate conditional on the rest.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit) |>
  modify_caption("**Multivariable linear regression.**")
mv
```

```{r fit-stats}
gl <- glance(fit)
tibble(
  Statistic = c("R-squared", "Adjusted R-squared", "Residual SE", "n", "Model df"),
  Value = c(sprintf("%.3f", gl$r.squared), sprintf("%.3f", gl$adj.r.squared),
            sprintf("%.2f", gl$sigma), as.character(gl$nobs), as.character(gl$df))
) |> gt() |> tab_header(title = md("**Model fit**"))
```

```{r mv-numbers}
tid <- tidy(fit, conf.int = TRUE)
g <- function(x) tid |> filter(term == x)
b_trait <- g("trait_anxiety"); b_res <- g("resilience_score")
b_dreem <- g("dreem_global"); b_age <- g("age_yrs"); b_sex <- g("sexMale")
b_priv <- g("school_legal_statusPrivate"); b_clerk <- g("year_groupClerkship")
b_clin <- g("year_groupClinical sciences")
fmt3 <- function(r) sprintf("%.3f (95%% CI %.3f to %.3f)", r$estimate, r$conf.low, r$conf.high)
```

The model explains `r sprintf("%.1f%%", gl$r.squared * 100)` of the variance in
overall quality of life (adjusted R-squared
`r sprintf("%.3f", gl$adj.r.squared)`) - the lowest of any outcome in this
dataset, and a useful reminder that a single global life rating is driven mostly
by things a medical-school survey does not measure. Holding the other six
predictors fixed, each trait-anxiety point is associated with `r fmt3(b_trait)`
points of the rating (about
`r sprintf("%.1f", abs(b_trait$estimate) * 60)` points across the observed
range), each DREEM point with `r fmt3(b_dreem)` (about
`r sprintf("%.2f", b_dreem$estimate * 20)` points per 20 DREEM points) and each
year of age with `r fmt3(b_age)` points. Resilience retains a small but
significant independent association (`r fmt3(b_res)` per point). Clerkship
students rate their overall quality of life `r fmt3(b_clerk)` points higher than
basic-sciences students (clinical sciences `r fmt3(b_clin)`) once age and the
rest are held fixed. Sex (`r fmt3(b_sex)`) and private schooling
(`r fmt3(b_priv)`) show no independent association. These are conditional
associations in a cross-sectional survey, not effects.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

**Resilience** moves furthest toward the null, from
`r sprintf("%.3f", uv_res$estimate)` crude to
`r sprintf("%.3f", b_res$estimate)` adjusted - only
`r sprintf("%.0f%%", 100 * b_res$estimate / uv_res$estimate)` of the crude
coefficient survives, and it is barely significant
(p = `r sprintf("%.3f", b_res$p.value)`) where crudely it was overwhelming. Most
of what the resilience scale contributed was trait anxiety. **The DREEM global
score** loses about half (`r sprintf("%.3f", uv_dreem$estimate)` to
`r sprintf("%.3f", b_dreem$estimate)`,
`r sprintf("%.0f%%", 100 * b_dreem$estimate / uv_dreem$estimate)` retained), for
the same reason. **Trait anxiety** shrinks least
(`r sprintf("%.3f", uv_trait$estimate)` to
`r sprintf("%.3f", b_trait$estimate)`,
`r sprintf("%.0f%%", 100 * b_trait$estimate / uv_trait$estimate)` retained) and
is the dominant predictor either way.

Two coefficients change significance in opposite directions. **Private
schooling** is significant crudely (`r sprintf("%.3f", uv_priv$estimate)`, p =
`r sprintf("%.3f", uv_priv$p.value)`) and null after adjustment
(`r sprintf("%.3f", b_priv$estimate)`, p = `r sprintf("%.2f", b_priv$p.value)`):
the private-school advantage was the DREEM difference between private and public
schools, not the schools themselves. **Clerkship** flips sign, from
`r sprintf("%.3f", uv_clerk$estimate)` crude and non-significant
(p = `r sprintf("%.2f", uv_clerk$p.value)`) to
`r sprintf("%.3f", b_clerk$estimate)` adjusted and significant
(p = `r sprintf("%.3f", b_clerk$p.value)`) - classic suppression by age, since
clerkship students are about four years older and age itself carries a negative
association. **Age** correspondingly moves *away* from the null
(`r sprintf("%.3f", uv_age$estimate)` to `r sprintf("%.3f", b_age$estimate)`)
once stage is held fixed. **Sex** barely moves in absolute terms
(`r sprintf("%.3f", uv_sex$estimate)` to `r sprintf("%.3f", b_sex$estimate)`) and
is non-significant in both models.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r vif-table}
vt <- performance::check_collinearity(fit) |> as_tibble()
vt |>
  transmute(Term, VIF = round(VIF, 2), Tolerance = round(Tolerance, 3)) |>
  arrange(desc(VIF)) |>
  gt() |> tab_header(title = md("**Variance inflation factors**"))
```

```{r het-test}
het <- performance::check_heteroscedasticity(fit)
```

**Linearity.** The fitted-versus-residual reference line is broadly flat but
curves down at the top of the fitted range: the model cannot predict above 10 and
students it places near the ceiling have systematically negative residuals.

**Homoscedasticity.** Breusch-Pagan strongly rejects constant variance
(p `r ifelse(attr(het, "p") < 0.001, "< 0.001", sprintf("= %.3f", attr(het, "p")))`).
Residual spread is much wider for students the model predicts to rate their life
poorly than for those it predicts near the ceiling, where almost everyone answers
8 or 9. As a sensitivity check we refitted with heteroscedasticity-consistent
(HC3) standard errors:

```{r robust}
rob <- lmtest::coeftest(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
broom::tidy(rob) |>
  transmute(term,
            estimate = round(estimate, 3),
            `robust SE` = round(std.error, 3),
            statistic = round(statistic, 2),
            p.value = format.pval(p.value, digits = 3, eps = 0.001)) |>
  gt() |> tab_header(title = md("**HC3 robust standard errors (sensitivity check)**"))
```

The robust standard errors are close to the model-based ones, but one conclusion
does change: **resilience** moves from p =
`r sprintf("%.3f", b_res$p.value)` under model-based standard errors to
p = `r sprintf("%.3f", broom::tidy(rob) |> filter(term == "resilience_score") |> pull(p.value))`
under HC3, crossing the 0.05 threshold. It was borderline to begin with and
should be described as such rather than as an established association. Trait
anxiety, the DREEM score, age and the clerkship contrast all keep their
significance status.

**Normality of residuals.** The Q-Q plot has a pronounced left tail. The outcome
is skewed toward the top of the scale (skewness
`r sprintf("%.2f", skew)`) and the residuals inherit it, with normality firmly
rejected
(p `r ifelse(attr(performance::check_normality(fit), "p") < 0.001, "< 0.001", sprintf("= %.3f", attr(performance::check_normality(fit), "p")))`).
The residuals are also visibly discrete because the outcome takes only nine
observed values. Coefficient standard errors survive this at
n = `r nrow(dat_model)`, but prediction intervals from this model are not
trustworthy.

**Influential points.** No observation exceeds the Cook's distance threshold; the
largest is `r sprintf("%.3f", max(cooks.distance(fit)))`. The few students who
rated their life 2 or 3 are the highest-leverage points but none is individually
decisive.

**Collinearity.** The highest VIF is `r sprintf("%.2f", max(vt$VIF))`
(`r vt |> arrange(desc(VIF)) |> slice(1) |> pull(Term)`), with trait anxiety at
`r sprintf("%.2f", vt |> filter(Term == "trait_anxiety") |> pull(VIF))` and
resilience at
`r sprintf("%.2f", vt |> filter(Term == "resilience_score") |> pull(VIF))`. All
values are below 2 and far from the threshold of 5, so the model is stable. This
is the payoff for excluding `bdi_depression` and `state_anxiety`: had all three
negative-affect scales been entered, their mutual correlations of 0.59 to 0.73
would have inflated the standard errors substantially and made the split between
them arbitrary.

# Discussion

Self-rated overall quality of life in Brazilian medical students is associated
most strongly with trait anxiety, and only weakly with anything the medical
school controls. A 20-point better perceived educational environment corresponds
to roughly `r sprintf("%.2f", b_dreem$estimate * 20)` of a point on the 1-10
scale - real but small next to the roughly
`r sprintf("%.1f", abs(b_trait$estimate) * 20)` points associated with a 20-point
difference in trait anxiety. The model explains
`r sprintf("%.0f%%", gl$r.squared * 100)` of the variance, the least of any
outcome in this dataset, which is the honest headline: how a student rates their
life as a whole depends chiefly on family, relationships, money and temperament,
of which this survey measured only temperament. Compared with
`medical_school_qol` - the same students' rating of their medical-school quality
of life - the DREEM association here is roughly half as strong, exactly as one
would expect if the learning environment affects the school-specific part of life
more than life in general.

**Limitations.** *Ceiling effect and a single-item outcome.* Overall quality of
life is one 1-10 question with no reliability estimate;
`r sprintf("%.0f%%", pct_high)` of students answered 8 or more, so the outcome
has little room to vary upward, measurement error attenuates every coefficient
toward the null, and R-squared is capped low for reasons that have nothing to do
with the predictors. *Clustering is not modelled.* Students came from 22 medical
schools, but the released file carries no school identifier - only
`school_legal_status` and `school_location` - so students within a school cannot
be treated as correlated; the standard errors are too narrow, particularly for
the school-level contrast, and no random intercept can be fitted.
*Cross-sectional design.* Trait anxiety and quality of life were measured on the
same day, and a student having a bad year will report both a low life rating and
high anxiety; the direction of the association cannot be established.
*Common-method bias.* Every psychosocial variable is self-reported on one form,
which inflates all associations at once. *Unmeasured confounders* - family
income, relationship status, living arrangements, physical illness - are absent
from the file and are precisely the determinants of a global life rating.

**Future work.** Add a small number of socio-demographic items (income,
relationship status, living arrangements) and repeat the rating at two time
points, retaining school identifiers so that between-school variation can be
separated out.

# Feasibility notes

*Not part of a student report.*

- **Knits cleanly, statistically stable.** n = 1,350 complete cases, seven
  predictors, max VIF `r sprintf("%.2f", max(vt$VIF))`, max Cook's distance
  `r sprintf("%.3f", max(cooks.distance(fit)))`, no sparse cells among the
  predictors.
- **Lowest R-squared of the six outcomes**
  (`r sprintf("%.3f", gl$r.squared)`, adjusted
  `r sprintf("%.3f", gl$adj.r.squared)`). Nothing is broken; the outcome is
  simply a one-item global life rating that this survey is not equipped to
  explain. Students should be warned in advance so they do not read a low
  R-squared as their own failure - but a student who wants an impressive-looking
  model will be disappointed.
- **Worst outcome distribution of the six after `bdi_depression`.** Nine observed
  values, `r sprintf("%.0f%%", pct_high)` of students at 8-10, skewness
  `r sprintf("%.2f", skew)`, and only
  `r sum(dat_model$overall_qol <= 4)` students at 4 or below. Normality is firmly
  rejected, residuals are gridded, and heteroscedasticity is strong. This is the
  **only outcome of the six where HC3 changes a conclusion**: resilience goes
  from p = `r sprintf("%.3f", b_res$p.value)` to
  p = `r sprintf("%.3f", broom::tidy(rob) |> filter(term == "resilience_score") |> pull(p.value))`.
  The linear model is a convenience here, not a good fit; ordinal regression
  would be the honest alternative.
- **Least tautological of the six.** This is the only outcome whose strongest
  predictor (trait anxiety, *r* = -0.37) is not near-collinear with it. Compare
  `whoqol_psychological`, where trait anxiety correlates -0.75. The trade-off is
  the low R-squared - the two facts are the same fact.
- **Good suppression example.** Clerkship flips from null to significantly
  positive once age is held fixed, and age strengthens correspondingly; the
  age/stage relationship in this file (mean age 20.9, 22.8, 24.8 across the three
  stages) makes this easy to explain.
- **Guardrails.** `dreem_global` must never appear with its five subscales; no
  WHOQOL domain as a predictor; only one of BDI / state / trait anxiety.
- **Verdict: workable and safe, but underpowered in explanatory terms.** A good
  choice for a student who understands that a small R-squared can be the right
  answer; a poor choice for one who wants a strong result.
