---
title: "Physical-health quality of life in Brazilian medical students"
subtitle: "Dataset 8: Resilience, learning environment and quality of life in Brazilian medical students"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "medstudent_qol.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "whoqol_physical"
predictors <- c("sex", "age_yrs", "year_group", "school_legal_status",
                "resilience_score", "dreem_global", "bdi_depression")

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex = factor(sex, levels = c("female", "male"),
                 labels = c("Female", "Male")),
    year_group = factor(year_group,
                        levels = c("basic_sciences", "clinical_sciences", "clerkship"),
                        labels = c("Basic sciences", "Clinical sciences", "Clerkship")),
    school_legal_status = factor(school_legal_status, levels = c("public", "private"),
                                 labels = c("Public", "Private")),
    school_location = factor(school_location, levels = c("state_capital", "other_city"),
                             labels = c("State capital", "Other city"))
  ) |>
  labelled::set_variable_labels(
    whoqol_physical     = "WHOQOL physical domain (0-100)",
    sex                 = "Sex",
    age_yrs             = "Age (years)",
    year_group          = "Stage of course",
    school_legal_status = "School legal status",
    resilience_score    = "Resilience score (16-98)",
    dreem_global        = "DREEM global score (0-200)",
    bdi_depression      = "BDI depressive symptoms (0-63)"
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)
```

# Introduction

Tempski and colleagues surveyed a random sample of 1,350 students at 22 Brazilian
medical schools in 2011-2012, measuring quality of life (WHOQOL-BREF), the
perceived educational environment (DREEM), resilience, depressive symptoms and
anxiety in a single cross-sectional questionnaire (Tempski et al., *PLOS ONE*
2015; doi:10.1371/journal.pone.0131535). The physical domain of the WHOQOL-BREF
covers sleep, energy, pain, mobility, everyday activity capacity and dependence
on medication - the concrete daily costs of medical training, and the domain most
likely to respond to changes in workload and rota design.

**Aim.** Among Brazilian medical students, we examined which of sex, age, stage
of the course, school legal status, resilience, the perceived educational
environment and depressive symptoms are independently associated with
physical-health quality of life, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** Multicentre cross-sectional survey of a random sample of 1,350
students drawn from 22 medical schools across Brazil, stratified by stage of the
course (basic sciences, clinical sciences, clerkship). All variables used here
are complete.

**Variables.** The outcome is `whoqol_physical`, the WHOQOL-BREF physical health
domain transformed to a 0-100 scale where higher is better. Predictors are sex
(reference: female, the larger group), age in years, stage of the course
(reference: basic sciences, the earliest stage), school legal status
(reference: public, the larger group and the traditional Brazilian medical
school), the resilience scale score (16-98, higher = more resilient), the DREEM
global score (sum of the five DREEM subscales, higher = a more positive
perceived educational environment), and the Beck Depression Inventory total
(0-63, higher = more depressive symptoms). All continuous predictors are
modelled per one-unit increase.

Three deliberate choices about construct overlap. First, `dreem_global` is used
**instead of** the five DREEM subscales, because the global score is their exact
arithmetic sum and the two sets cannot appear in one model. Second, no other
WHOQOL domain is used as a predictor: physical, psychological, social and
environment are subscales of one instrument, and regressing one on another
describes the questionnaire rather than the students. Third, of the three
negative-affect scales (`bdi_depression`, `state_anxiety`, `trait_anxiety`) only
the BDI is included. The three correlate at *r* = 0.59 to 0.73 with each other;
including all three would split one construct across three collinear terms and
make each coefficient uninterpretable. The BDI was chosen because somatic items
(sleep, appetite, fatigue) give it the clearest substantive link to physical
quality of life - a link that is also its weakness, discussed below.

**Statistical analysis.** Linear regression (`lm()`), justified because the
outcome is continuous and roughly symmetric. We fitted a univariable model for
each predictor first, then a single multivariable model containing all of them.
Predictors were chosen in advance on subject-matter grounds, not by significance.
Significance threshold 0.05; estimates are reported with 95% confidence
intervals. Assumptions were checked with `performance::check_model()`. Analysis
used `r R.version.string` with tidyverse, gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped).

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(
    missing = "no"
  ) |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

The median physical-domain score was 67.9 out of 100, higher than the
psychological domain in the same students. Two thirds attended public schools,
students were evenly spread across the three stages of the course, and the median
BDI score of 8 sits below the conventional threshold of 14 for moderate
depression - though the upper range reaches 42.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Physical-health quality of life against depressive symptoms, by stage of the course. Points are individual students; lines are linear smoothers."}
ggplot(dat_model, aes(x = bdi_depression, y = whoqol_physical,
                      colour = year_group)) +
  geom_point(alpha = 0.3, size = 1.2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.9) +
  geom_vline(xintercept = 14, linetype = "dashed", colour = "grey40") +
  annotate("text", x = 14.8, y = 15, hjust = 0, size = 3.2, colour = "grey30",
           label = "BDI 14 = moderate depression") +
  scale_colour_manual(values = c("#3B7EA1", "#D1741F", "#5B8C5A")) +
  labs(x = "BDI depressive symptoms (0-63)",
       y = "WHOQOL physical domain (0-100)",
       colour = "Stage of course") +
  theme(legend.position = "top")
```

Physical quality of life declines steadily and near-linearly with depressive
symptoms across the whole BDI range, and the clerkship line sits slightly above
the basic-sciences line at every level of depression, hinting that the later
stages of the course are not the worst for physical wellbeing.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    hide_n = TRUE
  ) |>
  modify_caption("**Univariable linear regression.**")
uv
```

```{r uv-numbers}
u <- function(v, term = v) lm(reformulate(v, outcome), data = dat_model) |>
  tidy(conf.int = TRUE) |> filter(term == !!term)
uv_bdi   <- u("bdi_depression")
uv_res   <- u("resilience_score")
uv_dreem <- u("dreem_global")
uv_sex   <- u("sex", "sexMale")
uv_priv  <- u("school_legal_status", "school_legal_statusPrivate")
uv_clerk <- u("year_group", "year_groupClerkship")
uv_age   <- u("age_yrs")
```

Depressive symptoms are the strongest crude predictor: each additional BDI point
is associated with a `r sprintf("%.2f", abs(uv_bdi$estimate))`-point *lower*
physical-domain score (95% CI
`r sprintf("%.2f to %.2f", uv_bdi$conf.low, uv_bdi$conf.high)`, p < 0.001), so a
student at the moderate-depression threshold of 14 scores about
`r sprintf("%.0f", abs(uv_bdi$estimate) * 14)` points lower than one scoring 0.
Resilience (`r sprintf("%.2f", uv_res$estimate)` points per point, 95% CI
`r sprintf("%.2f to %.2f", uv_res$conf.low, uv_res$conf.high)`) and the DREEM
global score (`r sprintf("%.2f", uv_dreem$estimate)` per point) are both
positively associated. Men score
`r sprintf("%.2f", uv_sex$estimate)` points higher than women crudely, clerkship
students `r sprintf("%.2f", uv_clerk$estimate)` points higher than basic-sciences
students, and private-school students `r sprintf("%.2f", uv_priv$estimate)`
points higher than public-school students. Age shows no crude association
(p = `r sprintf("%.2f", uv_age$p.value)`).

## Why adjustment is needed

The crude estimates are confounded by variables that sit upstream of both the
predictor and the outcome. **Depressive symptoms** confound the resilience
association: students with high BDI scores score much lower on the resilience
scale (*r* = -0.57) and also report worse physical quality of life, so a large
part of the crude resilience coefficient is depression rather than resilience.
**Depressive symptoms** also confound the DREEM association in the same
direction - a depressed student rates the learning environment more harshly
(*r* = -0.41) and reports worse physical wellbeing - so the crude environment
effect is partly the mood of the rater. **Stage of the course** confounds the
crude age association: clerkship students are older by construction, and the
stages differ in workload and clinical exposure, so without adjustment an age
coefficient is really a mixture of ageing and progression. **School legal status**
is a fourth: private-school students in Brazil come from wealthier families, and
family wealth plausibly affects both the ability to afford a private place and
the sleep, nutrition and health care that the physical domain measures. Fitting
all seven predictors together lets each be read holding the others fixed.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit) |>
  modify_caption("**Multivariable linear regression.**")
mv
```

```{r fit-stats}
gl <- glance(fit)
tibble(
  Statistic = c("R-squared", "Adjusted R-squared", "Residual SE", "n", "Model df"),
  Value = c(sprintf("%.3f", gl$r.squared), sprintf("%.3f", gl$adj.r.squared),
            sprintf("%.2f", gl$sigma), as.character(gl$nobs), as.character(gl$df))
) |> gt() |> tab_header(title = md("**Model fit**"))
```

```{r mv-numbers}
tid <- tidy(fit, conf.int = TRUE)
g <- function(x) tid |> filter(term == x)
b_bdi <- g("bdi_depression"); b_res <- g("resilience_score")
b_dreem <- g("dreem_global"); b_sex <- g("sexMale")
b_clerk <- g("year_groupClerkship"); b_clin <- g("year_groupClinical sciences")
b_priv <- g("school_legal_statusPrivate"); b_age <- g("age_yrs")
fmt <- function(r) sprintf("%.2f (95%% CI %.2f to %.2f)", r$estimate, r$conf.low, r$conf.high)
```

The model explains `r sprintf("%.1f%%", gl$r.squared * 100)` of the variance in
physical quality of life (adjusted R-squared
`r sprintf("%.3f", gl$adj.r.squared)`). Holding the other six predictors fixed,
each BDI point is associated with a `r fmt(b_bdi)`-point change in the physical
domain - still by far the largest contribution. The DREEM global score keeps an
independent association of `r fmt(b_dreem)` points per point, about
`r sprintf("%.1f", b_dreem$estimate * 20)` points across a 20-point difference in
perceived environment. Resilience retains a small independent association
(`r fmt(b_res)` points per point). Men score `r fmt(b_sex)` points higher than
women. Notably, the stage of the course strengthens after adjustment: clerkship
students score `r fmt(b_clerk)` points higher and clinical-sciences students
`r fmt(b_clin)` points higher than basic-sciences students, conditional on
depression and everything else. Age (`r fmt(b_age)`) and private schooling
(`r fmt(b_priv)`) show no independent association. All of these are conditional
associations in a single-timepoint survey, not effects of intervention.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

**Resilience** moves the furthest toward the null, from
`r sprintf("%.2f", uv_res$estimate)` crude to
`r sprintf("%.2f", b_res$estimate)` adjusted - only
`r sprintf("%.0f%%", 100 * b_res$estimate / uv_res$estimate)` of its crude size
survives. Almost all of the crude resilience association was depression: highly
resilient students are mostly students with low BDI scores. **The DREEM global
score** halves, from `r sprintf("%.2f", uv_dreem$estimate)` to
`r sprintf("%.2f", b_dreem$estimate)`
(`r sprintf("%.0f%%", 100 * b_dreem$estimate / uv_dreem$estimate)` retained), for
the same reason. By contrast **depressive symptoms** barely move, from
`r sprintf("%.2f", uv_bdi$estimate)` to `r sprintf("%.2f", b_bdi$estimate)`
(`r sprintf("%.0f%%", 100 * b_bdi$estimate / uv_bdi$estimate)` retained) - none
of the other six predictors explains away the BDI association.

Two predictors change significance. **Private schooling** is significant crudely
(`r sprintf("%.2f", uv_priv$estimate)` points, p =
`r sprintf("%.3f", uv_priv$p.value)`) but vanishes after adjustment
(`r sprintf("%.2f", b_priv$estimate)`, p = `r sprintf("%.2f", b_priv$p.value)`):
the crude advantage of private-school students was their lower depression scores,
not the school. **Stage of the course** moves *away* from the null - clerkship
goes from `r sprintf("%.2f", uv_clerk$estimate)` to
`r sprintf("%.2f", b_clerk$estimate)` points - because clerkship students are
also more depressed on average, and their physical-wellbeing advantage was being
masked by that. **Sex** shrinks by more than half
(`r sprintf("%.2f", uv_sex$estimate)` to `r sprintf("%.2f", b_sex$estimate)`) but
stays significant: women score worse on the physical domain partly, but not
wholly, because they report more depressive symptoms. **Age** barely moves
(`r sprintf("%.2f", uv_age$estimate)` to `r sprintf("%.2f", b_age$estimate)`) and
is non-significant in both models.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r vif-table}
performance::check_collinearity(fit) |>
  as_tibble() |>
  transmute(Term, VIF = round(VIF, 2), Tolerance = round(Tolerance, 3)) |>
  gt() |> tab_header(title = md("**Variance inflation factors**"))
```

```{r het-test}
het <- performance::check_heteroscedasticity(fit)
```

**Linearity.** The fitted-versus-residual panel is close to flat; the reference
line drifts down slightly at the highest fitted values because the WHOQOL scale
is capped at 100 and cannot accommodate students the model predicts above it.

**Homoscedasticity.** Breusch-Pagan rejects constant variance
(p = `r sprintf("%.3f", attr(het, "p"))`), with residual spread widening at low
fitted values. As a sensitivity check we refitted with
heteroscedasticity-consistent (HC3) standard errors:

```{r robust}
rob <- lmtest::coeftest(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
broom::tidy(rob) |>
  transmute(term,
            estimate = round(estimate, 3),
            `robust SE` = round(std.error, 3),
            statistic = round(statistic, 2),
            p.value = format.pval(p.value, digits = 3, eps = 0.001)) |>
  gt() |> tab_header(title = md("**HC3 robust standard errors (sensitivity check)**"))
```

Every conclusion is unchanged under HC3; the robust standard errors differ from
the model-based ones by only a few percent, so the heteroscedasticity is a
bounded-scale artefact rather than a threat to inference.

**Normality of residuals.** The Q-Q plot follows the diagonal through the middle
with a mild left tail. The formal test flags non-normality
(p = `r sprintf("%.3f", attr(performance::check_normality(fit), "p"))`), but at
n = `r nrow(dat_model)` this detects trivial departures and does not threaten the
coefficient standard errors.

**Influential points.** No observation exceeds the Cook's distance threshold; the
largest Cook's distance is `r sprintf("%.3f", max(cooks.distance(fit)))`, far
below 0.5.

**Collinearity.** The highest VIF is
`r sprintf("%.2f", max(performance::check_collinearity(fit)$VIF))`, all values
are below 2, and no term is near the conventional threshold of 5. This is the
payoff for using one negative-affect scale rather than three: had
`state_anxiety` and `trait_anxiety` been added alongside the BDI, the VIFs for
the three would have risen sharply and the split of the joint effect between them
would have been arbitrary.

# Discussion

Among Brazilian medical students, physical-health quality of life is most
strongly and independently associated with depressive symptoms; a student at the
moderate-depression threshold scores roughly
`r sprintf("%.0f", abs(b_bdi$estimate) * 14)` points lower on the 0-100 physical
domain than a student with no depressive symptoms, holding sex, age, stage,
school type, resilience and the perceived environment fixed. The perceived
educational environment retains a genuine independent association of about
`r sprintf("%.1f", b_dreem$estimate * 20)` points per 20 DREEM points, which is
the most actionable finding here because DREEM scores are a school-level target
in a way that a student's depression is not. Resilience, once depression is
accounted for, adds little.

**Limitations.** *Construct overlap between the BDI and the outcome.* The BDI
contains somatic items on sleep, fatigue, appetite and loss of libido, and the
WHOQOL physical domain asks about sleep, energy and everyday capacity; part of
the large BDI coefficient is the two instruments asking the same questions. The
overlap is smaller than for the psychological domain but it is real, and the
coefficient should not be read as "treating depression would raise physical
quality of life by this much". *Clustering is not modelled.* Students came from
22 medical schools, but the released file carries no school identifier - only
`school_legal_status` and `school_location` - so the shared teachers, curriculum,
rota and city of each school cannot be represented. Standard errors are therefore
too narrow, and the school-type coefficient is a crude two-category stand-in for
between-school variation that would have been better handled with a random
intercept. *Cross-sectional design.* Everything was measured at one sitting;
poor physical health causing depression is as plausible as the reverse.
*Common-method bias.* All psychosocial measures are self-reported on the same
questionnaire, so a globally negative respondent inflates every association at
once. *No objective measures* of sleep, exercise, workload or hours on duty are
available, so the mechanism behind the clerkship advantage cannot be examined.

**Future work.** Link the questionnaire to objective rota and sleep data and
retain the school identifier, so that between-school differences in workload can
be separated from between-student differences in mood.

# Feasibility notes

*Not part of a student report.*

- **Knits cleanly and is statistically well behaved.** n = 1,350 complete cases,
  seven predictors, max VIF
  `r sprintf("%.2f", max(performance::check_collinearity(fit)$VIF))`, max Cook's
  distance `r sprintf("%.3f", max(cooks.distance(fit)))`, no sparse cells.
- **Milder construct overlap than the psychological domain.** The BDI correlates
  -0.60 with this outcome versus -0.74 with `whoqol_psychological`, and trait
  anxiety -0.59 versus -0.75. This makes `whoqol_physical` a *safer* teaching
  outcome: a student can use a mood scale as a predictor without the model
  degenerating into one questionnaire predicting itself.
- **Real, interpretable adjustment story.** Resilience keeps only
  `r sprintf("%.0f%%", 100 * b_res$estimate / uv_res$estimate)` of its crude
  coefficient and private schooling loses significance entirely, while stage of
  the course moves *away* from the null. All three patterns are teachable, and
  the suppression example (clerkship) is unusually clean.
- **Guardrails still needed.** `dreem_global` must never appear with the five
  DREEM subscales (it is their exact sum), no other WHOQOL domain may be used as
  a predictor, and only one of BDI / state anxiety / trait anxiety should enter
  the model.
- **Heteroscedasticity is flagged** (BP p = `r sprintf("%.3f", attr(het, "p"))`)
  as it is for every outcome in this dataset - a floor/ceiling artefact of
  bounded 0-100 scales. HC3 changes nothing.
- **Verdict: the strongest all-round choice of the six.** Good effect sizes, low
  collinearity, a genuine confounding story, and less tautology than the
  psychological domain.
