---
title: "Environment-domain quality of life in Brazilian medical students"
subtitle: "Dataset 8: Resilience, learning environment and quality of life in Brazilian medical students"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "medstudent_qol.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "whoqol_environment"
predictors <- c("sex", "year_group", "school_legal_status", "school_location",
                "dreem_global", "resilience_score", "trait_anxiety")

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex = factor(sex, levels = c("female", "male"),
                 labels = c("Female", "Male")),
    year_group = factor(year_group,
                        levels = c("basic_sciences", "clinical_sciences", "clerkship"),
                        labels = c("Basic sciences", "Clinical sciences", "Clerkship")),
    school_legal_status = factor(school_legal_status, levels = c("public", "private"),
                                 labels = c("Public", "Private")),
    school_location = factor(school_location, levels = c("state_capital", "other_city"),
                             labels = c("State capital", "Other city"))
  ) |>
  labelled::set_variable_labels(
    whoqol_environment  = "WHOQOL environment domain (0-100)",
    sex                 = "Sex",
    year_group          = "Stage of course",
    school_legal_status = "School legal status",
    school_location     = "School location",
    dreem_global        = "DREEM global score (0-200)",
    resilience_score    = "Resilience score (16-98)",
    trait_anxiety       = "Trait anxiety (20-80)"
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)
```

# Introduction

Tempski and colleagues surveyed a random sample of 1,350 students at 22 Brazilian
medical schools in 2011-2012, measuring quality of life (WHOQOL-BREF), the
perceived educational environment (DREEM), resilience, depressive symptoms and
anxiety in a single cross-sectional questionnaire (Tempski et al., *PLOS ONE*
2015; doi:10.1371/journal.pone.0131535). The environment domain of the
WHOQOL-BREF asks about financial resources, physical safety, home environment,
transport, access to health care, opportunities for recreation and access to
information - the material circumstances of a student's life rather than their
mood. It is the domain most likely to differ between schools and cities, which
makes the two school-level variables in this file directly relevant.

**Aim.** Among Brazilian medical students, we examined which of sex, stage of the
course, school legal status, school location, the perceived educational
environment, resilience and trait anxiety are independently associated with
environment-domain quality of life, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** Multicentre cross-sectional survey of a random sample of 1,350
students drawn from 22 medical schools across Brazil, stratified by stage of the
course (basic sciences, clinical sciences, clerkship). All variables used here
are complete.

**Variables.** The outcome is `whoqol_environment`, the WHOQOL-BREF environment
domain transformed to a 0-100 scale where higher is better. Predictors are sex
(reference: female, the larger group), stage of the course (reference: basic
sciences, the earliest stage), school legal status (reference: public, the larger
group and the traditional Brazilian medical school), school location (reference:
state capital, the larger group and the better-resourced setting), the DREEM
global score (sum of the five DREEM subscales, higher = a more positive perceived
educational environment), the resilience scale score (16-98) and the trait
anxiety scale (20-80, higher = more anxious). All continuous predictors are
modelled per one-unit increase.

Deliberate choices about construct overlap. `dreem_global` is used **instead of**
its five subscales, which sum to it exactly. No other WHOQOL domain is used as a
predictor, because they are subscales of the same instrument as the outcome. Of
the three negative-affect scales, only `trait_anxiety` is included: it correlates
0.73 with `bdi_depression` and 0.70 with `state_anxiety`, and entering all three
would split one construct across three collinear terms. Trait anxiety was chosen
over state anxiety because it is the dispositional measure and therefore less
likely to be a same-day consequence of the material circumstances being measured.

**Statistical analysis.** Linear regression (`lm()`), justified because the
outcome is continuous and roughly symmetric. We fitted a univariable model for
each predictor first, then a single multivariable model containing all of them.
Predictors were chosen in advance on subject-matter grounds, not by significance.
Significance threshold 0.05; estimates are reported with 95% confidence
intervals. Assumptions were checked with `performance::check_model()`. Analysis
used `r R.version.string` with tidyverse, gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped).

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(
    missing = "no"
  ) |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

The median environment-domain score was 65.6 out of 100. Just under two thirds of
students attended public schools and just under three fifths attended a school in
a state capital, so both school-level contrasts have adequate numbers on each
side.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Environment-domain quality of life by school legal status and school location. Boxes show the median and interquartile range; the diamond marks the mean."}
ggplot(dat_model, aes(x = school_location, y = whoqol_environment,
                      fill = school_legal_status)) +
  geom_boxplot(alpha = 0.75, outlier.alpha = 0.25, width = 0.6,
               position = position_dodge(width = 0.7)) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 2.6,
               fill = "white", position = position_dodge(width = 0.7)) +
  scale_fill_manual(values = c("#3B7EA1", "#D1741F")) +
  labs(x = "School location", y = "WHOQOL environment domain (0-100)",
       fill = "School legal status") +
  theme(legend.position = "top")
```

Students at schools outside the state capitals report slightly *better*
environment-domain quality of life than those in the capitals, and the
private-school advantage is small; the distributions overlap heavily, so any
school-level differences are modest relative to between-student variation.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    hide_n = TRUE
  ) |>
  modify_caption("**Univariable linear regression.**")
uv
```

```{r uv-numbers}
u <- function(v, tm = v) lm(reformulate(v, outcome), data = dat_model) |>
  tidy(conf.int = TRUE) |> filter(term == tm)
uv_dreem <- u("dreem_global")
uv_trait <- u("trait_anxiety")
uv_res   <- u("resilience_score")
uv_sex   <- u("sex", "sexMale")
uv_priv  <- u("school_legal_status", "school_legal_statusPrivate")
uv_loc   <- u("school_location", "school_locationOther city")
uv_clerk <- u("year_group", "year_groupClerkship")
uv_clin  <- u("year_group", "year_groupClinical sciences")
```

The DREEM global score is the strongest crude predictor: each point is associated
with a `r sprintf("%.3f", uv_dreem$estimate)`-point higher environment score
(95% CI `r sprintf("%.3f to %.3f", uv_dreem$conf.low, uv_dreem$conf.high)`,
p < 0.001), so a 20-point better perceived educational environment corresponds to
about `r sprintf("%.1f", uv_dreem$estimate * 20)` points on the 0-100 scale.
Trait anxiety runs the other way (`r sprintf("%.2f", uv_trait$estimate)` per
point, 95% CI
`r sprintf("%.2f to %.2f", uv_trait$conf.low, uv_trait$conf.high)`) and
resilience is positively associated (`r sprintf("%.2f", uv_res$estimate)` per
point). Both school-level variables are significant crudely: students outside a
state capital score `r sprintf("%.2f", uv_loc$estimate)` points higher
(p < 0.001) and private-school students `r sprintf("%.2f", uv_priv$estimate)`
points higher (p = `r sprintf("%.4f", uv_priv$p.value)`). Sex shows no crude
association (p = `r sprintf("%.2f", uv_sex$p.value)`).

## Why adjustment is needed

The crude school-level contrasts are the ones most obviously in need of
adjustment, and there are specific variables here that would confound them.
**The DREEM global score** confounds the school legal status contrast: students
at private schools rate their educational environment far more favourably (mean
DREEM 128 versus 115 at public schools), and DREEM is itself strongly associated
with the environment domain, so a crude "private schools are better" finding may
be nothing more than the DREEM difference. **Trait anxiety** confounds both
psychosocial predictors:
dispositionally anxious students score lower on the resilience scale (*r* =
-0.61), rate their educational environment more harshly (*r* = -0.43) and report
worse material circumstances, so the crude resilience and DREEM coefficients each
contain a share of trait anxiety. **Sex** is a third: women in this sample report
substantially higher trait anxiety than men, which masks any direct sex
difference in the environment domain until anxiety is held fixed. Finally,
**stage of the course** is associated with both the perceived educational
environment and with living circumstances (later-stage students are more likely
to have moved out of the family home), so it must be held fixed before the DREEM
coefficient can be read. Fitting all seven predictors together allows each
estimate to be interpreted with the others held constant.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit) |>
  modify_caption("**Multivariable linear regression.**")
mv
```

```{r fit-stats}
gl <- glance(fit)
tibble(
  Statistic = c("R-squared", "Adjusted R-squared", "Residual SE", "n", "Model df"),
  Value = c(sprintf("%.3f", gl$r.squared), sprintf("%.3f", gl$adj.r.squared),
            sprintf("%.2f", gl$sigma), as.character(gl$nobs), as.character(gl$df))
) |> gt() |> tab_header(title = md("**Model fit**"))
```

```{r mv-numbers}
tid <- tidy(fit, conf.int = TRUE)
g <- function(x) tid |> filter(term == x)
b_dreem <- g("dreem_global"); b_trait <- g("trait_anxiety")
b_res <- g("resilience_score"); b_sex <- g("sexMale")
b_priv <- g("school_legal_statusPrivate"); b_loc <- g("school_locationOther city")
b_clerk <- g("year_groupClerkship"); b_clin <- g("year_groupClinical sciences")
fmt <- function(r) sprintf("%.2f (95%% CI %.2f to %.2f)", r$estimate, r$conf.low, r$conf.high)
fmt3 <- function(r) sprintf("%.3f (95%% CI %.3f to %.3f)", r$estimate, r$conf.low, r$conf.high)
```

The model explains `r sprintf("%.1f%%", gl$r.squared * 100)` of the variance
(adjusted R-squared `r sprintf("%.3f", gl$adj.r.squared)`) - much less than for
the psychological or physical domains, which is expected because material
circumstances depend heavily on family resources that this survey did not
measure. Holding the other predictors fixed, each DREEM point is associated with
`r fmt3(b_dreem)` points of environment-domain score (about
`r sprintf("%.1f", b_dreem$estimate * 20)` points per 20 DREEM points), and each
trait-anxiety point with `r fmt(b_trait)` points. Attending a school outside a
state capital remains independently associated with a `r fmt(b_loc)`-point
higher score. Men score `r fmt(b_sex)` points *lower* than women once anxiety is
held fixed. Clinical-sciences students score `r fmt(b_clin)` and clerkship
students `r fmt(b_clerk)` points higher than basic-sciences students. Private
schooling (`r fmt(b_priv)`) and resilience (`r fmt(b_res)`) show no independent
association. These are conditional associations from a cross-sectional survey.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

The most dramatic movement is **resilience**, which goes from a strongly
significant `r sprintf("%.2f", uv_res$estimate)` crude (p < 0.001) to
`r sprintf("%.2f", b_res$estimate)` adjusted and non-significant
(p = `r sprintf("%.2f", b_res$p.value)`) - it even changes sign. The entire crude
resilience association with material circumstances was trait anxiety: resilient
students are, on this instrument, mostly non-anxious students. **Private
schooling** does the same thing, from `r sprintf("%.2f", uv_priv$estimate)`
points (p = `r sprintf("%.4f", uv_priv$p.value)`) to
`r sprintf("%.2f", b_priv$estimate)` (p = `r sprintf("%.2f", b_priv$p.value)`);
once the DREEM score is held fixed the private-school advantage disappears,
because private-school students rate their educational environment about 13 DREEM
points higher on average and that, rather than the school's legal status, is what
tracked the environment domain. (School location is *not* the explanation here -
private and public schools are almost identically distributed between capitals
and other cities, 43% versus 41% outside a capital.) **Sex** moves in the
opposite direction, a textbook suppression: crude
`r sprintf("%.2f", uv_sex$estimate)` points and clearly non-significant
(p = `r sprintf("%.2f", uv_sex$p.value)`), adjusted
`r sprintf("%.2f", b_sex$estimate)` points and significant
(p = `r sprintf("%.3f", b_sex$p.value)`). Women's higher trait anxiety was
concealing a modest male *disadvantage* on the environment domain.

**School location** barely moves - `r sprintf("%.2f", uv_loc$estimate)` crude to
`r sprintf("%.2f", b_loc$estimate)` adjusted, retaining
`r sprintf("%.0f%%", 100 * b_loc$estimate / uv_loc$estimate)` and remaining
significant - so it is not explained away by any of the individual-level
predictors and is the closest thing in this analysis to a genuine school-level
finding. **The DREEM global score** shrinks moderately
(`r sprintf("%.3f", uv_dreem$estimate)` to
`r sprintf("%.3f", b_dreem$estimate)`,
`r sprintf("%.0f%%", 100 * b_dreem$estimate / uv_dreem$estimate)` retained), and
**stage of the course** strengthens slightly (clerkship
`r sprintf("%.2f", uv_clerk$estimate)` crude, non-significant, to
`r sprintf("%.2f", b_clerk$estimate)` adjusted and significant).

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r vif-table}
performance::check_collinearity(fit) |>
  as_tibble() |>
  transmute(Term, VIF = round(VIF, 2), Tolerance = round(Tolerance, 3)) |>
  gt() |> tab_header(title = md("**Variance inflation factors**"))
```

```{r het-test}
het <- performance::check_heteroscedasticity(fit)
```

**Linearity.** The fitted-versus-residual reference line is essentially flat
across the fitted range, with only slight curvature at the extremes; the linear
form is adequate.

**Homoscedasticity.** Breusch-Pagan rejects constant variance
(p = `r sprintf("%.3f", attr(het, "p"))`). As a sensitivity check we refitted
with heteroscedasticity-consistent (HC3) standard errors:

```{r robust}
rob <- lmtest::coeftest(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
broom::tidy(rob) |>
  transmute(term,
            estimate = round(estimate, 3),
            `robust SE` = round(std.error, 3),
            statistic = round(statistic, 2),
            p.value = format.pval(p.value, digits = 3, eps = 0.001)) |>
  gt() |> tab_header(title = md("**HC3 robust standard errors (sensitivity check)**"))
```

No coefficient changes its significance status under HC3, so the
heteroscedasticity - a floor/ceiling artefact of the bounded 0-100 scale - does
not affect the conclusions.

**Normality of residuals.** The Q-Q plot tracks the diagonal with a short left
tail. The formal test flags non-normality
(p = `r sprintf("%.3f", attr(performance::check_normality(fit), "p"))`), which at
n = `r nrow(dat_model)` reflects the test's power rather than a real problem.

**Influential points.** No observation exceeds the Cook's distance threshold; the
largest is `r sprintf("%.3f", max(cooks.distance(fit)))`.

**Collinearity.** The highest VIF is
`r sprintf("%.2f", max(performance::check_collinearity(fit)$VIF))` (trait
anxiety), with resilience at
`r sprintf("%.2f", performance::check_collinearity(fit) |> as_tibble() |> filter(Term == "resilience_score") |> pull(VIF))`.
All values are below 2 and comfortably under the threshold of 5. The two
school-level variables are only weakly associated with each other in this sample
(both VIFs near 1.1), so their separate contributions can be read with
confidence.

# Discussion

Environment-domain quality of life - money, safety, housing, transport, access to
care - is only weakly explained by anything measured in this survey
(`r sprintf("%.0f%%", gl$r.squared * 100)` of variance). Within that, the two
things that matter independently are how positively the student rates their
educational environment and, unusually for this dataset, a school-level variable:
students at schools outside the state capitals report about
`r sprintf("%.1f", b_loc$estimate)` points better environment-domain quality of
life than those in the capitals, and this is not explained away by school type,
stage, sex, anxiety or resilience. A plausible reading is cost of living and
commuting: capital cities in Brazil are more expensive and have longer, less safe
commutes, and both feed directly into the items this domain asks about.
Resilience, by contrast, has no independent association at all once trait anxiety
is held fixed - resilience training would not be expected to change a student's
material circumstances, and the data agree.

**Limitations.** *Clustering is not modelled, and this matters more here than for
any other outcome in the file.* The strongest finding is a school-level contrast,
but students were sampled from 22 schools and the released file carries no school
identifier - only `school_legal_status` and `school_location`. Two schools in the
same city share cost of living, transport and campus facilities, so the effective
number of independent units behind the location coefficient is closer to 22 than
to `r nrow(dat_model)`. The reported standard error for `school_location` is
therefore substantially too small and its p-value too optimistic; with a random
intercept for school this finding might well not survive. *No socioeconomic
measure.* Family income, which is the obvious driver of every item in the
environment domain, is not in the file, so residual confounding of both
school-level contrasts by family wealth is near-certain. *Cross-sectional
design and common-method bias.* DREEM, resilience, anxiety and WHOQOL were all
self-reported at one sitting, so a generally negative respondent depresses all of
them together. *Low explained variance.* An R-squared of
`r sprintf("%.2f", gl$r.squared)` means the model is describing a small slice of
what determines this outcome, and predictions for individuals would be poor.

**Future work.** Retain the school identifier and add a family-income measure,
then fit a mixed model with a random intercept for school to test whether the
capital-versus-other-city difference is a real contextual effect or an artefact
of a handful of schools.

# Feasibility notes

*Not part of a student report.*

- **Knits cleanly, statistically well behaved.** n = 1,350 complete cases, seven
  predictors, max VIF
  `r sprintf("%.2f", max(performance::check_collinearity(fit)$VIF))`, max Cook's
  distance `r sprintf("%.3f", max(cooks.distance(fit)))`.
- **The least tautological of the three WHOQOL outcomes.** Trait anxiety
  correlates only -0.41 with the environment domain (versus -0.75 with the
  psychological domain), and BDI -0.39. A student can use mood scales as
  predictors here without the analysis collapsing into one questionnaire
  predicting itself. That is a real pedagogical advantage.
- **Lowest R-squared of the six outcomes**
  (`r sprintf("%.2f", gl$r.squared)`). Some students will read this as failure;
  it should be framed as the honest result of asking about material
  circumstances with a questionnaire that measured none of them.
- **Best worked example of confounding in the dataset.** Three distinct patterns
  appear in one table: private schooling and resilience both lose significance,
  sex gains it through suppression, and school location is essentially immovable.
- **The clustering problem bites hardest here.** The headline finding
  (`school_location`) is a school-level contrast estimated as if the 1,350
  students were independent. With only 22 schools and no identifier this cannot
  be fixed, and it must be stated plainly in Limitations, not buried.
- **Guardrails.** `dreem_global` must never appear with its five subscales; no
  other WHOQOL domain as a predictor; only one of BDI / state / trait anxiety.
- **Heteroscedasticity is flagged** (BP p = `r sprintf("%.3f", attr(het, "p"))`),
  as for every outcome in this dataset; HC3 changes nothing.
- **Verdict: workable and pedagogically the most interesting**, provided the
  student is willing to write an honest Limitations section about the 22 schools.
