---
title: "Diastolic blood pressure in adults aged 30 and over in Bharatpur, Nepal"
subtitle: "Dataset 3: Nepal hypertension and cardiovascular risk-factor survey"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)
library(sandwich)
library(lmtest)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "nepal_hypertension.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
#
# Deliberately EXCLUDED from the predictor set:
#   systolic_bp           - the same measurement occasion as the outcome
#                           (r = 0.79); a co-outcome, not a risk factor.
#   hypertension          - a deterministic function of systolic_bp, diastolic_bp
#                           and antihypertensive_drug; circular.
#   bp_awareness,
#   antihypertensive_drug - consequences of a previous high reading, not causes
#                           of today's reading.
#   mets_min_per_week     - physical_activity is a deterministic cut of it at
#                           600 MET-min/week, so only one of the pair may enter.
#   cigarettes_per_week   - zero for every never-smoker, so it is smoking_status
#                           re-expressed on a different scale.
#   marital_status        - only 37 unmarried participants and largely a function
#                           of age and sex.
outcome    <- "diastolic_bp"
predictors <- c("sex", "age_group", "education", "occupation",
                "below_poverty_line", "smoking_status", "smokeless_tobacco",
                "alcohol_last_month", "fruit_veg_servings", "salt_g_per_day",
                "physical_activity", "family_history_htn", "bmi")

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex = factor(sex, levels = c("female", "male"), labels = c("Female", "Male")),
    # Youngest band is the reference so the age contrasts read upwards.
    age_group = factor(age_group,
                       levels = c("30_39", "40_49", "50_59", "60_plus"),
                       labels = c("30-39", "40-49", "50-59", "60 or more")),
    # No or informal schooling is both the largest group and the most deprived.
    education = factor(education,
                       levels = c("none_or_informal", "primary", "secondary", "higher"),
                       labels = c("None or informal", "Primary", "Secondary", "Higher")),
    # Household work is the largest occupational group.
    occupation = factor(occupation,
                        levels = c("household_work", "agriculture_or_labour",
                                   "job_or_self_employed", "unemployed"),
                        labels = c("Household work", "Agriculture or labour",
                                   "Job or self-employed", "Unemployed")),
    # Never-smoker is the clinically unexposed reference.
    smoking_status = factor(smoking_status,
                            levels = c("never", "former", "current"),
                            labels = c("Never", "Former", "Current")),
    # The more active group is the reference, so "Low" reads as the risk category.
    physical_activity = factor(physical_activity,
                               levels = c("moderate_or_high", "low"),
                               labels = c("Moderate or high", "Low")),
    across(c(below_poverty_line, smokeless_tobacco, alcohol_last_month,
             family_history_htn, hypertension, bp_awareness, antihypertensive_drug),
           \(x) factor(x, levels = c(0, 1), labels = c("No", "Yes")))
  ) |>
  labelled::set_variable_labels(
    diastolic_bp          = "Diastolic blood pressure (mmHg)",
    systolic_bp           = "Systolic blood pressure (mmHg)",
    sex                   = "Sex",
    age_group             = "Age band (years)",
    education             = "Education",
    occupation            = "Main occupation",
    below_poverty_line    = "Below poverty line",
    smoking_status        = "Smoking status",
    smokeless_tobacco     = "Smokeless tobacco use",
    alcohol_last_month    = "Alcohol in the last month",
    fruit_veg_servings    = "Fruit and vegetable servings per day",
    salt_g_per_day        = "Salt intake (g/day)",
    physical_activity     = "Physical activity level",
    family_history_htn    = "Family history of hypertension",
    bmi                   = "Body mass index (kg/m2)",
    antihypertensive_drug = "On antihypertensive medication"
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

# Kept alongside dat_model for the treatment-bias sensitivity analysis.
dat_untreated <- dat |>
  filter(antihypertensive_drug == "No") |>
  select(all_of(c(outcome, predictors))) |> drop_na()
```

# Introduction

This is a community-based cross-sectional survey of adults aged 30 years and
over living in Bharatpur, Chitwan district, Nepal, designed to estimate the
prevalence of hypertension and describe its behavioural and anthropometric
correlates (Dhungana and colleagues, *PLOS ONE* 2016, e0185806). Blood pressure
was measured twice with a digital device and averaged, and the WHO STEPS
instrument was used for tobacco, alcohol, diet and physical activity, with
height and weight measured directly. Diastolic pressure is analysed separately
from systolic here because the two behave differently with age -- diastolic
pressure plateaus and then falls in later life as arteries stiffen, whereas
systolic pressure keeps rising -- so the same predictor set can give a
noticeably different picture.

**Aim.** Among adults aged 30 and over in Bharatpur, we examined which of sex,
age band, education, occupation, poverty, smoking, smokeless tobacco, alcohol,
fruit and vegetable intake, salt intake, physical activity, family history of
hypertension and body mass index are independently associated with mean
diastolic blood pressure, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** A cross-sectional survey of `r nrow(dat_raw)` adults aged 30
years and over resident in Bharatpur Municipality, Chitwan, Nepal. The source
study used multistage cluster sampling: four wards were selected at random and
approximately 300 participants enrolled in each, with the sample size
calculation assuming a design effect of 1.5. **The ward identifier is absent
from this extract**, with direct consequences for the standard errors (see
Limitations). Each row is one participant.

**Variables.** The outcome is `diastolic_bp`, the mean of two diastolic readings
in mmHg. Predictors are sex (reference *Female*); age band in years (reference
*30-39*; age was recorded only in four bands, so a per-year effect cannot be
estimated); education (reference *None or informal*); main occupation (reference
*Household work*); family income per head below the poverty line (reference
*No*); smoking status (reference *Never*); current smokeless tobacco use
(reference *No*); any alcohol in the last month (reference *No*); fruit and
vegetable servings per day; estimated salt intake in g/day; physical activity
level (reference *Moderate or high*); family history of hypertension (reference
*No*); and body mass index in kg/m2.

Systolic pressure is deliberately **not** a predictor. It is measured on the
same occasion, on the same arm, from the same physiology, and correlates with
diastolic at r = `r round(cor(dat$systolic_bp, dat$diastolic_bp), 2)`; including
it would produce a large, meaningless coefficient and drive every other estimate
to zero. For the same reason `hypertension`, `bp_awareness` and
`antihypertensive_drug` are excluded: `hypertension` is defined as raised
measured pressure *or* current treatment, so it is a function of the outcome,
and the other two are consequences of having been found hypertensive in the
past. Two further pairs measure the same construct twice and only one member of
each may enter: `physical_activity` is a deterministic cut of
`mets_min_per_week` at 600 MET-minutes per week (the "Low" group spans 0-600,
"Moderate or high" spans 620-29,260) and the categorical version is used because
it is the WHO STEPS classification reported in the source paper; and
`cigarettes_per_week` is zero for every never-smoker, so `smoking_status` is
used instead because it distinguishes former from never smokers.

**Statistical analysis.** Linear regression via `lm()`, justified because the
outcome is continuous. We fitted a univariable model for each predictor first,
then a single multivariable model containing all of them. Predictors were chosen
in advance on subject-matter grounds, not by significance. Significance
threshold 0.05; estimates are reported with 95% confidence intervals.
Assumptions were checked with `performance::check_model()`, and because
non-constant error variance was flagged, HC3 heteroscedasticity-consistent
standard errors are reported alongside the ordinary ones. Analysis used
`r R.version.string` with tidyverse, gtsummary, performance and sandwich.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped). Every variable in this file is 100%
complete, so listwise deletion removes
`r if (n_dropped == 0) "no rows at all" else paste(n_dropped, "rows")`. There is
no missing-data story here: no imputation is required and complete-case
selection bias cannot arise.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(
    missing = "no"
  ) |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

```{r table-one-numbers}
dbp_mean <- mean(dat_model$diastolic_bp)
dbp_sd   <- sd(dat_model$diastolic_bp)
pct_female <- round(100 * mean(dat_model$sex == "Female"), 1)
pct_low_pa <- round(100 * mean(dat_model$physical_activity == "Low"), 1)
n_treated  <- sum(dat$antihypertensive_drug == "Yes")
dbp_treated   <- mean(dat$diastolic_bp[dat$antihypertensive_drug == "Yes"])
dbp_untreated <- mean(dat$diastolic_bp[dat$antihypertensive_drug == "No"])
```

Mean diastolic pressure was `r round(dbp_mean, 1)` mmHg (SD
`r round(dbp_sd, 1)`), a much narrower distribution than the systolic one. The
sample is `r pct_female`% female, `r pct_low_pa`% are classified as physically
inactive, and `r n_treated` participants
(`r round(100 * n_treated / nrow(dat), 1)`%) were on antihypertensive medication
when measured. Their mean diastolic pressure was `r round(dbp_treated, 1)` mmHg
against `r round(dbp_untreated, 1)` mmHg in the untreated -- a gap of
`r round(dbp_treated - dbp_untreated, 1)` mmHg that exists *despite* treatment
and therefore understates the underlying difference.

## Exploratory figure

```{r explore, fig.cap="Figure 2. Diastolic blood pressure by age band, separately for participants taking and not taking antihypertensive medication. The dashed line marks the 90 mmHg threshold."}
dat |>
  ggplot(aes(age_group, diastolic_bp, fill = antihypertensive_drug)) +
  geom_boxplot(outlier.alpha = 0.3, position = position_dodge(width = 0.8)) +
  geom_hline(yintercept = 90, linetype = "dashed") +
  scale_fill_viridis_d(option = "D", end = 0.75, alpha = 0.8) +
  labs(x = "Age band (years)", y = "Diastolic blood pressure (mmHg)",
       fill = "On antihypertensive\nmedication",
       caption = "Dashed line: 90 mmHg diastolic threshold")
```

Diastolic pressure rises across the age bands but far less steeply than systolic
does -- mean values run
`r paste(round(tapply(dat$diastolic_bp, dat$age_group, mean), 1), collapse = ", ")`
mmHg against roughly 115 to 134 mmHg for systolic -- and the rise is mostly
between the two youngest bands. At every age the treated participants sit
higher than the untreated (means around 87-88 mmHg regardless of age), and the
treated boxes are wider, which is what one expects when a heterogeneous group of
people on different drugs at different doses is compared with an untreated
group; note also that only `r sum(dat$antihypertensive_drug == "Yes" & dat$age_group == "30-39")`
of the 30-39 band are treated, so that box rests on very few people.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv
```

```{r uv-numbers}
uv_fit <- function(v) lm(reformulate(v, outcome), data = dat_model) |> tidy(conf.int = TRUE)
u_bmi   <- uv_fit("bmi") |> filter(term == "bmi")
u_age60 <- uv_fit("age_group") |> filter(term == "age_group60 or more")
u_sex   <- uv_fit("sex") |> filter(term == "sexMale")
u_alc   <- uv_fit("alcohol_last_month") |> filter(term == "alcohol_last_monthYes")
u_form  <- uv_fit("smoking_status") |> filter(term == "smoking_statusFormer")
u_cur   <- uv_fit("smoking_status") |> filter(term == "smoking_statusCurrent")
u_smok  <- uv_fit("smokeless_tobacco") |> filter(term == "smokeless_tobaccoYes")
u_pa    <- uv_fit("physical_activity") |> filter(term == "physical_activityLow")
u_salt  <- uv_fit("salt_g_per_day") |> filter(term == "salt_g_per_day")
u_job   <- uv_fit("occupation") |> filter(term == "occupationJob or self-employed")
u_fam   <- uv_fit("family_history_htn") |> filter(term == "family_history_htnYes")
```

Body mass index is again the clearest continuous predictor: each additional
kg/m2 is associated with a `r round(u_bmi$estimate, 2)` mmHg higher diastolic
pressure (95% CI `r round(u_bmi$conf.low, 2)` to `r round(u_bmi$conf.high, 2)`,
p < 0.001), which across the interquartile range of BMI in this sample amounts
to roughly `r round(u_bmi$estimate * 5.4, 1)` mmHg. The age gradient is much
shallower than for systolic pressure -- participants aged 60 or more average
`r round(u_age60$estimate, 1)` mmHg above the 30-39 band (95% CI
`r round(u_age60$conf.low, 1)` to `r round(u_age60$conf.high, 1)`), against
about 21 mmHg for systolic -- which is the expected physiology. Men average
`r round(u_sex$estimate, 1)` mmHg above women (95% CI
`r round(u_sex$conf.low, 1)` to `r round(u_sex$conf.high, 1)`), drinkers
`r round(u_alc$estimate, 1)` mmHg above non-drinkers, former smokers
`r round(u_form$estimate, 1)` mmHg above never-smokers, and smokeless tobacco
users `r round(u_smok$estimate, 1)` mmHg above non-users. Salt intake is once
more flat (`r round(u_salt$estimate, 2)` mmHg per g/day, 95% CI
`r round(u_salt$conf.low, 2)` to `r round(u_salt$conf.high, 2)`).

## Why adjustment is needed

Almost every crude estimate above is contaminated by two variables in this
dataset. **Age band** is the first. It is associated with diastolic pressure and
simultaneously with practically every exposure here: former smokers are
concentrated in the older bands (quitting is an older person's behaviour), 209
of the 263 participants aged 60 or more have no or informal schooling against
114 of 383 among the 30-39 year olds, and occupation shifts from paid employment
towards household work and unemployment with age. A crude coefficient for
education or occupation therefore contains an age effect. **Sex** is the second
and is even starker here, because tobacco and alcohol in this population are
overwhelmingly male behaviours:
`r round(100 * mean(dat_model$smoking_status[dat_model$sex == "Male"] == "Current"), 0)`%
of men are current smokers against
`r round(100 * mean(dat_model$smoking_status[dat_model$sex == "Female"] == "Current"), 0)`%
of women, and men have higher diastolic pressure regardless. The crude smokeless
tobacco and alcohol coefficients are therefore partly measuring maleness.
**Body mass index** is the third: it is the strongest single predictor of
diastolic pressure in the file and it varies systematically across occupation,
poverty and smoking groups (smokers are leaner), so omitting it lets those
coefficients absorb an adiposity effect -- sometimes in the wrong direction.
Mutual adjustment is what separates these three overlapping gradients from the
behaviours of interest.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv
```

```{r fit-stats}
gl <- glance(fit)
tibble(
  Statistic = c("R-squared", "Adjusted R-squared", "Residual SD (mmHg)",
                "Observations", "Model parameters (incl. intercept)",
                "Observations per parameter"),
  Value = c(round(gl$r.squared, 3), round(gl$adj.r.squared, 3),
            round(sigma(fit), 2), nobs(fit), length(coef(fit)),
            round(nobs(fit) / length(coef(fit)), 1))
) |>
  gt() |>
  tab_header(title = "Model fit")
```

```{r mv-numbers}
tid <- tidy(fit, conf.int = TRUE)
g <- function(t) tid |> filter(term == t)
m_bmi   <- g("bmi")
m_age60 <- g("age_group60 or more")
m_sex   <- g("sexMale")
m_alc   <- g("alcohol_last_monthYes")
m_form  <- g("smoking_statusFormer")
m_cur   <- g("smoking_statusCurrent")
m_smok  <- g("smokeless_tobaccoYes")
m_pa    <- g("physical_activityLow")
m_salt  <- g("salt_g_per_day")
m_job   <- g("occupationJob or self-employed")
m_fam   <- g("family_history_htnYes")
```

The model explains `r round(100 * gl$r.squared, 1)`% of the variance in
diastolic pressure (adjusted R-squared `r round(gl$adj.r.squared, 3)`), a little
better than the corresponding systolic model but still leaving the great
majority unexplained. With `r nobs(fit)` observations and
`r length(coef(fit))` parameters there are
`r round(nobs(fit) / length(coef(fit)), 1)` observations per parameter, so
overfitting is not a concern. Conditional on the other twelve predictors, each
additional kg/m2 of BMI is associated with a `r round(m_bmi$estimate, 2)` mmHg
higher diastolic pressure (95% CI `r round(m_bmi$conf.low, 2)` to
`r round(m_bmi$conf.high, 2)`); participants aged 60 or more average
`r round(m_age60$estimate, 1)` mmHg above the 30-39 band (95% CI
`r round(m_age60$conf.low, 1)` to `r round(m_age60$conf.high, 1)`); men average
`r round(m_sex$estimate, 1)` mmHg above women (95% CI
`r round(m_sex$conf.low, 1)` to `r round(m_sex$conf.high, 1)`, p =
`r round(m_sex$p.value, 3)`); and recent drinkers average
`r round(m_alc$estimate, 1)` mmHg above non-drinkers (95% CI
`r round(m_alc$conf.low, 1)` to `r round(m_alc$conf.high, 1)`).

As in the systolic model, the smoking coefficients are the wrong way round for a
causal reading: former smokers are `r round(m_form$estimate, 1)` mmHg above
never-smokers (95% CI `r round(m_form$conf.low, 1)` to
`r round(m_form$conf.high, 1)`) while current smokers are indistinguishable from
never-smokers (`r round(m_cur$estimate, 1)` mmHg, 95% CI
`r round(m_cur$conf.low, 1)` to `r round(m_cur$conf.high, 1)`). The natural
explanation is reverse causation: in a cross-sectional survey "former smoker"
partly identifies people who quit *after* being told they had a cardiovascular
problem. Salt intake remains null (`r round(m_salt$estimate, 2)` mmHg per g/day,
95% CI `r round(m_salt$conf.low, 2)` to `r round(m_salt$conf.high, 2)`), as does
family history (`r round(m_fam$estimate, 1)` mmHg).

### Sensitivity analysis: excluding treated participants

```{r sensitivity}
fit_untr <- lm(reformulate(predictors, outcome), data = dat_untreated)
tid_u <- tidy(fit_untr, conf.int = TRUE)
gu <- function(t) tid_u |> filter(term == t)
s_bmi   <- gu("bmi")
s_age60 <- gu("age_group60 or more")
s_sex   <- gu("sexMale")
s_alc   <- gu("alcohol_last_monthYes")

tibble(
  Term = c("BMI (per kg/m2)", "Age 60+ vs 30-39", "Male vs female",
           "Alcohol last month"),
  `All participants` = sprintf("%.2f (%.2f to %.2f)",
                               c(m_bmi$estimate, m_age60$estimate, m_sex$estimate, m_alc$estimate),
                               c(m_bmi$conf.low, m_age60$conf.low, m_sex$conf.low, m_alc$conf.low),
                               c(m_bmi$conf.high, m_age60$conf.high, m_sex$conf.high, m_alc$conf.high)),
  `Untreated only` = sprintf("%.2f (%.2f to %.2f)",
                             c(s_bmi$estimate, s_age60$estimate, s_sex$estimate, s_alc$estimate),
                             c(s_bmi$conf.low, s_age60$conf.low, s_sex$conf.low, s_alc$conf.low),
                             c(s_bmi$conf.high, s_age60$conf.high, s_sex$conf.high, s_alc$conf.high))
) |>
  gt() |>
  tab_header(title = "Adjusted estimates (mmHg, 95% CI)",
             subtitle = sprintf("All n = %d; untreated n = %d", nobs(fit), nobs(fit_untr)))
```

`r n_treated` participants were taking medication whose entire purpose is to
lower the quantity being measured, so `diastolic_bp` in this file is a *partly
treated* outcome. Refitting in the `r nobs(fit_untr)` untreated participants
moves the estimates very little: BMI goes from `r round(m_bmi$estimate, 2)` to
`r round(s_bmi$estimate, 2)` mmHg per kg/m2 and the oldest-band contrast from
`r round(m_age60$estimate, 1)` to `r round(s_age60$estimate, 1)` mmHg. The
stability is reassuring but should not be over-read: treatment is concentrated
in the older and heavier participants and pulls their readings down, so the
full-sample coefficients for age and BMI are attenuated rather than inflated.
Restricting to the untreated is also not a clean fix, because being untreated is
itself a consequence of blood pressure, so this restriction trades treatment
bias for selection bias. Both fits are shown for that reason.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

Several estimates move a long way, all toward the null. **Alcohol in the last
month** falls from `r round(u_alc$estimate, 1)` mmHg crude (95% CI
`r round(u_alc$conf.low, 1)` to `r round(u_alc$conf.high, 1)`, p < 0.001) to
`r round(m_alc$estimate, 1)` mmHg adjusted (95% CI
`r round(m_alc$conf.low, 1)` to `r round(m_alc$conf.high, 1)`, p =
`r round(m_alc$p.value, 3)`), a `r round(100 * (1 - m_alc$estimate / u_alc$estimate))`%
reduction: roughly half the crude alcohol association is being male and being
older. **Smokeless tobacco** goes from `r round(u_smok$estimate, 1)` mmHg crude
(p `r format.pval(u_smok$p.value, digits = 2, eps = 0.001)`) to
`r round(m_smok$estimate, 1)` mmHg adjusted (p =
`r round(m_smok$p.value, 2)`) and loses significance entirely -- the same
confounding by sex and age. **Low physical activity** goes from
`r round(u_pa$estimate, 1)` mmHg (p =
`r round(u_pa$p.value, 3)`) to `r round(m_pa$estimate, 2)` mmHg (p =
`r round(m_pa$p.value, 2)`), also losing significance, because inactivity in
this sample is largely a marker of being old and heavy. **Being in paid
employment** rather than doing household work moves from
`r round(u_job$estimate, 1)` mmHg (p = `r round(u_job$p.value, 3)`) to
`r round(m_job$estimate, 1)` mmHg (p = `r round(m_job$p.value, 2)`) -- a
sex effect in disguise, since paid employment here is disproportionately male.
**Sex** itself roughly halves, from `r round(u_sex$estimate, 1)` to
`r round(m_sex$estimate, 1)` mmHg, and only barely stays significant.

Two estimates barely move, and one of them moves *away* from the null.
**Age 60 or more** goes from `r round(u_age60$estimate, 1)` to
`r round(m_age60$estimate, 1)` mmHg, a shift of under 15%, as expected for a
variable that nothing else in the model can confound. **BMI** goes from
`r round(u_bmi$estimate, 2)` to `r round(m_bmi$estimate, 2)` mmHg per kg/m2 --
slightly *larger* after adjustment, which happens because BMI is negatively
associated with smoking and with the older age bands, so the crude estimate was
being pulled down by those. This is negative confounding rather than mediation
or collinearity; the maximum variance inflation factor in the model is
`r round(max(check_collinearity(fit)$VIF), 2)`, so no coefficient is unstable.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r het}
het <- check_heteroscedasticity(fit)
het_p <- as.numeric(het)
rob <- coeftest(fit, vcov = vcovHC(fit, type = "HC3"))
rob_ci <- coefci(fit, vcov. = vcovHC(fit, type = "HC3"))
keep <- c("bmi", "sexMale", "age_group60 or more", "alcohol_last_monthYes",
          "smoking_statusFormer", "salt_g_per_day")
tibble(
  Term = keep,
  Estimate = round(rob[keep, "Estimate"], 3),
  `SE (OLS)` = round(summary(fit)$coefficients[keep, "Std. Error"], 3),
  `SE (HC3)` = round(rob[keep, "Std. Error"], 3),
  `95% CI (HC3)` = sprintf("%.2f to %.2f", rob_ci[keep, 1], rob_ci[keep, 2]),
  `p (HC3)` = format.pval(rob[keep, "Pr(>|t|)"], digits = 2, eps = 0.001)
) |>
  gt() |>
  tab_header(title = "HC3 heteroscedasticity-consistent standard errors",
             subtitle = "Compared with ordinary least-squares standard errors")
```

The **posterior-predictive check** shows the model reproducing the bulk of the
diastolic distribution but missing the digit-preference spikes at 70, 80 and 90
mmHg that are visible in the raw data: readings were recorded to the nearest
whole number by a human operator and cluster on multiples of ten, which no
continuous model can reproduce. The **linearity** panel is close to flat. The
**homogeneity of variance** panel shows a mild fan, and
`check_heteroscedasticity()` flags it at p
`r format.pval(het_p, digits = 2, eps = 0.001)` -- borderline rather than gross,
but flagged. As heteroscedasticity biases standard errors and not coefficients,
HC3 robust standard errors are reported above; they are almost identical to the
OLS ones (BMI `r round(summary(fit)$coefficients["bmi", "Std. Error"], 3)`
versus `r round(rob["bmi", "Std. Error"], 3)`), so no substantive conclusion
changes, but the robust version is the one to quote.
**Collinearity** is not an issue: maximum VIF
`r round(max(check_collinearity(fit)$VIF), 2)`, which is only achievable because
`systolic_bp`, `mets_min_per_week` and `cigarettes_per_week` were kept out.
**Influential observations** shows no point beyond the Cook's distance contour.
**Normality of residuals** shows mild departure at both tails, partly from the
digit preference; with `r nobs(fit)` observations this does not threaten the
coefficient estimates.

# Discussion

Diastolic blood pressure in these Bharatpur adults is driven by body mass index
and, more weakly than systolic pressure, by age. Each kg/m2 of BMI corresponds
to about `r round(m_bmi$estimate, 2)` mmHg, so the difference between a BMI of
22 and 28 is roughly `r round(m_bmi$estimate * 6, 1)` mmHg; being 60 or over
rather than in one's thirties adds about `r round(m_age60$estimate, 0)` mmHg.
Alcohol retains a modest independent association of
`r round(m_alc$estimate, 1)` mmHg. Everything else -- salt, physical activity,
fruit and vegetables, family history, poverty, education, smokeless tobacco --
is null once age, sex and BMI are accounted for. Practically, this is another
argument that weight is the single actionable target in this population, and a
warning that the crude lifestyle associations that make attractive headlines
here are mostly demographic confounding.

**Limitations.** The most serious problem is **treatment bias in the outcome**.
`r n_treated` participants (`r round(100 * n_treated / nrow(dat), 1)`%) were on
antihypertensive medication when measured, so what this file records is
diastolic pressure *as currently treated*, and because treatment is concentrated
in those with the highest underlying pressure, coefficients for any risk factor
that also predicts treatment are attenuated by an unknown amount. The
sensitivity analysis bounds this rather than removing it. Second, **the cluster
sampling cannot be modelled**: only four wards were randomly selected and about
300 participants enrolled in each, with a design effect of 1.5 assumed in the
sample size calculation, but no ward identifier exists in this extract. The
confidence intervals reported above are therefore too narrow -- plausibly by
around 20% -- and no mixed or survey-weighted model can repair that. Four
clusters also make any claim of representativeness of Nepal, or even of
Bharatpur as a whole, unsupportable. Third, **age is available only in four
bands**, so no per-year slope is estimable and the top band is open-ended.
Fourth, this is cross-sectional, so associations are not causal effects, and the
former-smoker coefficient is a live demonstration of reverse causation. Fifth,
diastolic readings show clear terminal-digit preference at multiples of ten,
which adds measurement noise and slightly distorts the residual distribution.
Sixth, salt intake is estimated rather than measured by 24-hour urinary sodium,
so its null association may be measurement error rather than biology. Seventh,
adjusted R-squared is only `r round(gl$adj.r.squared, 3)`. On the credit side,
**the file has no missing data at all**: all `r nrow(dat_raw)` rows are complete
on all `r ncol(dat_raw)` variables.

**Future work.** Obtain the ward identifier and refit with a random intercept
for ward; and repeat the analysis on a bias-corrected outcome that adds a fixed
5-10 mmHg diastolic offset to treated participants.

# Feasibility notes

*Not part of a student report.*

- **Works, with the same large caveat as systolic.** Continuous outcome, 100%
  complete data, `r nobs(fit)` rows over `r length(coef(fit))` parameters, max
  VIF `r round(max(check_collinearity(fit)$VIF), 2)`, adjusted R-squared
  `r round(gl$adj.r.squared, 3)`, and genuine confounding that visibly moves the
  alcohol, smokeless tobacco, occupation and physical activity coefficients.
- **This file is very close to `systolic_bp.Rmd`.** If both are offered to
  students, expect near-identical reports. The one genuinely different finding
  is the much shallower age gradient (`r round(m_age60$estimate, 1)` mmHg here
  versus about 16 mmHg for systolic), which is real physiology and worth
  making the students explain. Otherwise consider offering only one of the two.
- **The treatment trap applies identically.** Treated mean diastolic
  `r round(dbp_treated, 1)` versus `r round(dbp_untreated, 1)` mmHg untreated.
  A student who adjusts for `antihypertensive_drug` is adjusting for a
  consequence of the outcome.
- **Forbidden predictors.** `systolic_bp` (r =
  `r round(cor(dat$systolic_bp, dat$diastolic_bp), 2)` with the outcome),
  `hypertension` (deterministic function of the outcome), `bp_awareness` and
  `antihypertensive_drug`.
- **Deterministic twins.** `physical_activity` is exactly
  `mets_min_per_week <= 600`; `cigarettes_per_week` is zero for all
  never-smokers. Never use both members of either pair.
- **Heteroscedasticity is flagged but only just** (p
  `r format.pval(het_p, digits = 2, eps = 0.001)`), which makes this a slightly
  awkward teaching example: a student re-running it after any change to the
  predictor set may see it pass. HC3 standard errors change nothing here.
- **Digit preference.** Diastolic values pile up on 70/80/90. Harmless for
  regression but it makes the QQ and posterior-predictive panels look worse than
  the model deserves, and students will over-interpret that.
- **No ward identifier**, so the assumed design effect of 1.5 cannot be
  honoured; standard errors are too narrow throughout.
- **Age in bands only**, so no "mmHg per year" figure is available.
