---
title: "Hypertension in adults aged 30 and over in Bharatpur, Nepal"
subtitle: "Dataset 3: Nepal hypertension and cardiovascular risk-factor survey"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "nepal_hypertension.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
#
# Deliberately EXCLUDED from the predictor set, and this is the central design
# decision in this file:
#   systolic_bp, diastolic_bp,
#   antihypertensive_drug - `hypertension` is DEFINED as systolic >= 140 or
#                           diastolic >= 90 or currently taking antihypertensive
#                           medication. All three are components of the outcome's
#                           own definition, so using any of them as a predictor is
#                           tautology, and antihypertensive_drug produces perfect
#                           separation (every treated person is hypertensive by
#                           definition).
#   bp_awareness          - having been told you have raised blood pressure is a
#                           consequence of having it. In this file all 241 aware
#                           participants are hypertensive and none of the 918
#                           unaware non-hypertensive, i.e. awareness perfectly
#                           predicts the outcome in one direction; it is also
#                           quasi-separating. Excluded.
#   mets_min_per_week     - physical_activity is a deterministic cut of it at
#                           600 MET-min/week, so only one of the pair may enter.
#   cigarettes_per_week   - zero for every never-smoker, so it is smoking_status
#                           re-expressed on a different scale.
#   marital_status        - only 37 unmarried participants and largely a function
#                           of age and sex.
outcome    <- "hypertension"
predictors <- c("sex", "age_group", "education", "occupation",
                "below_poverty_line", "smoking_status", "smokeless_tobacco",
                "alcohol_last_month", "fruit_veg_servings", "salt_g_per_day",
                "physical_activity", "family_history_htn", "bmi")

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex = factor(sex, levels = c("female", "male"), labels = c("Female", "Male")),
    # Youngest band is the reference so the age contrasts read upwards.
    age_group = factor(age_group,
                       levels = c("30_39", "40_49", "50_59", "60_plus"),
                       labels = c("30-39", "40-49", "50-59", "60 or more")),
    # No or informal schooling is both the largest group and the most deprived.
    education = factor(education,
                       levels = c("none_or_informal", "primary", "secondary", "higher"),
                       labels = c("None or informal", "Primary", "Secondary", "Higher")),
    # Household work is the largest occupational group.
    occupation = factor(occupation,
                        levels = c("household_work", "agriculture_or_labour",
                                   "job_or_self_employed", "unemployed"),
                        labels = c("Household work", "Agriculture or labour",
                                   "Job or self-employed", "Unemployed")),
    # Never-smoker is the clinically unexposed reference.
    smoking_status = factor(smoking_status,
                            levels = c("never", "former", "current"),
                            labels = c("Never", "Former", "Current")),
    # The more active group is the reference, so "Low" reads as the risk category.
    physical_activity = factor(physical_activity,
                               levels = c("moderate_or_high", "low"),
                               labels = c("Moderate or high", "Low")),
    # "Yes" is the second level throughout, so glm() models the odds of "Yes".
    across(c(below_poverty_line, smokeless_tobacco, alcohol_last_month,
             family_history_htn, hypertension, bp_awareness, antihypertensive_drug),
           \(x) factor(x, levels = c(0, 1), labels = c("No", "Yes")))
  ) |>
  labelled::set_variable_labels(
    hypertension          = "Hypertension",
    sex                   = "Sex",
    age_group             = "Age band (years)",
    education             = "Education",
    occupation            = "Main occupation",
    below_poverty_line    = "Below poverty line",
    smoking_status        = "Smoking status",
    smokeless_tobacco     = "Smokeless tobacco use",
    alcohol_last_month    = "Alcohol in the last month",
    fruit_veg_servings    = "Fruit and vegetable servings per day",
    salt_g_per_day        = "Salt intake (g/day)",
    physical_activity     = "Physical activity level",
    family_history_htn    = "Family history of hypertension",
    bmi                   = "Body mass index (kg/m2)"
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

n_event <- sum(dat_model$hypertension == "Yes")
```

# Introduction

This is a community-based cross-sectional survey of adults aged 30 years and
over living in Bharatpur, Chitwan district, Nepal, carried out to estimate the
prevalence of hypertension and identify its risk factors (Dhungana and
colleagues, *PLOS ONE* 2016, e0185806). Blood pressure was measured twice with a
digital device and averaged, the WHO STEPS instrument supplied the behavioural
information, and height and weight were measured directly. Hypertension is the
outcome the survey was designed around, and at
`r round(100 * n_event / nrow(dat_model), 1)`% prevalence it is a substantial
burden for a population that is largely undiagnosed and untreated.

**Aim.** Among adults aged 30 and over in Bharatpur, we examined which of sex,
age band, education, occupation, poverty, smoking, smokeless tobacco, alcohol,
fruit and vegetable intake, salt intake, physical activity, family history of
hypertension and body mass index are independently associated with the odds of
being hypertensive, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** A cross-sectional survey of `r nrow(dat_raw)` adults aged 30
years and over resident in Bharatpur Municipality, Chitwan, Nepal. The source
study used multistage cluster sampling: four wards were selected at random and
approximately 300 participants enrolled in each, with the sample size
calculation assuming a design effect of 1.5. **The ward identifier is not in
this extract** (see Limitations). Each row is one participant.

**Variables.** The outcome is `hypertension`, coded 1 (the event) if mean
systolic pressure was at least 140 mmHg **or** mean diastolic pressure at least
90 mmHg **or** the participant reported taking antihypertensive medication in
the previous two weeks. There are `r n_event` events among `r nrow(dat_model)`
participants (`r round(100 * n_event / nrow(dat_model), 1)`%). Predictors are
sex (reference *Female*); age band in years (reference *30-39*; age was recorded
only in four bands, so no odds ratio per year of age can be estimated);
education (reference *None or informal*); main occupation (reference *Household
work*); family income per head below the poverty line (reference *No*); smoking
status (reference *Never*); current smokeless tobacco use (reference *No*); any
alcohol in the last month (reference *No*); fruit and vegetable servings per
day; estimated salt intake in g/day; physical activity level (reference
*Moderate or high*); family history of hypertension (reference *No*); and body
mass index in kg/m2.

**The exclusions matter more here than the inclusions.** Because `hypertension`
is *defined* from `systolic_bp`, `diastolic_bp` and `antihypertensive_drug`,
none of those three can be a predictor: they are components of the outcome, not
risk factors for it. `antihypertensive_drug` is the extreme case -- every one of
the `r sum(dat$antihypertensive_drug == "Yes")` treated participants is
hypertensive by definition, so it separates the outcome perfectly and the model
will not converge to a finite coefficient. `bp_awareness` is only slightly less
circular: having been told you have raised blood pressure is a consequence of
having it, and in this file every one of the
`r sum(dat$bp_awareness == "Yes")` participants who are aware is hypertensive,
while `r sum(dat$bp_awareness == "No" & dat$hypertension == "Yes")` hypertensive
participants are unaware. Awareness therefore predicts the outcome perfectly in
one direction, which is complete separation (see Diagnostics for the fitted
coefficient). It is excluded on causal grounds first, but it would also break
the model. Two further pairs measure
the same construct twice and only one of each may enter: `physical_activity` is
a deterministic cut of `mets_min_per_week` at 600 MET-minutes per week (the
"Low" group spans 0-600 and "Moderate or high" spans 620-29,260), and the
categorical version is used because it is the WHO STEPS classification the
source paper reports; and `cigarettes_per_week` is zero for every never-smoker,
so `smoking_status` is used instead because it also separates former from never
smokers.

**Statistical analysis.** Logistic regression via `glm(family = binomial)`,
justified because the outcome is binary; the event modelled is being
hypertensive. We fitted a univariable model for each predictor first, then a
single multivariable model containing all of them. Predictors were chosen in
advance on subject-matter grounds, not by significance. Estimates are reported
as odds ratios with 95% confidence intervals; significance threshold 0.05. Model
adequacy was assessed with AIC, Tjur's R-squared, events per parameter and
`performance::check_model()`. Analysis used `r R.version.string` with tidyverse,
gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped). Every variable in this file is 100%
complete, so listwise deletion removes
`r if (n_dropped == 0) "no rows at all" else paste(n_dropped, "rows")`. There is
no missing-data story: no imputation, no complete-case selection bias, and the
full event count is available to the model.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(
    by = hypertension,
    missing = "no"
  ) |>
  add_n() |>
  add_overall() |>
  modify_caption("**Table 1. Sample characteristics by hypertension status.**")
```

```{r table-one-numbers}
prev <- round(100 * n_event / nrow(dat_model), 1)
prev_60 <- round(100 * mean(dat_model$hypertension[dat_model$age_group == "60 or more"] == "Yes"), 1)
prev_30 <- round(100 * mean(dat_model$hypertension[dat_model$age_group == "30-39"] == "Yes"), 1)
bmi_htn <- mean(dat_model$bmi[dat_model$hypertension == "Yes"])
bmi_no  <- mean(dat_model$bmi[dat_model$hypertension == "No"])
n_med_only <- sum(dat$antihypertensive_drug == "Yes" & dat$systolic_bp < 140 & dat$diastolic_bp < 90)
```

Overall prevalence was `r prev`%, rising from `r prev_30`% in the 30-39 band to
`r prev_60`% among those aged 60 or more. Hypertensive participants averaged
`r round(bmi_htn, 1)` kg/m2 against `r round(bmi_no, 1)` kg/m2 in the
normotensive, and were more often male, less educated and more often former
smokers. Note one consequence of the composite definition: `r n_med_only`
participants are counted as hypertensive purely because they take medication,
despite having entirely normal measured readings on the day. They are correctly
classified -- treated hypertension is still hypertension -- but they are the
reason the outcome cannot be reconstructed from blood pressure alone.

## Exploratory figure

```{r explore, fig.cap="Figure 4. Prevalence of hypertension by age band and sex, with 95% Wilson confidence intervals."}
dat_model |>
  count(age_group, sex, hypertension) |>
  group_by(age_group, sex) |>
  summarise(events = sum(n[hypertension == "Yes"]), total = sum(n),
            .groups = "drop") |>
  mutate(p = events / total,
         se = sqrt(p * (1 - p) / total),
         lo = pmax(0, p - 1.96 * se), hi = pmin(1, p + 1.96 * se)) |>
  ggplot(aes(age_group, p, colour = sex, group = sex)) +
  geom_line(linewidth = 1, position = position_dodge(width = 0.25)) +
  geom_point(size = 2.5, position = position_dodge(width = 0.25)) +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.15,
                position = position_dodge(width = 0.25)) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_colour_viridis_d(option = "D", end = 0.75) +
  labs(x = "Age band (years)", y = "Prevalence of hypertension", colour = "Sex")
```

```{r explore-numbers}
pr <- round(100 * tapply(dat_model$hypertension == "Yes",
                         list(dat_model$age_group, dat_model$sex), mean), 1)
```

Prevalence climbs steeply across the four age bands in both sexes, roughly
tripling in women (`r pr["30-39", "Female"]`% to `r pr["60 or more", "Female"]`%)
and rising less dramatically in men (`r pr["30-39", "Male"]`% to
`r pr["60 or more", "Male"]`%). Men are above women in every band, but the gap
is large only in the 30-39 group -- where it also rests on just
`r sum(dat_model$age_group == "30-39" & dat_model$sex == "Male")` men -- and has
essentially closed by age 60, so age is by far the dominant gradient and the sex
effect is concentrated at younger ages.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = glm,
    method.args = list(family = binomial),
    exponentiate = TRUE
  )
uv
```

```{r uv-numbers}
uv_fit <- function(v) {
  glm(reformulate(v, outcome), data = dat_model, family = binomial) |>
    tidy(conf.int = TRUE, exponentiate = TRUE)
}
u_age60 <- uv_fit("age_group") |> filter(term == "age_group60 or more")
u_bmi   <- uv_fit("bmi") |> filter(term == "bmi")
u_sex   <- uv_fit("sex") |> filter(term == "sexMale")
u_form  <- uv_fit("smoking_status") |> filter(term == "smoking_statusFormer")
u_cur   <- uv_fit("smoking_status") |> filter(term == "smoking_statusCurrent")
u_alc   <- uv_fit("alcohol_last_month") |> filter(term == "alcohol_last_monthYes")
u_smok  <- uv_fit("smokeless_tobacco") |> filter(term == "smokeless_tobaccoYes")
u_pa    <- uv_fit("physical_activity") |> filter(term == "physical_activityLow")
u_fam   <- uv_fit("family_history_htn") |> filter(term == "family_history_htnYes")
u_pov   <- uv_fit("below_poverty_line") |> filter(term == "below_poverty_lineYes")
u_sec   <- uv_fit("education") |> filter(term == "educationSecondary")
u_unemp <- uv_fit("occupation") |> filter(term == "occupationUnemployed")
u_fv    <- uv_fit("fruit_veg_servings") |> filter(term == "fruit_veg_servings")
u_salt  <- uv_fit("salt_g_per_day") |> filter(term == "salt_g_per_day")
```

Age is the dominant crude predictor: participants aged 60 or more have
`r round(u_age60$estimate, 2)` times the odds of hypertension of those aged
30-39 (95% CI `r round(u_age60$conf.low, 2)` to `r round(u_age60$conf.high, 2)`,
p < 0.001). Body mass index is next: each additional kg/m2 multiplies the odds
by `r round(u_bmi$estimate, 3)` (95% CI `r round(u_bmi$conf.low, 3)` to
`r round(u_bmi$conf.high, 3)`), which compounds to
`r round(u_bmi$estimate^5, 2)`-fold across a 5 kg/m2 difference. Former smokers
have `r round(u_form$estimate, 2)` times the odds of never-smokers (95% CI
`r round(u_form$conf.low, 2)` to `r round(u_form$conf.high, 2)`) and recent
drinkers `r round(u_alc$estimate, 2)` times the odds of non-drinkers (95% CI
`r round(u_alc$conf.low, 2)` to `r round(u_alc$conf.high, 2)`). Men have
`r round(u_sex$estimate, 2)` times the odds of women. Almost everything else is
also significant crudely -- smokeless tobacco
(`r round(u_smok$estimate, 2)`), low physical activity
(`r round(u_pa$estimate, 2)`), poverty (`r round(u_pov$estimate, 2)`), family
history (`r round(u_fam$estimate, 2)`), unemployment
(`r round(u_unemp$estimate, 2)`) and secondary education
(`r round(u_sec$estimate, 2)`, protective) -- which is precisely the situation
in which crude estimates should not be trusted. The only clear crude null is
salt intake (`r round(u_salt$estimate, 3)` per g/day, 95% CI
`r round(u_salt$conf.low, 3)` to `r round(u_salt$conf.high, 3)`).

## Why adjustment is needed

The crude odds ratios above are shot through with confounding by two variables
in this dataset. **Age band** is the first and largest. It has an enormous
independent effect on hypertension (crude odds ratio
`r round(u_age60$estimate, 2)` for the oldest band) and it is associated with
nearly every candidate exposure: former smokers are concentrated in the older
bands because quitting takes years, education is far lower in older cohorts (209
of the 263 participants aged 60 or more have no or informal schooling against
114 of 383 in the 30-39 band), unemployment rises with age, and physical
activity falls. So the crude "protective effect of secondary education" and the
crude "harm of inactivity" are both partly age. **Sex** is the second, because
in this population tobacco and alcohol are overwhelmingly male:
`r round(100 * mean(dat_model$smoking_status[dat_model$sex == "Male"] == "Current"), 0)`%
of men are current smokers against
`r round(100 * mean(dat_model$smoking_status[dat_model$sex == "Female"] == "Current"), 0)`%
of women, and
`r round(100 * mean(dat_model$alcohol_last_month[dat_model$sex == "Male"] == "Yes"), 0)`%
of men drank last month against
`r round(100 * mean(dat_model$alcohol_last_month[dat_model$sex == "Female"] == "Yes"), 0)`%
of women, while men have higher blood pressure for reasons independent of both.
Crude tobacco and alcohol odds ratios therefore partly measure being male.
**Body mass index** is the third: it is a strong independent risk factor and it
is lower in smokers and in poorer participants, so leaving it out lets the
tobacco and poverty coefficients absorb an adiposity effect pointing the other
way. Mutual adjustment is the only way to separate them.

## Multivariable model

```{r multivariable}
fit <- glm(reformulate(predictors, outcome), data = dat_model, family = binomial)

mv <- tbl_regression(fit, exponentiate = TRUE)
mv
```

```{r fit-stats}
n_par <- length(coef(fit)) - 1  # excluding the intercept
epp   <- n_event / n_par
null_fit <- glm(reformulate("1", outcome), data = dat_model, family = binomial)

tibble(
  Statistic = c("AIC (full model)", "AIC (intercept only)",
                "Tjur's R-squared", "Events (hypertensive)",
                "Non-events", "Model parameters (excluding intercept)",
                "Events per parameter"),
  Value = c(round(AIC(fit), 1), round(AIC(null_fit), 1),
            round(as.numeric(r2_tjur(fit)), 3), n_event,
            nrow(dat_model) - n_event, n_par, round(epp, 1))
) |>
  gt() |>
  tab_header(title = "Model fit")
```

```{r mv-numbers}
tid <- tidy(fit, conf.int = TRUE, exponentiate = TRUE)
g <- function(t) tid |> filter(term == t)
m_age60 <- g("age_group60 or more")
m_bmi   <- g("bmi")
m_sex   <- g("sexMale")
m_form  <- g("smoking_statusFormer")
m_cur   <- g("smoking_statusCurrent")
m_alc   <- g("alcohol_last_monthYes")
m_smok  <- g("smokeless_tobaccoYes")
m_pa    <- g("physical_activityLow")
m_fam   <- g("family_history_htnYes")
m_pov   <- g("below_poverty_lineYes")
m_sec   <- g("educationSecondary")
m_unemp <- g("occupationUnemployed")
m_fv    <- g("fruit_veg_servings")
m_salt  <- g("salt_g_per_day")
```

The full model has AIC `r round(AIC(fit), 1)` against `r round(AIC(null_fit), 1)`
for an intercept-only model, a drop of `r round(AIC(null_fit) - AIC(fit), 1)`,
so the predictor set carries real information; Tjur's R-squared is
`r round(as.numeric(r2_tjur(fit)), 3)`, meaning the average predicted probability
differs by about `r round(100 * as.numeric(r2_tjur(fit)), 0)` percentage points
between hypertensive and normotensive participants. With `r n_event` events and
`r n_par` non-intercept parameters there are `r round(epp, 1)` **events per
parameter**, comfortably above the conventional minimum of 10, so the model is
not overfitted and the coefficients are not the product of sparse cells.

Conditional on the other twelve predictors, the odds of hypertension are
`r round(m_age60$estimate, 2)` times higher in those aged 60 or more than in the
30-39 band (95% CI `r round(m_age60$conf.low, 2)` to
`r round(m_age60$conf.high, 2)`); each additional kg/m2 of body mass index
multiplies the odds by `r round(m_bmi$estimate, 3)` (95% CI
`r round(m_bmi$conf.low, 3)` to `r round(m_bmi$conf.high, 3)`), that is
`r round(m_bmi$estimate^5, 2)`-fold across 5 kg/m2; recent drinkers have
`r round(m_alc$estimate, 2)` times the odds of non-drinkers (95% CI
`r round(m_alc$conf.low, 2)` to `r round(m_alc$conf.high, 2)`); and participants
with secondary education have `r round(m_sec$estimate, 2)` times the odds of
those with no or informal schooling (95% CI `r round(m_sec$conf.low, 2)` to
`r round(m_sec$conf.high, 2)`).

Two results need caution. Former smokers retain `r round(m_form$estimate, 2)`
times the odds of never-smokers (95% CI `r round(m_form$conf.low, 2)` to
`r round(m_form$conf.high, 2)`) while current smokers are not distinguishable
from never-smokers (`r round(m_cur$estimate, 2)`, 95% CI
`r round(m_cur$conf.low, 2)` to `r round(m_cur$conf.high, 2)`). Read causally
this would say quitting is dangerous; read as reverse causation it says people
quit after being told they have high blood pressure, which is almost certainly
what is happening in a cross-sectional survey where
`r sum(dat$bp_awareness == "Yes")` participants already knew their diagnosis.
Second, **fruit and vegetable servings** come out *positively* associated with
hypertension (`r round(m_fv$estimate, 3)` per serving, 95% CI
`r round(m_fv$conf.low, 3)` to `r round(m_fv$conf.high, 3)`, p =
`r round(m_fv$p.value, 3)`) -- the wrong direction for any dietary story, and
most plausibly either recall bias among people who know they are hypertensive
and have been advised to eat more vegetables, or simple confounding by an
unmeasured aspect of diet. It should not be reported as a finding.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted odds ratios.**")
```

Adjustment changes the picture substantially, and almost all of the movement is
toward the null. **Smokeless tobacco** collapses from
`r round(u_smok$estimate, 2)` crude (95% CI `r round(u_smok$conf.low, 2)` to
`r round(u_smok$conf.high, 2)`, p = `r round(u_smok$p.value, 3)`) to
`r round(m_smok$estimate, 2)` adjusted (95% CI `r round(m_smok$conf.low, 2)` to
`r round(m_smok$conf.high, 2)`, p = `r round(m_smok$p.value, 2)`) -- essentially
the entire crude association was age and sex. **Low physical activity** falls
from `r round(u_pa$estimate, 2)` (p = `r round(u_pa$p.value, 4)`) to
`r round(m_pa$estimate, 2)` (p = `r round(m_pa$p.value, 2)`) and loses
significance, for the same reason. **Family history of hypertension** goes from
`r round(u_fam$estimate, 2)` (p = `r round(u_fam$p.value, 3)`) to
`r round(m_fam$estimate, 2)` (p = `r round(m_fam$p.value, 2)`), which is
disappointing for a variable with real biological content but reflects the fact
that people reporting a family history are also older and heavier. **Current
smoking** moves from `r round(u_cur$estimate, 2)` (p < 0.001) to
`r round(m_cur$estimate, 2)` (p = `r round(m_cur$p.value, 2)`), **poverty** from
`r round(u_pov$estimate, 2)` to `r round(m_pov$estimate, 2)`, and
**unemployment** from `r round(u_unemp$estimate, 2)` (p =
`r round(u_unemp$p.value, 3)`) to `r round(m_unemp$estimate, 2)` (p =
`r round(m_unemp$p.value, 2)`); all three lose significance. Even the two
strongest predictors attenuate: **age 60 or more** from
`r round(u_age60$estimate, 2)` to `r round(m_age60$estimate, 2)`, and **former
smoking** from `r round(u_form$estimate, 2)` to `r round(m_form$estimate, 2)`.
**Sex** moves from `r round(u_sex$estimate, 2)` to `r round(m_sex$estimate, 2)`
and crosses from significant to marginal (p = `r round(m_sex$p.value, 3)`).

Two coefficients move the other way, *away* from the null. **Body mass index**
goes from `r round(u_bmi$estimate, 3)` crude to `r round(m_bmi$estimate, 3)`
adjusted per kg/m2, and **fruit and vegetable servings** from
`r round(u_fv$estimate, 3)` to `r round(m_fv$estimate, 3)`, becoming *more*
significant after adjustment (p = `r round(u_fv$p.value, 3)` to p =
`r round(m_fv$p.value, 3)`). Both are negative confounding: BMI and fruit and
vegetable intake are both lower in smokers and in the poorest participants, who
are themselves at higher risk, so the crude estimates were pulled down. None of
this is collinearity -- the maximum variance inflation factor is
`r round(max(check_collinearity(fit)$VIF), 2)`. The overall message of Table 2
is that in a crude analysis of this dataset almost every risk factor looks
significant, and after mutual adjustment only age, BMI, alcohol, former smoking
and education survive.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r diag-numbers}
vifs <- check_collinearity(fit)
max_vif <- round(max(vifs$VIF), 2)
infl <- sum(cooks.distance(fit) > 4 / nobs(fit))
sep_fit <- glm(hypertension ~ bp_awareness, data = dat, family = binomial)
sep_coef <- coef(sep_fit)[2]
sep_se <- sqrt(diag(vcov(sep_fit)))[2]
```

For a logistic model the panels that matter are different from the linear case;
homogeneity of variance and normality of residuals do not apply and can be
ignored here. The **binned residuals** panel is the key one: the binned average
residuals fall almost entirely inside the error bounds across the range of
fitted probabilities, with only one or two bins straying, so there is no
systematic region of the probability scale where the model is badly calibrated
and no strong evidence that a continuous predictor needs a non-linear term. The
**posterior-predictive check** shows the predicted distribution of the binary
outcome matching the observed proportions. **Collinearity** is unproblematic:
the maximum VIF is `r max_vif`, which is only achievable because the definitional
components (`systolic_bp`, `diastolic_bp`, `antihypertensive_drug`) and the
deterministic twins (`mets_min_per_week`, `cigarettes_per_week`) were kept out;
including `antihypertensive_drug` instead produces complete separation rather
than merely a high VIF. **Influential observations** shows
`r infl` points above the conventional 4/n Cook's distance cut-off out of
`r nobs(fit)`, none of them extreme, which is unremarkable at this sample size.
**Events per parameter** is `r round(epp, 1)` (`r n_event` events,
`r n_par` parameters), well above 10, so the estimates are not sparse-data
biased and the Wald confidence intervals can be trusted.

As a demonstration of what the excluded variables would have done: regressing
hypertension on `bp_awareness` alone gives a log-odds coefficient of
`r round(sep_coef, 1)` with a standard error of `r round(sep_se, 0)`, that is an
odds ratio of the order of 10^8 with an infinitely wide confidence interval --
the classic signature of complete separation, and a reminder that this variable
is a consequence of the outcome, not a predictor of it.

# Discussion

Roughly two in five adults aged 30 and over in Bharatpur are hypertensive, and
after mutual adjustment the drivers are age, body mass index and alcohol, with a
protective association for education. The odds of hypertension are
`r round(m_age60$estimate, 1)` times higher in those aged 60 or more than in
those in their thirties, and each 5 kg/m2 of body mass index multiplies the odds
by about `r round(m_bmi$estimate^5, 2)`. Recent alcohol use nearly doubles the
odds (`r round(m_alc$estimate, 2)`). What does *not* survive adjustment is
equally important for programme design: smokeless tobacco, low physical
activity, poverty, unemployment and even family history all look like risk
factors crudely and are null once age, sex and BMI are held fixed. Practically,
this argues for age- and weight-based screening and for alcohol as the one
behavioural target with an independent signal here, rather than a broad
lifestyle campaign.

**Limitations.** First, and specific to this outcome, **the composite definition
makes several apparently obvious predictors unusable**. `hypertension` is raised
measured pressure *or* current treatment, so `systolic_bp`, `diastolic_bp` and
`antihypertensive_drug` are components of it and cannot appear on the right-hand
side; `bp_awareness` is a consequence of it and separates the outcome perfectly.
A student who includes any of them gets an impressive-looking model that means
nothing. Second, the composite also mixes two rather different states:
`r n_med_only` participants are classified hypertensive on the basis of
medication alone despite normal readings on the day, so the outcome is partly
"has been diagnosed and treated" and partly "has a high reading now" -- and
access to diagnosis and treatment is itself socially patterned, which will bias
the education and poverty coefficients in an unknown direction. Third, **the
cluster sampling cannot be modelled**: only four wards were randomly selected
with roughly 300 participants each, and the sample size calculation assumed a
design effect of 1.5, but no ward identifier exists in this extract. All
confidence intervals reported here are therefore too narrow, plausibly by around
20%, and no mixed-effects or survey-design correction can repair that; nor can
results from four clusters be generalised to Nepal or even to Bharatpur as a
whole. Fourth, **age is available only in four bands**, so no odds ratio per year
of age is estimable and the oldest band is open-ended. Fifth, the design is
cross-sectional, so nothing here is a causal effect: the former-smoker result and
the positive fruit-and-vegetable coefficient are both best explained by reverse
causation or by behaviour change after diagnosis, and the same mechanism could be
operating more subtly for salt and physical activity. Sixth, hypertension was
defined on a single visit with two readings, whereas clinical diagnosis requires
readings on separate occasions, so some participants are misclassified by
white-coat effect. On the credit side, **the file has no missing data at all**
(all `r nrow(dat_raw)` rows complete on all `r ncol(dat_raw)` variables) and the
event count of `r n_event` is generous relative to the number of parameters.

**Future work.** Refit with a ward random effect if the identifier can be
obtained; and split the outcome into "raised measured pressure" and "treated"
to see whether the socio-economic coefficients differ between the two, which
would test the access-to-treatment explanation directly.

# Feasibility notes

*Not part of a student report.*

- **The strongest of the four outcomes in this dataset.** `r n_event` events in
  `r nrow(dat_model)` participants, `r round(epp, 1)` events per parameter, no
  missing data, maximum VIF `r max_vif`, binned residuals acceptable, AIC drop of
  `r round(AIC(null_fit) - AIC(fit), 1)` against the null model, and adjustment
  that genuinely reorders the findings.
- **The definitional trap is severe and will catch students.** `hypertension` =
  raised measured BP **or** on treatment. `systolic_bp`, `diastolic_bp` and
  `antihypertensive_drug` are therefore components of the outcome.
  `hypertension ~ systolic_bp` alone gives an odds ratio of about 1.21 per mmHg
  with z = 16; `hypertension ~ antihypertensive_drug` and
  `hypertension ~ bp_awareness` give complete separation (fitted log-odds
  `r round(sep_coef, 1)`, standard error `r round(sep_se, 0)`). The brief must
  name these four variables explicitly as forbidden, because every one of them
  produces a "great" model.
- **`bp_awareness` deserves its own warning.** It is not literally part of the
  definition, so a student may argue for it, but all
  `r sum(dat$bp_awareness == "Yes")` aware participants are hypertensive and the
  model does not converge to a finite estimate. Exclude it.
- **Deterministic twins.** `physical_activity` is exactly
  `mets_min_per_week <= 600`; `cigarettes_per_week` is zero for all
  never-smokers. Never use both members of either pair.
- **Two counter-intuitive survivors.** Former smoking is strongly and
  independently associated with hypertension while current smoking is not
  (reverse causation after diagnosis), and fruit and vegetable intake is
  *positively* associated and becomes more significant after adjustment. Both
  are excellent teaching material but will be reported as findings by an
  uncritical student.
- **Massive crude-to-adjusted attenuation.** Smokeless tobacco
  (`r round(u_smok$estimate, 2)` to `r round(m_smok$estimate, 2)`), physical
  activity (`r round(u_pa$estimate, 2)` to `r round(m_pa$estimate, 2)`), family
  history (`r round(u_fam$estimate, 2)` to `r round(m_fam$estimate, 2)`) and
  unemployment (`r round(u_unemp$estimate, 2)` to
  `r round(m_unemp$estimate, 2)`) all lose significance. This is the best
  demonstration of confounding in the whole dataset.
- **No ward identifier**, so the assumed design effect of 1.5 cannot be honoured
  and all standard errors are too narrow.
- **Age in bands only**, so no odds ratio per year of age is available.
- **`r n_med_only` participants are hypertensive on medication alone** with
  normal readings. Worth pointing out to students who try to "check" the outcome
  against the blood pressure columns and find it does not match.
