---
title: "Systolic blood pressure in adults aged 30 and over in Bharatpur, Nepal"
subtitle: "Dataset 3: Nepal hypertension and cardiovascular risk-factor survey"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)
library(sandwich)
library(lmtest)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "nepal_hypertension.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
#
# Deliberately EXCLUDED from the predictor set:
#   diastolic_bp          - the same measurement occasion as the outcome (r = 0.79),
#                           not a risk factor for it.
#   hypertension          - a deterministic function of systolic_bp, diastolic_bp
#                           and antihypertensive_drug; circular.
#   bp_awareness,
#   antihypertensive_drug - both are consequences of having a high blood pressure
#                           reading in the past, not causes of today's reading.
#   mets_min_per_week     - physical_activity is a deterministic cut of it at
#                           600 MET-min/week, so only one of the pair may enter.
#   cigarettes_per_week   - zero for every never-smoker and essentially zero for
#                           former smokers, so it is smoking_status re-expressed.
#   marital_status        - only 37 unmarried participants, and the categories are
#                           largely a function of age and sex.
outcome    <- "systolic_bp"
predictors <- c("sex", "age_group", "education", "occupation",
                "below_poverty_line", "smoking_status", "smokeless_tobacco",
                "alcohol_last_month", "fruit_veg_servings", "salt_g_per_day",
                "physical_activity", "family_history_htn", "bmi")

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex = factor(sex, levels = c("female", "male"), labels = c("Female", "Male")),
    # Youngest band is the reference so the age contrasts read upwards.
    age_group = factor(age_group,
                       levels = c("30_39", "40_49", "50_59", "60_plus"),
                       labels = c("30-39", "40-49", "50-59", "60 or more")),
    # No or informal schooling is both the largest group and the most deprived.
    education = factor(education,
                       levels = c("none_or_informal", "primary", "secondary", "higher"),
                       labels = c("None or informal", "Primary", "Secondary", "Higher")),
    # Household work is the largest occupational group.
    occupation = factor(occupation,
                        levels = c("household_work", "agriculture_or_labour",
                                   "job_or_self_employed", "unemployed"),
                        labels = c("Household work", "Agriculture or labour",
                                   "Job or self-employed", "Unemployed")),
    # Never-smoker is the clinically unexposed reference.
    smoking_status = factor(smoking_status,
                            levels = c("never", "former", "current"),
                            labels = c("Never", "Former", "Current")),
    # The more active group is the reference, so "Low" reads as the risk category.
    physical_activity = factor(physical_activity,
                               levels = c("moderate_or_high", "low"),
                               labels = c("Moderate or high", "Low")),
    across(c(below_poverty_line, smokeless_tobacco, alcohol_last_month,
             family_history_htn, hypertension, bp_awareness, antihypertensive_drug),
           \(x) factor(x, levels = c(0, 1), labels = c("No", "Yes")))
  ) |>
  labelled::set_variable_labels(
    systolic_bp           = "Systolic blood pressure (mmHg)",
    diastolic_bp          = "Diastolic blood pressure (mmHg)",
    sex                   = "Sex",
    age_group             = "Age band (years)",
    education             = "Education",
    occupation            = "Main occupation",
    below_poverty_line    = "Below poverty line",
    smoking_status        = "Smoking status",
    smokeless_tobacco     = "Smokeless tobacco use",
    alcohol_last_month    = "Alcohol in the last month",
    fruit_veg_servings    = "Fruit and vegetable servings per day",
    salt_g_per_day        = "Salt intake (g/day)",
    physical_activity     = "Physical activity level",
    family_history_htn    = "Family history of hypertension",
    bmi                   = "Body mass index (kg/m2)",
    antihypertensive_drug = "On antihypertensive medication"
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

# Kept alongside dat_model for the treatment-bias sensitivity analysis.
dat_untreated <- dat |>
  filter(antihypertensive_drug == "No") |>
  select(all_of(c(outcome, predictors))) |> drop_na()
```

# Introduction

This is a community-based cross-sectional survey of adults aged 30 years and
over living in Bharatpur, Chitwan district, Nepal, carried out to estimate the
prevalence of hypertension and its behavioural and anthropometric correlates
(Dhungana and colleagues, *PLOS ONE* 2016, e0185806). Blood pressure was
measured twice on each participant with a digital sphygmomanometer and the two
readings averaged; alongside this, the WHO STEPS instrument was used to collect
tobacco, alcohol, diet and physical-activity information, and height and weight
were measured directly. Systolic blood pressure is analysed here on its
continuous scale because dichotomising it at 140 mmHg discards most of the
variation and, in this dataset, produces an outcome that is partly defined by
medication use.

**Aim.** Among adults aged 30 and over in Bharatpur, we examined which of sex,
age band, education, occupation, poverty, smoking, smokeless tobacco, alcohol,
fruit and vegetable intake, salt intake, physical activity, family history of
hypertension and body mass index are independently associated with mean systolic
blood pressure, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** A cross-sectional survey of `r nrow(dat_raw)` adults aged 30
years and over resident in Bharatpur Municipality, Chitwan, Nepal. The source
study used multistage cluster sampling: four wards were selected at random and
approximately 300 participants were enrolled in each, with the sample size
calculation assuming a design effect of 1.5. **The ward identifier is not
present in this extract**, which has direct consequences for the standard errors
reported below (see Limitations). Each row is one participant.

**Variables.** The outcome is `systolic_bp`, the mean of two systolic readings
in mmHg. Predictors are sex (reference *Female*); age band in years (reference
*30-39*; the survey recorded age only in four bands, so no per-year effect can
be estimated); education (reference *None or informal*); main occupation
(reference *Household work*); family income per head below the poverty line
(reference *No*); smoking status (reference *Never*); current smokeless tobacco
use (reference *No*); any alcohol in the last month (reference *No*); fruit and
vegetable servings per day (continuous); estimated salt intake in g/day
(continuous); physical activity level (reference *Moderate or high*); family
history of hypertension (reference *No*); and body mass index in kg/m2
(continuous). These thirteen cover the classical modifiable and non-modifiable
determinants of blood pressure that the STEPS instrument was designed to
capture.

Three pairs of variables in this file measure the same thing twice and only one
member of each pair may enter a model. `physical_activity` is a deterministic
cut of `mets_min_per_week` at 600 MET-minutes per week (the "Low" group spans
0-600, the "Moderate or high" group 620-29,260); the categorical version is
used here because it is the WHO STEPS classification the source paper reports
and because the continuous version is extremely right-skewed. `smoking_status`
and `cigarettes_per_week` are the same construct, with `cigarettes_per_week`
equal to zero for every never-smoker; `smoking_status` is used because it
separates former from never smokers, which turns out to matter. Finally
`hypertension` is *defined* as raised measured blood pressure or current
antihypertensive treatment, so it, `bp_awareness` and `antihypertensive_drug`
are all downstream of the outcome and are excluded as predictors.

**Statistical analysis.** Linear regression via `lm()`, justified because the
outcome is continuous. We fitted a univariable model for each predictor first,
then a single multivariable model containing all of them. Predictors were chosen
in advance on subject-matter grounds, not by significance. Significance threshold
0.05; estimates are reported with 95% confidence intervals. Assumptions were
checked with `performance::check_model()`, and because non-constant error
variance was flagged, heteroscedasticity-consistent (HC3) standard errors are
reported alongside the ordinary ones. Analysis used `r R.version.string` with
tidyverse, gtsummary, performance and sandwich.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped). Every variable in this file is 100%
complete, so listwise deletion removes
`r if (n_dropped == 0) "no rows at all" else paste(n_dropped, "rows")` and there
is no missing-data story to tell: no imputation, no complete-case selection
bias, nothing to sensitivity-test.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(
    missing = "no"
  ) |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

```{r table-one-numbers}
sbp_mean <- mean(dat_model$systolic_bp)
sbp_sd   <- sd(dat_model$systolic_bp)
pct_female <- round(100 * mean(dat_model$sex == "Female"), 1)
pct_60     <- round(100 * mean(dat_model$age_group == "60 or more"), 1)
n_treated  <- sum(dat$antihypertensive_drug == "Yes")
sbp_treated   <- mean(dat$systolic_bp[dat$antihypertensive_drug == "Yes"])
sbp_untreated <- mean(dat$systolic_bp[dat$antihypertensive_drug == "No"])
```

Mean systolic blood pressure was `r round(sbp_mean, 1)` mmHg (SD
`r round(sbp_sd, 1)`), with a long right tail reaching 200 mmHg. The sample is
`r pct_female`% female and `r pct_60`% aged 60 or over, reflecting the daytime
household recruitment used in the source survey; men are therefore
under-represented relative to the municipal population. `r n_treated`
participants (`r round(100 * n_treated / nrow(dat), 1)`%) were taking
antihypertensive medication at the time of measurement, and their mean systolic
pressure was `r round(sbp_treated, 1)` mmHg against
`r round(sbp_untreated, 1)` mmHg in the untreated -- still higher, but
pharmacologically lowered from whatever their untreated value would have been.
This is the central interpretive problem for this outcome and is returned to
throughout.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Systolic blood pressure against body mass index, separately for participants taking and not taking antihypertensive medication, with linear smoothers. The dashed line marks the 140 mmHg threshold."}
dat |>
  ggplot(aes(bmi, systolic_bp, colour = antihypertensive_drug)) +
  geom_point(alpha = 0.25, size = 1) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  geom_hline(yintercept = 140, linetype = "dashed") +
  scale_colour_viridis_d(option = "D", end = 0.75) +
  labs(x = "Body mass index (kg/m2)", y = "Systolic blood pressure (mmHg)",
       colour = "On antihypertensive\nmedication",
       caption = "Dashed line: 140 mmHg systolic threshold")
```

```{r explore-numbers}
sl_untr <- coef(lm(systolic_bp ~ bmi, data = filter(dat, antihypertensive_drug == "No")))["bmi"]
sl_tr   <- coef(lm(systolic_bp ~ bmi, data = filter(dat, antihypertensive_drug == "Yes")))["bmi"]
```

Systolic pressure rises steadily with body mass index in the untreated
(`r round(sl_untr, 2)` mmHg per kg/m2) but is almost flat in the treated
(`r round(sl_tr, 2)` mmHg per kg/m2), and the treated line sits well above the
untreated one across the whole BMI range. Both features are treatment artefacts:
the vertical gap is a *treated* gap and so understates the true difference
between people who need medication and people who do not, while the flat treated
slope is what happens when the highest readings in a group are the ones most
aggressively medicated.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv
```

```{r uv-numbers}
uv_fit <- function(v) lm(reformulate(v, outcome), data = dat_model) |> tidy(conf.int = TRUE)
u_bmi   <- uv_fit("bmi") |> filter(term == "bmi")
u_age60 <- uv_fit("age_group") |> filter(term == "age_group60 or more")
u_sex   <- uv_fit("sex") |> filter(term == "sexMale")
u_alc   <- uv_fit("alcohol_last_month") |> filter(term == "alcohol_last_monthYes")
u_form  <- uv_fit("smoking_status") |> filter(term == "smoking_statusFormer")
u_salt  <- uv_fit("salt_g_per_day") |> filter(term == "salt_g_per_day")
u_fam   <- uv_fit("family_history_htn") |> filter(term == "family_history_htnYes")
u_pa    <- uv_fit("physical_activity") |> filter(term == "physical_activityLow")
```

Body mass index is the strongest continuous predictor: each additional kg/m2 is
associated with a `r round(u_bmi$estimate, 2)` mmHg higher systolic pressure
(95% CI `r round(u_bmi$conf.low, 2)` to `r round(u_bmi$conf.high, 2)`,
p < 0.001), so across the interquartile range of BMI in this sample
(21.0 to 26.4 kg/m2) the crude difference is about
`r round(u_bmi$estimate * 5.4, 1)` mmHg. Age is larger still in absolute terms:
participants aged 60 or more average `r round(u_age60$estimate, 1)` mmHg above
those aged 30-39 (95% CI `r round(u_age60$conf.low, 1)` to
`r round(u_age60$conf.high, 1)`). Men average `r round(u_sex$estimate, 1)` mmHg
above women (95% CI `r round(u_sex$conf.low, 1)` to
`r round(u_sex$conf.high, 1)`). Several of the classical lifestyle risk factors
are already flat in the crude models: salt intake is
`r round(u_salt$estimate, 2)` mmHg per g/day (95% CI
`r round(u_salt$conf.low, 2)` to `r round(u_salt$conf.high, 2)`) and family
history of hypertension is `r round(u_fam$estimate, 1)` mmHg (95% CI
`r round(u_fam$conf.low, 1)` to `r round(u_fam$conf.high, 1)`), neither close to
significant. Low physical activity, by contrast, does look harmful crudely
(`r round(u_pa$estimate, 1)` mmHg, 95% CI `r round(u_pa$conf.low, 1)` to
`r round(u_pa$conf.high, 1)`, p `r format.pval(u_pa$p.value, digits = 2, eps = 0.001)`),
which is worth holding onto because it does not survive adjustment.

## Why adjustment is needed

The crude estimates above mix several strongly overlapping gradients, and two
variables in this dataset do most of the confounding. **Age band** is the
clearest: systolic pressure rises about `r round(u_age60$estimate, 0)` mmHg
between the youngest and oldest bands, and age is simultaneously associated with
almost every candidate exposure here -- former smokers are concentrated in the
older bands, education is far lower in the older bands (209 of the 263
participants aged 60 or more have no or informal schooling, against 114 of 383
in the 30-39 band), and BMI itself falls in the oldest band. Any crude
"education effect" on blood pressure is therefore partly an age effect wearing a
different label. **Sex** is the second: `r round(100 * mean(dat_model$smoking_status[dat_model$sex == "Male"] == "Current"), 0)`%
of men are current smokers against
`r round(100 * mean(dat_model$smoking_status[dat_model$sex == "Female"] == "Current"), 0)`%
of women, alcohol use is similarly concentrated in men, and men have higher
systolic pressure for reasons unrelated to either behaviour; so crude alcohol
and tobacco coefficients partly measure being male. **Body mass index** plays
the same role for the behavioural variables: it is a strong independent
predictor of blood pressure and it differs across occupation, poverty and
activity groups, so leaving it out would let a socio-economic coefficient absorb
an adiposity effect. Mutual adjustment separates these.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv
```

```{r fit-stats}
gl <- glance(fit)
tibble(
  Statistic = c("R-squared", "Adjusted R-squared", "Residual SD (mmHg)",
                "Observations", "Model parameters (incl. intercept)",
                "Observations per parameter"),
  Value = c(round(gl$r.squared, 3), round(gl$adj.r.squared, 3),
            round(sigma(fit), 2), nobs(fit), length(coef(fit)),
            round(nobs(fit) / length(coef(fit)), 1))
) |>
  gt() |>
  tab_header(title = "Model fit")
```

```{r mv-numbers}
tid <- tidy(fit, conf.int = TRUE)
g <- function(t) tid |> filter(term == t)
m_bmi   <- g("bmi")
m_age60 <- g("age_group60 or more")
m_sex   <- g("sexMale")
m_alc   <- g("alcohol_last_monthYes")
m_form  <- g("smoking_statusFormer")
m_cur   <- g("smoking_statusCurrent")
m_salt  <- g("salt_g_per_day")
m_fam   <- g("family_history_htnYes")
m_pa    <- g("physical_activityLow")
```

The model explains `r round(100 * gl$r.squared, 1)`% of the variance in systolic
pressure (adjusted R-squared `r round(gl$adj.r.squared, 3)`), leaving roughly
three quarters unexplained -- normal for individual-level blood pressure, which
is measured with substantial within-person variability. With `r nobs(fit)`
observations and `r length(coef(fit))` parameters there are
`r round(nobs(fit) / length(coef(fit)), 1)` observations per parameter, so the
model is not overfitted. Conditional on the other twelve predictors, each
additional kg/m2 of BMI is associated with a `r round(m_bmi$estimate, 2)` mmHg
higher systolic pressure (95% CI `r round(m_bmi$conf.low, 2)` to
`r round(m_bmi$conf.high, 2)`); participants aged 60 or more average
`r round(m_age60$estimate, 1)` mmHg above the 30-39 band (95% CI
`r round(m_age60$conf.low, 1)` to `r round(m_age60$conf.high, 1)`); men average
`r round(m_sex$estimate, 1)` mmHg above women (95% CI
`r round(m_sex$conf.low, 1)` to `r round(m_sex$conf.high, 1)`); and participants
who drank in the last month average `r round(m_alc$estimate, 1)` mmHg above
those who did not (95% CI `r round(m_alc$conf.low, 1)` to
`r round(m_alc$conf.high, 1)`).

The smoking result needs care. Former smokers average
`r round(m_form$estimate, 1)` mmHg above never-smokers (95% CI
`r round(m_form$conf.low, 1)` to `r round(m_form$conf.high, 1)`) while current
smokers are indistinguishable from never-smokers
(`r round(m_cur$estimate, 1)` mmHg, 95% CI `r round(m_cur$conf.low, 1)` to
`r round(m_cur$conf.high, 1)`). Read causally this is nonsense. Read as reverse
causation it is entirely expected: people quit smoking *because* they were told
they had high blood pressure or another cardiovascular problem, so "former
smoker" is partly a marker of having been diagnosed. Salt intake
(`r round(m_salt$estimate, 2)` mmHg per g/day, 95% CI
`r round(m_salt$conf.low, 2)` to `r round(m_salt$conf.high, 2)`), low physical
activity and family history of hypertension all remain null after adjustment.

### Sensitivity analysis: excluding treated participants

```{r sensitivity}
fit_untr <- lm(reformulate(predictors, outcome), data = dat_untreated)
tid_u <- tidy(fit_untr, conf.int = TRUE)
gu <- function(t) tid_u |> filter(term == t)
s_bmi   <- gu("bmi")
s_age60 <- gu("age_group60 or more")
s_sex   <- gu("sexMale")
s_alc   <- gu("alcohol_last_monthYes")

tibble(
  Term = c("BMI (per kg/m2)", "Age 60+ vs 30-39", "Male vs female",
           "Alcohol last month"),
  `All participants` = sprintf("%.2f (%.2f to %.2f)",
                               c(m_bmi$estimate, m_age60$estimate, m_sex$estimate, m_alc$estimate),
                               c(m_bmi$conf.low, m_age60$conf.low, m_sex$conf.low, m_alc$conf.low),
                               c(m_bmi$conf.high, m_age60$conf.high, m_sex$conf.high, m_alc$conf.high)),
  `Untreated only` = sprintf("%.2f (%.2f to %.2f)",
                             c(s_bmi$estimate, s_age60$estimate, s_sex$estimate, s_alc$estimate),
                             c(s_bmi$conf.low, s_age60$conf.low, s_sex$conf.low, s_alc$conf.low),
                             c(s_bmi$conf.high, s_age60$conf.high, s_sex$conf.high, s_alc$conf.high))
) |>
  gt() |>
  tab_header(title = "Adjusted estimates (mmHg, 95% CI)",
             subtitle = sprintf("All n = %d; untreated n = %d", nobs(fit), nobs(fit_untr)))
```

Because `r n_treated` participants are on medication that lowers the very
quantity being measured, `systolic_bp` in this file is a *partly treated*
outcome. Refitting the model in the `r nobs(fit_untr)` untreated participants
changes the picture only modestly: the age contrast falls from
`r round(m_age60$estimate, 1)` to `r round(s_age60$estimate, 1)` mmHg and the
BMI slope from `r round(m_bmi$estimate, 2)` to `r round(s_bmi$estimate, 2)` mmHg
per kg/m2, while sex and alcohol are essentially unchanged. The direction of the
bias is predictable: treatment is concentrated in the older, heavier, higher-risk
participants and pulls *their* readings down, so the full-sample estimates for
age and BMI are, if anything, conservative. Excluding the treated is not a free
fix, however -- it conditions on a variable caused by the outcome and so
introduces its own selection bias -- which is why both fits are shown rather
than one being declared correct.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

```{r compare-numbers}
u_sec  <- uv_fit("education") |> filter(term == "educationSecondary")
m_sec  <- g("educationSecondary")
u_high <- uv_fit("education") |> filter(term == "educationHigher")
m_high <- g("educationHigher")
u_cur  <- uv_fit("smoking_status") |> filter(term == "smoking_statusCurrent")
u_smok <- uv_fit("smokeless_tobacco") |> filter(term == "smokeless_tobaccoYes")
m_smok <- g("smokeless_tobaccoYes")
```

Three predictors move a long way, all toward the null, and mostly for the same
reason -- age. **Low physical activity** is `r round(u_pa$estimate, 1)` mmHg
crude (95% CI `r round(u_pa$conf.low, 1)` to `r round(u_pa$conf.high, 1)`,
p `r format.pval(u_pa$p.value, digits = 2, eps = 0.001)`) and only
`r round(m_pa$estimate, 1)` mmHg adjusted (95% CI
`r round(m_pa$conf.low, 1)` to `r round(m_pa$conf.high, 1)`,
p = `r round(m_pa$p.value, 2)`), losing significance entirely: inactivity in this
sample is concentrated in the oldest and heaviest participants, and once age and
BMI are in the model there is nothing left for it to explain. **Higher education** is `r round(u_high$estimate, 1)` mmHg
relative to no or informal schooling in the crude model
(95% CI `r round(u_high$conf.low, 1)` to `r round(u_high$conf.high, 1)`,
p `r format.pval(u_high$p.value, digits = 2, eps = 0.001)`) but only
`r round(m_high$estimate, 1)` mmHg after adjustment
(95% CI `r round(m_high$conf.low, 1)` to `r round(m_high$conf.high, 1)`,
p = `r round(m_high$p.value, 2)`): the crude "protective effect of schooling" is
almost entirely the fact that educated participants in this municipality are
young. **Secondary education** behaves identically, moving from
`r round(u_sec$estimate, 1)` to `r round(m_sec$estimate, 1)` mmHg and losing
significance. **Current smoking** moves in the opposite direction, from
`r round(u_cur$estimate, 1)` mmHg crude to `r round(m_cur$estimate, 1)` mmHg
adjusted, because smokers are male (which raises blood pressure) but leaner
(which lowers it), and the two crude biases had partly cancelled. **Smokeless
tobacco** goes from `r round(u_smok$estimate, 1)` mmHg crude to
`r round(m_smok$estimate, 1)` mmHg adjusted, again an age and sex artefact.

Two predictors barely move at all. **BMI** goes from `r round(u_bmi$estimate, 2)`
to `r round(m_bmi$estimate, 2)` mmHg per kg/m2, a shift of under 10%, so the
adiposity--blood-pressure relationship in this sample is not explained by the
demographic or behavioural covariates. **Alcohol in the last month** goes from
`r round(u_alc$estimate, 1)` to `r round(m_alc$estimate, 1)` mmHg and stays
significant. The predictors that were null crudely -- salt, fruit and vegetable
intake, family history, poverty -- stay null; adjustment does not rescue them.
Overall
the pattern is confounding by age and sex, not mediation or collinearity: the
maximum variance inflation factor in this model is
`r round(max(check_collinearity(fit)$VIF), 2)`.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r het}
het <- check_heteroscedasticity(fit)
het_p <- as.numeric(het)
rob <- coeftest(fit, vcov = vcovHC(fit, type = "HC3"))
rob_ci <- coefci(fit, vcov. = vcovHC(fit, type = "HC3"))
keep <- c("bmi", "sexMale", "age_group60 or more", "alcohol_last_monthYes",
          "smoking_statusFormer", "salt_g_per_day")
tibble(
  Term = keep,
  Estimate = round(rob[keep, "Estimate"], 3),
  `SE (OLS)` = round(summary(fit)$coefficients[keep, "Std. Error"], 3),
  `SE (HC3)` = round(rob[keep, "Std. Error"], 3),
  `95% CI (HC3)` = sprintf("%.2f to %.2f", rob_ci[keep, 1], rob_ci[keep, 2]),
  `p (HC3)` = format.pval(rob[keep, "Pr(>|t|)"], digits = 2, eps = 0.001)
) |>
  gt() |>
  tab_header(title = "HC3 heteroscedasticity-consistent standard errors",
             subtitle = "Compared with ordinary least-squares standard errors")
```

The **posterior-predictive check** shows the model reproducing the centre of the
systolic distribution but under-predicting the right tail: the observed density
has a fatter upper shoulder around 150-180 mmHg than the model generates,
because a linear model with additive terms cannot manufacture the skew that real
blood pressure has. The **linearity** panel is broadly flat with a mild upward
drift at high fitted values, the same phenomenon. The **homogeneity of variance**
panel is the clear failure: the spread of residuals widens as fitted values
increase, and `check_heteroscedasticity()` confirms it
(p `r format.pval(het_p, digits = 2, eps = 0.001)`). This is intrinsic to blood
pressure -- people with higher average pressure are also more variable -- and it
biases the standard errors, not the coefficients. HC3 robust standard errors are
therefore reported in the table above; they are close to the OLS ones (for
example BMI `r round(summary(fit)$coefficients["bmi", "Std. Error"], 3)` versus
`r round(rob["bmi", "Std. Error"], 3)`), so no conclusion changes, but the robust
version is the one to quote. **Collinearity** is not a problem: the maximum VIF
is `r round(max(check_collinearity(fit)$VIF), 2)`, well below any threshold of
concern, which is expected once `mets_min_per_week` and `cigarettes_per_week`
have been kept out. **Influential observations** shows a handful of points near
the Cook's distance contour -- the participants with systolic pressures near 200
mmHg -- but none beyond it, and refitting without them does not change the
conclusions. **Normality of residuals** shows the expected right-tail departure;
with `r nobs(fit)` observations this is not a threat to the coefficient
estimates.

# Discussion

Among adults aged 30 and over in Bharatpur, systolic blood pressure is driven
mainly by age and body mass index. Each additional kg/m2 of BMI corresponds to
about `r round(m_bmi$estimate, 1)` mmHg, so the difference between a BMI of 22
and a BMI of 28 is roughly `r round(m_bmi$estimate * 6, 0)` mmHg -- clinically
meaningful, and the only large modifiable effect in the model. Being 60 or over
rather than in one's thirties adds a further `r round(m_age60$estimate, 0)` mmHg.
The classic lifestyle levers that a public-health programme would want to pull --
salt reduction, physical activity, fruit and vegetable intake -- show no
*independent* association here (physical activity looks protective crudely but
that is entirely age and adiposity), and the tobacco results run backwards, so
this analysis
supports weight management and age-based screening rather than dietary
messaging.

**Limitations.** The most serious problem is **treatment bias in the outcome
itself**. `r n_treated` participants (`r round(100 * n_treated / nrow(dat), 1)`%)
were taking antihypertensive medication when their blood pressure was measured,
which lowers the recorded value by an amount that depends on the drug, the dose
and adherence, none of which is recorded. Measured systolic pressure in this file
is therefore not "blood pressure" but "blood pressure as currently treated", and
because treatment is concentrated in exactly the people whose underlying pressure
is highest, every coefficient for a risk factor that also predicts treatment is
attenuated. The sensitivity analysis above bounds the problem rather than solving
it. Second, **the cluster sampling is not recoverable**: the source study
randomly selected only four wards and enrolled roughly 300 participants in each,
assuming a design effect of 1.5, but no ward identifier exists in this extract.
Consequently the standard errors and confidence intervals reported here are too
narrow -- plausibly by a factor of around the square root of the design effect,
so roughly 20% -- and no mixed model or survey-design correction can be fitted to
repair them. Estimates from only four clusters also cannot be generalised to
Nepal, or even confidently to the whole of Bharatpur. Third, **age is available
only in four bands**, so no per-year slope can be estimated and the oldest band
("60 or more") is open-ended and internally heterogeneous. Fourth, the design is
cross-sectional, so all estimates are associations: the former-smoker result is a
clear demonstration of reverse causation, and the same logic applies in principle
to diet and activity, since people who have been told they have high blood
pressure change their behaviour. Fifth, salt intake was estimated rather than
measured by 24-hour urinary sodium, and the estimate has essentially no
association with blood pressure here, which is at least as likely to reflect
measurement error as a true null. Sixth, the model explains only
`r round(100 * gl$r.squared, 1)`% of the variance, and residual variance is
non-constant. On the credit side, **there is no missing data at all** -- all
`r nrow(dat_raw)` rows are complete on all `r ncol(dat_raw)` variables -- so
none of the usual complete-case caveats apply.

**Future work.** Refit with the ward identifier as a random effect or survey
cluster if the original data can be obtained, and repeat the analysis restricted
to participants never treated, using a formal bias-correction (adding a fixed
offset of 10-15 mmHg to treated participants' readings) as a further sensitivity
analysis.

# Feasibility notes

*Not part of a student report.*

- **Works, with one large caveat.** A clean continuous outcome, 100% complete
  data, `r nobs(fit)` rows and `r length(coef(fit))` parameters, no collinearity
  (max VIF `r round(max(check_collinearity(fit)$VIF), 2)`), and real confounding
  that genuinely moves the education coefficients. Adjusted R-squared
  `r round(gl$adj.r.squared, 3)`.
- **The treatment trap is unavoidable and must be signposted in the brief.**
  `r n_treated` participants are medicated; treated mean systolic
  `r round(sbp_treated, 1)` versus `r round(sbp_untreated, 1)` mmHg untreated.
  A student who ignores this will report attenuated effects as if they were
  unbiased. Worse, a student who "controls for" `antihypertensive_drug` is
  adjusting for a consequence of the outcome and will get an even more distorted
  answer.
- **Forbidden predictors.** `diastolic_bp` (r = 0.79 with the outcome, same
  measurement occasion), `hypertension` (deterministic function of the outcome),
  `bp_awareness` and `antihypertensive_drug` (both downstream). Every one of
  these will look spectacular in a model and every one is circular.
- **Deterministic twins.** `physical_activity` is exactly
  `mets_min_per_week <= 600`; `cigarettes_per_week` is zero for all never-smokers.
  Putting both members of either pair into a model is a guaranteed collinearity
  failure.
- **Heteroscedasticity is certain to be flagged** (p
  `r format.pval(het_p, digits = 2, eps = 0.001)`). This is a good teaching
  moment provided HC3 is available; the HC3 and OLS standard errors barely
  differ, so the lesson is "check it, report it, and note it changed nothing".
- **No ward identifier.** The design effect of 1.5 assumed in the source sample
  size calculation cannot be honoured. Standard errors are too narrow throughout
  and this must appear in every report from this dataset.
- **Age in bands only.** Students expecting a "mmHg per year of age" figure
  cannot have one.
- **Several null lifestyle predictors** (salt, physical activity, fruit and
  vegetables, family history). This is realistic but disappointing; the brief
  should warn that a Purpose 3 analysis can legitimately end with most
  predictors null.
