---
title: "Physical health-related quality of life (SF-12 PCS) in adults with chronic disease"
subtitle: "Dataset 8: Patient activation for self-management in chronic disease"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "patient_activation.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "sf12_physical"
predictors <- c("age_yrs", "sex", "bmi", "smoking_status",
                "disease", "charlson_index", "ipq_total")

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex = factor(sex, levels = c("male", "female"), labels = c("Male", "Female")),
    education_level = factor(education_level, levels = c("lower", "middle", "higher"),
                             labels = c("Lower", "Middle", "Higher")),
    financial_distress = factor(financial_distress, levels = c("none", "low", "high"),
                                labels = c("None", "Low", "High")),
    smoking_status = factor(smoking_status, levels = c("never", "former", "current"),
                            labels = c("Never", "Former", "Current")),
    disease = factor(disease,
                     levels = c("diabetes_t2", "copd", "heart_failure", "renal_disease"),
                     labels = c("Type 2 diabetes", "COPD", "Heart failure", "Renal disease")),
    disease_duration = factor(disease_duration,
                              levels = c("under_2_yrs", "2_to_5_yrs", "over_5_yrs"),
                              labels = c("<2 years", "2-5 years", ">5 years")),
    living_alone = factor(living_alone, levels = c(0, 1), labels = c("No", "Yes"))
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

# Missing questionnaire answers were coded 99 in the original deposit; the
# conversion to NA was already applied upstream, so no re-coding is done here.
# The only 99 anywhere in the file is a participant identifier (id == 99).
n99 <- dat_raw |>
  select(-id) |>
  select(where(is.numeric)) |>
  summarise(across(everything(), ~ sum(.x == 99, na.rm = TRUE))) |>
  rowSums()
```

# Introduction

Physical health-related quality of life is the outcome patients themselves care
most about in long-term illness, and it is only loosely tied to clinical
severity markers. These data come from a cross-sectional survey of 1,154 Dutch
adults treated in primary and secondary care for type 2 diabetes, COPD, chronic
heart failure or chronic renal disease (Bos-Touwen et al., *PLoS ONE*
2015;10(5):e0126400), in which the SF-12 was administered alongside demographic,
anthropometric and illness-belief measures. Understanding which patient
characteristics track physical functioning across four different conditions
helps distinguish disease-specific from generic drivers of disability.

**Aim.** Among Dutch adults with one of four chronic conditions, we examined
which of age, sex, body mass index, smoking status, index disease, comorbidity
burden and illness perceptions are independently associated with the SF-12
physical component score, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** Cross-sectional survey of adults with type 2 diabetes, COPD,
chronic heart failure or chronic renal disease recruited from Dutch general
practices and hospital outpatient clinics; `r nrow(dat_raw)` participants, all
measured on a single occasion.

**Variables.** The outcome is `sf12_physical`, the SF-12 physical component
score on a 0-100 scale, where higher means better physical health. Predictors
are `age_yrs` (years), `sex` (reference: male), `bmi` (kg/m²),
`smoking_status` (never / former / current, reference: never — the clinically
unexposed group), `disease` (reference: type 2 diabetes, the largest group),
`charlson_index` (points; higher = more comorbidity) and `ipq_total` (Brief
Illness Perception Questionnaire total, 0-80; higher = a more threatening view
of one's illness).

**Body size.** In this file `bmi` is exactly `weight_kg / (height_cm/100)^2`, so
`bmi`, `height_cm` and `weight_kg` can never all appear in the same model. Only
`bmi` is used.

**Sibling scale deliberately excluded.** `sf12_mental` is the other component of
the same 12-item instrument, computed from overlapping items and completed at
the same sitting; it correlates 0.75 with the outcome. Its variance-inflation
factor if entered would be only 1.55, so collinearity is not the reason to leave
it out — the reason is that adjusted R² would rise from 0.43 to 0.65 and the
model would largely be the SF-12 predicting itself, which is a restatement of
the outcome rather than an explanation. It is therefore excluded.
`n_comorbidities` correlates 0.78 with `charlson_index`, so only the Charlson
index is used.

**Statistical analysis.** Linear regression (`lm()`), justified because the
outcome is a continuous 0-100 score. We fitted a univariable model for each
predictor first, then a single multivariable model containing all of them.
Predictors were chosen in advance on subject-matter grounds, not by significance.
Significance threshold 0.05; estimates are reported with 95% confidence
intervals. Assumptions were checked with `performance::check_model()`. Analysis
used `r R.version.string` with tidyverse, gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped). Missing questionnaire answers were
coded 99 in the original deposit, but that conversion to `NA` has already been
applied in this CSV; a check across all numeric columns other than `id` found
`r n99` remaining values equal to 99, so no further re-coding was needed.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(missing = "no") |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

The analysed sample is elderly and mostly male, with a median BMI in the
overweight range and a median Charlson index of 3. Median SF-12 physical score
is 50 with an interquartile range of roughly 31 to 75, so physical functioning
in this sample is both poor on average and very widely spread.

## Exploratory figure

```{r explore, fig.cap="Figure 1. SF-12 physical component score by index disease. Boxes show medians and interquartile ranges; points are individual participants."}
ggplot(dat_model, aes(x = disease, y = sf12_physical)) +
  geom_boxplot(outlier.shape = NA, fill = "#a6bddb") +
  geom_jitter(width = 0.2, height = 0, alpha = 0.18, size = 0.8) +
  labs(x = "Index chronic condition", y = "SF-12 physical component score (0-100)")
```

Physical functioning differs markedly between the four pooled disease groups:
heart failure and COPD patients sit roughly 15-20 points below the diabetes
group, with renal disease in between — differences far larger than anything the
demographic variables produce.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv
```

Illness perception is the strongest continuous crude predictor: each additional
IPQ point is associated with a 1.10-point lower SF-12 physical score (95% CI
-1.21 to -0.99, p < 0.001), so patients at the top and bottom of the IPQ range
differ by tens of points on the SF-12. Comorbidity burden is also strong
(-5.12 per Charlson point, 95% CI -5.92 to -4.31, p < 0.001), as is age (-0.54
per year, 95% CI -0.68 to -0.40) and female sex (-7.51, 95% CI -10.72 to -4.30).
BMI is modestly associated (-0.66 per kg/m², 95% CI -1.01 to -0.31, p < 0.001).
Crudely, heart failure patients score 20.8 points and COPD patients 15.3 points
below the diabetes group (both p < 0.001). Smoking status shows no crude
association at all (both contrasts p > 0.6).

## Why adjustment is needed

The crude disease differences in Figure 1 are the clearest case for adjustment
in this dataset. `charlson_index` is a confounder of them: the heart-failure and
renal groups carry substantially more comorbidity than the diabetes group, and
comorbidity independently predicts worse physical functioning, so a large part
of the apparent "heart failure effect" is really the extra disease those
patients carry. `age_yrs` works the same way — the heart-failure and renal
groups are older, and age independently predicts worse physical function.
`ipq_total` is a third: patients with the more disabling conditions also hold
more threatening illness beliefs, and illness beliefs are themselves associated
with reported functioning, so the crude disease contrasts partly reflect
differences in how patients construe their illness. Because the four groups were
pooled by design, `disease` has to be in the model both as an adjustment
variable for the other predictors and as a comparison of interest in its own
right.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv
```

```{r fit-stats}
gl <- glance(fit)
tibble(
  Statistic = c("R-squared", "Adjusted R-squared", "Residual SE", "N", "Model df"),
  Value = c(round(gl$r.squared, 3), round(gl$adj.r.squared, 3),
            round(gl$sigma, 2), nrow(dat_model), gl$df)
) |>
  gt() |>
  tab_header(title = "Model fit")
```

Conditional on all other predictors, each IPQ point is associated with a
0.94-point lower SF-12 physical score (95% CI -1.04 to -0.83, p < 0.001) and
each Charlson point with a 3.72-point lower score (95% CI -4.49 to -2.94,
p < 0.001). Age remains associated (-0.40 per year, 95% CI -0.52 to -0.28), as
does female sex (-7.87, 95% CI -10.42 to -5.33) and BMI (-0.86 per kg/m²,
95% CI -1.13 to -0.59, p < 0.001) — note that the BMI association *strengthens*
after adjustment. COPD (-7.39, 95% CI -10.77 to -4.02) and heart failure
(-4.87, 95% CI -8.65 to -1.08, p = 0.012) remain below diabetes; renal disease
now sits 3.78 points *above* diabetes, at the edge of significance (95% CI -0.01
to 7.57, p = 0.051). Smoking status is not associated in either the crude or the
adjusted model. The model explains R² = `r round(gl$r.squared, 3)` of the
variance (adjusted R² = `r round(gl$adj.r.squared, 3)`), much the best fit of
the five outcomes in this dataset.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

The disease coefficients move enormously towards the null. Heart failure goes
from -20.8 to -4.9 and COPD from -15.3 to -7.4, so roughly three-quarters and
one-half of those crude deficits respectively are explained by age, comorbidity
and illness perceptions rather than the diagnosis itself. Renal disease does
more than attenuate — it crosses the null, from -10.6 crude to +3.8 adjusted:
once you compare renal patients with diabetes patients of the same age and
comorbidity burden and the same illness beliefs, the renal patients report
slightly *better* physical function. That sign reversal is the single most
striking result in this analysis and should be reported cautiously, because it
depends heavily on the Charlson index doing the work that a disease-specific
severity marker (eGFR, dialysis status) should be doing. The Charlson index
itself attenuates from -5.12 to -3.72 and age from -0.54 to -0.40, both by
sharing explanatory work with each other and with `disease`. Two predictors
barely move: female sex (-7.51 to -7.87) and, in the other direction, BMI
strengthens from -0.66 to -0.86 — negative confounding, because BMI is inversely
related to age in this sample (r = -0.13) and older age also lowers the SF-12.
No predictor changes significance status except renal disease, which moves from
clearly significant and negative to borderline and positive. All VIFs are below
1.7, so none of this is collinearity.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r diag-tests}
check_heteroscedasticity(fit)
check_normality(fit)
check_collinearity(fit)
```

Linearity is acceptable and the reference-line smooth in the residual plot is
close to flat. Homogeneity of variance passes
(`check_heteroscedasticity()` p = 0.290), so ordinary standard errors are
reported and HC3 robust re-estimation was not required. Normality of residuals
is formally rejected (p < 0.001): the SF-12 physical score in this file takes
only 39 distinct values because it is a coarse sum of a small number of items,
so the residual distribution is visibly lumpy, and the posterior-predictive
check shows the model producing a smoother distribution than the data actually
has. With n = `r nrow(dat_model)` this has little effect on the confidence
intervals, but it is a real mismatch between an interval-scale model and an
ordinal-scale instrument. Collinearity is fine: the largest VIF is `disease`
at 1.61, then `charlson_index` at 1.43. No observation exceeds the Cook's
distance threshold.

# Discussion

Among Dutch adults with one of four chronic conditions, comorbidity burden and
threatening illness beliefs are the characteristics most strongly and
independently associated with physical health-related quality of life, followed
by age, female sex and higher BMI. Once these are held constant, the large raw
gaps between the four disease groups shrink by half or more, and the renal
group's apparent disadvantage disappears entirely. Practically, this says that
in a mixed chronic-disease clinic, how many other conditions a patient has and
how they think about their illness tell you more about their physical
functioning than which diagnosis is written at the top of their chart.

**Limitations.** The design is cross-sectional, so the illness-perception
association could run in either direction — patients who function badly may
reasonably come to see their illness as more threatening. IPQ and SF-12 are both
self-reported on the same occasion, so common-method bias inflates that
association. Most importantly, the file contains **no disease-specific severity
markers**: there is no eGFR or dialysis status for the renal group, no NYHA
class for heart failure and no GOLD stage for COPD. The `disease` coefficients
after adjustment therefore rest entirely on the generic Charlson index as a
severity proxy, and the renal-disease sign reversal in particular is very likely
an artefact of unmeasured severity rather than a real advantage. The SF-12 takes
only 39 distinct values here, so treating it as continuous is an approximation
and residual normality fails. Complete-case analysis dropped `r n_dropped`
participants (`r round(100*n_dropped/nrow(dat), 1)`%). All estimates are
associations, not effects.

**Future work.** Linking these survey data to disease-specific severity
measures would show how much of the residual between-disease difference is
severity and how much is the condition itself.

# Feasibility notes

*Not part of a student report.*

- **Best-fitting of the five outcomes.** R² = `r round(gl$r.squared, 3)`,
  adjusted R² = `r round(gl$adj.r.squared, 3)`, with several large, clearly
  significant coefficients. Students get a satisfying model.
- **Complete-case N = `r nrow(dat_model)`** of `r nrow(dat)` (`r n_dropped`
  dropped, `r round(100*n_dropped/nrow(dat), 1)`%).
- **No collinearity problem**: max VIF 1.61 (`disease`). But the sibling trap is
  live — `sf12_mental` has VIF 1.55, i.e. it looks harmless on a VIF check, yet
  adding it takes adjusted R² from 0.426 to 0.650. A student who includes it
  will conclude that "mental health is the strongest predictor of physical
  health" when they have really fitted the SF-12 to itself. This must be ruled
  out in the brief.
- **`bmi` = `weight_kg`/(`height_cm`/100)² exactly** in this file. Putting all
  three in one model gives a rank-deficient / near-singular fit. Only `bmi` is
  used here.
- **Diagnostics**: homoscedasticity passes (p = 0.29), so no HC3 needed.
  Normality is rejected, driven by the SF-12 having only 39 achievable values
  — worth showing students the posterior-predictive panel for this.
- **The renal-disease sign reversal** (crude -10.6 to adjusted +3.8) is
  interesting but is almost certainly residual confounding by unmeasured renal
  severity. Strong students can discuss it; weak students will report it as a
  finding.
- **`smoking_status` is a dud predictor** (crude and adjusted p > 0.5). Useful
  as an honest "not everything is significant" example, but a student hoping for
  a smoking result will be disappointed.
