---
title: "Diastolic blood pressure in Kenyan outpatient attenders"
subtitle: "Dataset 4: Kenya hypertension screening survey"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "kenya_hypertension.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "diastolic_bp"
predictors <- c("age_group", "sex", "education", "bmi_category",
                "current_smoker", "current_alcohol", "adequate_activity",
                "diabetes")

lab <- function(x, l) { attr(x, "label") <- l; x }

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest band, or the clinically
    # "unexposed" group. 26-35 rather than 18-25 is the age reference because
    # 18-25 holds only 101 people and would make every age contrast imprecise.
    age_group = lab(fct_relevel(factor(age_group),
                                "26-35", "18-25", "36-45", "46-55", "56-65", ">65"),
                    "Age band (years)"),
    sex = lab(factor(sex, levels = c("female", "male"), labels = c("Female", "Male")),
              "Sex"),
    # No schooling is the reference so higher categories read as "more education
    # than the reference".
    education = lab(fct_relevel(factor(education),
                                "No School", "Primary", "Secondary", "Tertiary"),
                    "Education"),
    # Healthy weight is both the largest category and the clinical comparator.
    bmi_category = lab(fct_relevel(factor(bmi_category),
                                   "Healthy", "Underweight", "Overweight", "Obese"),
                       "BMI category (WHO)"),
    current_smoker = lab(factor(current_smoker, levels = c(0, 1), labels = c("No", "Yes")),
                         "Current smoker"),
    current_alcohol = lab(factor(current_alcohol, levels = c(0, 1), labels = c("No", "Yes")),
                          "Current alcohol use"),
    adequate_activity = lab(factor(adequate_activity, levels = c(0, 1), labels = c("No", "Yes")),
                            "Meets activity recommendation"),
    # "unknown" is kept as its own level rather than set to missing: it is a
    # substantive answer (never tested) and dropping it would cost 261 rows.
    diabetes = lab(fct_relevel(factor(diabetes), "no", "yes", "unknown"),
                   "Self-reported diabetes"),
    cardiovascular_disease = lab(fct_relevel(factor(cardiovascular_disease),
                                             "no", "yes", "unknown"),
                                 "Self-reported cardiovascular disease"),
    study_site = lab(fct_relevel(factor(study_site), "public", "private", "faith"),
                     "Facility type"),
    monthly_income = lab(fct_relevel(factor(monthly_income),
                                     "0  - 15,000", "16,000 - 50,000",
                                     "51,000 - 100,000", "100,000 - 200,000",
                                     "Above 200,000"),
                         "Monthly household income (KSh)"),
    prior_htn_diagnosis = lab(factor(prior_htn_diagnosis, levels = c(0, 1),
                                     labels = c("No", "Yes")),
                              "Previously diagnosed with hypertension"),
    systolic_bp  = lab(systolic_bp, "Systolic BP (mmHg)"),
    diastolic_bp = lab(diastolic_bp, "Diastolic BP (mmHg)")
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

# The income-augmented sample is carried through the whole document so that the
# cost of adding a 31%-missing variable can be quantified rather than asserted.
dat_inc <- dat |> select(all_of(c(outcome, predictors, "monthly_income"))) |> drop_na()

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
pv  <- function(p) ifelse(p < 0.001, "< 0.001", paste0("= ", fmt(p, 3)))
```

# Introduction

Diastolic blood pressure is the pressure in the arteries between beats, and in
younger and middle-aged adults it is the component of blood pressure most
closely tied to peripheral vascular resistance. It behaves quite differently
from systolic pressure across the life course: diastolic pressure typically
rises until the fifth or sixth decade and then falls as large arteries stiffen,
while systolic pressure keeps rising. This dataset is a cross-sectional
screening survey of `r nrow(dat_raw)` adults recruited at three types of Kenyan
outpatient facility - a government (public) facility, a private facility and a
faith-based mission facility - in which each participant had blood pressure
measured three times and completed a WHO STEPS-style questionnaire.

**Aim.** Among these Kenyan outpatient attenders, we examined which of age band,
sex, education, BMI category, current smoking, current alcohol use, physical
activity and self-reported diabetes are independently associated with average
diastolic blood pressure, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** A cross-sectional facility-based screening survey of
`r nrow(dat_raw)` adults aged 18 and over attending outpatient services at three
Kenyan facilities (`r sum(dat$study_site == "public")` public,
`r sum(dat$study_site == "private")` private,
`r sum(dat$study_site == "faith")` faith-based). Blood pressure was taken three
times per participant and the file carries the average of the three readings.

**Variables.** The outcome is `diastolic_bp`, the mean of three diastolic
readings in mmHg, treated as continuous (observed range
`r fmt(min(dat$diastolic_bp), 0)` to `r fmt(max(dat$diastolic_bp), 0)` mmHg).
Predictors are age band (reference *26-35*, the largest band); sex (reference
*Female*); education (reference *No School*); WHO BMI category (reference
*Healthy*); current smoking (reference *No*); current alcohol use (reference
*No*); meeting the physical-activity recommendation (reference *No*); and
self-reported diabetes (reference *no*, with *unknown* retained as a substantive
third level rather than treated as missing).

**BMI is categorical, not continuous.** The file contains only the WHO
`bmi_category` label, not height, weight or a BMI value, so no per-unit BMI
effect (mmHg per kg/m2) can be reported from these data. More generally **no
predictor in this file is continuous** - age is supplied only as a six-level
band - so this linear model is a cell-means (ANOVA-style) model and no
slope-per-unit statement of any kind is possible.

**Variables deliberately excluded.** `hypertension` is defined *from* the
measured average blood pressure and is therefore a deterministic function of
this outcome and `systolic_bp`; it can never sit on the right-hand side.
`systolic_bp` is excluded because it is the other half of the same measurement
(correlation with diastolic pressure r =
`r fmt(cor(dat$systolic_bp, dat$diastolic_bp), 2)`), a co-outcome rather than an
exposure; a model containing it would explain a great deal of variance and
answer nothing. `prior_htn_diagnosis` is kept out of the main model and examined
separately, because a prior diagnosis usually implies treatment, which lowers
measured pressure, so that variable sits downstream of the outcome as much as
upstream of it. `monthly_income` is excluded on missing-data grounds and its
cost is quantified below.

**Statistical analysis.** `lm()`, justified because the outcome is a continuous,
roughly symmetric measurement. We fitted a univariable model for each predictor
first, then a single multivariable model containing all eight. Predictors were
chosen in advance on subject-matter grounds, not by significance. Significance
threshold 0.05; estimates are reported with 95% confidence intervals.
Assumptions were checked with `performance::check_model()`, and because
non-constant variance was flagged we additionally report HC3
heteroscedasticity-consistent standard errors. Analysis used
`r R.version.string` with tidyverse, gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped,
`r fmt(100 * n_dropped / nrow(dat), 1)`%). The outcome and four predictors are
100% complete; the losses come from `bmi_category`
(`r sum(is.na(dat$bmi_category))` missing), `education`
(`r sum(is.na(dat$education))`) and `age_group`
(`r sum(is.na(dat$age_group))`). The one seriously incomplete variable is
`monthly_income`, missing for `r sum(is.na(dat$monthly_income))` participants
(`r fmt(100 * mean(is.na(dat$monthly_income)), 1)`%); it is not in the main
model and the section *The cost of `monthly_income`* below reports what
including it would do.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(missing = "no") |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics, complete-case analytic sample.**")
```

```{r table-one-numbers}
med_dbp <- median(dat_model$diastolic_bp)
iqr_dbp <- quantile(dat_model$diastolic_bp, c(0.25, 0.75))
```

Median diastolic pressure is `r fmt(med_dbp, 1)` mmHg (IQR
`r fmt(iqr_dbp[1], 1)` to `r fmt(iqr_dbp[2], 1)`), a much tighter distribution
than systolic pressure and closer to symmetric. The sample is predominantly
female (`r fmt(100 * mean(dat_model$sex == "Female"), 0)`%), better educated
than the Kenyan adult population
(`r fmt(100 * mean(dat_model$education == "Tertiary"), 0)`% tertiary) and
`r fmt(100 * mean(dat_model$bmi_category %in% c("Overweight", "Obese")), 0)`%
overweight or obese; current smoking
(`r fmt(100 * mean(dat_model$current_smoker == "Yes"), 1)`%) and alcohol use
(`r fmt(100 * mean(dat_model$current_alcohol == "Yes"), 1)`%) are rare, as
expected in a mostly female sample.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Mean systolic and diastolic blood pressure by age band, with 95% confidence intervals. Diastolic pressure peaks in the fifth decade and then falls; systolic pressure keeps rising."}
dat |>
  filter(!is.na(age_group)) |>
  mutate(age_group = fct_relevel(age_group, "18-25", "26-35", "36-45",
                                 "46-55", "56-65", ">65")) |>
  select(age_group, Systolic = systolic_bp, Diastolic = diastolic_bp) |>
  pivot_longer(-age_group, names_to = "Measure", values_to = "mmHg") |>
  group_by(age_group, Measure) |>
  summarise(mean = mean(mmHg),
            lo = mean(mmHg) - 1.96 * sd(mmHg) / sqrt(n()),
            hi = mean(mmHg) + 1.96 * sd(mmHg) / sqrt(n()),
            .groups = "drop") |>
  ggplot(aes(age_group, mean, colour = Measure, group = Measure)) +
  geom_line(linewidth = 1) +
  geom_pointrange(aes(ymin = lo, ymax = hi), size = 0.5) +
  scale_colour_manual(values = c(Systolic = "#2E6F9E", Diastolic = "#B25D9C")) +
  labs(x = "Age band (years)", y = "Mean blood pressure (mmHg)",
       colour = NULL,
       caption = "Whiskers are 95% CIs for the mean within each band")
```

Diastolic pressure rises from `r fmt(mean(dat$diastolic_bp[dat$age_group == "18-25"], na.rm = TRUE), 1)`
mmHg in the 18-25 band to a peak of
`r fmt(mean(dat$diastolic_bp[dat$age_group == "46-55"], na.rm = TRUE), 1)` mmHg
at 46-55 and then falls back to
`r fmt(mean(dat$diastolic_bp[dat$age_group == ">65"], na.rm = TRUE), 1)` mmHg in
the over-65s, while systolic pressure climbs throughout - so the age
relationship for this outcome is an inverted U rather than a straight line, and
pulse pressure widens from about
`r fmt(mean(dat$systolic_bp[dat$age_group == "18-25"] - dat$diastolic_bp[dat$age_group == "18-25"], na.rm = TRUE), 0)`
to `r fmt(mean(dat$systolic_bp[dat$age_group == ">65"] - dat$diastolic_bp[dat$age_group == ">65"], na.rm = TRUE), 0)`
mmHg, the classic signature of large-artery stiffening.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv

utid <- map_dfr(predictors, function(v) {
  tidy(lm(reformulate(v, outcome), data = dat_model), conf.int = TRUE) |>
    filter(term != "(Intercept)")
})
ug <- function(term, d = 2) {
  r <- utid[utid$term == term, ]
  stopifnot(nrow(r) == 1L)
  sprintf("%s mmHg (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d), fmt(r$conf.high, d))
}
up <- function(term) { x <- utid$p.value[utid$term == term]; stopifnot(length(x) == 1L); pv(x) }
```

Age is again the strongest crude predictor but the pattern is not monotonic:
relative to the 26-35 band, the 46-55 band averages `r ug("age_group46-55")`
higher diastolic pressure (p `r up("age_group46-55")`) and the 56-65 band
`r ug("age_group56-65")` (p `r up("age_group56-65")`), but the over-65s only
`r ug("age_group>65")` (p `r up("age_group>65")`) - the coefficient falls back
in the oldest band exactly as Figure 1 showed. Adiposity is the other clear
crude signal: obese participants average `r ug("bmi_categoryObese")` above
healthy-weight participants (p `r up("bmi_categoryObese")`) and overweight
participants `r ug("bmi_categoryOverweight")` (p `r up("bmi_categoryOverweight")`).
Current alcohol use is associated with `r ug("current_alcoholYes")`
(p `r up("current_alcoholYes")`) - a crude association it does *not* have with
systolic pressure - and men average `r ug("sexMale")` above women
(p `r up("sexMale")`). Primary schooling sits `r ug("educationPrimary")` above
no schooling (p `r up("educationPrimary")`), diabetes `r ug("diabetesyes")`
(p `r up("diabetesyes")`), and physical activity shows nothing.

## Why adjustment is needed

The crude estimates cannot be compared with one another because the predictors
are heavily interlocked. **Age band** confounds nearly everything, but in a way
that is easy to get wrong for this outcome: it is associated with diabetes
(`r fmt(100 * mean(dat_model$diabetes[dat_model$age_group == ">65"] == "yes"), 0)`%
of over-65s report diabetes versus
`r fmt(100 * mean(dat_model$diabetes[dat_model$age_group == "26-35"] == "yes"), 0)`%
of 26-35s), with education
(`r fmt(100 * mean(dat_model$education[dat_model$age_group == ">65"] == "No School"), 0)`%
of over-65s never attended school versus
`r fmt(100 * mean(dat_model$education[dat_model$age_group == "26-35"] == "No School"), 0)`%
of 26-35s) and with BMI category, and it has a non-monotonic relationship with
diastolic pressure. The crude diabetes coefficient of about
`r fmt(utid$estimate[utid$term == "diabetesyes"], 1)` mmHg is therefore mostly
an age and adiposity contrast between people with and without diabetes.
**Sex** confounds both behavioural exposures: only
`r sum(dat_model$current_alcohol == "Yes" & dat_model$sex == "Female")` of the
`r sum(dat_model$current_alcohol == "Yes")` current drinkers in the analytic
sample are women, and
`r sum(dat_model$current_smoker == "Yes" & dat_model$sex == "Male")` of
`r sum(dat_model$current_smoker == "Yes")` smokers are men, while men also have
higher diastolic pressure at a given age - so the crude alcohol and smoking
coefficients are partly sex coefficients and must be adjusted before they mean
anything. **BMI category** confounds the diabetes and activity contrasts, since
adiposity causes both insulin resistance and raised peripheral resistance.
Mutual adjustment lets each coefficient be read as a comparison between
participants alike on the other seven measured characteristics.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv

tid <- tidy(fit, conf.int = TRUE)
g <- function(term, d = 2) {
  r <- tid[tid$term == term, ]
  stopifnot(nrow(r) == 1L)
  sprintf("%s mmHg (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d), fmt(r$conf.high, d))
}
gp <- function(term) { x <- tid$p.value[tid$term == term]; stopifnot(length(x) == 1L); pv(x) }
```

```{r fit-stats}
s <- summary(fit)
tibble(
  Statistic = c("Complete-case N", "Rows dropped by listwise deletion",
                "Parameters (excl. intercept)", "R-squared",
                "Adjusted R-squared", "Residual SE (mmHg)", "AIC"),
  Value = c(as.character(nrow(dat_model)),
            as.character(n_dropped),
            as.character(length(coef(fit)) - 1L),
            fmt(s$r.squared, 3), fmt(s$adj.r.squared, 3), fmt(s$sigma, 2),
            fmt(AIC(fit), 1))
) |>
  gt() |>
  tab_header(title = md("**Model fit: multivariable linear model for diastolic BP**"))
```

The model was fitted on `r nrow(dat_model)` complete cases and explains
`r fmt(100 * s$r.squared, 1)`% of the variance in diastolic pressure (adjusted
R-squared `r fmt(s$adj.r.squared, 3)`, residual SE `r fmt(s$sigma, 1)` mmHg) -
noticeably less than the same predictors explain for systolic pressure, which is
expected given the inverted-U age pattern. Among participants of the same sex,
education, BMI category, smoking, drinking, activity and diabetes status, the
46-55 band averages `r g("age_group46-55")` higher diastolic pressure than the
26-35 band (p `r gp("age_group46-55")`) while the over-65s average only
`r g("age_group>65")` (p `r gp("age_group>65")`), so age adds little at the top
of the range. Obesity retains `r g("bmi_categoryObese")`
(p `r gp("bmi_categoryObese")`) and being overweight `r g("bmi_categoryOverweight")`
(p `r gp("bmi_categoryOverweight")`). Current alcohol use retains
`r g("current_alcoholYes")` (p `r gp("current_alcoholYes")`) and male sex
`r g("sexMale")` (p `r gp("sexMale")`). Diabetes is null after adjustment
(`r g("diabetesyes")`, p `r gp("diabetesyes")`). These are associations among
the participants observed, not effects of intervening on any of these variables.

## The cost of `monthly_income`

```{r income-cost}
fit_inc      <- lm(reformulate(c(predictors, "monthly_income"), outcome), data = dat_inc)
fit_noinc_cc <- lm(reformulate(predictors, outcome), data = dat_inc)

inc_tid <- tidy(fit_inc, conf.int = TRUE)
si <- summary(fit_inc)

tibble(
  Model = c("Main model (no income)",
            "Same model, income complete-cases only",
            "Model with monthly_income added"),
  N = c(nrow(dat_model), nrow(dat_inc), nrow(dat_inc)),
  `% of file` = fmt(100 * c(nrow(dat_model), nrow(dat_inc), nrow(dat_inc)) / nrow(dat), 1),
  Parameters = c(length(coef(fit)) - 1L, length(coef(fit_noinc_cc)) - 1L,
                 length(coef(fit_inc)) - 1L),
  `R-squared` = fmt(c(s$r.squared, summary(fit_noinc_cc)$r.squared, si$r.squared), 3),
  `Adj. R-squared` = fmt(c(s$adj.r.squared, summary(fit_noinc_cc)$adj.r.squared,
                           si$adj.r.squared), 3)
) |>
  gt() |>
  tab_header(title = md("**Table 2. What adding `monthly_income` costs.**"),
             subtitle = md("`monthly_income` is missing for 31% of participants; listwise deletion removes them from the whole model, not just from that one coefficient."))
```

```{r income-numbers}
inc_lost <- nrow(dat_model) - nrow(dat_inc)
inc_p    <- anova(fit_noinc_cc, fit_inc)$`Pr(>F)`[2]
ob_main  <- tid$estimate[tid$term == "bmi_categoryObese"]
ob_inc   <- inc_tid$estimate[inc_tid$term == "bmi_categoryObese"]
al_main  <- tid$estimate[tid$term == "current_alcoholYes"]
al_inc   <- inc_tid$estimate[inc_tid$term == "current_alcoholYes"]
al_p_inc <- inc_tid$p.value[inc_tid$term == "current_alcoholYes"]
miss_by_site <- dat |> group_by(study_site) |>
  summarise(pct = 100 * mean(is.na(monthly_income)))
```

This is the most important methodological point in this dataset.
`monthly_income` is missing for
`r fmt(100 * mean(is.na(dat$monthly_income)), 1)`% of participants, and because
`lm()` deletes rows listwise, adding it removes those participants from *every*
coefficient in the model, not just from the income coefficient. The
complete-case N falls from **`r nrow(dat_model)`** to **`r nrow(dat_inc)`**
(`r fmt(100 * nrow(dat_inc) / nrow(dat), 0)`% of the file): `r inc_lost` further
participants are discarded to buy
`r length(coef(fit_inc)) - length(coef(fit))` extra parameters. The F-test
comparing the two models on the *same* `r nrow(dat_inc)` rows gives
p `r pv(inc_p)`, and adjusted R-squared moves from
`r fmt(summary(fit_noinc_cc)$adj.r.squared, 3)` to `r fmt(si$adj.r.squared, 3)`.

The missingness is also plainly not random. Non-response to the income question
ranges from `r fmt(miss_by_site$pct[miss_by_site$study_site == "faith"], 0)`% at
the faith-based facility to
`r fmt(miss_by_site$pct[miss_by_site$study_site == "private"], 0)`% at the
private facility, and rises with education (from
`r fmt(100 * mean(is.na(dat$monthly_income[dat$education == "No School"])), 0)`%
among those with no schooling to
`r fmt(100 * mean(is.na(dat$monthly_income[dat$education == "Tertiary"])), 0)`%
among the tertiary educated), so the people who answer are systematically
different from those who do not and the reduced sample is not a random
subsample. The consequences are visible here: the alcohol coefficient moves from
`r fmt(al_main, 2)` mmHg (p `r pv(tid$p.value[tid$term == "current_alcoholYes"])`)
in the full complete-case sample to `r fmt(al_inc, 2)` mmHg
(p `r pv(al_p_inc)`) in the income-restricted one, while obesity is more stable
(`r fmt(ob_main, 2)` to `r fmt(ob_inc, 2)` mmHg). **The main model therefore
excludes `monthly_income`**; a student who wants the income gradient must either
accept a one-third smaller and non-random sample or use multiple imputation.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 3. Unadjusted and adjusted estimates.**")
```

The largest movement is **self-reported diabetes**, which falls from
`r ug("diabetesyes")` crude (p `r up("diabetesyes")`, borderline) to
`r g("diabetesyes")` adjusted (p `r gp("diabetesyes")`) - it not only loses
significance but changes sign, so the entire crude diabetes association with
diastolic pressure in these data is the age and adiposity difference between
participants with and without diabetes. **Current smoking** attenuates the same
way, from `r ug("current_smokerYes")` crude (p `r up("current_smokerYes")`) to
`r g("current_smokerYes")` adjusted (p `r gp("current_smokerYes")`); smokers
here are overwhelmingly male, so the crude estimate was largely the sex
difference. The **over-65 age contrast** halves, from `r ug("age_group>65")`
(p `r up("age_group>65")`) to `r g("age_group>65")` (p `r gp("age_group>65")`),
losing significance. Moving in the opposite direction, **male sex** grows
slightly away from the null, from `r ug("sexMale")` to `r g("sexMale")`, and
**primary education** from `r ug("educationPrimary")` to `r g("educationPrimary")`,
because both were partly masked by the confounders. Two predictors barely move:
**obesity** (`r ug("bmi_categoryObese")` to `r g("bmi_categoryObese")`) and
**current alcohol use** (`r ug("current_alcoholYes")` to
`r g("current_alcoholYes")`), so the alcohol-diastolic association is not simply
a sex artefact and is the one behavioural signal that survives adjustment.
Finally **"unknown" diabetes status** moves *away* from the null
(`r ug("diabetesunknown")` to `r g("diabetesunknown")`, p `r gp("diabetesunknown")`),
a reminder that this level means "never tested" rather than "no diabetes" and
should not be interpreted as a dose of disease.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r het-check}
het <- check_heteroscedasticity(fit)
print(het)
vifs <- check_collinearity(fit)
max_vif <- max(vifs$VIF)
```

```{r robust-se}
rob   <- lmtest::coeftest(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
robci <- lmtest::coefci(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
tibble(
  Term = rownames(rob),
  Estimate = fmt(rob[, 1], 2),
  `SE (OLS)` = fmt(summary(fit)$coefficients[, 2], 3),
  `SE (HC3)` = fmt(rob[, 2], 3),
  `95% CI (HC3)` = paste0(fmt(robci[, 1], 2), " to ", fmt(robci[, 2], 2)),
  `p (HC3)` = fmt(rob[, 4], 3)
) |>
  filter(Term != "(Intercept)") |>
  gt() |>
  tab_header(title = md("**Table 4. HC3 heteroscedasticity-consistent standard errors.**"))
```

```{r het-numbers}
sig_ols <- summary(fit)$coefficients[, 4] < 0.05
sig_hc3 <- rob[, 4] < 0.05
n_flip  <- sum(sig_ols != sig_hc3)
flip_terms <- rownames(rob)[sig_ols != sig_hc3]
max_se_diff <- 100 * max(abs(rob[, 2] - summary(fit)$coefficients[, 2]) /
                           summary(fit)$coefficients[, 2])
```

The posterior-predictive panel shows the model reproduces the centre of the
diastolic distribution reasonably but flattens both tails, which is what a
cell-means model with eight categorical predictors can do and no more. The
residual-versus-fitted panel fans out towards higher fitted values and
`check_heteroscedasticity()` confirms non-constant error variance (p < 0.001):
people with higher average diastolic pressure vary more from reading to reading
and from person to person. The normal-QQ panel is closer to the line than for
systolic pressure - the diastolic distribution is more symmetric - with a modest
tail at both ends. The "linearity" panel is uninformative because no predictor
is continuous: every fitted value is a cell mean. Collinearity is low (maximum
VIF `r fmt(max_vif, 2)`, for `r vifs$Term[which.max(vifs$VIF)]`), so the
adjusted estimates are stable, and no observation exceeds the Cook's distance
contours.

Because non-constant variance was detected, Table 4 reports HC3 robust standard
errors. Point estimates are unchanged; only the uncertainty is re-estimated. The
HC3 standard errors differ from the OLS ones by at most
`r fmt(max_se_diff, 0)`%, and
`r if (n_flip == 0) "no coefficient changes its significance status" else paste0(n_flip, " coefficient(s) change significance status (", paste(flip_terms, collapse = ", "), ")")`
under robust inference. The substantive conclusions - obesity, sex, alcohol and
mid-life age in; diabetes, smoking and activity out - do not depend on the
variance assumption.

## Sensitivity analysis: prior diagnosis and treatment

```{r prior-sens}
dat_prior <- dat |>
  select(all_of(c(outcome, predictors, "prior_htn_diagnosis"))) |> drop_na()
fit_prior <- lm(reformulate(c(predictors, "prior_htn_diagnosis"), outcome),
                data = dat_prior)
pt <- tidy(fit_prior, conf.int = TRUE) |> filter(term == "prior_htn_diagnosisYes")
dbp_dx  <- mean(dat$diastolic_bp[dat$prior_htn_diagnosis == "Yes"])
dbp_ndx <- mean(dat$diastolic_bp[dat$prior_htn_diagnosis == "No"])
```

Participants reporting a previous hypertension diagnosis average
`r fmt(dbp_dx, 1)` mmHg diastolic against `r fmt(dbp_ndx, 1)` mmHg in those who
do not, and adding `prior_htn_diagnosis` to the model gives an adjusted
difference of `r fmt(pt$estimate, 1)` mmHg (95% CI `r fmt(pt$conf.low, 1)` to
`r fmt(pt$conf.high, 1)`, p `r pv(pt$p.value)`), raising R-squared to
`r fmt(summary(fit_prior)$r.squared, 3)`. **This coefficient is not an effect of
diagnosis on blood pressure.** The relationship is bidirectional: a diagnosis
happens *because* pressure was high (pushing the coefficient up) and leads to
antihypertensive treatment that lowers measured pressure (pushing it down). The
file has **no medication variable**, so the two directions cannot be separated.
The variable is therefore excluded from the main model and reported here only as
a sensitivity analysis.

# Discussion

In this Kenyan outpatient sample, BMI category, sex, mid-life age and current
alcohol use are independently associated with diastolic blood pressure. Adjusted
for the other seven predictors, obese participants average
`r g("bmi_categoryObese")` above healthy-weight participants, current drinkers
`r g("current_alcoholYes")` above non-drinkers, and men `r g("sexMale")` above
women. The age relationship is an inverted U: pressure peaks in the 46-55 band
(`r g("age_group46-55")` above 26-35) and is essentially back to the reference
level by 65, so - unlike systolic pressure - diastolic pressure does not
identify the oldest attenders as highest risk. The crude diabetes association
disappears completely on adjustment. Practically, diastolic pressure is the
wrong screening quantity for older adults in this setting, and body size is the
modifiable characteristic most strongly tied to it.

**Limitations.** This is a **facility-based convenience sample** of outpatient
attenders, not a community sample, and is far better educated and more female
than the Kenyan adult population, so no mean here generalises. The design is
**cross-sectional**; nothing establishes temporality. **No predictor is
continuous** - age is banded and BMI is only a WHO category - so residual
confounding within band is guaranteed and no per-unit effect exists.
**Antihypertensive medication is not recorded**, the most serious unmeasured
confounder here: treated participants have lower measured pressure precisely
because their underlying risk is higher, which biases the age, obesity and
diabetes coefficients toward the null. Participants come from **three facilities
with no facility identifier**, so the clustering of attenders within a facility
cannot be modelled and the standard errors, robust or not, are probably too
small. Complete-case analysis dropped `r n_dropped` rows, and `monthly_income`
was excluded because including it would have cost a further
`r nrow(dat_model) - nrow(dat_inc)` rows, so household economic position is
unadjusted for. Alcohol is recorded only as a yes/no, with no quantity or
frequency, so the alcohol coefficient is a very crude exposure contrast built on
`r sum(dat_model$current_alcohol == "Yes")` drinkers. Sodium intake, family
history and kidney function are absent. Only `r fmt(100 * s$r.squared, 0)`% of
the variance is explained.

**Future work.** Refit with continuous age, measured BMI, quantity-based alcohol
exposure and a medication variable, and use a facility identifier to account for
clustering - none of which this extract carries.

# Feasibility notes

*NOT part of a student report.*

- **Workable, but the weaker of the two blood-pressure outcomes.** Same
  `r nrow(dat_model)` complete cases and the same predictor set as
  `systolic_bp`, but the model explains `r fmt(100 * s$r.squared, 1)`% of
  variance against about 18% for systolic pressure, because the age relationship
  is non-monotonic.
- **The inverted-U age pattern is the interesting feature and the main trap.**
  Diastolic pressure peaks at 46-55 and falls in the over-65s. Students who
  assume "older = higher" will misreport the age coefficients, and anyone who
  models age as a linear term (if they recode the bands to numbers) will find
  almost nothing. As banded factors it works, which is a lucky accident of how
  the file was coded.
- **FATAL trap: `hypertension` must never be a predictor.** It is defined from
  the measured pressures, so it is a coarsened function of this outcome.
  `systolic_bp` must also stay out (r =
  `r fmt(cor(dat$systolic_bp, dat$diastolic_bp), 2)` and a co-outcome). A model
  with either produces a big R-squared and answers nothing.
- **Missing-data teaching case.** `monthly_income` is
  `r fmt(100 * mean(is.na(dat$monthly_income)), 1)`% missing; adding it cuts the
  sample from `r nrow(dat_model)` to `r nrow(dat_inc)` and contributes nothing
  (p `r pv(inc_p)`). Here it also visibly destabilises the alcohol coefficient
  (`r fmt(al_main, 2)` to `r fmt(al_inc, 2)` mmHg), which makes the point better
  than the systolic version does.
- **Heteroscedasticity flagged (p < 0.001)**; HC3 changes nothing substantive
  (`r if (n_flip == 0) "no significance flips" else paste(n_flip, "significance flip(s)")`).
  Good, genuine practice material.
- **`current_smoker` is sparse** (`r sum(dat$current_smoker == "Yes")` smokers in
  the file, almost all male) and **`bmi_category` "Underweight" is thin**
  (`r sum(dat_model$bmi_category == "Underweight")` in the analytic sample). Keep
  both to one contrast each; no interactions.
- **Substantially redundant with `systolic_bp`.** Same predictors, same sample,
  largely the same story. A student should do one of the two, not both; systolic
  is the better choice unless the inverted-U age pattern is the point of the
  project.
- **Clustering cannot be modelled** (three facilities, no facility ID). No
  separation, no collinearity (max VIF `r fmt(max_vif, 2)`), no influential
  points.
