---
title: "Previous hypertension diagnosis in Kenyan outpatient attenders"
subtitle: "Dataset 4: Kenya hypertension screening survey"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "kenya_hypertension.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names. The predictor set is
# built around contact with care, not physiology, because that is what a
# self-reported earlier diagnosis actually measures.
outcome    <- "prior_htn_diagnosis"
predictors <- c("age_group", "sex", "education", "employment", "bmi_category",
                "diabetes", "cardiovascular_disease", "study_site")

lab <- function(x, l) { attr(x, "label") <- l; x }

dat <- dat_raw |>
  mutate(
    # Event level is "Yes"; a two-level factor with "No" first makes glm() model
    # the probability of a previous diagnosis, and lets Table 1 be grouped by
    # outcome.
    prior_htn_diagnosis = lab(factor(prior_htn_diagnosis, levels = c(0, 1),
                                     labels = c("No", "Yes")),
                              "Previously diagnosed with hypertension"),
    hypertension = lab(factor(hypertension, levels = c(0, 1), labels = c("No", "Yes")),
                       "Hypertension on measured BP"),
    # Deliberate reference levels: the largest band, or the clinically
    # "unexposed" group. 26-35 rather than 18-25 is the age reference because
    # 18-25 holds only 101 people and would make every age contrast imprecise.
    age_group = lab(fct_relevel(factor(age_group),
                                "26-35", "18-25", "36-45", "46-55", "56-65", ">65"),
                    "Age band (years)"),
    sex = lab(factor(sex, levels = c("female", "male"), labels = c("Female", "Male")),
              "Sex"),
    # No schooling is the reference so higher categories read as "more education
    # than the reference" - the natural framing for an access-to-care outcome.
    education = lab(fct_relevel(factor(education),
                                "No School", "Primary", "Secondary", "Tertiary"),
                    "Education"),
    employment = lab(fct_relevel(factor(employment), "Employed / Self Employed",
                                 "Unemployed", "Retired"),
                     "Employment status"),
    bmi_category = lab(fct_relevel(factor(bmi_category),
                                   "Healthy", "Underweight", "Overweight", "Obese"),
                       "BMI category (WHO)"),
    # "unknown" is kept as its own level rather than set to missing: for these
    # two variables it means "never tested / never told", which is itself an
    # access-to-care state and is exactly what this analysis is about.
    diabetes = lab(fct_relevel(factor(diabetes), "no", "yes", "unknown"),
                   "Self-reported diabetes"),
    cardiovascular_disease = lab(fct_relevel(factor(cardiovascular_disease),
                                             "no", "yes", "unknown"),
                                 "Self-reported cardiovascular disease"),
    # Public (government) facility is the reference: it is the default point of
    # care in Kenya and the natural comparator for the private and mission
    # sectors.
    study_site = lab(fct_relevel(factor(study_site), "public", "private", "faith"),
                     "Facility type"),
    monthly_income = lab(fct_relevel(factor(monthly_income),
                                     "0  - 15,000", "16,000 - 50,000",
                                     "51,000 - 100,000", "100,000 - 200,000",
                                     "Above 200,000"),
                         "Monthly household income (KSh)"),
    current_smoker = lab(factor(current_smoker, levels = c(0, 1), labels = c("No", "Yes")),
                         "Current smoker")
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)
n_events  <- sum(dat_model[[outcome]] == "Yes")

# The income-augmented sample is carried through the whole document so that the
# cost of adding a 31%-missing variable can be quantified rather than asserted.
dat_inc      <- dat |> select(all_of(c(outcome, predictors, "monthly_income"))) |> drop_na()
n_events_inc <- sum(dat_inc[[outcome]] == "Yes")

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
pv  <- function(p) ifelse(p < 0.001, "< 0.001", paste0("= ", fmt(p, 3)))
```

# Introduction

Roughly half of all hypertension in sub-Saharan Africa is undiagnosed, and the
diagnosis step - not treatment or control - is where most of the care cascade is
lost. This dataset is a cross-sectional screening survey of `r nrow(dat_raw)`
adults recruited at three types of Kenyan outpatient facility - a government
(public) facility, a private facility and a faith-based mission facility - in
which every participant had blood pressure measured three times and was asked
whether they had ever previously been told by a health worker that they had
hypertension.

That self-reported earlier diagnosis is a **different construct from the
measured hypertension recorded on the same day**. Measured hypertension is
physiology at one moment; a prior diagnosis is a record of having had contact
with a health system that measured your blood pressure and told you the result.
Modelling it asks who reaches diagnostic services, which is the more interesting
question this file can answer and the one for which the predictor set below is
assembled: age, sex, education, employment, body size, other diagnosed
conditions and facility type.

**Aim.** Among these Kenyan outpatient attenders, we examined which of age band,
sex, education, employment status, BMI category, self-reported diabetes,
self-reported cardiovascular disease and facility type are independently
associated with reporting a previous hypertension diagnosis, mutually adjusting
for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** A cross-sectional facility-based screening survey of
`r nrow(dat_raw)` adults aged 18 and over attending outpatient services at three
Kenyan facilities (`r sum(dat$study_site == "public")` public,
`r sum(dat$study_site == "private")` private,
`r sum(dat$study_site == "faith")` faith-based).

**Variables.** The outcome is `prior_htn_diagnosis`, self-report of having
previously been diagnosed with hypertension; the modelled event is **prior
diagnosis = Yes** (`r sum(dat$prior_htn_diagnosis == "Yes")` of `r nrow(dat)`,
`r fmt(100 * mean(dat$prior_htn_diagnosis == "Yes"), 1)`%). Predictors are age
band (reference *26-35*); sex (reference *Female*); education (reference *No
School*); employment status (reference *Employed / Self Employed*); WHO BMI
category (reference *Healthy*); self-reported diabetes (reference *no*);
self-reported cardiovascular disease (reference *no*); and facility type
(reference *public*, the default point of care in Kenya). For diabetes and
cardiovascular disease the *unknown* level is retained rather than treated as
missing: "never tested / never told" is itself a state of contact with care and
is precisely what this analysis is about.

**Measured `hypertension`, `systolic_bp` and `diastolic_bp` are excluded from
the main model** and examined in a sensitivity analysis. They are not clean
predictors of a prior diagnosis, because the relationship runs in both
directions at once: a person is diagnosed *because* their pressure was high in
the past, but a diagnosis leads to antihypertensive treatment, which *lowers*
the pressure measured today. Today's measurement is therefore partly a
consequence of the outcome. **The file contains no medication variable of any
kind**, so the two directions cannot be separated, and any coefficient for
measured pressure in this model is uninterpretable as a causal quantity.

**BMI is categorical, not continuous.** Only the WHO `bmi_category` label is
supplied, so no per-unit BMI odds ratio can be reported; and in fact **no
predictor in this file is continuous** - age is banded too - so every estimate
is a contrast between broad groups with certain residual confounding within
band.

**Facility type and clustering.** `study_site` is a facility-*type* label
(private, public, faith), not a facility identifier. It is a plausible
confounder for a care-related outcome, since the three sectors serve different
populations and have different case mixes, so it is included. But it is also the
only handle on the fact that participants are clustered within three recruitment
sites; with no facility ID, a random effect or cluster-robust standard errors
are impossible, and the reported standard errors are almost certainly too small.

**Statistical analysis.** Logistic regression via `glm(family = binomial)`,
justified because the outcome is binary; estimates are reported as odds ratios.
We fitted a univariable model for each predictor first, then a single
multivariable model containing all eight. Predictors were chosen in advance on
subject-matter grounds, not by significance. Significance threshold 0.05;
estimates are reported with 95% confidence intervals. Assumptions were checked
with `performance::check_model()`. Analysis used `r R.version.string` with
tidyverse, gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped,
`r fmt(100 * n_dropped / nrow(dat), 1)`%), containing `r n_events` events. The
outcome and four predictors are 100% complete; losses come from `bmi_category`
(`r sum(is.na(dat$bmi_category))` missing), `education`
(`r sum(is.na(dat$education))`), `age_group` (`r sum(is.na(dat$age_group))`) and
`employment` (`r sum(is.na(dat$employment))`). `monthly_income`, missing for
`r sum(is.na(dat$monthly_income))` participants
(`r fmt(100 * mean(is.na(dat$monthly_income)), 1)`%), is not in the main model
despite being substantively relevant to an access-to-care question; the section
*The cost of `monthly_income`* below reports exactly what including it would
cost and gain.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(
    by = prior_htn_diagnosis,
    missing = "no"
  ) |>
  add_n() |>
  add_overall() |>
  modify_caption("**Table 1. Sample characteristics by self-reported previous hypertension diagnosis.**")
```

```{r table-one-numbers}
prev <- 100 * mean(dat_model$prior_htn_diagnosis == "Yes")
```

`r fmt(prev, 1)`% of the analytic sample (`r n_events` of `r nrow(dat_model)`)
reported a previous hypertension diagnosis. Those reporting one are much older
(`r fmt(100 * mean(dat_model$age_group[dat_model$prior_htn_diagnosis == "Yes"] %in% c("56-65", ">65")), 0)`%
over 55 versus
`r fmt(100 * mean(dat_model$age_group[dat_model$prior_htn_diagnosis == "No"] %in% c("56-65", ">65")), 0)`%),
much more likely to report cardiovascular disease
(`r fmt(100 * mean(dat_model$cardiovascular_disease[dat_model$prior_htn_diagnosis == "Yes"] == "yes"), 0)`%
versus
`r fmt(100 * mean(dat_model$cardiovascular_disease[dat_model$prior_htn_diagnosis == "No"] == "yes"), 0)`%)
or diabetes, more often retired, and heavier. Sex differences are slight.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Proportion reporting a previous hypertension diagnosis, by age band and facility type."}
dat_model |>
  mutate(age_group = fct_relevel(age_group, "18-25", "26-35", "36-45",
                                 "46-55", "56-65", ">65")) |>
  group_by(age_group, study_site) |>
  summarise(prop = mean(prior_htn_diagnosis == "Yes"), n = n(), .groups = "drop") |>
  ggplot(aes(age_group, prop, fill = study_site)) +
  geom_col(position = position_dodge(preserve = "single")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 1)) +
  scale_fill_viridis_d(option = "D", end = 0.9) +
  labs(x = "Age band (years)", y = "Previously diagnosed (%)",
       fill = "Facility type",
       caption = "Bars are observed proportions; the 18-25 cells contain fewer than 40 people each")
```

The proportion reporting a previous diagnosis rises steeply with age in all
three sectors, from almost nobody under 25 to well over half of the over-65s,
and the faith-based facility sits above the other two in most age bands - but
its attenders are also older and sicker, which is exactly the confounding the
multivariable model has to resolve.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = glm,
    method.args = list(family = binomial),
    exponentiate = TRUE
  )
uv

utid <- map_dfr(predictors, function(v) {
  tidy(glm(reformulate(v, outcome), data = dat_model, family = binomial),
       conf.int = TRUE, exponentiate = TRUE) |>
    filter(term != "(Intercept)")
})
ug <- function(term, d = 2) {
  r <- utid[utid$term == term, ]
  stopifnot(nrow(r) == 1L)
  sprintf("OR %s (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d), fmt(r$conf.high, d))
}
up <- function(term) { x <- utid$p.value[utid$term == term]; stopifnot(length(x) == 1L); pv(x) }
```

Crudely, the strongest associations are with other diagnosed conditions and with
age. Self-reported cardiovascular disease carries `r ug("cardiovascular_diseaseyes")`
(p `r up("cardiovascular_diseaseyes")`) and self-reported diabetes
`r ug("diabetesyes")` (p `r up("diabetesyes")`). Relative to the 26-35 band, the
odds of reporting a previous diagnosis are `r ug("age_group46-55")` at 46-55,
`r ug("age_group56-65")` at 56-65 and `r ug("age_group>65")` above 65. Retired
participants have `r ug("employmentRetired")` relative to the employed
(p `r up("employmentRetired")`). Obesity carries `r ug("bmi_categoryObese")`
(p `r up("bmi_categoryObese")`) and underweight `r ug("bmi_categoryUnderweight")`
(p `r up("bmi_categoryUnderweight")`). Attenders at the faith-based facility have
`r ug("study_sitefaith")` the odds of those at the public facility
(p `r up("study_sitefaith")`), while private-facility attenders are
indistinguishable from public (`r ug("study_siteprivate")`,
p `r up("study_siteprivate")`). Neither sex (`r ug("sexMale")`,
p `r up("sexMale")`) nor education (tertiary `r ug("educationTertiary")`,
p `r up("educationTertiary")`) shows a crude association.

## Why adjustment is needed

The crude odds ratios are badly confounded because the predictors are strongly
interlocked. **Age band** is the central problem. It is the strongest
determinant of ever having been diagnosed - simply because older people have had
more years and more clinical encounters in which a blood pressure could be
taken - and it is associated with almost every other predictor here. Retirement
is nearly a deterministic function of age
(`r fmt(100 * mean(dat_model$age_group[dat_model$employment == "Retired"] %in% c("56-65", ">65")), 0)`%
of retired participants are over 55), so the crude retirement OR of
`r fmt(utid$estimate[utid$term == "employmentRetired"], 2)` is largely an age
effect wearing a different label. Education is steeply age-patterned in the
opposite direction:
`r fmt(100 * mean(dat_model$education[dat_model$age_group == ">65"] == "Tertiary"), 0)`%
of the over-65s have tertiary education against
`r fmt(100 * mean(dat_model$education[dat_model$age_group == "26-35"] == "Tertiary"), 0)`%
of the 26-35 band, so the crude education contrasts are the *youth* of the
educated cancelling out any access advantage they may have, and only adjustment
for age can separate the two. **Facility type** is the second key confounder,
and the reason it must be in the model: the three sectors serve very different
populations. The faith-based facility's attenders are older and far more likely
to report cardiovascular disease
(`r fmt(100 * mean(dat_model$cardiovascular_disease[dat_model$study_site == "faith"] == "yes"), 0)`%
versus
`r fmt(100 * mean(dat_model$cardiovascular_disease[dat_model$study_site == "public"] == "yes"), 0)`%
at the public facility), so its crude excess of previously diagnosed attenders
may be case mix rather than anything about the facility. **Comorbidity**
(diabetes, cardiovascular disease) confounds the socio-economic contrasts in
turn, because people already in chronic-disease care get their blood pressure
taken repeatedly, and comorbidity is unevenly distributed across age, body size
and facility. Mutual adjustment lets each odds ratio be read as a comparison
between attenders alike on the other seven measured characteristics.

## Multivariable model

```{r multivariable}
fit <- glm(reformulate(predictors, outcome), data = dat_model, family = binomial)

mv <- tbl_regression(fit, exponentiate = TRUE)
mv

tid <- tidy(fit, conf.int = TRUE, exponentiate = TRUE)
g <- function(term, d = 2) {
  r <- tid[tid$term == term, ]
  stopifnot(nrow(r) == 1L)
  sprintf("OR %s (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d), fmt(r$conf.high, d))
}
gp <- function(term) { x <- tid$p.value[tid$term == term]; stopifnot(length(x) == 1L); pv(x) }
```

```{r fit-stats}
n_par <- length(coef(fit))
epp   <- n_events / (n_par - 1)
tibble(
  Statistic = c("Complete-case N", "Rows dropped by listwise deletion",
                "Events (prior diagnosis = Yes)", "AIC", "Null deviance",
                "Residual deviance", "Model parameters (incl. intercept)",
                "Events per parameter"),
  Value = c(as.character(nrow(dat_model)), as.character(n_dropped),
            as.character(n_events), fmt(AIC(fit), 1),
            fmt(fit$null.deviance, 1), fmt(fit$deviance, 1),
            as.character(n_par), fmt(epp, 1))
) |>
  gt() |>
  tab_header(title = md("**Model fit: multivariable logistic model for previous hypertension diagnosis**"))
```

The model was fitted on `r nrow(dat_model)` complete cases with `r n_events`
events and `r n_par` parameters, giving `r fmt(epp, 1)` **events per
parameter** - above the conventional minimum of 10, so the model is identified
and the predictor set does not need trimming. AIC is `r fmt(AIC(fit), 1)`
against a null deviance of `r fmt(fit$null.deviance, 1)` and residual deviance
`r fmt(fit$deviance, 1)`, a large improvement on the null.

Conditional on the other seven predictors, age remains dominant:
`r g("age_group46-55")` at 46-55, `r g("age_group56-65")` at 56-65 and
`r g("age_group>65")` above 65, all relative to the 26-35 band. Self-reported
cardiovascular disease carries an extreme `r g("cardiovascular_diseaseyes")`
(p `r gp("cardiovascular_diseaseyes")`) - see the caution below - and diabetes
`r g("diabetesyes")` (p `r gp("diabetesyes")`). Obesity retains
`r g("bmi_categoryObese")` (p `r gp("bmi_categoryObese")`) and being overweight
`r g("bmi_categoryOverweight")` (p `r gp("bmi_categoryOverweight")`). Education
becomes strongly and positively associated once age is held constant: relative
to no schooling, primary `r g("educationPrimary")`, secondary
`r g("educationSecondary")` and tertiary `r g("educationTertiary")` (all
p < 0.05). Attenders at the faith-based facility have `r g("study_sitefaith")`
the adjusted odds of public-facility attenders (p `r gp("study_sitefaith")`) -
the reverse of the crude direction. Employment and sex are null after
adjustment. These are associations among the attenders observed, not effects of
intervening on any of these variables.

**Caution on the cardiovascular-disease coefficient.** An odds ratio of
`r fmt(tid$estimate[tid$term == "cardiovascular_diseaseyes"], 0)` is not a
plausible causal quantity and rests on a thin cell: only
`r sum(dat_model$cardiovascular_disease == "yes" & dat_model$prior_htn_diagnosis == "No")`
of the `r sum(dat_model$cardiovascular_disease == "yes")` participants reporting
cardiovascular disease do *not* also report a hypertension diagnosis. This is
close to - though not actually - quasi-complete separation, which is why the
confidence interval is so wide. Substantively, cardiovascular disease is not a
clean upstream exposure: it is largely diagnosed in the same clinical encounters
that diagnose hypertension, and it is often a consequence of hypertension, so
this coefficient reflects co-detection and shared care contact at least as much
as any effect of one condition on the other. It is retained because leaving it
out changes conclusions about facility type (see the sensitivity analysis), but
it should be described as a marker of engagement with chronic-disease care and
never as a risk factor.

## The cost of `monthly_income`

```{r income-cost}
fit_inc      <- glm(reformulate(c(predictors, "monthly_income"), outcome),
                    data = dat_inc, family = binomial)
fit_noinc_cc <- glm(reformulate(predictors, outcome), data = dat_inc, family = binomial)

inc_tid <- tidy(fit_inc, conf.int = TRUE, exponentiate = TRUE)
epp_inc <- n_events_inc / (length(coef(fit_inc)) - 1)

tibble(
  Model = c("Main model (no income)",
            "Same model, income complete-cases only",
            "Model with monthly_income added"),
  N = c(nrow(dat_model), nrow(dat_inc), nrow(dat_inc)),
  `% of file` = fmt(100 * c(nrow(dat_model), nrow(dat_inc), nrow(dat_inc)) / nrow(dat), 1),
  Events = c(n_events, n_events_inc, n_events_inc),
  Parameters = c(n_par - 1, length(coef(fit_noinc_cc)) - 1,
                 length(coef(fit_inc)) - 1),
  `Events/parameter` = fmt(c(epp, n_events_inc / (length(coef(fit_noinc_cc)) - 1),
                             epp_inc), 1),
  AIC = fmt(c(AIC(fit), AIC(fit_noinc_cc), AIC(fit_inc)), 1)
) |>
  gt() |>
  tab_header(title = md("**Table 2. What adding `monthly_income` costs.**"),
             subtitle = md("`monthly_income` is missing for 31% of participants; listwise deletion removes them from the whole model, not just from that one coefficient. AICs from different samples are not comparable."))
```

```{r income-numbers}
inc_lost <- nrow(dat_model) - nrow(dat_inc)
inc_p    <- anova(fit_noinc_cc, fit_inc, test = "LRT")$`Pr(>Chi)`[2]
faith_main <- tid$estimate[tid$term == "study_sitefaith"]
faith_inc  <- inc_tid$estimate[inc_tid$term == "study_sitefaith"]
tert_main  <- tid$estimate[tid$term == "educationTertiary"]
tert_inc   <- inc_tid$estimate[inc_tid$term == "educationTertiary"]
cvd_main   <- tid$estimate[tid$term == "cardiovascular_diseaseyes"]
cvd_inc    <- inc_tid$estimate[inc_tid$term == "cardiovascular_diseaseyes"]
miss_by_site <- dat |> group_by(study_site) |>
  summarise(pct = 100 * mean(is.na(monthly_income)))
```

This is the most important methodological point in this dataset, and it bites
hardest here because household income is genuinely relevant to an
access-to-care outcome - this is the one model where a student has a real reason
to want the variable. `monthly_income` is missing for
`r fmt(100 * mean(is.na(dat$monthly_income)), 1)`% of participants, and because
`glm()` deletes rows listwise, adding it removes those participants from *every*
coefficient in the model, not just from the income coefficient. The
complete-case N falls from **`r nrow(dat_model)`** to **`r nrow(dat_inc)`**
(`r fmt(100 * nrow(dat_inc) / nrow(dat), 0)`% of the file) and events from
**`r n_events`** to **`r n_events_inc`**: `r inc_lost` participants and
`r n_events - n_events_inc` events discarded to buy
`r length(coef(fit_inc)) - n_par` extra parameters, cutting events per parameter
from `r fmt(epp, 1)` to `r fmt(epp_inc, 1)`. The likelihood-ratio test comparing
the two models on the *same* `r nrow(dat_inc)` rows gives p `r pv(inc_p)`, so
even on its own restricted sample income does not improve the model at the 0.05
level, although it comes closer here than in any other outcome in this dataset.

The missingness is emphatically not random, and in a way that is fatal for this
particular research question. Non-response to the income question ranges from
`r fmt(miss_by_site$pct[miss_by_site$study_site == "faith"], 0)`% at the
faith-based facility to
`r fmt(miss_by_site$pct[miss_by_site$study_site == "private"], 0)`% at the
private facility, and rises with education (from
`r fmt(100 * mean(is.na(dat$monthly_income[dat$education == "No School"])), 0)`%
among those with no schooling to
`r fmt(100 * mean(is.na(dat$monthly_income[dat$education == "Tertiary"])), 0)`%
among the tertiary educated). In other words, the propensity to withhold income
is itself patterned by two of the model's own predictors - facility type and
education - so restricting to income responders systematically reshapes exactly
the contrasts this analysis is about. The effect is visible: the faith-facility
odds ratio moves from `r fmt(faith_main, 2)` to `r fmt(faith_inc, 2)`, tertiary
education from `r fmt(tert_main, 2)` to `r fmt(tert_inc, 2)`, and the
cardiovascular-disease coefficient from `r fmt(cvd_main, 1)` to
`r fmt(cvd_inc, 1)`. **The main model therefore excludes `monthly_income`**, and
a student who wants an income gradient for this outcome must use multiple
imputation rather than complete cases; taking the reduced sample at face value
would be a mistake.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 3. Unadjusted and adjusted odds ratios.**")
```

This is the most dramatic crude-versus-adjusted table in the dataset. **Facility
type reverses direction.** The faith-based facility looks like the sector with
*more* previously diagnosed attenders crudely (`r ug("study_sitefaith")`,
p `r up("study_sitefaith")`), and becomes the sector with *fewer* after
adjustment (`r g("study_sitefaith")`, p `r gp("study_sitefaith")`): its crude
excess was entirely case mix - older attenders with far more cardiovascular
disease - and once like is compared with like, its attenders are less likely to
have been diagnosed before. **Education reverses too, and more strongly.**
Tertiary education is crudely and non-significantly *protective*
(`r ug("educationTertiary")`, p `r up("educationTertiary")`) and becomes a
strong positive association after adjustment (`r g("educationTertiary")`,
p `r gp("educationTertiary")`); the same happens to secondary
(`r ug("educationSecondary")` to `r g("educationSecondary")`) and primary
(`r ug("educationPrimary")` to `r g("educationPrimary")`). The crude estimates
were suppressed by age: the educated are young, the young are undiagnosed, and
holding age constant reveals a clear educational gradient in ever having been
told you have hypertension - which is the substantive finding of this analysis.
**Retirement collapses**, from `r ug("employmentRetired")` crude
(p `r up("employmentRetired")`) to `r g("employmentRetired")` adjusted
(p `r gp("employmentRetired")`), losing significance entirely: retirement in
these data is essentially a restatement of old age. **Diabetes halves**, from
`r ug("diabetesyes")` to `r g("diabetesyes")`, though it remains significant.
Two predictors barely move: **obesity** (`r ug("bmi_categoryObese")` to
`r g("bmi_categoryObese")`) and **cardiovascular disease**
(`r ug("cardiovascular_diseaseyes")` to `r g("cardiovascular_diseaseyes")`),
the latter simply because nothing measured here can explain away a variable so
tightly bound to the outcome. The **age gradient** attenuates by roughly a third
(over-65s `r ug("age_group>65")` to `r g("age_group>65")`) but stays the largest
demographic signal. Collinearity is low throughout (see diagnostics), so these
reversals are genuine confounding, not variance inflation.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r diag-numbers}
vifs <- check_collinearity(fit)
max_vif <- max(vifs$VIF)
br <- binned_residuals(fit)
br_pct <- 100 * attributes(br)$resid_ok
min_cell <- sum(dat_model$cardiovascular_disease == "yes" &
                  dat_model$prior_htn_diagnosis == "No")
```

For a logistic model only some `check_model()` panels apply. **Binned
residuals**: `r fmt(br_pct, 1)`% of bins fall inside the error bounds against a
rule-of-thumb target of about 95%, so there is genuine lack of fit. As with the
other outcomes in this file the cause is structural - every predictor is a
coarse category, so only a limited number of distinct fitted probabilities can be
produced - and it cannot be repaired without continuous age and BMI.
**Collinearity** is low (maximum VIF `r fmt(max_vif, 2)`, for
`r vifs$Term[which.max(vifs$VIF)]`), which matters a great deal here: it means
the education and facility-type reversals in Table 3 are real confounding and
not artefacts of unstable estimation. **Separation** is the live risk in this
model: the cardiovascular-disease contrast rests on a cell of just
`r min_cell` participants (CVD reported, no prior hypertension diagnosis), which
is why its odds ratio is enormous and its confidence interval spans a factor of
`r fmt(tid$conf.high[tid$term == "cardiovascular_diseaseyes"] / tid$conf.low[tid$term == "cardiovascular_diseaseyes"], 0)`.
The model converges and no coefficient runs away to infinity, so this is
sparse-data bias rather than true separation, but the coefficient should be
reported with the caveat above and never interpreted as a magnitude.
**Influential observations**: no point exceeds the Cook's distance contours.
**Events per parameter** is `r fmt(epp, 1)`, adequate. Homogeneity of variance
and normality of residuals are printed for completeness but do not apply to a
binomial GLM.

## Sensitivity analyses

```{r sens-nocvd}
p_nocvd <- setdiff(predictors, "cardiovascular_disease")
fit_nocvd <- glm(reformulate(p_nocvd, outcome), data = dat_model, family = binomial)
nc <- tidy(fit_nocvd, conf.int = TRUE, exponentiate = TRUE)
ncg <- function(term, d = 2) {
  r <- nc[nc$term == term, ]
  stopifnot(nrow(r) == 1L)
  sprintf("OR %s (95%% CI %s to %s, p %s)", fmt(r$estimate, d), fmt(r$conf.low, d),
          fmt(r$conf.high, d), pv(r$p.value))
}
```

**Dropping cardiovascular disease.** Because that coefficient is so extreme, the
model was refitted without it on the same `r nrow(dat_model)` rows. The facility
result does not survive: the faith-facility odds ratio moves from
`r g("study_sitefaith")` back to `r ncg("study_sitefaith")`, i.e. no difference
from the public facility at all. The education gradient also weakens sharply
(tertiary `r ncg("educationTertiary")`). In other words, **the two headline
findings of this analysis exist only conditional on cardiovascular disease
status**, a variable that is arguably a consequence of the same care contact
that produces the outcome. Adjusting for it may be removing confounding by case
mix, or may be conditioning on a collider and manufacturing associations that do
not exist. This file cannot distinguish the two, and a student report should
present both models rather than choosing.

```{r sens-htn}
dat_h <- dat |> select(all_of(c(outcome, predictors, "hypertension"))) |> drop_na()
fit_h <- glm(reformulate(c(predictors, "hypertension"), outcome),
             data = dat_h, family = binomial)
ht <- tidy(fit_h, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term == "hypertensionYes")
tab <- table(measured = dat$hypertension, prior = dat$prior_htn_diagnosis)
aware <- 100 * tab["Yes", "Yes"] / sum(tab["Yes", ])
```

```{r prior-tab}
as.data.frame.matrix(tab) |>
  rownames_to_column("Measured hypertension") |>
  rename(`Prior diagnosis: No` = No, `Prior diagnosis: Yes` = Yes) |>
  gt() |>
  tab_header(title = md("**Table 4. Measured hypertension against self-reported prior diagnosis.**"))
```

**Adding today's measured hypertension.** The two variables agree on only
`r fmt(100 * (tab["No", "No"] + tab["Yes", "Yes"]) / sum(tab), 0)`% of
participants: just `r fmt(aware, 0)`% of those hypertensive on measurement
report a prior diagnosis, and `r tab["No", "Yes"]` of the
`r sum(tab[, "Yes"])` who report a prior diagnosis are normotensive today. Adding
`hypertension` to the model gives
`r sprintf("OR %s (95%% CI %s to %s)", fmt(ht$estimate, 2), fmt(ht$conf.low, 2), fmt(ht$conf.high, 2))`,
p `r pv(ht$p.value)`, and lowers AIC from `r fmt(AIC(fit), 1)` to
`r fmt(AIC(fit_h), 1)`. **This must not be read as an effect of blood pressure
on diagnosis.** Reverse causation is guaranteed here and runs in both
directions: a diagnosis is given because pressure was high in the *past*, which
inflates the association; and the diagnosis brings antihypertensive treatment
which lowers the pressure measured *today*, which deflates it. The file records
**no medication whatsoever**, so the two cannot be untangled, and the observed
OR of `r fmt(ht$estimate, 1)` is a net of the two opposing mechanisms with no
way to recover either. That the OR is only about
`r fmt(ht$estimate, 0)` - rather than the very large value one would expect if
diagnosis simply tracked pressure - is itself informative: it is consistent with
substantial treatment effect among the diagnosed, but it is equally consistent
with poor recall or with single-visit measurement error, and these data cannot
choose between those explanations.

# Discussion

Among these Kenyan outpatient attenders, having ever been told you have
hypertension is driven by age, by body size, by other diagnosed chronic
conditions and - once age is held constant - by education. The educational
gradient is the substantive finding and it is invisible in the crude analysis:
tertiary-educated attenders have `r g("educationTertiary")` the odds of
attenders who never went to school, of the same age, sex, employment status,
body size, comorbidity and facility type, whereas crudely they appear
marginally *less* likely to have been diagnosed. The reason is that education in
Kenya is heavily age-patterned, so the crude comparison is mostly a comparison
of young with old. Practically, this points to schooling - a proxy for health
literacy and for the ability to seek and use services - as an axis of inequality
in diagnosis, not merely in treatment. The awareness gap is the other headline:
only `r fmt(aware, 0)`% of participants who were hypertensive at this visit had
ever been told so.

**Limitations.** The outcome is **self-reported** and depends on recall and on
having understood what a clinician said; misclassification is certain and is
probably differential by education, which is the very predictor of interest -
better-educated participants may be better at recalling and reporting a
diagnosis, so part of the education gradient reported here may be reporting
behaviour rather than access. This is a **facility-based convenience sample**,
so it describes attenders at three specific facilities and nothing wider; by
construction everyone in it is already in contact with a health facility today,
which truncates the access spectrum this analysis is meant to describe. The
design is **cross-sectional**: comorbidity, body size and facility attendance
are all recorded at the same visit as the outcome, so temporal ordering is
assumed, not observed. The **cardiovascular-disease coefficient is
uninterpretable as a magnitude** and, as the sensitivity analysis shows, both
headline findings depend on adjusting for it; if it is a collider rather than a
confounder, those findings are artefacts. **No medication variable exists**,
which is a serious gap for any analysis linking diagnosis to measured pressure.
**No facility identifier exists** either, only the three-level facility type, so
clustering of attenders within facilities cannot be modelled and the standard
errors are too small. **`monthly_income` was excluded** because including it
would cost `r inc_lost` rows and `r n_events - n_events_inc` events, so household
economic position - arguably the most important determinant of access to
diagnosis - is simply unadjusted for, and its missingness is itself patterned by
facility and education. **No predictor is continuous**, so age and adiposity
effects are cell contrasts with residual confounding within band. Insurance
status, distance to facility, and the reason for today's visit are all absent,
and all three would plausibly dominate an access analysis.

**Future work.** Repeat with insurance status, distance to facility and reason
for attendance, a facility identifier for clustering, medication use, and
multiple imputation for income - none of which this extract carries.

# Feasibility notes

*NOT part of a student report.*

- **The most interesting outcome in this dataset, and the most fragile.** It is
  about access to and contact with care rather than physiology, it is fully
  observed (`r sum(is.na(dat_raw$prior_htn_diagnosis))` missing), it has
  `r n_events` events in `r nrow(dat_model)` complete cases and
  `r fmt(epp, 1)` events per parameter, and it produces the most striking
  crude-versus-adjusted table in the catalogue - two clean sign reversals
  (facility type `r ug("study_sitefaith")` to `r g("study_sitefaith")`; tertiary
  education `r ug("educationTertiary")` to `r g("educationTertiary")`) with
  maximum VIF only `r fmt(max_vif, 2)`, so they are real confounding.
- **MAJOR problem: both headline results depend on `cardiovascular_disease`.**
  Drop it and the faith-facility effect becomes `r ncg("study_sitefaith")` and
  tertiary education `r ncg("educationTertiary")` - both null. CVD is plausibly
  a collider (diagnosed in the same care contact as hypertension) rather than a
  confounder, so the analysis may be manufacturing its own findings. A student
  must be told to fit both models. This is the biggest threat to the outcome.
- **Near-sparse cell driving an OR of `r fmt(cvd_main, 0)`.** Only
  `r min_cell` of `r sum(dat_model$cardiovascular_disease == "yes")` participants
  reporting cardiovascular disease do not also report a hypertension diagnosis.
  Not formal separation - the model converges - but the CI spans a factor of
  `r fmt(tid$conf.high[tid$term == "cardiovascular_diseaseyes"] / tid$conf.low[tid$term == "cardiovascular_diseaseyes"], 0)`
  and the magnitude is meaningless.
- **Trap: measured `hypertension` / `systolic_bp` / `diastolic_bp` as
  predictors.** Not forbidden the way they are for the `hypertension` outcome -
  they are genuinely different constructs, agreeing on only
  `r fmt(100 * (tab["No", "No"] + tab["Yes", "Yes"]) / sum(tab), 0)`% of rows -
  but the relationship is bidirectional (diagnosis follows past high pressure;
  treatment lowers present pressure) and there is **no medication variable** to
  break the loop. Students will report the
  `r sprintf("OR %s", fmt(ht$estimate, 2))` as "high blood pressure causes
  diagnosis" unless explicitly warned.
- **THE missing-data teaching case, and it is worst here.** `monthly_income` is
  `r fmt(100 * mean(is.na(dat$monthly_income)), 1)`% missing; including it drops
  N from `r nrow(dat_model)` to `r nrow(dat_inc)`
  (`r fmt(100 * nrow(dat_inc) / nrow(dat), 0)`% of the file) and events from
  `r n_events` to `r n_events_inc` (LRT p `r pv(inc_p)`). Unlike the other three
  outcomes, income is *substantively central* to this research question, so the
  student genuinely wants it - and its missingness is patterned by facility
  (`r fmt(miss_by_site$pct[miss_by_site$study_site == "private"], 0)`% private vs
  `r fmt(miss_by_site$pct[miss_by_site$study_site == "faith"], 0)`% faith) and by
  education, i.e. by two of the model's own predictors. Best missing-data
  discussion in the catalogue; multiple imputation is the only honest route.
- **`study_site` is a facility *type*, not a facility ID.** Three sites, no
  identifier, so clustering cannot be modelled at all and standard errors are
  optimistic. It doubles as both a substantive predictor and the only (crude)
  handle on clustering, which students should be told is unsatisfactory.
- **Self-report outcome with differential misclassification by education** -
  the same variable the analysis reports as its main finding. Must be discussed,
  not hidden.
- **Binned residuals fail (`r fmt(br_pct, 1)`% of bins inside bounds)**, for the
  same structural reason as the other outcomes: no continuous predictors exist
  in this file, so only a handful of fitted probabilities are possible. Not
  repairable here.
- **`employment` is nearly collinear with age in substance if not in VIF**
  (`r fmt(100 * mean(dat_model$age_group[dat_model$employment == "Retired"] %in% c("56-65", ">65")), 0)`%
  of retired participants are over 55); its crude OR of
  `r fmt(utid$estimate[utid$term == "employmentRetired"], 2)` collapsing to
  `r fmt(tid$estimate[tid$term == "employmentRetired"], 2)` is a good teaching
  example but will confuse students who chose it as an exposure.
- **`current_smoker` was left out** of this model deliberately: only
  `r sum(dat_raw$current_smoker == 1)` smokers in the file, no substantive
  reason to expect it to predict *diagnosis*, and the events-per-parameter
  budget is better spent elsewhere.
