---
title: "Systolic blood pressure in Kenyan outpatient attenders"
subtitle: "Dataset 4: Kenya hypertension screening survey"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "kenya_hypertension.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "systolic_bp"
predictors <- c("age_group", "sex", "education", "bmi_category",
                "current_smoker", "current_alcohol", "adequate_activity",
                "diabetes")

lab <- function(x, l) { attr(x, "label") <- l; x }

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest band, or the clinically
    # "unexposed" group. 26-35 rather than 18-25 is the age reference because
    # 18-25 holds only 101 people and would make every age contrast imprecise.
    age_group = lab(fct_relevel(factor(age_group),
                                "26-35", "18-25", "36-45", "46-55", "56-65", ">65"),
                    "Age band (years)"),
    sex = lab(factor(sex, levels = c("female", "male"), labels = c("Female", "Male")),
              "Sex"),
    # No schooling is the reference so higher categories read as "more education
    # than the reference" and the gradient can be read straight down the column.
    education = lab(fct_relevel(factor(education),
                                "No School", "Primary", "Secondary", "Tertiary"),
                    "Education"),
    # Healthy weight is both the largest category and the clinical comparator.
    bmi_category = lab(fct_relevel(factor(bmi_category),
                                   "Healthy", "Underweight", "Overweight", "Obese"),
                       "BMI category (WHO)"),
    current_smoker = lab(factor(current_smoker, levels = c(0, 1), labels = c("No", "Yes")),
                         "Current smoker"),
    current_alcohol = lab(factor(current_alcohol, levels = c(0, 1), labels = c("No", "Yes")),
                          "Current alcohol use"),
    adequate_activity = lab(factor(adequate_activity, levels = c(0, 1), labels = c("No", "Yes")),
                            "Meets activity recommendation"),
    # "unknown" is kept as its own level rather than set to missing: it is a
    # substantive answer (never tested) and dropping it would cost 261 rows.
    diabetes = lab(fct_relevel(factor(diabetes), "no", "yes", "unknown"),
                   "Self-reported diabetes"),
    cardiovascular_disease = lab(fct_relevel(factor(cardiovascular_disease),
                                             "no", "yes", "unknown"),
                                 "Self-reported cardiovascular disease"),
    study_site = lab(fct_relevel(factor(study_site), "public", "private", "faith"),
                     "Facility type"),
    monthly_income = lab(fct_relevel(factor(monthly_income),
                                     "0  - 15,000", "16,000 - 50,000",
                                     "51,000 - 100,000", "100,000 - 200,000",
                                     "Above 200,000"),
                         "Monthly household income (KSh)"),
    prior_htn_diagnosis = lab(factor(prior_htn_diagnosis, levels = c(0, 1),
                                     labels = c("No", "Yes")),
                              "Previously diagnosed with hypertension"),
    systolic_bp  = lab(systolic_bp, "Systolic BP (mmHg)"),
    diastolic_bp = lab(diastolic_bp, "Diastolic BP (mmHg)")
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

# The income-augmented sample is carried through the whole document so that the
# cost of adding a 31%-missing variable can be quantified rather than asserted.
dat_inc <- dat |> select(all_of(c(outcome, predictors, "monthly_income"))) |> drop_na()

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
pv  <- function(p) ifelse(p < 0.001, "< 0.001", paste0("= ", fmt(p, 3)))
```

# Introduction

Hypertension is the leading modifiable cardiovascular risk factor in sub-Saharan
Africa, and Kenya's national NCD strategy relies on opportunistic screening of
people already attending health facilities rather than on population surveys.
This dataset is a cross-sectional screening survey of `r nrow(dat_raw)` adults
recruited at three types of outpatient facility in Kenya - a government (public)
facility, a private facility and a faith-based mission facility - in which every
participant had blood pressure measured three times and completed a WHO
STEPS-style questionnaire on demography, behaviour and self-reported morbidity.
Systolic pressure is modelled here as a continuous quantity because it is the
measurement that drives stroke and ischaemic heart disease risk, and
dichotomising it discards information.

**Aim.** Among these Kenyan outpatient attenders, we examined which of age band,
sex, education, BMI category, current smoking, current alcohol use, physical
activity and self-reported diabetes are independently associated with average
systolic blood pressure, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** A cross-sectional facility-based screening survey of
`r nrow(dat_raw)` adults aged 18 and over attending outpatient services at three
Kenyan facilities (`r sum(dat$study_site == "public")` public,
`r sum(dat$study_site == "private")` private,
`r sum(dat$study_site == "faith")` faith-based). Blood pressure was taken three
times per participant and the file carries the average of the three readings.

**Variables.** The outcome is `systolic_bp`, the mean of three systolic readings
in mmHg, treated as continuous (observed range
`r fmt(min(dat$systolic_bp), 0)` to `r fmt(max(dat$systolic_bp), 0)` mmHg).
Predictors are age band (reference *26-35*, the largest band); sex (reference
*Female*); education (reference *No School*); WHO BMI category (reference
*Healthy*); current smoking (reference *No*); current alcohol use (reference
*No*); meeting the physical-activity recommendation (reference *No*); and
self-reported diabetes (reference *no*, with *unknown* retained as a substantive
third level rather than treated as missing).

**BMI is categorical, not continuous.** The file contains only the WHO
`bmi_category` label, not the underlying height, weight or BMI value. It is
therefore impossible to report a per-unit BMI effect (mmHg per kg/m2) from these
data; every adiposity result below is a contrast between broad WHO bands, and
the bands are wide enough that residual confounding by adiposity within band is
certain. In fact **no predictor in this file is continuous** - even age is
supplied only as a six-level band - so this linear model is a cell-means
(ANOVA-style) model throughout and no slope-per-unit statement of any kind can
be made.

**Variables deliberately excluded.** `hypertension` is *not* an independent
variable: it is defined as measured average blood pressure crossing the
diagnostic threshold, so it is a deterministic function of the outcome and
`diastolic_bp`. Regressing systolic pressure on it would be regressing the
outcome on a coarsened copy of itself. `diastolic_bp` is excluded for the same
reason - it is the other half of the same measurement, correlates with systolic
pressure at r = `r fmt(cor(dat$systolic_bp, dat$diastolic_bp), 2)`, and is a
co-outcome rather than an exposure. `prior_htn_diagnosis` is excluded from the
main model and examined separately in a sensitivity analysis, because people who
already carry a diagnosis are likely to be on treatment, which *lowers* measured
pressure; that variable therefore sits downstream of the outcome as much as
upstream of it. `monthly_income` is excluded from the main model on missing-data
grounds and its cost is quantified below.

**Statistical analysis.** `lm()`, justified because the outcome is a continuous,
roughly unimodal measurement. We fitted a univariable model for each predictor
first, then a single multivariable model containing all eight. Predictors were
chosen in advance on subject-matter grounds, not by significance. Significance
threshold 0.05; estimates are reported with 95% confidence intervals.
Assumptions were checked with `performance::check_model()`, and because
non-constant variance was flagged we additionally report HC3
heteroscedasticity-consistent standard errors. Analysis used
`r R.version.string` with tidyverse, gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped,
`r fmt(100 * n_dropped / nrow(dat), 1)`%). Missingness in this file is
extremely uneven. The outcome and four of the predictors are 100% complete; the
losses come from `bmi_category`
(`r sum(is.na(dat$bmi_category))` missing), `education`
(`r sum(is.na(dat$education))`) and `age_group`
(`r sum(is.na(dat$age_group))`). The one seriously incomplete variable is
`monthly_income`, missing for `r sum(is.na(dat$monthly_income))` participants
(`r fmt(100 * mean(is.na(dat$monthly_income)), 1)`%), which is why it is not in
the main model; the section *The cost of `monthly_income`* below reports what
including it would do.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(missing = "no") |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics, complete-case analytic sample.**")
```

```{r table-one-numbers}
med_sbp <- median(dat_model$systolic_bp)
iqr_sbp <- quantile(dat_model$systolic_bp, c(0.25, 0.75))
pct_over <- 100 * mean(dat_model$bmi_category %in% c("Overweight", "Obese"))
```

Median systolic pressure is `r fmt(med_sbp, 1)` mmHg (IQR
`r fmt(iqr_sbp[1], 1)` to `r fmt(iqr_sbp[2], 1)`), with a long right tail
reaching `r fmt(max(dat_model$systolic_bp), 0)` mmHg. The sample is
predominantly female (`r fmt(100 * mean(dat_model$sex == "Female"), 0)`%) and
better educated than the Kenyan adult population
(`r fmt(100 * mean(dat_model$education == "Tertiary"), 0)`% tertiary), which is
a facility-attender rather than a community sample.
`r fmt(pct_over, 0)`% are overweight or obese, while current smoking
(`r fmt(100 * mean(dat_model$current_smoker == "Yes"), 1)`%) and current alcohol
use (`r fmt(100 * mean(dat_model$current_alcohol == "Yes"), 1)`%) are rare.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Systolic blood pressure by age band and WHO BMI category. Boxes are medians and IQRs; points are individual participants."}
ggplot(dat_model, aes(age_group, systolic_bp, fill = bmi_category)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.85, position = position_dodge(preserve = "single")) +
  geom_point(position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75),
             alpha = 0.18, size = 0.7, show.legend = FALSE) +
  scale_fill_viridis_d(option = "D", end = 0.9) +
  labs(x = "Age band (years)", y = "Systolic blood pressure (mmHg)",
       fill = "BMI category",
       caption = "Underweight cells are small (n = 83 in the whole file) and their boxes should not be over-read")
```

Systolic pressure rises across successive age bands within every BMI category,
and within each age band the obese sit above the healthy-weight group, so the
two gradients are at least partly independent; the spread of pressures also
widens noticeably in the older bands, which is the first hint of the
non-constant variance confirmed in the diagnostics.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv

utid <- map_dfr(predictors, function(v) {
  tidy(lm(reformulate(v, outcome), data = dat_model), conf.int = TRUE) |>
    filter(term != "(Intercept)")
})
ug <- function(term, d = 2) {
  r <- utid[utid$term == term, ]
  stopifnot(nrow(r) == 1L)
  sprintf("%s mmHg (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d), fmt(r$conf.high, d))
}
up <- function(term) { x <- utid$p.value[utid$term == term]; stopifnot(length(x) == 1L); pv(x) }
```

Age dominates. Relative to the 26-35 band, participants over 65 average
`r ug("age_group>65")` higher systolic pressure (p `r up("age_group>65")`), with
a monotonic gradient through the intermediate bands
(`r ug("age_group46-55")` at 46-55, `r ug("age_group56-65")` at 56-65). Obesity
is the next largest crude signal: obese participants average
`r ug("bmi_categoryObese")` higher pressure than healthy-weight participants
(p `r up("bmi_categoryObese")`), and underweight participants
`r ug("bmi_categoryUnderweight")` (p `r up("bmi_categoryUnderweight")`). Men
average `r ug("sexMale")` higher than women (p `r up("sexMale")`).
Self-reported diabetes carries `r ug("diabetesyes")` crudely
(p `r up("diabetesyes")`). Education is non-monotonic in the crude analysis:
primary schooling sits `r ug("educationPrimary")` above no schooling
(p `r up("educationPrimary")`) while tertiary sits `r ug("educationTertiary")`
(p `r up("educationTertiary")`). Smoking, alcohol and physical activity show no
crude association.

## Why adjustment is needed

The crude estimates above are not comparable with one another because the
predictors are strongly interlocked in this sample. **Age band** is the obvious
confounder: it is the strongest determinant of systolic pressure through
arterial stiffening, and it is also associated with almost every other predictor
here - diabetes prevalence rises steeply with age
(`r fmt(100 * mean(dat_model$diabetes[dat_model$age_group == ">65"] == "yes"), 0)`%
of the over-65s report diabetes versus
`r fmt(100 * mean(dat_model$diabetes[dat_model$age_group == "26-35"] == "yes"), 0)`%
of the 26-35 band), obesity accumulates with age, and education is far lower in
the older bands
(`r fmt(100 * mean(dat_model$education[dat_model$age_group == ">65"] == "No School"), 0)`%
of over-65s never attended school versus
`r fmt(100 * mean(dat_model$education[dat_model$age_group == "26-35"] == "No School"), 0)`%
of 26-35s). The crude diabetes coefficient of about
`r fmt(utid$estimate[utid$term == "diabetesyes"], 1)` mmHg is therefore in large
part an age difference between people with and without diabetes, and the crude
"protective" tertiary-education coefficient is in large part the youth of the
tertiary-educated. **Sex** confounds both behavioural exposures: smoking is
almost entirely male in this sample
(`r sum(dat_model$current_smoker == "Yes" & dat_model$sex == "Male")` of
`r sum(dat_model$current_smoker == "Yes")` smokers are men) and so is alcohol
use, while men also have higher pressure at a given age, so the crude smoking
and alcohol coefficients are partly sex coefficients. **BMI category** confounds
the diabetes and activity contrasts in turn, since adiposity causes both insulin
resistance and raised pressure. Mutual adjustment lets each coefficient be read
as a comparison between participants who are alike on the other seven measured
characteristics.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv

tid <- tidy(fit, conf.int = TRUE)
g <- function(term, d = 2) {
  r <- tid[tid$term == term, ]
  stopifnot(nrow(r) == 1L)
  sprintf("%s mmHg (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d), fmt(r$conf.high, d))
}
gp <- function(term) { x <- tid$p.value[tid$term == term]; stopifnot(length(x) == 1L); pv(x) }
```

```{r fit-stats}
s <- summary(fit)
tibble(
  Statistic = c("Complete-case N", "Rows dropped by listwise deletion",
                "Parameters (excl. intercept)", "R-squared",
                "Adjusted R-squared", "Residual SE (mmHg)", "AIC"),
  Value = c(as.character(nrow(dat_model)),
            as.character(n_dropped),
            as.character(length(coef(fit)) - 1L),
            fmt(s$r.squared, 3), fmt(s$adj.r.squared, 3), fmt(s$sigma, 2),
            fmt(AIC(fit), 1))
) |>
  gt() |>
  tab_header(title = md("**Model fit: multivariable linear model for systolic BP**"))
```

The model was fitted on `r nrow(dat_model)` complete cases and explains
`r fmt(100 * s$r.squared, 1)`% of the variance in systolic pressure (adjusted
R-squared `r fmt(s$adj.r.squared, 3)`, residual SE `r fmt(s$sigma, 1)` mmHg).
The gap between R-squared and adjusted R-squared is small, so the
`r length(coef(fit)) - 1L` parameters are not being bought at a meaningful cost
in overfitting. Among participants of the same sex, education, BMI category,
smoking, drinking, activity and diabetes status, the over-65s average
`r g("age_group>65")` higher systolic pressure than the 26-35 band
(p `r gp("age_group>65")`), and the 46-55 band `r g("age_group46-55")`
(p `r gp("age_group46-55")`). Obesity retains `r g("bmi_categoryObese")`
(p `r gp("bmi_categoryObese")`) and underweight `r g("bmi_categoryUnderweight")`
(p `r gp("bmi_categoryUnderweight")`). Men average `r g("sexMale")` higher than
women of the same age and body size (p `r gp("sexMale")`). Diabetes is
essentially null after adjustment (`r g("diabetesyes")`,
p `r gp("diabetesyes")`). These are associations among the participants
observed, not effects of intervening on any of these variables.

## The cost of `monthly_income`

```{r income-cost}
fit_inc      <- lm(reformulate(c(predictors, "monthly_income"), outcome), data = dat_inc)
fit_noinc_cc <- lm(reformulate(predictors, outcome), data = dat_inc)

inc_tid <- tidy(fit_inc, conf.int = TRUE)
si  <- summary(fit_inc)

tibble(
  Model = c("Main model (no income)",
            "Same model, income complete-cases only",
            "Model with monthly_income added"),
  N = c(nrow(dat_model), nrow(dat_inc), nrow(dat_inc)),
  `% of file` = fmt(100 * c(nrow(dat_model), nrow(dat_inc), nrow(dat_inc)) / nrow(dat), 1),
  `Parameters` = c(length(coef(fit)) - 1L, length(coef(fit_noinc_cc)) - 1L,
                   length(coef(fit_inc)) - 1L),
  `R-squared` = fmt(c(s$r.squared, summary(fit_noinc_cc)$r.squared, si$r.squared), 3),
  `Adj. R-squared` = fmt(c(s$adj.r.squared, summary(fit_noinc_cc)$adj.r.squared,
                           si$adj.r.squared), 3)
) |>
  gt() |>
  tab_header(title = md("**Table 2. What adding `monthly_income` costs.**"),
             subtitle = md("`monthly_income` is missing for 31% of participants; listwise deletion removes them from the whole model, not just from that one coefficient."))
```

```{r income-numbers}
inc_lost   <- nrow(dat_model) - nrow(dat_inc)
inc_p      <- anova(fit_noinc_cc, fit_inc)$`Pr(>F)`[2]
age65_main <- tid$estimate[tid$term == "age_group>65"]
age65_inc  <- inc_tid$estimate[inc_tid$term == "age_group>65"]
ob_main    <- tid$estimate[tid$term == "bmi_categoryObese"]
ob_inc     <- inc_tid$estimate[inc_tid$term == "bmi_categoryObese"]
miss_by_site <- dat |> group_by(study_site) |>
  summarise(pct = 100 * mean(is.na(monthly_income)))
```

This is the single most important methodological point in this dataset.
`monthly_income` is missing for
`r fmt(100 * mean(is.na(dat$monthly_income)), 1)`% of participants, and because
`lm()` deletes rows listwise, adding it removes those participants from *every*
coefficient in the model, not only from the income coefficient. The
complete-case N falls from **`r nrow(dat_model)`** to **`r nrow(dat_inc)`**
(`r fmt(100 * nrow(dat_inc) / nrow(dat), 0)`% of the file): `r inc_lost` further
participants are discarded, at a cost of `r length(coef(fit_inc)) - length(coef(fit))`
extra parameters. Income adds essentially nothing in exchange - the F-test
comparing the two models on the *same* `r nrow(dat_inc)` rows gives
p `r pv(inc_p)`, and adjusted R-squared moves from
`r fmt(summary(fit_noinc_cc)$adj.r.squared, 3)` to `r fmt(si$adj.r.squared, 3)`.

Worse, the missingness is clearly not random. Non-response to the income
question ranges from
`r fmt(miss_by_site$pct[miss_by_site$study_site == "faith"], 0)`% at the
faith-based facility to
`r fmt(miss_by_site$pct[miss_by_site$study_site == "private"], 0)`% at the
private facility, and rises with education (from
`r fmt(100 * mean(is.na(dat$monthly_income[dat$education == "No School"])), 0)`%
among those with no schooling to
`r fmt(100 * mean(is.na(dat$monthly_income[dat$education == "Tertiary"])), 0)`%
among the tertiary educated). The people who decline to state their income are
therefore systematically different from those who answer, so the reduced sample
is not a random subsample and complete-case estimates from it are not guaranteed
to be unbiased. The main coefficients happen to be fairly stable (the over-65
contrast moves from `r fmt(age65_main, 1)` to `r fmt(age65_inc, 1)` mmHg and
obesity from `r fmt(ob_main, 1)` to `r fmt(ob_inc, 1)` mmHg), which is
reassuring but is a result, not something that could have been assumed. **The
main model therefore excludes `monthly_income`**, and any student who wants the
income gradient must either accept a one-third smaller sample or use multiple
imputation.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 3. Unadjusted and adjusted estimates.**")
```

Two coefficients move a long way toward the null. **Self-reported diabetes**
falls from `r ug("diabetesyes")` crude (p `r up("diabetesyes")`) to
`r g("diabetesyes")` adjusted (p `r gp("diabetesyes")`), a drop of about
`r fmt(100 * (1 - tid$estimate[tid$term == "diabetesyes"] / utid$estimate[utid$term == "diabetesyes"]), 0)`%,
and loses significance entirely: essentially the whole crude diabetes-blood
pressure association in these data is the age and adiposity difference between
participants with and without diabetes, exactly as anticipated. **Tertiary
education** behaves the same way, moving from `r ug("educationTertiary")`
(p `r up("educationTertiary")`, nominally protective) to
`r g("educationTertiary")` (p `r gp("educationTertiary")`, indistinguishable
from no schooling) once age is held constant - the crude "benefit" of tertiary
education was the youth of the tertiary-educated. **Current smoking** moves in
the opposite direction and changes sign, from `r ug("current_smokerYes")` crude to
`r g("current_smokerYes")` adjusted; smokers here are almost all men, so the crude
estimate was carrying the male sex difference, and once sex is adjusted for
nothing is left. Two predictors barely move at all: the **over-65 age contrast**
goes from `r ug("age_group>65")` to `r g("age_group>65")` and **obesity** from
`r ug("bmi_categoryObese")` to `r g("bmi_categoryObese")`, which is what one
expects of variables that confound the others but are themselves confounded by
little that is measured here. One coefficient crosses *into* significance:
**secondary education** moves from `r ug("educationSecondary")`
(p `r up("educationSecondary")`) to `r g("educationSecondary")`
(p `r gp("educationSecondary")`), a precision gain rather than a change of
substance, and given the non-monotonic education pattern it should not be
over-interpreted.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r het-check}
het <- check_heteroscedasticity(fit)
print(het)
vifs <- check_collinearity(fit)
max_vif <- max(vifs$VIF)
```

```{r robust-se}
rob   <- lmtest::coeftest(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
robci <- lmtest::coefci(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
tibble(
  Term = rownames(rob),
  Estimate = fmt(rob[, 1], 2),
  `SE (OLS)` = fmt(summary(fit)$coefficients[, 2], 3),
  `SE (HC3)` = fmt(rob[, 2], 3),
  `95% CI (HC3)` = paste0(fmt(robci[, 1], 2), " to ", fmt(robci[, 2], 2)),
  `p (HC3)` = fmt(rob[, 4], 3)
) |>
  filter(Term != "(Intercept)") |>
  gt() |>
  tab_header(title = md("**Table 4. HC3 heteroscedasticity-consistent standard errors.**"))
```

```{r het-numbers}
worst <- which.max(abs(rob[, 2] - summary(fit)$coefficients[, 2]) /
                     summary(fit)$coefficients[, 2])
worst_term <- rownames(rob)[worst]
sig_ols <- summary(fit)$coefficients[, 4] < 0.05
sig_hc3 <- rob[, 4] < 0.05
n_flip  <- sum(sig_ols != sig_hc3)
```

The posterior-predictive panel shows the model captures the centre of the
distribution but cannot reproduce the long right tail: participants above about
170 mmHg are systematically under-predicted, which is unavoidable when every
predictor is a coarse category. The residual-versus-fitted panel fans out
clearly toward higher fitted values - the spread of systolic pressure genuinely
grows with its mean - and `check_heteroscedasticity()` confirms non-constant
error variance (p < 0.001). The normal-QQ panel shows a heavy right tail
consistent with the same high-pressure participants; with `r nobs(fit)`
observations this is not a threat to the coefficient estimates themselves, only
to small-sample normal-theory inference. The "linearity" panel is close to
uninformative here because there is no continuous predictor: every fitted value
is a cell mean, so the reference line simply traces the cell structure.
Collinearity is a non-issue (maximum VIF `r fmt(max_vif, 2)`, for
`r vifs$Term[which.max(vifs$VIF)]`), and no point exceeds the Cook's distance
contours, so no single participant drives the fit.

Because non-constant variance was detected, Table 4 reports HC3 robust standard
errors. Point estimates are unchanged by construction; only the uncertainty is
re-estimated. The HC3 standard errors differ from the OLS ones by at most
`r fmt(100 * max(abs(rob[, 2] - summary(fit)$coefficients[, 2]) / summary(fit)$coefficients[, 2]), 0)`%
(for `r worst_term`), and `r if (n_flip == 0) "no predictor changes its significance status" else paste(n_flip, "coefficients change significance status")`
under robust inference. The substantive conclusions - age, obesity and sex in,
diabetes and behaviour out - are unaffected by the variance assumption.

## Sensitivity analysis: prior diagnosis and treatment

```{r prior-sens}
dat_prior <- dat |>
  select(all_of(c(outcome, predictors, "prior_htn_diagnosis"))) |> drop_na()
fit_prior <- lm(reformulate(c(predictors, "prior_htn_diagnosis"), outcome),
                data = dat_prior)
pt <- tidy(fit_prior, conf.int = TRUE) |> filter(term == "prior_htn_diagnosisYes")
sbp_dx  <- mean(dat$systolic_bp[dat$prior_htn_diagnosis == "Yes"])
sbp_ndx <- mean(dat$systolic_bp[dat$prior_htn_diagnosis == "No"])
```

Participants who report a previous hypertension diagnosis average
`r fmt(sbp_dx, 1)` mmHg against `r fmt(sbp_ndx, 1)` mmHg in those who do not,
and adding `prior_htn_diagnosis` to the model gives an adjusted difference of
`r fmt(pt$estimate, 1)` mmHg (95% CI `r fmt(pt$conf.low, 1)` to
`r fmt(pt$conf.high, 1)`, p `r pv(pt$p.value)`) with R-squared rising to
`r fmt(summary(fit_prior)$r.squared, 3)`. **This coefficient must not be read as
an effect of diagnosis on blood pressure.** It runs in both directions at once:
being diagnosed is a consequence of having had high pressure in the past
(pushing the coefficient up), while diagnosis leads to antihypertensive
treatment which lowers measured pressure (pushing it down). The file contains
**no medication variable at all**, so the two cannot be separated, and the
observed positive coefficient is best read as evidence that treatment in this
population is not bringing pressure back to the level of the undiagnosed. It is
reported here as a sensitivity analysis and excluded from the main model.

# Discussion

In this Kenyan outpatient sample, age band, BMI category and sex are the
characteristics independently associated with systolic blood pressure. Adjusted
for the other seven predictors, the over-65s average
`r g("age_group>65")` higher pressure than 26-35 year olds and obese
participants `r g("bmi_categoryObese")` higher than healthy-weight
participants. The crude association with self-reported diabetes, which looked
substantial and highly significant, is almost entirely explained by the age and
adiposity differences between participants with and without diabetes.
Practically, this supports age and body size as the two variables that should
drive who is offered opportunistic screening in these facilities, and it warns
that a crude comparison of pressure across diabetes status or education will
badly mislead.

**Limitations.** This is a **facility-based convenience sample** of people
attending outpatient clinics, not a community sample: attenders are sicker and,
here, far better educated than the Kenyan adult population, so no prevalence or
mean from this file can be generalised to Kenya. The design is
**cross-sectional**, so nothing establishes temporality and no causal claim is
warranted. **No predictor is continuous**: age is banded and BMI is supplied
only as a WHO category, so residual confounding within band is guaranteed and no
per-unit effect can be estimated. **Antihypertensive medication is not
recorded**, which is the single most serious unmeasured confounder in an
analysis of measured blood pressure - treated participants have lower measured
pressure precisely because their underlying risk is higher, which biases the
diabetes, obesity and age coefficients toward the null by an unknown amount.
Participants were recruited from **three facilities but no facility identifier
is present** beyond the three-level `study_site` label, so the correlation
between participants attending the same facility cannot be modelled and the
standard errors reported here, robust or not, are likely too small.
Complete-case analysis dropped `r n_dropped` rows; `monthly_income` was excluded
because including it would have cost a further `r nrow(dat_model) - nrow(dat_inc)`
rows for no gain, but that means household economic position is simply
unadjusted for. Sodium intake, family history, kidney function and stress are
all absent. Finally, `r fmt(100 * s$r.squared, 0)`% explained variance means
that most of the variation in systolic pressure in this sample is not captured
by any measured characteristic.

**Future work.** Refit with continuous age and measured BMI, a facility
identifier for a mixed-effects model, and a medication variable, none of which
this extract carries.

# Feasibility notes

*NOT part of a student report.*

- **Works well as a student project.** Continuous, fully observed outcome
  (`r sum(is.na(dat$systolic_bp))` missing values), `r nrow(dat_model)`
  complete cases, one unambiguous dominant signal (age) plus a second clean one
  (BMI category), and at least two predictors that collapse on adjustment. That
  is exactly the Purpose 3 teaching shape.
- **FATAL trap: `hypertension` must never be a predictor here.** `hypertension`
  is defined *from* the measured average blood pressure. Putting it on the
  right-hand side is regressing the outcome on a threshold version of itself; it
  will produce a large R-squared and a meaningless model. `diastolic_bp` is the
  same measurement's other half (r = `r fmt(cor(dat$systolic_bp, dat$diastolic_bp), 2)`)
  and must also stay out. Both need calling out in the brief.
- **THE missing-data teaching case.** `monthly_income` is
  `r fmt(100 * mean(is.na(dat$monthly_income)), 1)`% missing and listwise
  deletion drops the sample from `r nrow(dat_model)` to `r nrow(dat_inc)`
  (`r fmt(100 * nrow(dat_inc) / nrow(dat), 0)`% of the file) for a variable that
  contributes nothing (p `r pv(inc_p)`). Missingness is strongly patterned by
  facility (`r fmt(miss_by_site$pct[miss_by_site$study_site == "private"], 0)`%
  private vs `r fmt(miss_by_site$pct[miss_by_site$study_site == "faith"], 0)`%
  faith) and by education, so it is not MCAR. This is the best missing-data
  example in the catalogue and should be a required part of the brief.
- **No continuous predictors exist in this file.** Age is banded, BMI is
  categorised. Students will try to report "mmHg per kg/m2" or "per year of age"
  and cannot. The model is a cell-means model; say so up front.
- **Heteroscedasticity is real and flagged (p < 0.001)** with a heavy right
  tail. Good: it gives a genuine reason to run `check_heteroscedasticity()` and
  HC3, and the conclusions survive
  (`r if (n_flip == 0) "no significance flips" else paste(n_flip, "significance flips")`).
- **`current_smoker` is sparse**: `r sum(dat$current_smoker == "Yes")` smokers in
  the whole file, `r sum(dat_model$current_smoker == "Yes" & dat_model$sex == "Male")`
  of `r sum(dat_model$current_smoker == "Yes")` of them men. The coefficient is
  imprecise and largely a sex contrast. Keep it to one degree of freedom and do
  not interact it with anything.
- **`bmi_category` "Underweight" is thin** (`r sum(dat_model$bmi_category == "Underweight")`
  in the analytic sample) and the strongly negative underweight coefficient is
  probably reverse causation (illness-related weight loss), not protection.
- **Clustering cannot be modelled.** Three facilities, no facility ID, so
  `study_site` is the only handle and a random effect is impossible. Standard
  errors are optimistic.
- **Modest R-squared (`r fmt(s$r.squared, 3)`)** will disappoint students who
  expect a "good" model; pre-empt it.
- No separation, no collinearity (max VIF `r fmt(max_vif, 2)`), no influential-point
  problems.
