---
title: "Diagnosed depression among adults in Finland, Poland and Spain"
subtitle: "Dataset 14: Quality of life among adults in Finland, Poland and Spain (COURAGE)"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "ageing_qol_europe.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names. The catalogue's own
# predictor list contains `depression_diagnosis`, which is the outcome here and
# is therefore removed. `sadness` and `sleep_difficulty` are removed as symptoms
# of the outcome. The cut is argued in Methods.
outcome    <- "depression_diagnosis"
predictors <- c("age_yrs", "sex", "country", "residence", "education",
                "marital_status", "physical_activity", "social_network_index",
                "body_pain", "arthritis", "angina", "diabetes")

lab <- function(x, l) { attr(x, "label") <- l; x }
yn  <- function(x, l) lab(factor(x, levels = c(0, 1), labels = c("No", "Yes")), l)

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest group, or the clinically
    # "unexposed" one where that is the more natural comparison.
    age_yrs = lab(age_yrs, "Age (years)"),
    sex     = lab(factor(sex, levels = c("female", "male"),
                         labels = c("Female", "Male")), "Sex"),
    country = lab(factor(country, levels = c("poland", "spain", "finland"),
                         labels = c("Poland", "Spain", "Finland")), "Country"),
    residence = lab(factor(residence, levels = c("urban", "rural"),
                           labels = c("Urban", "Rural")), "Residence"),
    # education: `none_or_informal` (n = 125) is collapsed into the lower
    # category because 124 of those 125 participants live in Spain and none live
    # in Finland. See Methods.
    education = lab(factor(if_else(education == "high_school_or_higher",
                                   "High school or higher",
                                   "None, primary or secondary"),
                           levels = c("None, primary or secondary",
                                      "High school or higher")),
                    "Education"),
    marital_status = lab(factor(marital_status,
      levels = c("married_or_cohabiting", "never_married",
                 "separated_or_divorced", "widowed"),
      labels = c("Married or cohabiting", "Never married",
                 "Separated or divorced", "Widowed")), "Marital status"),
    physical_activity = lab(factor(physical_activity,
      levels = c("high", "moderate", "low"),
      labels = c("High", "Moderate", "Low")), "Physical activity"),
    social_network_index = lab(social_network_index, "Social network index (0-100)"),
    body_pain = lab(factor(body_pain,
      levels = c("none", "pain_without_difficulty", "mild_difficulty",
                 "moderate_difficulty", "severe_or_extreme_difficulty"),
      labels = c("None", "Pain without difficulty", "Mild difficulty",
                 "Moderate difficulty", "Severe or extreme difficulty")),
      "Bodily pain"),
    sadness = lab(factor(sadness,
      levels = c("none", "mild", "moderate", "severe_or_extreme"),
      labels = c("None", "Mild", "Moderate", "Severe or extreme")), "Sadness"),
    arthritis = yn(arthritis, "Diagnosed arthritis"),
    angina    = yn(angina, "Diagnosed angina"),
    diabetes  = yn(diabetes, "Diagnosed diabetes"),
    depression_diagnosis = lab(depression_diagnosis, "Diagnosed depression")
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)
n_events  <- sum(dat_model[[outcome]])

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
pv  <- function(p) ifelse(p < 0.001, "<0.001", paste0("= ", fmt(p, 3)))

# Parameter accounting for the catalogue predictor list minus the outcome.
all_suggested <- c("age_yrs", "sex", "country", "residence", "marital_status",
                   "education", "bmi_category", "smoking_status", "alcohol_use",
                   "physical_activity", "body_pain", "sleep_difficulty", "sadness",
                   "distance_vision_difficulty", "social_network_index",
                   "neighbourhood_usability", "home_usability", "arthritis",
                   "angina", "diabetes")
fit_full   <- glm(reformulate(all_suggested, outcome), family = binomial,
                  data = mutate(dat_raw, across(where(is.character), factor)))
n_par_full <- length(coef(fit_full)) - 1L

# Sadness circularity check (reported in Methods, not used in the model).
sad_tab <- table(dat$sadness, dat$depression_diagnosis)
sad_chi <- suppressWarnings(chisq.test(sad_tab))
cramer_v <- sqrt(as.numeric(sad_chi$statistic) /
                   (sum(sad_tab) * (min(dim(sad_tab)) - 1)))

# Demonstration only: `sadness` is a symptom of the outcome and is NOT in the
# model. Fitted here purely to show how well it would have "worked".
dat_sad  <- dat |> select(all_of(c(outcome, predictors, "sadness"))) |> drop_na()
fit_base <- glm(reformulate(predictors, outcome), data = dat_sad, family = binomial)
fit_sad  <- glm(reformulate(c(predictors, "sadness"), outcome), data = dat_sad,
                family = binomial)
r2_base  <- as.numeric(r2_tjur(fit_base))
r2_sad   <- as.numeric(r2_tjur(fit_sad))
or_sad   <- exp(coef(fit_sad)[["sadnessSevere or extreme"]])

# Demonstration only: country x alcohol_use separates because the Polish
# frequent-heavy stratum has zero events.
sep_fit <- suppressWarnings(glm(depression_diagnosis ~ country * alcohol_use,
                                data = mutate(dat, alcohol_use = factor(alcohol_use)),
                                family = binomial))
sep_cf  <- summary(sep_fit)$coefficients["alcohol_usefrequent_heavy", ]
```

# Introduction

COURAGE in Europe is a cross-sectional survey of the general community-dwelling
adult population of Finland, Poland and Spain, designed to describe healthy
ageing and its social and environmental determinants; the source paper is Miret
and colleagues, *PLOS ONE* (2016). The file distributed with this catalogue holds
`r nrow(dat_raw)` participants aged `r min(dat$age_yrs)` to `r max(dat$age_yrs)`
(mean `r fmt(mean(dat$age_yrs), 1)` years), of whom
`r sum(dat_raw$depression_diagnosis)`
(`r fmt(100 * mean(dat_raw$depression_diagnosis), 1)`%) report having been
diagnosed with depression. Depression is the leading cause of years lived with
disability in European adults and is heavily under-diagnosed in older people, so
identifying which social, physical and national characteristics mark higher
diagnosed prevalence is a standard question in ageing research.

**Aim.** Among these adults in Finland, Poland and Spain, we examined which of
age, sex, country, urban or rural residence, education, marital status, physical
activity, social network, bodily pain, arthritis, angina and diabetes are
independently associated with having a diagnosis of depression, mutually
adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** COURAGE in Europe, a cross-sectional multistage stratified
cluster sample of the community-dwelling adult populations of Finland, Poland and
Spain, fielded between 2011 and 2012. The deposited S1 file used here contains
`r nrow(dat_raw)` participants and `r ncol(dat_raw)` variables:
`r sum(dat$country == "Poland")` from Poland, `r sum(dat$country == "Spain")` from
Spain and `r sum(dat$country == "Finland")` from Finland.

**Variables.** The outcome is `depression_diagnosis`, a binary indicator where the
**event is a reported diagnosis of depression** (`r sum(dat_raw$depression_diagnosis)`
events in the full file, `r fmt(100 * mean(dat_raw$depression_diagnosis), 1)`%).
Predictors are age (years), sex (reference female), country (reference Poland, the
largest group), residence (reference urban), education dichotomised as high school
or higher versus none, primary or secondary (reference the latter), marital status
(reference married or cohabiting), physical activity (reference high), the social
network index (0-100), bodily pain in five ordered categories (reference none) and
diagnosed arthritis, angina and diabetes (each reference no).

**`depression_diagnosis` appears in the catalogue's own predictor list and has
been removed from it.** It is the outcome here; regressing a variable on itself is
not an analysis. This is worth stating explicitly because the suggested-predictor
list shipped with the dataset does contain it, and a student copying that list
verbatim will produce a model with an infinite coefficient.

**`sadness` and `sleep_difficulty` are excluded as symptoms of the outcome.**
Depressed mood and sleep disturbance are two of the nine diagnostic criteria for a
major depressive episode, so a model in which they "predict" a depression
diagnosis is measuring the outcome twice rather than explaining it. The
association is strong but, importantly, **not deterministic**: among participants
reporting severe or extreme sadness,
`r fmt(100 * sad_tab["Severe or extreme", "1"] / sum(sad_tab["Severe or extreme", ]), 1)`%
report a depression diagnosis, against
`r fmt(100 * sad_tab["None", "1"] / sum(sad_tab["None", ]), 1)`% of those
reporting none - a strong gradient (Cramer's V `r fmt(cramer_v, 3)`,
chi-squared p `r pv(sad_chi$p.value)`) but nowhere near a one-to-one mapping, and
no cell of the cross-tabulation is empty. So this is *not* a separation problem
that R would warn about; it is a conceptual problem that only subject-matter
reasoning catches, which makes it more dangerous rather than less. Adding
`sadness` to the final model raises Tjur's R-squared from
`r fmt(r2_base, 3)` to `r fmt(r2_sad, 3)`
and gives severe or extreme sadness an odds ratio of about `r fmt(or_sad, 1)`
- i.e. it "works" beautifully, which is exactly why it is tempting. The
cross-tabulation is shown in the Results.

**The predictor set is deliberately cut.** Excluding the outcome, the catalogue
suggests 20 predictors. Most are multi-level factors, so entering all of them
costs **`r n_par_full` estimated parameters**, not 20: `body_pain` alone costs
four, and `marital_status`, `alcohol_use`, `sleep_difficulty`, `sadness` and
`distance_vision_difficulty` cost three each. With
`r sum(dat_raw$depression_diagnosis)` events that is
`r fmt(sum(dat_raw$depression_diagnosis) / n_par_full, 1)` events per parameter,
which still clears the conventional minimum of ten - so on this outcome too the
global events-per-parameter rule does not by itself force the cut, and a student
who fits everything will see no warning. What does force it is the combination of
four things: a `r n_par_full`-row odds ratio table cannot be interpreted; testing
`r n_par_full` coefficients at 0.05 means expecting about
`r fmt(0.05 * n_par_full, 1)` significant results by chance; two of the suggested
predictors are symptoms of the outcome; and the *global* events-per-parameter
figure hides the fact that individual levels are thin - the `frequent_heavy`
alcohol category has `r sum(dat_raw$alcohol_use == "frequent_heavy")` participants
and `r sum(dat_raw$alcohol_use == "frequent_heavy" & dat_raw$depression_diagnosis == 1)`
events, and crossed with country it contains a **zero-event cell**
(`r sum(dat_raw$alcohol_use == "frequent_heavy" & dat_raw$country == "poland")`
Polish heavy-frequent drinkers, none with a depression diagnosis), which produces
complete separation in any model that interacts the two. The set used here costs
**19 parameters**, i.e. `r fmt(n_events / 19, 1)` events per parameter, and was
fixed before any model was fitted.

Six further suggested predictors are dropped. `bmi_category`, `smoking_status` and
`alcohol_use` are behaviours whose relationship with a *diagnosis* of depression is
weak and heavily bidirectional, and `alcohol_use` carries the sparse-level problem
above. `distance_vision_difficulty` costs three parameters for a functional
limitation whose smallest stratum holds
`r sum(dat_raw$distance_vision_difficulty == "severe_or_extreme")` people.
`neighbourhood_usability` and `home_usability` are environmental scales whose
natural outcome in this dataset is quality of life rather than a clinical
diagnosis; `social_network_index` is retained as the one psychosocial scale with a
well-established literature linking it to depression.

**Level collapsing.** Five categorical predictors were checked for sparse strata.
`education` had `r sum(dat_raw$education == "none_or_informal")` participants with
no or informal education, of whom
`r sum(dat_raw$education == "none_or_informal" & dat_raw$country == "spain")` live
in Spain and **none** in Finland; it was collapsed to two levels, because a
category that exists in one country only cannot be separated from country.
`alcohol_use` would have needed collapsing (`frequent_heavy`,
`r sum(dat_raw$alcohol_use == "frequent_heavy")` participants) but is dropped for
the reasons above. `marital_status` (smallest level `r min(table(dat$marital_status))`),
`bmi_category` (`r min(table(dat$bmi_category))`) and `physical_activity`
(`r min(table(dat$physical_activity))`) were large enough to keep as supplied, and
the thinnest retained cell is
`r min(sapply(dat_model[, predictors][sapply(dat_model[, predictors], is.factor)], function(v) min(table(v, dat_model$depression_diagnosis)[, 2])))`
events, so there is no zero cell and no separation.

**Statistical analysis.** `glm(family = binomial)`, justified because the outcome
is binary. We fitted a univariable model for each predictor first, then a single
multivariable model containing all of them. Predictors were chosen in advance on
subject-matter grounds, not by significance. Coefficients are exponentiated and
reported as odds ratios with 95% confidence intervals; significance threshold
0.05. Model adequacy was assessed with `performance::check_model()`, focusing on
binned residuals, collinearity, influential observations and events per parameter.
Models are **unweighted**: the COURAGE sampling weights, strata and clusters are
not carried in this file, so no result here is a nationally representative
estimate. Analysis used `r R.version.string` with tidyverse, gtsummary and
performance.

**Missing data.** There is none. All `r ncol(dat_raw)` columns are 100% complete
in this file and `r sum(complete.cases(dat_raw))` of `r nrow(dat_raw)` rows are
complete on every variable, so the complete-case analysis uses
`r nrow(dat_model)` of `r nrow(dat)` rows (`r n_dropped` dropped), leaving
`r n_events` events. Listwise deletion costs nothing here, so the predictor cut
rests entirely on interpretability and causal-structure grounds.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  mutate(depression_diagnosis = factor(depression_diagnosis, levels = c(0, 1),
                                       labels = c("No depression diagnosis",
                                                  "Depression diagnosis"))) |>
  tbl_summary(
    by = depression_diagnosis,
    missing = "no"
  ) |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics by depression diagnosis.**")
```

Participants with a depression diagnosis are much more often women
(`r fmt(100 * mean(dat_model$sex[dat_model$depression_diagnosis == 1] == "Female"), 0)`%
versus `r fmt(100 * mean(dat_model$sex[dat_model$depression_diagnosis == 0] == "Female"), 0)`%),
much more often Spanish
(`r fmt(100 * mean(dat_model$country[dat_model$depression_diagnosis == 1] == "Spain"), 0)`%
versus `r fmt(100 * mean(dat_model$country[dat_model$depression_diagnosis == 0] == "Spain"), 0)`%),
and carry a far heavier pain burden
(`r fmt(100 * mean(dat_model$body_pain[dat_model$depression_diagnosis == 1] == "Severe or extreme difficulty"), 0)`%
report severe or extreme pain-related difficulty versus
`r fmt(100 * mean(dat_model$body_pain[dat_model$depression_diagnosis == 0] == "Severe or extreme difficulty"), 0)`%).
Median age is almost identical in the two groups
(`r median(dat_model$age_yrs[dat_model$depression_diagnosis == 1])` versus
`r median(dat_model$age_yrs[dat_model$depression_diagnosis == 0])` years), which
is the first sign that the crude age association is weak and will change under
adjustment.

## The `sadness` cross-tabulation

```{r sadness-tab}
as.data.frame.matrix(sad_tab) |>
  rownames_to_column("Sadness severity") |>
  transmute(
    `Sadness severity`,
    `No depression diagnosis` = `0`,
    `Depression diagnosis` = `1`,
    `Total` = `0` + `1`,
    `% with diagnosis` = fmt(100 * `1` / (`0` + `1`), 1)
  ) |>
  gt() |>
  tab_header(title = md("**Table 2. `sadness` against `depression_diagnosis` (excluded predictor).**"))
```

The gradient is steep and monotonic, from
`r fmt(100 * sad_tab["None", "1"] / sum(sad_tab["None", ]), 1)`% in the no-sadness
group to `r fmt(100 * sad_tab["Severe or extreme", "1"] / sum(sad_tab["Severe or extreme", ]), 1)`%
in the severe group, but the relationship is far from deterministic - most
participants reporting severe or extreme sadness have **no** recorded depression
diagnosis, and `r fmt(100 * sad_tab["None", "1"] / sum(sad_tab["None", ]), 1)`% of
those reporting no sadness do have one. Every cell is populated, so nothing here
would trigger a separation warning. `sadness` is excluded on the conceptual
grounds set out in Methods, not because the model refuses to fit it.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Proportion with a depression diagnosis by country and sex."}
dat_model |>
  group_by(country, sex) |>
  summarise(p = mean(depression_diagnosis), n = n(), .groups = "drop") |>
  ggplot(aes(country, p, fill = sex)) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  geom_text(aes(label = paste0("n=", n)), position = position_dodge(0.8),
            vjust = -0.4, size = 3, colour = "grey30") +
  scale_y_continuous(labels = scales::percent, expand = expansion(c(0, 0.12))) +
  scale_fill_manual(values = c(Female = "#B5495B", Male = "#3F7CAC")) +
  labs(x = "Country", y = "Proportion with a depression diagnosis", fill = "Sex")
```

Diagnosed depression is roughly twice as common in Spain
(`r fmt(100 * mean(dat_model$depression_diagnosis[dat_model$country == "Spain"]), 1)`%)
as in Poland
(`r fmt(100 * mean(dat_model$depression_diagnosis[dat_model$country == "Poland"]), 1)`%),
with Finland in between
(`r fmt(100 * mean(dat_model$depression_diagnosis[dat_model$country == "Finland"]), 1)`%),
and within every country women are diagnosed far more often than men; the sex gap
is proportionally similar in all three, which argues against strong effect
modification by country on this particular contrast.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = glm,
    method.args = list(family = binomial),
    exponentiate = TRUE
  )
uv

utid <- map_dfr(predictors, function(v) {
  tidy(glm(reformulate(v, outcome), data = dat_model, family = binomial),
       conf.int = TRUE, exponentiate = TRUE) |>
    filter(term != "(Intercept)")
})
ug <- function(term, d = 2) {
  r <- utid[utid$term == term, ]
  sprintf("OR %s (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d),
          fmt(r$conf.high, d))
}
up <- function(term) pv(utid$p.value[utid$term == term])
```

Sex shows the strongest crude association: men have `r ug("sexMale")`
(p `r up("sexMale")`) of a depression diagnosis relative to women, a reduction of
about `r fmt(100 * (1 - utid$estimate[utid$term == "sexMale"]), 0)`% in the odds.
Bodily pain shows a steep monotonic gradient, from
`r ug("body_painPain without difficulty")` for pain without difficulty
(p `r up("body_painPain without difficulty")`) to
`r ug("body_painSevere or extreme difficulty")` for severe or extreme
pain-related difficulty (p `r up("body_painSevere or extreme difficulty")`).
Separation or divorce carries `r ug("marital_statusSeparated or divorced")`
(p `r up("marital_statusSeparated or divorced")`) and widowhood
`r ug("marital_statusWidowed")` (p `r up("marital_statusWidowed")`) relative to
being married or cohabiting. Arthritis is associated with `r ug("arthritisYes")`
(p `r up("arthritisYes")`), angina with `r ug("anginaYes")`
(p `r up("anginaYes")`) and diabetes with `r ug("diabetesYes")`
(p `r up("diabetesYes")`). Spanish participants have `r ug("countrySpain")`
(p `r up("countrySpain")`) and Finnish participants `r ug("countryFinland")`
(p `r up("countryFinland")`) relative to Polish participants. Each additional
point of the social network index is associated with `r ug("social_network_index")`
(p `r up("social_network_index")`), which compounds to about
`r fmt(100 * (1 - utid$estimate[utid$term == "social_network_index"]^20), 0)`%
lower odds across a 20-point difference. Low physical activity carries
`r ug("physical_activityLow")` (p `r up("physical_activityLow")`) and higher
education `r ug("educationHigh school or higher")`
(p `r up("educationHigh school or higher")`). Two crude results should be flagged
in advance as unstable: **rural residence appears protective**
(`r ug("residenceRural")`, p `r up("residenceRural")`), and **each year of age is
associated with slightly *higher* odds** (`r ug("age_yrs")`,
p `r up("age_yrs")`) - both reverse or vanish on adjustment.

## Why adjustment is needed

**Country** is the confounder that matters most here, and it explains the rural
result directly. Rural residence is overwhelmingly a Polish phenomenon in this
sample:
`r fmt(100 * mean(dat_model$residence[dat_model$country == "Poland"] == "Rural"), 0)`%
of Polish participants live rurally against
`r fmt(100 * mean(dat_model$residence[dat_model$country == "Spain"] == "Rural"), 0)`%
of Spanish participants, and Poland has by far the lowest diagnosed prevalence
(`r fmt(100 * mean(dat_model$depression_diagnosis[dat_model$country == "Poland"]), 1)`%
against `r fmt(100 * mean(dat_model$depression_diagnosis[dat_model$country == "Spain"]), 1)`%
in Spain). The crude protective rural odds ratio is therefore a statement about
which country people live in, not about rurality. The same argument applies to
education and to the social network index, both of which differ markedly between
the three national samples.

**Age** is the second, and it works in the opposite direction from the usual case.
Older participants are far more likely to be widowed
(median age `r median(dat_model$age_yrs[dat_model$marital_status == "Widowed"])`
against `r median(dat_model$age_yrs[dat_model$marital_status == "Married or cohabiting"])`
for the married or cohabiting), to report bodily pain and to have arthritis, and
all of those are associated with a depression diagnosis. Age therefore *inflates*
the crude widowhood, arthritis and pain coefficients, while the crude age
coefficient itself is propped up by exactly those variables. Holding pain,
arthritis and marital status constant should push the age coefficient below one -
which is what happens, and is consistent with under-diagnosis of depression in the
oldest old. **Bodily pain** is a third: it is associated with arthritis, angina,
low physical activity and widowhood as well as with the outcome, so it will absorb
part of each of those coefficients. Mutual adjustment lets each odds ratio be read
as a comparison between participants of the same age, sex, country, residence,
education, marital status, activity level, social network, pain and chronic
disease status.

## Multivariable model

```{r multivariable}
fit <- glm(reformulate(predictors, outcome), data = dat_model, family = binomial)

mv <- tbl_regression(fit, exponentiate = TRUE)
mv

tid <- tidy(fit, conf.int = TRUE, exponentiate = TRUE)
g <- function(term, d = 2) {
  r <- tid[tid$term == term, ]
  sprintf("OR %s (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d),
          fmt(r$conf.high, d))
}
gp <- function(term) pv(tid$p.value[tid$term == term])
```

```{r fit-stats}
n_par  <- length(coef(fit)) - 1L
epp    <- n_events / n_par
null_fit <- glm(reformulate("1", outcome), data = dat_model, family = binomial)
tibble(
  Statistic = c("Complete-case N", "Events (depression_diagnosis = 1)",
                "Parameters (excl. intercept)", "Events per parameter",
                "AIC (null model)", "AIC (full model)", "Tjur's R-squared"),
  Value = c(as.character(nrow(dat_model)), as.character(n_events),
            as.character(n_par), fmt(epp, 1), fmt(AIC(null_fit), 1),
            fmt(AIC(fit), 1), fmt(as.numeric(r2_tjur(fit)), 3))
) |>
  gt() |>
  tab_header(title = md("**Model fit: multivariable logistic model for diagnosed depression**"))
```

The model has `r n_events` events for `r n_par` estimated parameters, i.e.
**`r fmt(epp, 1)` events per parameter**, four times the conventional minimum, so
over-fitting is not a concern. AIC falls from `r fmt(AIC(null_fit), 1)` for the
intercept-only model to `r fmt(AIC(fit), 1)`, a drop of
`r fmt(AIC(null_fit) - AIC(fit), 1)`, but Tjur's R-squared is only
`r fmt(as.numeric(r2_tjur(fit)), 3)` - the predictors carry real information yet
discriminate poorly between individuals, which is what one would expect of a
diagnosis that depends heavily on unmeasured help-seeking behaviour.

Among participants of the same age, sex, residence, education, marital status,
activity level, social network, pain and chronic disease status, **country is the
largest contrast in the model**: Spanish participants have `r g("countrySpain")`
(p `r gp("countrySpain")`) and Finnish participants `r g("countryFinland")`
(p `r gp("countryFinland")`) relative to Polish participants. Men have
`r g("sexMale")` (p `r gp("sexMale")`), i.e. roughly
`r fmt(100 * (1 - tid$estimate[tid$term == "sexMale"]), 0)`% lower odds than
women. Bodily pain retains its full gradient, from
`r g("body_painMild difficulty")` for mild difficulty
(p `r gp("body_painMild difficulty")`) to
`r g("body_painSevere or extreme difficulty")` for severe or extreme difficulty
(p `r gp("body_painSevere or extreme difficulty")`). Separation or divorce carries
`r g("marital_statusSeparated or divorced")`
(p `r gp("marital_statusSeparated or divorced")`), arthritis
`r g("arthritisYes")` (p `r gp("arthritisYes")`) and angina `r g("anginaYes")`
(p `r gp("anginaYes")`). Each point of the social network index is associated with
`r g("social_network_index")` (p `r gp("social_network_index")`), about
`r fmt(100 * (1 - tid$estimate[tid$term == "social_network_index"]^20), 0)`% lower
odds across 20 points. Age now runs the other way: each additional year is
associated with `r g("age_yrs")` (p `r gp("age_yrs")`). Rural residence
(`r g("residenceRural")`), education (`r g("educationHigh school or higher")`),
widowhood (`r g("marital_statusWidowed")`), diabetes (`r g("diabetesYes")`) and
both physical-activity contrasts (`r g("physical_activityLow")` for low activity)
are not independently associated. These are associations among the participants
observed, not effects of intervening.

**Statistical against practical significance.** With `r nrow(dat_model)`
observations and `r n_events` events, this model has power to detect associations
too small to be useful. Age is the clearest case: `r g("age_yrs")` per year is
significant (p `r gp("age_yrs")`) but corresponds to only
`r fmt(100 * (1 - tid$estimate[tid$term == "age_yrs"]^10), 0)`% lower odds per
decade, which no clinician would act on. Angina (`r g("anginaYes")`,
p `r gp("anginaYes")`) is a second: significant, but the confidence interval runs
from `r fmt(tid$conf.low[tid$term == "anginaYes"], 2)` to
`r fmt(tid$conf.high[tid$term == "anginaYes"], 2)`, so the data are compatible
with a nearly negligible association. Contrast those with the Spain contrast
(`r g("countrySpain")`), the sex contrast (`r g("sexMale")`) and severe pain
(`r g("body_painSevere or extreme difficulty")`), which are large enough to change
how one would think about service planning. At this sample size, statistical
significance separates "non-zero" from "zero", not "important" from
"unimportant".

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 3. Unadjusted and adjusted odds ratios.**")
```

**Age** is the coefficient that changes most fundamentally: it reverses direction,
from `r ug("age_yrs")` crude (p `r up("age_yrs")`, higher odds with age) to
`r g("age_yrs")` adjusted (p `r gp("age_yrs")`, lower odds with age). This is
exactly the confounding structure anticipated above - the crude positive slope was
carried by pain, arthritis and widowhood, all of which accumulate with age, and
once they are held constant the residual age association is protective, most
plausibly because depression is under-recognised in the oldest participants.
**Rural residence** does something similar: from `r ug("residenceRural")`
(p `r up("residenceRural")`, apparently protective and significant) to
`r g("residenceRural")` (p `r gp("residenceRural")`, exactly null). The whole crude
association was country. **Widowhood** collapses from `r ug("marital_statusWidowed")`
(p `r up("marital_statusWidowed")`) to `r g("marital_statusWidowed")`
(p `r gp("marital_statusWidowed")`), losing significance: widows are old and in
pain, and neither of those is caused by widowhood in any useful sense.
**Arthritis** attenuates from `r ug("arthritisYes")` to `r g("arthritisYes")`,
**diabetes** from `r ug("diabetesYes")` (p `r up("diabetesYes")`) to
`r g("diabetesYes")` (p `r gp("diabetesYes")`, null), and **low physical activity**
from `r ug("physical_activityLow")` (p `r up("physical_activityLow")`) to
`r g("physical_activityLow")` (p `r gp("physical_activityLow")`, null) - in each
case bodily pain and age absorb the crude association.

Two estimates move **away** from the null. **Spain** goes from `r ug("countrySpain")`
to `r g("countrySpain")`, and **Finland** from `r ug("countryFinland")` to
`r g("countryFinland")`: the Polish sample is more rural and more often in pain,
which was masking part of the national gap, so adjustment widens it. By contrast
the two strongest predictors **barely move at all**. Severe or extreme pain goes
from `r ug("body_painSevere or extreme difficulty")` to
`r g("body_painSevere or extreme difficulty")` - essentially unchanged, retaining
`r fmt(100 * (tid$estimate[tid$term == "body_painSevere or extreme difficulty"] - 1) / (utid$estimate[utid$term == "body_painSevere or extreme difficulty"] - 1), 0)`%
of its crude excess odds - and sex from `r ug("sexMale")` to `r g("sexMale")`. Neither
is explained by anything else measured here. One estimate crosses out of
significance without moving far: education, from
`r ug("educationHigh school or higher")` (p `r up("educationHigh school or higher")`)
to `r g("educationHigh school or higher")` (p `r gp("educationHigh school or higher")`),
because most of the crude educational gradient was country and age.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r binned-resid}
br <- binned_residuals(fit)
print(br)
```

```{r collin}
check_collinearity(fit)
```

```{r sep-check}
cf <- summary(fit)$coefficients
tibble(
  Check = c("Largest |coefficient| (log-odds)", "Largest standard error",
            "Largest VIF", "Largest Cook's distance", "Largest leverage",
            "Smallest event count in any predictor level",
            "Binned residuals inside bounds"),
  Value = c(fmt(max(abs(cf[-1, 1])), 2), fmt(max(cf[, 2]), 3),
            fmt(max(as.data.frame(check_collinearity(fit))$VIF), 2),
            fmt(max(cooks.distance(fit)), 4), fmt(max(hatvalues(fit)), 3),
            as.character(min(sapply(dat_model[, predictors][sapply(dat_model[, predictors], is.factor)],
                                    function(v) min(table(v, dat_model$depression_diagnosis)[, 2])))),
            paste0(fmt(100 * attributes(br)$resid_ok, 0), "%"))
) |>
  gt() |>
  tab_header(title = md("**Table 4. Separation, influence and collinearity checks.**"))
```

The binned-residual check is the diagnostic that matters for a logistic model and
it **fails badly**: only about `r fmt(100 * attributes(br)$resid_ok, 0)`% of binned
residuals fall inside the error bounds against a target of 95% or more, and
`performance` prints a "probably bad model fit" warning. This is the worst
calibration of the three models in this dataset, and with `r nrow(dat_model)`
observations it is not a small-bin artefact. The model systematically
under-predicts at some parts of the probability range and over-predicts at others,
which is consistent with a Tjur's R-squared of only
`r fmt(as.numeric(r2_tjur(fit)), 3)`: the linear predictor simply does not separate
the `r n_events` cases from the `r nrow(dat_model) - n_events` non-cases well.
Collinearity is not a problem: the largest VIF is
`r fmt(max(as.data.frame(check_collinearity(fit))$VIF), 2)` for
`r as.data.frame(check_collinearity(fit))$Term[which.max(as.data.frame(check_collinearity(fit))$VIF)]`,
comfortably below the conventional threshold of 5, although it is the highest of
the three models here and reflects the entanglement of country with residence and
education described earlier. Influence is negligible: the largest Cook's distance
is `r fmt(max(cooks.distance(fit)), 4)` and the largest leverage
`r fmt(max(hatvalues(fit)), 3)`. There is **no separation**: every coefficient is
finite, the largest standard error is
`r fmt(max(summary(fit)$coefficients[, 2]), 3)`, and every level of every
categorical predictor contains events (Table 4). Homogeneity of variance and
normality of residuals do not apply to a logistic model and are not interpreted;
the corresponding panels of `check_model()` should be ignored here.

# Discussion

Among these `r nrow(dat_model)` adults in Finland, Poland and Spain, the
characteristics most strongly and independently associated with having a
depression diagnosis are country, sex and bodily pain. Spanish participants have
about `r fmt(tid$estimate[tid$term == "countrySpain"], 1)` times the odds of
otherwise similar Polish participants, men about
`r fmt(100 * (1 - tid$estimate[tid$term == "sexMale"]), 0)`% lower odds than
women, and participants with severe or extreme pain-related difficulty about
`r fmt(tid$estimate[tid$term == "body_painSevere or extreme difficulty"], 1)` times
the odds of those with no pain. Separation or divorce, arthritis and a poorer
social network are also independently associated. The crude associations with
rural residence, widowhood, diabetes, low physical activity and education all
disappear on adjustment, and the crude *positive* age association reverses.
Practically, the results point at chronic pain and social disconnection as the two
modifiable characteristics most closely tracking diagnosed depression, and they
underline how much of the variation is between countries rather than between
individuals - a `r fmt(tid$estimate[tid$term == "countrySpain"], 1)`-fold
difference in diagnosis odds between two European countries, unexplained by any
individual characteristic measured here, is more likely to be a statement about
health systems than about mental health.

**The outcome is a *diagnosis*, not depression.** Being counted as a case requires
having consulted a clinician, having been assessed, and having been given a label.
Under-diagnosis in older adults, in men and in countries with less accessible
primary mental-health care will all be misclassified as absence of disease.
That makes the reversed age coefficient
(`r g("age_yrs")`) genuinely ambiguous: it is at least as plausible that
depression is *under-recognised* in the oldest participants as that they are less
depressed. The country contrasts are open to exactly the same reading. Bodily pain
is the other interpretive difficulty: pain and depression are bidirectionally
related, chronic pain is a recognised consequence of depression as well as a cause
of it, and in a cross-sectional survey nothing distinguishes the two directions.

**Limitations.** The models are **unweighted**: the COURAGE sampling weights,
strata and clusters are not carried in this file, so the sample must not be
described as representative of any national population, and the standard errors
ignore clustering within primary sampling units, which will make them too narrow.
The design is cross-sectional, so reverse causation is live for bodily pain,
physical activity, social network and marital status alike - depression plausibly
causes social withdrawal, inactivity and relationship breakdown as readily as the
reverse. The outcome is self-reported diagnosis, with the differential
misclassification by age, sex, country and healthcare access described above. Two
suggested predictors, `sadness` and `sleep_difficulty`, were excluded because they
are diagnostic criteria for the outcome rather than causes of it, and
`depression_diagnosis` itself had to be removed from the catalogue's suggested
predictor list; six others were excluded a priori, so residual confounding by
smoking, alcohol, body mass and vision is certain. The binned-residual check fails
(`r fmt(100 * attributes(br)$resid_ok, 0)`% inside bounds) and Tjur's R-squared is
`r fmt(as.numeric(r2_tjur(fit)), 3)`, so the model is poorly calibrated and
discriminates weakly; its coefficients describe average contrasts, not a usable
prediction rule. There is evidence of effect modification the additive model
ignores: country-by-age
(likelihood-ratio p `r pv(anova(fit, update(fit, . ~ . + country:age_yrs), test = "LRT")[["Pr(>Chi)"]][2])`)
and country-by-marital-status
(p `r pv(anova(fit, update(fit, . ~ . + country:marital_status), test = "LRT")[["Pr(>Chi)"]][2])`)
both improve the fit, so the single age slope reported here averages over three
different national gradients; country-by-sex does not
(p `r pv(anova(fit, update(fit, . ~ . + country:sex), test = "LRT")[["Pr(>Chi)"]][2])`),
consistent with Figure 1. The deposited file differs from the published paper: it
holds `r nrow(dat_raw)` rows with mean age `r fmt(mean(dat$age_yrs), 1)` years,
whereas the source paper reports 5,639 participants with mean age 46.3 years, so
results here should not be compared with the published analysis. The `education`
variable was collapsed to two levels, hiding any effect confined to the
`r sum(dat_raw$education == "none_or_informal")` participants with no formal
education. With `r nrow(dat_model)` observations, small and practically
uninteresting differences reach significance. All estimates are associations, not
causal effects.

**Future work.** Refit with the COURAGE survey design (weights, strata and
clusters) and with the study's symptom-based depression algorithm rather than
self-reported diagnosis, which would remove the help-seeking filter that dominates
the country and age contrasts here.

# Feasibility notes

*NOT part of a student report.*

- **USABLE WITH WARNINGS.** `r n_events` events in `r nrow(dat_model)` complete
  rows, no missing data anywhere, `r fmt(epp, 1)` events per parameter at 19
  parameters, no separation, no influential points (max Cook's
  `r fmt(max(cooks.distance(fit)), 4)`), max VIF
  `r fmt(max(as.data.frame(check_collinearity(fit))$VIF), 2)`. The warnings are
  about interpretation and calibration, not about whether the model can be fitted.
- **FATAL if uncorrected: `depression_diagnosis` is in `meta.json`'s own
  suggested predictor list for this dataset.** A student copying that list gets
  the outcome as a predictor of itself. `build/meta.json` should drop it from
  `predictors`, or the dataset page should carry an explicit note. This is the
  single most likely way this outcome is wrecked, and it is our error, not the
  student's.
- **MAJOR: `sadness` is a symptom of the outcome, and it is NOT detectable as a
  data problem.** Cramer's V `r fmt(cramer_v, 3)`;
  `r fmt(100 * sad_tab["Severe or extreme", "1"] / sum(sad_tab["Severe or extreme", ]), 1)`%
  of the severe-sadness group has a diagnosis against
  `r fmt(100 * sad_tab["None", "1"] / sum(sad_tab["None", ]), 1)`% of the
  no-sadness group, every cell populated, no separation, no warning of any kind.
  Adding it raises Tjur's R-squared from `r fmt(as.numeric(r2_tjur(fit)), 3)` to
  `r fmt(r2_sad, 3)` and gives severe sadness an odds ratio of about
  `r fmt(or_sad, 1)`. It will look like
  the best finding in the report. `sleep_difficulty` has the same problem more
  weakly (sleep disturbance is a diagnostic criterion for major depression). Both
  should be flagged on the dataset page as not-predictors for this outcome.
- **MAJOR: binned residuals fail badly** - only
  `r fmt(100 * attributes(br)$resid_ok, 0)`% inside bounds against a 95% target,
  the worst of the three outcomes, with a "probably bad model fit" warning. Tjur's
  R-squared `r fmt(as.numeric(r2_tjur(fit)), 3)`. A student following the
  diagnostics honestly must report a failed check and a weakly discriminating
  model.
- **MAJOR: 20 usable suggested predictors cost `r n_par_full` parameters, not
  20.** At `r sum(dat_raw$depression_diagnosis)` events that is
  `r fmt(sum(dat_raw$depression_diagnosis) / n_par_full, 1)` events per parameter,
  which *passes* the rule of ten, so the full model fits with no warning. The
  global EPP figure is therefore misleading here; what matters is the individual
  cell counts.
- **MAJOR: the file does not match the paper.** `r nrow(dat_raw)` rows, mean age
  `r fmt(mean(dat$age_yrs), 1)`, median `r median(dat$age_yrs)`, range
  `r min(dat$age_yrs)`-`r max(dat$age_yrs)`; the paper reports 5,639 participants
  with mean age 46.3. Verified directly from the CSV. The dataset page should
  carry an explicit caution.
- **MAJOR trap: separation via `country * alcohol_use`.** `frequent_heavy` has
  `r sum(dat_raw$alcohol_use == "frequent_heavy")` participants in total, and the
  Poland stratum has **`r sum(dat_raw$alcohol_use == "frequent_heavy" & dat_raw$country == "poland")`
  participants with 0 depression events**. Fitting
  `glm(depression_diagnosis ~ country * alcohol_use)` produces a coefficient of
  `r fmt(sep_cf[["Estimate"]], 1)` with a standard error of about
  `r fmt(sep_cf[["Std. Error"]], 0)` - textbook complete separation -
  and a "fitted probabilities numerically 0 or 1" warning. Even without the
  interaction, `frequent_heavy` carries only
  `r sum(dat_raw$alcohol_use == "frequent_heavy" & dat_raw$depression_diagnosis == 1)`
  events. `alcohol_use` should be collapsed or dropped.
- **MINOR but structural: `education == "none_or_informal"` is a Spain-only
  category** -
  `r sum(dat_raw$education == "none_or_informal" & dat_raw$country == "spain")` of
  `r sum(dat_raw$education == "none_or_informal")` are Spanish, **zero** are
  Finnish. Interacted with `country` it separates. Collapsed to two levels here.
- **MINOR: the crude age association reverses on adjustment**
  (`r ug("age_yrs")` to `r g("age_yrs")`). Excellent teaching material for
  confounding, but a weak student will report one direction in the univariable
  section and the opposite in the multivariable section without noticing.
- **MINOR: unmodelled effect modification by country.** country-by-age
  likelihood-ratio
  p `r pv(anova(fit, update(fit, . ~ . + country:age_yrs), test = "LRT")[["Pr(>Chi)"]][2])`,
  country-by-marital-status
  p `r pv(anova(fit, update(fit, . ~ . + country:marital_status), test = "LRT")[["Pr(>Chi)"]][2])`.
  Country is the strongest predictor in the model and is doing more than shifting
  the intercept.
- **Unweighted survey analysis.** Must be stated in Methods and Limitations.
