---
title: "Diagnosed hypertension among adults in Finland, Poland and Spain"
subtitle: "Dataset 14: Quality of life among adults in Finland, Poland and Spain (COURAGE)"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "ageing_qol_europe.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names. The catalogue suggests
# 21 predictors, which cost 39 parameters; the set below costs 15. The cut is
# argued in Methods.
outcome    <- "hypertension"
predictors <- c("age_yrs", "sex", "country", "residence", "education",
                "bmi_category", "smoking_status", "alcohol_use",
                "physical_activity", "diabetes")

lab <- function(x, l) { attr(x, "label") <- l; x }
yn  <- function(x, l) lab(factor(x, levels = c(0, 1), labels = c("No", "Yes")), l)

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest group, or the clinically
    # "unexposed" one where that is the more natural comparison.
    age_yrs = lab(age_yrs, "Age (years)"),
    sex     = lab(factor(sex, levels = c("female", "male"),
                         labels = c("Female", "Male")), "Sex"),
    country = lab(factor(country, levels = c("poland", "spain", "finland"),
                         labels = c("Poland", "Spain", "Finland")), "Country"),
    residence = lab(factor(residence, levels = c("urban", "rural"),
                           labels = c("Urban", "Rural")), "Residence"),
    # education: `none_or_informal` (n = 125) is collapsed into the lower
    # category because 124 of those 125 participants live in Spain and none live
    # in Finland. See Methods.
    education = lab(factor(if_else(education == "high_school_or_higher",
                                   "High school or higher",
                                   "None, primary or secondary"),
                           levels = c("None, primary or secondary",
                                      "High school or higher")),
                    "Education"),
    bmi_category = lab(factor(bmi_category,
      levels = c("normal_or_underweight", "overweight", "obese"),
      labels = c("Normal or underweight", "Overweight", "Obese")),
      "BMI category"),
    smoking_status = lab(factor(smoking_status,
      levels = c("never", "former", "current"),
      labels = c("Never", "Former", "Current")), "Smoking status"),
    # alcohol_use: `frequent_heavy` holds 35 participants, so the two heavy
    # categories are pooled. See Methods.
    alcohol_use = lab(factor(case_when(
      alcohol_use %in% c("frequent_heavy", "infrequent_heavy") ~ "Heavy",
      alcohol_use == "abstainer_or_occasional" ~ "Abstainer or occasional",
      alcohol_use == "non_heavy" ~ "Non-heavy"),
      levels = c("Non-heavy", "Abstainer or occasional", "Heavy")),
      "Alcohol use"),
    physical_activity = lab(factor(physical_activity,
      levels = c("high", "moderate", "low"),
      labels = c("High", "Moderate", "Low")), "Physical activity"),
    diabetes = yn(diabetes, "Diagnosed diabetes"),
    hypertension = lab(hypertension, "Diagnosed hypertension")
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)
n_events  <- sum(dat_model[[outcome]])

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
pv  <- function(p) ifelse(p < 0.001, "<0.001", paste0("= ", fmt(p, 3)))

# Parameter accounting for the full catalogue predictor list, used in Methods.
all_suggested <- c("age_yrs", "sex", "country", "residence", "marital_status",
                   "education", "bmi_category", "smoking_status", "alcohol_use",
                   "physical_activity", "body_pain", "sleep_difficulty", "sadness",
                   "distance_vision_difficulty", "social_network_index",
                   "neighbourhood_usability", "home_usability", "arthritis",
                   "angina", "diabetes", "depression_diagnosis")
fit_full   <- glm(reformulate(all_suggested, outcome), family = binomial,
                  data = mutate(dat_raw, across(where(is.character), factor)))
n_par_full <- length(coef(fit_full)) - 1L

# Demonstration only: angina is downstream of hypertension and is NOT in the model.
dat_ang <- dat |> select(all_of(c(outcome, predictors, "angina"))) |>
  mutate(angina = yn(angina, "Diagnosed angina")) |> drop_na()
or_angina <- exp(coef(glm(reformulate(c(predictors, "angina"), outcome),
                          data = dat_ang, family = binomial))[["anginaYes"]])
```

# Introduction

COURAGE in Europe is a cross-sectional survey of the general community-dwelling
adult population of Finland, Poland and Spain, designed to describe healthy
ageing and its social and environmental determinants; the source paper is Miret
and colleagues, *PLOS ONE* (2016). The file distributed with this catalogue holds
`r nrow(dat_raw)` participants aged `r min(dat$age_yrs)` to `r max(dat$age_yrs)`
(mean `r fmt(mean(dat$age_yrs), 1)` years), of whom `r sum(dat_raw$hypertension)`
(`r fmt(100 * mean(dat_raw$hypertension), 1)`%) report a diagnosis of
hypertension. Hypertension is the largest single contributor to cardiovascular
mortality in Europe and is both preventable and treatable, so identifying which
routinely collected characteristics mark higher diagnosed prevalence is directly
useful for targeting screening.

**Aim.** Among these adults in Finland, Poland and Spain, we examined which of
age, sex, country, urban or rural residence, education, BMI category, smoking
status, alcohol use, physical activity and diagnosed diabetes are independently
associated with having a diagnosis of hypertension, mutually adjusting for the
others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** COURAGE in Europe, a cross-sectional multistage stratified
cluster sample of the community-dwelling adult populations of Finland, Poland and
Spain, fielded between 2011 and 2012. The deposited S1 file used here contains
`r nrow(dat_raw)` participants and `r ncol(dat_raw)` variables:
`r sum(dat$country == "Poland")` from Poland, `r sum(dat$country == "Spain")` from
Spain and `r sum(dat$country == "Finland")` from Finland.

**Variables.** The outcome is `hypertension`, a binary indicator where the
**event is a reported diagnosis of hypertension** (`r sum(dat_raw$hypertension)`
events in the full file, `r fmt(100 * mean(dat_raw$hypertension), 1)`%). This is
diagnosed, not measured, hypertension - a point returned to in Limitations.
Predictors are age (years), sex (reference female), country (reference Poland,
the largest group), residence (reference urban), education dichotomised as high
school or higher versus none, primary or secondary (reference the latter), BMI
category (reference normal or underweight, the clinically unexposed group),
smoking status (reference never), alcohol use in three categories (reference
non-heavy drinking, the conventional reference in alcohol epidemiology because
abstainers include people who stopped because they were ill), physical activity
(reference high) and diagnosed diabetes (reference no).

**The predictor set is deliberately cut.** The catalogue suggests 21 predictors.
Most are multi-level factors, so entering all of them costs **`r n_par_full`
estimated parameters**, not 21: `body_pain` alone costs four, and
`marital_status`, `alcohol_use`, `sleep_difficulty`, `sadness` and
`distance_vision_difficulty` cost three each. With `r sum(dat_raw$hypertension)`
events that would be `r fmt(sum(dat_raw$hypertension) / n_par_full, 1)` events per
parameter, which clears the conventional minimum of ten comfortably, so it should
be said plainly that on this outcome the events-per-parameter rule is **not** what
forces the cut. Three other considerations do. First, a `r n_par_full`-row odds
ratio table cannot be interpreted, and the Purpose 3 question - which predictors
survive mutual adjustment - is only answerable for a set small enough to reason
about. Second, testing `r n_par_full` coefficients at the 0.05 level means
expecting about `r fmt(0.05 * n_par_full, 1)` significant results by chance alone.
Third, and decisively, most of the suggested predictors have no place in a model
for hypertension: they are either symptoms and disability measures that are
plausibly *consequences* of cardiovascular disease, or they are downstream of the
outcome itself. The set used here costs **15 parameters**, i.e.
`r fmt(n_events / 15, 1)` events per parameter, and was fixed before any model was
fitted.

Eleven suggested predictors are dropped. `angina` is excluded because angina is a
*consequence* of long-standing hypertension, not a cause of it: adding it to this
model gives an odds ratio of about
`r fmt(or_angina, 2)`,
which is conditioning on a downstream variable and would bias every other
coefficient. `body_pain`, `sleep_difficulty`, `sadness`,
`distance_vision_difficulty`, `arthritis` and `depression_diagnosis` are excluded
on the same logic - they are symptom and morbidity measures whose temporal
position relative to a hypertension diagnosis is unknown, and between them they
would cost 15 parameters for no gain in causal clarity. `marital_status` (3
parameters), `social_network_index`, `neighbourhood_usability` and
`home_usability` are excluded because they are not established determinants of
blood pressure and are more naturally read as determinants of quality of life,
which is a different outcome in this catalogue. `diabetes` is retained despite
sharing this ambiguity, because it is the standard comorbidity adjustment in
cardiovascular models and because its interpretation here is instructive: see
Limitations.

**Level collapsing.** Five categorical predictors were checked for sparse strata.
`alcohol_use` has a `frequent_heavy` level with only
`r sum(dat_raw$alcohol_use == "frequent_heavy")` participants and
`r sum(dat_raw$alcohol_use == "frequent_heavy" & dat_raw$hypertension == 1)`
events, so it was pooled with `infrequent_heavy` into a single "Heavy" category
(n = `r sum(dat_model$alcohol_use == "Heavy")`). `education` had
`r sum(dat_raw$education == "none_or_informal")` participants with no or informal
education, of whom
`r sum(dat_raw$education == "none_or_informal" & dat_raw$country == "spain")` live
in Spain and **none** in Finland; it was collapsed to two levels, because a
category that exists in only one of the three countries cannot be separated from
country - and a `country`-by-`education` model on the uncollapsed variable
produces outright separation (a coefficient of about -10.7 with a standard error
of about 197). `marital_status` (smallest level `r min(table(dat$marital_status))`),
`bmi_category` (`r min(table(dat$bmi_category))`) and `physical_activity`
(`r min(table(dat$physical_activity))`) were all large enough to keep as supplied,
and every level of every retained predictor contains at least
`r min(table(dat_model$alcohol_use, dat_model$hypertension)[, 2], table(dat_model$education, dat_model$hypertension)[, 2], table(dat_model$bmi_category, dat_model$hypertension)[, 2], table(dat_model$country, dat_model$hypertension)[, 2], table(dat_model$smoking_status, dat_model$hypertension)[, 2], table(dat_model$physical_activity, dat_model$hypertension)[, 2], table(dat_model$residence, dat_model$hypertension)[, 2])`
events, so there is no zero cell and no separation.

**Statistical analysis.** `glm(family = binomial)`, justified because the outcome
is binary. We fitted a univariable model for each predictor first, then a single
multivariable model containing all of them. Predictors were chosen in advance on
subject-matter grounds, not by significance. Coefficients are exponentiated and
reported as odds ratios with 95% confidence intervals; significance threshold
0.05. Model adequacy was assessed with `performance::check_model()`, focusing on
binned residuals, collinearity, influential observations and events per parameter.
Models are **unweighted**: the COURAGE sampling weights, strata and clusters are
not carried in this file, so no result here is a nationally representative
estimate. Analysis used `r R.version.string` with tidyverse, gtsummary and
performance.

**Missing data.** There is none. All `r ncol(dat_raw)` columns are 100% complete
in this file and `r sum(complete.cases(dat_raw))` of `r nrow(dat_raw)` rows are
complete on every variable, so the complete-case analysis uses
`r nrow(dat_model)` of `r nrow(dat)` rows (`r n_dropped` dropped), leaving
`r n_events` events. Listwise deletion costs nothing here, which means the
predictor cut above rests entirely on interpretability and causal-structure
grounds rather than on preserving sample size.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  mutate(hypertension = factor(hypertension, levels = c(0, 1),
                               labels = c("No hypertension", "Hypertension"))) |>
  tbl_summary(
    by = hypertension,
    missing = "no"
  ) |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics by hypertension status.**")
```

Participants with a hypertension diagnosis are considerably older (median
`r median(dat_model$age_yrs[dat_model$hypertension == 1])` versus
`r median(dat_model$age_yrs[dat_model$hypertension == 0])` years), far more often
obese (`r fmt(100 * mean(dat_model$bmi_category[dat_model$hypertension == 1] == "Obese"), 0)`%
versus `r fmt(100 * mean(dat_model$bmi_category[dat_model$hypertension == 0] == "Obese"), 0)`%)
and much more often diabetic
(`r fmt(100 * mean(dat_model$diabetes[dat_model$hypertension == 1] == "Yes"), 0)`%
versus `r fmt(100 * mean(dat_model$diabetes[dat_model$hypertension == 0] == "Yes"), 0)`%).
They are also less educated
(`r fmt(100 * mean(dat_model$education[dat_model$hypertension == 1] == "High school or higher"), 0)`%
with high school or higher versus
`r fmt(100 * mean(dat_model$education[dat_model$hypertension == 0] == "High school or higher"), 0)`%),
which is one of the associations that adjustment substantially changes. Sex and
urban/rural residence look similar in the two groups.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Proportion with diagnosed hypertension by age group and BMI category."}
dat_model |>
  mutate(age_grp = cut(age_yrs, breaks = c(18, 45, 60, 75, 105), right = FALSE,
                       labels = c("18-44", "45-59", "60-74", "75+"))) |>
  group_by(age_grp, bmi_category) |>
  summarise(p = mean(hypertension), n = n(), .groups = "drop") |>
  ggplot(aes(age_grp, p, fill = bmi_category)) +
  geom_col(position = position_dodge(0.85), width = 0.78) +
  geom_text(aes(label = paste0("n=", n)), position = position_dodge(0.85),
            vjust = -0.4, size = 2.7, colour = "grey30") +
  scale_y_continuous(labels = scales::percent, expand = expansion(c(0, 0.12))) +
  scale_fill_manual(values = c("Normal or underweight" = "#7FA9C9",
                               "Overweight" = "#E0A458", "Obese" = "#B5495B")) +
  labs(x = "Age group (years)", y = "Proportion with diagnosed hypertension",
       fill = "BMI category")
```

Diagnosed hypertension rises steeply with both age and BMI and the two compound:
prevalence is around
`r fmt(100 * mean(dat_model$hypertension[dat_model$age_yrs < 45 & dat_model$bmi_category == "Normal or underweight"]), 0)`%
among normal-weight adults under 45 and around
`r fmt(100 * mean(dat_model$hypertension[dat_model$age_yrs >= 75 & dat_model$bmi_category == "Obese"]), 0)`%
among obese adults aged 75 and over. The BMI gradient is visible within every age
band, so it is not simply an artefact of older participants being heavier.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = glm,
    method.args = list(family = binomial),
    exponentiate = TRUE
  )
uv

utid <- map_dfr(predictors, function(v) {
  tidy(glm(reformulate(v, outcome), data = dat_model, family = binomial),
       conf.int = TRUE, exponentiate = TRUE) |>
    filter(term != "(Intercept)")
})
ug <- function(term, d = 2) {
  r <- utid[utid$term == term, ]
  sprintf("OR %s (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d),
          fmt(r$conf.high, d))
}
up <- function(term) pv(utid$p.value[utid$term == term])
```

Obesity shows the strongest crude association: compared with normal-weight or
underweight participants, obese participants have `r ug("bmi_categoryObese")`
(p `r up("bmi_categoryObese")`) and overweight participants
`r ug("bmi_categoryOverweight")` (p `r up("bmi_categoryOverweight")`). Diagnosed
diabetes carries `r ug("diabetesYes")` (p `r up("diabetesYes")`). Each additional
year of age multiplies the odds by `r ug("age_yrs")` (p `r up("age_yrs")`), which
compounds to about
`r fmt(utid$estimate[utid$term == "age_yrs"]^10, 2)`-fold per decade. Education is
strongly protective in crude terms: having high school education or above is
associated with `r ug("educationHigh school or higher")`
(p `r up("educationHigh school or higher")`), a halving of the odds. Abstinence or
occasional drinking carries `r ug("alcohol_useAbstainer or occasional")`
(p `r up("alcohol_useAbstainer or occasional")`) relative to non-heavy drinking,
and low physical activity `r ug("physical_activityLow")`
(p `r up("physical_activityLow")`). Two crude results are worth flagging as
suspicious in advance: **current smoking appears protective**
(`r ug("smoking_statusCurrent")`, p `r up("smoking_statusCurrent")`) while former
smoking is associated with higher odds (`r ug("smoking_statusFormer")`,
p `r up("smoking_statusFormer")`); and **Finland shows no crude excess**
(`r ug("countryFinland")`, p `r up("countryFinland")`) although Spain does show a
deficit (`r ug("countrySpain")`, p `r up("countrySpain")`). Sex
(`r ug("sexMale")`, p `r up("sexMale")`) and rural residence
(`r ug("residenceRural")`, p `r up("residenceRural")`) show no crude association.

## Why adjustment is needed

**Age** is the dominant confounder and it explains both of the suspicious crude
results above. Current smokers in this sample have a median age of
`r median(dat_model$age_yrs[dat_model$smoking_status == "Current"])` years against
`r median(dat_model$age_yrs[dat_model$smoking_status == "Never"])` for never
smokers, so the apparently protective crude smoking odds ratio is a statement
about the youth of current smokers, not about tobacco. The same mechanism drives
the physical-activity association: participants in the low-activity category have
a median age of
`r median(dat_model$age_yrs[dat_model$physical_activity == "Low"])` against
`r median(dat_model$age_yrs[dat_model$physical_activity == "High"])` in the
high-activity category, and inactivity is as plausibly a consequence of
age-related morbidity as a cause of hypertension.

**Education** is the second confounder, and it works on the country and BMI
comparisons. Participants with high school education or above have a median age of
`r median(dat_model$age_yrs[dat_model$education == "High school or higher"])`
against `r median(dat_model$age_yrs[dat_model$education == "None, primary or secondary"])`
for the rest - educational expansion means the less educated are simply older -
and they are much less often obese
(`r fmt(100 * mean(dat_model$bmi_category[dat_model$education == "High school or higher"] == "Obese"), 0)`%
versus `r fmt(100 * mean(dat_model$bmi_category[dat_model$education == "None, primary or secondary"] == "Obese"), 0)`%).
So the crude halving of the odds with education carries both an age difference and
a body-mass difference. **Country** confounds in a third way: the Finnish sample is
the oldest of the three (median age
`r median(dat_model$age_yrs[dat_model$country == "Finland"])` against
`r median(dat_model$age_yrs[dat_model$country == "Poland"])` in Poland), which is
why Finland shows no crude excess despite the highest crude prevalence
(`r fmt(100 * mean(dat_model$hypertension[dat_model$country == "Finland"]), 1)`%).
Mutual adjustment lets each odds ratio be read as a comparison between
participants of the same age, sex, country, residence, education, body mass,
smoking, drinking, activity level and diabetes status.

## Multivariable model

```{r multivariable}
fit <- glm(reformulate(predictors, outcome), data = dat_model, family = binomial)

mv <- tbl_regression(fit, exponentiate = TRUE)
mv

tid <- tidy(fit, conf.int = TRUE, exponentiate = TRUE)
g <- function(term, d = 2) {
  r <- tid[tid$term == term, ]
  sprintf("OR %s (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d),
          fmt(r$conf.high, d))
}
gp <- function(term) pv(tid$p.value[tid$term == term])
```

```{r fit-stats}
n_par  <- length(coef(fit)) - 1L
epp    <- n_events / n_par
null_fit <- glm(reformulate("1", outcome), data = dat_model, family = binomial)
tibble(
  Statistic = c("Complete-case N", "Events (hypertension = 1)",
                "Parameters (excl. intercept)", "Events per parameter",
                "AIC (null model)", "AIC (full model)", "Tjur's R-squared"),
  Value = c(as.character(nrow(dat_model)), as.character(n_events),
            as.character(n_par), fmt(epp, 1), fmt(AIC(null_fit), 1),
            fmt(AIC(fit), 1), fmt(as.numeric(r2_tjur(fit)), 3))
) |>
  gt() |>
  tab_header(title = md("**Model fit: multivariable logistic model for diagnosed hypertension**"))
```

The model has `r n_events` events for `r n_par` estimated parameters, i.e.
**`r fmt(epp, 1)` events per parameter**, an order of magnitude above the
conventional minimum of ten, so over-fitting is not a concern here. AIC falls from
`r fmt(AIC(null_fit), 1)` for the intercept-only model to `r fmt(AIC(fit), 1)`, a
drop of `r fmt(AIC(null_fit) - AIC(fit), 1)`, and Tjur's R-squared is
`r fmt(as.numeric(r2_tjur(fit)), 3)`.

Among participants of the same age, sex, country, residence, education, smoking,
drinking, activity level and diabetes status, obesity is associated with
`r g("bmi_categoryObese")` (p `r gp("bmi_categoryObese")`) and being overweight
with `r g("bmi_categoryOverweight")` (p `r gp("bmi_categoryOverweight")`).
Diagnosed diabetes carries `r g("diabetesYes")` (p `r gp("diabetesYes")`). Each
additional year of age multiplies the odds by `r g("age_yrs")`
(p `r gp("age_yrs")`), about `r fmt(tid$estimate[tid$term == "age_yrs"]^10, 2)`-fold
per decade. Holding everything else constant, both non-Polish countries now show
lower odds than Poland: `r g("countrySpain")` for Spain
(p `r gp("countrySpain")`) and `r g("countryFinland")` for Finland
(p `r gp("countryFinland")`). Former smoking remains associated with
`r g("smoking_statusFormer")` (p `r gp("smoking_statusFormer")`) while current
smoking is now null (`r g("smoking_statusCurrent")`,
p `r gp("smoking_statusCurrent")`). Education retains a small independent
association (`r g("educationHigh school or higher")`,
p `r gp("educationHigh school or higher")`), as does abstinence or occasional
drinking (`r g("alcohol_useAbstainer or occasional")`,
p `r gp("alcohol_useAbstainer or occasional")`). Sex
(`r g("sexMale")`), rural residence (`r g("residenceRural")`), heavy drinking
(`r g("alcohol_useHeavy")`) and both physical-activity contrasts
(`r g("physical_activityLow")` for low activity) are not independently
associated. These are associations among the participants observed, not effects
of intervening.

**Statistical against practical significance.** With `r nrow(dat_model)`
observations and `r n_events` events, this model detects associations that are
too small to act on. The education contrast is the clearest case: it is
significant (`r g("educationHigh school or higher")`,
p `r gp("educationHigh school or higher")`) but a
`r fmt(100 * (1 - tid$estimate[tid$term == "educationHigh school or higher"]), 0)`%
relative reduction in the odds of a diagnosis, with a confidence interval whose
upper limit reaches
`r fmt(tid$conf.high[tid$term == "educationHigh school or higher"], 2)`, is not a
finding on which a screening programme would be built. The same applies to
abstinence (`r g("alcohol_useAbstainer or occasional")`,
p `r gp("alcohol_useAbstainer or occasional")`). Compare those with obesity
(`r g("bmi_categoryObese")`), which quadruples the odds, or age, which multiplies
them roughly `r fmt(tid$estimate[tid$term == "age_yrs"]^10, 1)`-fold per decade.
A p-value below 0.05 at this sample size answers "is it distinguishable from zero",
not "is it big enough to matter", and only the second question is useful.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted odds ratios.**")
```

**Current smoking** is the coefficient that moves furthest, and it crosses the
null: `r ug("smoking_statusCurrent")` crude (p `r up("smoking_statusCurrent")`,
apparently protective) becomes `r g("smoking_statusCurrent")` adjusted
(p `r gp("smoking_statusCurrent")`, null). This is age confounding in almost pure
form, as anticipated - current smokers are
`r median(dat_model$age_yrs[dat_model$smoking_status == "Never"]) - median(dat_model$age_yrs[dat_model$smoking_status == "Current"])`
years younger at the median than never smokers - and it is a good demonstration
of why a crude protective association with a known harm should never be reported.
**Physical activity** does the same thing from the other side: low activity moves
from `r ug("physical_activityLow")` (p `r up("physical_activityLow")`) to
`r g("physical_activityLow")` (p `r gp("physical_activityLow")`), i.e. all the way
to the null and out of significance, so the crude inactivity association was
entirely age and body mass. **Education** attenuates sharply, from
`r ug("educationHigh school or higher")` to
`r g("educationHigh school or higher")`, retaining significance but losing most of
its magnitude; the crude estimate was carrying age, country and obesity.
**Obesity** attenuates too, from `r ug("bmi_categoryObese")` to
`r g("bmi_categoryObese")`, and **diabetes** from `r ug("diabetesYes")` to
`r g("diabetesYes")`, both toward the null and both still large.

Two estimates move *away* from the null. **Finland** goes from
`r ug("countryFinland")` (p `r up("countryFinland")`, no association) to
`r g("countryFinland")` (p `r gp("countryFinland")`, clearly protective) - a change
of significance status caused by the Finnish sample being the oldest of the three,
which masked a genuinely lower age-specific prevalence. **Spain** moves the same
way, from `r ug("countrySpain")` to `r g("countrySpain")`. By contrast **age
itself barely moves**, from `r ug("age_yrs")` to `r g("age_yrs")`, and so does
**former smoking**, from `r ug("smoking_statusFormer")` to
`r g("smoking_statusFormer")` - neither is explained by the other measured
characteristics, which is what you would expect of age and of a cumulative
exposure that people typically quit *because* of a diagnosis.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r binned-resid}
br <- binned_residuals(fit)
print(br)
```

```{r collin}
check_collinearity(fit)
```

```{r sep-check}
cf <- summary(fit)$coefficients
tibble(
  Check = c("Largest |coefficient| (log-odds)", "Largest standard error",
            "Largest VIF", "Largest Cook's distance", "Largest leverage",
            "Smallest event count in any predictor level"),
  Value = c(fmt(max(abs(cf[-1, 1])), 2), fmt(max(cf[, 2]), 3),
            fmt(max(as.data.frame(check_collinearity(fit))$VIF), 2),
            fmt(max(cooks.distance(fit)), 4), fmt(max(hatvalues(fit)), 3),
            as.character(min(sapply(dat_model[, predictors][sapply(dat_model[, predictors], is.factor)],
                                    function(v) min(table(v, dat_model$hypertension)[, 2])))))
) |>
  gt() |>
  tab_header(title = md("**Table 3. Separation, influence and collinearity checks.**"))
```

The binned-residual check is the one that matters for a logistic model and it is
**not reassuring**: only about
`r fmt(100 * attributes(br)$resid_ok, 0)`%
of binned residuals fall inside the error bounds against a target of 95% or more,
and `performance` prints a "probably bad model fit" warning. With
`r nrow(dat_model)` observations the bins are large enough that this is unlikely
to be noise, and the most plausible explanation is that the linear-in-log-odds
assumption for age is imperfect - diagnosed hypertension prevalence plateaus in
the oldest participants rather than continuing to rise on the log-odds scale.
Collinearity is not a problem: the largest VIF is
`r fmt(max(as.data.frame(check_collinearity(fit))$VIF), 2)` for
`r as.data.frame(check_collinearity(fit))$Term[which.max(as.data.frame(check_collinearity(fit))$VIF)]`,
far below the conventional threshold of 5, so country, education and residence are
not competing badly despite their structural overlap. Influence is negligible: the
largest Cook's distance is `r fmt(max(cooks.distance(fit)), 4)` and the largest
leverage `r fmt(max(hatvalues(fit)), 3)`, so no participant drives the fit. There
is **no separation**: every coefficient is finite, the largest standard error is
`r fmt(max(summary(fit)$coefficients[, 2]), 3)`, and every level of every
categorical predictor contains events (Table 3). Homogeneity of variance and
normality of residuals do not apply to a logistic model and are not interpreted;
the corresponding panels of `check_model()` should be ignored here.

# Discussion

Among these `r nrow(dat_model)` adults in Finland, Poland and Spain, obesity, age
and diagnosed diabetes are the characteristics most strongly and independently
associated with having a hypertension diagnosis: obesity roughly
`r fmt(tid$estimate[tid$term == "bmi_categoryObese"], 1)`-fold, diabetes about
`r fmt(tid$estimate[tid$term == "diabetesYes"], 1)`-fold, and age about
`r fmt(tid$estimate[tid$term == "age_yrs"]^10, 1)`-fold per decade, each holding
the others constant. The crude protective association with current smoking and the
crude harmful association with inactivity both disappear once age is held
constant, and the education gradient shrinks to something statistically detectable
but practically small. Country remains a substantial independent contrast: at the
same age and body mass, Spanish and Finnish participants have roughly
`r fmt(100 * (1 - tid$estimate[tid$term == "countrySpain"]), 0)`% and
`r fmt(100 * (1 - tid$estimate[tid$term == "countryFinland"]), 0)`% lower odds of
a diagnosis than Polish participants. Practically, this supports age and BMI as
the primary axes for opportunistic blood-pressure screening in these populations.

**The outcome is a *diagnosis*, and that is the central interpretive problem.**
Undiagnosed hypertension is misclassified as absent, and being diagnosed requires
having had your blood pressure measured, which requires contact with a health
system. That contact is not distributed at random: it differs by country, by
education and by whether you already have another condition that brings you into
a clinic. This is **differential misclassification**, and it works in a direction
that can manufacture or destroy associations. The `diabetes` odds ratio of
`r fmt(tid$estimate[tid$term == "diabetesYes"], 2)` is the clearest example - some
of it is real shared aetiology through adiposity, but some of it is simply that
people with diagnosed diabetes attend clinics where blood pressure is measured
routinely, so their hypertension is more likely to have been *found*. The same
argument applies to the education and country contrasts: the apparently lower odds
in Spain and Finland could reflect genuinely lower prevalence, or different
diagnostic thresholds and screening intensity, and this dataset cannot separate
the two. A student reading these coefficients as prevalence differences is
over-reading them.

**Limitations.** The models are **unweighted**: the COURAGE sampling weights,
strata and clusters are not carried in this file, so the sample must not be
described as representative of any national population, and the standard errors
ignore clustering within primary sampling units, which will make them too narrow.
The outcome is self-reported *diagnosed* hypertension rather than measured blood
pressure, with the differential misclassification by healthcare access described
above; no measured systolic or diastolic pressure is available in this file to
check it. The design is cross-sectional, so reverse causation is plausible for
several predictors: people diagnosed with hypertension are advised to lose weight,
stop smoking and exercise, which would attenuate the BMI and activity associations
and could invert the smoking one. The deposited file differs from the published
paper: it holds `r nrow(dat_raw)` rows with mean age `r fmt(mean(dat$age_yrs), 1)`
years, whereas the source paper reports 5,639 participants with mean age 46.3
years, so results here should not be compared with the published analysis. The
binned-residual check fails, indicating that the model is miscalibrated in parts
of the predicted-probability range and most likely that age is not linear in the
log-odds. There is evidence of effect modification the additive model ignores: a
country-by-education interaction improves the fit
(likelihood-ratio p `r pv(anova(fit, update(fit, . ~ . + country:education), test = "LRT")[["Pr(>Chi)"]][2])`),
so the education gradient is not the same in all three countries; country-by-age
and country-by-BMI interactions do not
(p `r pv(anova(fit, update(fit, . ~ . + country:age_yrs), test = "LRT")[["Pr(>Chi)"]][2])`
and p `r pv(anova(fit, update(fit, . ~ . + country:bmi_category), test = "LRT")[["Pr(>Chi)"]][2])`).
Eleven suggested predictors were excluded a priori, so residual confounding is
certain - family history and salt intake, the two dominant determinants of blood
pressure, are not measured at all. The `education` variable was collapsed to two
levels, so any effect confined to the `r sum(dat_raw$education == "none_or_informal")`
participants with no formal education is invisible; that group has a crude
hypertension prevalence of
`r fmt(100 * mean(dat_raw$hypertension[dat_raw$education == "none_or_informal"]), 0)`%
and is almost entirely Spanish. With `r nrow(dat_model)` observations, small and
practically uninteresting differences reach significance. All estimates are
associations, not causal effects.

**Future work.** Refit with the COURAGE survey design (weights, strata and
clusters) and with the measured blood-pressure readings from the full COURAGE
files, which would replace diagnosed hypertension with measured hypertension and
remove the dominant misclassification problem.

# Feasibility notes

*NOT part of a student report.*

- **USABLE. The best-powered of the three outcomes in this dataset.**
  `r n_events` events in `r nrow(dat_model)` complete rows, no missing data
  anywhere in the file, `r fmt(epp, 1)` events per parameter at 15 parameters,
  Tjur's R-squared `r fmt(as.numeric(r2_tjur(fit)), 3)`, no separation, no
  influential points (max Cook's `r fmt(max(cooks.distance(fit)), 4)`), max VIF
  `r fmt(max(as.data.frame(check_collinearity(fit))$VIF), 2)`.
- **MAJOR: 21 suggested predictors cost `r n_par_full` parameters, not 21.** Most
  are multi-level factors. Because there are `r sum(dat_raw$hypertension)` events,
  even the full 21-predictor model runs at
  `r fmt(sum(dat_raw$hypertension) / n_par_full, 1)` events per parameter and
  produces **no warning of any kind** - which is precisely the trap. A student who
  fits everything gets an unreadable `r n_par_full`-row table in which about
  `r fmt(0.05 * n_par_full, 1)` coefficients are significant by chance, and no
  diagnostic will tell them anything is wrong. The brief must instruct students to
  count parameters (k levels cost k-1) and cut a priori.
- **MAJOR: the outcome is *diagnosed* hypertension.** Detection depends on
  healthcare contact, which differs by country and by education - exactly two of
  the predictors. This is differential misclassification and it is not
  recoverable from this file (no measured blood pressure is shipped). The
  `diabetes` odds ratio of `r fmt(tid$estimate[tid$term == "diabetesYes"], 2)` is
  partly a healthcare-contact artefact.
- **MAJOR: binned residuals fail** (about
  `r fmt(100 * attributes(br)$resid_ok, 0)`%
  inside bounds against a 95% target, with a "probably bad model fit" warning).
  Unlike small-sample cases this is not a bin-size artefact, because n is
  `r nrow(dat_model)` and the bins are large. A student following the diagnostics
  honestly must report a failed check; the likely cause is non-linear age.
- **MAJOR: the file does not match the paper.** `r nrow(dat_raw)` rows, mean age
  `r fmt(mean(dat$age_yrs), 1)`, median `r median(dat$age_yrs)`, range
  `r min(dat$age_yrs)`-`r max(dat$age_yrs)`; the paper reports 5,639 participants
  with mean age 46.3. Verified directly from the CSV. An 11.5-year gap in mean age
  is far too large to be a subsetting or rounding artefact and suggests the
  deposited S1 file is a different, older subsample. The dataset page should carry
  an explicit caution and students must not be asked to reproduce published
  numbers.
- **MAJOR trap: `angina` must not be a predictor here.** Angina is a consequence
  of long-standing hypertension. Added to this model it returns an odds ratio of
  about
  `r fmt(or_angina, 2)`
  and biases every other coefficient by conditioning on a downstream variable. A
  student building a "cardiovascular risk factor" model will reach for it.
- **MINOR: `alcohol_use` has a `frequent_heavy` level of only
  `r sum(dat_raw$alcohol_use == "frequent_heavy")` participants** with
  `r sum(dat_raw$alcohol_use == "frequent_heavy" & dat_raw$hypertension == 1)`
  hypertension events. Uncollapsed it is estimable but the contrast is very
  imprecise, and a `country * alcohol_use` interaction spends six parameters on
  cells of 9 to 15 people (standard errors up to about 1.0 on the log-odds
  scale). For the depression outcome the same level *does* separate. Pooled with
  `infrequent_heavy` here.
- **MINOR but structural: `education == "none_or_informal"` is a Spain-only
  category.**
  `r sum(dat_raw$education == "none_or_informal" & dat_raw$country == "spain")` of
  its `r sum(dat_raw$education == "none_or_informal")` members are Spanish and
  **zero** are Finnish, so `country * education` produces textbook separation - a
  coefficient of about -10.7 with a standard error of about 197, alongside a
  "fitted probabilities numerically 0 or 1" warning. This is the most likely way a
  student wrecks this outcome. Collapsed to two levels here.
- **MINOR: crude current-smoking OR is protective**
  (`r ug("smoking_statusCurrent")`, p `r up("smoking_statusCurrent")`) and becomes
  null on adjustment. Good teaching material, but a weak student will report
  "smoking protects against hypertension" from the univariable table.
- **MINOR: unmodelled effect modification by country.** country-by-education
  likelihood-ratio
  p `r pv(anova(fit, update(fit, . ~ . + country:education), test = "LRT")[["Pr(>Chi)"]][2])`;
  country-by-age and country-by-BMI are not significant. Worth a sentence, not a
  model change.
- **Unweighted survey analysis.** Must be stated in Methods and Limitations.
