---
title: "Quality of life among adults in Finland, Poland and Spain"
subtitle: "Dataset 14: Quality of life among adults in Finland, Poland and Spain (COURAGE)"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "ageing_qol_europe.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names. The catalogue suggests
# 21 predictors, which cost 39 parameters; the set below costs 25. The cut is
# argued in Methods.
outcome    <- "quality_of_life_score"
predictors <- c("age_yrs", "sex", "country", "education", "marital_status",
                "bmi_category", "physical_activity", "social_network_index",
                "home_usability", "body_pain", "sleep_difficulty",
                "arthritis", "angina", "diabetes", "depression_diagnosis")

lab <- function(x, l) { attr(x, "label") <- l; x }
yn  <- function(x, l) lab(factor(x, levels = c(0, 1), labels = c("No", "Yes")), l)

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest group, or the clinically
    # "unexposed" one where that is the more natural comparison.
    age_yrs = lab(age_yrs, "Age (years)"),
    sex     = lab(factor(sex, levels = c("female", "male"),
                         labels = c("Female", "Male")), "Sex"),
    country = lab(factor(country, levels = c("poland", "spain", "finland"),
                         labels = c("Poland", "Spain", "Finland")), "Country"),
    # education: `none_or_informal` (n = 125) is collapsed into the lower
    # category because 124 of those 125 participants live in Spain and none live
    # in Finland, so its contrast would be a Spain-only contrast entangled with
    # country. See Methods.
    education = lab(factor(if_else(education == "high_school_or_higher",
                                   "High school or higher",
                                   "None, primary or secondary"),
                           levels = c("None, primary or secondary",
                                      "High school or higher")),
                    "Education"),
    marital_status = lab(factor(marital_status,
      levels = c("married_or_cohabiting", "never_married",
                 "separated_or_divorced", "widowed"),
      labels = c("Married or cohabiting", "Never married",
                 "Separated or divorced", "Widowed")), "Marital status"),
    bmi_category = lab(factor(bmi_category,
      levels = c("normal_or_underweight", "overweight", "obese"),
      labels = c("Normal or underweight", "Overweight", "Obese")),
      "BMI category"),
    physical_activity = lab(factor(physical_activity,
      levels = c("high", "moderate", "low"),
      labels = c("High", "Moderate", "Low")), "Physical activity"),
    social_network_index = lab(social_network_index, "Social network index (0-100)"),
    home_usability = lab(home_usability, "Home usability and safety (0-100)"),
    body_pain = lab(factor(body_pain,
      levels = c("none", "pain_without_difficulty", "mild_difficulty",
                 "moderate_difficulty", "severe_or_extreme_difficulty"),
      labels = c("None", "Pain without difficulty", "Mild difficulty",
                 "Moderate difficulty", "Severe or extreme difficulty")),
      "Bodily pain"),
    sleep_difficulty = lab(factor(sleep_difficulty,
      levels = c("none", "mild", "moderate", "severe_or_extreme"),
      labels = c("None", "Mild", "Moderate", "Severe or extreme")),
      "Difficulty sleeping"),
    arthritis = yn(arthritis, "Diagnosed arthritis"),
    angina    = yn(angina, "Diagnosed angina"),
    diabetes  = yn(diabetes, "Diagnosed diabetes"),
    depression_diagnosis = yn(depression_diagnosis, "Diagnosed depression"),
    quality_of_life_score = lab(quality_of_life_score, "WHOQOL-AGE score (0-100)")
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
pv  <- function(p) ifelse(p < 0.001, "<0.001", paste0("= ", fmt(p, 3)))
skew <- function(x) { m <- mean(x); sum((x - m)^3) / length(x) / sd(x)^3 }

# Parameter accounting for the full catalogue predictor list, used in Methods.
all_suggested <- c("age_yrs", "sex", "country", "residence", "marital_status",
                   "education", "bmi_category", "smoking_status", "alcohol_use",
                   "physical_activity", "body_pain", "sleep_difficulty", "sadness",
                   "distance_vision_difficulty", "social_network_index",
                   "neighbourhood_usability", "home_usability", "arthritis",
                   "angina", "diabetes", "depression_diagnosis")
n_par_full <- length(coef(lm(reformulate(all_suggested, outcome),
                             data = mutate(dat_raw, across(where(is.character), factor))))) - 1L
```

# Introduction

COURAGE in Europe is a cross-sectional survey of the general community-dwelling
adult population of Finland, Poland and Spain, carried out to describe healthy
ageing and its social and environmental determinants; the analysis reported by
Miret and colleagues in *PLOS ONE* (2016) used it to model quality of life across
the three countries. The file distributed with this catalogue holds
`r nrow(dat_raw)` participants aged `r min(dat$age_yrs)` to `r max(dat$age_yrs)`
(mean `r fmt(mean(dat$age_yrs), 1)` years), each with a WHOQOL-AGE quality-of-life
score. Quality of life is the outcome that health systems for ageing populations
are ultimately trying to protect, and it is not well predicted by diagnosis
counts alone, so establishing which social, environmental and symptom
characteristics track with it is a standard descriptive question.

**Aim.** Among these adults in Finland, Poland and Spain, we examined which of
age, sex, country, education, marital status, BMI category, physical activity,
social network, home usability, bodily pain, sleep difficulty, arthritis, angina,
diabetes and diagnosed depression are independently associated with the
WHOQOL-AGE score, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** COURAGE in Europe, a cross-sectional multistage stratified
cluster sample of the community-dwelling adult populations of Finland, Poland and
Spain, fielded between 2011 and 2012. The deposited S1 file used here contains
`r nrow(dat_raw)` participants and `r ncol(dat_raw)` variables:
`r sum(dat$country == "Poland")` from Poland, `r sum(dat$country == "Spain")` from
Spain and `r sum(dat$country == "Finland")` from Finland;
`r fmt(100 * mean(dat$sex == "Female"), 1)`% are women.

**Variables.** The outcome is `quality_of_life_score`, the WHOQOL-AGE score,
a continuous index scaled from 0 (worst) to 100 (best) and treated as continuous.
In this file it runs from `r fmt(min(dat$quality_of_life_score), 1)` to
`r fmt(max(dat$quality_of_life_score), 1)` with mean
`r fmt(mean(dat$quality_of_life_score), 1)` and standard deviation
`r fmt(sd(dat$quality_of_life_score), 1)`; the distribution is
**left-skewed** (skewness `r fmt(skew(dat$quality_of_life_score), 2)`) and
bunched toward the upper end, with `r sum(dat$quality_of_life_score >= 95)`
participants scoring 95 or above and `r sum(dat$quality_of_life_score == 100)`
scoring exactly 100. That ceiling matters for the diagnostics and is returned to
below.

Predictors are age (years), sex (reference female), country (reference Poland,
the largest group), education dichotomised as high school or higher versus none,
primary or secondary (reference the latter), marital status (reference married or
cohabiting), BMI category (reference normal or underweight), physical activity
(reference high), the social network index (0-100), home usability and safety
(0-100), bodily pain in five ordered categories (reference none), difficulty
sleeping in four ordered categories (reference none), and diagnosed arthritis,
angina, diabetes and depression (each reference no).

**The predictor set is deliberately cut.** The catalogue suggests 21 predictors.
Most are multi-level factors, so entering all of them costs **`r n_par_full`
estimated parameters**, not 21: `body_pain` alone costs four, and
`marital_status`, `alcohol_use`, `sleep_difficulty`, `sadness` and
`distance_vision_difficulty` cost three each. With `r nrow(dat_raw)` observations
that model is not *arithmetically* infeasible - it is about
`r fmt(nrow(dat_raw) / n_par_full, 0)` observations per parameter, comfortably
inside the 10-15 rule - and it should be said plainly that the parameter budget is
not what forces the cut for a linear model on this sample. Three other things do.
First, a `r n_par_full`-row coefficient table cannot be read, and a Purpose 3
question ("which predictors survive mutual adjustment?") is only answerable for a
set small enough to think about. Second, testing `r n_par_full` coefficients at
the 0.05 level means expecting roughly `r fmt(0.05 * n_par_full, 1)` significant
results by chance alone. Third, several of the suggested predictors are not
plausibly causes of quality of life at all. The set used here costs **25
parameters**, i.e. `r fmt(nrow(dat_raw) / 25, 0)` observations per parameter, and
was fixed before any model was fitted.

Six suggested predictors are dropped, each for a stated reason. `sadness` is
excluded because it is the single most direct component of the construct being
measured - a quality-of-life score that asks how satisfied you are with your life
will move almost mechanically with how sad you report feeling - and because it
overlaps heavily with `depression_diagnosis`, which is retained as the more
clearly clinical measure. `neighbourhood_usability` is excluded because it
correlates `r fmt(cor(dat$neighbourhood_usability, dat$home_usability), 2)` with
`home_usability` and only `r fmt(cor(dat$neighbourhood_usability, dat$quality_of_life_score), 2)`
with the outcome, so it adds an environment scale without adding information;
`home_usability` is kept as the stronger of the two. `distance_vision_difficulty`
costs three parameters for a functional limitation whose smallest stratum holds
`r sum(dat_raw$distance_vision_difficulty == "severe_or_extreme")` people.
`residence` is excluded because urban/rural status here is largely a country-level
design feature (`r fmt(100 * mean(dat$residence[dat$country == "Spain"] == "urban"), 0)`%
of Spanish participants are urban against
`r fmt(100 * mean(dat$residence[dat$country == "Poland"] == "urban"), 0)`% of
Polish participants), so it competes with `country`. `smoking_status` and
`alcohol_use` are excluded as behaviours whose association with a self-reported
quality-of-life index is indirect, and in the case of `alcohol_use` because its
`frequent_heavy` level contains only
`r sum(dat_raw$alcohol_use == "frequent_heavy")` participants.

**Level collapsing.** Five categorical predictors were checked for sparse
strata. `education` had `r sum(dat_raw$education == "none_or_informal")`
participants with no or informal education, of whom
`r sum(dat_raw$education == "none_or_informal" & dat_raw$country == "spain")` live
in Spain and **none** in Finland; it was collapsed into the lower of the two
remaining categories, because a level that exists in one country only cannot be
separated from country. `marital_status` (smallest level
`r min(table(dat$marital_status))`), `bmi_category` (`r min(table(dat$bmi_category))`)
and `physical_activity` (`r min(table(dat$physical_activity))`) were all large
enough to keep as supplied. `alcohol_use` would have needed collapsing (its
`frequent_heavy` level has `r sum(dat_raw$alcohol_use == "frequent_heavy")`
participants) but is dropped from this model for the reason above.

**Statistical analysis.** `lm()`, justified because the outcome is a continuous
score. We fitted a univariable model for each predictor first, then a single
multivariable model containing all of them. Predictors were chosen in advance on
subject-matter grounds, not by significance. Significance threshold 0.05;
estimates are reported with 95% confidence intervals. Assumptions were checked
with `performance::check_model()`, and because non-constant variance was flagged
we additionally report HC3 heteroscedasticity-consistent standard errors. A
pre-specified sensitivity model omitting the three symptom-and-diagnosis
predictors (`body_pain`, `sleep_difficulty`, `depression_diagnosis`) is reported
alongside, because those three are arguably components of quality of life rather
than causes of it. Models are **unweighted**: the COURAGE sampling weights,
strata and clusters are not carried in this file, so no result here is a
nationally representative estimate. Analysis used `r R.version.string` with
tidyverse, gtsummary and performance.

**Missing data.** There is none. All `r ncol(dat_raw)` columns are 100% complete
in this file and `r sum(complete.cases(dat_raw))` of `r nrow(dat_raw)` rows are
complete on every variable, so the complete-case analysis uses
`r nrow(dat_model)` of `r nrow(dat)` rows (`r n_dropped` dropped) and listwise
deletion costs nothing. This is unusual and worth noting: it means the predictor
set could have been expanded at no cost in sample size, so the cut above rests
entirely on interpretability and causal-structure grounds.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(missing = "no") |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

The sample is old on average (median age `r median(dat_model$age_yrs)`, upper
quartile `r quantile(dat_model$age_yrs, 0.75)`) and predominantly female, and it
is dominated by the Polish and Spanish sites. Median WHOQOL-AGE score is
`r fmt(median(dat_model$quality_of_life_score), 1)` with an interquartile range of
`r fmt(quantile(dat_model$quality_of_life_score, 0.25), 1)` to
`r fmt(quantile(dat_model$quality_of_life_score, 0.75), 1)`. Symptom burden is
substantial: `r fmt(100 * mean(dat_model$body_pain != "None"), 0)`% report some
bodily pain and `r fmt(100 * mean(dat_model$sleep_difficulty != "None"), 0)`%
some difficulty sleeping, while
`r fmt(100 * mean(dat_model$depression_diagnosis == "Yes"), 1)`% report a
diagnosis of depression.

## Exploratory figure

```{r explore, fig.cap="Figure 1. WHOQOL-AGE score against age, by country, with loess smooths."}
ggplot(dat_model, aes(age_yrs, quality_of_life_score, colour = country)) +
  geom_point(alpha = 0.15, size = 0.8) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 1.1) +
  scale_colour_manual(values = c(Poland = "#B5495B", Spain = "#E0A458",
                                 Finland = "#3F7CAC")) +
  labs(x = "Age (years)", y = "WHOQOL-AGE score (0-100)", colour = "Country")
```

Quality of life declines with age in all three countries, but the three curves are
separated by roughly the same amount at every age and the Finnish curve is
markedly higher and flatter than the Polish one; the decline in Poland is steeper
and begins earlier, which is a first hint that country modifies the age gradient
rather than merely shifting it.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv

utid <- map_dfr(predictors, function(v) {
  tidy(lm(reformulate(v, outcome), data = dat_model), conf.int = TRUE) |>
    filter(term != "(Intercept)")
})
ug <- function(term, d = 2) {
  r <- utid[utid$term == term, ]
  sprintf("%s points (95%% CI %s to %s)", fmt(r$estimate, d),
          fmt(r$conf.low, d), fmt(r$conf.high, d))
}
up <- function(term) pv(utid$p.value[utid$term == term])
```

Bodily pain shows the largest crude association of any predictor: compared with
participants reporting no pain, those with severe or extreme pain-related
difficulty score `r ug("body_painSevere or extreme difficulty")` lower
(p `r up("body_painSevere or extreme difficulty")`), and the gradient is
monotonic across the intermediate categories. Severe or extreme sleep difficulty
is associated with `r ug("sleep_difficultySevere or extreme")`
(p `r up("sleep_difficultySevere or extreme")`) and a depression diagnosis with
`r ug("depression_diagnosisYes")` (p `r up("depression_diagnosisYes")`). Country
is the next strongest: Finnish participants score
`r ug("countryFinland")` higher than Polish participants
(p `r up("countryFinland")`) and Spanish participants
`r ug("countrySpain")` higher (p `r up("countrySpain")`). Each additional point of
the social network index is associated with `r ug("social_network_index")`
(p `r up("social_network_index")`) and each point of home usability with
`r ug("home_usability")` (p `r up("home_usability")`), so across the full 0-100
range of those scales the implied differences are large. Each additional year of
age is associated with `r ug("age_yrs")` (p `r up("age_yrs")`), about
`r fmt(-10 * utid$estimate[utid$term == "age_yrs"], 1)` points per decade.
Widowhood (`r ug("marital_statusWidowed")`, p `r up("marital_statusWidowed")`),
low physical activity (`r ug("physical_activityLow")`,
p `r up("physical_activityLow")`), arthritis (`r ug("arthritisYes")`,
p `r up("arthritisYes")`), angina (`r ug("anginaYes")`, p `r up("anginaYes")`),
diabetes (`r ug("diabetesYes")`, p `r up("diabetesYes")`) and obesity
(`r ug("bmi_categoryObese")`, p `r up("bmi_categoryObese")`) all show clear crude
deficits. Men score `r ug("sexMale")` **higher** than women
(p `r up("sexMale")`) - a result that does not survive adjustment.

## Why adjustment is needed

**Age** is the dominant confounder in this dataset and it operates on almost
every crude estimate. Older participants are far more likely to be widowed, to
report arthritis, angina and diabetes, to have bodily pain and to be in the
low physical-activity category, and they independently score lower on WHOQOL-AGE.
The crude `r ug("marital_statusWidowed")` deficit for widowhood is therefore
substantially a statement about age: the median age of widowed participants is
`r median(dat_model$age_yrs[dat_model$marital_status == "Widowed"])` against
`r median(dat_model$age_yrs[dat_model$marital_status == "Married or cohabiting"])`
for the married or cohabiting. The same argument applies to arthritis, angina and
diabetes, whose crude coefficients are largely age in disguise.

**Country** is the second, and it is structural rather than incidental. Mean
WHOQOL-AGE is `r fmt(mean(dat_model$quality_of_life_score[dat_model$country == "Finland"]), 1)`
in Finland, `r fmt(mean(dat_model$quality_of_life_score[dat_model$country == "Spain"]), 1)`
in Spain and `r fmt(mean(dat_model$quality_of_life_score[dat_model$country == "Poland"]), 1)`
in Poland, and the three national samples differ sharply in education
(`r fmt(100 * mean(dat_model$education[dat_model$country == "Finland"] == "High school or higher"), 0)`%
of Finnish participants have high school or higher against
`r fmt(100 * mean(dat_model$education[dat_model$country == "Poland"] == "High school or higher"), 0)`%
of Polish participants) and in age. Any crude education or social-network
coefficient is therefore partly a between-country comparison. **Sex** works
differently again: women report more bodily pain
(`r fmt(100 * mean(dat_model$body_pain[dat_model$sex == "Female"] == "Severe or extreme difficulty"), 1)`%
report severe or extreme pain difficulty against
`r fmt(100 * mean(dat_model$body_pain[dat_model$sex == "Male"] == "Severe or extreme difficulty"), 1)`%
of men), so the crude male advantage is a pain difference; holding pain constant
should reverse it. Mutual adjustment lets each coefficient be read as a comparison
between participants alike on the other measured characteristics.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv

tid <- tidy(fit, conf.int = TRUE)
g <- function(term, d = 2) {
  r <- tid[tid$term == term, ]
  sprintf("%s points (95%% CI %s to %s)", fmt(r$estimate, d),
          fmt(r$conf.low, d), fmt(r$conf.high, d))
}
gp <- function(term) pv(tid$p.value[tid$term == term])
```

```{r fit-stats}
s <- summary(fit)
tibble(
  Statistic = c("Complete-case N", "Parameters (excl. intercept)",
                "R-squared", "Adjusted R-squared", "Residual SE (points)"),
  Value = c(as.character(nrow(dat_model)),
            as.character(length(coef(fit)) - 1L),
            fmt(s$r.squared, 3), fmt(s$adj.r.squared, 3), fmt(s$sigma, 2))
) |>
  gt() |>
  tab_header(title = md("**Model fit: multivariable linear model for WHOQOL-AGE score**"))
```

The model explains `r fmt(100 * s$r.squared, 1)`% of the variance in the
WHOQOL-AGE score (adjusted R-squared `r fmt(s$adj.r.squared, 3)`), which is high
for a model of a self-reported score; the gap between the raw and adjusted values
is negligible because `r length(coef(fit)) - 1L` parameters on
`r nrow(dat_model)` observations is a modest spend. Residual standard deviation is
`r fmt(s$sigma, 1)` points against an outcome standard deviation of
`r fmt(sd(dat_model$quality_of_life_score), 1)`.

Among participants of the same age, sex, country, education, marital status, BMI
category, activity level, social network, home usability, sleep and diagnosed
conditions, severe or extreme pain-related difficulty is associated with
`r g("body_painSevere or extreme difficulty")` lower quality of life
(p `r gp("body_painSevere or extreme difficulty")`), and the gradient across pain
categories remains monotonic. Severe or extreme sleep difficulty is associated
with `r g("sleep_difficultySevere or extreme")`
(p `r gp("sleep_difficultySevere or extreme")`) and a depression diagnosis with
`r g("depression_diagnosisYes")` (p `r gp("depression_diagnosisYes")`). Being
Finnish rather than Polish remains worth `r g("countryFinland")`
(p `r gp("countryFinland")`) - the largest single contrast in the model after
severe pain, and one that no combination of individual-level characteristics in
this file explains away. Each point of the social network index is worth
`r g("social_network_index")` (p `r gp("social_network_index")`) and each point of
home usability `r g("home_usability")` (p `r gp("home_usability")`), i.e. about
`r fmt(20 * tid$estimate[tid$term == "social_network_index"], 1)` and
`r fmt(20 * tid$estimate[tid$term == "home_usability"], 1)` points respectively
across a 20-point difference on those scales. Low physical activity remains
associated with `r g("physical_activityLow")`
(p `r gp("physical_activityLow")`) and high school education or above with
`r g("educationHigh school or higher")` (p `r gp("educationHigh school or higher")`).
Age falls to `r g("age_yrs")` per year (p `r gp("age_yrs")`), about
`r fmt(-10 * tid$estimate[tid$term == "age_yrs"], 1)` points per decade. Arthritis
(`r g("arthritisYes")`, p `r gp("arthritisYes")`) and both BMI categories are no
longer independently associated, and men now score `r g("sexMale")` **lower** than
women (p `r gp("sexMale")`), the reverse of the crude direction. These are
associations among the participants observed, not effects of intervening.

**Statistical against practical significance.** With `r nrow(dat_model)`
observations this model has enough power to detect differences that would not
change anyone's decision. The sex coefficient is a good example: at
`r g("sexMale")` on a 0-100 scale it is formally significant
(p `r gp("sexMale")`) but amounts to roughly
`r fmt(100 * abs(tid$estimate[tid$term == "sexMale"]) / sd(dat_model$quality_of_life_score), 0)`%
of one outcome standard deviation, which is not a difference a clinician or a
policymaker would act on. The same caution applies to the angina
(`r g("anginaYes")`) and diabetes (`r g("diabetesYes")`) contrasts. Contrast those
with severe pain (`r g("body_painSevere or extreme difficulty")`, close to
`r fmt(abs(tid$estimate[tid$term == "body_painSevere or extreme difficulty"]) / sd(dat_model$quality_of_life_score), 1)`
standard deviations) or the Finland contrast (`r g("countryFinland")`), which are
large enough to matter substantively. Significance and importance are different
questions, and at this sample size only the second is informative.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

Several estimates move a long way. **Arthritis** collapses from
`r ug("arthritisYes")` crude to `r g("arthritisYes")` adjusted
(p `r up("arthritisYes")` to p `r gp("arthritisYes")`), i.e. essentially all the
way to the null and out of significance: the crude arthritis deficit was age and
bodily pain, and once pain is in the model arthritis has nothing left to explain.
**Widowhood** does the same thing and then some, from `r ug("marital_statusWidowed")`
crude to `r g("marital_statusWidowed")` adjusted - the sign reverses and the
estimate becomes non-significant, which is age confounding almost in its pure
form. **Obesity** moves from `r ug("bmi_categoryObese")` (p `r up("bmi_categoryObese")`)
to `r g("bmi_categoryObese")` (p `r gp("bmi_categoryObese")`), again explained by
pain and comorbidity. **Sex** is the most striking: `r ug("sexMale")` crude,
significant in the direction of men doing better, becomes `r g("sexMale")`
adjusted, significant in the direction of men doing worse. This is masking rather
than confounding in the usual sense - women's higher pain burden was hiding a
small female advantage - and it is a good illustration of why a crude sex
difference in a symptom-laden score should not be reported on its own.

By contrast, **country** barely moves: the Finland contrast goes only from
`r ug("countryFinland")` to `r g("countryFinland")`, retaining about
`r fmt(100 * tid$estimate[tid$term == "countryFinland"] / utid$estimate[utid$term == "countryFinland"], 0)`%
of its crude size, and it stays the second-largest term in the model. Nothing
measured at the individual level in this file accounts for the national gap.
**Severe bodily pain** also retains about
`r fmt(100 * tid$estimate[tid$term == "body_painSevere or extreme difficulty"] / utid$estimate[utid$term == "body_painSevere or extreme difficulty"], 0)`%
of its crude magnitude and **social network index** about
`r fmt(100 * tid$estimate[tid$term == "social_network_index"] / utid$estimate[utid$term == "social_network_index"], 0)`%.
Two contrasts change significance status in opposite directions. Reporting pain
*without* associated difficulty crosses **into** significance, from
`r ug("body_painPain without difficulty")` (p `r up("body_painPain without difficulty")`)
crude to `r g("body_painPain without difficulty")`
(p `r gp("body_painPain without difficulty")`) adjusted: participants in the
no-pain reference group have a median age of
`r median(dat_model$age_yrs[dat_model$body_pain == "None"])` against
`r median(dat_model$age_yrs[dat_model$body_pain == "Pain without difficulty"])` in
the pain-without-difficulty group, so holding age constant unmasks a small
deficit. Never-married status crosses the other way, from
`r ug("marital_statusNever married")` (p `r up("marital_statusNever married")`) to
`r g("marital_statusNever married")` (p `r gp("marital_statusNever married")`),
which is age confounding once more - never-married participants have a median age
of `r median(dat_model$age_yrs[dat_model$marital_status == "Never married"])`,
`r median(dat_model$age_yrs[dat_model$marital_status == "Married or cohabiting"]) - median(dat_model$age_yrs[dat_model$marital_status == "Never married"])`
years younger than the married or cohabiting.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r het-check}
het <- check_heteroscedasticity(fit)
print(het)
nor <- check_normality(fit)
print(nor)
```

```{r collin}
check_collinearity(fit)
```

```{r robust-se}
rob   <- lmtest::coeftest(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
robci <- lmtest::coefci(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
n_flip <- sum((summary(fit)$coefficients[, 4] < 0.05) != (rob[, 4] < 0.05))
tibble(
  Term = rownames(rob),
  Estimate = fmt(rob[, 1], 3),
  `SE (OLS)` = fmt(summary(fit)$coefficients[, 2], 3),
  `SE (HC3)` = fmt(rob[, 2], 3),
  `95% CI (HC3)` = paste0(fmt(robci[, 1], 2), " to ", fmt(robci[, 2], 2)),
  `p (HC3)` = fmt(rob[, 4], 3)
) |>
  filter(Term != "(Intercept)") |>
  gt() |>
  tab_header(title = md("**Table 3. HC3 heteroscedasticity-consistent standard errors.**"))
```

The linearity and posterior-predictive panels are acceptable but not clean. The
observed outcome has a ceiling - `r sum(dat_model$quality_of_life_score == 100)`
participants score exactly 100 and `r sum(dat_model$quality_of_life_score == 75)`
score exactly 75, a scoring artefact of the instrument - and the fitted normal
predictive distribution cannot reproduce either feature. The residual-versus-fitted
panel shows the spread narrowing at high fitted values, which is what a ceiling
produces, and `check_heteroscedasticity()` rejects constant variance
(p < 0.001). Residuals are left-skewed
(skewness `r fmt(skew(resid(fit)), 2)`, milder than the raw outcome's
`r fmt(skew(dat_model$quality_of_life_score), 2)`) and `check_normality()` also
rejects; with `r nrow(dat_model)` observations both tests would reject trivial
departures, so the QQ plot matters more than the p-value, and it shows a heavy
lower tail rather than a wholesale failure. Collinearity is not a problem: the
largest VIF is
`r fmt(max(as.data.frame(check_collinearity(fit))$VIF), 2)` for
`r as.data.frame(check_collinearity(fit))$Term[which.max(as.data.frame(check_collinearity(fit))$VIF)]`,
well below the conventional threshold of 5, and the pain, sleep and depression
terms are not competing as badly as their substantive overlap would suggest.
Influence is negligible: the largest Cook's distance is
`r fmt(max(cooks.distance(fit)), 4)` and the largest leverage
`r fmt(max(hatvalues(fit)), 3)`, so no participant drives the fit.

Table 3 reports HC3 robust standard errors because variance is non-constant. They
change very little - the largest ratio of HC3 to OLS standard error is
`r fmt(max(rob[, 2] / summary(fit)$coefficients[, 2]), 2)` - and
**`r n_flip`** coefficients change significance status under robust inference, so
the substantive conclusions stand. A transformation would help the skew (squaring
the score reduces residual skewness to about 0.21) but at the cost of coefficients
that no longer read in points of WHOQOL-AGE, and it does not change which
predictors matter; we therefore report the untransformed model and flag the
violation.

```{r sensitivity}
sens_preds <- setdiff(predictors, c("body_pain", "sleep_difficulty",
                                    "depression_diagnosis"))
fit_s <- lm(reformulate(sens_preds, outcome), data = dat_model)
ss <- summary(fit_s)
tids <- tidy(fit_s, conf.int = TRUE)
gs <- function(term, d = 2) {
  r <- tids[tids$term == term, ]
  sprintf("%s points (95%% CI %s to %s)", fmt(r$estimate, d),
          fmt(r$conf.low, d), fmt(r$conf.high, d))
}
tibble(
  Model = c("Full model (25 parameters)",
            "Without pain, sleep and depression (18 parameters)"),
  `R-squared` = c(fmt(s$r.squared, 3), fmt(ss$r.squared, 3)),
  `Age (per year)` = c(fmt(tid$estimate[tid$term == "age_yrs"], 3),
                       fmt(tids$estimate[tids$term == "age_yrs"], 3)),
  `Male vs female` = c(fmt(tid$estimate[tid$term == "sexMale"], 2),
                       fmt(tids$estimate[tids$term == "sexMale"], 2)),
  `Finland vs Poland` = c(fmt(tid$estimate[tid$term == "countryFinland"], 2),
                          fmt(tids$estimate[tids$term == "countryFinland"], 2)),
  `Obese vs normal` = c(fmt(tid$estimate[tid$term == "bmi_categoryObese"], 2),
                        fmt(tids$estimate[tids$term == "bmi_categoryObese"], 2))
) |>
  gt() |>
  tab_header(title = md("**Table 4. Sensitivity analysis: dropping the symptom and diagnosis predictors.**"))
```

Table 4 is the analysis that decides how much of this model to believe. Dropping
`body_pain`, `sleep_difficulty` and `depression_diagnosis` takes R-squared from
`r fmt(s$r.squared, 3)` to `r fmt(ss$r.squared, 3)`, so those three predictors
alone account for
`r fmt(100 * (s$r.squared - ss$r.squared), 1)` percentage points of explained
variance, roughly
`r fmt(100 * (s$r.squared - ss$r.squared) / s$r.squared, 0)`% of everything the
model explains. Without them, age recovers to `r gs("age_yrs")` per year, obesity becomes
`r gs("bmi_categoryObese")` and significant again, sex reverts to men scoring
`r gs("sexMale")` higher, and the Finland contrast grows to `r gs("countryFinland")`.
Every one of those changes is what you would expect if the symptom variables sit
*between* the demographic predictors and the outcome, or are part of the outcome
itself.

# Discussion

Among these `r nrow(dat_model)` adults in Finland, Poland and Spain, the
characteristics most strongly and independently associated with a lower
WHOQOL-AGE score are bodily pain (up to
`r fmt(-tid$estimate[tid$term == "body_painSevere or extreme difficulty"], 1)`
points for severe or extreme pain-related difficulty), difficulty sleeping
(up to `r fmt(-tid$estimate[tid$term == "sleep_difficultySevere or extreme"], 1)`
points) and a diagnosis of depression
(`r fmt(-tid$estimate[tid$term == "depression_diagnosisYes"], 1)` points), while a
richer social network and a more usable, safer home are associated with higher
scores. Country remains the second-largest contrast in the model: Finnish
participants score `r fmt(tid$estimate[tid$term == "countryFinland"], 1)` points
above otherwise similar Polish participants, and that gap is not explained by any
individual characteristic measured here. The crude associations with arthritis,
obesity, widowhood and being male all disappear on adjustment. Practically, the
model points at pain, sleep and mood - all modifiable, all routinely
under-treated in older adults - rather than at diagnosis lists, as the levers most
associated with self-reported quality of life; but see the reflection problem
below before reading that as a treatment recommendation.

**The reflection problem is the main interpretive weakness of this analysis.**
WHOQOL-AGE is a summary of how people rate their own physical health,
psychological state, social relationships and environment. `body_pain`,
`sleep_difficulty` and `sadness` are not distant causes of that rating - they are
close to being the things the rating is made of. Asking someone how much pain
interferes with their life and then asking them to rate their quality of life is
asking two versions of the same question, and regressing the second on the first
recovers the arithmetic of the instrument as much as any causal relationship. The
same argument, more weakly, applies to `social_network_index` and
`home_usability`, which measure the social-relationships and environment domains
that WHOQOL-AGE itself contains. Table 4 quantifies the problem: the three
symptom-and-diagnosis predictors carry
`r fmt(100 * (s$r.squared - ss$r.squared), 1)` percentage points of R-squared on
their own, and removing them changes the direction of the sex coefficient and
restores significance to obesity. Even where the relationship is genuinely causal
rather than definitional, the direction is unclear in a cross-sectional survey:
poor quality of life plausibly worsens sleep and amplifies reported pain as much
as the reverse. The honest reading is that the pain, sleep and depression
coefficients describe how the components of a quality-of-life score hang together,
and that the demographic and structural coefficients (age, sex, country,
education, obesity) are the ones with a defensible causal interpretation - and
those should be read from the sensitivity model in Table 4, not from the full
model, because the full model conditions on variables that sit on the pathway.

**Limitations.** The models are **unweighted**: the COURAGE sampling weights,
strata and clusters are not carried in this file, so the sample must not be
described as representative of any national population and the standard errors
ignore clustering within primary sampling units, which will make them too small.
The design is cross-sectional, so temporality is unknown and reverse causation is
live for every symptom predictor, as set out above; adjusting for pain, sleep and
depression is probably over-adjustment for mediators rather than confounder
control. The deposited file differs from the published paper: it holds
`r nrow(dat_raw)` rows with mean age `r fmt(mean(dat$age_yrs), 1)` years, whereas
the source paper reports 5,639 participants with mean age 46.3 years, so this is
not the analytic sample the authors modelled and results here should not be
compared with theirs. Residual variance is non-constant and residuals are
left-skewed because the outcome has a ceiling at 100 and a large spike at 75; HC3
standard errors address inference (changing `r n_flip` conclusions) but not the
model's inability to represent the top of the distribution. There is evidence of
effect modification by country that the additive model ignores: a country-by-age
interaction improves the fit (p `r pv(anova(fit, update(fit, . ~ . + country:age_yrs))$"Pr(>F)"[2])`)
and a country-by-education interaction does too
(p `r pv(anova(fit, update(fit, . ~ . + country:education))$"Pr(>F)"[2])`), so the
single age slope reported here is an average across three quite different
national gradients. Age enters linearly although a quadratic term is significant
(p `r pv(anova(fit, update(fit, . ~ . + I(age_yrs^2)))$"Pr(>F)"[2])`). The
education variable was collapsed to two levels, so any effect confined to the
`r sum(dat_raw$education == "none_or_informal")` participants with no formal
education is invisible. Six suggested predictors were excluded a priori, so
residual confounding by smoking, alcohol, vision and neighbourhood environment is
certain. With `r nrow(dat_model)` observations, small and practically
uninteresting differences reach significance. All estimates are associations, not
causal effects.

**Future work.** Refit with the COURAGE survey design (weights, strata and
clusters), with country-specific age slopes, and with the WHOQOL-AGE domains
separated from their putative determinants so that pain and sleep are not
simultaneously predictors and constituents of the outcome.

# Feasibility notes

*NOT part of a student report.*

- **USABLE, and the strongest of the three outcomes in this dataset.** No
  missing data at all (`r sum(complete.cases(dat_raw))`/`r nrow(dat_raw)` rows
  complete on every column), n = `r nrow(dat_model)`, R-squared
  `r fmt(s$r.squared, 3)`, no separation risk, no influential points
  (max Cook's `r fmt(max(cooks.distance(fit)), 4)`), max VIF
  `r fmt(max(as.data.frame(check_collinearity(fit))$VIF), 2)`.
- **MAJOR: the reflection problem.** `body_pain`, `sleep_difficulty` and
  `sadness` are components or near-components of the WHOQOL-AGE construct, and
  `social_network_index` and `home_usability` overlap with its social and
  environment domains. Dropping the first three takes R-squared from
  `r fmt(s$r.squared, 3)` to `r fmt(ss$r.squared, 3)`. A student who includes them
  and writes "pain reduces quality of life" has stated something close to a
  tautology. This is the single most important thing to warn about for this
  dataset, and it should probably be flagged on the dataset page.
- **MAJOR: 21 suggested predictors cost `r n_par_full` parameters, not 21.**
  Most are multi-level factors. For the linear outcome the parameter budget is
  not violated (`r fmt(nrow(dat_raw) / n_par_full, 0)` observations per
  parameter), so a student *can* fit all 21 without any warning appearing - which
  is exactly the trap. The result is an unreadable table in which roughly
  `r fmt(0.05 * n_par_full, 1)` coefficients are significant by chance. The brief
  must tell students to count parameters and cut.
- **MAJOR: the file does not match the paper.** `r nrow(dat_raw)` rows, mean age
  `r fmt(mean(dat$age_yrs), 1)` years, median `r median(dat$age_yrs)`, range
  `r min(dat$age_yrs)`-`r max(dat$age_yrs)`; the paper reports 5,639 participants
  with mean age 46.3. Verified directly from the CSV. The age discrepancy of
  ~11.5 years is far too large to be a rounding or subsetting artefact and
  suggests the deposited S1 file is a different (older) subsample from the one
  analysed. Students must not be asked to reproduce published numbers, and the
  dataset page should carry an explicit caution.
- **MAJOR: failed diagnostics.** Heteroscedasticity p < 0.001 and non-normality
  p < 0.001, both driven by the ceiling: `r sum(dat_model$quality_of_life_score == 100)`
  participants score exactly 100 and `r sum(dat_model$quality_of_life_score == 75)`
  score exactly 75 (a single scoring value accounting for
  `r fmt(100 * mean(dat_model$quality_of_life_score == 75), 1)`% of the sample).
  HC3 changes `r n_flip` conclusions, so this is a reportable violation rather
  than a threat to the findings.
- **MINOR but structural: `education == "none_or_informal"` is a Spain-only
  category.** `r sum(dat_raw$education == "none_or_informal" & dat_raw$country == "spain")`
  of its `r sum(dat_raw$education == "none_or_informal")` members are Spanish and
  **zero** are Finnish. Kept as three levels alongside `country` it is estimable
  but the contrast is a within-Spain contrast; interacted with `country` it
  produces outright separation. Collapsed to two levels here.
- **MINOR: `alcohol_use` has a `frequent_heavy` level of
  `r sum(dat_raw$alcohol_use == "frequent_heavy")` participants** (9 with
  depression, 11 with hypertension). Harmless in a linear model, dangerous in the
  logistic ones - see the other two documents.
- **MINOR: unmodelled effect modification by country.** country-by-age
  p `r pv(anova(fit, update(fit, . ~ . + country:age_yrs))$"Pr(>F)"[2])`,
  country-by-education
  p `r pv(anova(fit, update(fit, . ~ . + country:education))$"Pr(>F)"[2])`.
  The additive model is a simplification, not a description of three parallel
  populations. A good student can mention it; the workshop does not require
  interactions.
- **MINOR: non-linear age.** A quadratic age term is significant
  (p `r pv(anova(fit, update(fit, . ~ . + I(age_yrs^2)))$"Pr(>F)"[2])`), so the
  linear age coefficient understates the decline at the oldest ages.
- **Unweighted survey analysis.** Must be stated in Methods and Limitations for
  every outcome in this dataset.
