---
title: "Carotid atherosclerotic plaque in adults with and without rheumatoid arthritis"
subtitle: "Dataset 10: Cardiovascular risk and carotid artery thickness in rheumatoid arthritis"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "cimt_rheumatoid.csv"),
                    show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names. The predictor set is
# capped at six parameters because there are only 65 plaque events in the file:
# see the parameter-budget paragraph in Methods.
outcome    <- "carotid_plaque"
predictors <- c("age_yrs", "sex", "has_ra", "systolic_bp", "current_smoker",
                "total_cholesterol")

lab <- function(x, l) { attr(x, "label") <- l; x }

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the clinically "unexposed" group throughout.
    sex    = lab(factor(sex, levels = c("female", "male"),
                        labels = c("Female", "Male")), "Sex"),
    has_ra = lab(factor(has_ra, levels = c(0, 1),
                        labels = c("Control", "RA")), "Rheumatoid arthritis"),
    current_smoker    = lab(factor(current_smoker, levels = c(0, 1),
                                   labels = c("No", "Yes")), "Current smoker"),
    statins           = lab(factor(statins, levels = c(0, 1),
                                   labels = c("No", "Yes")), "Taking a statin"),
    antihypertensives = lab(factor(antihypertensives, levels = c(0, 1),
                                   labels = c("No", "Yes")),
                            "Taking antihypertensives"),
    log_crp = lab(log(crp), "log CRP (log mg/L)"),
    age_yrs = lab(age_yrs, "Age (years)"),
    bmi = lab(bmi, "BMI (kg/m2)"),
    systolic_bp = lab(systolic_bp, "Systolic BP (mmHg)"),
    total_cholesterol = lab(total_cholesterol, "Total cholesterol (mmol/L)"),
    hdl_cholesterol = lab(hdl_cholesterol, "HDL cholesterol (mmol/L)"),
    triglycerides = lab(triglycerides, "Triglycerides (mmol/L)"),
    glucose = lab(glucose, "Fasting glucose (mmol/L)"),
    carotid_plaque = lab(carotid_plaque, "Carotid plaque")
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)
n_events  <- sum(dat_model[[outcome]])

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
pv  <- function(p) ifelse(p < 0.001, "<0.001", paste0("= ", fmt(p, 3)))

# Differential missingness of the outcome, used in Methods and the feasibility notes.
miss_ctrl <- sum(is.na(dat_raw$carotid_plaque) & dat_raw$has_ra == 0)
miss_ra   <- sum(is.na(dat_raw$carotid_plaque) & dat_raw$has_ra == 1)
n_ctrl    <- sum(dat_raw$has_ra == 0)
n_ra      <- sum(dat_raw$has_ra == 1)

sex_vals    <- sort(unique(na.omit(dat_raw$sex)))
ht_min      <- min(dat_raw$height_cm, na.rm = TRUE)
ht_max      <- max(dat_raw$height_cm, na.rm = TRUE)
n_ht_metres <- sum(dat_raw$height_cm < 3, na.rm = TRUE)
friedewald  <- dat_raw$total_cholesterol - dat_raw$hdl_cholesterol -
               dat_raw$triglycerides / 2.2
ldl_r       <- cor(dat_raw$ldl_cholesterol, friedewald, use = "complete.obs")
```

# Introduction

The source study recruited adults with rheumatoid arthritis (RA) and healthy
controls without RA in the Netherlands, put every participant through the same
physical examination and fasting laboratory panel, and scanned both carotid
arteries by B-mode ultrasound (Hollan and colleagues, *PLOS ONE*, 2015). As well
as measuring wall thickness the ultrasound recorded whether a discrete
atherosclerotic **plaque** was present. Plaque is a later and clinically harder
endpoint than wall thickening - it is focal, it is what ruptures, and its
presence reclassifies cardiovascular risk in a way that a continuous thickness
measurement does not - so it is worth asking separately whether RA and the
conventional risk factors predict it.

**Aim.** Among these Dutch adults, we examined which of RA status, age, sex,
systolic blood pressure, current smoking and total cholesterol are independently
associated with the presence of carotid plaque, mutually adjusting for the
others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** A cross-sectional case-control comparison conducted in the
Netherlands and published in *PLOS ONE* in 2015. Cases were adults with
established rheumatoid arthritis; controls were adults without RA recruited from
the same setting. This extract contains `r nrow(dat)` participants (`r n_ra`
with RA, `r n_ctrl` controls) and 25 variables. Ages run from
`r min(dat$age_yrs, na.rm = TRUE)` to `r max(dat$age_yrs, na.rm = TRUE)` years.

**Variables.** The outcome is `carotid_plaque`, a binary indicator where the
**event is an atherosclerotic plaque visible on carotid ultrasound**. There are
`r sum(dat_raw$carotid_plaque == 1, na.rm = TRUE)` events among the
`r sum(!is.na(dat_raw$carotid_plaque))` participants scanned
(`r fmt(100 * mean(dat_raw$carotid_plaque, na.rm = TRUE), 1)`%). Predictors are
rheumatoid arthritis status (reference Control), age in years, sex (reference
Female), systolic blood pressure in mmHg, current smoking (reference No) and
total cholesterol in mmol/L. **RA status is the exposure of primary interest**;
the other five are the conventional determinants of atherosclerosis.

**The predictor set is deliberately small, and here is the arithmetic.** The
file contains `r sum(dat_raw$carotid_plaque == 1, na.rm = TRUE)` plaque events
and `r n_events` survive into the complete-case analytic sample. The
conventional rule of **ten events per estimated parameter** therefore allows
`r floor(n_events / 10)` parameters. All six predictors are continuous or
binary, so each costs exactly one degree of freedom and the model estimates
`r length(predictors)` parameters, giving
**`r fmt(n_events / length(predictors), 1)` events per parameter**. That is the
budget spent, with nothing to spare. The `meta.json` predictor list contains 16
variables; fitting all of them would cost 16-17 parameters and drop events per
parameter to about
`r fmt(sum(dat_raw$carotid_plaque == 1, na.rm = TRUE) / 17, 1)`, which is
over-fitting regardless of what the p-values then say. The six were chosen a
priori: age and sex because cases and controls were **not matched** on either
and both are the dominant determinants of plaque; RA status because it is the
exposure of interest; and blood pressure, smoking and total cholesterol because
they are the three modifiable risk factors that any cardiovascular model would
be expected to contain.

**Variables excluded from the candidate pool on structural grounds.**
`ldl_cholesterol` is not independently measured in this file - it is the
Friedewald calculation from total cholesterol, HDL and triglycerides
(correlation `r fmt(ldl_r, 4)` with that formula), so it must never appear
alongside all three. `waist_cm` duplicates `bmi`
(r `r fmt(cor(dat$bmi, dat$waist_cm, use = "complete.obs"), 2)`) and
`diastolic_bp` duplicates `systolic_bp`
(r `r fmt(cor(dat$systolic_bp, dat$diastolic_bp, use = "complete.obs"), 2)`);
`apo_a` and `apo_b` duplicate HDL and total cholesterol. `weight_kg` and
`height_cm` are excluded because `bmi` is exactly
`weight_kg/(height_cm/100)^2`. `statins` and `antihypertensives` are omitted
from *this* model for a different reason - they are confounded by indication and
each would cost a scarce parameter to estimate something that cannot be read as
a treatment effect (see Discussion). `prednisone` is omitted because it is
almost perfectly aligned with RA status
(`r sum(dat_raw$prednisone == 1 & dat_raw$has_ra == 0, na.rm = TRUE)` of
`r sum(dat_raw$prednisone == 1, na.rm = TRUE)` users are controls).

**Statistical analysis.** `glm(family = binomial)`, justified because the
outcome is binary. We fitted a univariable model for each predictor first, then
a single multivariable model containing all six. Predictors were chosen in
advance on subject-matter grounds, not by significance. Coefficients are
exponentiated and reported as odds ratios with 95% confidence intervals;
significance threshold 0.05. Model adequacy was assessed with
`performance::check_model()`, `binned_residuals()` and `check_collinearity()`,
focusing on calibration, collinearity, influential observations and events per
parameter. Analysis used `r R.version.string` with tidyverse, gtsummary and
performance.

**Data integrity.** The catalogue notes that in the original Zenodo deposit
`sex` was coded 0 = woman / 1 = man and the height column mixed metres and
centimetres. Both are corrected here: `sex` ships as the strings
`r paste(sex_vals, collapse = " and ")` and `height_cm` runs from `r ht_min` to
`r ht_max` cm with `r n_ht_metres` values below 3. One defect remains: row
`id = "control_row1"` is entirely blank apart from `id`, `has_ra` and
`hypertension_or_hyperchol`.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped,
`r fmt(100 * n_dropped / nrow(dat), 1)`%), leaving `r n_events` events. Almost
all the loss is the outcome itself: `carotid_plaque` is only
`r fmt(100 * mean(!is.na(dat_raw$carotid_plaque)), 1)`% complete
(`r sum(is.na(dat_raw$carotid_plaque))` missing). **That missingness is severely
differential with respect to the exposure of interest**: `r miss_ra` of `r n_ra`
RA cases (`r fmt(100 * miss_ra / n_ra, 1)`%) lack a plaque reading, against
`r miss_ctrl` of `r n_ctrl` controls (`r fmt(100 * miss_ctrl / n_ctrl, 1)`%).
Nearly a quarter of the control group is therefore absent from this analysis for
reasons we cannot observe, and if plaque status is related to whether a control
was scanned, the RA-versus-control comparison is biased by an unknown amount in
an unknown direction. This is stated again in the Limitations because it is the
most serious weakness of this outcome.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  mutate(carotid_plaque = factor(carotid_plaque, levels = c(0, 1),
                                 labels = c("No plaque", "Plaque"))) |>
  tbl_summary(
    by = carotid_plaque,
    missing = "no"
  ) |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics by carotid plaque status.**")
```

Participants with plaque are a decade older (median
`r median(dat_model$age_yrs[dat_model$carotid_plaque == 1])` versus
`r median(dat_model$age_yrs[dat_model$carotid_plaque == 0])` years), more often
male (`r fmt(100 * mean(dat_model$sex[dat_model$carotid_plaque == 1] == "Male"), 0)`%
versus `r fmt(100 * mean(dat_model$sex[dat_model$carotid_plaque == 0] == "Male"), 0)`%)
and have higher systolic pressure. RA status looks almost irrelevant:
`r fmt(100 * mean(dat_model$carotid_plaque[dat_model$has_ra == "RA"]), 1)`% of RA
cases have plaque against
`r fmt(100 * mean(dat_model$carotid_plaque[dat_model$has_ra == "Control"]), 1)`%
of controls.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Proportion with carotid plaque by age group and rheumatoid arthritis status. Group sizes are printed above each bar."}
dat_model |>
  mutate(age_grp = cut(age_yrs, breaks = c(19, 40, 50, 60, 71),
                       labels = c("20-39", "40-49", "50-59", "60-70"))) |>
  group_by(age_grp, has_ra) |>
  summarise(p = mean(carotid_plaque), n = n(), .groups = "drop") |>
  ggplot(aes(age_grp, p, fill = has_ra)) +
  geom_col(position = position_dodge(0.85), width = 0.78) +
  geom_text(aes(label = paste0("n=", n)), position = position_dodge(0.85),
            vjust = -0.4, size = 2.9, colour = "grey30") +
  scale_y_continuous(labels = scales::percent, expand = expansion(c(0, 0.12))) +
  scale_fill_manual(values = c(Control = "#3F7CAC", RA = "#B5495B")) +
  labs(x = "Age group (years)", y = "Proportion with carotid plaque",
       fill = "Group")
```

Plaque prevalence is driven overwhelmingly by age - it is
`r fmt(100 * mean(dat_model$carotid_plaque[dat_model$age_yrs < 40]), 0)`% below
age 40 and
`r fmt(100 * mean(dat_model$carotid_plaque[dat_model$age_yrs >= 60]), 0)`% at 60
and over - while the RA and control bars within each age band are close together
and cross over, with no consistent direction. (Age is grouped here for display
only; it enters the models as a continuous variable, which also avoids the
zero-event cells visible in the youngest band.)

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = glm,
    method.args = list(family = binomial),
    exponentiate = TRUE
  )
uv

utid <- map_dfr(predictors, function(v) {
  tidy(glm(reformulate(v, outcome), data = dat_model, family = binomial),
       conf.int = TRUE, exponentiate = TRUE) |>
    filter(term != "(Intercept)")
})
ug <- function(term, d = 2) {
  r <- utid[utid$term == term, ]
  sprintf("OR %s (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d),
          fmt(r$conf.high, d))
}
up <- function(term) pv(utid$p.value[utid$term == term])
```

Age dominates. Each additional year multiplies the odds of carotid plaque by
`r ug("age_yrs")` (p `r up("age_yrs")`), which compounds to roughly
`r fmt(utid$estimate[utid$term == "age_yrs"]^10, 1)`-fold per decade. Men have
`r ug("sexMale")` the odds of women (p `r up("sexMale")`). Each additional mmHg
of systolic pressure multiplies the odds by `r ug("systolic_bp")`
(p `r up("systolic_bp")`), about
`r fmt(utid$estimate[utid$term == "systolic_bp"]^10, 2)`-fold per 10 mmHg. The
remaining three show no crude association: total cholesterol
(`r ug("total_cholesterol")` per mmol/L, p `r up("total_cholesterol")`), current
smoking (`r ug("current_smokerYes")`, p `r up("current_smokerYes")`) and - the
exposure of interest - RA status (`r ug("has_raRA")`, p `r up("has_raRA")`),
whose confidence interval spans both a halving and a fifty-per-cent increase in
the odds.

## Why adjustment is needed

**Age** confounds almost everything here, and it confounds the smoking
association in the awkward direction. Plaque odds rise `r ug("age_yrs")` per
year, but current smokers in this sample are on average
`r fmt(mean(dat_model$age_yrs[dat_model$current_smoker == "No"]) - mean(dat_model$age_yrs[dat_model$current_smoker == "Yes"]), 1)`
years *younger* than non-smokers, and the imbalance is concentrated where it
matters: only
`r fmt(100 * mean(dat_model$current_smoker[dat_model$age_yrs >= 60] == "Yes"), 0)`%
of participants aged 60 and over smoke, against
`r fmt(100 * mean(dat_model$current_smoker[dat_model$age_yrs < 60] == "Yes"), 0)`%
of those under 60 (many older participants are ex-smokers, who are coded as
non-smokers here). Being younger protects strongly against plaque, so the crude
smoking odds ratio is dragged toward the null by the smokers' youth and a real
association is hidden - among participants under 50, for instance,
`r fmt(100 * mean(dat_model$carotid_plaque[dat_model$age_yrs < 50 & dat_model$current_smoker == "Yes"]), 1)`%
of smokers have plaque against
`r fmt(100 * mean(dat_model$carotid_plaque[dat_model$age_yrs < 50 & dat_model$current_smoker == "No"]), 1)`%
of non-smokers. **Sex** does the same: men are
`r fmt(mean(dat_model$age_yrs[dat_model$sex == "Male"]) - mean(dat_model$age_yrs[dat_model$sex == "Female"]), 1)`
years older on average as well as being at higher plaque risk for other reasons,
so part of the crude male excess is age. **Age also confounds the systolic
blood-pressure association**: arterial pressure rises with age
(r `r fmt(cor(dat_model$age_yrs, dat_model$systolic_bp), 2)` in this sample) and
so does plaque, so a crude blood-pressure gradient will partly be an age
gradient. And **age and sex jointly confound the RA comparison**, because the
groups were not matched on either: the RA group is
`r fmt(mean(dat_model$age_yrs[dat_model$has_ra == "RA"]) - mean(dat_model$age_yrs[dat_model$has_ra == "Control"]), 1)`
years older and more female. Mutual adjustment lets each odds ratio be read as a
comparison between participants of the same age, sex, blood pressure, smoking
status and cholesterol.

## Multivariable model

```{r multivariable}
fit <- glm(reformulate(predictors, outcome), data = dat_model, family = binomial)

mv <- tbl_regression(fit, exponentiate = TRUE)
mv

tid <- tidy(fit, conf.int = TRUE, exponentiate = TRUE)
g <- function(term, d = 2) {
  r <- tid[tid$term == term, ]
  sprintf("OR %s (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d),
          fmt(r$conf.high, d))
}
gp <- function(term) pv(tid$p.value[tid$term == term])
```

```{r fit-stats}
n_par    <- length(coef(fit)) - 1L
epp      <- n_events / n_par
null_fit <- glm(reformulate("1", outcome), data = dat_model, family = binomial)
tibble(
  Statistic = c("Complete-case N", "Events (plaque = 1)",
                "Parameters (excl. intercept)", "Events per parameter",
                "AIC (null model)", "AIC (full model)", "Tjur's R-squared"),
  Value = c(as.character(nrow(dat_model)), as.character(n_events),
            as.character(n_par), fmt(epp, 1), fmt(AIC(null_fit), 1),
            fmt(AIC(fit), 1), fmt(as.numeric(r2_tjur(fit)), 3))
) |>
  gt() |>
  tab_header(title = md("**Model fit: multivariable logistic model for carotid plaque**"))
```

The model has `r n_events` events for `r n_par` estimated parameters, i.e.
**`r fmt(epp, 1)` events per parameter**, which clears the conventional minimum
of ten but only because the predictor set was capped at six in advance. AIC
falls from `r fmt(AIC(null_fit), 1)` for the intercept-only model to
`r fmt(AIC(fit), 1)`, a drop of `r fmt(AIC(null_fit) - AIC(fit), 1)`, so the
predictors collectively carry real information; Tjur's R-squared is
`r fmt(as.numeric(r2_tjur(fit)), 3)`, which is modest - the model separates
plaque from no-plaque only partially.

Three predictors are independently associated with plaque. Among participants of
the same sex, RA status, blood pressure, smoking status and cholesterol, each
additional year of age multiplies the odds of plaque by `r g("age_yrs")`
(p `r gp("age_yrs")`), about `r fmt(tid$estimate[tid$term == "age_yrs"]^10, 1)`-fold
per decade. Among participants alike on everything else, men have
`r g("sexMale")` the odds of women (p `r gp("sexMale")`), and current smokers
`r g("current_smokerYes")` the odds of non-smokers (p `r gp("current_smokerYes")`).
The exposure of interest remains null: among participants of the same age, sex
and risk-factor profile, RA is associated with `r g("has_raRA")`
(p `r gp("has_raRA")`) - a point estimate suggesting somewhat *lower* odds, but
with a confidence interval that comfortably includes no association and a 35%
increase. Systolic blood pressure (`r g("systolic_bp")`, p `r gp("systolic_bp")`)
and total cholesterol (`r g("total_cholesterol")`, p `r gp("total_cholesterol")`)
are both null after adjustment. These are associations among the participants
observed, not effects of intervening.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted odds ratios for carotid plaque.**")
```

Two predictors move meaningfully in opposite directions. **Current smoking moves
away from the null and crosses into significance**, from `r ug("current_smokerYes")`
crude (p `r up("current_smokerYes")`) to `r g("current_smokerYes")` adjusted
(p `r gp("current_smokerYes")`) - the odds ratio roughly
`r fmt(tid$estimate[tid$term == "current_smokerYes"] / utid$estimate[utid$term == "current_smokerYes"], 1)`
times larger. This is the negative confounding by age described above: adding
age alone lifts the smoking odds ratio to
`r fmt(exp(coef(glm(carotid_plaque ~ current_smoker + age_yrs, data = dat_model, family = binomial)))["current_smokerYes"], 2)`,
and adding sex takes it the rest of the way. **Systolic blood pressure moves the
other way, toward the null and out of significance**, from `r ug("systolic_bp")`
(p `r up("systolic_bp")`) to `r g("systolic_bp")` (p `r gp("systolic_bp")`) - the
entire crude blood-pressure gradient is explained by age, which is unsurprising
given a correlation of `r fmt(cor(dat_model$age_yrs, dat_model$systolic_bp), 2)`
between the two. Total cholesterol does the same, from `r ug("total_cholesterol")`
to `r g("total_cholesterol")`.

**Age barely moves at all**: `r ug("age_yrs")` crude to `r g("age_yrs")`
adjusted, a change of
`r fmt(abs(100 * (tid$estimate[tid$term == "age_yrs"] / utid$estimate[utid$term == "age_yrs"] - 1)), 1)`%,
because nothing else in this model stands in for chronological age. Sex
attenuates only slightly, from `r ug("sexMale")` to `r g("sexMale")`, and keeps
its significance - part of the crude male excess was the men's greater age, but
most of it is not. The exposure of interest, RA status, moves from
`r ug("has_raRA")` to `r g("has_raRA")`: further below 1, consistent with the
same negative confounding seen for cIMT (the RA group is older), but nowhere
near significance in either column.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r binned-resid}
br <- binned_residuals(fit)
print(br)
```

```{r collin}
check_collinearity(fit)
```

```{r separation-check}
tibble(
  Check = c("Largest |coefficient| (log-odds scale)",
            "Largest standard error",
            "Largest variance inflation factor",
            "Largest Cook's distance",
            "Largest leverage",
            "Smallest cell: events in the smaller level of a binary predictor"),
  Value = c(fmt(max(abs(coef(fit)[-1])), 2),
            fmt(max(summary(fit)$coefficients[-1, 2]), 3),
            fmt(max(as.data.frame(check_collinearity(fit))$VIF), 2),
            fmt(max(cooks.distance(fit)), 3),
            fmt(max(hatvalues(fit)), 3),
            as.character(min(sum(dat_model$carotid_plaque[dat_model$sex == "Male"]),
                             sum(dat_model$carotid_plaque[dat_model$current_smoker == "Yes"]),
                             sum(dat_model$carotid_plaque[dat_model$has_ra == "Control"]))))
) |>
  gt() |>
  tab_header(title = md("**Table 3. Checks for separation and undue influence.**"))
```

For a logistic model, **homogeneity of variance and normality of residuals do
not apply and are not interpreted** - the panels `check_model()` draws for them
are ignored here. The checks that do matter are these four.

**Calibration (binned residuals).** This is the check that fails. Only about
`r fmt(100 * mean(as.data.frame(br)$group == "yes"), 0)`% of binned residuals
fall inside the error bounds against a target of 95% or more, so
`binned_residuals()` reports a probably-bad fit. Part of this is unavoidable
noise: with `r n_events` events spread over `r nrow(dat_model)` observations each
bin holds only a handful of events. Part of it is real - the pattern is
consistent with the linear-in-log-odds assumption for age being imperfect, since
plaque prevalence is essentially zero below age 40 and then accelerates.

**Collinearity is not a problem.** The largest variance inflation factor is
`r fmt(max(as.data.frame(check_collinearity(fit))$VIF), 2)`, far below the
conventional threshold of 5, so none of the six coefficients is competing with
another for the same variance. That is a consequence of the exclusions in
Methods rather than luck: adding `ldl_cholesterol` alongside total cholesterol,
HDL and triglycerides drives the lipid VIFs into the hundreds.

**No separation and no influential observations.** Every coefficient is finite,
the largest on the log-odds scale being `r fmt(max(abs(coef(fit)[-1])), 2)`, and
the largest standard error is `r fmt(max(summary(fit)$coefficients[-1, 2], na.rm = TRUE), 2)`
- both entirely ordinary, and there is no "fitted probabilities numerically 0 or
1" warning. Every level of every binary predictor contains events (the sparsest
combination still has
`r min(sum(dat_model$carotid_plaque[dat_model$sex == "Male"]), sum(dat_model$carotid_plaque[dat_model$current_smoker == "Yes"]), sum(dat_model$carotid_plaque[dat_model$has_ra == "Control"]))`).
The largest Cook's distance is `r fmt(max(cooks.distance(fit)), 3)` and the
largest leverage `r fmt(max(hatvalues(fit)), 3)`, so no single participant drives
the fit.

**Events per parameter.** `r fmt(epp, 1)`, just above the conventional ten. The
model is at its capacity; a seventh predictor would take it below.

# Discussion

In this Dutch case-control sample, carotid plaque is predicted by age, sex and
current smoking, and not by rheumatoid arthritis. Each decade of age multiplies
the odds of plaque by about `r fmt(tid$estimate[tid$term == "age_yrs"]^10, 1)`,
men have about `r fmt(tid$estimate[tid$term == "sexMale"], 1)` times the odds of
women, and current smokers about
`r fmt(tid$estimate[tid$term == "current_smokerYes"], 1)` times the odds of
non-smokers, each holding the others constant. RA status is null both crudely
(`r ug("has_raRA")`) and after adjustment (`r g("has_raRA")`). Practically, this
says that in these patients the presence of plaque tracks the ordinary
determinants of atherosclerosis, and that a clinician screening RA patients
should be looking at their age, sex and smoking rather than treating the RA
diagnosis itself as the risk marker.

The smoking result is the one worth dwelling on, because it only appears after
adjustment. Crudely, smokers look no different from non-smokers; that is an
artefact of the age distribution: smokers are
`r fmt(mean(dat_model$age_yrs[dat_model$current_smoker == "No"]) - mean(dat_model$age_yrs[dat_model$current_smoker == "Yes"]), 1)`
years younger on average and make up only
`r fmt(100 * mean(dat_model$current_smoker[dat_model$age_yrs >= 60] == "Yes"), 0)`%
of the over-60s, the age band where most of the plaque is, while
`current_smoker` counts ex-smokers as unexposed. Once age is held constant the
expected doubling of odds emerges. This
is a clean illustration of why an unadjusted analysis can be not merely
imprecise but directionally misleading.

A word is also owed to `statins` and `antihypertensives`, which are in the
dataset and were **not** used here. Both are treatment indicators, and treatment
is given because of risk: statin users in this sample are on average
`r fmt(mean(dat$age_yrs[dat$statins == "Yes"], na.rm = TRUE) - mean(dat$age_yrs[dat$statins == "No"], na.rm = TRUE), 1)`
years older than non-users and, despite being treated, still have total
cholesterol of
`r fmt(mean(dat$total_cholesterol[dat$statins == "Yes"], na.rm = TRUE), 1)`
mmol/L against
`r fmt(mean(dat$total_cholesterol[dat$statins == "No"], na.rm = TRUE), 1)` in
non-users. Put into a plaque model, they act as proxies for prior cardiovascular
risk and appear harmful; adjusted for the cholesterol and blood pressure they
were prescribed to lower, they are being conditioned on their own mediators.
Neither reading is a treatment effect, and with a six-parameter budget spending
two parameters on uninterpretable coefficients would be poor economy.

**Limitations.** The most serious problem is **differential missingness of the
outcome**. `r miss_ctrl` of `r n_ctrl` controls (`r fmt(100 * miss_ctrl / n_ctrl, 0)`%)
have no plaque reading against `r miss_ra` of `r n_ra` RA cases
(`r fmt(100 * miss_ra / n_ra, 1)`%), so the exposure group of interest is nearly
completely scanned while a quarter of the comparison group is not. Complete-case
analysis is unbiased here only if being scanned is unrelated to plaque status
among controls, which is untestable and unlikely if scanning was prompted by
clinical suspicion. The design is cross-sectional, so no temporal ordering is
established; the RA coefficient cannot be read as an effect of developing RA,
and smoking status is current status, not lifetime exposure, with ex-smokers
misclassified as unexposed - a misclassification that biases the smoking odds
ratio toward the null and means the true association is probably stronger than
reported. With only `r n_events` events the model can carry six parameters, so
adiposity, glucose, the full lipid panel and RA disease activity are all
omitted, and residual confounding is certain. RA disease duration, DAS28,
methotrexate and anti-TNF exposure are **not in this extract at all**, being
undefined for controls, so a null RA coefficient cannot rule out an association
confined to patients with active or long-standing disease. The binned-residual
check fails, indicating imperfect calibration. Complete-case analysis dropped
`r n_dropped` of `r nrow(dat)` rows. The sample is a single Dutch centre. All
estimates are associations, not causal effects.

**Future work.** Establish why a quarter of controls have no plaque reading and,
if the reason is administrative rather than clinical, refit with multiple
imputation of the outcome-adjacent covariates; separately, model plaque with age
as a spline to address the calibration failure.

# Feasibility notes

*NOT part of a student report.*

- **USABLE WITH WARNINGS.** `r nrow(dat_model)` complete cases, `r n_events`
  events, `r fmt(epp, 1)` events per parameter at six predictors. The model fits
  cleanly, has no separation, and finds a real and interpretable signal (age,
  sex, smoking). But two things must be flagged to students, below.
- **MAJOR: differential missingness of the outcome.** `carotid_plaque` is
  `r fmt(100 * mean(!is.na(dat_raw$carotid_plaque)), 1)`% complete overall, but
  that average hides everything: `r fmt(100 * miss_ctrl / n_ctrl, 1)`% missing
  among controls (`r miss_ctrl`/`r n_ctrl`) versus
  `r fmt(100 * miss_ra / n_ra, 1)`% among RA cases (`r miss_ra`/`r n_ra`). Since
  `has_ra` is the exposure of primary interest, the complete-case comparison is
  built on a control group that has lost a quarter of its members
  non-randomly. `index.html` presents this as a flat "8.9% missing", which
  understates the problem badly.
- **MAJOR: the parameter budget is genuinely binding.** Six predictors is the
  whole allowance (`r fmt(epp, 1)` EPP). The `meta.json` predictor list has 16
  entries; a student who uses all of them gets
  `r fmt(sum(dat_raw$carotid_plaque == 1, na.rm = TRUE) / 17, 1)` events per
  parameter and, because that model also loses rows to missingness, ends with
  about 408 observations and 64 events for 17 parameters.
- **MAJOR trap: `ldl_cholesterol` is calculated, not measured.** It is the
  Friedewald value (correlation `r fmt(ldl_r, 4)` with
  `total - HDL - triglycerides/2.2`). A student who puts total cholesterol, HDL,
  triglycerides and LDL in the same logistic model gets VIFs of roughly 478, 405,
  94 and 30 and lipid coefficients that are pure noise.
- **MAJOR trap: categorising age produces complete separation.** `age_yrs` is a
  natural candidate for banding, and there are **zero plaque events among the
  `r sum(dat_raw$age_yrs < 40 & !is.na(dat_raw$carotid_plaque), na.rm = TRUE)`
  scanned participants under 40** (`r sum(dat_raw$age_yrs < 30 & !is.na(dat_raw$carotid_plaque), na.rm = TRUE)`
  under 30 and the rest aged 30-39, none with plaque). Any model with an age
  band below 40 as a level will produce a divergent coefficient, an enormous
  standard error and a "fitted probabilities numerically 0 or 1" warning. Age
  must stay continuous.
- **MAJOR interpretive trap: `statins` and `antihypertensives` are confounded by
  indication**, and adjusting for the cholesterol and blood pressure they modify
  conditions on mediators. They are excluded here; a student who includes them
  and reports "statins are associated with plaque" has it backwards.
- **MAJOR: binned residuals fail** (~`r fmt(100 * mean(as.data.frame(br)$group == "yes"), 0)`%
  inside bounds against a 95% target). Partly the small event count per bin,
  partly genuine non-linearity in age. A student following the diagnostics
  honestly must report a failed check.
- **MINOR: `current_smoker` counts ex-smokers as unexposed** and smokers here are
  `r fmt(mean(dat_model$age_yrs[dat_model$current_smoker == "No"]) - mean(dat_model$age_yrs[dat_model$current_smoker == "Yes"]), 1)`
  years younger than non-smokers, so the crude smoking OR
  (`r ug("current_smokerYes")`) is null and only adjustment reveals the
  association (`r g("current_smokerYes")`). Good teaching material, but a student
  doing univariable screening first and dropping non-significant predictors will
  throw smoking away.
- **MINOR: `prednisone` is near-perfectly aligned with `has_ra`** -
  `r sum(dat_raw$prednisone == 1 & dat_raw$has_ra == 0, na.rm = TRUE)` of
  `r sum(dat_raw$prednisone == 1, na.rm = TRUE)` users is a control. Including
  both is close to including the same variable twice.
- **MINOR: the exposure of interest is null and stays null.** `has_ra`
  `r ug("has_raRA")` crude, `r g("has_raRA")` adjusted. This is a perfectly good
  finding but students hoping for a significant headline may be tempted to
  p-hack their way to one; the six-parameter budget makes that especially
  damaging here.
- **MINOR data defect: row `id = "control_row1"` is entirely blank.**
- **Verified clean:** `sex` ships as `r paste(sex_vals, collapse = "/")` and
  `height_cm` as `r ht_min`-`r ht_max` cm with `r n_ht_metres` values in metres,
  so both documented Zenodo defects are fixed in this catalogue's CSV. No
  RA-specific columns are present.
