---
title: "Carotid intima-media thickness in adults with and without rheumatoid arthritis"
subtitle: "Dataset 10: Cardiovascular risk and carotid artery thickness in rheumatoid arthritis"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "cimt_rheumatoid.csv"),
                    show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names. The predictor set is a
# deliberately chosen subset: see Methods for the collinearity reasoning behind
# dropping waist_cm, diastolic_bp, ldl_cholesterol, apo_a and apo_b.
outcome    <- "cimt_total"
predictors <- c("age_yrs", "sex", "has_ra", "bmi", "systolic_bp",
                "current_smoker", "statins", "antihypertensives",
                "total_cholesterol", "hdl_cholesterol", "triglycerides",
                "glucose", "log_crp")

lab <- function(x, l) { attr(x, "label") <- l; x }

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the clinically "unexposed" group throughout.
    sex    = lab(factor(sex, levels = c("female", "male"),
                        labels = c("Female", "Male")), "Sex"),
    has_ra = lab(factor(has_ra, levels = c(0, 1),
                        labels = c("Control", "RA")), "Rheumatoid arthritis"),
    current_smoker    = lab(factor(current_smoker, levels = c(0, 1),
                                   labels = c("No", "Yes")), "Current smoker"),
    statins           = lab(factor(statins, levels = c(0, 1),
                                   labels = c("No", "Yes")), "Taking a statin"),
    antihypertensives = lab(factor(antihypertensives, levels = c(0, 1),
                                   labels = c("No", "Yes")),
                            "Taking antihypertensives"),
    # CRP is strongly right-skewed (skewness ~5), so it enters as a predictor on
    # the log scale; see Methods.
    log_crp = lab(log(crp), "log CRP (log mg/L)"),
    age_yrs = lab(age_yrs, "Age (years)"),
    bmi = lab(bmi, "BMI (kg/m2)"),
    systolic_bp = lab(systolic_bp, "Systolic BP (mmHg)"),
    total_cholesterol = lab(total_cholesterol, "Total cholesterol (mmol/L)"),
    hdl_cholesterol = lab(hdl_cholesterol, "HDL cholesterol (mmol/L)"),
    triglycerides = lab(triglycerides, "Triglycerides (mmol/L)"),
    glucose = lab(glucose, "Fasting glucose (mmol/L)"),
    cimt_total = lab(cimt_total, "cIMT (mm)")
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
pv  <- function(p) ifelse(p < 0.001, "<0.001", paste0("= ", fmt(p, 3)))

# Data-integrity checks referred to in Methods and in the feasibility notes.
sex_vals    <- sort(unique(na.omit(dat_raw$sex)))
ht_min      <- min(dat_raw$height_cm, na.rm = TRUE)
ht_max      <- max(dat_raw$height_cm, na.rm = TRUE)
n_ht_metres <- sum(dat_raw$height_cm < 3, na.rm = TRUE)
bmi_gap     <- max(abs(dat_raw$bmi - dat_raw$weight_kg / (dat_raw$height_cm / 100)^2),
                   na.rm = TRUE)
friedewald  <- dat_raw$total_cholesterol - dat_raw$hdl_cholesterol -
               dat_raw$triglycerides / 2.2
ldl_r       <- cor(dat_raw$ldl_cholesterol, friedewald, use = "complete.obs")
ldl_gap     <- mean(abs(dat_raw$ldl_cholesterol - friedewald), na.rm = TRUE)
n_empty_row <- sum(rowSums(is.na(dat_raw)) >= ncol(dat_raw) - 3)
```

# Introduction

The source study recruited adults with rheumatoid arthritis (RA) and healthy
controls without RA in the Netherlands, put every participant through the same
physical examination and fasting laboratory panel, and measured carotid
intima-media thickness (cIMT) semi-automatically by B-mode ultrasound
(Hollan and colleagues, *PLOS ONE*, 2015). This teaching extract holds
`r nrow(dat)` participants, `r sum(dat_raw$has_ra == 1)` with RA and
`r sum(dat_raw$has_ra == 0)` controls. Carotid intima-media thickness is a
non-invasive structural marker of early atherosclerosis that predicts subsequent
myocardial infarction and stroke, and the long-standing clinical claim is that
the chronic systemic inflammation of RA accelerates it beyond what conventional
risk factors explain.

**Aim.** Among these Dutch adults, we examined which of RA status, age, sex,
adiposity, blood pressure, smoking, lipid-lowering and antihypertensive
treatment, the fasting lipid panel, glucose and C-reactive protein are
independently associated with carotid intima-media thickness, mutually adjusting
for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** A cross-sectional case-control comparison conducted in the
Netherlands and published in *PLOS ONE* in 2015. Cases were adults with
established rheumatoid arthritis; controls were adults without RA recruited from
the same setting. This extract contains `r nrow(dat)` participants and 25
variables: anthropometry, blood pressure, current medication, a fasting
lipid and glucose panel, C-reactive protein, and the two carotid ultrasound
measures. Ages run from `r min(dat$age_yrs, na.rm = TRUE)` to
`r max(dat$age_yrs, na.rm = TRUE)` years.

**Variables.** The outcome is `cimt_total`, mean carotid intima-media thickness
across both sides, in millimetres, treated as continuous (median
`r fmt(median(dat$cimt_total, na.rm = TRUE), 2)` mm, range
`r fmt(min(dat$cimt_total, na.rm = TRUE), 2)` to
`r fmt(max(dat$cimt_total, na.rm = TRUE), 2)` mm). Predictors are rheumatoid
arthritis status (reference Control), age in years, sex (reference Female),
BMI in kg/m2, systolic blood pressure in mmHg, current smoking (reference No),
statin use (reference No), antihypertensive use (reference No), total
cholesterol, HDL cholesterol and triglycerides in mmol/L, fasting glucose in
mmol/L, and log-transformed C-reactive protein. **RA status is the exposure of
primary interest**; the remainder are conventional cardiovascular risk factors
carried as covariates.

Cases and controls were **not matched on age or sex** (RA cases have median age
`r median(dat_model$age_yrs[dat_model$has_ra == "RA"])` versus
`r median(dat_model$age_yrs[dat_model$has_ra == "Control"])` years for controls,
and `r fmt(100 * mean(dat_model$sex[dat_model$has_ra == "RA"] == "Male"), 0)`%
versus `r fmt(100 * mean(dat_model$sex[dat_model$has_ra == "Control"] == "Male"), 0)`%
men), so age and sex are forced into every adjusted model regardless of their
own significance.

**Variables deliberately excluded, and why.** Six candidate predictors are left
out on structural grounds, fixed before any model was fitted.

- `weight_kg` and `height_cm`: `bmi` is exactly `weight_kg / (height_cm/100)^2`
  in this file (largest discrepancy across all rows
  `r formatC(bmi_gap, format = "e", digits = 1)` kg/m2, i.e. floating-point
  noise), so including any two of the three is a deterministic relationship.
- `waist_cm`: correlated `r fmt(cor(dat$bmi, dat$waist_cm, use = "complete.obs"), 2)`
  with BMI. Both measure adiposity; BMI is kept because it is the more
  completely recorded of the two.
- `diastolic_bp`: correlated
  `r fmt(cor(dat$systolic_bp, dat$diastolic_bp, use = "complete.obs"), 2)` with
  systolic blood pressure. Systolic pressure is kept because it is the stronger
  correlate of arterial structure at these ages.
- `ldl_cholesterol`: **LDL in this file is not independently measured, it is
  calculated.** Against the Friedewald formula
  (`total - HDL - triglycerides/2.2`) it correlates `r fmt(ldl_r, 4)` with a mean
  absolute discrepancy of `r fmt(ldl_gap, 3)` mmol/L, which is rounding. Putting
  total cholesterol, HDL, triglycerides *and* LDL in one model is therefore a
  near-exact linear dependency; doing so inflates the variance inflation factors
  for the lipid terms into the hundreds (documented in the feasibility notes).
- `apo_a` and `apo_b`: apolipoprotein B correlates
  `r fmt(cor(dat$apo_b, dat$total_cholesterol, use = "complete.obs"), 2)` with
  total cholesterol and apolipoprotein A correlates
  `r fmt(cor(dat$apo_a, dat$hdl_cholesterol, use = "complete.obs"), 2)` with
  HDL; they are alternative measurements of the same two lipid fractions rather
  than additional information.
- `hypertension_or_hyperchol`: every participant taking an antihypertensive or a
  statin is coded 1 on this variable, so it is a structural superset of two
  predictors already in the model.

**Parameter budget.** With `r nrow(dat_model)` complete cases and
`r length(predictors)` predictors, all of which cost one degree of freedom, the
model estimates `r length(predictors)` parameters, i.e.
`r fmt(nrow(dat_model) / length(predictors), 0)` observations per parameter. That
is comfortably inside the 10-15 observations per parameter that a linear model
needs, so the cap here is imposed by collinearity rather than by sample size.

**Statistical analysis.** `lm()`, justified because the outcome is a continuous
measurement in millimetres with a roughly symmetric distribution. We fitted a
univariable model for each predictor first, then a single multivariable model
containing all of them. Predictors were chosen in advance on subject-matter
grounds, not by significance. Significance threshold 0.05; estimates are
reported with 95% confidence intervals. Assumptions were checked with
`performance::check_model()`, supplemented by `check_collinearity()`, and
because non-constant variance was flagged we additionally report HC3
heteroscedasticity-consistent standard errors. C-reactive protein is used as a
predictor on the **log scale**: raw CRP has skewness
`r fmt(with(dat, {x <- na.omit(crp); mean((x - mean(x))^3) / sd(x)^3}), 1)` and a
range of `r min(dat$crp, na.rm = TRUE)` to `r max(dat$crp, na.rm = TRUE)`, so
untransformed it would let a handful of participants with CRP above 40 dominate
its coefficient. Analysis used `r R.version.string` with tidyverse, gtsummary
and performance.

**Data integrity.** The catalogue notes that in the original Zenodo deposit
`sex` was coded 0 = woman / 1 = man and the height column mixed metres and
centimetres. Both look corrected in the shipped CSV: `sex` takes the values
`r paste(sex_vals, collapse = " and ")`, and `height_cm` runs from `r ht_min` to
`r ht_max` cm with `r n_ht_metres` values below 3 (i.e. none left in metres).
One residual defect remains: row `id = "control_row1"` is **entirely empty**
apart from `id`, `has_ra` and `hypertension_or_hyperchol`, so the file's
`r nrow(dat)` rows are really `r nrow(dat) - n_empty_row` participants plus one
blank record. It is dropped by complete-case analysis and does not affect any
estimate here.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped,
`r fmt(100 * n_dropped / nrow(dat), 1)`%). The outcome itself is
`r fmt(100 * mean(!is.na(dat$cimt_total)), 1)`% complete
(`r sum(is.na(dat$cimt_total))` missing); the remaining attrition is spread
thinly across the predictors, the largest single contributors being `crp`
(`r sum(is.na(dat$crp))` missing), `statins` (`r sum(is.na(dat$statins))`) and
`current_smoker` (`r sum(is.na(dat$current_smoker))`). Only
`r sum(complete.cases(dat_raw))` of
`r nrow(dat)` rows are complete on *every* column in the file, so adding
predictors would cost further rows.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(
    by = has_ra,
    missing = "no"
  ) |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics of the analytic sample, by rheumatoid arthritis status.**")
```

The two groups are broadly similar in crude cIMT (median
`r fmt(median(dat_model$cimt_total[dat_model$has_ra == "RA"]), 2)` mm in RA
versus `r fmt(median(dat_model$cimt_total[dat_model$has_ra == "Control"]), 2)` mm
in controls) but not in the things that drive cIMT: RA cases are on average
`r fmt(mean(dat_model$age_yrs[dat_model$has_ra == "RA"]) - mean(dat_model$age_yrs[dat_model$has_ra == "Control"]), 1)`
years older, have systolic pressure about
`r fmt(mean(dat_model$systolic_bp[dat_model$has_ra == "RA"]) - mean(dat_model$systolic_bp[dat_model$has_ra == "Control"]), 0)`
mmHg higher, and are a more female group. Their C-reactive protein is markedly
higher (median `r median(exp(dat_model$log_crp[dat_model$has_ra == "RA"]))`
versus `r median(exp(dat_model$log_crp[dat_model$has_ra == "Control"]))` mg/L),
which is the expected signature of active systemic inflammation.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Carotid intima-media thickness against age, separately for participants with rheumatoid arthritis and controls, with linear smoothers and 95% confidence bands."}
ggplot(dat_model, aes(age_yrs, cimt_total, colour = has_ra, fill = has_ra)) +
  geom_point(alpha = 0.45, size = 1.5) +
  geom_smooth(method = "lm", alpha = 0.18, linewidth = 0.9) +
  scale_colour_manual(values = c(Control = "#3F7CAC", RA = "#B5495B")) +
  scale_fill_manual(values = c(Control = "#3F7CAC", RA = "#B5495B")) +
  labs(x = "Age (years)", y = "Carotid intima-media thickness (mm)",
       colour = "Group", fill = "Group")
```

Carotid intima-media thickness rises steadily and near-linearly with age in both
groups, and the two smoothers lie almost on top of one another, with the RA line
slightly *below* the control line across most of the age range - so whatever RA
contributes, it is not a visible upward shift in cIMT at a given age.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv

utid <- map_dfr(predictors, function(v) {
  tidy(lm(reformulate(v, outcome), data = dat_model), conf.int = TRUE) |>
    filter(term != "(Intercept)")
})
ug <- function(term, d = 4) {
  r <- utid[utid$term == term, ]
  sprintf("%s mm (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d),
          fmt(r$conf.high, d))
}
up <- function(term) pv(utid$p.value[utid$term == term])
```

Age is by far the strongest crude predictor: each additional year is associated
with `r ug("age_yrs")` greater cIMT (p `r up("age_yrs")`), i.e. about
`r fmt(10 * utid$estimate[utid$term == "age_yrs"], 2)` mm per decade against a
sample median of `r fmt(median(dat_model$cimt_total), 2)` mm. Systolic blood
pressure follows (`r ug("systolic_bp")` per mmHg, p `r up("systolic_bp")`), then
BMI (`r ug("bmi")` per kg/m2, p `r up("bmi")`), fasting glucose
(`r ug("glucose")` per mmol/L, p `r up("glucose")`), total cholesterol
(`r ug("total_cholesterol")` per mmol/L, p `r up("total_cholesterol")`) and
triglycerides (`r ug("triglycerides")`, p `r up("triglycerides")`). Men average
`r ug("sexMale")` more than women (p `r up("sexMale")`). The two treatment
indicators look like strong *risk* markers: statin users average
`r ug("statinsYes")` more cIMT (p `r up("statinsYes")`) and participants on
antihypertensives `r ug("antihypertensivesYes")` more
(p `r up("antihypertensivesYes")`). Crucially, **the exposure of interest shows
nothing crudely**: RA status is associated with `r ug("has_raRA")`
(p `r up("has_raRA")`), a difference that is both small and firmly
non-significant. Current smoking (p `r up("current_smokerYes")`), HDL
cholesterol (p `r up("hdl_cholesterol")`) and log CRP
(p `r up("log_crp")`) show no crude association.

## Why adjustment is needed

**Age** is the confounder that matters most, and it works directly against the
exposure of interest. Age is the dominant determinant of cIMT in this sample
(crude `r ug("age_yrs")` per year), and because cases and controls were not
age-matched the RA group is
`r fmt(mean(dat_model$age_yrs[dat_model$has_ra == "RA"]) - mean(dat_model$age_yrs[dat_model$has_ra == "Control"]), 1)`
years older on average. The RA group therefore carries an age-driven *upward*
push on cIMT that masks whatever the disease itself contributes; adding age
alone moves the RA coefficient from `r ug("has_raRA")` to
`r fmt(coef(lm(cimt_total ~ has_ra + age_yrs, data = dat_model))["has_raRA"], 4)`
mm. **Systolic blood pressure** does the same job for the same reason: RA cases
average
`r fmt(mean(dat_model$systolic_bp[dat_model$has_ra == "RA"]) - mean(dat_model$systolic_bp[dat_model$has_ra == "Control"]), 1)`
mmHg higher, and higher pressure independently thickens the arterial wall, so
part of the RA group's cIMT is a blood-pressure difference rather than a disease
effect. **Sex** is a third: men have thicker intima-media at every age
(`r ug("sexMale")` crude) and the RA group is more female
(`r fmt(100 * mean(dat_model$sex[dat_model$has_ra == "RA"] == "Male"), 0)`% men
versus `r fmt(100 * mean(dat_model$sex[dat_model$has_ra == "Control"] == "Male"), 0)`%),
which pushes the crude RA estimate in the opposite direction to age. Finally,
**age confounds the two treatment indicators**: statin users are on average
`r fmt(mean(dat_model$age_yrs[dat_model$statins == "Yes"]) - mean(dat_model$age_yrs[dat_model$statins == "No"]), 1)`
years older than non-users, which is enough on its own to generate the crude
"statins are associated with thicker arteries" result. Mutual adjustment lets
each estimate be read as a comparison between participants of the same age, sex,
body mass, blood pressure, lipid profile and treatment status.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv

tid <- tidy(fit, conf.int = TRUE)
g <- function(term, d = 4) {
  r <- tid[tid$term == term, ]
  sprintf("%s mm (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d),
          fmt(r$conf.high, d))
}
gp <- function(term) pv(tid$p.value[tid$term == term])
```

```{r fit-stats}
s <- summary(fit)
tibble(
  Statistic = c("Complete-case N", "Parameters (excl. intercept)",
                "Observations per parameter", "R-squared", "Adjusted R-squared",
                "Residual SE (mm)"),
  Value = c(as.character(nrow(dat_model)),
            as.character(length(coef(fit)) - 1L),
            fmt(nrow(dat_model) / (length(coef(fit)) - 1L), 1),
            fmt(s$r.squared, 3), fmt(s$adj.r.squared, 3), fmt(s$sigma, 4))
) |>
  gt() |>
  tab_header(title = md("**Model fit: multivariable linear model for carotid intima-media thickness**"))
```

The model explains `r fmt(100 * s$r.squared, 1)`% of the variance in cIMT
(adjusted R-squared `r fmt(s$adj.r.squared, 3)`), with a residual standard
deviation of `r fmt(s$sigma, 3)` mm - about a sixth of the observed cIMT range.
The gap between R-squared and adjusted R-squared is small because
`r length(coef(fit)) - 1L` parameters are spread over `r nrow(dat_model)`
observations.

Three predictors survive adjustment. Among participants of the same sex, RA
status, body mass, blood pressure, smoking status, treatment and laboratory
profile, each additional year of age is associated with `r g("age_yrs")` greater
cIMT (p `r gp("age_yrs")`), i.e. `r fmt(10 * tid$estimate[tid$term == "age_yrs"], 2)`
mm per decade. Among participants alike on everything else, each additional mmHg
of systolic pressure is associated with `r g("systolic_bp")`
(p `r gp("systolic_bp")`), or `r fmt(10 * tid$estimate[tid$term == "systolic_bp"], 3)`
mm per 10 mmHg. And the exposure of interest becomes significant only *after*
adjustment: among participants of the same age, sex and cardiovascular risk
profile, those with RA have `r g("has_raRA")` cIMT (p `r gp("has_raRA")`) - a
**thinner**, not thicker, intima-media. Nothing else reaches significance:
BMI (`r g("bmi")`, p `r gp("bmi")`), current smoking (`r g("current_smokerYes")`,
p `r gp("current_smokerYes")`), statins (`r g("statinsYes")`,
p `r gp("statinsYes")`), antihypertensives (`r g("antihypertensivesYes")`,
p `r gp("antihypertensivesYes")`), total cholesterol
(`r g("total_cholesterol")`, p `r gp("total_cholesterol")`), HDL
(`r g("hdl_cholesterol")`, p `r gp("hdl_cholesterol")`), triglycerides
(`r g("triglycerides")`, p `r gp("triglycerides")`), glucose (`r g("glucose")`,
p `r gp("glucose")`) and log CRP (`r g("log_crp")`, p `r gp("log_crp")`) are all
null. These are associations among the participants observed, not effects of
intervening.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates for carotid intima-media thickness.**")
```

The two treatment indicators move furthest, and both move *toward* the null.
Statin use falls from `r ug("statinsYes")` crude (p `r up("statinsYes")`) to
`r g("statinsYes")` adjusted (p `r gp("statinsYes")`), losing about
`r fmt(100 * (1 - tid$estimate[tid$term == "statinsYes"] / utid$estimate[utid$term == "statinsYes"]), 0)`%
of its magnitude and its significance; antihypertensive use falls from
`r ug("antihypertensivesYes")` (p `r up("antihypertensivesYes")`) to
`r g("antihypertensivesYes")` (p `r gp("antihypertensivesYes")`). This is
confounding by indication compounded by age: statin and antihypertensive users
are older and have higher blood pressure, and once age and pressure are held
constant the apparent "harm" of treatment disappears. Total cholesterol behaves
the same way, from `r ug("total_cholesterol")` (p `r up("total_cholesterol")`) to
`r g("total_cholesterol")` (p `r gp("total_cholesterol")`), as does glucose,
which crosses from `r ug("glucose")` (p `r up("glucose")`) to
`r g("glucose")` (p `r gp("glucose")`) - both crude estimates were essentially
age in disguise.

One estimate moves decisively *away* from the null, and it is the exposure of
interest. RA status goes from `r ug("has_raRA")` (p `r up("has_raRA")`) to
`r g("has_raRA")` (p `r gp("has_raRA")`), roughly
`r fmt(tid$estimate[tid$term == "has_raRA"] / utid$estimate[utid$term == "has_raRA"], 1)`
times larger in magnitude, and crosses *into* significance. This is negative
confounding: the RA group's greater age and higher blood pressure were
concealing a modest negative association, and removing them uncovers it. By
contrast **age barely moves** - `r ug("age_yrs")` crude to `r g("age_yrs")`
adjusted, a change of about
`r fmt(100 * (1 - tid$estimate[tid$term == "age_yrs"] / utid$estimate[utid$term == "age_yrs"]), 0)`% -
which is what one expects of a variable that nothing else in the model can stand
in for. Systolic blood pressure is intermediate: it more than halves, from
`r ug("systolic_bp")` to `r g("systolic_bp")`, but stays clearly significant,
so part of the crude blood-pressure gradient was age and part of it is its own.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r collin}
check_collinearity(fit)
```

```{r assumption-tests}
het  <- check_heteroscedasticity(fit)
print(het)
norm <- check_normality(fit)
print(norm)
```

```{r transform-check}
res_skew <- function(m) { r <- resid(m); mean((r - mean(r))^3) / sd(r)^3 }
fit_log  <- lm(reformulate(predictors, "log(cimt_total)"), data = dat_model)
fit_sqrt <- lm(reformulate(predictors, "sqrt(cimt_total)"), data = dat_model)
tibble(
  Outcome = c("cIMT (mm, untransformed)", "log(cIMT)", "sqrt(cIMT)"),
  `Residual skewness` = fmt(c(res_skew(fit), res_skew(fit_log), res_skew(fit_sqrt)), 2),
  `Shapiro-Wilk p` = formatC(c(shapiro.test(resid(fit))$p.value,
                               shapiro.test(resid(fit_log))$p.value,
                               shapiro.test(resid(fit_sqrt))$p.value),
                             format = "e", digits = 1),
  `R-squared` = fmt(c(summary(fit)$r.squared, summary(fit_log)$r.squared,
                      summary(fit_sqrt)$r.squared), 3)
) |>
  gt() |>
  tab_header(title = md("**Table 3. Would a transformation of the outcome help?**"))
```

```{r robust-se}
rob   <- lmtest::coeftest(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
robci <- lmtest::coefci(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
tibble(
  Term = rownames(rob),
  Estimate = fmt(rob[, 1], 4),
  `SE (OLS)` = fmt(summary(fit)$coefficients[, 2], 4),
  `SE (HC3)` = fmt(rob[, 2], 4),
  `95% CI (HC3)` = paste0(fmt(robci[, 1], 4), " to ", fmt(robci[, 2], 4)),
  `p (HC3)` = fmt(rob[, 4], 3)
) |>
  filter(Term != "(Intercept)") |>
  gt() |>
  tab_header(title = md("**Table 4. HC3 heteroscedasticity-consistent standard errors.**"))
```

**Linearity and predicted values.** The residual-versus-fitted panel is close to
flat, and Figure 1 already showed that the age-cIMT relationship is essentially
straight, so the linear specification is defensible. The posterior-predictive
check reproduces the bulk of the cIMT distribution well but slightly
under-predicts the right shoulder.

**Collinearity is not a problem in this model.** The largest variance inflation
factor is
`r fmt(max(as.data.frame(check_collinearity(fit))$VIF), 2)` (for
`r as.data.frame(check_collinearity(fit))$Term[which.max(as.data.frame(check_collinearity(fit))$VIF)]`)
and every VIF is below
`r fmt(ceiling(max(as.data.frame(check_collinearity(fit))$VIF) * 2) / 2, 1)`,
well short of the conventional threshold of 5. That is a direct consequence of
the exclusions set out in Methods: the same model with `waist_cm`,
`diastolic_bp` and `ldl_cholesterol` added has VIFs of about
`r fmt(max(as.data.frame(check_collinearity(lm(reformulate(c(predictors, "waist_cm", "diastolic_bp", "ldl_cholesterol"), outcome), data = dat |> select(all_of(c(outcome, predictors, "waist_cm", "diastolic_bp", "ldl_cholesterol"))) |> drop_na())))$VIF), 0)`
for the lipid terms.

**Variance is not constant.** `check_heteroscedasticity()` rejects constant
error variance (p < 0.001): participants predicted to have thicker
intima-media also vary more around their prediction, which is the usual pattern
for a measurement with a floor near zero. Table 4 therefore reports HC3 robust
standard errors. They change nothing substantive - the HC3 standard error for
the RA coefficient is `r fmt(rob["has_raRA", 2], 4)` against an OLS
`r fmt(summary(fit)$coefficients["has_raRA", 2], 4)`, and age, systolic pressure
and RA status keep their significance status under robust inference.

**Residuals are non-normal, but a transformation is the wrong fix.**
`check_normality()` rejects (p < 0.001). The cause is not skew - residual
skewness is only `r fmt(res_skew(fit), 2)` - but heavy tails: residual kurtosis
is `r fmt(with(list(r = resid(fit)), mean((r - mean(r))^4) / sd(r)^4), 1)`
against 3 for a normal distribution, driven by a small number of participants
with unusually thin or unusually thick arteries. Table 3 shows that
log-transforming the outcome makes matters clearly *worse* (residual skewness
`r fmt(res_skew(fit_log), 2)`, i.e. a strong left tail, and a smaller Shapiro-Wilk
p-value), and the square-root transform is no better; the outcome is therefore
left on its natural millimetre scale, where the coefficients are directly
interpretable. With `r nrow(dat_model)` observations the central limit theorem
covers the coefficient standard errors, and the HC3 intervals are the ones we
rely on.

**No influential observations.** The largest Cook's distance is
`r fmt(max(cooks.distance(fit)), 3)` and the largest leverage
`r fmt(max(hatvalues(fit)), 3)`, so no single participant drives the fit.

# Discussion

In this Dutch case-control sample, carotid intima-media thickness is dominated by
age and systolic blood pressure: about
`r fmt(10 * tid$estimate[tid$term == "age_yrs"], 2)` mm per decade of age and
`r fmt(10 * tid$estimate[tid$term == "systolic_bp"], 3)` mm per 10 mmHg of
systolic pressure, each holding the other constant. Rheumatoid arthritis itself
shows **no crude association** with cIMT, and after adjustment for age, sex and
conventional risk factors the RA group has intima-media that is
`r fmt(abs(tid$estimate[tid$term == "has_raRA"]), 3)` mm *thinner*
(p `r gp("has_raRA")`). Against a median cIMT of
`r fmt(median(dat_model$cimt_total), 2)` mm that is a difference of roughly
`r fmt(100 * abs(tid$estimate[tid$term == "has_raRA"]) / median(dat_model$cimt_total), 0)`%,
and it is around
`r fmt(abs(tid$estimate[tid$term == "has_raRA"]) / tid$estimate[tid$term == "age_yrs"], 1)`
years' worth of ageing - real but small. The practical reading is that in this
sample the hypothesis that RA accelerates subclinical atherosclerosis is not
supported; the cardiovascular risk carried by these RA patients is the risk
carried by their age and their blood pressure.

The result that most deserves caution is the pair of treatment indicators. Both
statin use and antihypertensive use look strongly *harmful* in the crude
analysis, and neither survives adjustment. That is confounding by indication in
its purest form: people are prescribed statins **because** their lipids and
overall cardiovascular risk are poor, and prescribed antihypertensives
**because** their blood pressure is high. In this sample statin users are on
average
`r fmt(mean(dat_model$age_yrs[dat_model$statins == "Yes"]) - mean(dat_model$age_yrs[dat_model$statins == "No"]), 1)`
years older and, despite treatment, have total cholesterol of
`r fmt(mean(dat_model$total_cholesterol[dat_model$statins == "Yes"]), 1)` mmol/L
against `r fmt(mean(dat_model$total_cholesterol[dat_model$statins == "No"]), 1)`
in non-users. Neither the crude nor the adjusted coefficient can be read as an
effect of treatment: the adjusted one conditions on the very lipid and pressure
values that treatment was given to change, which is adjusting for a mediator.
These indicators belong in the model as risk *markers* and should be described
that way, never as evidence about whether statins work.

**Limitations.** The design is cross-sectional, so no temporal ordering is
established and the RA coefficient cannot be read as an effect of developing RA.
Cases and controls were not matched on age or sex, so all comparisons rest on
statistical adjustment; residual confounding by age is likely because age enters
linearly while the age-cIMT relationship may curve at the extremes. The RA
group's disease-specific characteristics - disease duration, DAS28 activity
score, methotrexate, prednisone and anti-TNF exposure - are **not in this
extract at all** (they are undefined for the `r sum(dat_raw$has_ra == 0)`
controls), so we cannot distinguish RA patients with active from quiescent
disease, and the null RA finding may conceal a real association confined to
poorly controlled patients. Immunosuppressive treatment could itself reduce
cIMT, in which case the negative RA coefficient is a treatment effect masquerading
as a disease effect. Statin and antihypertensive use are confounded by indication
as described above, and adjusting for the lipid and blood-pressure values they
modify is over-adjustment. Error variance is non-constant and residuals are
heavy-tailed; HC3 standard errors address inference but not the model's fit in
the tails. Complete-case analysis dropped `r n_dropped` of `r nrow(dat)` rows.
The sample is a single Dutch centre and is not representative of any wider
population. All estimates are associations, not causal effects.

**Future work.** Refit with the RA-specific disease-activity and treatment
variables from the full deposit, restricted to the RA cases, to test whether
cumulative inflammatory burden rather than the RA label predicts intima-media
thickness.

# Feasibility notes

*NOT part of a student report.*

- **USABLE, and the best of the three outcomes in this dataset.** The outcome is
  `r fmt(100 * mean(!is.na(dat$cimt_total)), 1)`% complete, the complete-case N is
  `r nrow(dat_model)` of `r nrow(dat)`, R-squared is `r fmt(s$r.squared, 3)`, and
  there is a clean, strong, genuinely interpretable signal in age and systolic
  pressure. It is also a good teaching case because the exposure of interest is
  null crudely and significant only after adjustment.
- **MAJOR trap: `ldl_cholesterol` is the Friedewald calculation, not an
  independent measurement.** Against `total_cholesterol - hdl_cholesterol -
  triglycerides/2.2` it correlates `r fmt(ldl_r, 4)` (mean absolute discrepancy
  `r fmt(ldl_gap, 3)` mmol/L). A student who puts all four lipid variables in one
  model gets VIFs of roughly 500 (`total_cholesterol`), 410
  (`ldl_cholesterol`), 93 (`hdl_cholesterol`) and 36 (`triglycerides`), with
  standard errors an order of magnitude too large and coefficients that flip
  sign. This is the single most likely way a student wrecks this outcome. The
  `meta.json` predictor list currently suggests all four.
- **MAJOR interpretive trap: `statins` and `antihypertensives` are confounded by
  indication.** Both look harmful crudely (statins `r ug("statinsYes")`,
  p `r up("statinsYes")`) and both vanish on adjustment. A student who reports
  "statin use is associated with thicker arteries" without the indication caveat
  has written something clinically backwards. Worse, adjusting for cholesterol
  and blood pressure conditions on the mediators of treatment, so neither column
  of Table 2 is interpretable as a treatment effect.
- **MINOR-to-MAJOR: `bmi` and `waist_cm` (r
  `r fmt(cor(dat$bmi, dat$waist_cm, use = "complete.obs"), 2)`) and
  `systolic_bp` and `diastolic_bp` (r
  `r fmt(cor(dat$systolic_bp, dat$diastolic_bp, use = "complete.obs"), 2)`) are
  redundant pairs.** Including both members of each pair raises their VIFs to
  about 3.5 and 2.2 respectively - not fatal, but it splits the adiposity and
  pressure signal across two coefficients and neither reaches significance.
- **MINOR: `apo_a`/`apo_b` duplicate HDL and total cholesterol** (r
  `r fmt(cor(dat$apo_a, dat$hdl_cholesterol, use = "complete.obs"), 2)` and
  `r fmt(cor(dat$apo_b, dat$total_cholesterol, use = "complete.obs"), 2)`).
- **MINOR: `hypertension_or_hyperchol` is a structural superset of `statins` and
  `antihypertensives`** - every treated participant is coded 1 - so the three
  together are close to nested.
- **MINOR: `weight_kg` and `height_cm` must be excluded**; `bmi` is exactly
  `weight_kg/(height_cm/100)^2` (max discrepancy
  `r formatC(bmi_gap, format = "e", digits = 1)`).
- **MINOR failed diagnostics.** Non-constant variance (p < 0.001) and non-normal
  residuals (p < 0.001), the latter from kurtosis
  `r fmt(with(list(r = resid(fit)), mean((r - mean(r))^4) / sd(r)^4), 1)` rather
  than skew (skewness only `r fmt(res_skew(fit), 2)`). **Transformation does not
  help**: log(cIMT) gives residual skewness `r fmt(res_skew(fit_log), 2)` and a
  worse Shapiro-Wilk p. HC3 standard errors are the right response and change no
  conclusion. This is a useful counter-example to the reflex "skewed-looking
  outcome, take logs".
- **MINOR data defect: row `id = "control_row1"` is entirely blank** apart from
  `id`, `has_ra` and `hypertension_or_hyperchol`, so the advertised 470 rows are
  `r nrow(dat) - n_empty_row` participants plus one empty record. Harmless for
  complete-case work but it makes every "% complete" figure in the codebook one
  row pessimistic.
- **Verified clean:** `sex` ships as the strings
  `r paste(sex_vals, collapse = "/")` (the Zenodo 0/1 coding is fixed) and
  `height_cm` ranges `r ht_min`-`r ht_max` cm with `r n_ht_metres` values in
  metres (the mixed-unit problem is fixed). No RA-specific columns (DAS28,
  methotrexate, anti-TNF) are present, as intended.
- **No parameter-budget pressure**: `r nrow(dat_model)` observations for
  `r length(coef(fit)) - 1L` parameters,
  `r fmt(nrow(dat_model) / (length(coef(fit)) - 1L), 0)` per parameter. The
  constraint on this model is collinearity, not N.
