---
title: "C-reactive protein in adults with and without rheumatoid arthritis"
subtitle: "Dataset 10: Cardiovascular risk and carotid artery thickness in rheumatoid arthritis"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "cimt_rheumatoid.csv"),
                    show_col_types = FALSE)

# The outcome is CRP on the log scale (see the transformation section). `crp`
# appears in meta.json's suggested predictor list and is of course dropped from
# the predictors here: it is the outcome.
outcome    <- "log_crp"
predictors <- c("age_yrs", "sex", "has_ra", "bmi", "systolic_bp",
                "current_smoker", "statins", "antihypertensives",
                "total_cholesterol", "hdl_cholesterol", "triglycerides",
                "glucose")

lab <- function(x, l) { attr(x, "label") <- l; x }

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the clinically "unexposed" group throughout.
    sex    = lab(factor(sex, levels = c("female", "male"),
                        labels = c("Female", "Male")), "Sex"),
    has_ra = lab(factor(has_ra, levels = c(0, 1),
                        labels = c("Control", "RA")), "Rheumatoid arthritis"),
    current_smoker    = lab(factor(current_smoker, levels = c(0, 1),
                                   labels = c("No", "Yes")), "Current smoker"),
    statins           = lab(factor(statins, levels = c(0, 1),
                                   labels = c("No", "Yes")), "Taking a statin"),
    antihypertensives = lab(factor(antihypertensives, levels = c(0, 1),
                                   labels = c("No", "Yes")),
                            "Taking antihypertensives"),
    log_crp = lab(log(crp), "log CRP"),
    crp = lab(crp, "CRP (mg/L)"),
    age_yrs = lab(age_yrs, "Age (years)"),
    bmi = lab(bmi, "BMI (kg/m2)"),
    systolic_bp = lab(systolic_bp, "Systolic BP (mmHg)"),
    total_cholesterol = lab(total_cholesterol, "Total cholesterol (mmol/L)"),
    hdl_cholesterol = lab(hdl_cholesterol, "HDL cholesterol (mmol/L)"),
    triglycerides = lab(triglycerides, "Triglycerides (mmol/L)"),
    glucose = lab(glucose, "Fasting glucose (mmol/L)")
  )

# `crp` is carried alongside the log outcome for Table 1 and Figure 1 only; the
# models use `dat_fit`, which contains the log outcome and the predictors.
dat_model <- dat |> select(all_of(c("crp", outcome, predictors))) |> drop_na()
dat_fit   <- dat_model |> select(-crp)
n_dropped <- nrow(dat) - nrow(dat_model)

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
pv  <- function(p) ifelse(p < 0.001, "<0.001", paste0("= ", fmt(p, 3)))
skew <- function(x) { x <- na.omit(x); mean((x - mean(x))^3) / sd(x)^3 }

# gtsummary refuses `exponentiate = TRUE` for an lm(), which is correct in
# general but unhelpful when the outcome is already logged: exponentiating a
# log-outcome coefficient gives a ratio of geometric means, which is exactly how
# these estimates should be read. A custom tidier does the job, and the estimate
# column is widened to three decimals because several ratios sit close to 1.
tidy_exp <- function(x, exponentiate = FALSE, conf.level = 0.95, ...) {
  broom::tidy(x, conf.int = TRUE, conf.level = conf.level, exponentiate = TRUE)
}
est3 <- function(x) style_number(x, digits = 3)

crp_skew     <- skew(dat$crp)
logcrp_skew  <- skew(dat$log_crp)
p_at_floor   <- mean(dat_model$crp == 1)

sex_vals    <- sort(unique(na.omit(dat_raw$sex)))
ht_min      <- min(dat_raw$height_cm, na.rm = TRUE)
ht_max      <- max(dat_raw$height_cm, na.rm = TRUE)
n_ht_metres <- sum(dat_raw$height_cm < 3, na.rm = TRUE)
friedewald  <- dat_raw$total_cholesterol - dat_raw$hdl_cholesterol -
               dat_raw$triglycerides / 2.2
ldl_r       <- cor(dat_raw$ldl_cholesterol, friedewald, use = "complete.obs")
```

# Introduction

The source study recruited adults with rheumatoid arthritis (RA) and healthy
controls without RA in the Netherlands, and put every participant through the
same physical examination and fasting laboratory panel, including C-reactive
protein (Hollan and colleagues, *PLOS ONE*, 2015). This teaching extract holds
`r nrow(dat)` participants, `r sum(dat_raw$has_ra == 1)` with RA and
`r sum(dat_raw$has_ra == 0)` controls. C-reactive protein is the standard
clinical marker of systemic inflammation: it is the laboratory value used to
judge whether rheumatoid disease is active, and independently of that it is one
of the best-established biomarkers of cardiovascular risk, so the question of
what raises it in a mixed RA and non-RA population sits at the junction of the
two literatures.

**Aim.** Among these Dutch adults, we examined which of RA status, age, sex,
adiposity, blood pressure, smoking, lipid-lowering and antihypertensive
treatment, the fasting lipid panel and glucose are independently associated with
C-reactive protein, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** A cross-sectional case-control comparison conducted in the
Netherlands and published in *PLOS ONE* in 2015. Cases were adults with
established rheumatoid arthritis; controls were adults without RA recruited from
the same setting. This extract contains `r nrow(dat)` participants and 25
variables. Ages run from `r min(dat$age_yrs, na.rm = TRUE)` to
`r max(dat$age_yrs, na.rm = TRUE)` years.

**Variables.** The outcome is `crp`, C-reactive protein, recorded as whole
numbers running from `r min(dat$crp, na.rm = TRUE)` to
`r max(dat$crp, na.rm = TRUE)` with a median of
`r median(dat$crp, na.rm = TRUE)`; the units are not stated in the deposit but
the values are consistent with mg/L. **It is modelled on the natural-log scale**
for the reasons set out below, so every coefficient is reported as a
multiplicative ratio of geometric mean CRP. Predictors are rheumatoid arthritis
status (reference Control), age in years, sex (reference Female), BMI in kg/m2,
systolic blood pressure in mmHg, current smoking (reference No), statin use
(reference No), antihypertensive use (reference No), total cholesterol, HDL
cholesterol and triglycerides in mmol/L, and fasting glucose in mmol/L.

**`crp` is removed from the predictor set.** The catalogue's suggested predictor
list for this dataset includes `crp`, because that list is shared across all
three outcomes. Regressing CRP on itself is circular, so it is dropped here.
`r length(predictors)` predictors remain.

**Variables excluded on structural grounds.** `ldl_cholesterol` is not
independently measured in this file: against the Friedewald formula
(`total - HDL - triglycerides/2.2`) it correlates `r fmt(ldl_r, 4)`, so
including it alongside all three of its inputs creates a near-exact linear
dependency and inflates the lipid variance inflation factors into the hundreds.
`waist_cm` duplicates `bmi`
(r `r fmt(cor(dat$bmi, dat$waist_cm, use = "complete.obs"), 2)`) and
`diastolic_bp` duplicates `systolic_bp`
(r `r fmt(cor(dat$systolic_bp, dat$diastolic_bp, use = "complete.obs"), 2)`);
one of each pair is kept. `apo_a` and `apo_b` are alternative measurements of
the HDL and total-cholesterol fractions
(r `r fmt(cor(dat$apo_a, dat$hdl_cholesterol, use = "complete.obs"), 2)` and
`r fmt(cor(dat$apo_b, dat$total_cholesterol, use = "complete.obs"), 2)`).
`weight_kg` and `height_cm` are excluded because `bmi` is exactly
`weight_kg/(height_cm/100)^2`. `prednisone` is excluded because it is an
anti-inflammatory drug given almost exclusively to the RA group
(`r sum(dat_raw$prednisone == 1 & dat_raw$has_ra == 0, na.rm = TRUE)` of
`r sum(dat_raw$prednisone == 1, na.rm = TRUE)` users are controls) and would sit
directly on the pathway between RA and CRP.

**Parameter budget.** `r nrow(dat_model)` complete cases for
`r length(predictors)` parameters, all single degree-of-freedom terms, is
`r fmt(nrow(dat_model) / length(predictors), 0)` observations per parameter -
comfortably inside the 10-15 per parameter a linear model needs. The binding
constraint on this model is collinearity and interpretability, not sample size.

**Statistical analysis.** `lm()`, justified because the outcome is a continuous
laboratory measurement. Raw CRP is severely right-skewed (skewness
`r fmt(crp_skew, 1)`; the largest value, `r max(dat$crp, na.rm = TRUE)`, is
`r fmt(max(dat$crp, na.rm = TRUE) / median(dat$crp, na.rm = TRUE), 0)` times the
median), so a model on the raw scale would be dominated by a handful of
participants with an acute inflammatory episode. Taking natural logs reduces the
skewness to `r fmt(logcrp_skew, 1)` and, as reported in the Diagnostics, roughly
halves the residual skewness and doubles the explained variance, so **the
analysis is carried out on log CRP**. We fitted a univariable model for each
predictor first, then a single multivariable model containing all `r length(predictors)`.
Predictors were chosen in advance on subject-matter grounds, not by
significance. Significance threshold 0.05; estimates are reported with 95%
confidence intervals and, for readability, exponentiated to ratios of geometric
mean CRP. Assumptions were checked with `performance::check_model()` and
`check_collinearity()`, and because non-constant variance was flagged even after
transformation we additionally report HC3 heteroscedasticity-consistent standard
errors. Analysis used `r R.version.string` with tidyverse, gtsummary and
performance.

**Data integrity.** The catalogue notes that in the original Zenodo deposit
`sex` was coded 0 = woman / 1 = man and the height column mixed metres and
centimetres. Both are corrected here: `sex` ships as the strings
`r paste(sex_vals, collapse = " and ")` and `height_cm` runs from `r ht_min` to
`r ht_max` cm with `r n_ht_metres` values below 3. One defect remains: row
`id = "control_row1"` is entirely blank apart from `id`, `has_ra` and
`hypertension_or_hyperchol`.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped,
`r fmt(100 * n_dropped / nrow(dat), 1)`%). The outcome is
`r fmt(100 * mean(!is.na(dat$crp)), 1)`% complete (`r sum(is.na(dat$crp))`
missing); the rest of the attrition is spread thinly across the predictors, the
largest contributors being `statins` (`r sum(is.na(dat$statins))` missing),
`antihypertensives` (`r sum(is.na(dat$antihypertensives))`) and
`current_smoker` (`r sum(is.na(dat$current_smoker))`). Only
`r sum(complete.cases(dat_raw))` of `r nrow(dat)` rows are complete on *every*
column in the file.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  select(-all_of(outcome)) |>
  tbl_summary(
    by = has_ra,
    missing = "no"
  ) |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics of the analytic sample, by rheumatoid arthritis status.**")
```

The RA group has markedly higher CRP (median
`r median(dat_model$crp[dat_model$has_ra == "RA"])`, interquartile range
`r quantile(dat_model$crp[dat_model$has_ra == "RA"], 0.25)` to
`r quantile(dat_model$crp[dat_model$has_ra == "RA"], 0.75)`) than the controls
(median `r median(dat_model$crp[dat_model$has_ra == "Control"])`, interquartile
range `r quantile(dat_model$crp[dat_model$has_ra == "Control"], 0.25)` to
`r quantile(dat_model$crp[dat_model$has_ra == "Control"], 0.75)`), which is
exactly what an inflammatory arthritis should produce. The RA arm is also older
and more female, and differs modestly in BMI and blood pressure, but the two
groups' lipid profiles are close.

## Outcome distribution and the log transformation

```{r dist, fig.cap="Figure 1. Distribution of C-reactive protein on the raw scale (left) and after log transformation (right). The dashed line marks the lower reporting limit of 1."}
bind_rows(
  tibble(value = dat_model$crp, scale = "Raw CRP"),
  tibble(value = dat_model$log_crp, scale = "log CRP")
) |>
  mutate(scale = factor(scale, levels = c("Raw CRP", "log CRP"))) |>
  ggplot(aes(value, fill = scale)) +
  geom_histogram(bins = 30, colour = "white", linewidth = 0.2) +
  geom_vline(data = tibble(scale = factor(c("Raw CRP", "log CRP"),
                                          levels = c("Raw CRP", "log CRP")),
                           x = c(1, 0)),
             aes(xintercept = x), linetype = "dashed", colour = "grey30") +
  facet_wrap(~scale, scales = "free") +
  scale_fill_manual(values = c("Raw CRP" = "#B5495B", "log CRP" = "#3F7CAC"),
                    guide = "none") +
  labs(x = NULL, y = "Participants")
```

Raw CRP is extremely right-skewed (skewness `r fmt(crp_skew, 1)`), with a long
thin tail out to `r max(dat_model$crp)`; logging pulls the skewness down to
`r fmt(skew(dat_model$log_crp), 1)`. The transformation does not fix everything,
though: `r fmt(100 * p_at_floor, 0)`% of the analytic sample sits exactly at the
lower reporting limit of 1 (log 0), the spike at the left of the right-hand
panel. That point mass is a censoring artefact - these participants' true CRP is
"below the assay's reporting threshold", not "exactly 1" - and no transformation
can remove it.

## Exploratory figure

```{r explore, fig.cap="Figure 2. C-reactive protein (log scale) against BMI, separately for participants with rheumatoid arthritis and controls, with linear smoothers and 95% confidence bands."}
ggplot(dat_model, aes(bmi, crp, colour = has_ra, fill = has_ra)) +
  geom_point(alpha = 0.45, size = 1.5,
             position = position_jitter(width = 0, height = 0.12)) +
  geom_smooth(method = "lm", alpha = 0.18, linewidth = 0.9) +
  scale_y_continuous(trans = "log", breaks = c(1, 2, 5, 10, 20, 50)) +
  scale_colour_manual(values = c(Control = "#3F7CAC", RA = "#B5495B")) +
  scale_fill_manual(values = c(Control = "#3F7CAC", RA = "#B5495B")) +
  labs(x = "BMI (kg/m2)", y = "CRP (log scale)", colour = "Group",
       fill = "Group")
```

CRP rises with BMI at a similar rate in both groups, and the RA line sits
roughly a factor of two above the control line across the whole BMI range; the
flat band of points along the bottom of each panel is the
`r fmt(100 * p_at_floor, 0)`% of participants at the reporting floor.

## Univariable models

```{r univariable}
uv <- dat_fit |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    tidy_fun = tidy_exp,
    estimate_fun = est3
  ) |>
  modify_header(estimate = "**Ratio**")
uv

utid <- map_dfr(predictors, function(v) {
  tidy(lm(reformulate(v, outcome), data = dat_fit), conf.int = TRUE,
       exponentiate = TRUE) |>
    filter(term != "(Intercept)")
})
ug <- function(term, d = 3) {
  r <- utid[utid$term == term, ]
  sprintf("%s (95%% CI %s to %s)", fmt(r$estimate, d),
          fmt(r$conf.low, d), fmt(r$conf.high, d))
}
up <- function(term) pv(utid$p.value[utid$term == term])
```

Because the outcome is logged, each coefficient below is a **multiplicative
ratio of geometric mean CRP**; a ratio of 1 means no association. Rheumatoid
arthritis is by far the strongest crude predictor: RA cases have
`r ug("has_raRA")` times the geometric mean CRP of controls
(p `r up("has_raRA")`), i.e. roughly a
`r fmt(100 * (utid$estimate[utid$term == "has_raRA"] - 1), 0)`% higher CRP.
Adiposity follows - each additional kg/m2 of BMI is associated with
a ratio of `r ug("bmi")` (p `r up("bmi")`), which compounds to about
`r fmt(utid$estimate[utid$term == "bmi"]^5, 2)`-fold across 5 kg/m2 - then HDL
cholesterol, where each additional mmol/L is associated with a ratio of
`r ug("hdl_cholesterol")`
(p `r up("hdl_cholesterol")`), fasting glucose (`r ug("glucose")` per mmol/L,
p `r up("glucose")`), systolic pressure (`r ug("systolic_bp")` per mmHg,
p `r up("systolic_bp")`) and total cholesterol (`r ug("total_cholesterol")` per
mmol/L, p `r up("total_cholesterol")`). Age (p `r up("age_yrs")`), sex
(p `r up("sexMale")`), current smoking (p `r up("current_smokerYes")`), statin
use (p `r up("statinsYes")`), antihypertensive use
(p `r up("antihypertensivesYes")`) and triglycerides
(p `r up("triglycerides")`) show no crude association.

## Why adjustment is needed

**BMI** is the clearest confounder in this dataset, and it works on the lipid
associations. Adipose tissue secretes interleukin-6, which drives hepatic CRP
production, so BMI is a direct cause of higher CRP (a ratio of `r ug("bmi")` crude); it is
also strongly related to the lipid panel, being correlated
`r fmt(cor(dat_model$bmi, dat_model$hdl_cholesterol), 2)` with HDL cholesterol
and `r fmt(cor(dat_model$bmi, dat_model$triglycerides), 2)` with triglycerides.
The crude "low HDL means high CRP" association is therefore partly an adiposity
association wearing a lipid label: adding BMI alone moves the HDL ratio from
`r ug("hdl_cholesterol")` to
`r fmt(exp(coef(lm(log_crp ~ hdl_cholesterol + bmi, data = dat_fit)))["hdl_cholesterol"], 3)`.
The same argument applies to fasting glucose, which is correlated
`r fmt(cor(dat_model$bmi, dat_model$glucose), 2)` with BMI.

**RA status** is the second variable that must be handled explicitly, because it
is associated with almost everything else here. RA cases have higher CRP by
construction of the disease, and they also have higher BMI (mean
`r fmt(mean(dat_model$bmi[dat_model$has_ra == "RA"]), 1)` versus
`r fmt(mean(dat_model$bmi[dat_model$has_ra == "Control"]), 1)` kg/m2), higher
blood pressure, and a much lower rate of statin use
(`r fmt(100 * mean(dat_model$statins[dat_model$has_ra == "RA"] == "Yes"), 1)`%
versus
`r fmt(100 * mean(dat_model$statins[dat_model$has_ra == "Control"] == "Yes"), 1)`%
of controls). Any crude coefficient in this file is therefore partly a
case-control contrast. **Statin use** deserves its own warning: it is confounded
by indication, and here also confounded by group membership, since it is largely
the controls who are on statins. Mutual adjustment lets each ratio be read as a
comparison between participants of the same RA status, age, sex, body mass,
blood pressure, treatment and lipid profile.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_fit)

mv <- tbl_regression(fit, tidy_fun = tidy_exp, estimate_fun = est3) |>
  modify_header(estimate = "**Ratio**")
mv

tid <- tidy(fit, conf.int = TRUE, exponentiate = TRUE)
g <- function(term, d = 3) {
  r <- tid[tid$term == term, ]
  sprintf("%s (95%% CI %s to %s)", fmt(r$estimate, d),
          fmt(r$conf.low, d), fmt(r$conf.high, d))
}
gp <- function(term) pv(tid$p.value[tid$term == term])
```

```{r fit-stats}
s     <- summary(fit)
fit_raw <- lm(reformulate(predictors, "crp"), data = dat_model)
tibble(
  Statistic = c("Complete-case N", "Parameters (excl. intercept)",
                "Observations per parameter", "R-squared", "Adjusted R-squared",
                "Residual SE (log scale)",
                "R-squared, same model on untransformed CRP"),
  Value = c(as.character(nrow(dat_fit)),
            as.character(length(coef(fit)) - 1L),
            fmt(nrow(dat_fit) / (length(coef(fit)) - 1L), 1),
            fmt(s$r.squared, 3), fmt(s$adj.r.squared, 3), fmt(s$sigma, 3),
            fmt(summary(fit_raw)$r.squared, 3))
) |>
  gt() |>
  tab_header(title = md("**Model fit: multivariable linear model for log CRP**"))
```

The model explains `r fmt(100 * s$r.squared, 1)`% of the variance in log CRP
(adjusted R-squared `r fmt(s$adj.r.squared, 3)`). The same twelve predictors
applied to untransformed CRP explain only
`r fmt(100 * summary(fit_raw)$r.squared, 1)`%, so the transformation roughly
doubles the explained variance as well as fixing the skew.

Six predictors survive adjustment. The exposure of primary interest holds up:
among participants of the same age, sex, body mass, blood pressure, treatment
and lipid profile, RA cases have `r g("has_raRA")` times the geometric mean CRP
of controls (p `r gp("has_raRA")`). Among participants alike on everything else,
each additional kg/m2 of BMI is associated with a ratio of `r g("bmi")`
(p `r gp("bmi")`),
or about `r fmt(tid$estimate[tid$term == "bmi"]^5, 2)`-fold per 5 kg/m2; each
mmol/L of total cholesterol with a ratio of `r g("total_cholesterol")`
(p `r gp("total_cholesterol")`); each mmHg of systolic pressure with
a ratio of `r g("systolic_bp")` (p `r gp("systolic_bp")`); men have `r g("sexMale")` the
CRP of women (p `r gp("sexMale")`); and statin users `r g("statinsYes")` that of
non-users (p `r gp("statinsYes")`). Triglycerides fall just short
(`r g("triglycerides")`, p `r gp("triglycerides")`). Age (`r g("age_yrs")`,
p `r gp("age_yrs")`), current smoking (`r g("current_smokerYes")`,
p `r gp("current_smokerYes")`), antihypertensive use
(`r g("antihypertensivesYes")`, p `r gp("antihypertensivesYes")`), HDL
cholesterol (`r g("hdl_cholesterol")`, p `r gp("hdl_cholesterol")`) and glucose
(`r g("glucose")`, p `r gp("glucose")`) are not independently associated. These
are associations among the participants observed, not effects of intervening.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted ratios of geometric mean CRP.**")
```

Two predictors move substantially toward the null and lose significance. **HDL
cholesterol** falls from `r ug("hdl_cholesterol")` crude
(p `r up("hdl_cholesterol")`) to `r g("hdl_cholesterol")` adjusted
(p `r gp("hdl_cholesterol")`); adding BMI alone accounts for most of that shift,
so the crude association was largely adiposity. **Fasting glucose** does the
same, from `r ug("glucose")` (p `r up("glucose")`) to `r g("glucose")`
(p `r gp("glucose")`), attenuated first by BMI and then by RA status. In both
cases the explanation is confounding by body mass rather than a change in
precision: the confidence intervals are of similar width in the two columns.

Two predictors move the other way, **away from the null and into significance**.
Sex goes from `r ug("sexMale")` (p `r up("sexMale")`) to `r g("sexMale")`
(p `r gp("sexMale")`); the masking variable is HDL cholesterol, since men have
substantially lower HDL (mean
`r fmt(mean(dat_model$hdl_cholesterol[dat_model$sex == "Male"]), 2)` versus
`r fmt(mean(dat_model$hdl_cholesterol[dat_model$sex == "Female"]), 2)` mmol/L)
and low HDL travels with high CRP, so the men's lower intrinsic CRP was hidden
until HDL was held constant. Statin use goes from `r ug("statinsYes")`
(p `r up("statinsYes")`) to `r g("statinsYes")` (p `r gp("statinsYes")`); no
single covariate drives this - the lipid and glucose terms each strengthen it
while RA status weakens it - and given that statin users are disproportionately
controls, this coefficient should be treated as fragile.

The predictor of primary interest attenuates but survives comfortably: RA status
goes from `r ug("has_raRA")` to `r g("has_raRA")`, losing about
`r fmt(100 * (utid$estimate[utid$term == "has_raRA"] - tid$estimate[tid$term == "has_raRA"]) / (utid$estimate[utid$term == "has_raRA"] - 1), 0)`%
of its excess over 1, with BMI and systolic pressure accounting for most of the
attenuation. **Total cholesterol barely moves**, from
`r ug("total_cholesterol")` to `r g("total_cholesterol")`, and if anything
strengthens - nothing else in the model stands in for it. Age is unchanged and
null in both columns (`r ug("age_yrs")` to `r g("age_yrs")`), which is worth
noting because age is the dominant predictor of both other outcomes in this
dataset.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r collin}
check_collinearity(fit)
```

```{r assumption-tests}
het  <- check_heteroscedasticity(fit)
print(het)
norm <- check_normality(fit)
print(norm)
```

```{r transform-table}
res_skew <- function(m) { r <- resid(m); mean((r - mean(r))^3) / sd(r)^3 }
tibble(
  Outcome = c("CRP (untransformed)", "log CRP"),
  `Outcome skewness` = fmt(c(skew(dat_model$crp), skew(dat_model$log_crp)), 2),
  `Residual skewness` = fmt(c(res_skew(fit_raw), res_skew(fit)), 2),
  `Shapiro-Wilk p` = formatC(c(shapiro.test(resid(fit_raw))$p.value,
                               shapiro.test(resid(fit))$p.value),
                             format = "e", digits = 1),
  `R-squared` = fmt(c(summary(fit_raw)$r.squared, summary(fit)$r.squared), 3)
) |>
  gt() |>
  tab_header(title = md("**Table 3. Untransformed versus log-transformed outcome.**"))
```

```{r robust-se}
rob   <- lmtest::coeftest(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
robci <- lmtest::coefci(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
tibble(
  Term = rownames(rob),
  `Ratio` = fmt(exp(rob[, 1]), 3),
  `SE (OLS)` = fmt(summary(fit)$coefficients[, 2], 3),
  `SE (HC3)` = fmt(rob[, 2], 3),
  `95% CI (HC3), ratio scale` = paste0(fmt(exp(robci[, 1]), 3), " to ",
                                       fmt(exp(robci[, 2]), 3)),
  `p (HC3)` = fmt(rob[, 4], 3)
) |>
  filter(Term != "(Intercept)") |>
  gt() |>
  tab_header(title = md("**Table 4. HC3 heteroscedasticity-consistent standard errors, on the ratio scale.**"))
```

**The transformation was the right call, and Table 3 is the evidence.** Residual
skewness drops from `r fmt(res_skew(fit_raw), 2)` on the raw scale to
`r fmt(res_skew(fit), 2)` after logging, the Shapiro-Wilk p-value improves by
more than twenty orders of magnitude, and R-squared rises from
`r fmt(summary(fit_raw)$r.squared, 3)` to `r fmt(summary(fit)$r.squared, 3)`. On
the raw scale the residual-versus-fitted plot is a wedge dominated by a dozen
participants with CRP above 20; on the log scale it is broadly flat.

**But the log model still fails both formal checks, and the reason is the
reporting floor.** `check_heteroscedasticity()` rejects constant variance
(p < 0.001) and `check_normality()` rejects normal residuals (p < 0.001). The
posterior-predictive and normal-QQ panels show why: `r fmt(100 * p_at_floor, 0)`%
of participants sit exactly at CRP = 1, so the outcome is not really continuous
at its lower end but a point mass with a continuous distribution above it. No
model that assumes a smooth conditional distribution can reproduce that spike,
and no transformation will help. The honest description is that CRP here is
**left-censored at the assay's reporting limit** and that a tobit or
censored-normal model would be the correct specification; that is beyond what
this analysis attempts, so Table 4 reports HC3 robust standard errors instead.
They change nothing substantive - the HC3 standard error for the RA coefficient
is `r fmt(rob["has_raRA", 2], 3)` against an OLS
`r fmt(summary(fit)$coefficients["has_raRA", 2], 3)`, and every predictor keeps
its significance status under robust inference except triglycerides, which was
borderline in both.

**Collinearity is not a problem.** The largest variance inflation factor is
`r fmt(max(as.data.frame(check_collinearity(fit))$VIF), 2)`, for
`r as.data.frame(check_collinearity(fit))$Term[which.max(as.data.frame(check_collinearity(fit))$VIF)]`,
and every VIF is below 2 - well short of the conventional threshold of 5. That
follows directly from the exclusions in Methods; the same model with
`ldl_cholesterol`, `waist_cm` and `diastolic_bp` added has lipid VIFs of about
`r fmt(max(as.data.frame(check_collinearity(lm(reformulate(c(predictors, "ldl_cholesterol", "waist_cm", "diastolic_bp"), outcome), data = dat |> select(all_of(c(outcome, predictors, "ldl_cholesterol", "waist_cm", "diastolic_bp"))) |> drop_na())))$VIF), 0)`.

**No influential observations.** The largest Cook's distance is
`r fmt(max(cooks.distance(fit)), 3)` and the largest leverage
`r fmt(max(hatvalues(fit)), 3)`, so no single participant drives the fit -
another benefit of the log scale, since on the raw scale the participant with
CRP `r max(dat_model$crp)` is enormously influential.

# Discussion

In this Dutch case-control sample, C-reactive protein is raised most clearly by
rheumatoid arthritis and by adiposity. RA cases have `r g("has_raRA")` times the
geometric mean CRP of controls of the same age, sex, body mass, blood pressure
and lipid profile, and each 5 kg/m2 of BMI multiplies CRP by about
`r fmt(tid$estimate[tid$term == "bmi"]^5, 2)`. Systolic blood pressure and total
cholesterol contribute smaller independent associations, and men have about
`r fmt(100 * (1 - tid$estimate[tid$term == "sexMale"]), 0)`% lower CRP than
women once lipids are held constant. Practically, the finding that matters is
that a substantial part of the raised CRP in RA patients is not the arthritis:
adjustment for conventional cardiovascular risk factors removes about
`r fmt(100 * (utid$estimate[utid$term == "has_raRA"] - tid$estimate[tid$term == "has_raRA"]) / (utid$estimate[utid$term == "has_raRA"] - 1), 0)`%
of the crude excess, most of it attributable to body mass and blood pressure.
Clinicians reading a raised CRP in an RA patient as evidence of active joint
disease should note how much of it adiposity can explain.

The statin coefficient is the one that must not be over-read. Taken at face
value, the statin ratio of `r g("statinsYes")` says statin users have around
`r fmt(100 * (1 - tid$estimate[tid$term == "statinsYes"]), 0)`% lower CRP, and
there is a real pharmacological literature reporting exactly that. But statins
in this dataset are **confounded by indication in both directions at once**.
People are prescribed statins because their lipids and cardiovascular risk are
poor: users here are on average
`r fmt(mean(dat_model$age_yrs[dat_model$statins == "Yes"]) - mean(dat_model$age_yrs[dat_model$statins == "No"]), 1)`
years older and, despite treatment, still have total cholesterol of
`r fmt(mean(dat_model$total_cholesterol[dat_model$statins == "Yes"]), 1)` mmol/L
against `r fmt(mean(dat_model$total_cholesterol[dat_model$statins == "No"]), 1)`
in non-users - that pushes their apparent CRP up. At the same time, statin use
is concentrated in the control group
(`r fmt(100 * mean(dat_model$statins[dat_model$has_ra == "Control"] == "Yes"), 1)`%
of controls versus
`r fmt(100 * mean(dat_model$statins[dat_model$has_ra == "RA"] == "Yes"), 1)`% of
RA cases), and controls have much lower CRP for reasons that have nothing to do
with statins - that pushes it down. The adjusted coefficient additionally
conditions on total cholesterol and HDL, which are the very quantities statins
are given to change, so it is adjusting for a mediator. The same logic applies
to `antihypertensives`. Neither column of Table 2 is a treatment effect, and a
randomised comparison is the only thing that would be.

**Limitations.** The design is cross-sectional, so nothing here establishes
temporal order; in particular the BMI-CRP association is bidirectional in
principle, though the adipose-tissue mechanism makes the BMI-to-CRP direction the
more plausible one. **CRP is left-censored**: `r fmt(100 * p_at_floor, 0)`% of
participants sit at the reporting limit of 1, and the linear model treats these
as exact values, which biases coefficients toward the null and is the direct
cause of the failed normality and variance checks. CRP is also an *acute-phase*
reactant, so a single measurement partly reflects incidental infection at the
time of sampling rather than chronic inflammatory burden; that is
non-differential noise which will attenuate all associations. RA disease
activity, disease duration and disease-modifying treatment - the variables that
actually determine CRP within the RA group - are **not in this extract at all**,
being undefined for the `r sum(dat_raw$has_ra == 0)` controls, so the RA
coefficient is an average across patients whose disease may be quiescent or
raging. `prednisone`, an anti-inflammatory that directly suppresses CRP, was
excluded as a pathway variable, so its effect is folded into the RA coefficient.
Statins and antihypertensives are confounded by indication as described above.
Complete-case analysis dropped `r n_dropped` of `r nrow(dat)` rows. The sample is
a single Dutch centre. All estimates are associations, not causal effects.

**Future work.** Refit as a censored-normal (tobit) model with the reporting
limit at 1, and, within the RA cases only, replace the RA indicator with DAS28
disease activity from the full deposit to test whether CRP tracks disease
activity or simply the RA label.

# Feasibility notes

*NOT part of a student report.*

- **USABLE WITH WARNINGS.** `r nrow(dat_fit)` complete cases, R-squared
  `r fmt(s$r.squared, 3)` on the log scale, a strong and substantively sensible
  signal (RA status and BMI), and no collinearity or influence problems. The
  warnings are about the outcome's distribution, not about the model.
- **FATAL if untransformed.** Raw CRP has skewness `r fmt(crp_skew, 1)` and runs
  from `r min(dat$crp, na.rm = TRUE)` to `r max(dat$crp, na.rm = TRUE)`. The
  untransformed model has residual skewness `r fmt(res_skew(fit_raw), 2)`,
  R-squared `r fmt(summary(fit_raw)$r.squared, 3)` (against
  `r fmt(s$r.squared, 3)` logged), and is dominated by a dozen participants in an
  acute-phase response. A student who reports untransformed CRP has an
  uninterpretable model, not merely an assumption violation. **The catalogue's
  "consider log-transforming" note should be firmed up to "log-transform".**
- **MAJOR: CRP is left-censored at the reporting limit.**
  `r fmt(100 * p_at_floor, 0)`% of the analytic sample has CRP exactly 1, which
  is "below the assay threshold" rather than a measured value. This point mass is
  why `check_normality()` and `check_heteroscedasticity()` still reject **after**
  logging, and it is not fixable by transformation. Students will report failed
  diagnostics no matter what they do; the correct answer is to name the censoring
  and move on (or fit a tobit model, which is beyond the course). This should be
  in the caution list.
- **MAJOR: `crp` is in `meta.json`'s suggested predictor list** and must be
  dropped when CRP is the outcome. A student who leaves it in gets R-squared
  = 1. This is the one genuinely circular pairing in the dataset.
- **MAJOR trap: `ldl_cholesterol` is the Friedewald calculation, not a
  measurement** (correlation `r fmt(ldl_r, 4)` with
  `total - HDL - triglycerides/2.2`). Including it with all three inputs sends
  the lipid VIFs into the hundreds.
- **MAJOR interpretive trap: `statins` and `antihypertensives` are confounded by
  indication**, and here the confounding is doubly awkward because statin use is
  concentrated in the *control* group
  (`r fmt(100 * mean(dat_model$statins[dat_model$has_ra == "Control"] == "Yes"), 1)`%
  of controls versus
  `r fmt(100 * mean(dat_model$statins[dat_model$has_ra == "RA"] == "Yes"), 1)`%
  of RA cases). The statin coefficient crosses into significance on adjustment
  (`r ug("statinsYes")` to `r g("statinsYes")`) and looks like a beautiful
  anti-inflammatory effect. It is not evidence of one.
- **MINOR: `prednisone` sits on the causal pathway.** It is an anti-inflammatory
  given almost exclusively to RA patients
  (`r sum(dat_raw$prednisone == 1 & dat_raw$has_ra == 0, na.rm = TRUE)` of
  `r sum(dat_raw$prednisone == 1, na.rm = TRUE)` users are controls). Adding it
  to a model with `has_ra` is adjusting for a mediator and shrinks the RA
  coefficient for the wrong reason.
- **MINOR: redundant pairs.** `bmi`/`waist_cm`
  (r `r fmt(cor(dat$bmi, dat$waist_cm, use = "complete.obs"), 2)`),
  `systolic_bp`/`diastolic_bp`
  (r `r fmt(cor(dat$systolic_bp, dat$diastolic_bp, use = "complete.obs"), 2)`),
  `apo_a`/`hdl_cholesterol`
  (r `r fmt(cor(dat$apo_a, dat$hdl_cholesterol, use = "complete.obs"), 2)`),
  `apo_b`/`total_cholesterol`
  (r `r fmt(cor(dat$apo_b, dat$total_cholesterol, use = "complete.obs"), 2)`).
  With the chosen subset the maximum VIF is
  `r fmt(max(as.data.frame(check_collinearity(fit))$VIF), 2)`.
- **MINOR: the units of `crp` are not documented.** The codebook says only
  "C-reactive protein (inflammatory marker)". The integer-valued 1-71 range is
  consistent with mg/L but nothing in the deposit confirms it, so students cannot
  compare their numbers with published reference ranges with any confidence.
- **MINOR: CRP is an acute-phase reactant measured once.** Incidental infection
  at the time of sampling is measurement error that will attenuate every
  association; there is no way to detect or adjust for it here.
- **MINOR data defect: row `id = "control_row1"` is entirely blank.**
- **Verified clean:** `sex` ships as `r paste(sex_vals, collapse = "/")` and
  `height_cm` as `r ht_min`-`r ht_max` cm with `r n_ht_metres` values in metres,
  so both documented Zenodo defects are fixed in this catalogue's CSV. No
  RA-specific columns are present.
