---
title: "Regression diagnostics: SHIN checks for simple linear regression"
subtitle: "Reading check_model() diagnostic plots for shape, homoscedasticity, influential points, and normality"
author: "RSR Resources"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 7,
  fig.align = "center"
)
```

# Setup

```{r libraries}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, performance, qqplotr)
```

## SUNSHINE and SHIN

We created an acronym to help you remember the eight checks you should keep in mind when fitting a linear regression model.

- **S** (**Sample size**): do you have enough observations for the model?
- **U** (**Uncorrelated errors**): are residuals independent of each other?
- **N** (**No multicollinearity**): are the predictors not too correlated?
- **S** (**Shape / linearity**): is the relationship actually a straight line?
- **H** (**Homogeneity of variance**): does the spread of residuals stay constant?
- **I** (**Influential points**): is any single observation pulling the line around?
- **N** (**Normality of residuals**): are the prediction errors roughly bell-shaped?
- **E** (**Exogeneity**): are the predictors not caused by the outcome?

This script focuses on **SHIN** (Shape, Homoscedasticity, Influential points, and Normality), because those are the four assumptions you can check directly in R with `check_model()` on a simple linear regression.

The first **N** in SUNSHINE (no multicollinearity) is also relevant, and `check_model()` can plot it, but it only applies once you have two or more predictors.
That is multiple linear regression, which we have not reached yet.


# S: Shape (Linearity)

**The question:** is the relationship actually a straight line?

## Study hours and quiz score (non-linear relationship)

Weekly study hours vs. quiz score.
Going from no studying to moderate studying helps a lot; extra hours past that add little because scores are near the maximum of the scale.
```{r shape-bad-data}
shape_bad <- data.frame(
  study_hours = c(2, 4, 6, 9, 11, 13, 15, 17, 18, 20, 21, 23, 26, 29, 31, 34, 37, 39, 41, 43, 46, 48, 51, 53, 54, 56, 58, 60, 61, 62),
  quiz_score  = c(19.5, 30, 46.2, 55.9, 65.9, 68.3, 76.5, 76.4, 81.6, 81.3, 86.3, 84.4, 91.2, 88.8, 93, 90.4, 95.6, 92.2, 95.9, 92.6, 96.2, 92.9, 97.5, 92.9, 97.4, 93.5, 97.1, 93.4, 97.8, 93.2)
)
```

Look at the raw scatter first; you can often already see the bend.

```{r shape-bad-scatter}
ggplot(shape_bad, aes(study_hours, quiz_score)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#C83B3B") +
  labs(title = "Study hours vs quiz score: non-linear relationship fitted with a straight line") +
  theme_minimal(base_size = 14)
```

Next, inspect the diagnostic panel, especially the Linearity panel.

```{r shape-bad-fit}
shape_bad_model <- lm(quiz_score ~ study_hours, data = shape_bad)
summary(shape_bad_model)
```

```{r shape-bad-check}
check_model(shape_bad_model)
```

The green line in the *Linearity* panel arches, indicating a non-linear relationship.

So your model overestimates the quiz score for students who study for a few hours, and underestimates it for students who study for a lot of hours.

In such cases, you may need to transform your data or use a different kind of model.

## Exercise hours and VO₂ max (linear relationship)

Weekly exercise hours vs. VO₂ max: a linear relationship.

```{r shape-good-data}
shape_good <- data.frame(
  exercise_hours = c(1, 1.5, 2, 2.5, 3, 3.5, 4, 4.5, 5, 5.5, 6, 6.5, 7, 7.5, 8, 8.5, 9, 9.5, 10, 10.5, 2, 3, 4.5, 6, 7, 8, 9, 1.5, 5, 8.5),
  vo2_max        = c(31.6, 29.4, 34.2, 30, 33.8, 30.6, 36.4, 34.2, 40, 34.8, 38.6, 35.4, 41.2, 36, 43.8, 40.6, 44.4, 41.2, 45, 41.8, 31.2, 34.8, 34.2, 40.6, 37.2, 41.8, 41.4, 32.4, 33, 42.6)
)
```

```{r shape-good-scatter}
ggplot(shape_good, aes(exercise_hours, vo2_max)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#C83B3B") +
  labs(title = "Exercise hours vs VO₂ max: linear relationship") +
  theme_minimal(base_size = 14)

```

```{r shape-good-fit}
shape_good_model <- lm(vo2_max ~ exercise_hours, data = shape_good)
summary(shape_good_model)
```

```{r shape-good-check}
check_model(shape_good_model)
```

The *Linearity* green line is flat and near zero, showing no leftover pattern.

Note: For this check, and with all others, it is sometimes difficult to be completely sure whether a model 'passes' a test. 

There are statistical tests that can help, but they are increasinly looked down upon, in part because they are too sensitive on large datasets.

Your best bet is to solicit the judgment of others, and simply report your diagnostic plots (usually as an appendix) when you are publishing a paper, for example.

# H: Homoscedasticity (Equal Variance)

The question here is: does the *spread* of the residuals stay constant across the range?

For this, we read the **Homogeneity of variance** panel (scale-location).

A flat green line means even spread. If the line slopes upward, then the spread of the residuals increases with the fitted value.

## Income and medical spending (heteroscedasticity)

Household income vs. out-of-pocket medical spending.

Wealthier households have far more discretionary spending, so the spread gets larger at the high end, giving you a fan-shaped scatter.

```{r hom-bad-data}
hom_bad <- data.frame(
  income_k      = c(20, 25, 28, 32, 35, 38, 42, 45, 48, 52, 55, 58, 62, 65, 68, 72, 75, 80, 85, 90, 95, 100, 110, 120, 22, 40, 56, 70, 88, 105),
  medical_spend = c(464, 475, 573, 622, 495, 786, 593, 943, 744, 828, 682, 1052, 1000, 919, 842, 1294, 851, 1526, 961, 994, 1573, 1052, 2024, 2030, 555, 560, 1152, 718, 1729, 855)
)
```

```{r hom-bad-scatter}
ggplot(hom_bad, aes(income_k, medical_spend)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#C83B3B") +
  labs(title = "Income vs medical spending: residual spread increases with income") +
  theme_minimal(base_size = 14)
```

```{r hom-bad-fit}
hom_bad_model <- lm(medical_spend ~ income_k, data = hom_bad)
summary(hom_bad_model)
```

```{r hom-bad-check}
check_model(hom_bad_model)
```

The *Homogeneity of variance* green line slopes upward, indicating that the spread of the residuals increases with the fitted value. 

This is a sign of heteroscedasticity.

It means that your model's standard errors are too large for high-income households, and too small for low-income households.

## Exercise duration and calories (homoscedastic variance)

Exercise duration vs. calories burned.

The spread looks the same for short and long workouts.

```{r hom-good-data}
hom_good <- data.frame(
  exercise_min = c(15, 20, 22, 25, 28, 30, 33, 35, 37, 40, 42, 45, 48, 50, 52, 55, 58, 60, 63, 65, 18, 27, 34, 39, 44, 49, 54, 57, 62, 36),
  calories     = c(149, 158, 193, 185, 218, 210, 250, 246, 281, 271, 299, 298, 338, 334, 366, 360, 393, 386, 426, 419, 167, 197, 255, 261, 311, 323, 377, 374, 419, 247)
)
```

```{r hom-good-scatter}
ggplot(hom_good, aes(exercise_min, calories)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#C83B3B") +
  labs(title = "Exercise duration vs calories: even spread across the range") +
  theme_minimal(base_size = 14)
```

```{r hom-good-fit}
hom_good_model <- lm(calories ~ exercise_min, data = hom_good)
summary(hom_good_model)
```

```{r hom-good-check}
check_model(hom_good_model)
```

Here the *Homogeneity* green line is roughly flat.

This means that the spread of the residuals is roughly constant across the range of the predictor.

# I: Influential Points

The question here is: is any single observation pulling the line around?

For this, we read the **Influential observations** panel (residuals vs leverage).

Points that sit *outside the dashed contour* combine an unusual x-value (leverage) with a large residual (off the line).

## BMI and cholesterol with data-entry error (influential point)

A cholesterol screening with a likely data-entry error: BMI = 42 paired with cholesterol = 130, which is unusually low for that BMI.

```{r inf-bad-data}
inf_bad <- data.frame(
  bmi         = c(19, 20, 21, 22, 23, 23.5, 24, 24.5, 25, 25.5, 26, 26.5, 27, 27.5, 28, 28.5, 29, 29.5, 30, 30.5, 31, 20.5, 23, 25.5, 27, 29, 22, 24.5, 26.5, 28.5, 42),
  cholesterol = c(216, 204, 229, 211, 228, 214, 236, 226, 249, 225, 243, 225, 252, 238, 258, 240, 265, 243, 261, 245, 276, 220, 214, 241, 234, 265, 213, 242, 229, 256, 130)
)
```

```{r inf-bad-scatter}
ggplot(inf_bad, aes(bmi, cholesterol)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#C83B3B") +
  labs(title = "BMI vs cholesterol: influential observation at BMI 42") +
  theme_minimal(base_size = 14)
```

To diagnose, we look at the *Influential observations* panel in the check_model() output.

```{r inf-bad-check}
inf_bad_model  <- lm(cholesterol ~ bmi, data = inf_bad)
check_model(inf_bad_model)
```

In the *Influential observations* panel, one point lies far to the right, outside the dashed contour.
It has both high leverage (BMI 42 is well above the rest of the data) and a large residual (far from the fitted line).
Removing it would noticeably change the fit.

## BMI and cholesterol without outliers (no influential points)

The same screening, without the erroneous record.

```{r inf-good-data}
inf_good <- inf_bad %>% filter(bmi != 42)
```

```{r inf-good-scatter}
ggplot(inf_good, aes(bmi, cholesterol)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#C83B3B") +
  labs(title = "BMI vs cholesterol: no influential observations") +
  theme_minimal(base_size = 14)
```

```{r inf-good-fit}
inf_good_model <- lm(cholesterol ~ bmi, data = inf_good)
summary(inf_good_model)
```

```{r inf-good-check}
check_model(inf_good_model)
```

Every point stays well inside the dashed contour in the *Influential observations* panel.
No single observation stands out as influential.

# N: Normality of Residuals

**The question:** are the prediction errors roughly bell-shaped?
We read the **Normality of residuals** panel (the Q-Q plot).
Points close to the reference line are good; points diverging at the ends indicate skew or heavy tails.

## Poverty rate and ER visits (non-normal residuals)

Neighbourhood poverty rate vs. ER visits per 1,000.
Most neighbourhoods follow the trend, but several have unusually high visit rates, producing strongly right-skewed residuals.

```{r norm-bad-data}
norm_bad <- data.frame(
  poverty_rate = c(5, 7, 8, 10, 12, 13, 15, 16, 18, 19, 20, 22, 23, 25, 27, 28, 30, 32, 34, 35, 37, 38, 40, 42, 6, 14, 21, 29, 36, 41),
  er_visits    = c(113, 105, 107, 112, 213, 122, 119, 124, 205, 154, 150, 159, 243, 155, 176, 178, 280, 191, 207, 199, 280, 195, 222, 220, 99, 129, 219, 181, 211, 296)
)
```

```{r norm-bad-scatter}
ggplot(norm_bad, aes(poverty_rate, er_visits)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#C83B3B") +
  labs(title = "Poverty rate vs ER visits: right-skewed residuals") +
  theme_minimal(base_size = 14)
```

```{r norm-bad-fit}
norm_bad_model <- lm(er_visits ~ poverty_rate, data = norm_bad)
summary(norm_bad_model)
```

```{r norm-bad-check}
check_model(norm_bad_model)
```

In the *Normality* Q-Q panel, the points bend away from the reference line at the upper end.
That is the long right tail of the skewed residuals.

## Height and lung capacity (normal residuals)

Height vs. lung capacity.
Errors scatter symmetrically around the line.

```{r norm-good-data}
norm_good <- data.frame(
  height_cm     = c(154.7, 155.6, 155.8, 156.3, 157.6, 158, 158.6, 159.4, 159.6, 160.6, 160.4, 161, 162, 162.4, 163.3, 164.2, 164.8, 164.7, 165.8, 166.4, 166.6, 167.1, 167.8, 169, 168.9, 169.7, 169.9, 171.4, 171.4, 171.7, 173, 173.5, 173.7, 174.7, 174.9, 175.4, 175.9, 176.8, 177.4, 177.8, 179.1, 179, 180.1, 180.5, 181.2, 182.2, 181.9, 182.9, 183.8, 184, 185.1, 185.8, 185.8, 185.9, 187.1, 188, 188.4, 188.3, 189, 190.1),
  lung_capacity = c(3.54, 3.88, 3.61, 3.92, 3.98, 3.75, 3.43, 3.89, 3.59, 3.67, 3.47, 3.7, 3.98, 3.41, 3.78, 4.08, 3.85, 3.91, 3.77, 3.93, 3.82, 3.72, 3.99, 4.2, 4.27, 4.1, 3.78, 4.13, 4.15, 4.4, 4.08, 4.16, 4.2, 4.21, 4.54, 4.67, 4.55, 4.54, 4.54, 4.34, 4.62, 4.53, 4.49, 4.57, 4.67, 4.41, 4.25, 4.92, 4.64, 4.8, 4.63, 4.8, 5.03, 4.88, 4.94, 4.92, 5.01, 4.93, 4.8, 5.17)
)
```

```{r norm-good-scatter}
ggplot(norm_good, aes(height_cm, lung_capacity)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#C83B3B") +
  labs(title = "Height vs lung capacity: symmetric scatter around the line") +
  theme_minimal(base_size = 14)
```

```{r norm-good-fit}
norm_good_model <- lm(lung_capacity ~ height_cm, data = norm_good)
summary(norm_good_model)
```

```{r norm-good-check}
check_model(norm_good_model)
```

The Q-Q points sit on (or very near) the diagonal across the whole range, and inside the shaded band.
That is what approximately normal residuals look like.
