---
title: "Haemoglobin concentration in Peruvian children aged 6-59 months"
subtitle: "Dataset 5: Peru DHS child anaemia and growth extract"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "peru_child_anaemia.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "haemoglobin_g_dl"
predictors <- c("child_age_months", "child_sex", "natural_region",
                "wealth_index", "mother_education", "mother_indigenous")

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    child_sex = factor(child_sex, levels = c("female", "male"),
                       labels = c("Female", "Male")),
    # Lima metropolitan is the best-resourced, lowest-altitude region and is the
    # natural comparator for the three other regions.
    natural_region = factor(natural_region,
                            levels = c("lima_metropolitan", "rest_of_coast",
                                       "highlands", "jungle"),
                            labels = c("Lima metropolitan", "Rest of coast",
                                       "Highlands", "Jungle")),
    # Poorest is both the largest quintile in this extract and the group with the
    # greatest expected burden, so every other quintile is read against it.
    wealth_index = factor(wealth_index,
                          levels = c("poorest", "poorer", "middle",
                                     "richer", "richest"),
                          labels = c("Poorest", "Poorer", "Middle",
                                     "Richer", "Richest")),
    # Lowest schooling is the reference so higher categories read as "more
    # education than the reference".
    mother_education = factor(mother_education,
                              levels = c("primary_or_none", "secondary", "higher"),
                              labels = c("Primary or none", "Secondary", "Higher")),
    mother_indigenous = factor(mother_indigenous, levels = c(0, 1),
                               labels = c("No", "Yes"))
  ) |>
  labelled::set_variable_labels(
    haemoglobin_g_dl  = "Haemoglobin (g/dL)",
    child_age_months  = "Child age (months)",
    child_sex         = "Child sex",
    natural_region    = "Natural region",
    wealth_index      = "Household wealth quintile",
    mother_education  = "Mother's education",
    mother_indigenous = "Mother indigenous"
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)
```

# Introduction

The Peruvian Demographic and Health Survey (ENDES) measures altitude-adjusted
haemoglobin in every surveyed child aged 6-59 months, alongside detailed
household, maternal and geographic information. Childhood anaemia remains one of
Peru's most persistent public-health problems: roughly a third of preschool
children in this extract fall below the 11 g/dL threshold, with a strong
socio-economic and regional patterning that has been described in DHS-based
analyses of ENDES (for example Gonzales and colleagues' work on anaemia and
altitude in Peruvian children). Haemoglobin is analysed here on its continuous
scale rather than dichotomised, which preserves information and avoids the loss
of power that comes with cut-points.

**Aim.** Among Peruvian children aged 6-59 months, we examined which of child
age, child sex, natural region, household wealth, maternal education and
maternal indigenous identity are independently associated with altitude-adjusted
haemoglobin concentration, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** A cross-sectional extract of `r nrow(dat_raw)` children aged
6-59 months from the Peru Demographic and Health Survey (ENDES), a nationally
implemented household survey. Each row is one child with a valid
altitude-adjusted haemoglobin measurement, linked to their mother's interview
and to household characteristics.

**Variables.** The outcome is `haemoglobin_g_dl`, altitude-adjusted haemoglobin
in g/dL, measured by HemoCue on a capillary blood sample. Predictors are child
age in months (continuous, 6-59); child sex (reference *Female*); natural region
of Peru (reference *Lima metropolitan*, the lowest-altitude and best-resourced
region, versus rest of coast, highlands and jungle); household wealth quintile
(reference *Poorest*); mother's highest education (reference *Primary or none*);
and whether the mother belongs to an indigenous ethnic group (reference *No*).
These six were chosen in advance because they capture the three routes by which
haemoglobin varies in this age group: biological maturation (age, sex),
geography and altitude-related erythropoiesis (region), and household
socio-economic position (wealth, education, ethnicity).

**Statistical analysis.** Linear regression via `lm()`, justified because the
outcome is continuous and approximately symmetric. We fitted a univariable model
for each predictor first, then a single multivariable model containing all of
them. Predictors were chosen in advance on subject-matter grounds, not by
significance. Significance threshold 0.05; estimates are reported with 95%
confidence intervals. Assumptions were checked with `performance::check_model()`.
Analysis used `r R.version.string` with tidyverse, gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped). All six predictors and the outcome
are complete in this extract, so listwise deletion removes
`r if (n_dropped == 0) "no rows at all" else paste(n_dropped, "rows")` and
selection bias from missingness is not a concern for this particular model.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(
    missing = "no"
  ) |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

```{r table-one-numbers}
hb_med <- median(dat_model$haemoglobin_g_dl)
hb_q   <- quantile(dat_model$haemoglobin_g_dl, c(.25, .75))
pct_poorest <- round(100 * mean(dat_model$wealth_index == "Poorest"), 1)
pct_high    <- round(100 * mean(dat_model$natural_region == "Highlands"), 1)
```

Median haemoglobin was `r hb_med` g/dL (IQR `r hb_q[1]` to `r hb_q[2]`), which
sits only marginally above the 11 g/dL anaemia threshold, so a large share of
this sample is close to or below the cut-point. The sample is weighted towards
the poorer end of the wealth distribution (`r pct_poorest`% in the poorest
quintile) and towards the highlands (`r pct_high`%), reflecting the DHS design
which oversamples rural and less-populated departments.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Haemoglobin against child age, by natural region, with LOESS smoothers."}
ggplot(dat_model, aes(child_age_months, haemoglobin_g_dl, colour = natural_region)) +
  geom_point(alpha = 0.05, size = 0.5) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 1) +
  geom_hline(yintercept = 11, linetype = "dashed") +
  scale_colour_viridis_d(option = "D", end = 0.9) +
  labs(x = "Child age (months)", y = "Haemoglobin (g/dL)",
       colour = "Natural region",
       caption = "Dashed line: 11 g/dL anaemia threshold")
```

Haemoglobin rises steeply across the second year of life and then flattens after
about 24 months in every region, and the four regional curves are separated by
roughly 0.4 g/dL across the whole age range, with the highlands and jungle
sitting below Lima and the rest of the coast.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv
```

```{r uv-numbers}
uv_age <- lm(haemoglobin_g_dl ~ child_age_months, data = dat_model) |>
  tidy(conf.int = TRUE) |> filter(term == "child_age_months")
```

Child age is the strongest single predictor: each additional month of age is
associated with a `r round(uv_age$estimate, 4)` g/dL higher haemoglobin (95% CI
`r round(uv_age$conf.low, 4)` to `r round(uv_age$conf.high, 4)`, p < 0.001),
which over the 53-month span of the sample amounts to about
`r round(uv_age$estimate * 53, 2)` g/dL between the youngest and oldest children
-- clinically the single largest contrast in the table. All other predictors
also reach significance in crude models, which is unsurprising at this sample
size and is a reason to read the effect sizes rather than the p-values.

## Why adjustment is needed

The crude regional and ethnic contrasts are almost certainly confounded by
household socio-economic position and by child age. **Household wealth** is
associated with region -- the highlands and jungle contain almost no children in
the richest quintile whereas Lima contains almost no children in the poorest --
and wealth is independently associated with haemoglobin through diet quality,
access to iron supplementation and infection burden; so part of any raw
"highlands effect" is really a poverty effect. **Child age** confounds the
socio-economic contrasts in the opposite direction: haemoglobin climbs sharply
with age in this window, and the age distribution is not identical across wealth
quintiles or regions, so a crude comparison between, say, indigenous and
non-indigenous mothers partly compares younger children with older ones.
**Maternal education** plays the same role for the indigenous contrast:
indigenous mothers in this extract have systematically less schooling, and
schooling predicts feeding practice and health-service use, so an unadjusted
indigenous coefficient absorbs the education effect. Mutual adjustment separates
these overlapping gradients.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv
```

```{r fit-stats}
gl <- glance(fit)
tibble(
  Statistic = c("R-squared", "Adjusted R-squared", "Residual SD (g/dL)",
                "Observations", "Model parameters"),
  Value = c(round(gl$r.squared, 3), round(gl$adj.r.squared, 3),
            round(sigma(fit), 3), nobs(fit), length(coef(fit)))
) |>
  gt() |>
  tab_header(title = "Model fit")
```

```{r mv-numbers}
tid <- tidy(fit, conf.int = TRUE)
g <- function(t) tid |> filter(term == t)
mv_age  <- g("child_age_months")
mv_jung <- g("natural_regionJungle")
mv_rich <- g("wealth_indexRichest")
mv_indig <- g("mother_indigenousYes")
mv_sex  <- g("child_sexMale")
```

The model explains `r round(100 * gl$r.squared, 1)`% of the variance in
haemoglobin (adjusted R-squared `r round(gl$adj.r.squared, 3)`), so the great
majority of between-child variation is left unexplained by these six
characteristics. Conditional on the other five predictors, each additional month
of age is associated with a `r round(mv_age$estimate, 4)` g/dL higher
haemoglobin (95% CI `r round(mv_age$conf.low, 4)` to
`r round(mv_age$conf.high, 4)`); children in the jungle average
`r round(mv_jung$estimate, 2)` g/dL relative to Lima
(95% CI `r round(mv_jung$conf.low, 2)` to `r round(mv_jung$conf.high, 2)`); and
children in the richest quintile average `r round(mv_rich$estimate, 2)` g/dL
compared with the poorest (95% CI `r round(mv_rich$conf.low, 2)` to
`r round(mv_rich$conf.high, 2)`). Boys average `r round(mv_sex$estimate, 3)`
g/dL relative to girls, a difference that only just clears p = 0.05 and is far
too small to matter clinically.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

```{r compare-numbers}
uv_fit <- function(v) {
  lm(reformulate(v, outcome), data = dat_model) |> tidy(conf.int = TRUE)
}
u_indig <- uv_fit("mother_indigenous") |> filter(term == "mother_indigenousYes")
u_rich  <- uv_fit("wealth_index") |> filter(term == "wealth_indexRichest")
u_jung  <- uv_fit("natural_region") |> filter(term == "natural_regionJungle")
u_sec   <- uv_fit("mother_education") |> filter(term == "mother_educationSecondary")
mv_sec  <- g("mother_educationSecondary")
u_age   <- uv_age
```

Several estimates move substantially, and all of them toward the null. The
**richest-versus-poorest** wealth contrast shrinks from
`r round(u_rich$estimate, 2)` g/dL crude to `r round(mv_rich$estimate, 2)` g/dL
adjusted, a `r round(100 * (1 - mv_rich$estimate / u_rich$estimate))`% reduction,
because much of the raw wealth gradient is really a regional and educational
gradient. The **maternal indigenous** coefficient moves from
`r round(u_indig$estimate, 2)` g/dL crude to `r round(mv_indig$estimate, 2)` g/dL
adjusted, consistent with indigenous households being concentrated in the poorer
quintiles and in the highlands: roughly two thirds of the crude indigenous
deficit is confounding by wealth, education and region. The **jungle** contrast
likewise halves, from `r round(u_jung$estimate, 2)` to
`r round(mv_jung$estimate, 2)` g/dL. By contrast **child age** barely moves
(`r round(u_age$estimate, 4)` to `r round(mv_age$estimate, 4)` g/dL per month),
as expected for a variable essentially unrelated to the household covariates.
One predictor changes significance and even sign: **secondary versus primary or
no maternal education** is `r round(u_sec$estimate, 3)` g/dL crude
(p `r format.pval(u_sec$p.value, digits = 2, eps = 0.001)`) but
`r round(mv_sec$estimate, 3)` g/dL adjusted
(p = `r round(mv_sec$p.value, 2)`) -- the crude "benefit" of secondary schooling
was wealth and region in disguise. With n = `r nobs(fit)` the fact that most
other predictors stay significant is unremarkable: at this size even trivial
differences clear p = 0.05, so the comparison worth making is between effect
sizes, not between p-values.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r diag-numbers}
vifs <- check_collinearity(fit)
max_vif <- round(max(vifs$VIF), 2)
```

The **posterior-predictive check** and **linearity** panel show the fitted model
tracking the observed haemoglobin distribution well, with only a mild
under-prediction in the upper tail. The **homogeneity of variance** panel is
close to flat, so constant residual variance is a defensible assumption. The
**collinearity** panel is entirely in the low range (maximum VIF
`r max_vif`), so the wealth, region and education coefficients can be
interpreted separately despite their correlation. **Influential observations**:
no point approaches Cook's distance thresholds, which is expected with
`r nobs(fit)` observations. The **normality of residuals** panel shows mild
departure in the tails only; with a sample this large the central limit theorem
makes the coefficient standard errors robust to that. The most substantive
diagnostic problem is not visible in these panels at all: the LOESS curves in
Figure 1 show haemoglobin rising steeply then plateauing with age, so a single
linear age term is a simplification and a spline or quadratic term would fit
better.

# Discussion

Haemoglobin in Peruvian preschool children is dominated by age: the youngest
children are the anaemic ones, and haemoglobin climbs roughly
`r round(mv_age$estimate * 12, 2)` g/dL per year of life over this window,
independent of household circumstances. Region and wealth add smaller but real
gradients -- children in the jungle and in the poorest quintile average a few
tenths of a gram per decilitre lower than their Lima and richest-quintile
counterparts. Practically, this points supplementation and screening effort at
children in their second year of life, and at jungle and highland communities,
rather than at broad socio-economic categories.

**Limitations.** This is a cross-sectional survey, so all estimates are
associations and not causal effects; a child's haemoglobin today reflects diet,
infection and supplementation history that this extract does not measure at all
(iron supplementation, recent fever or diarrhoea, dietary diversity, maternal
anaemia). The source ENDES analysis used the DHS complex design -- cluster
(HV001), stratum (HV022) and sampling weights -- whereas this extract contains
only the weight variable and it is deliberately not used here; consequently the
standard errors are too narrow, the estimates are not design-corrected, and
**this sample must not be described as nationally representative**. With
`r nrow(dat_model)` rows almost every comparison is statistically significant,
so conclusions should lead with effect sizes: a sex difference of
`r round(mv_sex$estimate, 3)` g/dL is clinically meaningless whatever its
p-value.
Haemoglobin was measured by capillary HemoCue, which is noisier than venous
laboratory measurement, and the altitude adjustment is a formula rather than a
measured correction, so residual altitude confounding may persist inside the
highlands category. The linear age term understates the curvature visible in
Figure 1. Finally, wealth, region, education and ethnicity are strongly
intertwined in Peru, and although VIFs are acceptable, the mutually adjusted
coefficients should be read as "holding the others fixed", a contrast that is
partly hypothetical for combinations that barely exist (there are almost no
richest-quintile jungle households).

**Future work.** Refit with the DHS survey design and a spline for age, and add
maternal haemoglobin and supplementation history if the fuller ENDES recode file
can be obtained.

# Feasibility notes

*Not part of a student report.*

- **Works well.** This is the cleanest of the five outcomes in this dataset.
  Complete data on outcome and all six predictors (`r n_dropped` rows dropped),
  a symmetric continuous outcome, no collinearity problems (max VIF
  `r max_vif`), and a genuinely interesting Purpose 3 story with confounding that
  actually moves estimates.
- **Deterministic-twin trap.** `anaemia` is exactly `haemoglobin_g_dl < 11` and
  must never appear as a predictor here. `stunting`/`height_for_age_z` are also
  co-outcomes of undernutrition rather than exposures and were excluded on
  those grounds, not for statistical reasons.
- **Low R-squared.** Adjusted R-squared is only
  `r round(gl$adj.r.squared, 3)`. Students who expect a "good model" will be
  disappointed; this needs framing in advance as normal for individual-level
  biomarker data.
- **Non-linear age.** A single linear term for `child_age_months` visibly
  misfits (steep rise then plateau). This is a *teaching opportunity* rather
  than a defect, but a student who never plots the smoother will miss it.
- **Sample size.** At `r nrow(dat_model)` rows everything is significant. The
  brief must force effect-size-led interpretation or students will simply list
  stars.
- **Design not honoured.** Only the weight is present, not cluster or stratum,
  so a correct survey analysis is impossible from this extract. This is fine for
  a course exercise provided the limitation is stated.
