---
title: "Height-for-age Z-score in Peruvian children aged 6-59 months"
subtitle: "Dataset 5: Peru DHS child anaemia and growth extract"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "peru_child_anaemia.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "height_for_age_z"
predictors <- c("child_age_months", "child_sex", "birth_weight",
                "mother_education", "wealth_index", "natural_region",
                "mother_indigenous")

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    child_sex = factor(child_sex, levels = c("female", "male"),
                       labels = c("Female", "Male")),
    # Normal birth weight is the clinically unexposed group and by far the
    # largest, so both low birth weight and macrosomia read against it.
    birth_weight = factor(birth_weight,
                          levels = c("normal", "low", "macrosomia"),
                          labels = c("Normal", "Low (<2.5 kg)", "Macrosomia")),
    # Lowest schooling is the reference so higher categories read as "more
    # education than the reference".
    mother_education = factor(mother_education,
                              levels = c("primary_or_none", "secondary", "higher"),
                              labels = c("Primary or none", "Secondary", "Higher")),
    # Poorest is both the largest quintile in this extract and the group with the
    # greatest expected burden, so every other quintile is read against it.
    wealth_index = factor(wealth_index,
                          levels = c("poorest", "poorer", "middle",
                                     "richer", "richest"),
                          labels = c("Poorest", "Poorer", "Middle",
                                     "Richer", "Richest")),
    # Lima metropolitan is the best-resourced, lowest-altitude region and is the
    # natural comparator for the three other regions.
    natural_region = factor(natural_region,
                            levels = c("lima_metropolitan", "rest_of_coast",
                                       "highlands", "jungle"),
                            labels = c("Lima metropolitan", "Rest of coast",
                                       "Highlands", "Jungle")),
    mother_indigenous = factor(mother_indigenous, levels = c(0, 1),
                               labels = c("No", "Yes"))
  ) |>
  labelled::set_variable_labels(
    height_for_age_z  = "Height-for-age Z-score",
    child_age_months  = "Child age (months)",
    child_sex         = "Child sex",
    birth_weight      = "Birth-weight category (maternal report)",
    mother_education  = "Mother's education",
    wealth_index      = "Household wealth quintile",
    natural_region    = "Natural region",
    mother_indigenous = "Mother indigenous"
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)
```

# Introduction

Chronic undernutrition, measured as height-for-age relative to the WHO growth
standard, remains a major child-health problem in Peru despite two decades of
economic growth and targeted nutrition programmes. The Peruvian Demographic and
Health Survey (ENDES) anthropometry module measures every surveyed child aged
6-59 months, giving a large cross-sectional picture of how linear growth varies
with birth history, maternal circumstances, household wealth and geography.
Height-for-age Z-score (HAZ) is the continuous form of the indicator that is
usually dichotomised into "stunting"; analysing it continuously uses the whole
distribution rather than only the tail below -2.

**Aim.** Among Peruvian children aged 6-59 months, we examined which of child
age, child sex, birth-weight category, maternal education, household wealth,
natural region and maternal indigenous identity are independently associated
with height-for-age Z-score, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** A cross-sectional extract of `r nrow(dat_raw)` children aged
6-59 months from the Peru Demographic and Health Survey (ENDES), a nationally
implemented household survey. Each row is one child with measured length or
height, linked to the mother's interview and household characteristics.

**Variables.** The outcome is `height_for_age_z`, the WHO height-for-age
Z-score (dimensionless; 0 is the reference-population median, and one unit is
one reference standard deviation). Predictors are child age in months
(continuous, 6-59); child sex (reference *Female*); maternally reported
birth-weight category (reference *Normal*, versus *Low* i.e. below 2.5 kg and
*Macrosomia*); mother's highest education (reference *Primary or none*);
household wealth quintile (reference *Poorest*); natural region (reference
*Lima metropolitan*); and maternal indigenous identity (reference *No*). These
seven span birth endowment, maternal human capital, household resources and
geography, the four channels through which linear growth is conventionally
modelled.

`stunting` is excluded as a predictor because it is a deterministic function of
this outcome (`stunting = height_for_age_z < -2`); haemoglobin and anaemia are
also excluded because they are co-outcomes of undernutrition rather than
exposures.

**Statistical analysis.** Linear regression via `lm()`, justified because the
outcome is continuous and approximately normally distributed. We fitted a
univariable model for each predictor first, then a single multivariable model
containing all of them. Predictors were chosen in advance on subject-matter
grounds, not by significance. Significance threshold 0.05; estimates are
reported with 95% confidence intervals. Assumptions were checked with
`performance::check_model()`. Analysis used `r R.version.string` with tidyverse,
gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped,
`r round(100 * n_dropped / nrow(dat), 1)`%). Essentially all of the loss comes
from missing maternally reported birth weight, which is missing for children
whose mothers could not recall or did not report a size at birth.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(
    missing = "no"
  ) |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

```{r table-one-numbers}
haz_med <- round(median(dat_model$height_for_age_z), 2)
haz_q   <- round(quantile(dat_model$height_for_age_z, c(.25, .75)), 2)
pct_stunt <- round(100 * mean(dat_model$height_for_age_z < -2), 1)
pct_low   <- round(100 * mean(dat_model$birth_weight == "Low (<2.5 kg)"), 1)
```

Median HAZ was `r haz_med` (IQR `r haz_q[1]` to `r haz_q[2]`): the whole
distribution is shifted roughly three quarters of a standard deviation below the
WHO reference median, and `r pct_stunt`% of children fall below the -2 stunting
threshold. `r pct_low`% of children were reported as low birth weight, close to
the figure usually cited for Peru.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Height-for-age Z-score against child age, by household wealth quintile, with LOESS smoothers."}
ggplot(dat_model, aes(child_age_months, height_for_age_z, colour = wealth_index)) +
  geom_point(alpha = 0.04, size = 0.5) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 1) +
  geom_hline(yintercept = -2, linetype = "dashed") +
  scale_colour_viridis_d(option = "C", end = 0.9) +
  labs(x = "Child age (months)", y = "Height-for-age Z-score",
       colour = "Wealth quintile",
       caption = "Dashed line: -2 stunting threshold")
```

The wealth quintiles are separated by about one full Z-score across the whole
age range and the ordering never crosses, but the age pattern is *not*
monotonic: HAZ dips through the second year of life and partially recovers
afterwards, so a straight line in age will describe this poorly.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv
```

```{r uv-numbers}
uv_fit <- function(v) tidy(lm(reformulate(v, outcome), data = dat_model), conf.int = TRUE)
u_low   <- uv_fit("birth_weight")   |> filter(term == "birth_weightLow (<2.5 kg)")
u_rich  <- uv_fit("wealth_index")   |> filter(term == "wealth_indexRichest")
u_indig <- uv_fit("mother_indigenous") |> filter(term == "mother_indigenousYes")
u_high  <- uv_fit("natural_region") |> filter(term == "natural_regionHighlands")
u_coast <- uv_fit("natural_region") |> filter(term == "natural_regionRest of coast")
u_edu   <- uv_fit("mother_education") |> filter(term == "mother_educationHigher")
u_age   <- uv_fit("child_age_months") |> filter(term == "child_age_months")
u_sex   <- uv_fit("child_sex") |> filter(term == "child_sexMale")
```

Low birth weight is the strongest single child-level predictor: children
reported as weighing under 2.5 kg at birth average
`r round(u_low$estimate, 2)` Z-scores lower than normal-birth-weight children
(95% CI `r round(u_low$conf.low, 2)` to `r round(u_low$conf.high, 2)`,
p < 0.001), a deficit of more than half a reference standard deviation that
persists years after birth. The socio-economic gradients are larger still --
richest versus poorest quintile is `r round(u_rich$estimate, 2)` Z-scores -- and
child sex is the only predictor that is *not* significant even crudely
(`r round(u_sex$estimate, 3)`, p = `r round(u_sex$p.value, 3)`), which is worth
noting in a dataset this large.

## Why adjustment is needed

The crude estimates confuse several strongly overlapping gradients.
**Household wealth** confounds the maternal-indigenous contrast: indigenous
mothers in this extract are heavily concentrated in the poorest quintiles, and
poverty independently limits dietary quality, sanitation and health-service
access, so a crude indigenous-versus-not comparison is largely a
poor-versus-rich comparison. **Maternal education** confounds the wealth
gradient in turn: better-educated mothers live in wealthier households *and*
independently practise better infant feeding and seek care earlier, so part of
the raw wealth effect belongs to schooling. **Natural region** confounds both:
the highlands and jungle are simultaneously poorer, more indigenous and less
schooled than Lima, and also differ in altitude, food systems and service
coverage. Finally **birth weight** is itself socially patterned, so the crude
birth-weight coefficient partly reflects the circumstances that produced the low
birth weight rather than the birth weight itself. Mutual adjustment is required
before any of these can be read as an independent association.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv
```

```{r fit-stats}
gl <- glance(fit)
tibble(
  Statistic = c("R-squared", "Adjusted R-squared", "Residual SD (Z-scores)",
                "Observations", "Model parameters"),
  Value = c(round(gl$r.squared, 3), round(gl$adj.r.squared, 3),
            round(sigma(fit), 3), nobs(fit), length(coef(fit)))
) |>
  gt() |>
  tab_header(title = "Model fit")
```

```{r mv-numbers}
tid <- tidy(fit, conf.int = TRUE)
g <- function(t) tid |> filter(term == t)
mv_age   <- g("child_age_months")
mv_sex   <- g("child_sexMale")
mv_low   <- g("birth_weightLow (<2.5 kg)")
mv_macro <- g("birth_weightMacrosomia")
mv_rich  <- g("wealth_indexRichest")
mv_high  <- g("natural_regionHighlands")
mv_coast <- g("natural_regionRest of coast")
mv_indig <- g("mother_indigenousYes")
mv_edu   <- g("mother_educationHigher")
```

The model explains `r round(100 * gl$r.squared, 1)`% of the variance in HAZ
(adjusted R-squared `r round(gl$adj.r.squared, 3)`), which is respectable for
individual-level anthropometry but still leaves four fifths of the variation
unexplained. Holding the other six predictors constant, low birth weight is
associated with a `r round(mv_low$estimate, 2)` Z-score deficit (95% CI
`r round(mv_low$conf.low, 2)` to `r round(mv_low$conf.high, 2)`) and macrosomia
with a `r round(mv_macro$estimate, 2)` Z-score advantage (95% CI
`r round(mv_macro$conf.low, 2)` to `r round(mv_macro$conf.high, 2)`), a spread
of about one Z-score across the birth-weight categories. The wealth gradient
remains monotonic and substantial, `r round(mv_rich$estimate, 2)` Z-scores from
poorest to richest (95% CI `r round(mv_rich$conf.low, 2)` to
`r round(mv_rich$conf.high, 2)`). Children in the highlands remain
`r round(mv_high$estimate, 2)` Z-scores below Lima children of the same wealth,
education, ethnicity and birth weight. Child age contributes almost nothing on a
linear scale (`r round(mv_age$estimate, 4)` Z-scores per month) and child sex is
not significant (p = `r round(mv_sex$p.value, 2)`).

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

The estimates that move most are the social ones. **Maternal indigenous
identity** collapses from `r round(u_indig$estimate, 2)` Z-scores crude to
`r round(mv_indig$estimate, 2)` adjusted -- roughly
`r round(100 * (1 - mv_indig$estimate / u_indig$estimate))`% of the crude
association was confounding by wealth, education and region, and only a modest
independent deficit survives. **Higher maternal education** halves, from
`r round(u_edu$estimate, 2)` to `r round(mv_edu$estimate, 2)`, and the
**richest-versus-poorest** wealth contrast falls from
`r round(u_rich$estimate, 2)` to `r round(mv_rich$estimate, 2)`; both move
toward the null because these three social variables are near-substitutes for
one another. The **highlands** contrast also more than halves, from
`r round(u_high$estimate, 2)` to `r round(mv_high$estimate, 2)`. By contrast
**low birth weight** barely moves (`r round(u_low$estimate, 2)` to
`r round(mv_low$estimate, 2)`), so its association with later linear growth is
largely independent of the household circumstances measured here. One estimate
changes significance and sign: **rest of coast versus Lima** is
`r round(u_coast$estimate, 3)` crude (p = `r round(u_coast$p.value, 4)`) but
`r round(mv_coast$estimate, 3)` adjusted (p = `r round(mv_coast$p.value, 2)`) --
once wealth and education are held constant, coastal children outside Lima are
indistinguishable from Lima children. Child sex is non-significant in both
models. The pattern throughout is attenuation toward the null, the signature of
mutual confounding among correlated social exposures rather than of mediation.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r diag-numbers}
vifs <- check_collinearity(fit)
max_vif <- round(max(vifs$VIF), 2)
```

The **posterior-predictive check** shows the model reproducing the shape and
location of the observed HAZ distribution closely, which is expected given that
Z-scores are constructed to be roughly normal. **Linearity** and **homogeneity
of variance** panels are close to flat with no fanning, so constant residual
variance is defensible. **Normality of residuals** is good in the body of the
distribution with slight heaviness in the lower tail (a handful of children with
Z-scores below -5, which are biologically implausible and are likely
measurement or age-reporting errors rather than genuine values). The
**collinearity** panel is entirely in the low range (maximum VIF `r max_vif`)
despite the strong pairwise associations between wealth, education, region and
ethnicity. **Influential observations** show no point beyond the Cook's distance
contours. The one assumption that is genuinely violated is not shown in these
panels: Figure 1 makes clear that HAZ falls and then recovers across the age
range, so the linear age term is mis-specified; a quadratic or spline should be
used instead.

# Discussion

Linear growth in Peruvian preschool children tracks household circumstances more
than anything else measured here: moving from the poorest to the richest
quintile is associated with about `r round(mv_rich$estimate, 2)` Z-scores of
height-for-age, and living in the highlands with about
`r abs(round(mv_high$estimate, 2))` Z-scores less than Lima, after mutual
adjustment. Birth endowment matters too and matters independently: the
`r round(mv_low$estimate, 2)` Z-score deficit for low birth weight is barely
touched by adjustment, so children who start small stay small regardless of the
household they grow up in. Practically, that argues for both antenatal
interventions that raise birth weight and continued postnatal nutrition support
in highland and poorest-quintile communities.

**Limitations.** This is cross-sectional: HAZ is measured once, birth weight is
recalled, and no causal interpretation is warranted. Birth weight is a
**maternally reported three-level category**, not a measured weight in
kilograms, so it is subject to recall and heaping bias and the "macrosomia"
group in particular may capture perception of a large baby rather than a
measured >4 kg weight. The source ENDES analysis used the DHS complex design --
cluster (HV001), stratum (HV022) and sampling weights -- whereas this extract
carries only the weight variable, which is deliberately not used; standard errors
are therefore too narrow, estimates are not design-corrected, and **this sample
must not be described as nationally representative**. With `r nrow(dat_model)`
rows almost every comparison reaches statistical significance, so interpretation
must lead with effect sizes rather than p-values. Key confounders are absent
entirely: maternal height (the single strongest determinant of child height),
birth order and interval, infection history, and dietary diversity. The linear
age term is mis-specified. A small number of extreme Z-scores below -5 are
probably data errors and were retained rather than trimmed. Finally, the
`r n_dropped` rows dropped for missing birth weight are unlikely to be missing
completely at random, since recall of birth size is worse among less-educated
mothers, so the complete-case sample is mildly better-off than the full extract.

**Future work.** Refit with the DHS survey design, a spline for age, and
maternal height and birth order from the fuller ENDES recode file.

# Feasibility notes

*Not part of a student report.*

- **Works well.** Probably the best of the five outcomes for a student. Large
  effect sizes that are easy to talk about, a genuinely instructive confounding
  story (indigenous identity attenuating by
  `r round(100 * (1 - mv_indig$estimate / u_indig$estimate))`%), no collinearity
  problems (max VIF `r max_vif`), and adjusted R-squared
  `r round(gl$adj.r.squared, 3)`, the highest of the five models.
- **Deterministic-twin trap.** `stunting` is exactly `height_for_age_z < -2`.
  It must never enter this model. A student who runs "all variables" without
  thinking will get an R-squared near 0.7 and a nonsense model.
- **Non-linear age.** The age-HAZ relationship is U-shaped (nadir around
  18 months), so the linear term is close to useless
  (`r round(mv_age$estimate, 4)` per month). Fine as a teaching point, but
  students must be told to plot before modelling.
- **Birth weight is categorical, not kg.** The codebook wording invites students
  to treat `birth_weight` as continuous. It is a three-level string, and
  "macrosomia" is *not* an ordered extension of "low"-"normal" in the
  risk sense for this outcome (both tails are informative, in opposite
  directions).
- **Implausible values.** A few HAZ values below -5 and above +3.5 survive in
  the file. They do not break the model but they should be discussed.
- **Missingness.** `r n_dropped` rows (`r round(100 * n_dropped / nrow(dat), 1)`%)
  are lost to missing birth weight. Not a threat, but it makes this model's N
  differ from the haemoglobin model's, which confuses students who compare
  tables across outcomes.
- **Design not honoured.** Only the weight is present, not cluster or stratum,
  so a design-correct analysis is impossible here.
