---
title: "Gas transfer (DLCO % predicted) in adults with and without HIV"
subtitle: "Dataset 11: HIV infection and exercise capacity"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "hiv_lung_6mwt.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names. dlco_pct_predicted is
# the outcome and is therefore removed from the predictor set.
outcome    <- "dlco_pct_predicted"
predictors <- c("hiv_positive", "age_yrs", "sex", "current_smoker", "pack_years",
                "systolic_bp", "diastolic_bp", "haemoglobin_g_dl",
                "fev1_pct_predicted", "fvc_pct_predicted")

lab <- function(x, l) { attr(x, "label") <- l; x }

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    hiv_positive = lab(factor(hiv_positive, levels = c(0, 1), labels = c("No", "Yes")),
                       "Living with HIV"),
    sex = lab(factor(sex, levels = c("male", "female"), labels = c("Male", "Female")),
              "Sex"),
    # Smoking is parameterised as cumulative dose plus a current-smoking indicator;
    # see Methods for why the three-level factor is not used as-is.
    current_smoker = lab(factor(if_else(smoking_status == "current", "Yes", "No"),
                                levels = c("No", "Yes")),
                         "Current smoker"),
    smoking_status = lab(factor(smoking_status, levels = c("never", "former", "current"),
                                labels = c("Never", "Former", "Current")),
                         "Smoking status"),
    age_yrs = lab(age_yrs, "Age (years)"),
    pack_years = lab(pack_years, "Smoking (pack-years)"),
    systolic_bp = lab(systolic_bp, "Systolic BP (mmHg)"),
    diastolic_bp = lab(diastolic_bp, "Diastolic BP (mmHg)"),
    haemoglobin_g_dl = lab(haemoglobin_g_dl, "Haemoglobin (g/dL)"),
    fev1_pct_predicted = lab(fev1_pct_predicted, "FEV1 (% predicted)"),
    fvc_pct_predicted  = lab(fvc_pct_predicted, "FVC (% predicted)"),
    dlco_pct_predicted = lab(dlco_pct_predicted, "DLCO (% predicted)")
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
pv  <- function(p) ifelse(p < 0.001, "<0.001", paste0("= ", fmt(p, 3)))
```

# Introduction

The Pittsburgh/UCSF lung-function study enrolled adults living with HIV and
HIV-uninfected comparators from two US clinical centres and assessed all of them
identically with a six-minute walk test, spirometry, diffusing capacity and the
St George's Respiratory Questionnaire (Fitzpatrick and colleagues, *PLOS ONE*
2019). This teaching extract holds `r nrow(dat)` participants,
`r sum(dat$hiv_positive == "Yes")` of them living with HIV. The diffusing
capacity of the lung for carbon monoxide, expressed as a percentage of the value
predicted for a person of the same age, sex and height, measures how efficiently
gas crosses from alveolus to blood; it is abnormal in emphysema, interstitial
disease and pulmonary vascular disease, and in the source paper it is the
lung-function measurement most strongly related to how far participants could
walk.

**Aim.** Among these adults, we examined which of HIV status, age, sex, current
smoking, cumulative pack-years, blood pressure, haemoglobin and the two
spirometric volumes (FEV1 and FVC, each as a percentage of predicted) are
independently associated with DLCO % predicted, mutually adjusting for the
others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** A cross-sectional analysis of participants enrolled at two US
clinical centres (Pittsburgh and San Francisco) into a cohort study of lung
function in HIV. The extract contains `r nrow(dat)` adults aged
`r min(dat$age_yrs)` to `r max(dat$age_yrs)` and 21 variables;
`r sum(dat$hiv_positive == "Yes")` (`r fmt(100 * mean(dat$hiv_positive == "Yes"), 1)`%)
are living with HIV. All measurements were made at a single study visit.

**Variables.** The outcome is `dlco_pct_predicted`, single-breath diffusing
capacity for carbon monoxide as a percentage of the predicted value (observed
range `r fmt(min(dat$dlco_pct_predicted, na.rm = TRUE), 1)` to
`r fmt(max(dat$dlco_pct_predicted, na.rm = TRUE), 1)`), treated as continuous.
Predictors are HIV status (reference HIV-negative, the clinically unexposed
group), age (years), sex (reference male, the larger group), current smoking
(reference not a current smoker), cumulative smoking exposure (pack-years),
systolic and diastolic blood pressure (mmHg), haemoglobin (g/dL), and FEV1 and
FVC each as a percentage of predicted. DLCO itself is of course removed from the
predictor set.

**Both spirometric volumes are retained, and the reason is worth stating.** FEV1
and FVC as percentages of predicted correlate at
`r fmt(cor(dat$fev1_pct_predicted, dat$fvc_pct_predicted, use = "complete.obs"), 2)`
in this file, which is high enough to inflate their standard errors. They are
nonetheless kept together here because, unlike in the model where FEV1 is the
outcome, neither is a re-reading of the outcome: DLCO comes from a separate
single-breath manoeuvre. Keeping both lets the model distinguish restriction
(low FVC) from obstruction (low FEV1 relative to FVC), which is exactly the
distinction that matters for gas transfer. The cost is reported explicitly in
the Diagnostics section as a variance inflation factor.

**Smoking is entered once, as dose plus a current-smoking indicator.** The
shipped `smoking_status` variable has three levels (never
`r sum(dat$smoking_status == "Never")`, former
`r sum(dat$smoking_status == "Former")`, current
`r sum(dat$smoking_status == "Current")`) and `pack_years` is **exactly zero for
all `r sum(dat$pack_years == 0)` never-smokers and strictly positive for every
ever-smoker**, so the two encode the same never/ever distinction twice. Entering
the untouched factor alongside pack-years would force the former- and
current-smoker contrasts to be estimated at zero pack-years, a value no
ever-smoker in this sample has. We therefore keep `pack_years` as the
cumulative-dose term and add one binary indicator for *current* smoking. That
distinction is more than cosmetic for this outcome: current smoking raises
carboxyhaemoglobin, which lowers measured DLCO on the day of the test
independently of any lifetime damage.

**Variables deliberately excluded.** `cd4_count`, `on_antiretroviral` and
`viral_load_detectable` are **missing by design** for participants without HIV;
including any of them deletes most or all of the HIV-negative comparison group.
`mmrc_dyspnoea` is excluded because all `r sum(is.na(dat$mmrc_dyspnoea))` of its
missing values fall in HIV-positive participants, and because breathlessness is
downstream of impaired gas transfer. `fev1_fvc_ratio` is excluded because it is
very largely a function of the two %-predicted values already in the model
(regressing it on them gives R-squared
`r fmt(summary(lm(fev1_fvc_ratio ~ fev1_pct_predicted + fvc_pct_predicted, data = dat))$r.squared, 2)`).
The SGRQ scores and `six_min_walk_m` are consequences of lung function and are
analysed as separate outcomes.

**Statistical analysis.** `lm()`, justified because the outcome is a continuous,
approximately symmetric percentage (skewness
`r fmt(mean((dat_model$dlco_pct_predicted - mean(dat_model$dlco_pct_predicted))^3) / sd(dat_model$dlco_pct_predicted)^3, 2)`).
We fitted a univariable model for each predictor first, then a single
multivariable model containing all of them. Predictors were chosen in advance on
subject-matter grounds, not by significance. Significance threshold 0.05;
estimates are reported with 95% confidence intervals. Assumptions were checked
with `performance::check_model()`, and `check_collinearity()` was inspected
specifically because of the FEV1/FVC pair. Analysis used `r R.version.string`
with tidyverse, gtsummary and performance.

**Parameter budget.** The model estimates
`r length(coef(lm(reformulate(predictors, outcome), data = dat_model))) - 1L`
parameters on `r nrow(dat_model)` complete cases, about
`r fmt(nrow(dat_model) / (length(coef(lm(reformulate(predictors, outcome), data = dat_model))) - 1L), 0)`
observations per parameter, well inside the 10-15 rule.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped). The outcome itself is
`r fmt(100 * mean(!is.na(dat$dlco_pct_predicted)), 1)`% complete
(`r sum(is.na(dat$dlco_pct_predicted))` missing); the loss is mostly haemoglobin
(`r sum(is.na(dat$haemoglobin_g_dl))` missing), spirometry
(`r sum(is.na(dat$fev1_pct_predicted))`) and blood pressure
(`r sum(is.na(dat$systolic_bp))`). The analytic sample keeps
`r sum(dat_model$hiv_positive == "Yes")` participants with HIV and
`r sum(dat_model$hiv_positive == "No")` without.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(missing = "no") |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics (complete-case analytic sample).**")
```

Median DLCO is `r fmt(median(dat_model$dlco_pct_predicted), 0)`% of predicted and
`r sum(dat_model$dlco_pct_predicted < 80)`
(`r fmt(100 * mean(dat_model$dlco_pct_predicted < 80), 0)`%) of participants fall
below the conventional 80% lower limit, so impaired gas transfer is the modal
abnormality in this sample even though median FEV1
(`r fmt(median(dat_model$fev1_pct_predicted), 0)`% of predicted) and median FVC
(`r fmt(median(dat_model$fvc_pct_predicted), 0)`%) are essentially normal.

## Exploratory figure

```{r explore, fig.cap="Figure 1. DLCO (% predicted) against cumulative smoking exposure, by HIV status."}
ggplot(dat_model, aes(pack_years, dlco_pct_predicted, colour = hiv_positive)) +
  geom_point(alpha = 0.5, size = 1.6) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.9) +
  scale_colour_manual(values = c(No = "#3F6F9F", Yes = "#B5495B")) +
  labs(x = "Cumulative smoking exposure (pack-years)", y = "DLCO (% predicted)",
       colour = "Living with HIV")
```

Gas transfer falls as lifetime tobacco exposure rises in both groups, and the
HIV-positive line sits below the HIV-negative one across the whole range, so the
two exposures appear to act additively rather than one explaining the other.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv

utid <- map_dfr(predictors, function(v) {
  tidy(lm(reformulate(v, outcome), data = dat_model), conf.int = TRUE) |>
    filter(term != "(Intercept)")
})
ug <- function(term, d = 2) {
  r <- utid[utid$term == term, ]
  sprintf("%s percentage points (95%% CI %s to %s)", fmt(r$estimate, d),
          fmt(r$conf.low, d), fmt(r$conf.high, d))
}
up <- function(term) pv(utid$p.value[utid$term == term])
```

Crudely, participants living with HIV have lower diffusing capacity,
`r ug("hiv_positiveYes")` of predicted (p `r up("hiv_positiveYes")`). The
spirometric volumes track the outcome closely — `r ug("fev1_pct_predicted")` per
percentage point of predicted FEV1 (p `r up("fev1_pct_predicted")`) and
`r ug("fvc_pct_predicted")` per percentage point of predicted FVC
(p `r up("fvc_pct_predicted")`) — as does smoking, with each pack-year
associated with `r ug("pack_years")` (p `r up("pack_years")`) and current
smokers `r ug("current_smokerYes")` below non-current smokers
(p `r up("current_smokerYes")`). Each additional year of age is associated with
`r ug("age_yrs")` (p `r up("age_yrs")`) and women average `r ug("sexFemale")`
above men (p `r up("sexFemale")`). **Haemoglobin shows no crude association
whatever**, `r ug("haemoglobin_g_dl")` (p `r up("haemoglobin_g_dl")`); this turns
out to be the most interesting number in the table. Neither blood pressure is
crudely associated.

## Why adjustment is needed

**Cumulative smoking** is the primary confounder of the HIV association.
Participants with HIV have smoked more (mean
`r fmt(mean(dat$pack_years[dat$hiv_positive == "Yes"]), 1)` pack-years versus
`r fmt(mean(dat$pack_years[dat$hiv_positive == "No"]), 1)` in the HIV-negative
group), and pack-years is one of the strongest crude predictors of DLCO, so part
of the crude HIV deficit is a tobacco deficit and the whole question is how much
of it survives. **Sex** is the second, and it is the variable that makes this
model worth fitting. Men are over-represented in the HIV-positive group
(`r fmt(100 * mean(dat$sex[dat$hiv_positive == "Yes"] == "Male"), 0)`% versus
`r fmt(100 * mean(dat$sex[dat$hiv_positive == "No"] == "Male"), 0)`%), men have
markedly higher haemoglobin (mean
`r fmt(mean(dat$haemoglobin_g_dl[dat$sex == "Male"], na.rm = TRUE), 1)` versus
`r fmt(mean(dat$haemoglobin_g_dl[dat$sex == "Female"], na.rm = TRUE), 1)` g/dL)
and men in this sample have distinctly lower DLCO % predicted (mean
`r fmt(mean(dat$dlco_pct_predicted[dat$sex == "Male"], na.rm = TRUE), 0)`% versus
`r fmt(mean(dat$dlco_pct_predicted[dat$sex == "Female"], na.rm = TRUE), 0)`%).
Because higher haemoglobin should physiologically *raise* measured DLCO while
being commoner in the sex with *lower* DLCO, sex masks the haemoglobin
association almost perfectly in the crude analysis. Mutual adjustment lets each
estimate be read as a comparison between participants alike on the other
measured characteristics.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv

tid <- tidy(fit, conf.int = TRUE)
g <- function(term, d = 2) {
  r <- tid[tid$term == term, ]
  sprintf("%s percentage points (95%% CI %s to %s)", fmt(r$estimate, d),
          fmt(r$conf.low, d), fmt(r$conf.high, d))
}
gp <- function(term) pv(tid$p.value[tid$term == term])
```

```{r fit-stats}
s <- summary(fit)
tibble(
  Statistic = c("Complete-case N", "Parameters (excl. intercept)",
                "Observations per parameter",
                "R-squared", "Adjusted R-squared", "Residual SE (% predicted)"),
  Value = c(as.character(nrow(dat_model)),
            as.character(length(coef(fit)) - 1L),
            fmt(nrow(dat_model) / (length(coef(fit)) - 1L), 1),
            fmt(s$r.squared, 3), fmt(s$adj.r.squared, 3), fmt(s$sigma, 1))
) |>
  gt() |>
  tab_header(title = md("**Model fit: multivariable linear model for DLCO (% predicted)**"))
```

This is the best-fitting of the four models in this dataset: it explains
`r fmt(100 * s$r.squared, 1)`% of the variance in DLCO % predicted (adjusted
R-squared `r fmt(s$adj.r.squared, 3)`), with a residual standard deviation of
`r fmt(s$sigma, 1)` percentage points. Among participants of the same age, sex,
smoking history, blood pressure, haemoglobin and spirometry, those living with
HIV have lower diffusing capacity, `r g("hiv_positiveYes")` of predicted
(p `r gp("hiv_positiveYes")`), i.e. about
`r fmt(abs(tid$estimate[tid$term == "hiv_positiveYes"]), 1)` percentage points
lower. Among participants alike on everything else, each
additional year of age is associated with `r g("age_yrs")`
(p `r gp("age_yrs")`), each pack-year with `r g("pack_years")`
(p `r gp("pack_years")`), and current smoking with `r g("current_smokerYes")`
(p `r gp("current_smokerYes")`) — so both the acute and the cumulative smoking
terms are independently associated with gas transfer, which was not true for any
other outcome in this dataset. Women average `r g("sexFemale")` above men
(p `r gp("sexFemale")`), and each g/dL of haemoglobin is associated with
`r g("haemoglobin_g_dl")` (p `r gp("haemoglobin_g_dl")`). FEV1 contributes
`r g("fev1_pct_predicted")` (p `r gp("fev1_pct_predicted")`) and FVC
`r g("fvc_pct_predicted", 3)` (p `r gp("fvc_pct_predicted")`), just short of
conventional significance. Systolic pressure is
weakly associated (`r g("systolic_bp")`, p `r gp("systolic_bp")`) and diastolic
pressure is not. These are associations among the participants observed, not
effects of intervening.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

**Haemoglobin is the striking one.** Crudely it has no association at all with
diffusing capacity — `r ug("haemoglobin_g_dl")`, p `r up("haemoglobin_g_dl")`,
an estimate indistinguishable from zero — yet adjusted it is one of the
strongest terms in the model, `r g("haemoglobin_g_dl")`,
p `r gp("haemoglobin_g_dl")`. This is textbook masking rather than confounding
in the usual direction: haemoglobin genuinely raises measured DLCO (more
haemoglobin in the capillary bed means more carbon monoxide taken up), but the
people with high haemoglobin here are men, and men have lower DLCO % predicted
for other reasons. The two influences cancelled almost exactly in the crude
estimate. It also tells us something about the data: **the DLCO values in this
file are evidently *not* haemoglobin-corrected**, since a properly corrected DLCO
would show no residual haemoglobin association.

Two predictors attenuate substantially but survive. **HIV status** moves from
`r ug("hiv_positiveYes")` crude to `r g("hiv_positiveYes")` adjusted, so roughly
`r fmt(100 * (1 - tid$estimate[tid$term == "hiv_positiveYes"] / utid$estimate[utid$term == "hiv_positiveYes"]), 0)`%
of the crude deficit is explained by the other measured characteristics — mostly
smoking — but a deficit of about
`r fmt(abs(tid$estimate[tid$term == "hiv_positiveYes"]), 0)` percentage points
remains and stays significant. **Pack-years** moves from `r ug("pack_years")` to
`r g("pack_years")` for the same reason in reverse, sharing its crude effect with
current smoking and the spirometric volumes. **FVC** moves the furthest of any
term, from `r ug("fvc_pct_predicted")` crude (p `r up("fvc_pct_predicted")`) to
`r g("fvc_pct_predicted", 3)` adjusted (p `r gp("fvc_pct_predicted")`), losing
significance; this is collinearity with FEV1, which itself falls from
`r ug("fev1_pct_predicted")` to `r g("fev1_pct_predicted")`.

**Sex** is the one estimate that does not move at all, from `r ug("sexFemale")`
crude to `r g("sexFemale")` adjusted — the female advantage in gas transfer is
not explained by any other measured characteristic. **Age** attenuates modestly,
from `r ug("age_yrs")` to `r g("age_yrs")`, and stays clearly significant. One
estimate crosses
into significance: **systolic blood pressure**, from `r ug("systolic_bp")`
(p `r up("systolic_bp")`) to `r g("systolic_bp")` (p `r gp("systolic_bp")`),
which appears only once diastolic pressure is held constant and should be read as
an artefact of adjusting one blood-pressure component for the other rather than a
finding about the pulmonary circulation.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r collin}
check_collinearity(fit)
```

```{r assumption-tests}
print(check_heteroscedasticity(fit))
print(check_normality(fit))
```

The diagnostics are satisfactory. The residual-versus-fitted panel is flat and
`check_heteroscedasticity()` does not reject constant variance
(p `r pv(as.numeric(check_heteroscedasticity(fit)))`); the posterior-predictive
check reproduces the observed distribution well, which it should, since this is
the best-fitting model of the four. The normal-QQ panel shows a modest left tail
from a handful of participants with severely impaired gas transfer, and
`check_normality()` rejects at the 0.05 level
(p `r pv(as.numeric(check_normality(fit)))`); with `r nrow(dat_model)`
observations this has negligible practical consequence and the outcome is close
to symmetric, so no transformation is indicated.

Collinearity behaves exactly as anticipated in Methods. The two largest variance
inflation factors are
`r fmt(check_collinearity(fit)$VIF[check_collinearity(fit)$Term == "fev1_pct_predicted"], 2)`
for FEV1 and
`r fmt(check_collinearity(fit)$VIF[check_collinearity(fit)$Term == "fvc_pct_predicted"], 2)`
for FVC, against a maximum of
`r fmt(max(check_collinearity(fit)$VIF[!check_collinearity(fit)$Term %in% c("fev1_pct_predicted", "fvc_pct_predicted")]), 2)`
for every other term. Both are below the conventional threshold of 5, so the
model is estimable and we have kept both, but the practical cost is visible in
Table 2: the FVC standard error is
`r fmt(summary(fit)$coefficients["fvc_pct_predicted", 2], 3)` against
`r fmt(summary(lm(reformulate(setdiff(predictors, "fev1_pct_predicted"), outcome), data = dat_model))$coefficients["fvc_pct_predicted", 2], 3)`
in the same model with FEV1 removed, and the FVC coefficient loses significance
as a result. Neither spirometric coefficient should be quoted on its own. No
observation exceeds the Cook's distance contours (maximum
`r fmt(max(cooks.distance(fit)), 3)`).

# Discussion

In this two-centre US sample, adults living with HIV had diffusing capacity about
`r fmt(abs(tid$estimate[tid$term == "hiv_positiveYes"]), 1)` percentage points of
predicted lower than HIV-negative adults of the same age, sex, smoking history,
blood pressure, haemoglobin and spirometry — a little under half of the crude
difference, the rest being smoking. This is the clearest HIV-associated lung
abnormality in the dataset: HIV status is *not* associated with FEV1 in the same
participants, but it is associated with gas transfer, and gas transfer is in turn
the lung measurement most strongly related to how far they can walk. Practically,
that argues for measuring DLCO rather than spirometry alone when assessing lung
health in people living with HIV, since a normal FEV1 will not detect this.
Cumulative and current smoking each contribute independently, so both a smoking
history and current smoking status are worth recording.

**Limitations.** Participants were recruited at **two clinical centres,
Pittsburgh and UCSF, and the site variable is not in this deposit**, although the
source paper shows site matters; DLCO measurement in particular is
equipment-dependent and calibration differences between laboratories are a
plausible source of between-site variation that this model cannot address, so
site is both an unmeasured confounder and an ignored clustering variable. The
design is cross-sectional. The haemoglobin finding implies that the DLCO values
are uncorrected for haemoglobin; readers should not compare these percentages
with haemoglobin-corrected values from other cohorts. Smoking is self-reported
with no biochemical verification, so residual confounding of the HIV coefficient
by under-reported smoking is likely, and it would act in the direction of
exaggerating the HIV deficit. FEV1 and FVC compete for the same variance (VIFs
`r fmt(check_collinearity(fit)$VIF[check_collinearity(fit)$Term == "fev1_pct_predicted"], 1)`
and
`r fmt(check_collinearity(fit)$VIF[check_collinearity(fit)$Term == "fvc_pct_predicted"], 1)`),
so the FVC null should not be read as evidence that lung volume is irrelevant.
Injecting drug use, prior *Pneumocystis* pneumonia and pulmonary hypertension —
all commoner in HIV and all causes of reduced DLCO — are not measured, so the
mechanism behind the HIV association cannot be identified here. Complete-case
analysis dropped `r n_dropped` of `r nrow(dat)` rows
(`r fmt(100 * n_dropped / nrow(dat), 1)`%). All estimates are associations, not
causal effects.

**Future work.** Refit with study site as a fixed effect and with
haemoglobin-corrected DLCO, and test formally whether reduced diffusing capacity
mediates the HIV association with six-minute walk distance seen in
`six_min_walk_m.Rmd`.

# Feasibility notes

*NOT part of a student report.*

```{r feas}
cc_cd4 <- dat |> select(all_of(c(outcome, predictors, "cd4_count"))) |> drop_na()
cc_arv <- dat |> select(all_of(c(outcome, predictors, "on_antiretroviral"))) |> drop_na()
cc_mmr <- dat |> select(all_of(c(outcome, predictors, "mmrc_dyspnoea"))) |> drop_na()
fit_nofev <- lm(reformulate(setdiff(predictors, "fev1_pct_predicted"), outcome),
                data = dat_model)
```

- **USABLE, and the strongest of the four outcomes for teaching.** Complete-case
  N `r nrow(dat_model)` of `r nrow(dat)`, R-squared `r fmt(s$r.squared, 3)` (the
  highest of the four models), diagnostics pass, the exposure of interest is
  significant crudely and adjusted, and the dataset contains a beautiful worked
  example of masking (haemoglobin). Recommend it.
- **FATAL if any HIV-restricted variable is used.** `cd4_count`,
  `on_antiretroviral` and `viral_load_detectable` exist only for the
  `r sum(dat$hiv_positive == "Yes")` participants with HIV. Adding
  `on_antiretroviral` takes the complete-case N from `r nrow(dat_model)` to
  `r nrow(cc_arv)` with **`r sum(cc_arv$hiv_positive == "No")` HIV-negative
  participants remaining**, so `hiv_positive` becomes a constant and `lm()`
  refuses to fit; adding `cd4_count` takes N to `r nrow(cc_cd4)` with only
  `r sum(cc_cd4$hiv_positive == "No")` HIV-negative participants left, which is
  the dangerous case because nothing errors. `six_min_walk_m.Rmd` in this folder
  tabulates the damage in full. Given that this outcome shows the clearest HIV
  effect in the dataset, destroying the comparison here is the costliest mistake
  a student can make.
- **MAJOR: FEV1 and FVC compete.** VIFs
  `r fmt(check_collinearity(fit)$VIF[check_collinearity(fit)$Term == "fev1_pct_predicted"], 2)`
  and
  `r fmt(check_collinearity(fit)$VIF[check_collinearity(fit)$Term == "fvc_pct_predicted"], 2)`
  (r =
  `r fmt(cor(dat_model$fev1_pct_predicted, dat_model$fvc_pct_predicted), 2)`),
  against a maximum of
  `r fmt(max(check_collinearity(fit)$VIF[!check_collinearity(fit)$Term %in% c("fev1_pct_predicted", "fvc_pct_predicted")]), 2)`
  elsewhere. Below the threshold of 5, so nothing is unestimable, but FVC goes
  from p `r up("fvc_pct_predicted")` crude to p `r gp("fvc_pct_predicted")`
  adjusted purely because of it. Dropping FEV1 restores FVC to
  `r fmt(coef(fit_nofev)["fvc_pct_predicted"], 3)` with a standard error of
  `r fmt(summary(fit_nofev)$coefficients["fvc_pct_predicted", 2], 3)`. Students
  need to be told that a non-significant collinear pair is not evidence of no
  association.
- **MAJOR (interpretive, not statistical): haemoglobin is completely masked in
  the crude analysis.** Crude
  `r ug("haemoglobin_g_dl")` (p `r up("haemoglobin_g_dl")`), adjusted
  `r g("haemoglobin_g_dl")` (p `r gp("haemoglobin_g_dl")`). This is a real and
  well-understood physiological effect (uncorrected DLCO depends on haemoglobin)
  masked by sex, and it is the best teaching example in the whole dataset — but a
  student who reports only the univariable table will conclude haemoglobin is
  irrelevant, and one who reports only the multivariable table without noticing
  the crude estimate will miss why it matters.
- **MINOR: `mmrc_dyspnoea` is differentially missing** — all
  `r sum(is.na(dat$mmrc_dyspnoea))` missing values are HIV-positive participants;
  adding it would take N to `r nrow(cc_mmr)`. It is also downstream of the
  outcome.
- **MINOR: `fev1_fvc_ratio` is near-redundant** with the two %-predicted
  spirometry variables (R-squared
  `r fmt(summary(lm(fev1_fvc_ratio ~ fev1_pct_predicted + fvc_pct_predicted, data = dat))$r.squared, 2)`
  regressing it on them). Adding all three to one model is a collinearity
  disaster waiting to happen; the catalogue lists all three side by side without
  comment.
- **MINOR: `pack_years` and `smoking_status` are structurally entangled**
  (`pack_years == 0` is exactly `smoking_status == "never"` on all
  `r nrow(dat)` rows). Unusually for this dataset, *both* smoking terms are
  independently significant for this outcome
  (pack-years `r g("pack_years")`, current smoking
  `r g("current_smokerYes")`), so the parameterisation choice has real
  consequences here and must be justified.
- **MINOR: mildly non-normal residuals** (p
  `r pv(as.numeric(check_normality(fit)))`); variance is constant (p
  `r pv(as.numeric(check_heteroscedasticity(fit)))`).
- **Structural: two-site recruitment with no site variable.** More consequential
  for DLCO than for the other outcomes, because diffusing-capacity measurement is
  equipment- and calibration-dependent. Unfixable from the shipped file.
