---
title: "Airflow obstruction (FEV1 % predicted) in adults with and without HIV"
subtitle: "Dataset 11: HIV infection and exercise capacity"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "hiv_lung_6mwt.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names. fvc_pct_predicted is
# deliberately NOT a predictor - see Methods.
outcome    <- "fev1_pct_predicted"
predictors <- c("hiv_positive", "age_yrs", "sex", "current_smoker", "pack_years",
                "systolic_bp", "diastolic_bp", "haemoglobin_g_dl",
                "dlco_pct_predicted")

lab <- function(x, l) { attr(x, "label") <- l; x }

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    hiv_positive = lab(factor(hiv_positive, levels = c(0, 1), labels = c("No", "Yes")),
                       "Living with HIV"),
    sex = lab(factor(sex, levels = c("male", "female"), labels = c("Male", "Female")),
              "Sex"),
    # Smoking is parameterised as cumulative dose plus a current-smoking indicator;
    # see Methods for why the three-level factor is not used as-is.
    current_smoker = lab(factor(if_else(smoking_status == "current", "Yes", "No"),
                                levels = c("No", "Yes")),
                         "Current smoker"),
    smoking_status = lab(factor(smoking_status, levels = c("never", "former", "current"),
                                labels = c("Never", "Former", "Current")),
                         "Smoking status"),
    age_yrs = lab(age_yrs, "Age (years)"),
    pack_years = lab(pack_years, "Smoking (pack-years)"),
    systolic_bp = lab(systolic_bp, "Systolic BP (mmHg)"),
    diastolic_bp = lab(diastolic_bp, "Diastolic BP (mmHg)"),
    haemoglobin_g_dl = lab(haemoglobin_g_dl, "Haemoglobin (g/dL)"),
    fvc_pct_predicted  = lab(fvc_pct_predicted, "FVC (% predicted)"),
    dlco_pct_predicted = lab(dlco_pct_predicted, "DLCO (% predicted)"),
    fev1_pct_predicted = lab(fev1_pct_predicted, "FEV1 (% predicted)")
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
pv  <- function(p) ifelse(p < 0.001, "<0.001", paste0("= ", fmt(p, 3)))
```

# Introduction

The Pittsburgh/UCSF lung-function study enrolled adults living with HIV and
HIV-uninfected comparators from two US clinical centres and assessed all of them
identically with a six-minute walk test, spirometry, diffusing capacity and the
St George's Respiratory Questionnaire (Fitzpatrick and colleagues, *PLOS ONE*
2019). This teaching extract holds `r nrow(dat)` participants,
`r sum(dat$hiv_positive == "Yes")` of them living with HIV. FEV1 expressed as a
percentage of the value predicted for a person of the same age, sex and height is
the standard measure of airflow limitation, and an excess burden of chronic
obstructive lung disease in people living with HIV — beyond what their smoking
histories would explain — is one of the more contested claims in the HIV
comorbidity literature.

**Aim.** Among these adults, we examined which of HIV status, age, sex, current
smoking, cumulative pack-years, blood pressure, haemoglobin and diffusing
capacity are independently associated with FEV1 % predicted, mutually adjusting
for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** A cross-sectional analysis of participants enrolled at two US
clinical centres (Pittsburgh and San Francisco) into a cohort study of lung
function in HIV. The extract contains `r nrow(dat)` adults aged
`r min(dat$age_yrs)` to `r max(dat$age_yrs)` and 21 variables;
`r sum(dat$hiv_positive == "Yes")` (`r fmt(100 * mean(dat$hiv_positive == "Yes"), 1)`%)
are living with HIV. All measurements were made at a single study visit.

**Variables.** The outcome is `fev1_pct_predicted`, forced expiratory volume in
one second as a percentage of the predicted value (observed range
`r fmt(min(dat$fev1_pct_predicted, na.rm = TRUE), 1)` to
`r fmt(max(dat$fev1_pct_predicted, na.rm = TRUE), 1)`), treated as continuous.
Predictors are HIV status (reference HIV-negative, the clinically unexposed
group), age (years), sex (reference male, the larger group), current smoking
(reference not a current smoker), cumulative smoking exposure (pack-years),
systolic and diastolic blood pressure (mmHg), haemoglobin (g/dL) and DLCO as a
percentage of predicted.

**FVC is deliberately not a predictor.** `fev1_pct_predicted` and
`fvc_pct_predicted` are two numbers read off the *same* forced expiratory
manoeuvre — the volume blown out in the first second, and the total volume blown
out — and in this file they correlate at
`r fmt(cor(dat$fev1_pct_predicted, dat$fvc_pct_predicted, use = "complete.obs"), 3)`.
Putting FVC on the right-hand side does not adjust for a confounder; it asks
"given how big this person's total blow was, how big was the first second of it",
which is the definition of the FEV1/FVC ratio rather than a research question
about who has airflow obstruction. The Feasibility notes show what happens if it
is included: R-squared rises from
`r fmt(summary(lm(reformulate(predictors, outcome), data = dat_model))$r.squared, 3)`
to
`r fmt(summary(lm(reformulate(c(predictors, "fvc_pct_predicted"), outcome), data = dat |> select(all_of(c(outcome, predictors, "fvc_pct_predicted"))) |> drop_na()))$r.squared, 3)`
and every substantive coefficient is crushed. `fev1_fvc_ratio` is excluded for
the same reason, and more strongly: it *is* the outcome divided by FVC.

**Smoking is entered once, as dose plus a current-smoking indicator.** The
shipped `smoking_status` variable has three levels (never
`r sum(dat$smoking_status == "Never")`, former
`r sum(dat$smoking_status == "Former")`, current
`r sum(dat$smoking_status == "Current")`) and `pack_years` is **exactly zero for
all `r sum(dat$pack_years == 0)` never-smokers and strictly positive for every
ever-smoker**, so the two encode the same never/ever distinction twice. Entering
the untouched factor alongside pack-years would force the former- and
current-smoker contrasts to be estimated at zero pack-years, a value no
ever-smoker in this sample has. We therefore keep `pack_years` as the
cumulative-dose term and add one binary indicator for *current* smoking, which
captures the acute airway inflammation that lifetime dose does not.

**Variables deliberately excluded.** `cd4_count`, `on_antiretroviral` and
`viral_load_detectable` are **missing by design** for participants without HIV;
including any of them deletes most or all of the HIV-negative comparison group
and with it the exposure contrast this analysis exists to make.
`mmrc_dyspnoea` is excluded because all `r sum(is.na(dat$mmrc_dyspnoea))` of its
missing values fall in HIV-positive participants, and because breathlessness is
a consequence of airflow limitation rather than a cause of it. The SGRQ scores
and `six_min_walk_m` are likewise downstream of lung function and are analysed
as separate outcomes.

**Statistical analysis.** `lm()`, justified because the outcome is a continuous,
roughly symmetric percentage (skewness
`r fmt(mean((dat_model$fev1_pct_predicted - mean(dat_model$fev1_pct_predicted))^3) / sd(dat_model$fev1_pct_predicted)^3, 2)`).
We fitted a univariable model for each predictor first, then a single
multivariable model containing all of them. Predictors were chosen in advance on
subject-matter grounds, not by significance. Significance threshold 0.05;
estimates are reported with 95% confidence intervals. Assumptions were checked
with `performance::check_model()` and `check_collinearity()`. Analysis used
`r R.version.string` with tidyverse, gtsummary and performance.

**Parameter budget.** The model estimates
`r length(coef(lm(reformulate(predictors, outcome), data = dat_model))) - 1L`
parameters on `r nrow(dat_model)` complete cases, about
`r fmt(nrow(dat_model) / (length(coef(lm(reformulate(predictors, outcome), data = dat_model))) - 1L), 0)`
observations per parameter, well inside the 10-15 rule.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped). The outcome itself is
`r fmt(100 * mean(!is.na(dat$fev1_pct_predicted)), 1)`% complete
(`r sum(is.na(dat$fev1_pct_predicted))` missing); the rest of the loss is
haemoglobin (`r sum(is.na(dat$haemoglobin_g_dl))` missing), blood pressure
(`r sum(is.na(dat$systolic_bp))`) and DLCO
(`r sum(is.na(dat$dlco_pct_predicted))`). The analytic sample keeps
`r sum(dat_model$hiv_positive == "Yes")` participants with HIV and
`r sum(dat_model$hiv_positive == "No")` without.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(missing = "no") |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics (complete-case analytic sample).**")
```

Median FEV1 is `r fmt(median(dat_model$fev1_pct_predicted), 0)`% of predicted, so
the typical participant has normal airflow, but the range runs from
`r fmt(min(dat_model$fev1_pct_predicted), 0)`% to
`r fmt(max(dat_model$fev1_pct_predicted), 0)`% and
`r sum(dat_model$fev1_pct_predicted < 80)`
(`r fmt(100 * mean(dat_model$fev1_pct_predicted < 80), 0)`%) fall below the
conventional 80% cut-off. Median DLCO is lower, at
`r fmt(median(dat_model$dlco_pct_predicted), 0)`% of predicted, which is the
signature of this population: gas transfer is more impaired than airflow.

## Exploratory figure

```{r explore, fig.cap="Figure 1. FEV1 (% predicted) against diffusing capacity, by HIV status."}
ggplot(dat_model, aes(dlco_pct_predicted, fev1_pct_predicted, colour = hiv_positive)) +
  geom_point(alpha = 0.5, size = 1.6) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.9) +
  scale_colour_manual(values = c(No = "#3F6F9F", Yes = "#B5495B")) +
  labs(x = "DLCO (% predicted)", y = "FEV1 (% predicted)",
       colour = "Living with HIV")
```

FEV1 rises steadily with diffusing capacity, and the two fitted lines lie almost
on top of one another, which already suggests that HIV status will add little
once gas transfer is accounted for.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv

utid <- map_dfr(predictors, function(v) {
  tidy(lm(reformulate(v, outcome), data = dat_model), conf.int = TRUE) |>
    filter(term != "(Intercept)")
})
ug <- function(term, d = 2) {
  r <- utid[utid$term == term, ]
  sprintf("%s percentage points (95%% CI %s to %s)", fmt(r$estimate, d),
          fmt(r$conf.low, d), fmt(r$conf.high, d))
}
up <- function(term) pv(utid$p.value[utid$term == term])
```

Diffusing capacity is by far the strongest crude correlate: each additional
percentage point of predicted DLCO is associated with
`r ug("dlco_pct_predicted")` of predicted FEV1 (p `r up("dlco_pct_predicted")`).
Smoking follows, with each pack-year associated with `r ug("pack_years")`
(p `r up("pack_years")`) and current smokers `r ug("current_smokerYes")`
(p `r up("current_smokerYes")`) below non-current smokers. Each additional year
of age is associated with `r ug("age_yrs")` (p `r up("age_yrs")`), women average
`r ug("sexFemale")` relative to men (p `r up("sexFemale")`), and each g/dL of
haemoglobin `r ug("haemoglobin_g_dl")` (p `r up("haemoglobin_g_dl")`). **HIV
status shows no crude association at all**: `r ug("hiv_positiveYes")`
(p `r up("hiv_positiveYes")`). Neither blood pressure is associated.

## Why adjustment is needed

**Cumulative smoking** confounds nearly everything here. Participants with HIV
have smoked more (mean
`r fmt(mean(dat$pack_years[dat$hiv_positive == "Yes"]), 1)` pack-years versus
`r fmt(mean(dat$pack_years[dat$hiv_positive == "No"]), 1)`), pack-years is one of
the strongest crude predictors of FEV1, and it is also strongly related to
diffusing capacity, so it sits between HIV status, DLCO and the outcome. **Sex**
is the second variable that must be handled explicitly, and it does something
counter-intuitive: because FEV1 is already expressed as a percentage of a
sex-specific predicted value, the residual sex difference reflects who smokes and
how much rather than body size, and men in this sample are both more likely to be
in the HIV-positive group
(`r fmt(100 * mean(dat$sex[dat$hiv_positive == "Yes"] == "Male"), 0)`% versus
`r fmt(100 * mean(dat$sex[dat$hiv_positive == "No"] == "Male"), 0)`%) and to have
higher haemoglobin. **Haemoglobin** is a third: it differs sharply by sex (median
`r fmt(median(dat$haemoglobin_g_dl[dat$sex == "Male"], na.rm = TRUE), 1)` g/dL in
men versus
`r fmt(median(dat$haemoglobin_g_dl[dat$sex == "Female"], na.rm = TRUE), 1)` in
women) and is lower in people with chronic infection, so a crude haemoglobin-FEV1
association is largely a sex and chronic-illness association. Mutual adjustment
lets each estimate be read as a comparison between participants alike on the
other measured characteristics.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv

tid <- tidy(fit, conf.int = TRUE)
g <- function(term, d = 2) {
  r <- tid[tid$term == term, ]
  sprintf("%s percentage points (95%% CI %s to %s)", fmt(r$estimate, d),
          fmt(r$conf.low, d), fmt(r$conf.high, d))
}
gp <- function(term) pv(tid$p.value[tid$term == term])
```

```{r fit-stats}
s <- summary(fit)
tibble(
  Statistic = c("Complete-case N", "Parameters (excl. intercept)",
                "Observations per parameter",
                "R-squared", "Adjusted R-squared", "Residual SE (% predicted)"),
  Value = c(as.character(nrow(dat_model)),
            as.character(length(coef(fit)) - 1L),
            fmt(nrow(dat_model) / (length(coef(fit)) - 1L), 1),
            fmt(s$r.squared, 3), fmt(s$adj.r.squared, 3), fmt(s$sigma, 1))
) |>
  gt() |>
  tab_header(title = md("**Model fit: multivariable linear model for FEV1 (% predicted)**"))
```

The model explains `r fmt(100 * s$r.squared, 1)`% of the variance in FEV1 %
predicted (adjusted R-squared `r fmt(s$adj.r.squared, 3)`), with a residual
standard deviation of `r fmt(s$sigma, 1)` percentage points. Diffusing capacity
dominates: among participants of the same age, sex, HIV status, smoking history,
blood pressure and haemoglobin, each additional percentage point of predicted
DLCO is associated with `r g("dlco_pct_predicted")` of predicted FEV1
(p `r gp("dlco_pct_predicted")`). Haemoglobin is independently associated at
`r g("haemoglobin_g_dl")` per g/dL (p `r gp("haemoglobin_g_dl")`). Cumulative
smoking is borderline, `r g("pack_years")` per pack-year
(p `r gp("pack_years")`). **HIV status remains null**: `r g("hiv_positiveYes")`
(p `r gp("hiv_positiveYes")`) — the point estimate has changed sign relative to
the crude estimate but the confidence interval comfortably spans zero in both.
Age (`r g("age_yrs")`), sex (`r g("sexFemale")`), current smoking
(`r g("current_smokerYes")`) and both blood pressures are not independently
associated. These are associations among the participants observed, not effects
of intervening.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

Three predictors move a long way toward the null and lose significance.
**Current smoking** collapses from `r ug("current_smokerYes")` crude
(p `r up("current_smokerYes")`) to `r g("current_smokerYes")` adjusted
(p `r gp("current_smokerYes")`) — essentially all of it was cumulative dose,
which the pack-years term now holds constant. **Age** falls from
`r ug("age_yrs")` (p `r up("age_yrs")`) to `r g("age_yrs")`
(p `r gp("age_yrs")`), because older participants in this sample have both more
pack-years and lower diffusing capacity. **Sex** reverses direction, from women
averaging `r ug("sexFemale")` crude (p `r up("sexFemale")`) to
`r g("sexFemale")` adjusted (p `r gp("sexFemale")`); women in this sample have
smoked less (mean `r fmt(mean(dat$pack_years[dat$sex == "Female"]), 1)` versus
`r fmt(mean(dat$pack_years[dat$sex == "Male"]), 1)` pack-years) and have better
gas transfer (mean DLCO
`r fmt(mean(dat$dlco_pct_predicted[dat$sex == "Female"], na.rm = TRUE), 0)`%
versus `r fmt(mean(dat$dlco_pct_predicted[dat$sex == "Male"], na.rm = TRUE), 0)`%
of predicted), and once those are held constant the crude female advantage
disappears.

One estimate moves *away* from the null. **Haemoglobin** goes from
`r ug("haemoglobin_g_dl")` crude (p `r up("haemoglobin_g_dl")`) to
`r g("haemoglobin_g_dl")` adjusted (p `r gp("haemoglobin_g_dl")`) — already
significant crudely, and strengthened by adjustment because sex was partly
masking it: men have higher haemoglobin but, once smoking and DLCO are held
constant, no better airflow. A negative coefficient
here is not a physiological claim about oxygen carriage; it is more likely
confounding by body habitus and by the smoking-related polycythaemia that
accompanies chronic airflow limitation.

Two estimates barely move. **DLCO** goes from `r ug("dlco_pct_predicted")` to
`r g("dlco_pct_predicted")` and remains the dominant term, and **HIV status**
stays null throughout (`r ug("hiv_positiveYes")` crude,
`r g("hiv_positiveYes")` adjusted) — this is a genuine negative finding rather
than an artefact of confounding, and it is the interesting result of this
analysis.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r collin}
check_collinearity(fit)
```

```{r assumption-tests}
print(check_heteroscedasticity(fit))
print(check_normality(fit))
```

The diagnostics are good. The residual-versus-fitted panel is flat and
`check_heteroscedasticity()` does not reject constant variance
(p `r pv(as.numeric(check_heteroscedasticity(fit)))`), and the
posterior-predictive check reproduces the observed distribution closely. The
normal-QQ panel shows mild deviation in the tails and `check_normality()` rejects
at the 0.05 level (p `r pv(as.numeric(check_normality(fit)))`); with
`r nrow(dat_model)` observations this has little practical bearing on the
confidence intervals, and the outcome is close enough to symmetric (skewness
`r fmt(mean((dat_model$fev1_pct_predicted - mean(dat_model$fev1_pct_predicted))^3) / sd(dat_model$fev1_pct_predicted)^3, 2)`)
that no transformation is indicated.

Collinearity is unremarkable *because FVC was excluded*. The largest variance
inflation factor in the model is
`r fmt(max(check_collinearity(fit)$VIF), 2)`, for systolic blood pressure,
nowhere near the conventional threshold of 5, and the DLCO term sits at
`r fmt(check_collinearity(fit)$VIF[check_collinearity(fit)$Term == "dlco_pct_predicted"], 2)`.
Had FVC been included, the model would still not have shown alarming VIFs — the
problem it creates is not inflation but redundancy, and it is documented in the
Feasibility notes. No observation exceeds the Cook's distance contours (maximum
`r fmt(max(cooks.distance(fit)), 3)`).

# Discussion

In this two-centre US sample, **HIV status was not associated with airflow
obstruction**, either crudely (`r ug("hiv_positiveYes")`) or after adjustment for
age, sex, smoking, blood pressure, haemoglobin and diffusing capacity
(`r g("hiv_positiveYes")`). What does predict FEV1 is diffusing capacity — each
percentage point of predicted DLCO is worth about
`r fmt(tid$estimate[tid$term == "dlco_pct_predicted"], 2)` percentage points of
predicted FEV1 — and, marginally, cumulative tobacco dose. This is a clean
negative finding: in a population where gas transfer *is* clearly worse among
people with HIV (see `dlco_pct_predicted.Rmd` in this folder), spirometric
airflow is not. Practically, it argues that the lung problem in treated HIV is
not primarily an obstructive one, and that a normal FEV1 in a person with HIV
does not rule out clinically important lung disease.

**Limitations.** Participants were recruited at **two clinical centres,
Pittsburgh and UCSF, and the site variable is not in this deposit**, although the
source paper shows site matters; it is therefore an unmeasured confounder and an
ignored clustering variable, so the standard errors are probably optimistic.
The design is cross-sectional, so nothing here speaks to the *rate* of FEV1
decline, which is where an HIV effect would most plausibly show up; a null
cross-sectional difference is entirely compatible with faster decline over time.
Smoking is measured as self-reported status and pack-years with no biochemical
verification, and misreporting would bias the smoking coefficients toward the
null and leave residual confounding in the HIV estimate. Adjusting for DLCO is
arguably adjusting for a parallel consequence of the same lung injury rather
than a confounder, which makes the DLCO coefficient descriptive rather than
causal. Occupational exposures, cannabis and crack cocaine smoking, childhood
respiratory illness and prior pneumonia — all relevant to airflow in this
population, and the last of these specifically commoner in HIV — are not
measured. Complete-case analysis dropped `r n_dropped` of `r nrow(dat)` rows
(`r fmt(100 * n_dropped / nrow(dat), 1)`%). Residuals are mildly non-normal. All
estimates are associations, not causal effects.

**Future work.** Refit with study site as a fixed effect and with repeated
spirometry from the parent cohort so that the rate of FEV1 decline, rather than a
single cross-sectional level, becomes the outcome.

# Feasibility notes

*NOT part of a student report.*

```{r feas-fvc}
# What including FVC does to this model.
dm_fvc  <- dat |> select(all_of(c(outcome, predictors, "fvc_pct_predicted"))) |> drop_na()
fit_fvc <- lm(reformulate(c(predictors, "fvc_pct_predicted"), outcome), data = dm_fvc)
tf <- tidy(fit_fvc, conf.int = TRUE)
sf <- summary(fit_fvc)

tibble(
  Term = c("FVC (% predicted)", "DLCO (% predicted)", "Smoking (pack-years)",
           "Living with HIV (yes)", "Haemoglobin (g/dL)"),
  `Without FVC` = c("-",
                    fmt(tid$estimate[tid$term == "dlco_pct_predicted"], 3),
                    fmt(tid$estimate[tid$term == "pack_years"], 3),
                    fmt(tid$estimate[tid$term == "hiv_positiveYes"], 3),
                    fmt(tid$estimate[tid$term == "haemoglobin_g_dl"], 3)),
  `With FVC` = c(fmt(tf$estimate[tf$term == "fvc_pct_predicted"], 3),
                 fmt(tf$estimate[tf$term == "dlco_pct_predicted"], 3),
                 fmt(tf$estimate[tf$term == "pack_years"], 3),
                 fmt(tf$estimate[tf$term == "hiv_positiveYes"], 3),
                 fmt(tf$estimate[tf$term == "haemoglobin_g_dl"], 3))
) |>
  gt() |>
  tab_header(title = md("**Feasibility Table F1. Effect of adding FVC (% predicted) to the model for FEV1 (% predicted).**")) |>
  tab_source_note(md(paste0("R-squared without FVC: **", fmt(s$r.squared, 3),
                            "**; with FVC: **", fmt(sf$r.squared, 3),
                            "**. Residual SE falls from ", fmt(s$sigma, 1),
                            " to ", fmt(sf$sigma, 1), " percentage points.")))
```

```{r feas-hiv}
cc_cd4 <- dat |> select(all_of(c(outcome, predictors, "cd4_count"))) |> drop_na()
cc_arv <- dat |> select(all_of(c(outcome, predictors, "on_antiretroviral"))) |> drop_na()
cc_mmr <- dat |> select(all_of(c(outcome, predictors, "mmrc_dyspnoea"))) |> drop_na()
```

- **USABLE, with one large caveat: the headline result is a null.** The
  diagnostics pass, the complete-case N is `r nrow(dat_model)` of `r nrow(dat)`,
  collinearity is unremarkable and the model is honest — but the exposure of
  primary interest, `hiv_positive`, is not associated with the outcome either
  crudely (p `r up("hiv_positiveYes")`) or adjusted
  (p `r gp("hiv_positiveYes")`). Students who expect to "find something" about
  HIV will be disappointed, and weak ones will start dredging. It is a genuinely
  good negative result and worth flagging in the brief as such.
- **MAJOR trap: `fvc_pct_predicted` on the right-hand side.** FEV1 and FVC as
  percentages of predicted correlate at
  `r fmt(cor(dat$fev1_pct_predicted, dat$fvc_pct_predicted, use = "complete.obs"), 3)`
  and are two readings from the same manoeuvre. Feasibility Table F1 shows that
  adding FVC lifts R-squared from `r fmt(s$r.squared, 3)` to
  `r fmt(sf$r.squared, 3)` with an FVC coefficient of
  `r fmt(tf$estimate[tf$term == "fvc_pct_predicted"], 2)` — indistinguishable
  from 1, i.e. the model has discovered that FEV1 % predicted is approximately
  FVC % predicted — while the DLCO coefficient falls from
  `r fmt(tid$estimate[tid$term == "dlco_pct_predicted"], 2)` to
  `r fmt(tf$estimate[tf$term == "dlco_pct_predicted"], 2)` and haemoglobin
  becomes non-significant. Note that the VIFs stay modest, so a student who only
  checks collinearity will see nothing wrong: this trap is invisible to VIF and
  has to be caught by thinking about what the variables are. `fev1_fvc_ratio` is
  worse still, being the outcome divided by FVC.
- **FATAL if any HIV-restricted variable is used.** `cd4_count`,
  `on_antiretroviral` and `viral_load_detectable` exist only for the
  `r sum(dat$hiv_positive == "Yes")` participants with HIV. Adding
  `on_antiretroviral` takes the complete-case N from `r nrow(dat_model)` to
  `r nrow(cc_arv)` with **`r sum(cc_arv$hiv_positive == "No")` HIV-negative
  participants remaining**, so `lm()` cannot fit `hiv_positive` at all; adding
  `cd4_count` takes N to `r nrow(cc_cd4)` with only
  `r sum(cc_cd4$hiv_positive == "No")` HIV-negative participants left, and fails
  silently. `six_min_walk_m.Rmd` in this folder tabulates the damage.
- **MINOR: `mmrc_dyspnoea` is differentially missing** — all
  `r sum(is.na(dat$mmrc_dyspnoea))` missing values are HIV-positive participants,
  and adding it would take N to `r nrow(cc_mmr)`. It is also downstream of the
  outcome.
- **MINOR: `pack_years` and `smoking_status` are structurally entangled**
  (`pack_years == 0` is exactly `smoking_status == "never"` on all
  `r nrow(dat)` rows). Either parameterisation works; the choice must be
  justified. Note that with this outcome the *current smoking* term is entirely
  absorbed by pack-years on adjustment
  (`r ug("current_smokerYes")` to `r g("current_smokerYes")`), so a student who
  uses only `smoking_status` will report no smoking association at all and one
  who uses only `pack_years` will report a borderline one.
- **MINOR: mildly non-normal residuals** (p
  `r pv(as.numeric(check_normality(fit)))`); variance is constant (p
  `r pv(as.numeric(check_heteroscedasticity(fit)))`) and no transformation is
  needed.
- **Complete-case N `r nrow(dat_model)` of `r nrow(dat)`**
  (`r fmt(100 * n_dropped / nrow(dat), 1)`% lost); the outcome itself is only
  `r sum(is.na(dat$fev1_pct_predicted))` rows short of complete.
- **Structural: two-site recruitment with no site variable.** Unfixable from the
  shipped file; belongs in Limitations. It matters more here than elsewhere,
  because a null result is exactly what unmeasured between-site heterogeneity
  would manufacture.
