---
title: "Respiratory quality of life (SGRQ total) in adults with and without HIV"
subtitle: "Dataset 11: HIV infection and exercise capacity"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "hiv_lung_6mwt.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names. The three SGRQ
# subscales are NOT predictors: the total is an arithmetic function of them.
outcome    <- "sgrq_total"
predictors <- c("hiv_positive", "age_yrs", "sex", "current_smoker", "pack_years",
                "systolic_bp", "diastolic_bp", "haemoglobin_g_dl",
                "fev1_pct_predicted", "fvc_pct_predicted", "dlco_pct_predicted")

lab <- function(x, l) { attr(x, "label") <- l; x }

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    hiv_positive = lab(factor(hiv_positive, levels = c(0, 1), labels = c("No", "Yes")),
                       "Living with HIV"),
    sex = lab(factor(sex, levels = c("male", "female"), labels = c("Male", "Female")),
              "Sex"),
    # Smoking is parameterised as cumulative dose plus a current-smoking indicator;
    # see Methods for why the three-level factor is not used as-is.
    current_smoker = lab(factor(if_else(smoking_status == "current", "Yes", "No"),
                                levels = c("No", "Yes")),
                         "Current smoker"),
    smoking_status = lab(factor(smoking_status, levels = c("never", "former", "current"),
                                labels = c("Never", "Former", "Current")),
                         "Smoking status"),
    age_yrs = lab(age_yrs, "Age (years)"),
    pack_years = lab(pack_years, "Smoking (pack-years)"),
    systolic_bp = lab(systolic_bp, "Systolic BP (mmHg)"),
    diastolic_bp = lab(diastolic_bp, "Diastolic BP (mmHg)"),
    haemoglobin_g_dl = lab(haemoglobin_g_dl, "Haemoglobin (g/dL)"),
    fev1_pct_predicted = lab(fev1_pct_predicted, "FEV1 (% predicted)"),
    fvc_pct_predicted  = lab(fvc_pct_predicted, "FVC (% predicted)"),
    dlco_pct_predicted = lab(dlco_pct_predicted, "DLCO (% predicted)"),
    six_min_walk_m = lab(six_min_walk_m, "Six-minute walk distance (m)"),
    sgrq_total = lab(sgrq_total, "SGRQ total score (0-100)")
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
pv  <- function(p) ifelse(p < 0.001, "<0.001", paste0("= ", fmt(p, 3)))
```

# Introduction

The Pittsburgh/UCSF lung-function study enrolled adults living with HIV and
HIV-uninfected comparators from two US clinical centres and assessed all of them
identically with a six-minute walk test, spirometry, diffusing capacity and the
St George's Respiratory Questionnaire (Fitzpatrick and colleagues, *PLOS ONE*
2019). This teaching extract holds `r nrow(dat)` participants,
`r sum(dat$hiv_positive == "Yes")` of them living with HIV. The SGRQ is the
standard patient-reported measure of respiratory health status: it runs from 0
to 100 with higher scores meaning worse quality of life, a four-point change is
the accepted minimal important difference, and it is the outcome that determines
whether measured lung impairment actually matters to the person who has it.

**Aim.** Among these adults, we examined which of HIV status, age, sex, current
smoking, cumulative pack-years, blood pressure, haemoglobin and the three lung
function measurements (FEV1, FVC and DLCO, each as a percentage of predicted)
are independently associated with the SGRQ total score, mutually adjusting for
the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** A cross-sectional analysis of participants enrolled at two US
clinical centres (Pittsburgh and San Francisco) into a cohort study of lung
function in HIV. The extract contains `r nrow(dat)` adults aged
`r min(dat$age_yrs)` to `r max(dat$age_yrs)` and 21 variables;
`r sum(dat$hiv_positive == "Yes")` (`r fmt(100 * mean(dat$hiv_positive == "Yes"), 1)`%)
are living with HIV. All measurements were made at a single study visit.

**Variables.** The outcome is `sgrq_total`, the St George's Respiratory
Questionnaire total score on a 0-100 scale where **higher is worse** (observed
range `r fmt(min(dat$sgrq_total), 1)` to `r fmt(max(dat$sgrq_total), 1)`),
treated as continuous. Predictors are HIV status (reference HIV-negative, the
clinically unexposed group), age (years), sex (reference male, the larger
group), current smoking (reference not a current smoker), cumulative smoking
exposure (pack-years), systolic and diastolic blood pressure (mmHg), haemoglobin
(g/dL), and FEV1, FVC and DLCO each expressed as a percentage of the predicted
value for a person of the same age, sex and height.

**The three SGRQ subscales are excluded, and this is not negotiable.** The
questionnaire's total score is a fixed weighted combination of its symptoms,
activity and impacts components: regressing `sgrq_total` on the three subscales
in this file returns R-squared
`r fmt(summary(lm(sgrq_total ~ sgrq_symptoms + sgrq_activity + sgrq_impacts, data = dat))$r.squared, 4)`,
i.e. they reproduce the outcome to within rounding error. Putting any of them on
the right-hand side is regressing a score on its own components, which is
circular and tells us nothing.

**Smoking is entered once, as dose plus a current-smoking indicator.** The
shipped `smoking_status` variable has three levels (never
`r sum(dat$smoking_status == "Never")`, former
`r sum(dat$smoking_status == "Former")`, current
`r sum(dat$smoking_status == "Current")`) and `pack_years` is **exactly zero for
all `r sum(dat$pack_years == 0)` never-smokers and strictly positive for every
ever-smoker**, so the two encode the same never/ever distinction twice. Entering
the untouched factor alongside pack-years would force the former- and
current-smoker contrasts to be estimated at zero pack-years, a value no
ever-smoker in this sample has. We therefore keep `pack_years` as the
cumulative-dose term and add one binary indicator for *current* smoking, which
is the piece of information pack-years cannot carry.

**Variables deliberately excluded.** `cd4_count`, `on_antiretroviral` and
`viral_load_detectable` are **missing by design** for participants without HIV;
any of them in the model deletes most or all of the HIV-negative comparison
group. `mmrc_dyspnoea` is excluded on two grounds: it is missing for
`r sum(is.na(dat$mmrc_dyspnoea))` participants, **all of them HIV-positive**, and
it is a breathlessness rating that overlaps conceptually with the SGRQ symptoms
domain. `six_min_walk_m` is excluded because it is analysed as a separate outcome
and is best regarded as a parallel consequence of lung disease rather than a
determinant of quality of life.

**Statistical analysis.** `lm()`, justified because the outcome is a continuous
score with many distinct values. We fitted a univariable model for each predictor
first, then a single multivariable model containing all of them. Predictors were
chosen in advance on subject-matter grounds, not by significance. Significance
threshold 0.05; estimates are reported with 95% confidence intervals.
Assumptions were checked with `performance::check_model()`, and because
non-constant variance was flagged we additionally report HC3
heteroscedasticity-consistent standard errors. Analysis used
`r R.version.string` with tidyverse, gtsummary and performance.

**Parameter budget.** The model estimates
`r length(coef(lm(reformulate(predictors, outcome), data = dat_model))) - 1L`
parameters on `r nrow(dat_model)` complete cases, about
`r fmt(nrow(dat_model) / (length(coef(lm(reformulate(predictors, outcome), data = dat_model))) - 1L), 0)`
observations per parameter, comfortably inside the 10-15 rule. No cut to the
suggested predictor list was needed.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped). The outcome itself is complete
(`r sum(!is.na(dat$sgrq_total))`/`r nrow(dat)`); the loss is entirely from
predictors — haemoglobin (`r sum(is.na(dat$haemoglobin_g_dl))` missing),
spirometry (`r sum(is.na(dat$fev1_pct_predicted))`), blood pressure
(`r sum(is.na(dat$systolic_bp))`) and DLCO
(`r sum(is.na(dat$dlco_pct_predicted))`). The analytic sample keeps
`r sum(dat_model$hiv_positive == "Yes")` participants with HIV and
`r sum(dat_model$hiv_positive == "No")` without.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(missing = "no") |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics (complete-case analytic sample).**")
```

The SGRQ distribution is heavily floor-weighted: the median total is
`r fmt(median(dat_model$sgrq_total), 1)`,
`r sum(dat_model$sgrq_total == 0)` participants score exactly zero (no
respiratory impact at all) and the interquartile range runs only to
`r fmt(quantile(dat_model$sgrq_total, 0.75), 1)`, yet the maximum is
`r fmt(max(dat_model$sgrq_total), 1)`. The sample is middle-aged (median age
`r median(dat_model$age_yrs)` years), mostly male, and about
`r fmt(100 * mean(dat_model$current_smoker == "Yes"), 0)`% are current smokers.

## Exploratory figure

```{r explore, fig.cap="Figure 1. SGRQ total score against FEV1 (% predicted), by HIV status."}
ggplot(dat_model, aes(fev1_pct_predicted, sgrq_total, colour = hiv_positive)) +
  geom_point(alpha = 0.5, size = 1.6) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.9) +
  scale_colour_manual(values = c(No = "#3F6F9F", Yes = "#B5495B")) +
  labs(x = "FEV1 (% predicted)", y = "SGRQ total score (0-100, higher = worse)",
       colour = "Living with HIV")
```

Respiratory quality of life worsens as FEV1 falls in both groups, but the cloud
is strongly asymmetric — a dense floor of participants reporting no symptoms at
every level of lung function, and a scatter of high scores concentrated at low
FEV1 — which is the pattern that will show up later as non-constant residual
variance.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv

utid <- map_dfr(predictors, function(v) {
  tidy(lm(reformulate(v, outcome), data = dat_model), conf.int = TRUE) |>
    filter(term != "(Intercept)")
})
ug <- function(term, d = 2) {
  r <- utid[utid$term == term, ]
  sprintf("%s points (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d), fmt(r$conf.high, d))
}
up <- function(term) pv(utid$p.value[utid$term == term])
```

Crudely, participants living with HIV score `r ug("hiv_positiveYes")` higher on
the SGRQ, i.e. worse respiratory quality of life (p `r up("hiv_positiveYes")`),
a difference just at the four-point minimal important threshold. Airflow is the
strongest crude correlate: each additional percentage point of predicted FEV1 is
associated with `r ug("fev1_pct_predicted")` (p `r up("fev1_pct_predicted")`), so
a 20-point deficit in FEV1 corresponds to about
`r fmt(abs(20 * utid$estimate[utid$term == "fev1_pct_predicted"]), 1)` SGRQ
points. FVC (`r ug("fvc_pct_predicted")`, p `r up("fvc_pct_predicted")`) and DLCO
(`r ug("dlco_pct_predicted")`, p `r up("dlco_pct_predicted")`) show weaker
gradients in the same direction. Current smokers score
`r ug("current_smokerYes")` worse (p `r up("current_smokerYes")`), each
pack-year `r ug("pack_years")` (p `r up("pack_years")`), women
`r ug("sexFemale")` worse than men (p `r up("sexFemale")`), and each additional
g/dL of haemoglobin is associated with `r ug("haemoglobin_g_dl")`
(p `r up("haemoglobin_g_dl")`). Age
(`r ug("age_yrs")`, p `r up("age_yrs")`) and both blood pressures show no crude
association.

## Why adjustment is needed

**Current smoking** is the obvious confounder of the HIV association.
`r fmt(100 * mean(dat$current_smoker[dat$hiv_positive == "Yes"] == "Yes"), 0)`%
of participants with HIV are current smokers against
`r fmt(100 * mean(dat$current_smoker[dat$hiv_positive == "No"] == "Yes"), 0)`% of
the HIV-negative group, and current smoking is itself associated with
`r ug("current_smokerYes")` on the SGRQ, partly through cough and sputum
questions in the symptoms domain that ask directly about smoking-related
complaints. **Airflow obstruction** is the second: FEV1 is the strongest crude
predictor of the SGRQ here, and it differs between the two groups, so any crude
HIV difference is partly a lung-function difference. **Sex** is a third and works
in the opposite direction from the naive expectation: women in this sample have
*better* measured lung function (median FEV1
`r fmt(median(dat$fev1_pct_predicted[dat$sex == "Female"], na.rm = TRUE), 0)`%
versus `r fmt(median(dat$fev1_pct_predicted[dat$sex == "Male"], na.rm = TRUE), 0)`%
of predicted) but report *worse* respiratory quality of life, so leaving sex out
would make the lung-function coefficients look smaller than they are. Mutual
adjustment lets each estimate be read as a comparison between participants alike
on the other measured characteristics.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv

tid <- tidy(fit, conf.int = TRUE)
g <- function(term, d = 2) {
  r <- tid[tid$term == term, ]
  sprintf("%s points (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d), fmt(r$conf.high, d))
}
gp <- function(term) pv(tid$p.value[tid$term == term])
```

```{r fit-stats}
s <- summary(fit)
tibble(
  Statistic = c("Complete-case N", "Parameters (excl. intercept)",
                "Observations per parameter",
                "R-squared", "Adjusted R-squared", "Residual SE (SGRQ points)"),
  Value = c(as.character(nrow(dat_model)),
            as.character(length(coef(fit)) - 1L),
            fmt(nrow(dat_model) / (length(coef(fit)) - 1L), 1),
            fmt(s$r.squared, 3), fmt(s$adj.r.squared, 3), fmt(s$sigma, 1))
) |>
  gt() |>
  tab_header(title = md("**Model fit: multivariable linear model for SGRQ total score**"))
```

The model explains `r fmt(100 * s$r.squared, 1)`% of the variance in the SGRQ
total (adjusted R-squared `r fmt(s$adj.r.squared, 3)`), with a residual standard
deviation of `r fmt(s$sigma, 1)` points — nearly four times the minimal important
difference, so individual scores are not predictable from these characteristics.

Among participants of the same age, sex, smoking history, blood pressure,
haemoglobin and lung function, those living with HIV score
`r g("hiv_positiveYes")` worse (p `r gp("hiv_positiveYes")`). Airflow remains the
dominant correlate: each percentage point of predicted FEV1 is associated with
`r g("fev1_pct_predicted")` (p `r gp("fev1_pct_predicted")`). The FVC
coefficient, `r g("fvc_pct_predicted")` (p `r gp("fvc_pct_predicted")`), has
flipped to the *opposite* sign from its crude estimate; the two together are
effectively encoding the FEV1/FVC ratio, so what the model is really saying is
that quality of life tracks airflow *obstruction* rather than lung size (see
Diagnostics). Among participants alike on everything else, women score
`r g("sexFemale")` worse than men (p `r gp("sexFemale")`), current smokers
`r g("current_smokerYes")` worse than non-current smokers
(p `r gp("current_smokerYes")`), and each mmHg of systolic pressure
`r g("systolic_bp")` (p `r gp("systolic_bp")`). Age (`r g("age_yrs")`),
pack-years (`r g("pack_years")`), diastolic pressure, haemoglobin and DLCO
(`r g("dlco_pct_predicted")`) are not independently associated. These are
associations among the participants observed, not effects of intervening.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

The most dramatic movement is **FVC**, which reverses sign entirely: crude
`r ug("fvc_pct_predicted")` (p `r up("fvc_pct_predicted")`), adjusted
`r g("fvc_pct_predicted")` (p `r gp("fvc_pct_predicted")`). This is collinearity
with FEV1, not a substantive finding. The two correlate at
`r fmt(cor(dat_model$fev1_pct_predicted, dat_model$fvc_pct_predicted), 2)` and
have VIFs of
`r fmt(check_collinearity(fit)$VIF[check_collinearity(fit)$Term == "fev1_pct_predicted"], 1)`
and
`r fmt(check_collinearity(fit)$VIF[check_collinearity(fit)$Term == "fvc_pct_predicted"], 1)`;
when a negative FEV1 coefficient and a positive FVC coefficient of similar size
sit side by side, the pair is behaving as a difference term, which is to say as
the FEV1/FVC ratio. **DLCO** is the other big mover, from
`r ug("dlco_pct_predicted")` crude (p `r up("dlco_pct_predicted")`) to
`r g("dlco_pct_predicted")` adjusted (p `r gp("dlco_pct_predicted")`) — the
entire crude gas-transfer association is explained by the spirometry and smoking
variables it travels with. **Pack-years** moves the same way, from
`r ug("pack_years")` to `r g("pack_years")`, losing significance, while its
companion **current smoking** attenuates only from `r ug("current_smokerYes")` to
`r g("current_smokerYes")` and keeps it — of the two smoking terms it is present
smoking, not lifetime dose, that carries the symptom burden.

Two estimates barely move. **HIV status** goes from `r ug("hiv_positiveYes")` to
`r g("hiv_positiveYes")`, if anything slightly *stronger* after adjustment, which
means the HIV difference in respiratory quality of life is not explained by the
heavier smoking or the lung function of the HIV-positive group. **Sex** likewise
holds at `r ug("sexFemale")` crude and `r g("sexFemale")` adjusted. One estimate
crosses into significance: **systolic blood pressure**, from `r ug("systolic_bp")`
(p `r up("systolic_bp")`) to `r g("systolic_bp")` (p `r gp("systolic_bp")`),
which appears only once diastolic pressure is held constant and is best read as
a fragile artefact of adjusting one blood-pressure component for the other rather
than a discovery about respiratory health.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r collin}
check_collinearity(fit)
```

```{r assumption-tests}
print(check_heteroscedasticity(fit))
print(check_normality(fit))
```

```{r robust-se}
rob   <- lmtest::coeftest(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
robci <- lmtest::coefci(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
tibble(
  Term = rownames(rob),
  Estimate = fmt(rob[, 1], 3),
  `SE (OLS)` = fmt(summary(fit)$coefficients[, 2], 3),
  `SE (HC3)` = fmt(rob[, 2], 3),
  `95% CI (HC3)` = paste0(fmt(robci[, 1], 2), " to ", fmt(robci[, 2], 2)),
  `p (HC3)` = fmt(rob[, 4], 3)
) |>
  filter(Term != "(Intercept)") |>
  gt() |>
  tab_header(title = md("**Table 3. HC3 heteroscedasticity-consistent standard errors.**"))
```

```{r sqrt-check}
fit_sqrt <- lm(reformulate(predictors, "sqrt(sgrq_total)"), data = dat_model)
```

The diagnostics are the weakest part of this analysis, and the cause is the shape
of the outcome. The SGRQ total is strongly right-skewed (skewness
`r fmt(mean((dat_model$sgrq_total - mean(dat_model$sgrq_total))^3) / sd(dat_model$sgrq_total)^3, 2)`)
with a floor at zero that `r sum(dat_model$sgrq_total == 0)` participants sit on
exactly. The posterior-predictive check shows the consequence directly: the model
predicts a symmetric spread including negative scores, which the 0-100 scale
cannot produce. The residual-versus-fitted panel fans out as fitted values rise —
participants predicted to have good respiratory quality of life are tightly
clustered near the floor, those predicted to be worse are widely scattered — and
`check_heteroscedasticity()` rejects constant variance
(p `r pv(as.numeric(check_heteroscedasticity(fit)))`). `check_normality()` also
rejects (p `r pv(as.numeric(check_normality(fit)))`).

Table 3 therefore reports HC3 robust standard errors. They change the numbers a
little — the FEV1 standard error rises from
`r fmt(summary(fit)$coefficients["fev1_pct_predicted", 2], 3)` to
`r fmt(rob["fev1_pct_predicted", 2], 3)` — but no substantive conclusion changes:
HIV status, sex, current smoking, systolic pressure, FEV1 and FVC all keep their
significance status under robust inference. A square-root transformation fixes
the variance problem (heteroscedasticity p
`r pv(as.numeric(check_heteroscedasticity(fit_sqrt)))` on the transformed scale)
but not the normality problem
(p `r pv(as.numeric(check_normality(fit_sqrt)))`), because the pile-up of exact
zeros survives any monotone transformation; and it costs the direct reading of
coefficients in SGRQ points against a known four-point important difference. We
have kept the untransformed model with robust standard errors and flagged the
violation rather than hiding it.

Collinearity is the other flagged check. The two largest VIFs are
`r fmt(check_collinearity(fit)$VIF[check_collinearity(fit)$Term == "fev1_pct_predicted"], 2)`
for FEV1 and
`r fmt(check_collinearity(fit)$VIF[check_collinearity(fit)$Term == "fvc_pct_predicted"], 2)`
for FVC, against a maximum of
`r fmt(max(check_collinearity(fit)$VIF[!check_collinearity(fit)$Term %in% c("fev1_pct_predicted", "fvc_pct_predicted")]), 2)`
elsewhere in the model. Both are below the conventional threshold of 5 and both
terms are significant, so nothing is unestimable, but the sign flip on FVC is a
direct symptom of the two competing and neither coefficient should be quoted
alone. No observation exceeds the Cook's distance contours (maximum
`r fmt(max(cooks.distance(fit)), 3)`).

# Discussion

In this two-centre US sample, adults living with HIV reported respiratory
quality of life about `r fmt(tid$estimate[tid$term == "hiv_positiveYes"], 1)`
SGRQ points worse than HIV-negative adults of the same age, sex, smoking history,
blood pressure, haemoglobin and lung function — a difference at the accepted
minimal important threshold of four points, and one that is not explained by the
heavier smoking or poorer lung function of the HIV-positive group. Airflow is the
dominant physiological correlate: the FEV1 and FVC coefficients together amount
to saying that obstruction, rather than lung size, is what patients feel.
Diffusing capacity, which is the strongest lung-function correlate of *walking
distance* in this same dataset, has no independent association with *symptoms* at
all — a useful reminder that measured impairment and perceived health are
different outcomes. Practically, respiratory symptoms in people with HIV should
not be dismissed as smoking alone.

**Limitations.** Participants were recruited at **two clinical centres,
Pittsburgh and UCSF, and the site variable is not in this deposit**, although the
source paper shows site matters; it is therefore both an unmeasured confounder
and an ignored clustering variable, so the standard errors here are probably
optimistic. Residual variance is non-constant and residuals are non-normal
because the SGRQ has a hard floor that `r sum(dat_model$sgrq_total == 0)`
participants occupy; HC3 standard errors repair inference but not the model's
inability to respect the scale bounds. The design is cross-sectional and the
outcome is self-reported at the same visit as everything else, so common-method
bias and reverse causation are both live: a participant feeling unwell may
report worse symptoms *and* perform worse on spirometry on the day. Current
smoking and the SGRQ symptoms domain overlap by construction, since the
questionnaire asks about cough and sputum, so part of the smoking coefficient is
measurement rather than effect. FEV1 and FVC compete for the same variance and
the FVC sign flip should not be interpreted substantively. Depression, which is
common in this population and strongly influences questionnaire-based quality of
life, is not measured. Complete-case analysis dropped `r n_dropped` of
`r nrow(dat)` rows (`r fmt(100 * n_dropped / nrow(dat), 1)`%). All estimates are
associations, not causal effects.

**Future work.** Refit with study site as a fixed effect and with a mental-health
measure in the adjustment set, and consider a two-part or beta-regression model
that respects the floor at zero rather than forcing a symmetric error
distribution onto a bounded score.

# Feasibility notes

*NOT part of a student report.*

```{r feas-subscales}
fit_bad <- lm(sgrq_total ~ sgrq_symptoms + sgrq_activity + sgrq_impacts, data = dat)
r2_bad  <- summary(fit_bad)$r.squared

# What the HIV-restricted columns cost, in one line each.
cc_cd4 <- dat |> select(all_of(c(outcome, predictors, "cd4_count"))) |> drop_na()
cc_arv <- dat |> select(all_of(c(outcome, predictors, "on_antiretroviral"))) |> drop_na()
cc_mmr <- dat |> select(all_of(c(outcome, predictors, "mmrc_dyspnoea"))) |> drop_na()
```

- **USABLE WITH WARNINGS.** The exposure of interest is significant and stable
  under adjustment, and the substantive story (obstruction and HIV, not gas
  transfer) is a good one — but the outcome distribution breaks two assumption
  checks and the fix costs interpretability.
- **FATAL: the three SGRQ subscales must never be predictors.** `sgrq_total` is
  the questionnaire's fixed weighted combination of `sgrq_symptoms`,
  `sgrq_activity` and `sgrq_impacts`; regressing it on the three returns
  R-squared `r fmt(r2_bad, 4)`. A student who reaches for "other respiratory
  measures in the file" will produce a model that explains essentially all of the
  variance and whose coefficients are the SGRQ scoring algorithm
  (`r fmt(coef(fit_bad)["sgrq_symptoms"], 2)` for symptoms,
  `r fmt(coef(fit_bad)["sgrq_activity"], 2)` for activity,
  `r fmt(coef(fit_bad)["sgrq_impacts"], 2)` for impacts — the published SGRQ
  weights), not epidemiology. The catalogue lists all four SGRQ columns together
  and does not warn about this.
- **FATAL if any HIV-restricted variable is used.** `cd4_count`,
  `on_antiretroviral` and `viral_load_detectable` exist only for the
  `r sum(dat$hiv_positive == "Yes")` participants with HIV. Adding
  `on_antiretroviral` here takes the complete-case N from `r nrow(dat_model)` to
  `r nrow(cc_arv)` with **`r sum(cc_arv$hiv_positive == "No")` HIV-negative
  participants left**, so `hiv_positive` becomes a constant and the model will
  not fit at all; adding `cd4_count` takes N to `r nrow(cc_cd4)` with only
  `r sum(cc_cd4$hiv_positive == "No")` HIV-negative participants, which is worse
  because it fails silently. `six_min_walk_m.Rmd` in this folder quantifies the
  damage in a table.
- **MAJOR: heteroscedasticity (p
  `r pv(as.numeric(check_heteroscedasticity(fit)))`) and non-normality (p
  `r pv(as.numeric(check_normality(fit)))`),** both driven by the floor at zero
  (`r sum(dat_model$sgrq_total == 0)` participants score exactly 0,
  `r fmt(100 * mean(dat_model$sgrq_total == 0), 0)`% of the sample) and the long
  right tail (skewness
  `r fmt(mean((dat_model$sgrq_total - mean(dat_model$sgrq_total))^3) / sd(dat_model$sgrq_total)^3, 2)`).
  A square-root transform fixes the variance (p
  `r pv(as.numeric(check_heteroscedasticity(fit_sqrt)))`) but not the normality
  (p `r pv(as.numeric(check_normality(fit_sqrt)))`). Any honest student report
  has to record a failed check here.
- **MAJOR: the FVC coefficient reverses sign on adjustment**
  (`r ug("fvc_pct_predicted")` crude to `r g("fvc_pct_predicted")` adjusted,
  both significant) because FEV1 and FVC are collinear (r
  `r fmt(cor(dat_model$fev1_pct_predicted, dat_model$fvc_pct_predicted), 2)`,
  VIFs
  `r fmt(check_collinearity(fit)$VIF[check_collinearity(fit)$Term == "fev1_pct_predicted"], 2)`
  and
  `r fmt(check_collinearity(fit)$VIF[check_collinearity(fit)$Term == "fvc_pct_predicted"], 2)`).
  This is the single most misinterpretable result in the document: a weak student
  will report "larger lungs are associated with worse quality of life". It is
  salvageable as a teaching point about what a pair of opposite-signed collinear
  terms means, but it needs to be flagged.
- **MINOR: `mmrc_dyspnoea` is both differentially missing and conceptually
  circular here.** All `r sum(is.na(dat$mmrc_dyspnoea))` missing values are in
  HIV-positive participants; adding it would take N to `r nrow(cc_mmr)`. It also
  measures breathlessness, which the SGRQ activity domain measures again.
- **MINOR: `pack_years` and `smoking_status` are structurally entangled**
  (`pack_years == 0` is exactly `smoking_status == "never"` on all
  `r nrow(dat)` rows). The parameterisation chosen here is documented in
  Methods; the alternative is defensible but must be justified rather than
  stumbled into.
- **Complete-case N `r nrow(dat_model)` of `r nrow(dat)`**
  (`r fmt(100 * n_dropped / nrow(dat), 1)`% lost), matching the catalogue's
  advertised 410/427.
- **Structural: two-site recruitment with no site variable.** Unfixable from the
  shipped file; belongs in Limitations. Site also plausibly affects a
  self-reported questionnaire outcome more than a physiological one.
