---
title: "Six-minute walk distance in adults with and without HIV"
subtitle: "Dataset 11: HIV infection and exercise capacity"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "hiv_lung_6mwt.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "six_min_walk_m"
predictors <- c("hiv_positive", "age_yrs", "sex", "current_smoker", "pack_years",
                "systolic_bp", "diastolic_bp", "haemoglobin_g_dl",
                "fev1_pct_predicted", "fvc_pct_predicted", "dlco_pct_predicted")

lab <- function(x, l) { attr(x, "label") <- l; x }

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    hiv_positive = lab(factor(hiv_positive, levels = c(0, 1), labels = c("No", "Yes")),
                       "Living with HIV"),
    sex = lab(factor(sex, levels = c("male", "female"), labels = c("Male", "Female")),
              "Sex"),
    # Smoking is parameterised as cumulative dose plus a current-smoking indicator;
    # see Methods for why the three-level factor is not used as-is.
    current_smoker = lab(factor(if_else(smoking_status == "current", "Yes", "No"),
                                levels = c("No", "Yes")),
                         "Current smoker"),
    smoking_status = lab(factor(smoking_status, levels = c("never", "former", "current"),
                                labels = c("Never", "Former", "Current")),
                         "Smoking status"),
    age_yrs = lab(age_yrs, "Age (years)"),
    pack_years = lab(pack_years, "Smoking (pack-years)"),
    systolic_bp = lab(systolic_bp, "Systolic BP (mmHg)"),
    diastolic_bp = lab(diastolic_bp, "Diastolic BP (mmHg)"),
    haemoglobin_g_dl = lab(haemoglobin_g_dl, "Haemoglobin (g/dL)"),
    fev1_pct_predicted = lab(fev1_pct_predicted, "FEV1 (% predicted)"),
    fvc_pct_predicted  = lab(fvc_pct_predicted, "FVC (% predicted)"),
    dlco_pct_predicted = lab(dlco_pct_predicted, "DLCO (% predicted)"),
    six_min_walk_m = lab(six_min_walk_m, "Six-minute walk distance (m)")
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
pv  <- function(p) ifelse(p < 0.001, "<0.001", paste0("= ", fmt(p, 3)))
```

# Introduction

The Pittsburgh/UCSF lung-function study enrolled adults living with HIV and
HIV-uninfected comparators from two US clinical centres and put every
participant through the same battery: a six-minute walk test, spirometry,
diffusing capacity for carbon monoxide and the St George's Respiratory
Questionnaire (Fitzpatrick and colleagues, *PLOS ONE* 2019). This teaching
extract holds `r nrow(dat)` participants, `r sum(dat$hiv_positive == "Yes")` of
them living with HIV. Six-minute walk distance is a cheap, well-standardised
measure of integrated cardiopulmonary and musculoskeletal function that predicts
hospitalisation and death in chronic lung disease, and reduced exercise capacity
is one of the more consistent reports in ageing HIV cohorts.

**Aim.** Among these adults, we examined which of HIV status, age, sex, current
smoking, cumulative pack-years, blood pressure, haemoglobin and the three lung
function measurements (FEV1, FVC and DLCO, each as a percentage of predicted)
are independently associated with six-minute walk distance, mutually adjusting
for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** A cross-sectional analysis of participants enrolled at two US
clinical centres (Pittsburgh and San Francisco) into a cohort study of lung
function in HIV. The extract contains `r nrow(dat)` adults aged
`r min(dat$age_yrs)` to `r max(dat$age_yrs)` and 21 variables;
`r sum(dat$hiv_positive == "Yes")` (`r fmt(100 * mean(dat$hiv_positive == "Yes"), 1)`%)
are living with HIV and `r sum(dat$hiv_positive == "No")` are not. All
measurements were made at a single study visit.

**Variables.** The outcome is `six_min_walk_m`, the distance in metres covered
in a standardised six-minute walk test (observed range
`r fmt(min(dat$six_min_walk_m), 0)` to `r fmt(max(dat$six_min_walk_m), 0)` m),
treated as continuous. Predictors are HIV status (reference HIV-negative, the
clinically unexposed group), age (years), sex (reference male, the larger group),
current smoking (reference not a current smoker), cumulative smoking exposure
(pack-years), systolic and diastolic blood pressure (mmHg), haemoglobin (g/dL),
and FEV1, FVC and DLCO each expressed as a percentage of the predicted value for
a person of the same age, sex and height.

**Smoking is entered once, as dose plus a current-smoking indicator.** The
shipped `smoking_status` variable has three levels (never
`r sum(dat$smoking_status == "Never")`, former
`r sum(dat$smoking_status == "Former")`, current
`r sum(dat$smoking_status == "Current")`), and `pack_years` is **exactly zero for
all `r sum(dat$pack_years == 0)` never-smokers and strictly positive for every
ever-smoker**. The two variables therefore encode the same never/ever
distinction twice, and putting the untouched factor alongside pack-years forces
the former- and current-smoker contrasts to be estimated at zero pack-years, a
value no ever-smoker in this sample has. We instead keep `pack_years` as the
cumulative-dose term and add a single binary indicator for *current* smoking,
which is the one piece of information pack-years cannot carry (current smoking
has acute effects on carboxyhaemoglobin and airway inflammation that are
independent of lifetime dose). This costs two parameters instead of three and
imposes the sensible continuity assumption that a former smoker with negligible
lifetime exposure resembles a never-smoker.

**Variables deliberately excluded.** `cd4_count`, `on_antiretroviral` and
`viral_load_detectable` are **missing by design** for participants without HIV
and are excluded from every model; the Feasibility notes below quantify what
happens if they are not. `mmrc_dyspnoea` is excluded because it is missing for
`r sum(is.na(dat$mmrc_dyspnoea))` participants, **all of them HIV-positive**, so
it would delete a third of the exposed group. `fev1_fvc_ratio` is excluded
because it is very largely a function of the two %-predicted values already in
the model (regressing it on them gives R-squared
`r fmt(summary(lm(fev1_fvc_ratio ~ fev1_pct_predicted + fvc_pct_predicted, data = dat))$r.squared, 2)`).
The three SGRQ subscales and the SGRQ total are downstream symptom reports rather
than characteristics of the participant, and are analysed as outcomes elsewhere
in this folder rather than used as predictors here.

**Statistical analysis.** `lm()`, justified because the outcome is a continuous
distance in metres with a symmetric distribution (skewness
`r fmt(mean((dat$six_min_walk_m - mean(dat$six_min_walk_m))^3) / sd(dat$six_min_walk_m)^3, 2)`).
We fitted a univariable model for each predictor first, then a single
multivariable model containing all of them. Predictors were chosen in advance on
subject-matter grounds, not by significance: HIV status is the exposure of
interest, and the remainder are the demographic, smoking, haemodynamic and
lung-function characteristics that the source paper treats as determinants of
walk distance. Significance threshold 0.05; estimates are reported with 95%
confidence intervals. Assumptions were checked with
`performance::check_model()`, with `check_collinearity()` inspected specifically
because FEV1 and FVC are two readings from the same spirometry manoeuvre.
Analysis used `r R.version.string` with tidyverse, gtsummary and performance.

**Parameter budget.** The model estimates
`r length(coef(lm(reformulate(predictors, outcome), data = dat_model))) - 1L`
parameters on `r nrow(dat_model)` complete cases, i.e. about
`r fmt(nrow(dat_model) / (length(coef(lm(reformulate(predictors, outcome), data = dat_model))) - 1L), 0)`
observations per parameter, comfortably inside the 10-15 rule for a linear
model. No cut to the suggested predictor list was needed.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped). The loss comes from haemoglobin
(`r sum(is.na(dat$haemoglobin_g_dl))` missing), spirometry
(`r sum(is.na(dat$fev1_pct_predicted))` missing), blood pressure
(`r sum(is.na(dat$systolic_bp))` missing) and DLCO
(`r sum(is.na(dat$dlco_pct_predicted))` missing); the outcome itself is complete.
The analytic sample retains `r sum(dat_model$hiv_positive == "Yes")` participants
with HIV and `r sum(dat_model$hiv_positive == "No")` without, so the comparison
the dataset exists for is essentially intact.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(missing = "no") |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics (complete-case analytic sample).**")
```

The sample is middle-aged (median age `r median(dat_model$age_yrs)` years),
predominantly male and heavily exposed to tobacco: about
`r fmt(100 * mean(dat_model$current_smoker == "Yes"), 0)`% are current smokers.
Median walk distance is `r fmt(median(dat_model$six_min_walk_m), 0)` m, with a
wide spread from `r fmt(min(dat_model$six_min_walk_m), 0)` to
`r fmt(max(dat_model$six_min_walk_m), 0)` m, and median lung function sits close
to predicted for FEV1 and FVC but somewhat below predicted for DLCO
(`r fmt(median(dat_model$dlco_pct_predicted), 0)`% of predicted).

## Exploratory figure

```{r explore, fig.cap="Figure 1. Six-minute walk distance against diffusing capacity, by HIV status."}
ggplot(dat_model, aes(dlco_pct_predicted, six_min_walk_m, colour = hiv_positive)) +
  geom_point(alpha = 0.5, size = 1.6) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.9) +
  scale_colour_manual(values = c(No = "#3F6F9F", Yes = "#B5495B")) +
  labs(x = "DLCO (% predicted)", y = "Six-minute walk distance (m)",
       colour = "Living with HIV")
```

Walk distance rises steadily with diffusing capacity in both groups, and the two
fitted lines are close to parallel with the HIV-positive line sitting below the
HIV-negative one across the whole range of DLCO, which is what an additive model
with an HIV main effect assumes.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv

utid <- map_dfr(predictors, function(v) {
  tidy(lm(reformulate(v, outcome), data = dat_model), conf.int = TRUE) |>
    filter(term != "(Intercept)")
})
ug <- function(term, d = 2) {
  r <- utid[utid$term == term, ]
  sprintf("%s m (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d), fmt(r$conf.high, d))
}
up <- function(term) pv(utid$p.value[utid$term == term])
```

Crudely, participants living with HIV walk less far than HIV-negative
participants: `r ug("hiv_positiveYes")` (p `r up("hiv_positiveYes")`), a deficit
of about `r fmt(abs(utid$estimate[utid$term == "hiv_positiveYes"]), 0)` m. All three lung-function
measures show strong crude gradients: each additional percentage point of
predicted FEV1 is worth `r ug("fev1_pct_predicted")` (p
`r up("fev1_pct_predicted")`), of predicted FVC `r ug("fvc_pct_predicted")`
(p `r up("fvc_pct_predicted")`) and of predicted DLCO
`r ug("dlco_pct_predicted")` (p `r up("dlco_pct_predicted")`). Smoking matters in
both parameterisations: each pack-year is associated with
`r ug("pack_years")` (p `r up("pack_years")`), and current smokers walk shorter
distances than non-current smokers, `r ug("current_smokerYes")` (p
`r up("current_smokerYes")`). Each additional year of age is associated with
`r ug("age_yrs")` (p `r up("age_yrs")`). Sex (`r ug("sexFemale")`,
p `r up("sexFemale")`), blood pressure and haemoglobin show no crude
association.

## Why adjustment is needed

**Cumulative smoking** is the clearest confounder of the HIV association.
Participants with HIV in this sample have smoked more
(mean `r fmt(mean(dat$pack_years[dat$hiv_positive == "Yes"]), 1)` pack-years
versus `r fmt(mean(dat$pack_years[dat$hiv_positive == "No"]), 1)` in the
HIV-negative group), and pack-years is itself one of the strongest crude
predictors of walk distance (`r ug("pack_years")` per pack-year). Some of the
crude HIV deficit is therefore a tobacco deficit. **Diffusing capacity** is the
second variable that must be handled explicitly: median DLCO is
`r fmt(median(dat$dlco_pct_predicted[dat$hiv_positive == "Yes"], na.rm = TRUE), 0)`%
of predicted in participants with HIV against
`r fmt(median(dat$dlco_pct_predicted[dat$hiv_positive == "No"], na.rm = TRUE), 0)`%
in those without, and DLCO is strongly related to walk distance, so it lies on
the path between the two. Whether that makes it a confounder or a *mediator* is
a genuine ambiguity: if HIV impairs gas transfer and impaired gas transfer
shortens walk distance, then adjusting for DLCO removes part of the very effect
we are trying to estimate, and the adjusted HIV coefficient should be read as the
part of the HIV deficit *not* explained by measured lung function. **Sex** works
on the smoking and haemoglobin coefficients rather than on HIV: men in this
sample have higher haemoglobin (median
`r fmt(median(dat$haemoglobin_g_dl[dat$sex == "Male"], na.rm = TRUE), 1)` versus
`r fmt(median(dat$haemoglobin_g_dl[dat$sex == "Female"], na.rm = TRUE), 1)` g/dL)
and different smoking histories, and sex also drives absolute walking capacity
through stature and leg length. Mutual adjustment lets each coefficient be read
as a comparison between participants alike on the other measured
characteristics.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv

tid <- tidy(fit, conf.int = TRUE)
g <- function(term, d = 2) {
  r <- tid[tid$term == term, ]
  sprintf("%s m (95%% CI %s to %s)", fmt(r$estimate, d), fmt(r$conf.low, d), fmt(r$conf.high, d))
}
gp <- function(term) pv(tid$p.value[tid$term == term])
```

```{r fit-stats}
s <- summary(fit)
tibble(
  Statistic = c("Complete-case N", "Parameters (excl. intercept)",
                "Observations per parameter",
                "R-squared", "Adjusted R-squared", "Residual SE (m)"),
  Value = c(as.character(nrow(dat_model)),
            as.character(length(coef(fit)) - 1L),
            fmt(nrow(dat_model) / (length(coef(fit)) - 1L), 1),
            fmt(s$r.squared, 3), fmt(s$adj.r.squared, 3), fmt(s$sigma, 1))
) |>
  gt() |>
  tab_header(title = md("**Model fit: multivariable linear model for six-minute walk distance**"))
```

The model explains `r fmt(100 * s$r.squared, 1)`% of the variance in walk
distance (adjusted R-squared `r fmt(s$adj.r.squared, 3)`), leaving a residual
standard deviation of `r fmt(s$sigma, 0)` m — larger than the whole HIV
difference, so the model orders groups rather than individuals. Among
participants of the same age, sex, smoking history, blood pressure, haemoglobin
and lung function, those living with HIV still walk shorter distances,
`r g("hiv_positiveYes")` (p `r gp("hiv_positiveYes")`), i.e. about
`r fmt(abs(tid$estimate[tid$term == "hiv_positiveYes"]), 0)` m less. Among
participants alike on everything else, each
additional pack-year is associated with `r g("pack_years")`
(p `r gp("pack_years")`), so a 40 pack-year history corresponds to about
`r fmt(abs(40 * tid$estimate[tid$term == "pack_years"]), 0)` m. Diffusing
capacity is the only lung-function measure that survives adjustment:
`r g("dlco_pct_predicted")` per percentage point of predicted DLCO
(p `r gp("dlco_pct_predicted")`), against `r g("fev1_pct_predicted")` for FEV1
(p `r gp("fev1_pct_predicted")`) and `r g("fvc_pct_predicted")` for FVC
(p `r gp("fvc_pct_predicted")`). Age (`r g("age_yrs")`), sex (`r g("sexFemale")`),
current smoking (`r g("current_smokerYes")`), both blood pressures and
haemoglobin are not independently associated. These are associations among the
participants observed, not effects of intervening.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

Three predictors move a long way toward the null and lose significance. **Age**
falls from `r ug("age_yrs")` crude (p `r up("age_yrs")`) to `r g("age_yrs")`
adjusted (p `r gp("age_yrs")`) — almost the whole crude age effect on walking is
carried by lung function and smoking, which both worsen with age. **Current
smoking** falls from `r ug("current_smokerYes")` (p `r up("current_smokerYes")`)
to `r g("current_smokerYes")` (p `r gp("current_smokerYes")`), because current
smokers also have the most pack-years and the worst diffusing capacity; the
cumulative-dose term absorbs it. **FEV1** falls from `r ug("fev1_pct_predicted")`
to `r g("fev1_pct_predicted")` and **FVC** from `r ug("fvc_pct_predicted")` to
`r g("fvc_pct_predicted")`, both losing significance. That is collinearity rather
than confounding: FEV1 and FVC correlate at
`r fmt(cor(dat_model$fev1_pct_predicted, dat_model$fvc_pct_predicted), 2)` in
this sample, so once both are in the model neither can be estimated precisely
and the shared spirometric signal is split between them.

Two estimates move much less. **HIV status** attenuates from
`r ug("hiv_positiveYes")` to `r g("hiv_positiveYes")`, i.e. roughly
`r fmt(100 * (1 - tid$estimate[tid$term == "hiv_positiveYes"] / utid$estimate[utid$term == "hiv_positiveYes"]), 0)`%
of the crude deficit is explained by smoking and lung function, but a deficit of
about `r fmt(abs(tid$estimate[tid$term == "hiv_positiveYes"]), 0)` m remains and
stays significant. **Pack-years** attenuates from `r ug("pack_years")` to
`r g("pack_years")` and also stays significant. Nothing crosses *into*
significance. The overall pattern is that the walk-distance signal in this
dataset is carried by HIV status, cumulative tobacco dose and diffusing capacity,
with the spirometric volumes contributing little once DLCO is present.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r collin}
check_collinearity(fit)
```

```{r assumption-tests}
print(check_heteroscedasticity(fit))
print(check_normality(fit))
```

The diagnostics are largely reassuring. The residual-versus-fitted panel is flat
with no fanning and `check_heteroscedasticity()` does not reject constant
variance (p
`r pv(as.numeric(check_heteroscedasticity(fit)))`), which is
expected for a bounded distance measure. The posterior-predictive check tracks
the observed distribution closely. The normal-QQ panel is the one blemish: a
handful of participants who walked very short distances sit well below the line,
and `check_normality()` rejects normality
(p `r pv(as.numeric(check_normality(fit)))`). With
`r nrow(dat_model)` observations this mild left tail has little practical
consequence for the confidence intervals, and no transformation is indicated
because the outcome itself is close to symmetric.

Collinearity is the check that matters here, and it behaves exactly as
anticipated: the two largest variance inflation factors are
`r fmt(check_collinearity(fit)$VIF[check_collinearity(fit)$Term == "fev1_pct_predicted"], 2)`
for FEV1 and
`r fmt(check_collinearity(fit)$VIF[check_collinearity(fit)$Term == "fvc_pct_predicted"], 2)`
for FVC, three times higher than anything else in the model, because the two are
readings from the same forced expiratory manoeuvre. Both sit just below the
conventional threshold of 5, so the model is estimable, but the practical
consequence is visible in Table 2: the standard errors on the two spirometry
terms are inflated and neither reaches significance despite both having strong
crude associations. Every other VIF is below
`r fmt(max(check_collinearity(fit)$VIF[!check_collinearity(fit)$Term %in% c("fev1_pct_predicted", "fvc_pct_predicted")]), 2)`.
No observation exceeds the Cook's distance contours (maximum
`r fmt(max(cooks.distance(fit)), 3)`), so no single participant drives the fit.

# Discussion

In this two-centre US sample, adults living with HIV walked about
`r fmt(abs(tid$estimate[tid$term == "hiv_positiveYes"]), 0)` m less in six
minutes than HIV-negative adults of the same age, sex, smoking history, blood
pressure, haemoglobin and lung function. That is a clinically meaningful gap:
the minimal important difference for the six-minute walk test in chronic lung
disease is usually put at 25-35 m. Cumulative tobacco exposure and diffusing
capacity are the other independent correlates, and diffusing capacity displaces
both spirometric volumes once it is in the model — the same conclusion the source
paper reaches. Practically, this points at gas transfer rather than airflow
obstruction as the lung mechanism worth measuring when exercise capacity is the
concern, and it says that the HIV deficit is not simply explained by the heavier
smoking histories of people living with HIV.

**Limitations.** Participants were recruited at **two clinical centres,
Pittsburgh and UCSF, and the site variable is not in this deposit**, although the
source paper shows that walk distance differs between them; site is therefore an
unmeasured confounder and also a clustering variable that the model ignores, so
the standard errors are very likely too small. The design is cross-sectional, so
temporality is unknown and reverse causation is plausible for smoking, since
people who cannot walk far may also be the people who continue to smoke.
Adjusting for DLCO is probably adjusting for a *mediator* of the HIV effect
rather than a confounder, which means the adjusted HIV coefficient is a
"direct effect not through measured lung function" and understates the total
association. Nothing about socioeconomic position, comorbidity, body composition,
injecting drug use or musculoskeletal disease is measured, and all of these bear
on how far someone walks. FEV1 and FVC compete for the same variance (VIFs
`r fmt(check_collinearity(fit)$VIF[check_collinearity(fit)$Term == "fev1_pct_predicted"], 1)`
and
`r fmt(check_collinearity(fit)$VIF[check_collinearity(fit)$Term == "fvc_pct_predicted"], 1)`),
so neither spirometric coefficient should be interpreted on its own. Complete-case
analysis dropped `r n_dropped` of `r nrow(dat)` rows
(`r fmt(100 * n_dropped / nrow(dat), 1)`%). Residuals are mildly non-normal in
the left tail. All estimates are associations, not causal effects.

**Future work.** Refit with study site as a fixed effect and with body
composition and comorbidity data from the parent cohort, and separate the total
HIV association from the part mediated by diffusing capacity using a formal
mediation analysis rather than a single adjusted coefficient.

# Feasibility notes

*NOT part of a student report.*

```{r feas-hiv-vars}
# Quantifies what happens when the HIV-specific columns are used as predictors.
base_p <- predictors
dm_base <- dat |> select(all_of(c(outcome, base_p))) |> drop_na()
fit_base <- lm(reformulate(base_p, outcome), data = dm_base)
tb <- tidy(fit_base, conf.int = TRUE)

dmg <- function(v) {
  dm <- dat |> select(all_of(c(outcome, base_p, v))) |> drop_na()
  # lm() errors outright when hiv_positive has collapsed to a single level, which
  # is itself the finding, so the failure is caught and reported rather than fixed.
  f <- try(lm(reformulate(c(base_p, v), outcome), data = dm), silent = TRUE)
  ok <- !inherits(f, "try-error")
  r  <- if (ok) tidy(f, conf.int = TRUE) |> filter(term == "hiv_positiveYes") else NULL
  est_ok <- ok && nrow(r) == 1 && !is.na(r$estimate[1])
  tibble(
    Added = v,
    N = nrow(dm),
    `HIV-negative retained` = sum(dm$hiv_positive == "No"),
    `HIV-positive retained` = sum(dm$hiv_positive == "Yes"),
    `HIV coefficient` = if (est_ok) sprintf("%s (SE %s)", fmt(r$estimate, 1), fmt(r$std.error, 1))
                        else "not estimable",
    `p` = if (est_ok) fmt(r$p.value, 3) else "model fails"
  )
}
r0 <- tibble(Added = "(none - model as fitted)", N = nrow(dm_base),
             `HIV-negative retained` = sum(dm_base$hiv_positive == "No"),
             `HIV-positive retained` = sum(dm_base$hiv_positive == "Yes"),
             `HIV coefficient` = sprintf("%s (SE %s)",
                                         fmt(tb$estimate[tb$term == "hiv_positiveYes"], 1),
                                         fmt(tb$std.error[tb$term == "hiv_positiveYes"], 1)),
             p = fmt(tb$p.value[tb$term == "hiv_positiveYes"], 3))
feas_tab <- bind_rows(r0, map_dfr(c("cd4_count", "on_antiretroviral",
                                    "viral_load_detectable", "mmrc_dyspnoea"), dmg))
feas_tab |>
  gt() |>
  tab_header(title = md("**Feasibility Table F1. Cost of adding an HIV-restricted variable to the model.**"))

n_cd4 <- feas_tab$N[feas_tab$Added == "cd4_count"]
neg_cd4 <- feas_tab$`HIV-negative retained`[feas_tab$Added == "cd4_count"]
n_arv <- feas_tab$N[feas_tab$Added == "on_antiretroviral"]
n_mmrc <- feas_tab$N[feas_tab$Added == "mmrc_dyspnoea"]
```

```{r feas-smoking}
# The alternative smoking parameterisation: the untouched three-level factor
# alongside pack_years, fitted on the same rows for a fair comparison.
alt_pred <- c(setdiff(predictors, "current_smoker"), "smoking_status")
alt_dat  <- dat |> select(all_of(c(outcome, alt_pred))) |> drop_na()
alt_fit  <- lm(reformulate(alt_pred, outcome), data = alt_dat)
```

```{r feas-dichotomy}
# What the <400 m dichotomisation would cost.
dm_bin <- dat_model |> mutate(walk_lt400 = as.integer(six_min_walk_m < 400))
n_ev <- sum(dm_bin$walk_lt400)
gfit <- glm(reformulate(predictors, "walk_lt400"), data = dm_bin, family = binomial)
gt_ <- tidy(gfit, conf.int = TRUE, exponentiate = TRUE)
ghiv <- gt_[gt_$term == "hiv_positiveYes", ]
br <- binned_residuals(gfit)
br_pct <- 100 * mean(br$group == "yes")
```

- **USABLE, and the best of the four outcomes in this dataset.** The outcome is
  complete (`r sum(!is.na(dat$six_min_walk_m))`/`r nrow(dat)`), symmetric,
  clinically meaningful, and the exposure of interest survives adjustment. This
  is the outcome the source paper analysed and it behaves.
- **FATAL if any HIV-restricted variable is used as a predictor.** `cd4_count`,
  `on_antiretroviral` and `viral_load_detectable` exist only for participants
  with HIV. Feasibility Table F1 shows the damage precisely: adding
  `on_antiretroviral` or `viral_load_detectable` leaves **zero HIV-negative
  participants** (N falls from `r nrow(dm_base)` to `r n_arv` and
  `r feas_tab$N[feas_tab$Added == "viral_load_detectable"]` respectively), so
  `hiv_positive` becomes a constant and `lm()` refuses to run at all
  ("contrasts can be applied only to factors with 2 or more levels") — the
  comparison the dataset exists for has been deleted. A student who codes HIV
  status as 0/1 numeric instead of a factor gets no error at all, just a silent
  `NA` coefficient. Adding `cd4_count` is more insidious because it does not
  produce an obvious error: N falls to `r n_cd4` and the HIV-negative group is
  cut from `r sum(dm_base$hiv_positive == "No")` to `r neg_cd4`, the HIV
  coefficient's standard error more than doubles, and the association goes from
  p `r pv(tb$p.value[tb$term == "hiv_positiveYes"])` to
  p = `r feas_tab$p[feas_tab$Added == "cd4_count"]`. A student would report "HIV
  was not associated with walk distance" and be wrong. This is the single most
  likely way this dataset is wrecked and it must be flagged in the brief.
- **MAJOR: `mmrc_dyspnoea` is differentially missing.** All
  `r sum(is.na(dat$mmrc_dyspnoea))` missing values are in HIV-positive
  participants (`r fmt(100 * mean(is.na(dat$mmrc_dyspnoea[dat$hiv_positive == "Yes"])), 0)`%
  of that group, 0% of the HIV-negative group). Adding it drops N to
  `r n_mmrc` and the HIV coefficient loses significance
  (p = `r feas_tab$p[feas_tab$Added == "mmrc_dyspnoea"]`). It is also arguably a
  symptom consequence of low exercise capacity rather than a cause of it. Do not
  suggest it as a predictor.
- **MAJOR: FEV1 and FVC compete.** VIFs
  `r fmt(check_collinearity(fit)$VIF[check_collinearity(fit)$Term == "fev1_pct_predicted"], 2)`
  and
  `r fmt(check_collinearity(fit)$VIF[check_collinearity(fit)$Term == "fvc_pct_predicted"], 2)`
  (r = `r fmt(cor(dat_model$fev1_pct_predicted, dat_model$fvc_pct_predicted), 2)`),
  against a maximum of
  `r fmt(max(check_collinearity(fit)$VIF[!check_collinearity(fit)$Term %in% c("fev1_pct_predicted", "fvc_pct_predicted")]), 2)`
  elsewhere. Below the threshold of 5 so no action is forced, but both crude
  associations (p `r up("fev1_pct_predicted")` and p
  `r up("fvc_pct_predicted")`) disappear on adjustment purely for this reason.
  Students will misread that as "lung volumes don't matter".
- **MINOR: `pack_years` and `smoking_status` are structurally entangled.**
  `pack_years == 0` is exactly `smoking_status == "never"` on all
  `r sum((dat$pack_years == 0) == (dat$smoking_status == "Never"))` of
  `r nrow(dat)` rows. Entering both untouched is estimable — VIFs
  `r fmt(check_collinearity(alt_fit)$VIF[check_collinearity(alt_fit)$Term == "smoking_status"], 2)`
  for `smoking_status` and
  `r fmt(check_collinearity(alt_fit)$VIF[check_collinearity(alt_fit)$Term == "pack_years"], 2)`
  for `pack_years` — but the former/current contrasts are then evaluated at zero
  pack-years, which no ever-smoker has. The parameterisation used here
  (`pack_years` + current-smoker indicator) costs one parameter fewer and gives
  AIC `r fmt(AIC(fit), 1)` against `r fmt(AIC(alt_fit), 1)` for the untouched
  three-level factor, i.e. no practical difference in fit and a cleaner
  interpretation. Either choice is defensible; silently using both without
  comment is not.
- **MINOR: non-normal residuals** (p
  `r pv(as.numeric(check_normality(fit)))`), a left tail from a
  small number of very poor walkers. Variance is constant (p
  `r pv(as.numeric(check_heteroscedasticity(fit)))`) and no
  transformation is warranted.
- **Complete-case N `r nrow(dat_model)` of `r nrow(dat)`**
  (`r fmt(100 * n_dropped / nrow(dat), 1)`% lost), matching the catalogue's
  advertised 410/427 for the >= 95%-complete variable set. Confirmed for this
  predictor set.
- **On the `< 400 m` dichotomisation.** It works, but it is strictly worse. In
  the complete-case sample there are `r n_ev` events
  (`r fmt(100 * n_ev / nrow(dm_bin), 1)`%) for
  `r length(coef(gfit)) - 1L` parameters, i.e.
  `r fmt(n_ev / (length(coef(gfit)) - 1L), 1)` events per parameter — right at
  the conventional minimum, with no room for a further covariate. The HIV odds
  ratio is `r fmt(ghiv$estimate, 2)` (95% CI `r fmt(ghiv$conf.low, 2)` to
  `r fmt(ghiv$conf.high, 2)`, p `r pv(ghiv$p.value)`), i.e. **the association
  that is clearly significant on the linear scale is no longer significant after
  dichotomisation**, which is the standard power loss from throwing away a
  continuous outcome. Binned residuals also fail
  (`r fmt(br_pct, 0)`% inside the bounds against a 95% target). The catalogue is
  right to present this outcome as linear; the logistic version should be
  offered only as an explicit lesson in what dichotomisation costs.
- No separation issues arise (linear model), and no influential observations
  (max Cook's distance `r fmt(max(cooks.distance(fit)), 3)`).
- **Structural: two-site recruitment with no site variable.** The deposit does
  not carry Pittsburgh/UCSF. The model treats `r nrow(dat_model)` participants as
  independent when they come from two centres with different populations, so
  standard errors are optimistic and site confounding cannot be excluded. This is
  unfixable from the shipped file and belongs in every Limitations section.
