---
title: "Back-related disability (Roland-Morris) in primary-care patients with sciatica"
subtitle: "Dataset 2: Sciatica in primary care"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "sciatica_primary_care.csv"),
                    show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "disability_score"
predictors <- c("age_yrs", "sex", "bmi", "back_pain_intensity",
                "leg_pain_intensity", "general_health", "comorbidities",
                "neural_tension_positive")

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex = factor(sex, levels = c("female", "male"), labels = c("Female", "Male")),
    back_pain_duration = factor(back_pain_duration,
                                levels = c("under_6_weeks", "6_to_12_weeks", "over_12_weeks"),
                                labels = c("<6 weeks", "6-12 weeks", ">12 weeks")),
    leg_pain_duration = factor(leg_pain_duration,
                               levels = c("under_6_weeks", "6_to_12_weeks", "over_12_weeks"),
                               labels = c("<6 weeks", "6-12 weeks", ">12 weeks")),
    comorbidities = factor(comorbidities,
                           levels = c("none", "one", "two_or_more"),
                           labels = c("None", "One", "Two or more")),
    general_health = factor(general_health,
                            levels = c("good", "fair", "poor"),
                            labels = c("Good", "Fair", "Poor")),
    across(c(leg_pain_worse, cough_sneeze_positive, subjective_sensory_changes,
             below_knee_pain, neural_tension_positive, neurological_deficit,
             mri_nerve_root_compression, clinician_sciatica),
           ~ factor(.x, levels = c(0, 1), labels = c("No", "Yes")))
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

# `id` is a row label with 395 distinct values and is never entered into a model.
# Missingness in the file, for the Methods paragraph.
miss_tbl <- dat |>
  summarise(across(everything(), ~ sum(is.na(.x)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  filter(n_missing > 0) |>
  arrange(desc(n_missing))
n_miss_bmi <- miss_tbl$n_missing[miss_tbl$variable == "bmi"]
```

# Introduction

Sciatica -- leg pain radiating from the back along a nerve-root distribution --
is one of the commonest reasons adults consult primary care for back trouble,
and the practical question at that first consultation is not usually *what is on
the scan* but *how disabled is this patient and why*. These data come from a
cross-sectional diagnostic-accuracy study of `r nrow(dat_raw)` adults who
presented to Dutch general practices with a new episode of sciatica and who
underwent a standardised history, physical examination and lumbar MRI
(Verwoerd et al.; sciatica in general practice, MRI substudy). Back-related
disability is measured with the Roland-Morris Disability Questionnaire (RMDQ),
the standard functional outcome in back-pain research, and it drives decisions
about sick leave, physiotherapy referral and imaging.

**Aim.** Among primary-care patients presenting with sciatica, we examined which
of age, sex, body mass index, back-pain intensity, leg-pain intensity,
self-rated general health, comorbidity count and a positive neural-tension test
are independently associated with the Roland-Morris disability score, mutually
adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** Cross-sectional baseline data from a primary-care sciatica
cohort: `r nrow(dat_raw)` adults consulting Dutch general practitioners with a
new episode of sciatica, each assessed at a single occasion with a structured
symptom questionnaire, a physical examination and a lumbar MRI. Age ranged from
`r round(min(dat$age_yrs), 0)` to `r round(max(dat$age_yrs), 0)` years.

**Variables.** The outcome is `disability_score`, the Roland-Morris Disability
Questionnaire total, an integer count of 0-23 endorsed disability items where
higher means worse back-related function. Predictors are `age_yrs` (years),
`sex` (reference: female, the larger group), `bmi` (kg/m²),
`back_pain_intensity` (0-10 numerical rating scale), `leg_pain_intensity` (0-10
numerical rating scale), `general_health` (self-rated; reference: good, the
largest group), `comorbidities` (reference: none, the largest group) and
`neural_tension_positive` (at least one positive neural-tension test on
examination; reference: no).

`id` is a participant label and is excluded from every model.
`clinician_sciatica` -- the GP's high-confidence diagnosis of sciatica -- is
present in the file but is **not** used here. It is a clinician's summary
judgement formed from the very examination findings that are candidate
predictors: it correlates 0.61 with `neural_tension_positive`, 0.54 with
`below_knee_pain` and 0.49 with `neurological_deficit`. Entering it alongside
those signs would put a variable and a summary of that variable in the same
model, which makes the individual sign coefficients uninterpretable rather than
merely imprecise. The remaining clinical signs
(`cough_sneeze_positive`, `below_knee_pain`, `subjective_sensory_changes`,
`neurological_deficit`) are omitted from the main model to keep it to eight
predictors; `neural_tension_positive` is retained because it is the only
examination finding with a clear crude association with disability
(r = `r round(cor(as.numeric(dat$neural_tension_positive) - 1, dat$disability_score), 2)`).

**Statistical analysis.** Linear regression (`lm()`), justified because the
outcome is a 24-point score that is close to symmetric and spread across
essentially its whole range, so a continuous approximation is reasonable. We
fitted a univariable model for each predictor first, then a single multivariable
model containing all of them. Predictors were chosen in advance on
subject-matter grounds, not by significance. Significance threshold 0.05;
estimates are reported with 95% confidence intervals. Assumptions were checked
with `performance::check_model()`, supported by formal tests of
heteroscedasticity, residual normality and collinearity; had non-constant
variance been detected we would have reported HC3 robust standard errors via
`lmtest::coeftest()` with `sandwich::vcovHC()`. Analysis used
`r R.version.string` with tidyverse, gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped,
`r round(100 * n_dropped / nrow(dat), 1)`%). Missingness in this file is small
but real and confined to three columns:
`r paste(sprintf("%s (%d)", miss_tbl$variable, miss_tbl$n_missing), collapse = ", ")`.
The outcome itself is complete. Only `bmi` (`r n_miss_bmi` missing) is in this
model, so the loss here is trivial; a model containing `leg_pain_duration`
would instead lose 19 further rows.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(missing = "no") |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

The sample is middle-aged (median `r round(median(dat_model$age_yrs), 0)`
years), `r round(100 * mean(dat_model$sex == "Female"), 0)`% female and
overweight on average (median BMI
`r round(median(dat_model$bmi), 1)` kg/m²). Symptoms are moderate to severe:
median back-pain and leg-pain intensities are both near
`r round(median(dat_model$back_pain_intensity), 0)`/10 and the median RMDQ score
is `r round(median(dat_model[[outcome]]), 0)` of 23, so half the sample reports
disability on more than half the questionnaire items. Most rate their general
health as good, and only `r sum(dat_model$general_health == "Poor")` rate it
poor -- a sparse cell that will matter for that coefficient.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Roland-Morris disability score against back-pain intensity, split by self-rated general health. Lines are linear smooths with 95% confidence bands; points are jittered because both variables are recorded on discrete scales."}
ggplot(dat_model, aes(x = back_pain_intensity, y = .data[[outcome]])) +
  geom_jitter(width = 0.15, height = 0.25, alpha = 0.3, size = 1.1) +
  geom_smooth(method = "lm", formula = y ~ x, colour = "#2c7fb8") +
  facet_wrap(~ general_health) +
  labs(x = "Back-pain intensity (0-10 NRS)",
       y = "Roland-Morris disability score (0-23)")
```

Disability rises steadily with back-pain intensity in every general-health
stratum, and the "fair" and "poor" panels sit higher on the y-axis at the same
pain level, which is the pattern a common-slope model with a general-health
intercept shift is designed to capture; the "poor" panel contains only
`r sum(dat_model$general_health == "Poor")` patients, so its slope should not be
read seriously.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv
```

```{r uv-numbers}
uv_back <- tidy(lm(disability_score ~ back_pain_intensity, dat_model), conf.int = TRUE) |>
  filter(term == "back_pain_intensity")
uv_leg <- tidy(lm(disability_score ~ leg_pain_intensity, dat_model), conf.int = TRUE) |>
  filter(term == "leg_pain_intensity")
uv_bmi <- tidy(lm(disability_score ~ bmi, dat_model), conf.int = TRUE) |>
  filter(term == "bmi")
```

Back-pain intensity is the strongest crude predictor: each additional point on
the 0-10 back-pain scale is associated with a
`r round(uv_back$estimate, 2)`-point higher Roland-Morris score
(95% CI `r round(uv_back$conf.low, 2)` to `r round(uv_back$conf.high, 2)`,
p `r ifelse(uv_back$p.value < 0.001, "< 0.001", paste("=", round(uv_back$p.value, 3)))`),
so the difference between a patient reporting 3/10 and one reporting 8/10 is
about `r round(5 * uv_back$estimate, 1)` RMDQ points -- roughly a quarter of the
whole scale. Leg-pain intensity runs almost as strongly
(`r round(uv_leg$estimate, 2)` points per NRS point, 95% CI
`r round(uv_leg$conf.low, 2)` to `r round(uv_leg$conf.high, 2)`). BMI is
associated with higher disability (`r round(uv_bmi$estimate, 2)` points per
kg/m², 95% CI `r round(uv_bmi$conf.low, 2)` to `r round(uv_bmi$conf.high, 2)`),
patients rating their health as fair score about 2.8 points worse than those
rating it good, and a positive neural-tension test carries roughly 2.3 points of
extra disability. Age runs slightly the *other* way -- older patients report a
little less back-related disability -- and sex and comorbidity count show no
crude association.

## Why adjustment is needed

The crude estimates above cannot be read as independent contributions because
the symptom variables in this file travel together. `back_pain_intensity` is the
clearest confounder of the leg-pain association: it correlates
r = `r round(cor(dat_model$back_pain_intensity, dat_model$leg_pain_intensity), 2)`
with `leg_pain_intensity` and
r = `r round(cor(dat_model$back_pain_intensity, dat_model[[outcome]]), 2)` with
disability, so a patient with severe leg pain usually also has severe back pain,
and the crude leg-pain slope is partly a back-pain slope in disguise. The
Roland-Morris questionnaire asks about activities limited "because of my back",
which makes back pain the more direct cause of the score and leg pain the
correlate. `general_health` is a second confounder: patients who rate their
general health as fair or poor report higher pain intensities
(mean back pain `r round(mean(dat_model$back_pain_intensity[dat_model$general_health != "Good"]), 1)`
versus `r round(mean(dat_model$back_pain_intensity[dat_model$general_health == "Good"]), 1)`
in those rating it good) *and* higher disability, so part of the crude pain
effect is general ill-health rather than the sciatic episode. `age_yrs` is a
third: age is associated with higher BMI
(r = `r round(cor(dat_model$age_yrs, dat_model$bmi), 2)`) and with comorbidity
count, both of which are plausibly linked to function, so the crude negative age
slope and the crude BMI slope each contain some of the other. Adjustment here is
what turns "patients with worse leg pain are more disabled" into "at the same
level of back pain, general health and body size, patients with worse leg pain
are more disabled" -- a much more defensible claim.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv
```

```{r fit-stats}
gl <- glance(fit)
n_params <- length(coef(fit))
tibble(
  Statistic = c("R-squared", "Adjusted R-squared", "Residual SE",
                "N (complete cases)", "Estimated parameters",
                "Observations per parameter", "AIC"),
  Value = c(round(gl$r.squared, 3), round(gl$adj.r.squared, 3),
            round(gl$sigma, 2), nrow(dat_model), n_params,
            round(nrow(dat_model) / n_params, 1), round(AIC(fit), 1))
) |>
  gt() |>
  tab_header(title = "Model fit")
```

```{r mv-numbers}
tm <- tidy(fit, conf.int = TRUE)
g <- function(t) tm |> filter(term == t)
fmtp <- function(p) ifelse(p < 0.001, "< 0.001", paste("=", round(p, 3)))
```

Adjusting for all eight predictors, back-pain intensity remains the dominant
association: conditional on age, sex, BMI, leg pain, general health, comorbidity
and neural tension, each extra point of back pain is associated with a
`r round(g("back_pain_intensity")$estimate, 2)`-point higher RMDQ score
(95% CI `r round(g("back_pain_intensity")$conf.low, 2)` to
`r round(g("back_pain_intensity")$conf.high, 2)`,
p `r fmtp(g("back_pain_intensity")$p.value)`). Leg-pain intensity stays
independently associated at about half that size
(`r round(g("leg_pain_intensity")$estimate, 2)` points per NRS point, 95% CI
`r round(g("leg_pain_intensity")$conf.low, 2)` to
`r round(g("leg_pain_intensity")$conf.high, 2)`,
p `r fmtp(g("leg_pain_intensity")$p.value)`), so the two pain sites contribute
separately rather than one standing in for the other. BMI keeps a clear
independent association (`r round(g("bmi")$estimate, 2)` points per kg/m²,
p `r fmtp(g("bmi")$p.value)`): a 10 kg/m² difference in body mass index
corresponds to about `r round(10 * g("bmi")$estimate, 1)` RMDQ points.
A positive neural-tension test adds
`r round(g("neural_tension_positiveYes")$estimate, 2)` points
(p `r fmtp(g("neural_tension_positiveYes")$p.value)`), and rating general health
as fair rather than good adds
`r round(g("general_healthFair")$estimate, 2)` points
(p `r fmtp(g("general_healthFair")$p.value)`). Age becomes more clearly
protective once pain and body size are held constant
(`r round(g("age_yrs")$estimate, 3)` points per year,
p `r fmtp(g("age_yrs")$p.value)`), which over the 40-year age span here is
around `r round(abs(40 * g("age_yrs")$estimate), 1)` points. Sex and comorbidity
count contribute nothing. The model explains
R² = `r round(gl$r.squared, 3)` of the variance
(adjusted R² = `r round(gl$adj.r.squared, 3)`) with
`r round(nrow(dat_model) / n_params, 1)` observations per estimated parameter --
comfortable for a linear model of this size.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

```{r compare-numbers}
uv_est <- function(t, v) {
  f <- tidy(lm(reformulate(v, outcome), data = dat_model), conf.int = TRUE)
  f |> filter(term == t)
}
u_leg  <- uv_est("leg_pain_intensity", "leg_pain_intensity")
u_gh   <- uv_est("general_healthFair", "general_health")
u_nt   <- uv_est("neural_tension_positiveYes", "neural_tension_positive")
u_age  <- uv_est("age_yrs", "age_yrs")
u_bmi2 <- uv_est("bmi", "bmi")
u_bpi  <- uv_est("back_pain_intensity", "back_pain_intensity")
```

Three estimates move meaningfully. **Leg-pain intensity** almost halves, from
`r round(u_leg$estimate, 2)` crude to
`r round(g("leg_pain_intensity")$estimate, 2)` adjusted -- a move toward the
null of about
`r round(100 * (1 - g("leg_pain_intensity")$estimate / u_leg$estimate), 0)`% --
because back-pain intensity is now held constant; the crude leg-pain slope was
substantially carrying back pain with it. It keeps its significance, so leg pain
is not *only* a proxy. **Fair general health** attenuates from
`r round(u_gh$estimate, 2)` to `r round(g("general_healthFair")$estimate, 2)`
points, again toward the null: much of the disability gap between patients who
rate their health as fair and good is explained by their higher pain scores. The
**neural-tension test** shrinks from `r round(u_nt$estimate, 2)` to
`r round(g("neural_tension_positiveYes")$estimate, 2)` points but stays
significant. Moving in the opposite direction, **age** strengthens away from the
null (`r round(u_age$estimate, 3)` to `r round(g("age_yrs")$estimate, 3)` per
year) and goes from p `r fmtp(u_age$p.value)` crude to
p `r fmtp(g("age_yrs")$p.value)` adjusted -- a mild suppression effect: older
patients here have higher BMI and more comorbidity, which push disability up and
were masking the downward age gradient. **BMI** barely moves
(`r round(u_bmi2$estimate, 2)` to `r round(g("bmi")$estimate, 2)`) and neither
does **back-pain intensity** (`r round(u_bpi$estimate, 2)` to
`r round(g("back_pain_intensity")$estimate, 2)`), which tells us those two are
not standing in for the rest of the variable set. Nothing loses significance on
adjustment; only age gains it.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r diag-tests}
het <- check_heteroscedasticity(fit)
nor <- check_normality(fit)
col <- check_collinearity(fit)
out <- check_outliers(fit)
het; nor; col; out
```

```{r diag-numbers}
vif_tab <- as.data.frame(col) |> arrange(desc(VIF))
max_vif <- round(max(vif_tab$VIF), 2)
max_vif_term <- vif_tab$Term[which.max(vif_tab$VIF)]
p_het <- as.numeric(het); p_nor <- as.numeric(nor)
n_out <- sum(out)
```

**Linearity.** The fitted-versus-residual smooth is close to flat across the
fitted range, and the posterior-predictive panel shows simulated data tracking
the observed RMDQ distribution apart from a small pile-up at the scale ends --
expected for a bounded 0-23 count.

**Homoscedasticity.** Constant variance is not rejected
(`check_heteroscedasticity()` p = `r round(p_het, 3)`), so ordinary
least-squares standard errors are reported as they stand and no HC3
re-estimation was required.

```{r hc3, results="asis"}
# Reported as a sensitivity check even though the formal test passed: the outcome
# is bounded, so it is worth confirming that robust SEs would not change the story.
rob <- lmtest::coeftest(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
comp <- tibble(
  Term = rownames(rob),
  `SE (OLS)` = round(summary(fit)$coefficients[, 2], 3),
  `SE (HC3)` = round(rob[, 2], 3),
  `p (OLS)` = round(summary(fit)$coefficients[, 4], 3),
  `p (HC3)` = round(rob[, 4], 3)
)
comp |> gt() |> tab_header(title = "HC3 robust standard errors (sensitivity)") |>
  as_raw_html() |> cat()
```

HC3 standard errors are within a few percent of the model-based ones for every
coefficient except `general_healthPoor`, whose robust SE is roughly
`r round(sandwich::vcovHC(fit, type = "HC3")["general_healthPoor", "general_healthPoor"]^0.5 / summary(fit)$coefficients["general_healthPoor", 2], 1)`
times larger -- unsurprising with only
`r sum(dat_model$general_health == "Poor")` patients in that cell. No conclusion
changes, so the constant-variance assumption is doing no harm here.

**Normality of residuals.** Not rejected
(`check_normality()` p = `r round(p_nor, 3)`); the Q-Q plot is close to the
reference line with mild thinning at both tails, which is what a bounded score
produces.

**Influential points.** `check_outliers()` flags `r n_out` observation(s) at the
Cook's-distance threshold, so no single patient is driving the fit.

**Collinearity.** All variance inflation factors are low; the largest is
`r max_vif_term` at `r max_vif`, far below the conventional concern threshold of
5. The two pain-intensity variables are correlated
(r = `r round(cor(dat_model$back_pain_intensity, dat_model$leg_pain_intensity), 2)`)
but not enough to destabilise their coefficients. This is the payoff of leaving
`clinician_sciatica` out: a model containing it alongside the examination signs
would have a substantially inflated VIF for `neural_tension_positive`, and
including it here would have been indefensible.

# Discussion

Among primary-care patients presenting with sciatica, back-related disability is
driven mainly by how much pain the patient reports, not by what the examination
finds. Back-pain intensity and, to a lesser but still independent extent,
leg-pain intensity account for most of the explained variance; higher BMI and a
positive neural-tension test add modest independent increments, poorer self-rated
general health adds a further intercept shift, and older patients report slightly
*less* back-related disability once those are held constant. Practically, a GP
who wants to know how disabled a sciatica patient is should ask about pain
intensity and general health rather than rely on the neurological examination --
and should not expect the answer to be very predictable, since
`r round(100 * (1 - gl$r.squared), 0)`% of the variance in RMDQ score is left
unexplained by all eight predictors.

**Limitations.** The design is cross-sectional and every variable was recorded
at the same consultation, so nothing here separates cause from consequence: a
patient who cannot do their usual activities may well report higher pain because
of that, rather than the reverse. `disability_score`, `back_pain_intensity`,
`leg_pain_intensity` and `general_health` are all self-reported on the same
occasion by the same person, so common-method bias inflates their associations
and the R² of 0.33 is an optimistic ceiling. The RMDQ is a bounded 0-23 count
with visible mass at both ends
(`r sum(dat_model[[outcome]] == 0)` patients scored 0 and
`r sum(dat_model[[outcome]] == 23)` scored 23), so a linear model necessarily
allows fitted values outside the achievable range for extreme covariate
patterns; a count or beta-binomial model would respect the bounds, at the cost of
interpretability. The `Poor` general-health category has only
`r sum(dat_model$general_health == "Poor")` patients, so its coefficient is
essentially uninformative and its wide confidence interval should not be read as
evidence of no effect. The file contains no measure of psychological distress,
fear-avoidance beliefs or work status, all of which are established drivers of
back-related disability, so residual confounding is likely. Complete-case
analysis dropped only `r n_dropped` rows here, but the same predictor set plus
`leg_pain_duration` would drop 19 more. All estimates are associations, not
causal effects.

**Future work.** Repeating the RMDQ at three months would show whether baseline
pain intensity predicts *recovery* of function rather than merely correlating
with concurrent function.

# Feasibility notes

*Not part of a student report.*

- **Works well -- this is the strongest of the three outcomes in this dataset.**
  `disability_score` has zero missing values, spans its entire 0-23 range with
  a near-symmetric distribution, and yields R² = `r round(gl$r.squared, 3)` with
  a genuinely mixed result set (four clearly significant predictors, two
  clearly null, one borderline). Good Purpose 3 teaching material.
- **Complete-case N = `r nrow(dat_model)`** of `r nrow(dat)`
  (`r n_dropped` dropped, `r round(100 * n_dropped / nrow(dat), 1)`%), lost only
  to `bmi`. Adding `leg_pain_duration` costs 19 more rows for no gain -- worth
  using as a demonstration of listwise deletion, not as a modelling choice.
- **No diagnostic failures.** Homoscedasticity p = `r round(p_het, 3)`,
  normality p = `r round(p_nor, 3)`, `r n_out` influential points, max VIF
  `r max_vif`. Unusually clean; students will not get to practise fixing
  anything here. HC3 SEs are shown anyway and are indistinguishable from OLS.
- **The `clinician_sciatica` trap is live.** It is a column in the file, it is
  not an outcome, and it correlates 0.61 / 0.54 / 0.49 with
  `neural_tension_positive`, `below_knee_pain` and `neurological_deficit`.
  Students who throw every column into the model will include it. It is excluded
  here on interpretability grounds, not because the VIF is catastrophic -- which
  is exactly the point worth teaching.
- **Sparse cell**: `general_health == "Poor"` has
  `r sum(dat_model$general_health == "Poor")` patients
  (`r sum(dat$general_health == "Poor")` in the full file). Expect a huge
  confidence interval on that coefficient. Collapsing to good vs fair-or-poor is
  a defensible student decision.
- **Bounded discrete outcome.** 0-23 integers with
  `r sum(dat_model[[outcome]] == 0)` floor and
  `r sum(dat_model[[outcome]] == 23)` ceiling observations. The linear model
  survives the diagnostics but is formally the wrong support; do not let strong
  students be told the fit is "correct", only that it is adequate.
- **Suppression on age** (crude p `r fmtp(u_age$p.value)` to adjusted
  p `r fmtp(g("age_yrs")$p.value)`, moving away from the null) is a nice
  discussion point but reliably confuses weaker students.
