---
title: "Nerve-root compression on MRI in primary-care patients with sciatica"
subtitle: "Dataset 2: Sciatica in primary care"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "sciatica_primary_care.csv"),
                    show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "mri_nerve_root_compression"
predictors <- c("age_yrs", "sex", "bmi", "leg_pain_worse",
                "cough_sneeze_positive", "subjective_sensory_changes",
                "below_knee_pain", "neural_tension_positive",
                "neurological_deficit")

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex = factor(sex, levels = c("female", "male"), labels = c("Female", "Male")),
    back_pain_duration = factor(back_pain_duration,
                                levels = c("under_6_weeks", "6_to_12_weeks", "over_12_weeks"),
                                labels = c("<6 weeks", "6-12 weeks", ">12 weeks")),
    leg_pain_duration = factor(leg_pain_duration,
                               levels = c("under_6_weeks", "6_to_12_weeks", "over_12_weeks"),
                               labels = c("<6 weeks", "6-12 weeks", ">12 weeks")),
    comorbidities = factor(comorbidities,
                           levels = c("none", "one", "two_or_more"),
                           labels = c("None", "One", "Two or more")),
    general_health = factor(general_health,
                            levels = c("good", "fair", "poor"),
                            labels = c("Good", "Fair", "Poor")),
    across(c(leg_pain_worse, cough_sneeze_positive, subjective_sensory_changes,
             below_knee_pain, neural_tension_positive, neurological_deficit,
             mri_nerve_root_compression, clinician_sciatica),
           ~ factor(.x, levels = c(0, 1), labels = c("No", "Yes")))
  )

# `clinician_sciatica` is kept in dat_model only so that the collinearity
# demonstration below runs on identical rows; it is NOT in `predictors`.
dat_model <- dat |>
  select(all_of(c(outcome, predictors, "clinician_sciatica"))) |>
  drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

miss_tbl <- dat |>
  summarise(across(everything(), ~ sum(is.na(.x)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  filter(n_missing > 0) |>
  arrange(desc(n_missing))

n_event  <- sum(dat_model[[outcome]] == "Yes")
n_nonev  <- sum(dat_model[[outcome]] == "No")
n_rarer  <- min(n_event, n_nonev)

signs <- c("leg_pain_worse", "cough_sneeze_positive", "subjective_sensory_changes",
           "below_knee_pain", "neural_tension_positive", "neurological_deficit")
num01 <- function(x) as.numeric(x) - 1
cor_tab <- tibble(
  sign = signs,
  r_mri = map_dbl(signs, ~ cor(num01(dat[[.x]]), num01(dat[[outcome]]))),
  r_clin = map_dbl(signs, ~ cor(num01(dat[[.x]]), num01(dat$clinician_sciatica)))
)
```

# Introduction

The clinical question behind this dataset is a diagnostic one: when a GP sees a
patient with leg pain and finds the classical signs of sciatica, how well do
those signs predict what the scan will actually show? These data come from a
cross-sectional study of `r nrow(dat_raw)` adults presenting to Dutch general
practices with a new episode of sciatica, all of whom had a standardised history,
a physical examination and a lumbar MRI regardless of the clinician's suspicion
(Verwoerd et al.; sciatica in general practice, MRI substudy). Because the MRI
was done in everyone, the usual verification bias that plagues diagnostic studies
is absent, and the signs can be assessed against the imaging finding they are
supposed to detect.

**Aim.** Among primary-care patients presenting with sciatica, we examined which
of age, sex, body mass index and five history and examination findings
(leg pain worse than back pain, pain reproduced by coughing or sneezing,
subjective sensory changes, pain radiating below the knee, a positive
neural-tension test and a neurological deficit) are independently associated with
nerve-root compression on MRI, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** Cross-sectional data from a primary-care sciatica cohort:
`r nrow(dat_raw)` adults consulting Dutch general practitioners with a new
episode of sciatica, each assessed at a single occasion with a structured
history, a physical examination and a lumbar MRI. Age ranged from
`r round(min(dat$age_yrs), 0)` to `r round(max(dat$age_yrs), 0)` years.

**Variables.** The outcome is `mri_nerve_root_compression`, coded 1 when
nerve-root compression is present on MRI; **the modelled event is presence of
compression**, which occurs in `r n_event` of `r nrow(dat_model)` analysed
patients (`r round(100 * n_event / nrow(dat_model), 1)`%). Predictors are
`age_yrs` (years), `sex` (reference: female, the larger group), `bmi` (kg/m²),
`leg_pain_worse` (patient reports leg pain worse than back pain; reference: no),
`cough_sneeze_positive` (coughing or sneezing reproduces the leg pain;
reference: no), `subjective_sensory_changes` (reference: no), `below_knee_pain`
(pain radiates below the knee; reference: no), `neural_tension_positive` (at
least one positive neural-tension test; reference: no) and
`neurological_deficit` (at least one deficit on examination; reference: no).
Every binary predictor takes "no / sign absent" as its reference so that odds
ratios above 1 mean the sign is associated with compression.

**`clinician_sciatica` is deliberately excluded from the main model.** It is the
GP's own high-confidence diagnosis of sciatica -- not an independent
observation, but a summary judgement formed from the very signs listed above. In
this file it correlates
r = `r round(cor_tab$r_clin[cor_tab$sign == "neural_tension_positive"], 2)` with
`neural_tension_positive`,
r = `r round(cor_tab$r_clin[cor_tab$sign == "below_knee_pain"], 2)` with
`below_knee_pain` and
r = `r round(cor_tab$r_clin[cor_tab$sign == "neurological_deficit"], 2)` with
`neurological_deficit`. Entering it alongside those signs would ask each sign to
explain variation in the outcome that its own summary has already absorbed; the
coefficients would become answers to a question nobody asked. A sensitivity
model containing it, with the resulting variance inflation factors, is reported
in the Diagnostics section rather than being used for inference. `id` is a row
label and is excluded from every model.

**Statistical analysis.** Logistic regression (`glm(family = binomial)`),
justified because the outcome is binary. We fitted a univariable model for each
predictor first, then a single multivariable model containing all of them.
Predictors were chosen in advance on subject-matter grounds, not by significance.
Estimates are reported as odds ratios with 95% confidence intervals;
significance threshold 0.05. Model fit is summarised with AIC, Tjur's R² and the
number of events per estimated parameter. Assumptions were checked with
`performance::check_model()`, focusing on binned residuals, collinearity and
influential observations. Analysis used `r R.version.string` with tidyverse,
gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped,
`r round(100 * n_dropped / nrow(dat), 1)`%). Missingness in this file is small
but real and confined to
`r paste(sprintf("%s (%d)", miss_tbl$variable, miss_tbl$n_missing), collapse = ", ")`.
The outcome and all six clinical signs are complete; the only loss here is from
`bmi`. Had `leg_pain_duration` been included the analysed sample would have
fallen to `r nrow(drop_na(dat[, c(outcome, predictors, "leg_pain_duration")]))`.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  select(-clinician_sciatica) |>
  tbl_summary(by = all_of(outcome), missing = "no") |>
  add_n() |>
  add_overall() |>
  modify_caption("**Table 1. Sample characteristics by MRI nerve-root compression status.**")
```

Compression is present in `r round(100 * n_event / nrow(dat_model), 1)`% of the
analysed sample, so neither class is rare: the smaller class contains
`r n_rarer` patients. The most striking contrast in Table 1 is not a clinical
sign at all but sex --
`r round(100 * mean(dat_model[[outcome]][dat_model$sex == "Male"] == "Yes"), 0)`%
of men have compression versus
`r round(100 * mean(dat_model[[outcome]][dat_model$sex == "Female"] == "Yes"), 0)`%
of women. The clinical signs differ far less between the two MRI groups than one
would expect: for example
`r round(100 * mean(dat_model$below_knee_pain[dat_model[[outcome]] == "Yes"] == "Yes"), 0)`%
of patients with compression report pain below the knee, compared with
`r round(100 * mean(dat_model$below_knee_pain[dat_model[[outcome]] == "No"] == "Yes"), 0)`%
of those without.

## Exploratory figure

```{r explore, fig.height=5.5, fig.cap="Figure 1. Proportion of patients with nerve-root compression on MRI, by each clinical sign and by sex. Bars are proportions with 95% Wilson confidence intervals; the dashed line is the overall prevalence of compression. Numbers above the bars are group sizes."}
prop_ci <- function(x, n) {
  z <- 1.96; p <- x / n
  c(lo = (p + z^2 / (2 * n) - z * sqrt(p * (1 - p) / n + z^2 / (4 * n^2))) / (1 + z^2 / n),
    hi = (p + z^2 / (2 * n) + z * sqrt(p * (1 - p) / n + z^2 / (4 * n^2))) / (1 + z^2 / n))
}

plot_dat <- dat_model |>
  select(all_of(c(outcome, signs, "sex"))) |>
  mutate(sex = factor(sex, levels = c("Female", "Male"), labels = c("No", "Yes"))) |>
  pivot_longer(-all_of(outcome), names_to = "predictor", values_to = "level") |>
  group_by(predictor, level) |>
  summarise(n = n(), k = sum(.data[[outcome]] == "Yes"), .groups = "drop") |>
  mutate(p = k / n,
         lo = map2_dbl(k, n, ~ prop_ci(.x, .y)["lo"]),
         hi = map2_dbl(k, n, ~ prop_ci(.x, .y)["hi"]),
         predictor = recode(predictor, sex = "sex (Yes = male)"))

ggplot(plot_dat, aes(x = level, y = p)) +
  geom_hline(yintercept = mean(dat_model[[outcome]] == "Yes"),
             linetype = "dashed", colour = "grey50") +
  geom_col(fill = "#2c7fb8", width = 0.6) +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.15) +
  geom_text(aes(y = hi + 0.06, label = n), size = 3) +
  facet_wrap(~ predictor, nrow = 2) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(x = "Sign present?", y = "Nerve-root compression on MRI (%)")
```

Only the sex panel shows a wide separation from the dashed overall prevalence,
with `leg_pain_worse` and `cough_sneeze_positive` a distant second and third,
while `subjective_sensory_changes` and
`neural_tension_positive`, the signs a clinician would weight most heavily,
barely move the proportion at all. This is the central finding of the analysis
and the reason the model that follows explains so little.

## Univariable models

```{r univariable}
uv <- dat_model |>
  select(all_of(c(outcome, predictors))) |>
  tbl_uvregression(
    y = all_of(outcome),
    method = glm,
    method.args = list(family = binomial),
    exponentiate = TRUE
  )
uv
```

```{r uv-numbers}
uv_fit <- function(v) {
  tidy(glm(reformulate(v, outcome), data = dat_model, family = binomial),
       conf.int = TRUE, exponentiate = TRUE)
}
fmtp <- function(p) ifelse(p < 0.001, "< 0.001", paste("=", round(p, 3)))
u <- function(term, v) uv_fit(v) |> filter(.data$term == !!term)
u_sex <- u("sexMale", "sex")
u_lpw <- u("leg_pain_worseYes", "leg_pain_worse")
u_cs  <- u("cough_sneeze_positiveYes", "cough_sneeze_positive")
u_nd  <- u("neurological_deficitYes", "neurological_deficit")
u_bkp <- u("below_knee_painYes", "below_knee_pain")
u_nt  <- u("neural_tension_positiveYes", "neural_tension_positive")
u_ssc <- u("subjective_sensory_changesYes", "subjective_sensory_changes")
```

Sex is the strongest crude predictor of what the scan shows: men have
`r round(u_sex$estimate, 2)` times the odds of nerve-root compression compared
with women (95% CI `r round(u_sex$conf.low, 2)` to `r round(u_sex$conf.high, 2)`,
p `r fmtp(u_sex$p.value)`), a substantially larger effect than any clinical sign.
Among the signs, reporting leg pain as worse than back pain carries an odds
ratio of `r round(u_lpw$estimate, 2)` (95% CI `r round(u_lpw$conf.low, 2)` to
`r round(u_lpw$conf.high, 2)`, p `r fmtp(u_lpw$p.value)`), pain reproduced by
coughing or sneezing `r round(u_cs$estimate, 2)`
(p `r fmtp(u_cs$p.value)`), a neurological deficit `r round(u_nd$estimate, 2)`
(p `r fmtp(u_nd$p.value)`), pain below the knee `r round(u_bkp$estimate, 2)`
(p `r fmtp(u_bkp$p.value)`) and a positive neural-tension test
`r round(u_nt$estimate, 2)` (p `r fmtp(u_nt$p.value)`). Subjective sensory
changes are not associated at all (OR `r round(u_ssc$estimate, 2)`,
p `r fmtp(u_ssc$p.value)`).

```{r sign-gap}
cor_show <- cor_tab |>
  mutate(across(c(r_mri, r_clin), ~ round(.x, 2))) |>
  rename(`Clinical sign` = sign,
         `r with MRI compression` = r_mri,
         `r with clinician diagnosis` = r_clin)
cor_show |> gt() |>
  tab_header(title = "The signs predict the clinician, not the scan",
             subtitle = "Point-biserial correlations, full file (n = 395)")
```

```{r tjur-gap}
mri_signs <- glm(reformulate(signs, outcome), data = dat_model, family = binomial)
clin_signs <- glm(reformulate(signs, "clinician_sciatica"), data = dat_model,
                  family = binomial)
r2_mri  <- as.numeric(r2_tjur(mri_signs))
r2_clin <- as.numeric(r2_tjur(clin_signs))
```

The table above is worth pausing on, because it is the most interesting thing in
this dataset. Each of the six signs correlates only weakly with the MRI finding
(r between `r round(min(cor_tab$r_mri), 2)` and
`r round(max(cor_tab$r_mri), 2)`), yet the same signs correlate strongly with the
clinician's diagnosis of sciatica (up to
r = `r round(max(cor_tab$r_clin), 2)`). Fitting the same six signs to the two
different outcomes makes the size of the gap explicit: they explain Tjur's
R² = `r round(r2_clin, 2)` of the clinician's judgement but only
R² = `r round(r2_mri, 2)` of the MRI finding. The examination is highly
reproducible *as a description of what the clinician concludes* and only weakly
informative *about what the scan shows*. That gap is the finding, and it is a
better research question than any single coefficient in the model below.

## Why adjustment is needed

The crude odds ratios cannot be read as independent contributions because the
signs cluster within patients and because both the signs and the outcome vary
with basic demographics. `sex` is the clearest confounder: men are more likely to
have compression on MRI
(`r round(100 * mean(dat_model[[outcome]][dat_model$sex == "Male"] == "Yes"), 0)`%
versus
`r round(100 * mean(dat_model[[outcome]][dat_model$sex == "Female"] == "Yes"), 0)`%)
and they also present differently -- for example
`r round(100 * mean(dat_model$leg_pain_worse[dat_model$sex == "Male"] == "Yes"), 0)`%
of men report leg pain worse than back pain versus
`r round(100 * mean(dat_model$leg_pain_worse[dat_model$sex == "Female"] == "Yes"), 0)`%
of women -- so an unadjusted sign estimate partly measures the sex mix of the
patients who have that sign. `below_knee_pain` is a second confounder, this time
of the examination findings: it correlates
r = `r round(cor(num01(dat_model$below_knee_pain), num01(dat_model$neural_tension_positive)), 2)`
with `neural_tension_positive` and
r = `r round(cor(num01(dat_model$below_knee_pain), num01(dat_model$neurological_deficit)), 2)`
with `neurological_deficit`, and it is itself associated with compression, so
patients with a positive neural-tension test are simply more likely to be
patients whose pain radiates below the knee. `age_yrs` is a third: older patients
are more likely to show degenerative nerve-root compression
(r = `r round(cor(dat_model$age_yrs, num01(dat_model[[outcome]])), 2)`) and less
likely to report cough-provoked pain
(r = `r round(cor(dat_model$age_yrs, num01(dat_model$cough_sneeze_positive)), 2)`),
so age pushes the crude cough-sneeze estimate around. Mutual adjustment is what
lets us ask whether any single sign carries information about the scan that the
other signs and the patient's demographics do not already provide.

## Multivariable model

```{r multivariable}
fit <- glm(reformulate(predictors, outcome), data = dat_model, family = binomial)

mv <- tbl_regression(fit, exponentiate = TRUE)
mv
```

```{r fit-stats}
n_params <- length(coef(fit))
epp <- n_rarer / n_params
tibble(
  Statistic = c("AIC", "Null deviance", "Residual deviance", "Tjur's R-squared",
                "N (complete cases)", "Events (compression present)",
                "Non-events", "Rarer class", "Estimated parameters",
                "Events per parameter (rarer class)"),
  Value = c(round(AIC(fit), 1), round(fit$null.deviance, 1),
            round(fit$deviance, 1), round(as.numeric(r2_tjur(fit)), 3),
            nrow(dat_model), n_event, n_nonev, n_rarer, n_params,
            round(epp, 1))
) |>
  gt() |>
  tab_header(title = "Model fit")
```

```{r mv-numbers}
tm <- tidy(fit, conf.int = TRUE, exponentiate = TRUE)
g <- function(t) tm |> filter(term == t)
auc_val <- as.numeric(performance_roc(fit))
```

Adjusting for all nine predictors, four remain independently associated with
nerve-root compression. Sex is by far the strongest: conditional on age, BMI and
all six clinical signs, men have `r round(g("sexMale")$estimate, 2)` times the
odds of compression (95% CI `r round(g("sexMale")$conf.low, 2)` to
`r round(g("sexMale")$conf.high, 2)`, p `r fmtp(g("sexMale")$p.value)`). Pain
reproduced by coughing or sneezing carries an odds ratio of
`r round(g("cough_sneeze_positiveYes")$estimate, 2)` (95% CI
`r round(g("cough_sneeze_positiveYes")$conf.low, 2)` to
`r round(g("cough_sneeze_positiveYes")$conf.high, 2)`,
p `r fmtp(g("cough_sneeze_positiveYes")$p.value)`), leg pain worse than back pain
`r round(g("leg_pain_worseYes")$estimate, 2)`
(p `r fmtp(g("leg_pain_worseYes")$p.value)`) and a neurological deficit
`r round(g("neurological_deficitYes")$estimate, 2)`
(p `r fmtp(g("neurological_deficitYes")$p.value)`). The two signs that a
clinician would probably weight most heavily contribute nothing once the others
are held constant: `neural_tension_positive` has an adjusted odds ratio of
`r round(g("neural_tension_positiveYes")$estimate, 2)`
(95% CI `r round(g("neural_tension_positiveYes")$conf.low, 2)` to
`r round(g("neural_tension_positiveYes")$conf.high, 2)`,
p `r fmtp(g("neural_tension_positiveYes")$p.value)`) -- an odds ratio of
essentially exactly 1 -- and `subjective_sensory_changes`
`r round(g("subjective_sensory_changesYes")$estimate, 2)`
(p `r fmtp(g("subjective_sensory_changesYes")$p.value)`). `below_knee_pain`,
`age_yrs` and `bmi` are also not independently associated.

Fit is modest. AIC is `r round(AIC(fit), 1)` against a null deviance of
`r round(fit$null.deviance, 1)` and residual deviance
`r round(fit$deviance, 1)`, Tjur's R² is
`r round(as.numeric(r2_tjur(fit)), 3)`, and the area under the ROC curve is
`r round(auc_val, 3)` -- better than chance but nowhere near a usable diagnostic
rule. Sample size is not the constraint: with `r n_rarer` patients in the rarer
class and `r n_params` estimated parameters there are `r round(epp, 1)` events
per parameter, comfortably above the conventional minimum of 10, so the wide
confidence intervals on the null signs reflect genuinely weak associations rather
than an underpowered model.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted odds ratios.**")
```

Two estimates move dramatically toward the null and lose significance. The
**neural-tension test** falls from a crude odds ratio of
`r round(u_nt$estimate, 2)` (p `r fmtp(u_nt$p.value)`) to
`r round(g("neural_tension_positiveYes")$estimate, 2)`
(p `r fmtp(g("neural_tension_positiveYes")$p.value)`) -- essentially the whole
association disappears. **Pain below the knee** falls from
`r round(u_bkp$estimate, 2)` (p `r fmtp(u_bkp$p.value)`) to
`r round(g("below_knee_painYes")$estimate, 2)`
(p `r fmtp(g("below_knee_painYes")$p.value)`). Both are explained by the same
thing: these two signs travel with the other signs (`below_knee_pain` correlates
`r round(cor(num01(dat_model$below_knee_pain), num01(dat_model$neurological_deficit)), 2)`
with `neurological_deficit` and
`r round(cor(num01(dat_model$below_knee_pain), num01(dat_model$neural_tension_positive)), 2)`
with `neural_tension_positive`), so once the more informative signs are in the
model there is nothing left for them to explain. This is confounding-by-clustering
of the signs, not collinearity in the technical sense -- as the VIFs below show.

Moving in the other direction, **sex** strengthens slightly away from the null
(`r round(u_sex$estimate, 2)` to `r round(g("sexMale")$estimate, 2)`) and
**cough/sneeze provocation** does the same
(`r round(u_cs$estimate, 2)` to
`r round(g("cough_sneeze_positiveYes")$estimate, 2)`), because the signs that
women are more likely to report were masking a little of both. **Neurological
deficit** barely moves (`r round(u_nd$estimate, 2)` to
`r round(g("neurological_deficitYes")$estimate, 2)`) and stays significant, and
**subjective sensory changes** is null before and after
(`r round(u_ssc$estimate, 2)` to
`r round(g("subjective_sensory_changesYes")$estimate, 2)`). The overall pattern
-- three of six signs surviving, none with an odds ratio above 3 -- is a
quantitative statement of how loosely the clinical examination maps onto the
imaging finding.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r diag-tests}
col <- check_collinearity(fit)
col
binned_residuals(fit)
```

```{r diag-numbers}
vif_tab <- as.data.frame(col) |> arrange(desc(VIF))
max_vif <- round(max(vif_tab$VIF), 2)
max_vif_term <- vif_tab$Term[which.max(vif_tab$VIF)]
n_out <- sum(check_outliers(fit))
br <- binned_residuals(fit)
br_ok <- attributes(br)$p_value
```

**Binned residuals.** Observed and predicted probabilities agree acceptably
across the range of fitted values; the binned-residual plot shows no systematic
curvature, so the linear-in-logit assumption for `age_yrs` and `bmi` is not
obviously violated.

**Collinearity in the main model.** All variance inflation factors are low. The
largest is `r max_vif_term` at `r max_vif` -- nowhere near the conventional
threshold of 5. This is worth stating explicitly because the six clinical signs
correlate with each other at r = 0.10 to 0.39, which sounds alarming but is far
too weak to destabilise a nine-parameter model. The attenuation of
`neural_tension_positive` and `below_knee_pain` in the previous section is
therefore genuine confounding by the other signs, not a variance-inflation
artefact.

**Influential observations.** `check_outliers()` flags `r n_out` observation(s)
at the Cook's-distance threshold, and there is no separation: every fitted
probability lies between
`r round(min(fitted(fit)), 2)` and `r round(max(fitted(fit)), 2)`, and every
coefficient has a finite standard error.

**Events per parameter.** `r n_rarer` in the rarer class over `r n_params`
parameters gives `r round(epp, 1)` events per parameter, above the usual
minimum of 10.

```{r clinician-trap}
fit_trap <- glm(reformulate(c(predictors, "clinician_sciatica"), outcome),
                data = dat_model, family = binomial)
vif_trap <- as.data.frame(check_collinearity(fit_trap)) |> arrange(desc(VIF))

trap_tab <- as.data.frame(col) |>
  select(Term, `VIF (main model)` = VIF) |>
  full_join(vif_trap |> select(Term, `VIF (+ clinician_sciatica)` = VIF),
            by = "Term") |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  arrange(desc(`VIF (+ clinician_sciatica)`))

trap_or <- tidy(fit_trap, exponentiate = TRUE, conf.int = TRUE)
trap_tab |> gt() |>
  tab_header(title = "What happens if `clinician_sciatica` is added",
             subtitle = "Sensitivity analysis only - not the reported model")
```

**The `clinician_sciatica` sensitivity model.** Adding the clinician's diagnosis
alongside the signs it is derived from does exactly what the Methods section
predicted. The VIF for `neural_tension_positive` rises from
`r round(vif_tab$VIF[vif_tab$Term == "neural_tension_positive"], 2)` to
`r round(vif_trap$VIF[vif_trap$Term == "neural_tension_positive"], 2)` and
`below_knee_pain` from
`r round(vif_tab$VIF[vif_tab$Term == "below_knee_pain"], 2)` to
`r round(vif_trap$VIF[vif_trap$Term == "below_knee_pain"], 2)`; the odds ratio
for `below_knee_pain` moves from
`r round(g("below_knee_painYes")$estimate, 2)` to
`r round(trap_or$estimate[trap_or$term == "below_knee_painYes"], 2)` and
`neural_tension_positive` from
`r round(g("neural_tension_positiveYes")$estimate, 2)` to
`r round(trap_or$estimate[trap_or$term == "neural_tension_positiveYes"], 2)` --
i.e. it flips to the protective side. AIC improves marginally
(`r round(AIC(fit), 1)` to `r round(AIC(fit_trap), 1)`), which is precisely the
trap: a model-selection criterion rewards adding a variable that makes the
scientific question unanswerable. The VIFs alone (max
`r round(max(vif_trap$VIF), 2)`) would not have raised an alarm. The reason to
exclude `clinician_sciatica` is that it is a summary of the other predictors, not
that a diagnostic test says so.

# Discussion

Among primary-care patients presenting with sciatica, the classical clinical
signs are only weakly informative about whether the MRI will show nerve-root
compression. Three of the six signs -- pain provoked by coughing or sneezing,
leg pain reported as worse than back pain, and a neurological deficit -- remain
independently associated after mutual adjustment, but with odds ratios below 3
and an overall Tjur's R² of `r round(as.numeric(r2_tjur(fit)), 3)` and AUC of
`r round(auc_val, 3)`. The single strongest predictor of the scan finding is
being male, which is not a clinical sign at all. Two signs that clinicians weight
heavily -- the neural-tension test and subjective sensory changes -- carry no
independent information about the imaging finding whatsoever.

The result becomes much more interesting when set against the clinician's own
diagnosis. The same six signs explain Tjur's R² = `r round(r2_clin, 2)` of the
GP's high-confidence sciatica diagnosis but only `r round(r2_mri, 2)` of the MRI
finding. The examination is therefore internally coherent -- clinicians use the
signs consistently and their diagnosis is close to a summary of them -- while
being only loosely coupled to the anatomy. In practical terms this means a
confident clinical diagnosis of sciatica should not be read as a confident
prediction of nerve-root compression, and it explains why MRI changes management
in a meaningful minority of these patients.

**Limitations.** This is a cross-sectional diagnostic association study; nothing
here is causal, and odds ratios are not likelihood ratios, so the model does not
directly give the quantities a clinician would use at the bedside. Nerve-root
compression on MRI is itself an imperfect reference standard: it is common in
asymptomatic people and the file records only its presence, not the level, side
or severity, nor whether it corresponds to the symptomatic dermatome -- a
patient counted as "compression present" may have it on the wrong side. The six
signs are recorded as coarse binaries ("at least one neural-tension test
positive", "at least one deficit"), which discards severity and almost certainly
attenuates the associations. No inter-observer reliability data are available, so
some of the weak signal may be measurement error in the signs rather than a true
absence of association. Complete-case analysis dropped `r n_dropped` rows for
missing BMI, which is negligible. The sample is Dutch primary care and the
prevalence of compression here
(`r round(100 * n_event / nrow(dat_model), 1)`%) will not transfer to secondary
care, where selection is stronger.

**Future work.** Linking the side and level of compression to the side and
dermatome of the symptoms would test whether the weak signal reflects a genuinely
loose sign-anatomy relationship or merely a crude outcome definition.

# Feasibility notes

*Not part of a student report.*

- **Fully workable as a logistic regression, and the most scientifically
  interesting of the three outcomes -- but the signal is weak.** Tjur's
  R² = `r round(as.numeric(r2_tjur(fit)), 3)`, AUC = `r round(auc_val, 3)`,
  three of nine predictors significant. Students expecting a "good model" will be
  disappointed; students who understand that a null result about a diagnostic
  test *is* a result will get the most out of this.
- **Outcome balance is excellent.** `r n_event` with compression,
  `r n_nonev` without; rarer class `r n_rarer`. With `r n_params` parameters that
  is `r round(epp, 1)` events per parameter. No separation, no zero cells, all
  fitted probabilities between `r round(min(fitted(fit)), 2)` and
  `r round(max(fitted(fit)), 2)`.
- **Complete-case N = `r nrow(dat_model)`** of `r nrow(dat)` (`r n_dropped`
  dropped, `r round(100 * n_dropped / nrow(dat), 1)`%), all from `bmi`. Adding
  `leg_pain_duration` would cost 19 more.
- **No collinearity problem in the main model**: max VIF `r max_vif`
  (`r max_vif_term`). The six signs inter-correlate at only r = 0.10 to 0.39,
  which is not enough to inflate variances. Students who *expect* a collinearity
  problem among "all those overlapping signs" will not find one -- a useful
  corrective.
- **The `clinician_sciatica` trap is the sharpest in this dataset and students
  will fall into it.** It is a column in the file, it is not an outcome, and it
  is essentially a summary of the signs (they explain Tjur's
  R² = `r round(r2_clin, 2)` of it). Added to the model it inflates the VIF for
  `neural_tension_positive` from
  `r round(vif_tab$VIF[vif_tab$Term == "neural_tension_positive"], 2)` to
  `r round(vif_trap$VIF[vif_trap$Term == "neural_tension_positive"], 2)`, flips
  the `neural_tension_positive` odds ratio to
  `r round(trap_or$estimate[trap_or$term == "neural_tension_positiveYes"], 2)`,
  drives `below_knee_pain` to
  `r round(trap_or$estimate[trap_or$term == "below_knee_painYes"], 2)` -- and
  *improves* AIC from `r round(AIC(fit), 1)` to `r round(AIC(fit_trap), 1)`.
  The VIFs never exceed `r round(max(vif_trap$VIF), 2)`, so no automated check
  catches it. The argument has to be made on subject-matter grounds.
- **The sign-versus-MRI gap is the teaching gold here.** Signs correlate
  r = `r round(min(cor_tab$r_mri), 2)` to `r round(max(cor_tab$r_mri), 2)` with
  the MRI outcome but up to `r round(max(cor_tab$r_clin), 2)` with the
  clinician's diagnosis. Point students at that contrast rather than at the
  individual odds ratios.
- **Sex dominates.** OR `r round(g("sexMale")$estimate, 2)`, larger than any
  clinical sign, and it is not obvious a priori. Expect students to discover it
  and struggle to explain it; the honest answer is that this file has no
  anatomical or occupational variable to mediate it.
