---
title: "Feeling protected from COVID-19 at work"
subtitle: "Dataset 2: COVID-19 workplace protection survey"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "workplace_protection.csv"),
                    show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names. `id` is a row
# identifier and is deliberately excluded from every model.
outcome    <- "feels_protected"
predictors <- c("vaccination_status", "management_role", "sex",
                "college_degree", "age_group", "country_grp", "sector_grp")

# The same predictor set with country and sector left at their original, sparse
# granularity. Used once, to show the reader why the collapsing was done.
predictors_sparse <- c("vaccination_status", "management_role", "sex",
                       "college_degree", "age_group", "country", "sector_raw")

dat <- dat_raw |>
  mutate(
    # Outcome: answering "yes" to feeling protected from COVID-19 at work.
    feels_protected = factor(feels_protected, levels = c(0, 1),
                             labels = c("No", "Yes")),
    # Deliberate reference levels: the largest group in each variable, except
    # age_group, where the youngest band is the natural baseline.
    vaccination_status = factor(
      vaccination_status,
      levels = c("vaccinated", "no_vaccine_access", "vaccine_refusal"),
      labels = c("Vaccinated", "No vaccine access", "Vaccine refusal")),
    management_role = factor(management_role, levels = c(0, 1),
                             labels = c("No", "Yes")),
    sex = factor(sex, levels = c("female", "male"),
                 labels = c("Female", "Male")),
    college_degree = factor(college_degree, levels = c(0, 1),
                            labels = c("No", "Yes")),
    age_group = factor(age_group,
                       levels = c("under_35", "35_to_45", "45_to_55", "over_55"),
                       labels = c("<35", "35-45", "45-55", ">55")),
    # Original 7-level country and 10-level sector, kept for the sparse-model
    # demonstration and for the exploratory figure.
    country = factor(country,
                     levels = c("canada", "ireland", "united_kingdom", "china",
                                "argentina", "nigeria", "poland"),
                     labels = c("Canada", "Ireland", "United Kingdom", "China",
                                "Argentina", "Nigeria", "Poland")),
    sector_raw = factor(occupational_sector,
                        levels = c("health_care", "other", "education",
                                   "professional", "public_service", "retail",
                                   "admin", "construction", "transport",
                                   "manufacturing"),
                        labels = c("Health care", "Other", "Education",
                                   "Professional", "Public service", "Retail",
                                   "Admin", "Construction", "Transport",
                                   "Manufacturing")),
    # Collapsed versions actually used in the models: see Methods.
    country_grp = fct_other(country, keep = c("Canada", "Ireland"),
                            other_level = "Other country"),
    sector_grp = factor(
      case_when(occupational_sector == "health_care" ~ "Health care",
                occupational_sector == "education"   ~ "Education",
                TRUE                                 ~ "Other sector"),
      levels = c("Health care", "Education", "Other sector"))
  ) |>
  mutate(country_grp = relevel(country_grp, ref = "Canada"))

dat_model <- dat |>
  select(all_of(c(outcome, predictors, "country", "sector_raw"))) |>
  drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

n_events    <- sum(dat_model[[outcome]] == "Yes")   # feels protected
n_nonevents <- sum(dat_model[[outcome]] == "No")    # does not feel protected
n_rarer     <- min(n_events, n_nonevents)

# Counts behind the collapsing decision, quoted in Methods.
cnt_country <- dat |> count(country)
cnt_sector  <- dat |> count(sector_raw)
n_small_country <- sum(cnt_country$n < 25)
n_small_sector  <- sum(cnt_sector$n < 28)

# Helper used for every odds ratio quoted in the prose, so that no number in the
# text is typed by hand.
or_txt <- function(model, term, digits = 2) {
  tt <- tidy(model, exponentiate = TRUE, conf.int = TRUE) |>
    filter(term == !!term)
  p <- tt$p.value
  p_lab <- if (p < 0.001) "p < 0.001" else paste0("p = ", format(round(p, 3), nsmall = 3))
  sprintf("OR %s, 95%% CI %s to %s, %s",
          format(round(tt$estimate, digits), nsmall = digits),
          format(round(tt$conf.low, digits), nsmall = digits),
          format(round(tt$conf.high, digits), nsmall = digits),
          p_lab)
}

# Univariable fits are needed both for the table and for the prose, so they are
# built once and stored.
uv_fits <- map(predictors, ~ glm(reformulate(.x, outcome),
                                 data = dat_model, family = binomial)) |>
  set_names(predictors)
```

# Introduction

Workplaces were the main setting in which working-age adults were exposed to
SARS-CoV-2, and whether staff *believed* their employer's control measures were
working mattered for adherence, for morale and for staff retention. These data
come from the international validation study of a multi-lingual online
questionnaire on COVID-19 prevention and control measures in workplaces (Ingram
et al., *BMC Public Health* 2022;22:74), in which workers in eight countries
were recruited by non-probability convenience sampling. The catalogue file
contains the `r nrow(dat_raw)` records in the deposited complete-demographics
sheet, each a single worker reporting whether they feel protected from COVID-19
at work.

**Aim.** Among these employed adults, we examined which of vaccination status,
holding a management role, sex, education, age group, country of residence and
occupational sector are independently associated with the odds of feeling
protected from COVID-19 at work, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** International cross-sectional online survey of workers,
recruited through non-probability convenience sampling in eight countries; the
deposited demographic sheet supplies `r nrow(dat_raw)` respondents, each
measured once. There is no follow-up and no sampling frame.

**Variables.** The outcome is `feels_protected`, a yes/no answer to whether the
respondent feels protected from COVID-19 at work; "Yes" is treated as the event,
so odds ratios above 1 mean *more* likely to feel protected. In the analysed
file `r n_events` of `r nrow(dat_model)` respondents
(`r round(100 * n_events / nrow(dat_model), 1)`%) feel protected and
`r n_nonevents` (`r round(100 * n_nonevents / nrow(dat_model), 1)`%) do not.
Predictors are all categorical: `vaccination_status` (reference: vaccinated, the
largest group; the other levels are no vaccine access and vaccine refusal),
`management_role` (reference: no), `sex` (reference: female, the larger group),
`college_degree` (reference: no degree), `age_group` in years (reference: <35,
the youngest band), `country_grp` (reference: Canada, the largest group) and
`sector_grp` (reference: health care, the largest group). There are no
continuous predictors, so no units apply. `id` is a row identifier and is
excluded from every model.

**Collapsing sparse categories.** Two predictors arrive with far more levels
than this sample can support. `country` has 7 levels, of which
`r n_small_country` contain fewer than 25 respondents (United Kingdom
`r cnt_country$n[cnt_country$country == "United Kingdom"]`, China
`r cnt_country$n[cnt_country$country == "China"]`, Argentina
`r cnt_country$n[cnt_country$country == "Argentina"]`, Nigeria
`r cnt_country$n[cnt_country$country == "Nigeria"]`, Poland
`r cnt_country$n[cnt_country$country == "Poland"]`), against Canada
`r cnt_country$n[cnt_country$country == "Canada"]` and Ireland
`r cnt_country$n[cnt_country$country == "Ireland"]`.
`occupational_sector` has 10 levels, of which `r n_small_sector` contain fewer
than 28 respondents, the smallest being transport
(`r cnt_sector$n[cnt_sector$sector_raw == "Transport"]`) and manufacturing
(`r cnt_sector$n[cnt_sector$sector_raw == "Manufacturing"]`), and manufacturing
splits `r sum(dat$sector_raw == "Manufacturing" & dat$feels_protected == "No")`
against `r sum(dat$sector_raw == "Manufacturing" & dat$feels_protected == "Yes")`
on the outcome — one cell of size one. Entered raw, these two variables alone
would contribute 15 of the 23 estimated coefficients and would produce odds
ratios with confidence intervals spanning one to two orders of magnitude. We
therefore collapsed them *before* looking at any model: `country` into Canada /
Ireland / other country, and `occupational_sector` into health care / education
/ other sector. The kept levels are the two largest countries and the two
largest sectors; education is kept separate from "other" because teaching was a
distinct exposure setting during the pandemic and it is the second-largest
sector. The collapsing rule is based on cell sizes and subject matter, not on
which grouping gave the nicest p-value. The uncollapsed model is fitted once
below purely to show the reader what was avoided.

**Statistical analysis.** Logistic regression (`glm(family = binomial)`),
justified because the outcome is binary. We fitted a univariable model for each
predictor first, then a single multivariable model containing all of them.
Predictors were chosen in advance on subject-matter grounds, not by
significance. Results are reported as odds ratios with 95% profile-likelihood
confidence intervals; significance threshold 0.05. Model fit is summarised with
AIC and events per parameter. Assumptions were checked with
`performance::check_model()`. Analysis used `r R.version.string` with tidyverse,
gtsummary and performance.

**Missing data.** There is none. All nine columns are 100% complete, so the
complete-case model uses all `r nrow(dat_model)` of `r nrow(dat_raw)` rows and
`r n_dropped` rows were dropped. No imputation, no missingness mechanism to
argue about, and no difference between the crude and adjusted analytic samples.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  select(all_of(c(outcome, predictors))) |>
  tbl_summary(
    by = feels_protected,
    missing = "no"
  ) |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics by whether the respondent feels protected from COVID-19 at work.**")
```

`r n_events` of `r nrow(dat_model)` respondents
(`r round(100 * n_events / nrow(dat_model), 1)`%) feel protected, so the two
outcome groups are reasonably balanced and the rarer class contains
`r n_rarer` people. The sample is not balanced on its predictors, however: it is
`r round(100 * mean(dat_model$sex == "Female"), 0)`% female,
`r round(100 * mean(dat_model$college_degree == "Yes"), 0)`% college-educated,
`r round(100 * mean(dat_model$vaccination_status == "Vaccinated"), 0)`%
vaccinated, and `r round(100 * mean(dat_model$country_grp != "Other country"), 0)`%
of respondents live in Canada or Ireland. The clearest group differences in
Table 1 are that managers and men are over-represented among those who feel
protected, and education-sector workers among those who do not.

## Exploratory figure

```{r explore, fig.height=6.2, fig.cap="Figure 1. Proportion feeling protected from COVID-19 at work, by country (7 original levels) and by occupational sector (10 original levels). Points are observed proportions, bars are 95% Wilson confidence intervals, the number beside each label is the group size, and the dashed line is the overall proportion. Groups plotted in grey have fewer than 30 respondents."}
prop_dat <- bind_rows(
  dat_model |> transmute(variable = "Country (7 levels)",
                         level = as.character(country),
                         event = feels_protected == "Yes"),
  dat_model |> transmute(variable = "Occupational sector (10 levels)",
                         level = as.character(sector_raw),
                         event = feels_protected == "Yes")
) |>
  group_by(variable, level) |>
  summarise(total = n(), x = sum(event), .groups = "drop") |>
  # Wilson score interval: better behaved than Wald when a cell is tiny, which
  # is exactly the situation this figure is drawn to expose.
  mutate(
    p      = x / total,
    denom  = 1 + 1.96^2 / total,
    centre = (p + 1.96^2 / (2 * total)) / denom,
    half   = 1.96 * sqrt(p * (1 - p) / total + 1.96^2 / (4 * total^2)) / denom,
    lo     = pmax(0, centre - half),
    hi     = pmin(1, centre + half),
    label  = fct_reorder(paste0(level, " (n = ", total, ")"), p),
    sparse = total < 30
  )

ggplot(prop_dat, aes(x = p, y = label, colour = sparse)) +
  geom_vline(xintercept = n_events / nrow(dat_model),
             linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0.2) +
  geom_point(size = 2.2) +
  scale_colour_manual(values = c(`FALSE` = "#2b6a99", `TRUE` = "grey65"),
                      guide = "none") +
  scale_x_continuous(labels = scales::percent, limits = c(0, 1)) +
  facet_grid(rows = vars(variable), scales = "free_y", space = "free_y") +
  labs(x = "Feels protected from COVID-19 at work", y = NULL)
```

The figure shows both a signal and a problem. The signal is that education,
retail and transport workers report feeling protected far less often than health
care workers, and that the education contrast rests on a group of
`r sum(dat_model$sector_raw == "Education")` people, which is large enough to
take seriously. The problem is everything drawn in grey: the confidence interval
for manufacturing runs from about
`r round(100 * prop_dat$lo[prop_dat$level == "Manufacturing"], 0)`% to
`r round(100 * prop_dat$hi[prop_dat$level == "Manufacturing"], 0)`%,
and China, Argentina, Nigeria and Poland have intervals between
`r round(100 * min((prop_dat |> filter(level %in% c("China", "Argentina", "Nigeria", "Poland")))$hi - (prop_dat |> filter(level %in% c("China", "Argentina", "Nigeria", "Poland")))$lo), 0)`
and
`r round(100 * max((prop_dat |> filter(level %in% c("China", "Argentina", "Nigeria", "Poland")))$hi - (prop_dat |> filter(level %in% c("China", "Argentina", "Nigeria", "Poland")))$lo), 0)`
percentage points wide, against
`r round(100 * (prop_dat$hi[prop_dat$level == "Canada"] - prop_dat$lo[prop_dat$level == "Canada"]), 0)`
for Canada. Those groups cannot support their own coefficient, which is why
country and sector were collapsed before modelling.

## Univariable models

```{r univariable}
uv <- dat_model |>
  select(all_of(c(outcome, predictors))) |>
  tbl_uvregression(
    y = all_of(outcome),
    method = glm,
    method.args = list(family = binomial),
    exponentiate = TRUE
  )
uv
```

Two predictors dominate the crude picture. Holding a management role more than
triples the odds of feeling protected
(`r or_txt(uv_fits$management_role, "management_roleYes")`), and men have roughly
double the odds of women (`r or_txt(uv_fits$sex, "sexMale")`). Working in the
education sector rather than health care roughly halves them
(`r or_txt(uv_fits$sector_grp, "sector_grpEducation")`), while the pooled "other
sector" group is indistinguishable from health care
(`r or_txt(uv_fits$sector_grp, "sector_grpOther sector")`). Feeling protected
rises steadily with age: relative to workers under 35, the 45-55 band has
`r or_txt(uv_fits$age_group, "age_group45-55")` and the over-55 band
`r or_txt(uv_fits$age_group, "age_group>55")`.

The predictor of most substantive interest behaves least interestingly.
Vaccination status shows no crude association with feeling protected: relative
to vaccinated respondents, those with no access to a vaccine have
`r or_txt(uv_fits$vaccination_status, "vaccination_statusNo vaccine access")`
and those who refused vaccination have
`r or_txt(uv_fits$vaccination_status, "vaccination_statusVaccine refusal")`.
Both point estimates sit below 1 and both intervals comfortably include it.
Country contributes nothing either, once collapsed
(Ireland versus Canada, `r or_txt(uv_fits$country_grp, "country_grpIreland")`),
and holding a college degree is at most weakly and negatively associated
(`r or_txt(uv_fits$college_degree, "college_degreeYes")`).

## Why adjustment is needed

The crude estimates above are mutually contaminated in at least two concrete
ways in *this* file. First, **`sex` confounds the `management_role`
association**. The sample is `r round(100 * mean(dat_model$sex == "Female"), 0)`%
female overall but only
`r round(100 * mean(dat_model$sex[dat_model$management_role == "Yes"] == "Female"), 0)`%
female among managers, and men are crudely more likely to feel protected
regardless of their role, so part of the apparent management effect is simply
that managers are disproportionately men. The reverse also holds: part of the
crude male advantage is that men are more often managers. Second,
**`sector_grp` confounds both `sex` and `age_group`**. Health care and education
have very different sex compositions here
(`r round(100 * mean(dat_model$sex[dat_model$sector_grp == "Health care"] == "Female"), 0)`%
female in health care versus
`r round(100 * mean(dat_model$sex[dat_model$sector_grp == "Other sector"] == "Female"), 0)`%
in the other-sector group) and very different outcome rates, so a sex comparison
that ignores sector is partly a comparison of industries. Third, **`country_grp`
and `sector_grp` confound each other severely**, because recruitment was
country-specific and sector-specific at the same time: only
`r round(100 * mean(dat_model$sector_grp[dat_model$country_grp == "Canada"] == "Education"), 0)`%
of Canadian and
`r round(100 * mean(dat_model$sector_grp[dat_model$country_grp == "Ireland"] == "Education"), 0)`%
of Irish respondents work in education, against
`r round(100 * mean(dat_model$sector_grp[dat_model$country_grp == "Other country"] == "Education"), 0)`%
of the other-country group, and education is the sector with much the lowest
rate of feeling protected. A crude country comparison is therefore substantially
a comparison of which industries happened to be recruited where — which is why
the Ireland coefficient changes direction once sector is held constant. Only a
model holding all seven predictors constant lets any one of them be read as a
within-stratum comparison.

## Multivariable model

```{r multivariable}
fit <- glm(reformulate(predictors, outcome), data = dat_model, family = binomial)

mv <- tbl_regression(fit, exponentiate = TRUE)
mv
```

```{r fit-stats}
n_params <- length(coef(fit))
fit_null <- glm(reformulate("1", outcome), data = dat_model, family = binomial)
lrt <- anova(fit_null, fit, test = "LRT")

tibble(
  Statistic = c("AIC (fitted model)", "AIC (intercept-only model)",
                "Null deviance", "Residual deviance",
                "Likelihood-ratio test vs null (p)", "Tjur's R2",
                "N", "Events (feels protected)", "Non-events",
                "Rarer outcome class",
                "Model parameters (incl. intercept)",
                "Events per parameter"),
  Value = c(round(AIC(fit), 1), round(AIC(fit_null), 1),
            round(fit$null.deviance, 1), round(fit$deviance, 1),
            format.pval(lrt$`Pr(>Chi)`[2], digits = 3, eps = 1e-10),
            round(performance::r2_tjur(fit), 3),
            nrow(dat_model), n_events, n_nonevents, n_rarer, n_params,
            round(n_rarer / (n_params - 1), 1))
) |>
  gt() |>
  tab_header(title = "Model fit")
```

The model estimates `r n_params - 1` coefficients besides the intercept, and the
smaller outcome class contains `r n_rarer` observations, giving about
`r round(n_rarer / (n_params - 1), 1)` events per parameter — comfortably above
the conventional minimum of 10, so the collapsed model is not overfitted. It
beats the intercept-only model (AIC `r round(AIC(fit), 1)` versus
`r round(AIC(fit_null), 1)`; likelihood-ratio test p < 0.001), but explains
little: Tjur's R² is `r round(performance::r2_tjur(fit), 3)`, meaning the average
predicted probability of feeling protected differs by only about
`r round(100 * performance::r2_tjur(fit), 0)` percentage points between those who
do and those who do not.

Conditional on all the other predictors, three associations survive. Managers
have `r or_txt(fit, "management_roleYes")` — the single strongest and most robust
finding in the dataset. Men have `r or_txt(fit, "sexMale")`. Education-sector
workers have `r or_txt(fit, "sector_grpEducation")` relative to health care, and
the other-sector group `r or_txt(fit, "sector_grpOther sector")`. Age retains a
gradient but only the oldest band clears the threshold
(`r or_txt(fit, "age_group>55")` for over-55 versus under-35, with
`r or_txt(fit, "age_group45-55")` at 45-55). Neither country contrast is
significant (Ireland `r or_txt(fit, "country_grpIreland")`), and neither is
holding a degree (`r or_txt(fit, "college_degreeYes")`).

Vaccination status remains null after adjustment, and the two non-reference
levels remain distinct from one another in the direction one would expect but
without precision to speak of: no vaccine access
`r or_txt(fit, "vaccination_statusNo vaccine access")` and vaccine refusal
`r or_txt(fit, "vaccination_statusVaccine refusal")`. A likelihood-ratio test of
dropping the whole variable gives
p = `r round(anova(update(fit, . ~ . - vaccination_status), fit, test = "LRT")$"Pr(>Chi)"[2], 3)`,
so on these data vaccination status adds nothing to a model of feeling
protected. It is worth stating explicitly that the two levels were *not*
collapsed: lacking access to a vaccine and refusing one are different states —
one is a supply failure, the other a choice — and pooling them would assume the
very thing an analyst would want to test.

```{r sparse-demo}
fit_sparse <- glm(reformulate(predictors_sparse, outcome),
                  data = dat_model, family = binomial)

sparse_or <- tidy(fit_sparse, exponentiate = TRUE, conf.int = TRUE) |>
  filter(str_detect(term, "^country|^sector_raw")) |>
  transmute(Term = term,
            OR = round(estimate, 2),
            `95% CI` = paste0(round(conf.low, 2), " to ", round(conf.high, 2)),
            `CI width (x)` = round(conf.high / conf.low, 1))

sparse_or |>
  gt() |>
  tab_header(
    title = "What the uncollapsed model looks like",
    subtitle = paste0("Country (7 levels) and occupational sector (10 levels) ",
                      "entered raw; ", length(coef(fit_sparse)) - 1,
                      " coefficients, ",
                      round(n_rarer / (length(coef(fit_sparse)) - 1), 1),
                      " events per parameter"))
```

This is the model that was avoided. Fifteen of its
`r length(coef(fit_sparse)) - 1` coefficients come from country and sector
alone, events per parameter falls to
`r round(n_rarer / (length(coef(fit_sparse)) - 1), 1)` — exactly on the
conventional floor — and several intervals are unusable: the manufacturing
coefficient has a confidence interval spanning a
`r round(sparse_or$"CI width (x)"[sparse_or$Term == "sector_rawManufacturing"], 0)`-fold
range and the China coefficient a roughly
`r round(sparse_or$"CI width (x)"[sparse_or$Term == "countryChina"], 0)`-fold
range, because each rests on 13 and 24 people respectively. Honesty requires
noting that the uncollapsed model does fit better in a formal sense (AIC
`r round(AIC(fit_sparse), 1)` versus `r round(AIC(fit), 1)`; likelihood-ratio
test for the extra 11 coefficients
p = `r round(anova(fit, fit_sparse, test = "LRT")$"Pr(>Chi)"[2], 3)`), driven by
the three or four small groups with extreme proportions. That is a real cost of
collapsing, and it is a genuine trade-off rather than a free lunch: the
collapsed model gives interpretable, stable estimates for the contrasts it
reports, at the price of averaging over heterogeneity it cannot resolve at this
sample size.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted odds ratios.**")
```

The honest summary of this table is that adjustment changes remarkably little.
Two estimates do move meaningfully. The **education sector** contrast moves
*away* from the null, from
`r round(exp(coef(uv_fits$sector_grp))["sector_grpEducation"], 2)` crude to
`r round(exp(coef(fit))["sector_grpEducation"], 2)` adjusted: education workers
are more often female and less often managers than health care workers, and both
of those characteristics independently lower the odds of feeling protected, so
the crude comparison *understated* the sector gap. The **45-55 age band** moves
towards the null and loses significance, from
`r round(exp(coef(uv_fits$age_group))["age_group45-55"], 2)`
(p = `r format(round(tidy(uv_fits$age_group)$p.value[3], 3), nsmall = 3)`) to
`r round(exp(coef(fit))["age_group45-55"], 2)`
(p = `r format(round(tidy(fit)$p.value[which(names(coef(fit)) == "age_group45-55")], 3), nsmall = 3)`);
that crude association was partly the fact that middle-aged respondents are more
often managers. The **other-sector** contrast also moves from
`r round(exp(coef(uv_fits$sector_grp))["sector_grpOther sector"], 2)` to
`r round(exp(coef(fit))["sector_grpOther sector"], 2)`, though it never reaches
significance in either column. **Ireland versus Canada** changes direction, from
`r round(exp(coef(uv_fits$country_grp))["country_grpIreland"], 2)` to
`r round(exp(coef(fit))["country_grpIreland"], 2)`, and the other-country
contrast from
`r round(exp(coef(uv_fits$country_grp))["country_grpOther country"], 2)` to
`r round(exp(coef(fit))["country_grpOther country"], 2)`; neither is significant
in either column, but the flip is exactly what the sector composition predicts —
the Irish respondents are disproportionately health care workers and the
other-country respondents disproportionately education workers, so the crude
country comparison was carrying the sector effect.

Against that, the headline predictors barely move at all. Management goes from
`r format(round(exp(coef(uv_fits$management_role))["management_roleYes"], 2), nsmall = 2)` to
`r format(round(exp(coef(fit))["management_roleYes"], 2), nsmall = 2)` — a change of about
`r round(100 * abs(exp(coef(fit))["management_roleYes"] / exp(coef(uv_fits$management_role))["management_roleYes"] - 1), 0)`%
— and sex from
`r format(round(exp(coef(uv_fits$sex))["sexMale"], 2), nsmall = 2)` to
`r format(round(exp(coef(fit))["sexMale"], 2), nsmall = 2)`. Vaccination status is null in both
columns and country is null in both columns. So the mutual confounding argued
for above is real but small: it matters for the sector and age contrasts and is
essentially irrelevant for the two strongest predictors. Nothing here is
collinearity — every VIF is below 1.7 — and because the file is complete, the
unadjusted and adjusted columns are computed on exactly the same
`r nrow(dat_model)` people, so none of the movement is a change of sample.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r diag-tests}
check_collinearity(fit)
```

```{r diag-binned, fig.height=4.5, fig.cap="Figure 2. Binned residuals against the fitted probability of feeling protected. Grey band is the 95% error bound; points outside it are shown in red."}
br <- binned_residuals(fit)
br_df <- as.data.frame(br)
pct_inside <- 100 * mean(br_df$group == "yes")
plot(br)
```

For a logistic model the relevant panels are the binned residual plot,
collinearity and influential observations; normality of residuals and
homogeneity of variance are linear-model assumptions and do not apply to a
binary outcome, so they are neither produced nor interpreted here.

*Binned residuals.* Every predictor in this model is categorical, so there is no
linear-in-the-logit assumption to violate in the usual sense — each predictor is
fitted saturated, and the model can only misfit through omitted interactions.
`r round(pct_inside, 0)`% of the `r nrow(br_df)` bins fall inside the error
bounds, against the 95% one would like to see, and the bins that fall outside
sit at the top of the predicted-probability range with positive residuals
(largest average residual `r round(max(br_df$ybar), 2)` at a mean fitted
probability of `r round(br_df$xbar[which.max(br_df$ybar)], 2)`). In other words
the model slightly under-predicts the probability of feeling protected among the
respondents it already rates most likely to feel protected — the signature of an
omitted interaction, most plausibly management role by sector. There is no
systematic curvature across the range, so this is a mild misfit rather than a
structural failure, but it is worth reporting rather than declaring clean.

*Collinearity.* All VIFs are low. The largest are `country_grp`
(`r round(max(check_collinearity(fit)$VIF), 2)`) and `sector_grp`
(`r round(sort(check_collinearity(fit)$VIF, decreasing = TRUE)[2], 2)`),
reflecting the fact that recruitment was country- and sector-specific at once,
but both are far below any threshold of concern.

*Influential observations.* The largest Cook's distance is
`r format(round(max(cooks.distance(fit)), 4), nsmall = 4)`, orders of magnitude
below the usual 0.5 or 1 cut-offs, and no point crosses the contour in the
influence panel. With `r nrow(dat_model)` rows and only categorical predictors,
no single respondent can move an estimate.

*Separation and cell sizes.* Every coefficient in the fitted model is finite
with a finite confidence interval, and the smallest predictor-by-outcome cell
anywhere in the fitted model is
`r min(map_int(predictors, ~ min(table(dat_model[[.x]], dat_model[[outcome]]))))`
(vaccine refusal, not feeling protected), so there is no separation and no zero
cell. This is precisely what the collapsing bought: the uncollapsed model has a
cell of size
`r min(table(dat_model$sector_raw, dat_model$feels_protected))` (manufacturing,
not feeling protected), which is quasi-separation in all but name.

*Events per parameter* is `r round(n_rarer / (n_params - 1), 1)`, adequate.

# Discussion

Among workers in this international convenience sample, three things are
independently associated with feeling protected from COVID-19 at work: being a
manager, being a man, and not working in education. The management association
is the strongest and the most stable — managers have
`r round(exp(coef(fit))["management_roleYes"], 1)` times the adjusted odds of
non-managers — and it is barely touched by adjustment for sex, age, sector or
country. Practically this says that the people who design and communicate
workplace COVID-19 controls are considerably more confident in them than the
people subject to them, which is a finding about organisational communication
rather than about the controls themselves. The original questionnaire validation
paper reported the same pattern for men, older workers and managerial staff, so
this analysis reproduces it rather than adding to it.

What the data do *not* show is any association between vaccination status and
feeling protected: neither lacking access to a vaccine nor refusing one shifts
the odds (joint p =
`r round(anova(update(fit, . ~ . - vaccination_status), fit, test = "LRT")$"Pr(>Chi)"[2], 2)`).
Personal vaccination and perceived workplace protection appear to be separate
judgements. Overall discrimination is weak (Tjur's R² =
`r round(performance::r2_tjur(fit), 3)`), so these are descriptions of
associations in one sample, not a model that would predict anything.

**Limitations.** The sampling is the central problem. Respondents were recruited
online by non-probability convenience sampling across eight countries, and
`r round(100 * mean(dat_model$country_grp != "Other country"), 0)`% of them are
in Canada or Ireland while
`r round(100 * mean(dat_model$sector_grp == "Health care"), 0)`% work in health
care. There is no sampling frame and no weighting, so the results generalise to
nobody in particular: they describe these `r nrow(dat_model)` volunteers and no
identifiable population. Country contrasts in particular should not be read as
national differences, because the composition of each country's respondents
differs in ways the file cannot capture. Second, sparse categories forced
collapsing: `country` went from 7 levels to 3 and `occupational_sector` from 10
to 3, and the uncollapsed model fits significantly better
(p = `r round(anova(fit, fit_sparse, test = "LRT")$"Pr(>Chi)"[2], 3)`), so real
heterogeneity has been averaged away. The "other country" bucket is especially
heterogeneous — the United Kingdom is at
`r round(100 * mean(dat_model$feels_protected[dat_model$country == "United Kingdom"] == "Yes"), 0)`%
feeling protected and China at
`r round(100 * mean(dat_model$feels_protected[dat_model$country == "China"] == "Yes"), 0)`%,
and pooling them into one coefficient is a compromise, not a description.
Third, the outcome is a single self-reported yes/no item collected at the same
sitting as the predictors; it measures perception, not the presence or quality
of any actual control measure, and no measure of the workplace's actual
protective measures is available in this file. Fourth, the design is
cross-sectional, so nothing here is causal — becoming a manager may raise
confidence, or confident people may become managers. Fifth, all the obvious
confounders of the management association (seniority, income, whether the job is
remote-capable, workplace size, timing relative to local case waves) are absent
from the file, so residual confounding is likely. The one thing that is *not* a
limitation is missing data: the file is 100% complete on all columns.

**Future work.** Linking these perception data to an objective audit of the
control measures actually in place would show whether the manager-worker gap
reflects better-protected managers or merely better-informed ones.

# Feasibility notes

*Not part of a student report.*

- **The outcome works.** `r n_events` protected versus `r n_nonevents` not, out
  of `r nrow(dat_model)` complete cases — a
  `r round(100 * n_events / nrow(dat_model), 0)`/`r round(100 * n_nonevents / nrow(dat_model), 0)`
  split, `r n_rarer` in the rarer class. With the collapsed predictor set the
  model has `r n_params - 1` coefficients and
  `r round(n_rarer / (n_params - 1), 1)` events per parameter, max VIF
  `r round(max(check_collinearity(fit)$VIF), 2)`, max Cook's distance
  `r format(round(max(cooks.distance(fit)), 4), nsmall = 4)`, no separation.
  Nothing breaks.
- **Sparse categories are the real hazard, and a student will hit them.**
  `country` (7 levels, `r n_small_country` with n < 25) and
  `occupational_sector` (10 levels, `r n_small_sector` with n < 28) are the two
  most tempting predictors in the file and both are unusable raw. The
  uncollapsed model has `r length(coef(fit_sparse)) - 1` coefficients,
  `r round(n_rarer / (length(coef(fit_sparse)) - 1), 1)` events per parameter,
  a 1-vs-12 outcome cell in manufacturing, and confidence intervals spanning up
  to a
  `r round(max(sparse_or$"CI width (x)"), 0)`-fold range. The brief must tell
  students to collapse, or they will report an odds ratio of
  `r round(sparse_or$OR[sparse_or$Term == "sector_rawManufacturing"], 1)` with a
  straight face.
- **Collapsing is not free.** The uncollapsed model has the *lower* AIC
  (`r round(AIC(fit_sparse), 1)` versus `r round(AIC(fit), 1)`) and the
  likelihood-ratio test favours it
  (p = `r round(anova(fit, fit_sparse, test = "LRT")$"Pr(>Chi)"[2], 3)`). A
  sharp student will notice this and needs a defensible answer: precision and
  interpretability versus fit, decided before seeing the models. Instructors
  should be ready for the question.
- **Weak signal, and the confounding story is thin.** Tjur's R² =
  `r round(performance::r2_tjur(fit), 3)`. Only three of seven predictors are
  significant after adjustment, and the two strongest — management role and sex
  — move by no more than about
  `r ceiling(100 * max(abs(exp(coef(fit))["management_roleYes"] / exp(coef(uv_fits$management_role))["management_roleYes"] - 1), abs(exp(coef(fit))["sexMale"] / exp(coef(uv_fits$sex))["sexMale"] - 1)))`%
  on adjustment. The unadjusted-versus-adjusted comparison is therefore much
  less instructive here than in datasets where crude associations collapse; the
  only real movers are the education-sector contrast (which moves *away* from
  the null, from
  `r round(exp(coef(uv_fits$sector_grp))["sector_grpEducation"], 2)` to
  `r round(exp(coef(fit))["sector_grpEducation"], 2)`) and the 45-55 age band
  (`r round(exp(coef(uv_fits$age_group))["age_group45-55"], 2)` to
  `r round(exp(coef(fit))["age_group45-55"], 2)`, losing significance).
- **The most interesting predictor is null.** `vaccination_status` is the
  variable students will want to write about, and it does nothing: joint LRT
  p = `r round(anova(update(fit, . ~ . - vaccination_status), fit, test = "LRT")$"Pr(>Chi)"[2], 2)`,
  both levels' intervals straddling 1 crudely and adjusted. It must not be
  collapsed to vaccinated/unvaccinated to manufacture a result — no access and
  refusal are different states — and a student who does so is answering a worse
  question.
- **No missing data at all.** Every column is 100% complete, so this dataset
  cannot be used to teach missingness, complete-case bias or imputation.
- **One outcome only.** `feels_protected` is the sole outcome column; the other
  seven usable columns are all categorical predictors and none is continuous.
  There is no linear-regression option, no second logistic outcome, and no
  continuous predictor, so students cannot practise per-unit interpretation,
  linearity checks, or a continuous-outcome workflow. A whole project on this
  file is one model, repeated.
- **Convenience sample across eight countries.** No sampling frame, no weights,
  `r round(100 * mean(dat_model$country_grp != "Other country"), 0)`% Canada or
  Ireland, `r round(100 * mean(dat_model$sector_grp == "Health care"), 0)`%
  health care. Any generalisation statement a student writes will be wrong
  unless it is "these respondents only".
- **Also note:** the source paper analyses 627 active workers; the deposited
  sheet used here supplies `r nrow(dat_raw)`. Students comparing their output to
  the paper will not reproduce it exactly.
