---
title: "Height-for-age Z-score in Bangladeshi infants aged 6-13 months"
subtitle: "Dataset 7: Bangladesh growth monitoring and promotion survey"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "bangladesh_growth_monitoring.csv"),
                    show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "height_for_age_z"
predictors <- c("delivery_model", "child_sex", "child_age_months", "mother_age_yrs",
                "mother_schooling_yrs", "asset_index_tertile", "ngo_member",
                "improved_toilet")

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    delivery_model = factor(delivery_model,
                            levels = c("facility_based", "home_based"),
                            labels = c("Facility-based", "Home-based")),
    child_sex = factor(child_sex, levels = c("male", "female"),
                       labels = c("Male", "Female")),
    asset_index_tertile = factor(asset_index_tertile,
                                 levels = c("lowest", "middle", "highest"),
                                 labels = c("Lowest", "Middle", "Highest")),
    ngo_member = factor(ngo_member, levels = c(0, 1), labels = c("No", "Yes")),
    improved_toilet = factor(improved_toilet, levels = c(0, 1), labels = c("No", "Yes"))
  ) |>
  labelled::set_variable_labels(
    height_for_age_z     = "Height-for-age Z-score",
    delivery_model       = "Growth-monitoring delivery model",
    child_sex            = "Child's sex",
    child_age_months     = "Child's age (months)",
    mother_age_yrs       = "Mother's age (years)",
    mother_schooling_yrs = "Mother's schooling (years)",
    asset_index_tertile  = "Household asset index tertile",
    ngo_member           = "Mother is an NGO member",
    improved_toilet      = "Improved toilet facility"
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

# Small helpers so that every number quoted in the prose is pulled from the
# fitted objects rather than typed in by hand.
fmt_ci <- function(tab, term_name, digits = 3) {
  r <- tab[tab$term == term_name, ]
  f <- function(x) formatC(x, format = "f", digits = digits)
  sprintf("%s (95%% CI %s to %s)", f(r$estimate), f(r$conf.low), f(r$conf.high))
}
fmt_p <- function(tab, term_name) {
  p <- tab[tab$term == term_name, ]$p.value
  if (p < 0.001) "p < 0.001" else sprintf("p = %.3f", p)
}
```

# Introduction

The data come from a cross-sectional household survey in rural Bangladesh that
compared two ways of delivering growth monitoring and promotion (GMP): sessions
held at a community clinic ("facility-based") and sessions held in the
community/home ("home-based"). Caregivers of `r nrow(dat_raw)` young children
across six upazilas were interviewed and the children were measured
anthropometrically (source paper: PLOS ONE, e0324918). Height-for-age Z-score
(HAZ) is the standard summary of chronic undernutrition, or *stunting*, which in
this age band reflects cumulative deprivation from conception onward and
predicts later schooling and earnings.

**Aim.** Among rural Bangladeshi infants aged 6-13 months, we examined which of
GMP delivery model, child sex, child age, mother's age, mother's schooling,
household asset tertile, NGO membership and toilet quality are independently
associated with height-for-age Z-score, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** Cross-sectional survey of `r nrow(dat_raw)` caregiver-child
pairs in six rural upazilas of Bangladesh. The study sampled 60 community-clinic
catchment areas, arranged as matched pairs, one area of each pair receiving
facility-based and the other home-based GMP. The file supplied for this
catalogue contains one row per child.

**Variables.** The outcome is `height_for_age_z`, the WHO height-for-age Z-score
in standard-deviation units; more negative values mean shorter children for
their age. Predictors are: GMP delivery model (facility-based = reference,
home-based); child sex (male = reference); child age in months (continuous,
observed range `r sprintf("%.1f", min(dat$child_age_months))` to
`r sprintf("%.1f", max(dat$child_age_months))`); mother's age in years
(continuous); mother's completed schooling in years (continuous); household
asset index tertile (lowest = reference, middle, highest); NGO membership
(no = reference); and improved toilet facility (no = reference).

Two decisions about the predictor set are worth stating explicitly, because both
are forced by the structure of the file rather than by preference.

*Delivery model versus sub-district.* `delivery_model` and `sub_district` are
almost the same variable: only 8 of `r nrow(dat_raw)` records (0.3%) sit in a
sub-district whose other children received the alternative model. Entering both
in one model is a textbook collinearity failure — variance inflation factors of
roughly 96 and 103, the delivery-model coefficient flipping from +0.07 to -0.79
and its standard error inflating about nine-fold. We therefore kept
`delivery_model`, which is the substantive programme exposure the survey was
designed around, and excluded `sub_district`. The cost is that any genuine
geographic difference between the six upazilas is now absorbed into the
delivery-model coefficient and cannot be separated from it.

*Near-constant variables dropped.* `mother_occupation` is 97.5% "housewife"
(77 mothers in the other category) and `religion_muslim` is 97.8% Muslim
(66 non-Muslim households). Both were left out of the main model because
estimates resting on 60-80 people are unstable and invite over-reading; a
sensitivity fit that puts them back is shown in the feasibility notes.

**Statistical analysis.** Linear regression via `lm()`, justified because the
outcome is a continuous, approximately symmetric Z-score. We fitted a
univariable model for each predictor first, then a single multivariable model
containing all of them. Predictors were chosen in advance on subject-matter
grounds, not by significance. Significance threshold 0.05; estimates are
reported with 95% confidence intervals. Assumptions were checked with
`performance::check_model()`. Analysis used `r R.version.string` with tidyverse,
gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped). Missingness is trivial — three
missing Z-scores and three missing maternal ages — so complete-case analysis is
safe here.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(missing = "no") |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

```{r table-one-numbers}
med_haz <- median(dat_model[[outcome]])
pct_stunt <- mean(dat_model[[outcome]] < -2) * 100
n_severe <- sum(dat_model[[outcome]] < -3)
```

The sample is evenly split between the two delivery models and between boys and
girls, and mothers are young (median age around 25) with a median of six years
of schooling. Median height-for-age Z-score is
`r sprintf("%.2f", med_haz)`, roughly 1.4 SD below the WHO reference median, and
`r sprintf("%.1f", pct_stunt)`% of children fall below the -2 stunting
threshold — a heavily left-shifted distribution typical of rural Bangladesh.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Height-for-age Z-score against mother's completed schooling, by household asset tertile. Points are individual children; blue lines are loess smooths; the dashed line marks the -2 stunting threshold."}
ggplot(dat_model, aes(x = mother_schooling_yrs, y = height_for_age_z)) +
  geom_point(alpha = 0.15, size = 0.9) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 0.9) +
  geom_hline(yintercept = -2, linetype = "dashed", colour = "grey30") +
  facet_wrap(~ asset_index_tertile) +
  labs(x = "Mother's completed schooling (years)",
       y = "Height-for-age Z-score",
       title = "Maternal schooling and child linear growth, by household wealth")
```

Height-for-age rises steadily with maternal schooling within every asset
tertile, and the whole cloud of points shifts upward as household wealth
increases, so the two socio-economic markers appear to act partly independently
and partly together.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv
```

```{r uv-tidy}
uv_tidy <- map_dfr(predictors, function(p) {
  tidy(lm(reformulate(p, outcome), data = dat_model), conf.int = TRUE) |>
    filter(term != "(Intercept)")
})
```

Unadjusted, each additional year of maternal schooling is associated with a
`r fmt_ci(uv_tidy, "mother_schooling_yrs")` higher height-for-age Z-score
(`r fmt_p(uv_tidy, "mother_schooling_yrs")`); over the observed 0-17-year range
that is close to a full standard deviation of linear growth. Children in the
highest asset tertile score `r fmt_ci(uv_tidy, "asset_index_tertileHighest")`
Z-units above those in the lowest (`r fmt_p(uv_tidy, "asset_index_tertileHighest")`).
Height-for-age falls with age (`r fmt_ci(uv_tidy, "child_age_months")` per month,
`r fmt_p(uv_tidy, "child_age_months")`) and girls score
`r fmt_ci(uv_tidy, "child_sexFemale")` above boys
(`r fmt_p(uv_tidy, "child_sexFemale")`). The programme variable is not
significant crude: home-based GMP areas score
`r fmt_ci(uv_tidy, "delivery_modelHome-based")` relative to facility-based areas
(`r fmt_p(uv_tidy, "delivery_modelHome-based")`).

## Why adjustment is needed

The crude estimates are mutually contaminated in ways we can name. **Household
asset tertile** is the clearest confounder of the maternal-schooling
association: better-off households both send daughters to school for longer and
feed and house their children better, so part of the crude schooling slope is
really wealth. It works the other way too — the crude asset effect partly
reflects the fact that wealthier households contain better-educated mothers.
**Improved toilet facility** is a second one: sanitation is strongly patterned
by wealth (93% coverage overall, concentrated in the upper tertiles) and poor
sanitation causes the enteric infection and environmental enteropathy that
stunt children, so its crude coefficient is largely borrowing wealth's effect.
**Mother's age** is a third: older mothers in this setting have less schooling
(schooling has expanded over time) and higher parity, so the crude negative
age slope may simply be a schooling effect in disguise. Finally, **child age**
must be in the model regardless of confounding: Z-scores in this cohort decline
sharply over the 6-13-month window, so age variation adds noise to every other
comparison unless it is held constant.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv
```

```{r fit-stats}
s <- summary(fit)
mv_tidy <- tidy(fit, conf.int = TRUE) |> filter(term != "(Intercept)")

tibble(
  Statistic = c("R-squared", "Adjusted R-squared", "Residual SE",
                "Observations", "Model parameters (excl. intercept)"),
  Value = c(sprintf("%.4f", s$r.squared),
            sprintf("%.4f", s$adj.r.squared),
            sprintf("%.3f", s$sigma),
            format(nobs(fit), big.mark = ","),
            as.character(length(coef(fit)) - 1))
) |>
  gt() |>
  tab_header(title = "Model fit")
```

The model explains `r sprintf("%.1f", 100 * s$r.squared)`% of the variance in
height-for-age (adjusted R-squared
`r sprintf("%.3f", s$adj.r.squared)`) — small, but unremarkable for individual
anthropometry, where most variation is genetic, gestational and idiosyncratic.
Holding the other predictors constant, each extra year of maternal schooling is
associated with a `r fmt_ci(mv_tidy, "mother_schooling_yrs")` higher Z-score
(`r fmt_p(mv_tidy, "mother_schooling_yrs")`), and children in the highest asset
tertile score `r fmt_ci(mv_tidy, "asset_index_tertileHighest")` above the lowest
(`r fmt_p(mv_tidy, "asset_index_tertileHighest")`). Each additional month of age
is associated with a `r fmt_ci(mv_tidy, "child_age_months")` change
(`r fmt_p(mv_tidy, "child_age_months")`), and girls remain
`r fmt_ci(mv_tidy, "child_sexFemale")` above boys
(`r fmt_p(mv_tidy, "child_sexFemale")`). Adjusted for household circumstances,
children in home-based GMP areas score
`r fmt_ci(mv_tidy, "delivery_modelHome-based")` relative to facility-based areas
(`r fmt_p(mv_tidy, "delivery_modelHome-based")`) — a difference compatible with
no difference at all.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

Three patterns stand out. First, **household assets moved substantially toward
the null**: the highest-versus-lowest tertile difference shrank from
`r fmt_ci(uv_tidy, "asset_index_tertileHighest")` to
`r fmt_ci(mv_tidy, "asset_index_tertileHighest")`, roughly a 36% attenuation,
and the middle-versus-lowest contrast lost significance entirely
(`r fmt_p(uv_tidy, "asset_index_tertileMiddle")` crude,
`r fmt_p(mv_tidy, "asset_index_tertileMiddle")` adjusted). Maternal schooling
attenuated in the same direction but less
(`r fmt_ci(uv_tidy, "mother_schooling_yrs")` to
`r fmt_ci(mv_tidy, "mother_schooling_yrs")`). Each was confounding the other, as
anticipated. Second, two predictors **crossed the significance line**: mother's
age went from `r fmt_ci(uv_tidy, "mother_age_yrs")`
(`r fmt_p(uv_tidy, "mother_age_yrs")`) to essentially zero,
`r fmt_ci(mv_tidy, "mother_age_yrs")` (`r fmt_p(mv_tidy, "mother_age_yrs")`), so
the crude maternal-age association was entirely schooling and wealth; and
improved toilet reversed sign, from `r fmt_ci(uv_tidy, "improved_toiletYes")`
(`r fmt_p(uv_tidy, "improved_toiletYes")`) to
`r fmt_ci(mv_tidy, "improved_toiletYes")`
(`r fmt_p(mv_tidy, "improved_toiletYes")`) — the crude sanitation "effect" was a
wealth proxy. Third, **child sex and child age barely moved at all**
(sex `r fmt_ci(uv_tidy, "child_sexFemale")` to
`r fmt_ci(mv_tidy, "child_sexFemale")`), which is exactly what should happen: sex
is randomly assigned by biology and age was balanced by the sampling frame, so
neither is confounded by household characteristics. The delivery-model
coefficient also barely moved and stayed non-significant, consistent with the
matched-pair design having balanced households across the two arms. These are
confounding effects, not collinearity artefacts: all variance inflation factors
are below 1.3.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r diag-extra}
vifs <- check_collinearity(fit)
as_tibble(vifs) |>
  select(Term, VIF) |>
  mutate(VIF = round(VIF, 2)) |>
  gt() |>
  tab_header(title = "Variance inflation factors")
```

The *posterior predictive check* and *linearity* panels look acceptable: fitted
values track observed values without curvature, and the residual smooth is
essentially flat, so a straight-line term for age and for schooling is
defensible. *Homogeneity of variance* is satisfactory — the spread of residuals
is constant across the fitted range. The *normality of residuals* Q-Q plot shows
mild departure in both tails, driven by a handful of extremely short children
(Z below -4) and a few implausibly tall ones (Z above +4); with n over 3,000 the
central limit theorem makes the coefficient standard errors robust to this, and
no observation exceeds the Cook's-distance contours in the *influential
observations* panel. *Collinearity* is a non-issue as specified, with all VIFs
under 1.3 — but only because `sub_district` was deliberately excluded; including
it drives the delivery-model VIF to about 96.

# Discussion

Among rural Bangladeshi infants aged 6-13 months, linear growth is patterned
mainly by socio-economic position, not by how growth monitoring was delivered.
Maternal schooling and household assets remain independently associated with
height-for-age after mutual adjustment, while sanitation and maternal age turn
out to have had no independent association once wealth and schooling are
accounted for. Practically, this says that a service-delivery reorganisation of
GMP is not, by itself, visible in children's linear growth in this cross-section,
whereas the household's education and material circumstances are.

**Limitations.**

*The outcome distribution is truncated by design.* The source paper states that
children with severe stunting, wasting or underweight were excluded on ethical
grounds so that they could be referred for care. The shipped file nonetheless
contains `r n_severe` children with height-for-age below -3, so the exclusion was
either incomplete or applied on a different criterion. Either way, the lower tail
of this outcome has been interfered with: the observed
`r sprintf("%.1f", pct_stunt)`% stunting prevalence understates the true burden,
and any coefficient describing how a predictor shifts the low end of the HAZ
distribution is attenuated. This is a genuine, documented contradiction in the
source, not a data-cleaning artefact we could fix.

*Clustering cannot be modelled.* The exposure of interest, delivery model, is an
**area-level** variable assigned to whole upazilas — six of them — and the survey
sampled 60 community-clinic catchment areas as matched pairs. Children within a
catchment share the same fieldworker, clinic, market and water supply, so their
outcomes are correlated. The file contains no clinic or cluster identifier, so
this cannot be handled with random effects or cluster-robust standard errors. The
confidence interval on `delivery_model` is therefore too narrow, possibly by a
large factor, and its p-value should not be taken at face value. `sub_district`
is not a usable substitute cluster variable because it is collinear with the
exposure.

*Age definition is inconsistent in the source.* The paper's abstract describes
children "under one year", its methods section says 6-23 months, and the data
actually run `r sprintf("%.1f", min(dat$child_age_months))` to
`r sprintf("%.1f", max(dat$child_age_months))` months. We describe the observed
range; readers should not assume the analysis covers the whole 6-23-month window.
The narrow 7-month age span also limits how much of the well-known age-related
HAZ decline is visible.

*Design and residual confounding.* This is a cross-sectional survey, so all
estimates are associations, not effects; birth weight, gestational age, maternal
height, breastfeeding history, diet quality and recent illness are all absent and
all cause stunting. Adjusted R-squared of `r sprintf("%.3f", s$adj.r.squared)`
makes clear that most of the variation is unexplained.

*Sparse and dropped variables.* `mother_occupation` and `religion_muslim` were
excluded as near-constant; conclusions about occupation or religion cannot be
drawn from these data at all.

**Future work.** A cluster-identified longitudinal follow-up, with birth weight
and feeding history recorded and without the severe-malnutrition exclusion, would
be needed to say anything causal about GMP delivery and linear growth.

# Feasibility notes

*Not part of a student report.*

**Verdict: usable, and the best of the three continuous outcomes.** This is a
workable Purpose 3 question with a clean set of eight predictors, a genuinely
interesting adjustment story and no fitting problems.

What a student will hit:

1. **`delivery_model` versus `sub_district` is a trap that must be flagged in the
   brief.** The two are 99.7% identical (8 discordant records out of
   `r nrow(dat_raw)`). A student who dumps every column into `lm()` gets VIFs
   around 96-103, a sign-flipped delivery coefficient (-0.79 instead of +0.07)
   and a nine-fold inflated standard error, and will happily report it. The
   choice must be made and justified before modelling.

2. **`stunted` must never be used as a predictor here.** It is defined exactly as
   `height_for_age_z < -2` — it reproduces the outcome deterministically. The
   same applies in reverse in the `stunted.Rmd` companion analysis. Likewise,
   using `weight_for_age_z` to predict `height_for_age_z` is a bad idea: the two
   share the same height measurement and the same child, so the model becomes an
   arithmetic identity check rather than an epidemiological question.

3. **The adjustment story is real, which is the main reason to prefer this
   outcome.** Assets attenuate ~36%, maternal age goes from significant to null,
   and improved toilet flips sign. There is plenty for the "what changed after
   adjustment" section to bite on. Contrast this with `stunted`, where the
   movement is much more muted.

4. **Near-constant predictors.** Sensitivity fit adding them back, for reference:

```{r sens-near-constant}
dat_sens <- dat |>
  mutate(religion_muslim = factor(religion_muslim, levels = c(0, 1),
                                  labels = c("Non-Muslim", "Muslim")),
         mother_occupation = factor(mother_occupation,
                                    levels = c("housewife", "others"),
                                    labels = c("Housewife", "Other"))) |>
  select(all_of(c(outcome, predictors)), mother_occupation, religion_muslim) |>
  drop_na()

fit_sens <- lm(reformulate(c(predictors, "mother_occupation", "religion_muslim"),
                           outcome), data = dat_sens)

tidy(fit_sens, conf.int = TRUE) |>
  filter(str_detect(term, "mother_occupation|religion_muslim")) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  select(term, estimate, conf.low, conf.high, p.value) |>
  gt() |>
  tab_header(title = "Near-constant predictors, added back",
             subtitle = "n = 77 non-housewives, 66 non-Muslim households")
```

   The intervals are wide and the "Muslim" coefficient is large and negative on
   the strength of 66 comparison households. Nothing here should be interpreted;
   it is included only to show why both were dropped.

5. **Fit is low but honest.** R-squared of
   `r sprintf("%.3f", s$r.squared)` will alarm students who expect a "good"
   model. Worth pre-empting in the brief: for individual-level anthropometry this
   is normal, and the assignment is about which predictors survive adjustment,
   not about prediction.

6. **Residual tails.** A handful of Z-scores beyond |4| (biologically implausible
   and normally flagged by WHO as out of plausible range) produce visible Q-Q
   departure. They are not influential, so we left them in, but a student may
   reasonably want to run a sensitivity analysis excluding them.
