---
title: "Underweight in Bangladeshi infants aged 6-13 months"
subtitle: "Dataset 7: Bangladesh growth monitoring and promotion survey"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "bangladesh_growth_monitoring.csv"),
                    show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "underweight"
predictors <- c("delivery_model", "child_sex", "child_age_months", "mother_age_yrs",
                "mother_schooling_yrs", "asset_index_tertile", "ngo_member",
                "improved_toilet")

dat <- dat_raw |>
  mutate(
    # DERIVED OUTCOME: the file ships no underweight indicator, so we cut the
    # weight-for-age Z-score at the WHO threshold. Nine children sit exactly on
    # -2.00; the "<=" convention counts them as underweight.
    underweight = factor(as.integer(weight_for_age_z <= -2),
                         levels = c(0, 1), labels = c("Not underweight", "Underweight")),
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    delivery_model = factor(delivery_model,
                            levels = c("facility_based", "home_based"),
                            labels = c("Facility-based", "Home-based")),
    child_sex = factor(child_sex, levels = c("male", "female"),
                       labels = c("Male", "Female")),
    asset_index_tertile = factor(asset_index_tertile,
                                 levels = c("lowest", "middle", "highest"),
                                 labels = c("Lowest", "Middle", "Highest")),
    ngo_member = factor(ngo_member, levels = c(0, 1), labels = c("No", "Yes")),
    improved_toilet = factor(improved_toilet, levels = c(0, 1), labels = c("No", "Yes"))
  ) |>
  labelled::set_variable_labels(
    underweight          = "Underweight (weight-for-age Z <= -2)",
    delivery_model       = "Growth-monitoring delivery model",
    child_sex            = "Child's sex",
    child_age_months     = "Child's age (months)",
    mother_age_yrs       = "Mother's age (years)",
    mother_schooling_yrs = "Mother's schooling (years)",
    asset_index_tertile  = "Household asset index tertile",
    ngo_member           = "Mother is an NGO member",
    improved_toilet      = "Improved toilet facility"
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

n_event <- sum(dat_model[[outcome]] == "Underweight")
n_nonevent <- nrow(dat_model) - n_event

# Small helpers so that every number quoted in the prose is pulled from the
# fitted objects rather than typed in by hand. Estimates are exponentiated to
# odds ratios.
fmt_or <- function(tab, term_name, digits = 3) {
  r <- tab[tab$term == term_name, ]
  f <- function(x) formatC(exp(x), format = "f", digits = digits)
  sprintf("OR %s (95%% CI %s to %s)", f(r$estimate), f(r$conf.low), f(r$conf.high))
}
fmt_p <- function(tab, term_name) {
  p <- tab[tab$term == term_name, ]$p.value
  if (p < 0.001) "p < 0.001" else sprintf("p = %.3f", p)
}
```

# Introduction

The data come from a cross-sectional household survey in rural Bangladesh that
compared two ways of delivering growth monitoring and promotion (GMP): sessions
held at a community clinic ("facility-based") and sessions held in the
community/home ("home-based"). Caregivers of `r nrow(dat_raw)` young children
across six upazilas were interviewed and the children were weighed and measured
(source paper: PLOS ONE, e0324918). Underweight — weight-for-age more than two
standard deviations below the WHO reference median — is the indicator plotted on
the child's growth card and the one that triggers referral, so it is the binary
outcome most directly connected to what a growth-monitoring programme is
supposed to detect.

**Aim.** Among rural Bangladeshi infants aged 6-13 months, we examined which of
GMP delivery model, child sex, child age, mother's age, mother's schooling,
household asset tertile, NGO membership and toilet quality are independently
associated with the odds of being underweight, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** Cross-sectional survey of `r nrow(dat_raw)` caregiver-child
pairs in six rural upazilas of Bangladesh. The study sampled 60 community-clinic
catchment areas, arranged as matched pairs, one area of each pair receiving
facility-based and the other home-based GMP. The file supplied for this
catalogue contains one row per child.

**Variables.** The outcome, `underweight`, is **derived** rather than shipped: a
child is classified as underweight when `weight_for_age_z <= -2`. Nine children
have a Z-score of exactly -2.00 and are counted as underweight under this
convention; the WHO definition strictly uses "below -2", so a student adopting
`< -2` would obtain `r sum(dat_raw$weight_for_age_z < -2, na.rm = TRUE)` events
instead of `r n_event`. **The event modelled is being underweight**, and "not
underweight" is the reference. Predictors are: GMP delivery model
(facility-based = reference, home-based); child sex (male = reference); child age
in months (continuous, observed range
`r sprintf("%.1f", min(dat$child_age_months))` to
`r sprintf("%.1f", max(dat$child_age_months))`); mother's age in years
(continuous); mother's completed schooling in years (continuous); household asset
index tertile (lowest = reference, middle, highest); NGO membership
(no = reference); and improved toilet facility (no = reference).

*Weight-for-age Z-score is excluded by construction.* The outcome is a
deterministic cut of `weight_for_age_z`, so including the parent Z-score would
give complete separation and meaningless coefficients. `weight_for_height_z` is
also excluded because it is computed from the identical weight measurement, and
`height_for_age_z` / `stunted` because they are built from the same child's
length measurement and overlap heavily with the outcome
(`r sum(dat_raw$weight_for_age_z <= -2 & dat_raw$stunted == 1, na.rm = TRUE)`
children are both underweight and stunted).

*Delivery model versus sub-district.* `delivery_model` and `sub_district` are
almost the same variable: only 8 of `r nrow(dat_raw)` records (0.3%) sit in a
sub-district whose other children received the alternative model. Fitting both
together yields variance inflation factors of roughly 96 and 103 and inflates the
delivery-model standard error about nine-fold. We kept `delivery_model`, the
substantive programme exposure, and excluded `sub_district`; the cost is that
geographic variation between the six upazilas is absorbed into the
delivery-model coefficient.

*Near-constant variables dropped.* `mother_occupation` is 97.5% "housewife" and
`religion_muslim` is 97.8% Muslim; both were left out of the main model and are
shown in a sensitivity fit in the feasibility notes.

**Statistical analysis.** Logistic regression via `glm(family = binomial)`,
justified because the outcome is binary. We fitted a univariable model for each
predictor first, then a single multivariable model containing all of them.
Predictors were chosen in advance on subject-matter grounds, not by significance.
Estimates are reported as odds ratios with 95% confidence intervals;
significance threshold 0.05. Model fit is summarised with AIC and the
events-per-parameter ratio. Assumptions were checked with
`performance::check_model()`. Analysis used `r R.version.string` with tidyverse,
gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped), of whom `r n_event` were underweight.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(
    by = underweight,
    missing = "no"
  ) |>
  add_n() |>
  add_overall() |>
  modify_caption("**Table 1. Sample characteristics, by underweight status.**")
```

```{r table-one-numbers}
prev <- 100 * n_event / nrow(dat_model)
n_severe_waz <- sum(dat$weight_for_age_z < -3, na.rm = TRUE)
n_severe_haz <- sum(dat$height_for_age_z < -3, na.rm = TRUE)
```

`r sprintf("%.1f", prev)`% of children (`r n_event` of `r nrow(dat_model)`) were
underweight. Underweight children come disproportionately from the lowest asset
tertile and have mothers with fewer years of schooling and, notably, higher
maternal age — a crude contrast that turns out not to survive adjustment.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Prevalence of underweight by mother's completed schooling (grouped) and household asset tertile. Error bars are 95% Wilson confidence intervals; numbers above the bars are group sizes."}
prev_dat <- dat_model |>
  mutate(school_grp = cut(mother_schooling_yrs,
                          breaks = c(-Inf, 0, 5, 10, Inf),
                          labels = c("None", "1-5 yrs", "6-10 yrs", "11+ yrs"))) |>
  count(school_grp, asset_index_tertile, underweight) |>
  pivot_wider(names_from = underweight, values_from = n, values_fill = 0) |>
  rename(no = `Not underweight`, yes = Underweight) |>
  mutate(n = no + yes,
         p = yes / n,
         z = 1.96,
         centre = (p + z^2 / (2 * n)) / (1 + z^2 / n),
         halfw  = z * sqrt(p * (1 - p) / n + z^2 / (4 * n^2)) / (1 + z^2 / n),
         lo = pmax(0, centre - halfw),
         hi = pmin(1, centre + halfw))

ggplot(prev_dat, aes(x = school_grp, y = 100 * p, fill = asset_index_tertile)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  geom_errorbar(aes(ymin = 100 * lo, ymax = 100 * hi),
                position = position_dodge(width = 0.8), width = 0.2) +
  geom_text(aes(y = 100 * hi + 2.5, label = n),
            position = position_dodge(width = 0.8), size = 2.6) +
  scale_fill_viridis_d(option = "D", end = 0.85, name = "Asset tertile") +
  labs(x = "Mother's completed schooling",
       y = "Underweight prevalence (%)",
       title = "Underweight falls with both maternal schooling and household wealth")
```

Underweight prevalence falls with maternal schooling within each asset tertile
and with asset tertile within each schooling band, which is a visual argument
that the two are partly independent — and also that each will absorb some of the
other's crude association once both are in the model.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = glm,
    method.args = list(family = binomial),
    exponentiate = TRUE
  )
uv
```

```{r uv-tidy}
uv_tidy <- map_dfr(predictors, function(p) {
  tidy(glm(reformulate(p, outcome), data = dat_model, family = binomial),
       conf.int = TRUE) |>
    filter(term != "(Intercept)")
})
```

Unadjusted, household wealth is the strongest predictor: children in the highest
asset tertile have `r fmt_or(uv_tidy, "asset_index_tertileHighest")` compared
with the lowest (`r fmt_p(uv_tidy, "asset_index_tertileHighest")`), a reduction
of well over half in the odds of being underweight, with the middle tertile at
`r fmt_or(uv_tidy, "asset_index_tertileMiddle")`
(`r fmt_p(uv_tidy, "asset_index_tertileMiddle")`). Each additional year of
maternal schooling is associated with
`r fmt_or(uv_tidy, "mother_schooling_yrs")`
(`r fmt_p(uv_tidy, "mother_schooling_yrs")`). Each additional year of *mother's
age* is associated with higher odds,
`r fmt_or(uv_tidy, "mother_age_yrs")` (`r fmt_p(uv_tidy, "mother_age_yrs")`).
Older children have higher odds, `r fmt_or(uv_tidy, "child_age_months")` per
month (`r fmt_p(uv_tidy, "child_age_months")`), and girls lower odds than boys,
`r fmt_or(uv_tidy, "child_sexFemale")` (`r fmt_p(uv_tidy, "child_sexFemale")`).
The programme variable is flat: home-based areas show
`r fmt_or(uv_tidy, "delivery_modelHome-based")`
(`r fmt_p(uv_tidy, "delivery_modelHome-based")`), an odds ratio essentially equal
to 1.

## Why adjustment is needed

**Mother's age** is the clearest confounded variable in this analysis, and the
mechanism is concrete: female schooling has expanded rapidly across birth cohorts
in Bangladesh, so a 35-year-old mother in this sample has systematically fewer
years of education than a 20-year-old, and low maternal education is strongly
associated with underweight. The crude maternal-age odds ratio is therefore
largely a maternal-education effect wearing a different label. **Household asset
tertile** confounds the schooling association in the same way and is confounded
by it in return: wealthier households educate daughters longer and also buy more
and better food, so neither crude estimate can be read on its own. **Improved
toilet facility** is a third: latrine ownership is concentrated in the upper
asset tertiles, and poor sanitation causes the enteric disease that suppresses
weight gain, so its crude odds ratio cannot be separated from wealth without
adjustment. **Child age** must be held constant on structural grounds, since
weight-for-age declines across the observed window.

## Multivariable model

```{r multivariable}
fit <- glm(reformulate(predictors, outcome), data = dat_model, family = binomial)

mv <- tbl_regression(fit, exponentiate = TRUE)
mv
```

```{r fit-stats}
mv_tidy <- tidy(fit, conf.int = TRUE) |> filter(term != "(Intercept)")
n_par <- length(coef(fit)) - 1
epp   <- min(n_event, n_nonevent) / n_par

tibble(
  Statistic = c("AIC", "Null deviance", "Residual deviance",
                "Events (underweight)", "Non-events",
                "Model parameters (excl. intercept)", "Events per parameter"),
  Value = c(sprintf("%.1f", AIC(fit)),
            sprintf("%.1f", fit$null.deviance),
            sprintf("%.1f", fit$deviance),
            format(n_event, big.mark = ","),
            format(n_nonevent, big.mark = ","),
            as.character(n_par),
            sprintf("%.1f", epp))
) |>
  gt() |>
  tab_header(title = "Model fit")
```

AIC is `r sprintf("%.1f", AIC(fit))`, with residual deviance
`r sprintf("%.1f", fit$deviance)` against a null deviance of
`r sprintf("%.1f", fit$null.deviance)`. With `r n_event` events and
`r n_par` estimated parameters there are `r sprintf("%.1f", epp)` events per
parameter, far above the conventional minimum of 10, so no coefficient is
resting on a sparse cell.

Holding the other predictors constant, children in the highest asset tertile have
`r fmt_or(mv_tidy, "asset_index_tertileHighest")` of being underweight relative
to the lowest tertile (`r fmt_p(mv_tidy, "asset_index_tertileHighest")`), and
those in the middle tertile `r fmt_or(mv_tidy, "asset_index_tertileMiddle")`
(`r fmt_p(mv_tidy, "asset_index_tertileMiddle")`). Each extra year of maternal
schooling is associated with `r fmt_or(mv_tidy, "mother_schooling_yrs")`
(`r fmt_p(mv_tidy, "mother_schooling_yrs")`), each additional month of child age
with `r fmt_or(mv_tidy, "child_age_months")`
(`r fmt_p(mv_tidy, "child_age_months")`), and girls with
`r fmt_or(mv_tidy, "child_sexFemale")` relative to boys
(`r fmt_p(mv_tidy, "child_sexFemale")`). Mother's age is no longer associated
with the outcome, `r fmt_or(mv_tidy, "mother_age_yrs")`
(`r fmt_p(mv_tidy, "mother_age_yrs")`). Adjusted for household circumstances,
home-based GMP areas show `r fmt_or(mv_tidy, "delivery_modelHome-based")`
(`r fmt_p(mv_tidy, "delivery_modelHome-based")`).

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted odds ratios.**")
```

**One predictor changed significance status, and it is the most instructive
result in the table.** Mother's age went from
`r fmt_or(uv_tidy, "mother_age_yrs")`
(`r fmt_p(uv_tidy, "mother_age_yrs")`) to
`r fmt_or(mv_tidy, "mother_age_yrs")` (`r fmt_p(mv_tidy, "mother_age_yrs")`) — on
the log-odds scale, 58% of the crude association vanished once maternal schooling
and household wealth were held constant. Older mothers do not have more
underweight children; poorer, less-educated mothers do, and they happen to be
older.

**Two predictors moved substantially toward the null but stayed significant.**
The highest-versus-lowest asset odds ratio went from
`r fmt_or(uv_tidy, "asset_index_tertileHighest")` to
`r fmt_or(mv_tidy, "asset_index_tertileHighest")` (23% attenuation on the
log-odds scale), and maternal schooling from
`r fmt_or(uv_tidy, "mother_schooling_yrs")` to
`r fmt_or(mv_tidy, "mother_schooling_yrs")` (21%). This is the reciprocal
confounding anticipated above: each was carrying part of the other's association.

**One moved away from the null without reaching significance.** Improved toilet
reversed direction from `r fmt_or(uv_tidy, "improved_toiletYes")`
(`r fmt_p(uv_tidy, "improved_toiletYes")`) to
`r fmt_or(mv_tidy, "improved_toiletYes")`
(`r fmt_p(mv_tidy, "improved_toiletYes")`). Once wealth is held constant, the
households that still lack a latrine are not the poorest ones, and whatever
crude protection sanitation appeared to offer was wealth all along.

**Three predictors barely moved**: delivery model
(`r fmt_or(uv_tidy, "delivery_modelHome-based")` to
`r fmt_or(mv_tidy, "delivery_modelHome-based")`), child sex
(`r fmt_or(uv_tidy, "child_sexFemale")` to
`r fmt_or(mv_tidy, "child_sexFemale")`) and child age
(`r fmt_or(uv_tidy, "child_age_months")` to
`r fmt_or(mv_tidy, "child_age_months")`). Sex and age are not determined by
household circumstances, so there is no confounding to remove; delivery model was
allocated to matched-pair areas, which is presumably why the arms are balanced.
None of this is a collinearity artefact — all variance inflation factors are
below 1.3.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r diag-extra}
vifs <- check_collinearity(fit)
as_tibble(vifs) |>
  select(Term, VIF) |>
  mutate(VIF = round(VIF, 2)) |>
  gt() |>
  tab_header(title = "Variance inflation factors")

bin <- binned_residuals(fit)
pct_ok <- 100 * attributes(bin)$resid_ok

fit_quad <- glm(reformulate(c("delivery_model", "child_sex",
                              "poly(child_age_months, 2)", "poly(mother_age_yrs, 2)",
                              "poly(mother_schooling_yrs, 2)", "asset_index_tertile",
                              "ngo_member", "improved_toilet"), outcome),
                data = dat_model, family = binomial)
pct_ok_quad <- 100 * attributes(binned_residuals(fit_quad))$resid_ok
```

For a logistic model only some panels apply; homogeneity of variance and
normality of residuals do not and should be ignored.

*Binned residuals* is the panel that matters, and the model does not pass it
cleanly: only `r sprintf("%.0f%%", pct_ok)` of bins fall inside the error bounds
against a nominal 95%. Refitting with quadratic terms for child age, mother's age
and schooling gives `r sprintf("%.0f%%", pct_ok_quad)` and a *worse* AIC
(`r sprintf("%.1f", AIC(fit_quad))` versus `r sprintf("%.1f", AIC(fit))`), so a
wrong functional form is not the explanation. The likelier cause is the
unmodelled clustering described under limitations, which produces extra-binomial
variation of exactly this kind. We retain the linear specification and report the
caveat rather than chasing a transformation that does not help.

*Collinearity* is unproblematic as specified, all VIFs under 1.3 — but only
because `sub_district` was deliberately excluded. The *posterior predictive
check* matches the observed outcome distribution, and no observation lies outside
the Cook's-distance contours in the *influential observations* panel. With
`r sprintf("%.1f", epp)` events per parameter the model is well clear of the
sparse-data regime.

# Discussion

About one child in five in this sample of rural Bangladeshi infants was
underweight by 6-13 months. Household wealth and maternal schooling are the
predictors that survive mutual adjustment, together with child age and sex. The
crude association with mother's age disappears entirely once education and wealth
are accounted for, which is a useful reminder that maternal age in this setting
is largely a marker of cohort and schooling. How growth monitoring was delivered
shows no association with the odds of being underweight, crude or adjusted.
Practically, this points at household poverty and women's education rather than
at service reorganisation.

**Limitations.**

*The outcome is truncated by design, so prevalence is understated.* The source
paper states that children with severe stunting, wasting or underweight were
excluded on ethical grounds so they could be referred for treatment. The shipped
file nonetheless contains `r n_severe_waz` children with weight-for-age below -3
and `r n_severe_haz` with height-for-age below -3, so the exclusion was either
incomplete or applied on a different criterion. For a binary outcome this bites
hard: the children who would most clearly be classified as underweight have been
selectively thinned, so the `r sprintf("%.1f", prev)`% prevalence reported here
understates the true burden and every odds ratio is attenuated toward 1. This is
a documented contradiction in the source, not something we could repair.

*The outcome is derived, not measured.* We cut `weight_for_age_z` at -2. Nine
children sit exactly on the boundary, and reasonable people use `<` rather than
`<=`; the choice changes the event count by nine and is recorded in the methods
so that it can be checked. Dichotomising also discards information: a child at
-2.01 and a child at -5.1 are treated identically.

*Clustering cannot be modelled.* Delivery model is an **area-level** exposure
assigned to whole upazilas — six of them — and the survey sampled 60
community-clinic catchment areas as matched pairs. Children in the same catchment
share fieldworkers, scales, clinics and food markets, so their outcomes are
correlated. The file contains no clinic or cluster identifier, so neither random
effects nor cluster-robust standard errors can be fitted, and the confidence
interval on `delivery_model` is too narrow. `sub_district` cannot substitute as a
cluster variable because it is collinear with the exposure.

*Age definition is inconsistent in the source.* The paper's abstract says
children "under one year", its methods say 6-23 months, and the data run
`r sprintf("%.1f", min(dat$child_age_months))` to
`r sprintf("%.1f", max(dat$child_age_months))` months. We describe the observed
range only; underweight prevalence over a full 6-23-month window would be higher.

*Design and residual confounding.* Cross-sectional data give associations, not
effects. Birth weight, gestational age, maternal height and BMI, breastfeeding
and complementary feeding practice, and recent illness are all absent and all
cause low weight-for-age.

*Sparse and dropped variables.* `mother_occupation` and `religion_muslim` were
excluded as near-constant, so nothing can be said about either.

**Future work.** A longitudinal design with cluster identifiers, birth
anthropometry and feeding data, and without the severe-malnutrition exclusion,
would be needed for any causal statement about GMP delivery and underweight.

# Feasibility notes

*Not part of a student report.*

**Verdict: the strongest of the three binary outcomes, and the one to recommend
if a student wants a logistic model.**

1. **The crude-versus-adjusted section actually has content**, which is not true
   of `stunted`. Mother's age crosses from significant to null
   (`r sprintf("%.3f", exp(uv_tidy$estimate[uv_tidy$term == "mother_age_yrs"]))`,
   `r fmt_p(uv_tidy, "mother_age_yrs")` to
   `r sprintf("%.3f", exp(mv_tidy$estimate[mv_tidy$term == "mother_age_yrs"]))`,
   `r fmt_p(mv_tidy, "mother_age_yrs")`); the highest asset tertile attenuates
   from `r sprintf("%.3f", exp(uv_tidy$estimate[uv_tidy$term == "asset_index_tertileHighest"]))`
   to `r sprintf("%.3f", exp(mv_tidy$estimate[mv_tidy$term == "asset_index_tertileHighest"]))`;
   maternal schooling from
   `r sprintf("%.3f", exp(uv_tidy$estimate[uv_tidy$term == "mother_schooling_yrs"]))`
   to `r sprintf("%.3f", exp(mv_tidy$estimate[mv_tidy$term == "mother_schooling_yrs"]))`;
   improved toilet flips from
   `r sprintf("%.3f", exp(uv_tidy$estimate[uv_tidy$term == "improved_toiletYes"]))`
   to `r sprintf("%.3f", exp(mv_tidy$estimate[mv_tidy$term == "improved_toiletYes"]))`.
   There is a real story to write.

2. **The outcome must be derived, so the brief has to say so explicitly**, along
   with the `<=` versus `<` convention. Students who silently use a different cut
   will report a different event count (`r n_event` versus
   `r sum(dat_raw$weight_for_age_z < -2, na.rm = TRUE)`).

3. **`weight_for_age_z` must never appear as a predictor** — it defines the
   outcome and produces complete separation. `weight_for_height_z` shares the
   identical weight measurement and is nearly as bad. This is the single most
   likely student error on this outcome and needs to be stated in the brief.

4. **`delivery_model` versus `sub_district`.** 99.7% overlap (8 discordant
   records out of `r nrow(dat_raw)`); together, VIFs around 96 and 103.

5. **Fitting is clean apart from binned residuals.**
   `r sprintf("%.1f", epp)` events per parameter, no separation, no sparse cells,
   VIFs under 1.3. The binned-residual check fails the nominal 95% standard
   (`r sprintf("%.0f%%", pct_ok)` inside bounds) and quadratic terms make it
   worse, so it should be reported as unmodelled clustering rather than treated
   as a fixable specification problem.

6. **Near-constant predictors.** Sensitivity fit adding them back:

```{r sens-near-constant}
dat_sens <- dat |>
  mutate(religion_muslim = factor(religion_muslim, levels = c(0, 1),
                                  labels = c("Non-Muslim", "Muslim")),
         mother_occupation = factor(mother_occupation,
                                    levels = c("housewife", "others"),
                                    labels = c("Housewife", "Other"))) |>
  select(all_of(c(outcome, predictors)), mother_occupation, religion_muslim) |>
  drop_na()

fit_sens <- glm(reformulate(c(predictors, "mother_occupation", "religion_muslim"),
                            outcome), data = dat_sens, family = binomial)

tidy(fit_sens, conf.int = TRUE, exponentiate = TRUE) |>
  filter(str_detect(term, "mother_occupation|religion_muslim")) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  select(term, estimate, conf.low, conf.high, p.value) |>
  gt() |>
  tab_header(title = "Near-constant predictors, added back (odds ratios)",
             subtitle = "n = 77 non-housewives, 66 non-Muslim households")
```

   Very wide intervals resting on tiny comparison groups; shown only to justify
   dropping them.

7. **Overlap with `weight_for_age_z`.** Same measurement, dichotomised. If the
   assignment does not require a logistic model, the continuous version is
   strictly better powered.
