---
title: "Wasting in Bangladeshi infants aged 6-13 months"
subtitle: "Dataset 7: Bangladesh growth monitoring and promotion survey"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "bangladesh_growth_monitoring.csv"),
                    show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "wasting"
predictors <- c("delivery_model", "child_sex", "child_age_months", "mother_age_yrs",
                "mother_schooling_yrs", "asset_index_tertile", "ngo_member",
                "improved_toilet")

dat <- dat_raw |>
  mutate(
    # DERIVED OUTCOME: the file ships no wasting indicator, so we cut the
    # weight-for-height Z-score at the WHO threshold. Five children sit exactly
    # on -2.00; the "<=" convention counts them as wasted.
    wasting = factor(as.integer(weight_for_height_z <= -2),
                     levels = c(0, 1), labels = c("Not wasted", "Wasted")),
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    delivery_model = factor(delivery_model,
                            levels = c("facility_based", "home_based"),
                            labels = c("Facility-based", "Home-based")),
    child_sex = factor(child_sex, levels = c("male", "female"),
                       labels = c("Male", "Female")),
    asset_index_tertile = factor(asset_index_tertile,
                                 levels = c("lowest", "middle", "highest"),
                                 labels = c("Lowest", "Middle", "Highest")),
    ngo_member = factor(ngo_member, levels = c(0, 1), labels = c("No", "Yes")),
    improved_toilet = factor(improved_toilet, levels = c(0, 1), labels = c("No", "Yes"))
  ) |>
  labelled::set_variable_labels(
    wasting              = "Wasted (weight-for-height Z <= -2)",
    delivery_model       = "Growth-monitoring delivery model",
    child_sex            = "Child's sex",
    child_age_months     = "Child's age (months)",
    mother_age_yrs       = "Mother's age (years)",
    mother_schooling_yrs = "Mother's schooling (years)",
    asset_index_tertile  = "Household asset index tertile",
    ngo_member           = "Mother is an NGO member",
    improved_toilet      = "Improved toilet facility"
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)

n_event <- sum(dat_model[[outcome]] == "Wasted")
n_nonevent <- nrow(dat_model) - n_event

# Small helpers so that every number quoted in the prose is pulled from the
# fitted objects rather than typed in by hand. Estimates are exponentiated to
# odds ratios.
fmt_or <- function(tab, term_name, digits = 3) {
  r <- tab[tab$term == term_name, ]
  f <- function(x) formatC(exp(x), format = "f", digits = digits)
  sprintf("OR %s (95%% CI %s to %s)", f(r$estimate), f(r$conf.low), f(r$conf.high))
}
fmt_p <- function(tab, term_name) {
  p <- tab[tab$term == term_name, ]$p.value
  if (p < 0.001) "p < 0.001" else sprintf("p = %.3f", p)
}
```

# Introduction

The data come from a cross-sectional household survey in rural Bangladesh that
compared two ways of delivering growth monitoring and promotion (GMP): sessions
held at a community clinic ("facility-based") and sessions held in the
community/home ("home-based"). Caregivers of `r nrow(dat_raw)` young children
across six upazilas were interviewed and the children were weighed and measured
(source paper: PLOS ONE, e0324918). Wasting — weight-for-height more than two
standard deviations below the WHO reference median — is the indicator of *acute*
undernutrition, carries the highest short-term mortality risk of the three
standard indices, and is the one a growth-monitoring programme would hope to
detect early.

**Aim.** Among rural Bangladeshi infants aged 6-13 months, we examined which of
GMP delivery model, child sex, child age, mother's age, mother's schooling,
household asset tertile, NGO membership and toilet quality are independently
associated with the odds of wasting, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** Cross-sectional survey of `r nrow(dat_raw)` caregiver-child
pairs in six rural upazilas of Bangladesh. The study sampled 60 community-clinic
catchment areas, arranged as matched pairs, one area of each pair receiving
facility-based and the other home-based GMP. The file supplied for this
catalogue contains one row per child.

**Variables.** The outcome, `wasting`, is **derived** rather than shipped: a
child is classified as wasted when `weight_for_height_z <= -2`. Five children
have a Z-score of exactly -2.00 and count as wasted under this convention; the
WHO definition strictly uses "below -2", which would give
`r sum(dat_raw$weight_for_height_z < -2, na.rm = TRUE)` events instead of
`r n_event`. **The event modelled is being wasted**, and "not wasted" is the
reference. Predictors are: GMP delivery model (facility-based = reference,
home-based); child sex (male = reference); child age in months (continuous,
observed range `r sprintf("%.1f", min(dat$child_age_months))` to
`r sprintf("%.1f", max(dat$child_age_months))`); mother's age in years
(continuous); mother's completed schooling in years (continuous); household asset
index tertile (lowest = reference, middle, highest); NGO membership
(no = reference); and improved toilet facility (no = reference).

*Weight-for-height Z-score is excluded by construction.* The outcome is a
deterministic cut of `weight_for_height_z`, so including the parent Z-score
would give complete separation. `weight_for_age_z` is also excluded because it
is computed from the identical weight measurement, and `height_for_age_z` /
`stunted` because they use the same length measurement on the same child.

*Delivery model versus sub-district.* `delivery_model` and `sub_district` are
almost the same variable: only 8 of `r nrow(dat_raw)` records (0.3%) sit in a
sub-district whose other children received the alternative model. Fitting both
together yields variance inflation factors of roughly 96 and 103 and inflates
the delivery-model standard error about nine-fold. We kept `delivery_model`, the
substantive programme exposure, and excluded `sub_district`; geographic variation
between the six upazilas is therefore absorbed into the delivery-model
coefficient.

*Near-constant variables dropped.* `mother_occupation` is 97.5% "housewife" and
`religion_muslim` is 97.8% Muslim. With only `r n_event` events in this analysis
these variables are especially dangerous — the smallest cross-classified cell
would contain a handful of children — so both were left out of the main model and
are shown in a sensitivity fit in the feasibility notes.

**Statistical analysis.** Logistic regression via `glm(family = binomial)`,
justified because the outcome is binary. We fitted a univariable model for each
predictor first, then a single multivariable model containing all of them.
Predictors were chosen in advance on subject-matter grounds, not by significance.
Estimates are reported as odds ratios with 95% confidence intervals;
significance threshold 0.05. Model fit is summarised with AIC and the
events-per-parameter ratio. Assumptions were checked with
`performance::check_model()`. Analysis used `r R.version.string` with tidyverse,
gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped), of whom `r n_event` were wasted.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(
    by = wasting,
    missing = "no"
  ) |>
  add_n() |>
  add_overall() |>
  modify_caption("**Table 1. Sample characteristics, by wasting status.**")
```

```{r table-one-numbers}
prev <- 100 * n_event / nrow(dat_model)
n_severe_whz <- sum(dat$weight_for_height_z < -3, na.rm = TRUE)
n_severe_haz <- sum(dat$height_for_age_z < -3, na.rm = TRUE)
```

`r sprintf("%.1f", prev)`% of children (`r n_event` of `r nrow(dat_model)`) were
wasted — far fewer events than for stunting or underweight in the same children,
which is both epidemiologically expected and the main statistical constraint on
this analysis. Wasted children come somewhat more often from the lowest asset
tertile, but the differences across the other predictors in Table 1 are small.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Prevalence of wasting by household asset tertile, separately for the two growth-monitoring delivery models. Error bars are 95% Wilson confidence intervals; numbers above the bars are group sizes. Note the wide intervals, a direct consequence of the small number of wasted children."}
prev_dat <- dat_model |>
  count(delivery_model, asset_index_tertile, wasting) |>
  pivot_wider(names_from = wasting, values_from = n, values_fill = 0) |>
  rename(no = `Not wasted`, yes = Wasted) |>
  mutate(n = no + yes,
         p = yes / n,
         z = 1.96,
         centre = (p + z^2 / (2 * n)) / (1 + z^2 / n),
         halfw  = z * sqrt(p * (1 - p) / n + z^2 / (4 * n^2)) / (1 + z^2 / n),
         lo = pmax(0, centre - halfw),
         hi = pmin(1, centre + halfw))

ggplot(prev_dat, aes(x = asset_index_tertile, y = 100 * p,
                     fill = asset_index_tertile)) +
  geom_col(width = 0.65) +
  geom_errorbar(aes(ymin = 100 * lo, ymax = 100 * hi), width = 0.15) +
  geom_text(aes(y = 100 * hi + 0.8, label = paste0("n=", n)), size = 3) +
  facet_wrap(~ delivery_model) +
  scale_fill_viridis_d(option = "D", end = 0.85, guide = "none") +
  labs(x = "Household asset index tertile",
       y = "Wasting prevalence (%)",
       title = "Wasting by household wealth and delivery model")
```

Wasting prevalence is lower in the highest asset tertile in both arms, but the
confidence intervals are wide and overlap substantially, so the wealth gradient
here is much less clear-cut than it is for stunting or underweight.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = glm,
    method.args = list(family = binomial),
    exponentiate = TRUE
  )
uv
```

```{r uv-tidy}
uv_tidy <- map_dfr(predictors, function(p) {
  tidy(glm(reformulate(p, outcome), data = dat_model, family = binomial),
       conf.int = TRUE) |>
    filter(term != "(Intercept)")
})
```

Only three predictors reach significance unadjusted. Children in the highest
asset tertile have `r fmt_or(uv_tidy, "asset_index_tertileHighest")` of being
wasted compared with the lowest tertile
(`r fmt_p(uv_tidy, "asset_index_tertileHighest")`), while the middle tertile is
not distinguishable from the lowest
(`r fmt_or(uv_tidy, "asset_index_tertileMiddle")`,
`r fmt_p(uv_tidy, "asset_index_tertileMiddle")`). Each additional month of child
age is associated with `r fmt_or(uv_tidy, "child_age_months")`
(`r fmt_p(uv_tidy, "child_age_months")`), and each additional year of maternal
schooling with `r fmt_or(uv_tidy, "mother_schooling_yrs")`
(`r fmt_p(uv_tidy, "mother_schooling_yrs")`) — a much weaker schooling gradient
than for stunting, where the equivalent odds ratio is near 0.90. The programme
variable points toward more wasting in home-based areas but does not reach
significance: `r fmt_or(uv_tidy, "delivery_modelHome-based")`
(`r fmt_p(uv_tidy, "delivery_modelHome-based")`). Child sex
(`r fmt_p(uv_tidy, "child_sexFemale")`), mother's age
(`r fmt_p(uv_tidy, "mother_age_yrs")`), NGO membership
(`r fmt_p(uv_tidy, "ngo_memberYes")`) and improved toilet
(`r fmt_p(uv_tidy, "improved_toiletYes")`) are all non-significant crude.

## Why adjustment is needed

**Household asset tertile** is the confounder that does most of the work here.
It determines whether a mother went to school, whether the household owns a
latrine and whether the household can buy food through a seasonal price spike,
so it stands upstream of both the exposures and the outcome. In particular, the
crude maternal-schooling association almost certainly contains a wealth
component: educated women live in better-off households, and it is the household
budget rather than the mother's years of study that most plausibly protects
against acute weight loss. **Improved toilet facility** is the second: latrine
ownership is concentrated in the upper asset tertiles (93% coverage overall) and
poor sanitation causes the diarrhoeal illness that produces acute weight loss, so
its crude odds ratio conflates sanitation with wealth. **Mother's age** is a
third, since older mothers belong to less-educated cohorts and have higher
parity. And **child age** must be held constant on structural grounds, because
weight-for-height falls measurably even across the seven-month window observed
here, so any group with an older age profile will look more wasted.

## Multivariable model

```{r multivariable}
fit <- glm(reformulate(predictors, outcome), data = dat_model, family = binomial)

mv <- tbl_regression(fit, exponentiate = TRUE)
mv
```

```{r fit-stats}
mv_tidy <- tidy(fit, conf.int = TRUE) |> filter(term != "(Intercept)")
n_par <- length(coef(fit)) - 1
epp   <- min(n_event, n_nonevent) / n_par

tibble(
  Statistic = c("AIC", "Null deviance", "Residual deviance",
                "Events (wasted)", "Non-events",
                "Model parameters (excl. intercept)", "Events per parameter"),
  Value = c(sprintf("%.1f", AIC(fit)),
            sprintf("%.1f", fit$null.deviance),
            sprintf("%.1f", fit$deviance),
            format(n_event, big.mark = ","),
            format(n_nonevent, big.mark = ","),
            as.character(n_par),
            sprintf("%.1f", epp))
) |>
  gt() |>
  tab_header(title = "Model fit")
```

AIC is `r sprintf("%.1f", AIC(fit))`, with residual deviance
`r sprintf("%.1f", fit$deviance)` against a null deviance of
`r sprintf("%.1f", fit$null.deviance)` — the model reduces deviance by only
`r sprintf("%.1f", fit$null.deviance - fit$deviance)` points across
`r n_par` parameters. With `r n_event` events and `r n_par` parameters there are
`r sprintf("%.1f", epp)` events per parameter. That clears the conventional
minimum of 10, so the estimates are not in the sparse-data regime, but it is less
than a third of the ratio available for the stunting model and the confidence
intervals are correspondingly wide.

Holding the other predictors constant, only two associations remain: children in
the highest asset tertile have `r fmt_or(mv_tidy, "asset_index_tertileHighest")`
relative to the lowest tertile
(`r fmt_p(mv_tidy, "asset_index_tertileHighest")`), and each additional month of
child age is associated with `r fmt_or(mv_tidy, "child_age_months")`
(`r fmt_p(mv_tidy, "child_age_months")`). Maternal schooling is no longer
associated with wasting, `r fmt_or(mv_tidy, "mother_schooling_yrs")`
(`r fmt_p(mv_tidy, "mother_schooling_yrs")`). Adjusted for household
circumstances, home-based GMP areas show
`r fmt_or(mv_tidy, "delivery_modelHome-based")`
(`r fmt_p(mv_tidy, "delivery_modelHome-based")`), still not significant.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted odds ratios.**")
```

**One predictor changed significance status, and it is the whole story of this
table.** Maternal schooling went from
`r fmt_or(uv_tidy, "mother_schooling_yrs")`
(`r fmt_p(uv_tidy, "mother_schooling_yrs")`) to
`r fmt_or(mv_tidy, "mother_schooling_yrs")`
(`r fmt_p(mv_tidy, "mother_schooling_yrs")`) — on the log-odds scale 53% of the
crude association disappeared, and the confidence interval now spans 1. The
apparent protective effect of maternal education against acute undernutrition was
largely household wealth. This contrasts sharply with stunting, where the
schooling association survives adjustment easily, and it fits the standard
interpretation that education acts on *chronic* growth through care practices
while acute weight loss is driven by short-term resource shocks.

**One predictor reversed direction without ever being significant.** Improved
toilet went from `r fmt_or(uv_tidy, "improved_toiletYes")`
(`r fmt_p(uv_tidy, "improved_toiletYes")`) to
`r fmt_or(mv_tidy, "improved_toiletYes")`
(`r fmt_p(mv_tidy, "improved_toiletYes")`), and NGO membership from
`r fmt_or(uv_tidy, "ngo_memberYes")` to
`r fmt_or(mv_tidy, "ngo_memberYes")`. Both crude estimates were wealth proxies.

**Everything else barely moved.** The highest asset tertile went from
`r fmt_or(uv_tidy, "asset_index_tertileHighest")` to
`r fmt_or(mv_tidy, "asset_index_tertileHighest")`, child age from
`r fmt_or(uv_tidy, "child_age_months")` to
`r fmt_or(mv_tidy, "child_age_months")`, child sex from
`r fmt_or(uv_tidy, "child_sexFemale")` to
`r fmt_or(mv_tidy, "child_sexFemale")`, and delivery model from
`r fmt_or(uv_tidy, "delivery_modelHome-based")` to
`r fmt_or(mv_tidy, "delivery_modelHome-based")`. Age and sex are not determined by
household circumstances, so there is nothing to adjust away; the asset tertile
barely moves because, unlike in the stunting and underweight models, maternal
schooling has little independent association with wasting to take back. None of
this reflects collinearity — all variance inflation factors are below 1.3.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r diag-extra}
vifs <- check_collinearity(fit)
as_tibble(vifs) |>
  select(Term, VIF) |>
  mutate(VIF = round(VIF, 2)) |>
  gt() |>
  tab_header(title = "Variance inflation factors")

bin <- binned_residuals(fit)
pct_ok <- 100 * attributes(bin)$resid_ok

fit_quad <- glm(reformulate(c("delivery_model", "child_sex",
                              "poly(child_age_months, 2)", "poly(mother_age_yrs, 2)",
                              "poly(mother_schooling_yrs, 2)", "asset_index_tertile",
                              "ngo_member", "improved_toilet"), outcome),
                data = dat_model, family = binomial)
pct_ok_quad <- 100 * attributes(binned_residuals(fit_quad))$resid_ok
```

For a logistic model only some panels apply; homogeneity of variance and
normality of residuals do not and should be ignored.

*Binned residuals* is the panel that matters and this model performs worst of the
three binary outcomes on it: only `r sprintf("%.0f%%", pct_ok)` of bins lie
inside the error bounds, against a nominal 95%. Refitting with quadratic terms
for child age, mother's age and schooling leaves it at
`r sprintf("%.0f%%", pct_ok_quad)` and worsens AIC
(`r sprintf("%.1f", AIC(fit_quad))` versus `r sprintf("%.1f", AIC(fit))`), so the
problem is not a wrong functional form. Two things are contributing: the small
number of events (`r n_event`) means each bin holds only a handful of wasted
children, so the binned average is noisy; and the clustering described under
limitations produces genuine extra-binomial variation. The specification is
retained and the caveat reported.

*Collinearity* is unproblematic as specified, all VIFs under 1.3 — but only
because `sub_district` was deliberately excluded. The *posterior predictive
check* reproduces the observed outcome distribution, and no observation lies
outside the Cook's-distance contours in the *influential observations* panel.
With `r sprintf("%.1f", epp)` events per parameter the model avoids the
sparse-data regime, but the wide confidence intervals throughout Table 2 are the
visible cost of having only `r n_event` events.

# Discussion

`r sprintf("%.1f", prev)`% of these rural Bangladeshi infants were wasted at
6-13 months. After mutual adjustment only two predictors retain an association:
household wealth (highest versus lowest asset tertile) and the child's age. The
crude maternal-schooling association disappears once wealth is held constant,
which is the clearest substantive finding here — acute undernutrition in this
population tracks household resources rather than maternal education. How growth
monitoring was delivered shows no association with wasting, before or after
adjustment. The honest summary is that this survey has little to say about
wasting: most of what determines whether a child is acutely thin — a recent
diarrhoeal episode, a seasonal food gap, an interrupted feed — was not measured.

**Limitations.**

*Few events, so power is limited.* Only `r n_event` of `r nrow(dat_model)`
children were wasted. Almost every confidence interval in Table 2 is wide enough
to contain both a clinically important effect and no effect at all, so
"not significant" here should be read as "not informative", not as "no
association". This is the single biggest constraint on the analysis.

*The outcome is truncated by design, so prevalence is understated.* The source
paper states that children with severe stunting, wasting or underweight were
excluded on ethical grounds so they could be referred for treatment. The shipped
file nonetheless contains `r n_severe_whz` children with weight-for-height below
-3 and `r n_severe_haz` with height-for-age below -3, so the exclusion was either
incomplete or applied on a different criterion. For wasting this is the most
damaging of all three outcomes: the excluded group is precisely the group being
counted, so the `r sprintf("%.1f", prev)`% prevalence understates the true burden,
the already small event count is smaller than it should be, and every odds ratio
is attenuated toward 1.

*The outcome is derived, not measured.* We cut `weight_for_height_z` at -2. Five
children sit exactly on the boundary; using `<` rather than `<=` would give
`r sum(dat_raw$weight_for_height_z < -2, na.rm = TRUE)` events instead of
`r n_event`. Dichotomising also discards information about where in the
distribution each child sits.

*Clustering cannot be modelled.* Delivery model is an **area-level** exposure
assigned to whole upazilas — six of them — and the survey sampled 60
community-clinic catchment areas as matched pairs. Children in the same catchment
share fieldworkers, scales, clinics, water supply and food markets, so their
outcomes are correlated; wasting in particular is driven by local, seasonal
conditions that operate at exactly the catchment level. The file contains no
clinic or cluster identifier, so neither random effects nor cluster-robust
standard errors can be fitted, and the confidence interval on `delivery_model`
is too narrow. `sub_district` cannot substitute as a cluster variable because it
is collinear with the exposure.

*Age definition is inconsistent in the source.* The paper's abstract says
children "under one year", its methods say 6-23 months, and the data run
`r sprintf("%.1f", min(dat$child_age_months))` to
`r sprintf("%.1f", max(dat$child_age_months))` months. We describe the observed
range only.

*Design and residual confounding.* Cross-sectional data give associations, not
effects. Wasting is the index most sensitive to *recent* events, and neither
recent illness, feeding practice, nor season of measurement was recorded —
a serious omission, since wasting in Bangladesh is strongly seasonal.

*Sparse and dropped variables.* `mother_occupation` and `religion_muslim` were
excluded as near-constant; with `r n_event` events, estimating anything about
either would be reckless.

**Future work.** Repeated measurement across seasons, with cluster identifiers
and recent-illness histories, and without the severe-malnutrition exclusion,
would be needed before this dataset could say anything useful about acute
undernutrition.

# Feasibility notes

*Not part of a student report.*

**Verdict: the weakest of the six outcomes. Usable, but only just, and a student
who picks it should be warned what they are signing up for.**

1. **Only `r n_event` events.** Events per parameter is
   `r sprintf("%.1f", epp)` — above the rule-of-thumb 10, so nothing breaks, but
   the model is underpowered and most of the report will consist of
   non-significant results with wide intervals. Students often read that as
   failure.

2. **Only two significant adjusted predictors** (highest asset tertile and child
   age). The "which predictors survive adjustment" discussion is very short.

3. **On the other hand, there is exactly one good teaching moment**: maternal
   schooling loses significance on adjustment
   (`r sprintf("%.3f", exp(uv_tidy$estimate[uv_tidy$term == "mother_schooling_yrs"]))`,
   `r fmt_p(uv_tidy, "mother_schooling_yrs")` to
   `r sprintf("%.3f", exp(mv_tidy$estimate[mv_tidy$term == "mother_schooling_yrs"]))`,
   `r fmt_p(mv_tidy, "mother_schooling_yrs")`), a 53% attenuation on the log-odds
   scale, and the mechanism (wealth confounding) is easy to explain. If the
   assignment is graded on the confounding narrative rather than on the number of
   findings, this is salvageable.

4. **Worst binned residuals of the six models**
   (`r sprintf("%.0f%%", pct_ok)` of bins inside bounds). Quadratic terms do not
   help and make AIC worse. Students must be told to attribute this to sparse
   events plus clustering, not to hunt for transformations.

5. **The outcome must be derived**, and the `<=` versus `<` convention changes
   the event count from `r n_event` to
   `r sum(dat_raw$weight_for_height_z < -2, na.rm = TRUE)`. The brief must state
   the intended cut.

6. **`weight_for_height_z` must never appear as a predictor** — it defines the
   outcome and gives complete separation. `weight_for_age_z` shares the same
   weight measurement and is nearly as bad.

7. **`delivery_model` versus `sub_district`.** 99.7% overlap (8 discordant
   records out of `r nrow(dat_raw)`); together, VIFs of roughly 96 and 103.

8. **Near-constant predictors.** Sensitivity fit adding them back — note how much
   worse the intervals are here than in the other outcomes, because there are so
   few events:

```{r sens-near-constant}
dat_sens <- dat |>
  mutate(religion_muslim = factor(religion_muslim, levels = c(0, 1),
                                  labels = c("Non-Muslim", "Muslim")),
         mother_occupation = factor(mother_occupation,
                                    levels = c("housewife", "others"),
                                    labels = c("Housewife", "Other"))) |>
  select(all_of(c(outcome, predictors)), mother_occupation, religion_muslim) |>
  drop_na()

fit_sens <- glm(reformulate(c(predictors, "mother_occupation", "religion_muslim"),
                            outcome), data = dat_sens, family = binomial)

tidy(fit_sens, conf.int = TRUE, exponentiate = TRUE) |>
  filter(str_detect(term, "mother_occupation|religion_muslim")) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  select(term, estimate, conf.low, conf.high, p.value) |>
  gt() |>
  tab_header(title = "Near-constant predictors, added back (odds ratios)",
             subtitle = "8 wasted children among 77 non-housewives; 2 among 66 non-Muslim households")
```

   Two wasted children in the entire non-Muslim group. Nothing here is
   interpretable.

9. **Overlap with `weight_for_height_z`.** Same measurement, dichotomised, and
   the continuous version has far more power. If a student is set on acute
   undernutrition, push them to the Z-score.
