---
title: "<Outcome name> in <population>"
subtitle: "Dataset <N>: <dataset title>"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "<slug>.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "<outcome_var>"
predictors <- c(<predictor vector>)

dat <- dat_raw |>
  mutate(
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    # <recode categorical predictors into labelled factors here>
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)
```

# Introduction

<2-3 sentences: what the study was, who was enrolled, when and where, and why
this outcome matters. Name the source paper.>

**Aim.** Among <population>, we examined which of <predictor set> are
independently associated with <outcome>, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** <Dataset, sampling, N, dates, place.>

**Variables.** The outcome is `<outcome_var>` (<units / definition and, for a
binary outcome, which level is the event>). Predictors are <list with units for
continuous variables and reference levels for categorical ones>.

**Statistical analysis.** <lm() / glm(family = binomial)>, justified because the
outcome is <continuous / binary>. We fitted a univariable model for each
predictor first, then a single multivariable model containing all of them.
Predictors were chosen in advance on subject-matter grounds, not by significance.
Significance threshold 0.05; estimates are reported with 95% confidence
intervals. Assumptions were checked with `performance::check_model()`. Analysis
used `r R.version.string` with tidyverse, gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped).

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(
    by = <grouping var or delete `by =` for a continuous outcome>,
    missing = "no"
  ) |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

<One or two sentences interpreting Table 1.>

## Exploratory figure

```{r explore, fig.cap="Figure 1. <caption>"}
# Continuous outcome: scatter + geom_smooth. Binary outcome: grouped bar of proportions.
```

<One sentence interpreting the figure.>

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = <lm / glm>,
    method.args = list(<family = binomial>),
    exponentiate = <TRUE for logistic, FALSE for linear>
  )
uv
```

<Interpret the main predictor in full: estimate, direction, unit of change,
95% CI, significance.>

## Why adjustment is needed

<A paragraph naming at least two specific variables in THIS dataset that could
confound the crude associations, and saying how each relates to both the
predictor and the outcome.>

## Multivariable model

```{r multivariable}
fit <- <lm / glm>(reformulate(predictors, outcome), data = dat_model
                  <, family = binomial>)

mv <- tbl_regression(fit, exponentiate = <TRUE/FALSE>)
mv
```

```{r fit-stats}
# Linear: R-squared and adjusted R-squared. Logistic: AIC and events per parameter.
```

<Interpret key coefficients with conditional language.>

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

<Cover: at least two predictors whose estimates moved meaningfully and whether
they moved toward or away from the null; at least one that barely moved; any
that changed significance; and the likely explanation (confounding, mediation,
collinearity, or precision).>

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

<Interpret each relevant panel. For logistic models, focus on binned residuals,
collinearity, influential observations and events per parameter; homogeneity of
variance and normality of residuals do not apply.>

# Discussion

<Main finding in plain language. What it means practically.>

**Limitations.** <Specific to this analysis: failed assumption checks,
unmeasured confounders, cross-sectional design, missing data, sparse subgroups,
clustering not accounted for, associations rather than causal effects.>

**Future work.** <One sentence.>

# Feasibility notes

<NOT part of a student report. Record here anything that made this outcome
awkward or unworkable: separation, sparse cells, deterministic relationships
between variables, failed diagnostics, collinearity, tiny effective sample size
after listwise deletion, or an outcome that is not really a research question.>
