---
title: "Anaemia in Peruvian children aged 6-59 months"
subtitle: "Dataset 6: Peru DHS child anaemia and growth extract"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "peru_child_anaemia.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "anaemia"
predictors <- c("child_age_months", "child_sex", "natural_region",
                "wealth_index", "mother_education", "mother_indigenous",
                "unimproved_water")

dat <- dat_raw |>
  mutate(
    # Event level is "Yes"; a two-level factor with "No" first makes glm() model
    # the probability of anaemia, and lets Table 1 be grouped by outcome.
    anaemia = factor(anaemia, levels = c(0, 1), labels = c("No", "Yes")),
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    child_sex = factor(child_sex, levels = c("female", "male"),
                       labels = c("Female", "Male")),
    # Lima metropolitan is the best-resourced, lowest-altitude region and is the
    # natural comparator for the three other regions.
    natural_region = factor(natural_region,
                            levels = c("lima_metropolitan", "rest_of_coast",
                                       "highlands", "jungle"),
                            labels = c("Lima metropolitan", "Rest of coast",
                                       "Highlands", "Jungle")),
    # Poorest is both the largest quintile in this extract and the group with the
    # greatest expected burden, so every other quintile is read against it.
    wealth_index = factor(wealth_index,
                          levels = c("poorest", "poorer", "middle",
                                     "richer", "richest"),
                          labels = c("Poorest", "Poorer", "Middle",
                                     "Richer", "Richest")),
    # Lowest schooling is the reference so higher categories read as "more
    # education than the reference".
    mother_education = factor(mother_education,
                              levels = c("primary_or_none", "secondary", "higher"),
                              labels = c("Primary or none", "Secondary", "Higher")),
    mother_indigenous = factor(mother_indigenous, levels = c(0, 1),
                               labels = c("No", "Yes")),
    unimproved_water = factor(unimproved_water, levels = c(0, 1),
                              labels = c("No", "Yes"))
  ) |>
  labelled::set_variable_labels(
    anaemia           = "Anaemia (Hb < 11 g/dL)",
    child_age_months  = "Child age (months)",
    child_sex         = "Child sex",
    natural_region    = "Natural region",
    wealth_index      = "Household wealth quintile",
    mother_education  = "Mother's education",
    mother_indigenous = "Mother indigenous",
    unimproved_water  = "Unimproved water source"
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)
n_events  <- sum(dat_model[[outcome]] == "Yes")
```

# Introduction

Childhood anaemia is Peru's most prevalent nutritional condition: roughly four
in ten children aged 6-59 months fall below the 11 g/dL altitude-adjusted
haemoglobin threshold, and national reduction targets have been repeatedly
missed. The Peruvian Demographic and Health Survey (ENDES) measures haemoglobin
in every surveyed child of this age and links it to household wealth, maternal
characteristics, geography and water and sanitation conditions, which is the
data structure used in published ENDES analyses of the determinants of childhood
anaemia (for example Gonzales and colleagues on anaemia and altitude in Peruvian
children).

**Aim.** Among Peruvian children aged 6-59 months, we examined which of child
age, child sex, natural region, household wealth, maternal education, maternal
indigenous identity and household water source are independently associated with
anaemia, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** A cross-sectional extract of `r nrow(dat_raw)` children aged
6-59 months from the Peru Demographic and Health Survey (ENDES), a nationally
implemented household survey, each linked to their mother's interview and
household record.

**Variables.** The outcome is `anaemia`, a binary indicator defined as
altitude-adjusted haemoglobin below 11 g/dL; the modelled event is **anaemia =
Yes**. Predictors are child age in months (continuous, 6-59); child sex
(reference *Female*); natural region (reference *Lima metropolitan*); household
wealth quintile (reference *Poorest*); mother's highest education (reference
*Primary or none*); maternal indigenous identity (reference *No*); and an
unimproved household water source (reference *No*, i.e. improved source). These
seven cover child biology, geography, household socio-economic position and the
environmental route to anaemia through waterborne infection and parasitism.

`haemoglobin_g_dl` is **excluded by construction**: `anaemia` is exactly
`haemoglobin_g_dl < 11`, so including it would produce a deterministic, perfectly
separating model. Height-for-age and stunting are also excluded because they are
co-outcomes of undernutrition rather than exposures. `survey_weight` is never a
predictor.

**Statistical analysis.** Logistic regression via `glm(family = binomial)`,
justified because the outcome is binary; estimates are reported as odds ratios.
We fitted a univariable model for each predictor first, then a single
multivariable model containing all of them. Predictors were chosen in advance on
subject-matter grounds, not by significance. Significance threshold 0.05;
estimates are reported with 95% confidence intervals. Assumptions were checked
with `performance::check_model()`. Analysis used `r R.version.string` with
tidyverse, gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped,
`r round(100 * n_dropped / nrow(dat), 1)`%). The only predictor with appreciable
missingness is the water-source variable.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(
    by = anaemia,
    missing = "no"
  ) |>
  add_n() |>
  add_overall() |>
  modify_caption("**Table 1. Sample characteristics by anaemia status.**")
```

```{r table-one-numbers}
prev <- round(100 * mean(dat_model$anaemia == "Yes"), 1)
med_age_yes <- median(dat_model$child_age_months[dat_model$anaemia == "Yes"])
med_age_no  <- median(dat_model$child_age_months[dat_model$anaemia == "No"])
```

`r prev`% of children (`r n_events` of `r nrow(dat_model)`) were anaemic.
Anaemic children are markedly younger (median `r med_age_yes` versus
`r med_age_no` months) and are over-represented in the poorest quintiles, in the
highlands and jungle, and in indigenous households.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Prevalence of anaemia by household wealth quintile and natural region."}
dat_model |>
  group_by(natural_region, wealth_index) |>
  summarise(prop = mean(anaemia == "Yes"), n = n(), .groups = "drop") |>
  ggplot(aes(wealth_index, prop, fill = natural_region)) +
  geom_col(position = position_dodge(preserve = "single")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 0.65)) +
  scale_fill_viridis_d(option = "D", end = 0.9) +
  labs(x = "Household wealth quintile", y = "Anaemic (%)",
       fill = "Natural region",
       caption = "Bars are observed proportions; some region-by-quintile cells are small")
```

Anaemia prevalence falls steadily across wealth quintiles within every region,
and the highlands and jungle sit above Lima and the rest of the coast at almost
every level of wealth, so the two gradients are partly independent of each other.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = glm,
    method.args = list(family = binomial),
    exponentiate = TRUE
  )
uv
```

```{r uv-numbers}
uv_fit <- function(v) tidy(glm(reformulate(v, outcome), data = dat_model,
                               family = binomial),
                           conf.int = TRUE, exponentiate = TRUE)
u_age   <- uv_fit("child_age_months")  |> filter(term == "child_age_months")
u_rich  <- uv_fit("wealth_index")      |> filter(term == "wealth_indexRichest")
u_high  <- uv_fit("natural_region")    |> filter(term == "natural_regionHighlands")
u_indig <- uv_fit("mother_indigenous") |> filter(term == "mother_indigenousYes")
u_water <- uv_fit("unimproved_water")  |> filter(term == "unimproved_waterYes")
u_edu   <- uv_fit("mother_education")  |> filter(term == "mother_educationHigher")
u_sec   <- uv_fit("mother_education")  |> filter(term == "mother_educationSecondary")
u_sex   <- uv_fit("child_sex")         |> filter(term == "child_sexMale")
```

Child age dominates: each additional month of age is associated with
`r round(100 * (1 - u_age$estimate), 1)`% lower odds of anaemia (OR
`r round(u_age$estimate, 3)`, 95% CI `r round(u_age$conf.low, 3)` to
`r round(u_age$conf.high, 3)`, p < 0.001), which compounds to roughly a
`r round(u_age$estimate^12, 2)`-fold change in odds per year of life. Among the
categorical predictors, maternal indigenous identity (OR
`r round(u_indig$estimate, 2)`), an unimproved water source (OR
`r round(u_water$estimate, 2)`) and highland residence (OR
`r round(u_high$estimate, 2)`) all roughly double the crude odds, while the
richest quintile has about `r round(100 * (1 - u_rich$estimate))`% lower odds
than the poorest. Child sex is the only predictor that does not reach
significance (OR `r round(u_sex$estimate, 2)`, p = `r round(u_sex$p.value, 3)`).

## Why adjustment is needed

The crude odds ratios above are not comparable to each other because the
exposures overlap heavily. **Child age** confounds nearly everything: anaemia
prevalence falls from about two thirds in infancy to under a quarter at four
years, and the age distribution differs across regions and wealth quintiles, so
any crude social contrast partly compares younger with older children.
**Household wealth** confounds both the water-source and the indigenous
contrasts: unimproved water supply and indigenous ethnicity are both
concentrated in the poorest quintiles, and poverty independently affects diet
quality, supplementation uptake and infection burden, so the crude OR of
`r round(u_water$estimate, 2)` for unimproved water is partly a poverty
estimate. **Natural region** confounds the wealth gradient in turn, since the
highlands and jungle are simultaneously poorer and, for the highlands,
higher-altitude with a residual altitude effect that the haemoglobin adjustment
formula may not fully remove. Mutual adjustment is needed before any of these
can be described as independent.

## Multivariable model

```{r multivariable}
fit <- glm(reformulate(predictors, outcome), data = dat_model, family = binomial)

mv <- tbl_regression(fit, exponentiate = TRUE)
mv
```

```{r fit-stats}
n_par <- length(coef(fit))
epp   <- n_events / (n_par - 1)
tibble(
  Statistic = c("AIC", "Null deviance", "Residual deviance", "Observations",
                "Events (anaemia = Yes)", "Model parameters (incl. intercept)",
                "Events per parameter"),
  Value = c(round(AIC(fit), 1), round(fit$null.deviance, 1),
            round(fit$deviance, 1), nobs(fit), n_events, n_par,
            round(epp, 1))
) |>
  gt() |>
  tab_header(title = "Model fit")
```

```{r mv-numbers}
tid <- tidy(fit, conf.int = TRUE, exponentiate = TRUE)
g <- function(t) tid |> filter(term == t)
mv_age   <- g("child_age_months")
mv_sex   <- g("child_sexMale")
mv_high  <- g("natural_regionHighlands")
mv_jung  <- g("natural_regionJungle")
mv_rich  <- g("wealth_indexRichest")
mv_indig <- g("mother_indigenousYes")
mv_water <- g("unimproved_waterYes")
mv_edu   <- g("mother_educationHigher")
mv_sec   <- g("mother_educationSecondary")
```

The model has AIC `r round(AIC(fit), 1)` with `r n_par` parameters and
`r n_events` events, giving `r round(epp, 1)` events per parameter -- far above
the conventional minimum of 10, so the model is comfortably identified and there
is no need to reduce the predictor set. Conditional on the other six predictors,
each additional month of age is associated with `r round(100 * (1 - mv_age$estimate), 1)`%
lower odds of anaemia (OR `r round(mv_age$estimate, 3)`, 95% CI
`r round(mv_age$conf.low, 3)` to `r round(mv_age$conf.high, 3)`). Highland
residence carries `r round(mv_high$estimate, 2)` times the odds of anaemia
relative to Lima (95% CI `r round(mv_high$conf.low, 2)` to
`r round(mv_high$conf.high, 2)`) and the jungle `r round(mv_jung$estimate, 2)`
times, at the same age, sex, wealth, education, ethnicity and water source. The
wealth gradient remains monotonic: the richest quintile has
`r round(100 * (1 - mv_rich$estimate))`% lower adjusted odds than the poorest.
An unimproved water source retains an adjusted OR of
`r round(mv_water$estimate, 2)` (95% CI `r round(mv_water$conf.low, 2)` to
`r round(mv_water$conf.high, 2)`), and maternal indigenous identity
`r round(mv_indig$estimate, 2)`.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted odds ratios.**")
```

Several odds ratios move sharply toward the null. **Maternal indigenous
identity** falls from OR `r round(u_indig$estimate, 2)` crude to
`r round(mv_indig$estimate, 2)` adjusted, and **unimproved water** from
`r round(u_water$estimate, 2)` to `r round(mv_water$estimate, 2)`: in both cases
roughly half the crude excess odds was wealth, region and education, and only a
modest independent association remains. The **richest-versus-poorest** contrast
also attenuates, from OR `r round(u_rich$estimate, 2)` to
`r round(mv_rich$estimate, 2)`. By contrast **child age** barely moves at all
(OR `r round(u_age$estimate, 3)` crude to `r round(mv_age$estimate, 3)`
adjusted), because age is essentially unrelated to the social and environmental
covariates -- the textbook picture of a predictor that needs no adjustment.
Two coefficients change interpretation entirely. **Higher maternal education**
is strongly protective crudely (OR `r round(u_edu$estimate, 2)`,
p < 0.001) but is null after adjustment (OR `r round(mv_edu$estimate, 2)`,
p = `r round(mv_edu$p.value, 2)`); and **secondary education** actually flips
direction, from OR `r round(u_sec$estimate, 2)` (protective, p =
`r round(u_sec$p.value, 3)`) to `r round(mv_sec$estimate, 2)`
(p = `r round(mv_sec$p.value, 3)`). Maternal education is so tightly bound to
wealth and region that once those are held constant it carries essentially no
independent information about anaemia, and the small residual reversal should not
be over-interpreted. The **highlands** coefficient is the interesting exception
in the opposite direction: it attenuates only modestly, from
`r round(u_high$estimate, 2)` to `r round(mv_high$estimate, 2)`, so the highland
excess is not simply poverty.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r diag-numbers}
vifs <- check_collinearity(fit)
max_vif <- round(max(vifs$VIF), 2)
br <- binned_residuals(fit)
br_pct <- round(100 * attributes(br)$resid_ok, 1)
```

For a logistic model the relevant panels are the following. **Binned
residuals**: `r br_pct`% of bins fall inside the error bounds; the standard
rule of thumb is that about 95% should, so this model shows a real, if modest,
lack of fit. The pattern is driven by the linear age term -- anaemia prevalence
falls steeply from 6 to 24 months and then flattens, which a single log-odds
slope cannot follow -- and a spline or age-group factor would fix it.
**Collinearity** is low throughout (maximum VIF `r max_vif`), so despite the
strong pairwise associations between wealth, education, region and ethnicity the
adjusted odds ratios are estimated stably; the education reversal noted above is
a substantive confounding result and not a variance-inflation artefact.
**Influential observations** show no points outside the Cook's distance
contours, which is expected with `r nobs(fit)` rows. **Events per parameter** is
`r round(epp, 1)`, so overfitting and sparse-data bias are not concerns.
Homogeneity of variance and normality of residuals are printed by
`check_model()` for completeness but do not apply to a binomial GLM.

# Discussion

Anaemia in Peruvian preschool children is overwhelmingly a condition of the
second year of life: adjusted odds fall by about
`r round(100 * (1 - mv_age$estimate), 1)`% per month of age, so an 8-month-old
has several times the odds of a 4-year-old in the same household. On top of that
age effect sit a clear wealth gradient and a substantial highland and jungle
excess that survives adjustment for wealth, education and ethnicity. Water
source contributes an independent but smaller amount. Practically, this supports
concentrating iron supplementation and screening on infants and one-year-olds,
with geographic targeting of the highlands and jungle rather than
socio-economic targeting alone.

**Limitations.** Cross-sectional design: haemoglobin, water source and household
wealth are all measured at the same visit, so no temporal ordering is
established and no causal claim is warranted. The outcome is a **dichotomised
version of a continuous measurement**, which discards information and makes
results sensitive to the 11 g/dL cut-point and to the altitude-adjustment
formula; children in the highlands sit near the cut-point in large numbers, so
misclassification is non-random with respect to a key predictor. The source
ENDES analysis used the DHS complex design -- cluster (HV001), stratum (HV022)
and sampling weights -- whereas this extract contains only the weight, which is
deliberately not used here; standard errors are therefore too narrow, estimates
are not design-corrected, and **this sample must not be described as nationally
representative**. With `r nrow(dat_model)` rows almost everything is
statistically significant, so interpretation must lead with effect sizes: the
sex difference, for example, is significant in neither model and would be
irrelevant even if it were. Major determinants are simply absent: iron
supplementation, recent infection, dietary diversity, maternal anaemia, helminth
infection and exact altitude of residence. Finally, the binned-residual check
shows the linear age term does not fit well, so the age odds ratio should be
read as an average slope rather than a constant effect.

**Future work.** Refit with the DHS survey design and a spline or age-group
factor for child age, and if possible model haemoglobin continuously rather than
dichotomised.

# Feasibility notes

*Not part of a student report.*

- **Works, with one large caveat.** `r n_events` events, `r round(epp, 1)`
  events per parameter, no separation, no sparse cells, maximum VIF
  `r max_vif`. Statistically this is a safe logistic exercise.
- **FATAL trap: the deterministic twin.** `anaemia` is *exactly*
  `haemoglobin_g_dl < 11`. A student who puts haemoglobin in the model gets
  perfect separation and either an error or absurd odds ratios. This must be
  called out explicitly in the brief for this dataset; it is the single most
  likely way for a student to destroy this analysis.
- **Redundant with the haemoglobin outcome.** This is the same research question
  as `haemoglobin_g_dl`, run on a thrown-away version of the same measurement.
  It is a fine logistic-regression exercise, but a student should be told to do
  one or the other, not both, and that the continuous version is the better
  analysis.
- **Binned residuals fail (`r br_pct`% within bounds).** Caused by the
  non-linear age effect. Good teaching material, but students will need help
  reading the panel and will otherwise report "diagnostics fine".
- **Education sign flip.** Secondary education reverses direction between crude
  and adjusted models. It is real confounding, not collinearity (VIF
  `r max_vif`), but it will confuse students and needs supervision.
- **Sample size.** At `r nrow(dat_model)` rows everything is significant; the
  brief must force effect-size-led interpretation.
- **Design not honoured.** Only the weight is present, not cluster or stratum,
  so a design-correct analysis is impossible from this extract.
