---
title: "Stunting in Peruvian children aged 6-59 months"
subtitle: "Dataset 6: Peru DHS child anaemia and growth extract"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "peru_child_anaemia.csv"), show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names.
outcome    <- "stunting"
predictors <- c("child_age_months", "child_sex", "birth_weight",
                "mother_education", "wealth_index", "natural_region",
                "mother_indigenous")

dat <- dat_raw |>
  mutate(
    # Event level is "Yes"; a two-level factor with "No" first makes glm() model
    # the probability of stunting, and lets Table 1 be grouped by outcome.
    stunting = factor(stunting, levels = c(0, 1), labels = c("No", "Yes")),
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    child_sex = factor(child_sex, levels = c("female", "male"),
                       labels = c("Female", "Male")),
    # Normal birth weight is the clinically unexposed group and by far the
    # largest, so both low birth weight and macrosomia read against it.
    birth_weight = factor(birth_weight,
                          levels = c("normal", "low", "macrosomia"),
                          labels = c("Normal", "Low (<2.5 kg)", "Macrosomia")),
    # Lowest schooling is the reference so higher categories read as "more
    # education than the reference".
    mother_education = factor(mother_education,
                              levels = c("primary_or_none", "secondary", "higher"),
                              labels = c("Primary or none", "Secondary", "Higher")),
    # Poorest is both the largest quintile in this extract and the group with the
    # greatest expected burden, so every other quintile is read against it.
    wealth_index = factor(wealth_index,
                          levels = c("poorest", "poorer", "middle",
                                     "richer", "richest"),
                          labels = c("Poorest", "Poorer", "Middle",
                                     "Richer", "Richest")),
    # Lima metropolitan is the best-resourced, lowest-altitude region and is the
    # natural comparator for the three other regions.
    natural_region = factor(natural_region,
                            levels = c("lima_metropolitan", "rest_of_coast",
                                       "highlands", "jungle"),
                            labels = c("Lima metropolitan", "Rest of coast",
                                       "Highlands", "Jungle")),
    mother_indigenous = factor(mother_indigenous, levels = c(0, 1),
                               labels = c("No", "Yes"))
  ) |>
  labelled::set_variable_labels(
    stunting          = "Stunting (HAZ < -2)",
    child_age_months  = "Child age (months)",
    child_sex         = "Child sex",
    birth_weight      = "Birth-weight category (maternal report)",
    mother_education  = "Mother's education",
    wealth_index      = "Household wealth quintile",
    natural_region    = "Natural region",
    mother_indigenous = "Mother indigenous"
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)
n_events  <- sum(dat_model[[outcome]] == "Yes")
```

# Introduction

Stunting -- height-for-age more than two standard deviations below the WHO
reference median -- is the standard indicator of chronic childhood
undernutrition and a strong predictor of later cognitive and economic outcomes.
Peru has halved its national stunting prevalence since 2005, but the remaining
burden is heavily concentrated in highland and jungle communities and in the
poorest households. The Peruvian Demographic and Health Survey (ENDES) measures
every surveyed child aged 6-59 months and links the measurement to birth
history, maternal characteristics, household wealth and geography.

**Aim.** Among Peruvian children aged 6-59 months, we examined which of child
age, child sex, birth-weight category, maternal education, household wealth,
natural region and maternal indigenous identity are independently associated
with stunting, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** A cross-sectional extract of `r nrow(dat_raw)` children aged
6-59 months from the Peru Demographic and Health Survey (ENDES), a nationally
implemented household survey, each linked to their mother's interview and
household record.

**Variables.** The outcome is `stunting`, defined as height-for-age Z-score
below -2; the modelled event is **stunting = Yes**. Predictors are child age in
months (continuous, 6-59); child sex (reference *Female*); maternally reported
birth-weight category (reference *Normal*, versus *Low* i.e. below 2.5 kg and
*Macrosomia*); mother's highest education (reference *Primary or none*);
household wealth quintile (reference *Poorest*); natural region (reference *Lima
metropolitan*); and maternal indigenous identity (reference *No*).

`height_for_age_z` is **excluded by construction**: `stunting` is exactly
`height_for_age_z < -2`, so including the Z-score would give a perfectly
separating, meaningless model. Haemoglobin and anaemia are also excluded because
they are co-outcomes of undernutrition rather than exposures. `survey_weight` is
never a predictor.

**Statistical analysis.** Logistic regression via `glm(family = binomial)`,
justified because the outcome is binary; estimates are reported as odds ratios.
We fitted a univariable model for each predictor first, then a single
multivariable model containing all of them. Predictors were chosen in advance on
subject-matter grounds, not by significance. Significance threshold 0.05;
estimates are reported with 95% confidence intervals. Assumptions were checked
with `performance::check_model()`. Analysis used `r R.version.string` with
tidyverse, gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped,
`r round(100 * n_dropped / nrow(dat), 1)`%). Essentially all the loss is missing
maternally reported birth weight.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(
    by = stunting,
    missing = "no"
  ) |>
  add_n() |>
  add_overall() |>
  modify_caption("**Table 1. Sample characteristics by stunting status.**")
```

```{r table-one-numbers}
prev <- round(100 * mean(dat_model$stunting == "Yes"), 1)
prev_poorest <- round(100 * mean(dat_model$stunting[dat_model$wealth_index == "Poorest"] == "Yes"), 1)
prev_richest <- round(100 * mean(dat_model$stunting[dat_model$wealth_index == "Richest"] == "Yes"), 1)
n_macro_ev <- sum(dat_model$stunting == "Yes" & dat_model$birth_weight == "Macrosomia")
```

`r prev`% of children (`r n_events` of `r nrow(dat_model)`) were stunted. The
socio-economic concentration is extreme: `r prev_poorest`% of children in the
poorest quintile versus `r prev_richest`% in the richest, an approximately
`r round(prev_poorest / prev_richest, 1)`-fold difference in prevalence. Stunted
children are also over-represented in the highlands and jungle, in indigenous
households and among those reported as low birth weight.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Prevalence of stunting by household wealth quintile and reported birth-weight category."}
dat_model |>
  group_by(wealth_index, birth_weight) |>
  summarise(prop = mean(stunting == "Yes"), n = n(), .groups = "drop") |>
  ggplot(aes(wealth_index, prop, fill = birth_weight)) +
  geom_col(position = position_dodge(preserve = "single")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_viridis_d(option = "C", end = 0.85) +
  labs(x = "Household wealth quintile", y = "Stunted (%)",
       fill = "Birth weight",
       caption = "Bars are observed proportions; the richest / macrosomia cell is small")
```

Both gradients run in the same direction: within every wealth quintile
low-birth-weight children are stunted two to three times as often as
normal-birth-weight children (and macrosomic children least often of all), and
within every birth-weight category prevalence falls steeply as wealth rises --
from roughly half of low-birth-weight children in the poorest quintile to under
a tenth in the richest.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = glm,
    method.args = list(family = binomial),
    exponentiate = TRUE
  )
uv
```

```{r uv-numbers}
uv_fit <- function(v) tidy(glm(reformulate(v, outcome), data = dat_model,
                               family = binomial),
                           conf.int = TRUE, exponentiate = TRUE)
u_age   <- uv_fit("child_age_months")  |> filter(term == "child_age_months")
u_sex   <- uv_fit("child_sex")         |> filter(term == "child_sexMale")
u_low   <- uv_fit("birth_weight")      |> filter(term == "birth_weightLow (<2.5 kg)")
u_macro <- uv_fit("birth_weight")      |> filter(term == "birth_weightMacrosomia")
u_rich  <- uv_fit("wealth_index")      |> filter(term == "wealth_indexRichest")
u_high  <- uv_fit("natural_region")    |> filter(term == "natural_regionHighlands")
u_indig <- uv_fit("mother_indigenous") |> filter(term == "mother_indigenousYes")
u_edu   <- uv_fit("mother_education")  |> filter(term == "mother_educationHigher")
```

Low birth weight is the strongest child-level predictor: reported birth weight
under 2.5 kg carries `r round(u_low$estimate, 2)` times the crude odds of
stunting (95% CI `r round(u_low$conf.low, 2)` to `r round(u_low$conf.high, 2)`,
p < 0.001), while macrosomia is strongly protective (OR
`r round(u_macro$estimate, 2)`). The social contrasts are larger still: highland
residence carries `r round(u_high$estimate, 1)` times the odds of Lima, maternal
indigenous identity `r round(u_indig$estimate, 1)` times, and the richest
quintile has `r round(100 * (1 - u_rich$estimate))`% lower odds than the
poorest. Every predictor including child sex is significant crudely, which at
this sample size says little on its own.

## Why adjustment is needed

The crude odds ratios overlap heavily and cannot be read side by side.
**Household wealth** confounds the regional and ethnic contrasts: the highlands
and jungle contain most of the poorest quintile while Lima contains almost none
of it, and poverty independently determines diet, sanitation and health-service
access, so a crude highlands OR of `r round(u_high$estimate, 1)` is substantially
a poverty estimate. **Maternal education** does the same for wealth and
ethnicity: better-educated mothers live in richer households, are less often
indigenous, and independently practise better infant feeding, so schooling and
wealth each absorb part of the other's crude effect. **Birth weight** is itself
socially patterned, so the crude birth-weight OR partly reflects the maternal
circumstances that produced the small baby. And **child age** matters because
stunting prevalence peaks in the second year of life and the age composition is
not identical across quintiles or regions. Mutual adjustment is required before
any single association can be described as independent.

## Multivariable model

```{r multivariable}
fit <- glm(reformulate(predictors, outcome), data = dat_model, family = binomial)

mv <- tbl_regression(fit, exponentiate = TRUE)
mv
```

```{r fit-stats}
n_par <- length(coef(fit))
epp   <- n_events / (n_par - 1)
tibble(
  Statistic = c("AIC", "Null deviance", "Residual deviance", "Observations",
                "Events (stunting = Yes)", "Model parameters (incl. intercept)",
                "Events per parameter"),
  Value = c(round(AIC(fit), 1), round(fit$null.deviance, 1),
            round(fit$deviance, 1), nobs(fit), n_events, n_par,
            round(epp, 1))
) |>
  gt() |>
  tab_header(title = "Model fit")
```

```{r mv-numbers}
tid <- tidy(fit, conf.int = TRUE, exponentiate = TRUE)
g <- function(t) tid |> filter(term == t)
mv_age   <- g("child_age_months")
mv_sex   <- g("child_sexMale")
mv_low   <- g("birth_weightLow (<2.5 kg)")
mv_macro <- g("birth_weightMacrosomia")
mv_rich  <- g("wealth_indexRichest")
mv_high  <- g("natural_regionHighlands")
mv_indig <- g("mother_indigenousYes")
mv_edu   <- g("mother_educationHigher")
```

The model has AIC `r round(AIC(fit), 1)` with `r n_par` parameters and
`r n_events` events, giving `r round(epp, 1)` events per parameter -- well above
the conventional minimum of 10, so the model is stably estimated. Conditional on
the other six predictors, low birth weight retains `r round(mv_low$estimate, 2)`
times the odds of stunting (95% CI `r round(mv_low$conf.low, 2)` to
`r round(mv_low$conf.high, 2)`) and macrosomia `r round(mv_macro$estimate, 2)`
times (95% CI `r round(mv_macro$conf.low, 2)` to
`r round(mv_macro$conf.high, 2)`), so birth-weight category alone spans a
roughly `r round(mv_low$estimate / mv_macro$estimate)`-fold range in adjusted
odds. The wealth gradient remains monotonic, with the richest quintile at
`r round(mv_rich$estimate, 2)` times the odds of the poorest (95% CI
`r round(mv_rich$conf.low, 2)` to `r round(mv_rich$conf.high, 2)`), and highland
residence at `r round(mv_high$estimate, 2)` times Lima. Each additional month of
age is associated with `r round(100 * (1 - mv_age$estimate), 1)`% lower adjusted
odds, and boys have `r round(mv_sex$estimate, 2)` times the odds of girls
(p = `r round(mv_sex$p.value, 3)`).

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted odds ratios.**")
```

The social variables move most, all toward the null. **Maternal indigenous
identity** collapses from OR `r round(u_indig$estimate, 2)` crude to
`r round(mv_indig$estimate, 2)` adjusted: most of the raw excess was wealth,
education and region, though a real independent association survives.
**Highland residence** falls from `r round(u_high$estimate, 2)` to
`r round(mv_high$estimate, 2)` and **higher maternal education** rises from
`r round(u_edu$estimate, 2)` to `r round(mv_edu$estimate, 2)`, both large moves
toward 1. The **richest-versus-poorest** contrast also attenuates markedly, from
`r round(u_rich$estimate, 2)` to `r round(mv_rich$estimate, 2)`, because the
wealth quintiles differ systematically in maternal schooling and region. In
contrast, **low birth weight** barely moves (`r round(u_low$estimate, 2)` to
`r round(mv_low$estimate, 2)`) and **child age** barely moves
(`r round(u_age$estimate, 3)` to `r round(mv_age$estimate, 3)`): both are
essentially uncorrelated with the household covariates, so no adjustment was
needed for them. **Macrosomia** is the one estimate that moves appreciably in a
partly explanatory direction, from `r round(u_macro$estimate, 2)` to
`r round(mv_macro$estimate, 2)`; large babies are commoner in better-off
households, so part of its crude protective effect was socio-economic. No
predictor changes significance status. Given `r nrow(dat_model)` observations
that is unsurprising, and the useful comparison is between the *sizes* of the
attenuations, not between p-values.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r diag-numbers}
vifs <- check_collinearity(fit)
max_vif <- round(max(vifs$VIF), 2)
br <- binned_residuals(fit)
br_pct <- round(100 * attributes(br)$resid_ok, 1)
```

For a logistic model the relevant panels are as follows. **Binned residuals**:
only `r br_pct`% of bins fall inside the error bounds against a rule-of-thumb
target of about 95%, which is a clear signal of mis-specification. The cause is
visible in the data: stunting prevalence *rises* from about 12% at 6-11 months
to about 18% at 12-23 months and then falls back to about 9% by four years, so a
single linear log-odds slope in age cannot follow it, and the model
systematically under-predicts risk in the second year of life. A spline or an
age-group factor would fix this. **Collinearity** is low throughout (maximum VIF
`r max_vif`), so the wealth, education, region and ethnicity coefficients are
separately estimable despite their strong pairwise associations. **Influential
observations** show no points beyond the Cook's distance contours. **Events per
parameter** is `r round(epp, 1)`, so sparse-data bias and overfitting are not
concerns, although some individual cells are thin -- there are only
`r n_macro_ev` stunted macrosomic children in the whole file, and only a handful
of stunted children in the Lima / richest cell, so any interaction involving
those groups would be unstable. Homogeneity of variance and normality of
residuals are printed by `check_model()` for completeness but do not apply to a
binomial GLM.

# Discussion

Stunting in Peruvian preschool children is driven jointly by birth endowment and
household circumstances. A child reported as low birth weight has about
`r round(mv_low$estimate, 1)` times the odds of stunting compared with a
normal-birth-weight child of the same sex, age, region, ethnicity, maternal
schooling and household wealth -- and that estimate is essentially unchanged by
adjustment, so it is not merely a marker of poverty. On top of that, moving from
the poorest to the richest quintile is associated with roughly
`r round(100 * (1 - mv_rich$estimate))`% lower adjusted odds, and highland
residence with about `r round(mv_high$estimate, 1)` times the odds of Lima.
Practically this supports a two-pronged approach: antenatal care that reduces
low birth weight, and continued postnatal nutrition programming targeted at
poor highland and jungle households.

**Limitations.** Cross-sectional design; all estimates are associations, not
causal effects. The outcome is a **dichotomised continuous measurement** -- the
same information analysed on the continuous HAZ scale has more power and does
not depend on an arbitrary cut-point, so the continuous analysis should be
preferred where possible. Birth weight is a **maternally reported three-level
category**, not a measured weight in kilograms, and is therefore subject to
recall bias that is likely worse among less-educated mothers, exactly the group
with the highest stunting risk; this would bias the birth-weight OR toward the
null. The source ENDES analysis used the DHS complex design -- cluster (HV001),
stratum (HV022) and sampling weights -- whereas this extract carries only the
weight variable, which is deliberately not used here; standard errors are
therefore too narrow, estimates are not design-corrected, and **this sample must
not be described as nationally representative**. With `r nrow(dat_model)` rows
almost every comparison is statistically significant, so interpretation must
lead with effect sizes rather than p-values. The single strongest known
determinant of child height, **maternal height**, is not in this extract at all,
and neither are birth order, birth interval, infection history or dietary
diversity, so residual confounding is certain. The binned-residual check fails,
so the age term is mis-specified. Finally, the `r n_dropped` rows dropped for
missing birth weight are unlikely to be missing completely at random.

**Future work.** Refit with the DHS survey design, an age spline, and maternal
height and birth order from the fuller ENDES recode file.

# Feasibility notes

*Not part of a student report.*

- **Works statistically.** `r n_events` events, `r round(epp, 1)` events per
  parameter, no separation, maximum VIF `r max_vif`, large and interpretable
  odds ratios. This is a safe logistic exercise.
- **FATAL trap: the deterministic twin.** `stunting` is *exactly*
  `height_for_age_z < -2`. Including the Z-score gives perfect separation. This
  must be flagged in the brief; it is the most likely way a student wrecks this
  analysis.
- **Redundant with the HAZ outcome.** This is the same question as
  `height_for_age_z`, run on a coarsened version of the same measurement.
  Students should do one or the other; the continuous version is statistically
  better and the binary version is the better logistic-regression *exercise*.
- **Binned residuals fail badly (`r br_pct`% within bounds).** Worse than the
  anaemia model, because the age-stunting relationship is non-monotonic (peak
  around 18 months). Usable as a teaching point, but a student will need
  guidance or will simply report that diagnostics were fine.
- **Some thin cells.** Only `r n_macro_ev` stunted macrosomic children, and only
  single-digit stunted counts in the Lima / richest-quintile cell. The main
  effects model is fine; any interaction with those groups is not.
- **Birth weight is categorical, not kg.** The codebook wording invites students
  to treat `birth_weight` as continuous. It is a three-level string, and the two
  non-reference levels point in *opposite* directions.
- **Sample size.** At `r nrow(dat_model)` rows everything is significant; the
  brief must force effect-size-led interpretation.
- **Design not honoured.** Only the weight is present, not cluster or stratum,
  so a design-correct analysis is impossible from this extract.
