---
title: "Trabecular bone strength at the distal tibia in healthy adults"
subtitle: "Dataset 12: Muscle power and tibial bone strength in healthy adults"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "tibial_bone_strength.csv"),
                    show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names. The predictor set is
# capped at five parameters because there are only 142 participants, and
# `relative_peak_power` is deliberately excluded because it is an exact function
# of `peak_power_w` and `body_mass_kg` (verified below).
outcome    <- "bone_strength_index"
predictors <- c("sex", "age_band", "body_mass_kg", "peak_power_w")

lab <- function(x, l) { attr(x, "label") <- l; x }
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
pv  <- function(p) ifelse(p < 0.001, "<0.001", paste0("= ", fmt(p, 3)))

# --- Verifying the algebraic relationships between the shipped columns -------
# 1. relative_peak_power is peak_power_w / body_mass_kg, rounded to 2 dp.
rpp_err <- max(abs(dat_raw$relative_peak_power -
                     round(dat_raw$peak_power_w / dat_raw$body_mass_kg, 2)))

# 2. peak_power_w is the Sayers (1999) countermovement-jump equation,
#    PP = 51.9 * jump height (cm) + 48.9 * body mass (kg) - 2007.
#    Inverting it recovers the jump height each participant actually achieved.
jump_cm  <- (dat_raw$peak_power_w - 48.9 * dat_raw$body_mass_kg + 2007) / 51.9
pp_err   <- max(abs(dat_raw$peak_power_w -
                      (51.9 * round(jump_cm, 2) + 48.9 * dat_raw$body_mass_kg - 2007)))
# The recovered heights land on a half-inch grid, confirming the reconstruction.
halfinch_err <- max(abs(jump_cm / 2.54 * 2 - round(jump_cm / 2.54 * 2)))
n_jump_values <- n_distinct(round(jump_cm, 2))

dat <- dat_raw |>
  mutate(
    jump_cm = lab(jump_cm, "Recovered jump height (cm)"),
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex = lab(factor(sex, levels = c("female", "male"), labels = c("Female", "Male")),
              "Sex"),
    # Decades 3 to 6 hold 9, 6, 5 and 2 participants respectively, far too few to
    # estimate separately, so they are collapsed. Decade 2 is the reference
    # because it is by far the largest band.
    age_band = lab(factor(case_when(
      age_decade == 1 ~ "Decade 1",
      age_decade == 2 ~ "Decade 2",
      age_decade >= 3 ~ "Decades 3-6"),
      levels = c("Decade 2", "Decade 1", "Decades 3-6")),
      "Age band (10-year bins)"),
    body_mass_kg = lab(body_mass_kg, "Body mass (kg)"),
    peak_power_w = lab(peak_power_w, "Estimated peak jump power (W)"),
    relative_peak_power = lab(relative_peak_power, "Relative peak power (W/kg)"),
    bone_strength_index = lab(bone_strength_index, "Bone strength index (mg/mm4)")
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)
```

# Introduction

This is a cross-sectional laboratory study of `r nrow(dat_raw)` healthy adults
recruited as a convenience sample from the faculty, staff and students of
California State University East Bay, reported by Nickols-Richardson and
colleagues in the *Journal of Musculoskeletal and Neuronal Interactions* (2022).
Each participant performed a countermovement vertical jump, from which peak leg
power was estimated, and had the tibia scanned by peripheral quantitative
computed tomography (pQCT). The outcome here is the **bone strength index at the
4% distal tibial site**, a composite of trabecular density and cross-sectional
area that indexes the ability of the bone end to resist compression; the distal
tibia is a common fracture site and a standard target for exercise
interventions, so knowing which everyday characteristics track with its strength
is a useful descriptive question.

**Aim.** Among these healthy adults, we examined which of sex, age band, body
mass and estimated peak jump power are independently associated with the
trabecular bone strength index at the distal tibia, mutually adjusting for the
others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** A convenience sample of `r nrow(dat_raw)` faculty, staff and
students at a mid-sized regional university in the United States, measured once
in a laboratory session. The published report gives a mean age of about 25 years
and describes the sample as healthy and free of conditions affecting bone. The
file shipped with this catalogue holds `r ncol(dat_raw)` columns and
`r nrow(dat_raw)` rows, and nothing is missing.

**Variables.** The outcome is `bone_strength_index`, the compressive bone
strength index at the 4% distal tibial site in mg/mm4, treated as continuous
(range `r fmt(min(dat$bone_strength_index), 1)` to
`r fmt(max(dat$bone_strength_index), 1)`). Predictors are sex (reference female,
the larger group), age band (reference Decade 2, the modal band), body mass in
kilograms, and estimated peak jump power in watts. Age is supplied only as a
1-to-6 decade bin, not as exact years; the bins are consecutive ten-year bands,
and given the reported mean age of about 25 years, Decade 2 corresponds to
participants in their twenties. Because the file does not state the boundaries,
the bands are labelled by bin number rather than by age in years.

**Predictor budget.** With `r nrow(dat_raw)` participants and a conventional
allowance of 10 to 15 observations per estimated parameter, this dataset supports
roughly `r floor(nrow(dat_raw)/15)` to `r floor(nrow(dat_raw)/10)` parameters,
and the smaller end of that range is the honest target for a sample recruited by
convenience. The model below spends **`r length(predictors) + 1` parameters**
(sex 1, age band 2, body mass 1, peak power 1), i.e.
`r fmt(nrow(dat_raw) / (length(predictors) + 1), 0)` observations per parameter.
Entering `age_decade` as an unmodified six-level factor would cost five
parameters for age alone and take the model to eight; it is also not estimable in
any useful sense, because decades 3, 4, 5 and 6 contain
`r sum(dat$age_decade == 3)`, `r sum(dat$age_decade == 4)`,
`r sum(dat$age_decade == 5)` and `r sum(dat$age_decade == 6)` participants. Those
four bands are therefore collapsed into a single "Decades 3-6" level
(n = `r sum(dat$age_decade >= 3)`). The budget was fixed before any model was
fitted.

**Choosing between the two power variables.** `peak_power_w` and
`relative_peak_power` must never appear in the same model, because
`relative_peak_power` is exactly `peak_power_w / body_mass_kg` rounded to two
decimal places: across all `r nrow(dat_raw)` rows the largest discrepancy is
`r rpp_err`. The redundancy runs deeper than that. `peak_power_w` was not
measured; it was calculated from the Sayers (1999) equation, *peak power (W) =
51.9 x jump height (cm) + 48.9 x body mass (kg) - 2007*, so body mass is already
inside it with a fixed coefficient. Inverting the equation recovers each
participant's jump height, and the recovered values sit on a half-inch grid
(largest deviation `r fmt(halfinch_err, 5)` of a half-inch, with only
`r n_jump_values` distinct heights in the file), which confirms the
reconstruction; feeding those heights back through the equation reproduces
`peak_power_w` to within `r pp_err` W. Three parameterisations are therefore
available: body mass with absolute power, body mass with relative power, or
power alone. We use **body mass and absolute peak power**, for three reasons.
First, it is the parameterisation of the source paper, so the results here are
directly comparable with the published analysis. Second, because power is a
fixed linear combination of body mass and jump height, a model containing body
mass and absolute power is *algebraically the same model* as one containing body
mass and jump height, with the power coefficient equal to the jump-height
coefficient divided by 51.9; this is demonstrated in the sensitivity section and
it gives the adjusted power coefficient a clean physical reading. Third,
`relative_peak_power` is a ratio whose denominator is another candidate
predictor, and entering a ratio alongside its own denominator produces
coefficients that answer no clear question. The cost of this choice is a
correlation of `r fmt(cor(dat$body_mass_kg, dat$peak_power_w), 2)` between body
mass and peak power, which inflates their standard errors; the resulting variance
inflation factors are reported with the diagnostics.

**Statistical analysis.** `lm()`, justified because the outcome is a continuous
measurement with no floor or ceiling. We fitted a univariable model for each
predictor first, then a single multivariable model containing all of them.
Predictors were chosen in advance on subject-matter grounds, not by significance.
Significance threshold 0.05; estimates are reported with 95% confidence
intervals. Assumptions were checked with `performance::check_model()`,
collinearity with `check_collinearity()`, and influence with Cook's distance;
because non-constant variance was flagged we additionally report HC3
heteroscedasticity-consistent standard errors. Analysis used
`r R.version.string` with tidyverse, gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped). Every variable in the file is 100%
complete, so listwise deletion costs nothing here and the complete-case N equals
the recruited N.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(missing = "no") |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

The sample is young and female-weighted: `r sum(dat_model$sex == "Female")`
women and `r sum(dat_model$sex == "Male")` men, with
`r sum(dat_model$age_band == "Decade 2")` of `r nrow(dat_model)`
(`r fmt(100 * mean(dat_model$age_band == "Decade 2"), 0)`%) in the modal age
band and only `r sum(dat_model$age_band == "Decades 3-6")` participants in the
whole of decades 3 to 6. The bone strength index has a median of
`r fmt(median(dat_model$bone_strength_index), 1)` mg/mm4 with a long right tail
reaching `r fmt(max(dat_model$bone_strength_index), 1)`, roughly five times the
minimum of `r fmt(min(dat_model$bone_strength_index), 1)`.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Trabecular bone strength index against estimated peak jump power, by sex."}
ggplot(dat_model, aes(peak_power_w, bone_strength_index, colour = sex)) +
  geom_point(alpha = 0.7, size = 1.9) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.9) +
  scale_colour_manual(values = c(Female = "#C46B4A", Male = "#3F8F7A")) +
  labs(x = "Estimated peak jump power (W)",
       y = "Bone strength index (mg/mm4)", colour = "Sex")
```

Bone strength index rises steadily with estimated peak power in both sexes, and
the two fitted lines are close to parallel and close to each other, which
suggests that much of the crude difference between men and women is a difference
in power rather than a separate sex effect.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv

utid <- map_dfr(predictors, function(v) {
  tidy(lm(reformulate(v, outcome), data = dat_model), conf.int = TRUE) |>
    filter(term != "(Intercept)")
})
ug <- function(term, d = 2) {
  r <- utid[utid$term == term, ]
  sprintf("%s mg/mm4 (95%% CI %s to %s)", fmt(r$estimate, d),
          fmt(r$conf.low, d), fmt(r$conf.high, d))
}
up <- function(term) pv(utid$p.value[utid$term == term])
```

Every predictor except the youngest age band shows a crude association. Each
additional watt of estimated peak power is associated with
`r ug("peak_power_w", 4)` higher bone strength index (p `r up("peak_power_w")`),
which amounts to `r fmt(1000 * utid$estimate[utid$term == "peak_power_w"], 1)`
mg/mm4 per 1000 W; each additional kilogram of body mass with
`r ug("body_mass_kg")` (p `r up("body_mass_kg")`); and men average
`r ug("sexMale")` higher than women (p `r up("sexMale")`). Participants in
decades 3 to 6 average `r ug("age_bandDecades 3-6")` relative to decade 2
(p `r up("age_bandDecades 3-6")`), while decade 1 is indistinguishable from
decade 2 (`r ug("age_bandDecade 1")`, p `r up("age_bandDecade 1")`).

## Why adjustment is needed

**Body mass** confounds the crude peak-power association in a way that is
unusually explicit in this dataset: heavier participants have higher estimated
peak power partly because they jumped higher, but also partly by construction,
since body mass enters the Sayers equation with a coefficient of 48.9 W per
kilogram. Body mass is separately associated with bone strength because bone
adapts to the load it habitually carries, so a crude power coefficient is
carrying a mass effect on both counts. **Sex** confounds the same association
from the other direction: men in this sample are heavier (median
`r fmt(median(dat_model$body_mass_kg[dat_model$sex == "Male"]), 1)` versus
`r fmt(median(dat_model$body_mass_kg[dat_model$sex == "Female"]), 1)` kg) and
more powerful (median
`r fmt(median(dat_model$peak_power_w[dat_model$sex == "Male"]), 0)` versus
`r fmt(median(dat_model$peak_power_w[dat_model$sex == "Female"]), 0)` W), and
men have larger, stronger tibiae for reasons of skeletal size and hormonal
history that have nothing to do with jumping, so the crude power slope is partly
a comparison of men with women. **Age band** works on both: the older
participants here are both less powerful (mean peak power
`r fmt(mean(dat_model$peak_power_w[dat_model$age_band == "Decades 3-6"]), 0)` W
in decades 3-6 versus
`r fmt(mean(dat_model$peak_power_w[dat_model$age_band == "Decade 2"]), 0)` W in
decade 2) and losing trabecular bone with age. Mutual adjustment lets each
coefficient be read as a comparison between participants of the same sex, age
band and body mass.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv

tid <- tidy(fit, conf.int = TRUE)
g <- function(term, d = 2) {
  r <- tid[tid$term == term, ]
  sprintf("%s mg/mm4 (95%% CI %s to %s)", fmt(r$estimate, d),
          fmt(r$conf.low, d), fmt(r$conf.high, d))
}
gp <- function(term) pv(tid$p.value[tid$term == term])
```

```{r fit-stats}
s <- summary(fit)
tibble(
  Statistic = c("Complete-case N", "Parameters (excl. intercept)",
                "Observations per parameter",
                "R-squared", "Adjusted R-squared", "Residual SE (mg/mm4)"),
  Value = c(as.character(nrow(model.frame(fit))),
            as.character(length(coef(fit)) - 1L),
            fmt(nrow(model.frame(fit)) / (length(coef(fit)) - 1L), 1),
            fmt(s$r.squared, 3), fmt(s$adj.r.squared, 3), fmt(s$sigma, 2))
) |>
  gt() |>
  tab_header(title = md("**Model fit: multivariable linear model for bone strength index**"))
```

The model explains `r fmt(100 * s$r.squared, 1)`% of the variance in the bone
strength index (adjusted R-squared `r fmt(s$adj.r.squared, 3)`), and the gap
between the two is small because only `r length(coef(fit)) - 1L` parameters are
spent on `r nrow(model.frame(fit))` observations. **Estimated peak power is the
one predictor that survives adjustment for body mass**: among participants of the
same sex, age band and body mass, each additional watt is associated with
`r g("peak_power_w", 4)` (p `r gp("peak_power_w")`), i.e.
`r fmt(1000 * tid$estimate[tid$term == "peak_power_w"], 1)` mg/mm4 per 1000 W.
Men average `r g("sexMale")` more than women of the same age, mass and power
(p `r gp("sexMale")`). Body mass reverses sign and becomes null
(`r g("body_mass_kg")`, p `r gp("body_mass_kg")`), and the deficit in decades 3
to 6 falls to `r g("age_bandDecades 3-6")` (p `r gp("age_bandDecades 3-6")`),
just short of the 0.05 threshold. The negative adjusted body-mass coefficient
must be read with care and is explained in the sensitivity section: holding
*estimated* peak power constant while increasing body mass by one kilogram
implicitly requires the participant to have jumped lower, so this coefficient is
not "heavier people have weaker bone". These are associations among the
participants observed, not effects of intervening.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

Two estimates move a long way. **Body mass** collapses from
`r ug("body_mass_kg")` crude to `r g("body_mass_kg")` adjusted, crossing the null
and losing significance entirely (p `r up("body_mass_kg")` to
p `r gp("body_mass_kg")`): almost the whole crude association between body mass
and trabecular bone strength is shared with peak power, which is unsurprising
given that the two are correlated at
`r fmt(cor(dat_model$body_mass_kg, dat_model$peak_power_w), 2)` and that body
mass is a component of the power calculation. **Sex** halves, from
`r ug("sexMale")` to `r g("sexMale")`, moving toward the null but remaining
clearly significant, so roughly half of the crude male advantage is explained by
men being heavier and more powerful and half is not. The **decades 3-6** contrast
also attenuates markedly, from `r ug("age_bandDecades 3-6")` to
`r g("age_bandDecades 3-6")`, and crosses from significant
(p `r up("age_bandDecades 3-6")`) to non-significant
(p `r gp("age_bandDecades 3-6")`) - the older participants have lower bone
strength largely because they are less powerful, not independently of it. By
contrast **peak power barely moves**, from `r ug("peak_power_w", 4)` crude to
`r g("peak_power_w", 4)` adjusted; it loses about
`r fmt(100 * (1 - tid$estimate[tid$term == "peak_power_w"] / utid$estimate[utid$term == "peak_power_w"]), 0)`%
of its magnitude and gains width in its interval because it is competing with
body mass for the same variance, but it stays significant throughout. The
**decade 1** contrast is negligible in both columns.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r collin}
check_collinearity(fit)
```

```{r influence}
cd <- cooks.distance(fit)
i_max <- which.max(cd)
fit_drop <- lm(reformulate(predictors, outcome), data = dat_model[-i_max, ])
tibble(
  Term = names(coef(fit)),
  `Full sample` = fmt(coef(fit), 4),
  `Most influential dropped` = fmt(coef(fit_drop), 4),
  `Change (%)` = fmt(100 * (coef(fit_drop) - coef(fit)) / abs(coef(fit)), 1)
) |>
  filter(Term != "(Intercept)") |>
  gt() |>
  tab_header(title = md(paste0(
    "**Table 3. Coefficients with and without the single most influential ",
    "observation (Cook's D = ", fmt(max(cd), 3), ").**")))
```

```{r het-check}
het <- check_heteroscedasticity(fit)
print(het)
nor <- check_normality(fit)
print(nor)
```

```{r robust-se}
rob   <- lmtest::coeftest(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
robci <- lmtest::coefci(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
tibble(
  Term = rownames(rob),
  Estimate = fmt(rob[, 1], 4),
  `SE (OLS)` = fmt(summary(fit)$coefficients[, 2], 4),
  `SE (HC3)` = fmt(rob[, 2], 4),
  `95% CI (HC3)` = paste0(fmt(robci[, 1], 3), " to ", fmt(robci[, 2], 3)),
  `p (HC3)` = fmt(rob[, 4], 3)
) |>
  filter(Term != "(Intercept)") |>
  gt() |>
  tab_header(title = md("**Table 4. HC3 heteroscedasticity-consistent standard errors.**"))
```

**Linearity and homogeneity of variance.** The residual-versus-fitted panel shows
a mild fan: spread grows with the fitted value, and
`check_heteroscedasticity()` flags non-constant error variance
(p `r pv(as.numeric(het))`). This is the expected consequence of a
right-skewed outcome whose variability scales with its level. Table 4 therefore
reports HC3 robust standard errors, and the substantive conclusions are
unchanged: peak power keeps p `r pv(rob["peak_power_w", 4])`, and the decades 3-6
contrast in fact becomes significant under robust inference
(p `r pv(rob["age_bandDecades 3-6", 4])`) because its OLS standard error was
conservative.

**Normality of residuals.** The normal-QQ panel is close to the reference line
and `check_normality()` does **not** reject (p `r pv(as.numeric(nor))`), so
despite the skewed raw outcome the residuals are acceptably normal. This is the
one diagnostic that passes cleanly.

**Collinearity.** The variance inflation factors are
`r fmt(check_collinearity(fit)$VIF[check_collinearity(fit)$Term == "peak_power_w"], 2)`
for peak power and
`r fmt(check_collinearity(fit)$VIF[check_collinearity(fit)$Term == "body_mass_kg"], 2)`
for body mass, with sex at
`r fmt(check_collinearity(fit)$VIF[check_collinearity(fit)$Term == "sex"], 2)` and
age band at
`r fmt(check_collinearity(fit)$VIF[check_collinearity(fit)$Term == "age_band"], 2)`.
These are below the conventional threshold of 5 but they are the direct price of
the parameterisation chosen in Methods, and they are the reason the body-mass
coefficient has a confidence interval wide enough to span zero comfortably.

**Influential observations.** This matters more than usual at n =
`r nrow(model.frame(fit))`, where one participant is 0.7% of the sample. The
largest Cook's distance is `r fmt(max(cd), 3)`, far below the conventional
concern threshold of 0.5, and the largest leverage is
`r fmt(max(hatvalues(fit)), 3)`; `r sum(cd > 4 / nrow(dat_model))` observations
exceed the stricter 4/n rule but none dominates. Table 3 confirms this directly:
refitting without the single most influential participant changes the peak-power
coefficient by only
`r fmt(abs(100 * (coef(fit_drop)["peak_power_w"] - coef(fit)["peak_power_w"]) / abs(coef(fit)["peak_power_w"])), 1)`%
and the sex coefficient by
`r fmt(abs(100 * (coef(fit_drop)["sexMale"] - coef(fit)["sexMale"]) / abs(coef(fit)["sexMale"])), 1)`%.
The only coefficient that swings proportionally is the decade 1 contrast, which
moves from `r fmt(coef(fit)["age_bandDecade 1"], 2)` to
`r fmt(coef(fit_drop)["age_bandDecade 1"], 2)` mg/mm4 - a large percentage change
only because the estimate is close to zero to begin with, and immaterial next to
its standard error of `r fmt(s$coefficients["age_bandDecade 1", 2], 2)`.

## Sensitivity: alternative parameterisations

```{r sensitivity}
alt <- list(
  `Chosen: body mass + absolute power`  = c("sex", "age_band", "body_mass_kg", "peak_power_w"),
  `Equivalent: body mass + jump height` = c("sex", "age_band", "body_mass_kg", "jump_cm"),
  `Alternative: body mass + relative power` = c("sex", "age_band", "body_mass_kg",
                                                "relative_peak_power"),
  `Absolute power only (no body mass)`  = c("sex", "age_band", "peak_power_w"),
  `WRONG: all three power/mass terms`   = c("sex", "age_band", "body_mass_kg",
                                            "peak_power_w", "relative_peak_power")
)
alt_fits <- map(alt, ~ lm(reformulate(.x, outcome), data = dat))

map2_dfr(names(alt_fits), alt_fits, function(nm, f) {
  tibble(
    Model = nm,
    `Parameters` = length(coef(f)) - 1L,
    `R-squared` = fmt(summary(f)$r.squared, 3),
    `Adjusted R-squared` = fmt(summary(f)$adj.r.squared, 3),
    `Residual SE` = fmt(summary(f)$sigma, 2),
    `Max VIF` = fmt(max(check_collinearity(f)$VIF), 1)
  )
}) |>
  gt() |>
  tab_header(title = md("**Table 5. The same data under five parameterisations of body mass and power.**"))
```

```{r reparam}
f_pp <- alt_fits[["Chosen: body mass + absolute power"]]
f_jh <- alt_fits[["Equivalent: body mass + jump height"]]
f_bad <- alt_fits[["WRONG: all three power/mass terms"]]
b_pp <- coef(f_pp)["peak_power_w"]
b_jh <- coef(f_jh)["jump_cm"]
```

Table 5 makes the argument of the Methods section concrete. The chosen model and
the jump-height model have **identical** R-squared
(`r fmt(summary(f_pp)$r.squared, 6)` versus
`r fmt(summary(f_jh)$r.squared, 6)`) and identical residual standard error,
because they are the same model written twice: multiplying the peak-power
coefficient by 51.9 gives `r fmt(51.9 * b_pp, 8)`, against a jump-height
coefficient of `r fmt(b_jh, 8)` mg/mm4 per centimetre - the same number to eight
decimal places, not merely a similar one. Read this way, the adjusted "power"
effect is really a jump-height effect, and the negative body-mass coefficient in
the main model is an artefact of holding a mass-containing composite fixed: in
the jump-height parameterisation, body mass is associated with
`r fmt(coef(f_jh)["body_mass_kg"], 2)` mg/mm4 per kilogram
(p `r pv(summary(f_jh)$coefficients["body_mass_kg", 4])`), which is positive, as
biology would expect.

Substituting relative for absolute power changes little
(R-squared `r fmt(summary(alt_fits[[3]])$r.squared, 3)` against
`r fmt(summary(f_pp)$r.squared, 3)`) and lowers the maximum VIF, so a student who
chose that parameterisation would reach the same substantive conclusion by a
different route. The last row is the trap: entering body mass, absolute power
*and* relative power together buys almost nothing in fit
(R-squared `r fmt(summary(f_bad)$r.squared, 3)`) while pushing the maximum
variance inflation factor to `r fmt(max(check_collinearity(f_bad)$VIF), 0)`,
whereupon peak power loses significance
(p `r pv(summary(f_bad)$coefficients["peak_power_w", 4])`) and relative power
takes a meaningless negative coefficient of
`r fmt(coef(f_bad)["relative_peak_power"], 2)`. Nothing in that column should be
interpreted.

```{r log-sens}
fit_log <- lm(reformulate(predictors, paste0("log(", outcome, ")")), data = dat_model)
het_log <- check_heteroscedasticity(fit_log)
```

Finally, because the outcome is right-skewed and the variance is non-constant, we
refitted on the log scale. Heteroscedasticity is resolved
(p `r pv(as.numeric(het_log))`), the fit is essentially unchanged
(R-squared `r fmt(summary(fit_log)$r.squared, 3)`), and the conclusions hold:
each 1000 W of peak power is associated with
`r fmt(100 * (exp(1000 * coef(fit_log)["peak_power_w"]) - 1), 1)`% higher bone
strength index (p `r pv(summary(fit_log)$coefficients["peak_power_w", 4])`) and
men are `r fmt(100 * (exp(coef(fit_log)["sexMale"]) - 1), 1)`% higher than women
(p `r pv(summary(fit_log)$coefficients["sexMale", 4])`). The natural-scale model
is retained as primary because its coefficients are in the units clinicians and
the source paper use, with HC3 standard errors doing the inferential work.

# Discussion

Among these `r nrow(dat_model)` healthy adults, **estimated peak leg power is the
characteristic most strongly and independently associated with trabecular bone
strength at the distal tibia**: about
`r fmt(1000 * tid$estimate[tid$term == "peak_power_w"], 0)` mg/mm4 per 1000 W,
holding sex, age band and body mass constant. Men average about
`r fmt(tid$estimate[tid$term == "sexMale"], 0)` mg/mm4 more than women who are
alike on the other three. The crude association with body mass does not survive
adjustment; once the sensitivity analysis is read alongside it, the picture is
that *how hard a person can push off the ground*, rather than *how much they
weigh*, is what tracks with the strength of the bone end. Practically, this is
consistent with the idea that high-rate loading rather than static weight-bearing
is what the distal tibia responds to, which is the rationale behind jumping and
plyometric exercise programmes for bone. It does not demonstrate that such
programmes work.

**Limitations.** The design is **cross-sectional**, so temporality is unknown and
reverse causation is entirely plausible: people with stronger tibiae may simply
be the people who have always jumped, played sport and stayed active, in which
case the arrow runs from bone to power as much as from power to bone. The sample
is a **convenience sample of one university's staff and students**, young
(`r fmt(100 * mean(dat_model$age_band == "Decade 2"), 0)`% in a single decade
band) and healthy, so nothing here generalises to older adults or to anyone at
real fracture risk - the age coefficients in particular rest on
`r sum(dat_model$age_band == "Decades 3-6")` participants above decade 2 and are
wide. **Peak power was estimated, not measured**: it is the Sayers regression
equation applied to jump height and body mass, and the jump heights themselves
are recorded to the nearest half inch, giving only `r n_jump_values` distinct
values. Measurement error in a predictor attenuates its coefficient toward the
null, so the true association with genuinely measured peak power is if anything
*stronger* than reported here; it also means the "power" variable is partly a
restatement of body mass, which is why the two compete
(VIF `r fmt(max(check_collinearity(fit)$VIF), 1)`). **Age is a decade bin, not a
number**, so any age gradient is measured crudely and decades 3 to 6 had to be
merged. Physical activity, calcium and vitamin D intake, smoking, menstrual and
hormonal history, tibial length and limb dominance are **not measured at all**,
and several of them plausibly cause both muscle power and bone strength, so
residual confounding is certain. Residual variance is non-constant; HC3 standard
errors address inference but not the model's fit to the upper tail. All estimates
are associations, not causal effects.

**Future work.** A prospective study with directly measured force-plate power,
exact ages and an activity history would settle whether the power-bone
association is a training effect or a shared constitutional one.

# Feasibility notes

*NOT part of a student report.*

- **USABLE.** This is a clean, well-behaved linear outcome with a genuine
  finding, no missing data and a defensible model at
  `r fmt(nrow(dat_model) / (length(coef(fit)) - 1L), 1)` observations per
  parameter. It is one of the simpler datasets in the catalogue, and the
  simplicity is the point: the whole difficulty is in choosing the predictor
  parameterisation, not in cleaning.
- **MAJOR trap: `peak_power_w` and `relative_peak_power` are the same variable.**
  `relative_peak_power` is exactly `peak_power_w / body_mass_kg` rounded to 2 dp
  (largest discrepancy across all `r nrow(dat_raw)` rows: `r rpp_err`). Putting
  both in one model, with body mass, gives VIFs of
  `r fmt(max(check_collinearity(f_bad)$VIF), 0)`,
  `r fmt(sort(check_collinearity(f_bad)$VIF, decreasing = TRUE)[2], 0)` and
  `r fmt(sort(check_collinearity(f_bad)$VIF, decreasing = TRUE)[3], 0)`, flips
  the sign of relative power and kills the significance of peak power. The
  catalogue's own suggested predictor list contains **both**, so a student
  following it literally will hit this. This must be flagged in `meta.json`.
- **MAJOR trap: `peak_power_w` already contains `body_mass_kg`.** It is the
  Sayers (1999) equation, `PP = 51.9*jump_cm + 48.9*body_mass - 2007`, inverted;
  the recovered jump heights land on a half-inch grid to within
  `r fmt(halfinch_err, 5)` and reproduce `peak_power_w` to
  `r pp_err` W. Consequences: (a) `cor(body_mass, peak_power) =
  `r fmt(cor(dat$body_mass_kg, dat$peak_power_w), 2)``, VIF
  `r fmt(max(check_collinearity(fit)$VIF), 1)`; (b) the adjusted body-mass
  coefficient is **negative** (`r fmt(coef(fit)["body_mass_kg"], 2)`) purely
  because it is conditioning on a mass-containing composite, which a student will
  almost certainly misinterpret as "heavier people have weaker bone". The source
  paper reports the same negative standardised coefficient (-0.102) without
  explaining it. This is the single most likely way a student writes a
  confidently wrong discussion.
- **MAJOR: `age_decade` cannot be used as a six-level factor.** Cell sizes are
  `r paste(as.integer(table(dat$age_decade)), collapse = ", ")` for decades 1
  through 6, so decade 6 holds `r sum(dat$age_decade == 6)` participants and
  there are no men at all in decades 4 or 6. As a six-level factor the largest
  standard error is
  `r fmt(max(summary(lm(bone_strength_index ~ sex + factor(age_decade) + body_mass_kg + peak_power_w, dat))$coefficients[, 2]), 1)`
  mg/mm4 and the extra three degrees of freedom buy nothing (F test against the
  collapsed model p =
  `r fmt(anova(fit, lm(bone_strength_index ~ sex + factor(age_decade) + body_mass_kg + peak_power_w, dat))$"Pr(>F)"[2], 2)`).
  Collapsing to three bands is required, not optional.
- **MINOR: heteroscedasticity** (p `r pv(as.numeric(het))`). Normality of
  residuals passes (p `r pv(as.numeric(nor))`), so this is the milder of the
  two problems and a log transform fixes it (p `r pv(as.numeric(het_log))`)
  without changing conclusions. A student who reports the untransformed model
  without HC3 has one failed check on the record.
- **MINOR: no influential-point problem.** Maximum Cook's distance
  `r fmt(max(cd), 3)`, maximum leverage `r fmt(max(hatvalues(fit)), 3)`, and
  dropping the most influential participant moves the peak-power coefficient by
  `r fmt(abs(100 * (coef(fit_drop)["peak_power_w"] - coef(fit)["peak_power_w"]) / abs(coef(fit)["peak_power_w"])), 1)`%.
  Worth checking because n = `r nrow(dat_model)`, but it comes out clean for this
  outcome (contrast `polar_strength_strain_index`, where the same check moves a
  coefficient by 22%).
- **MINOR: no binary outcome exists in this file, and none should be
  manufactured.** A student will be tempted to split the bone strength index at
  its median or a percentile to run a logistic model. There is **no clinical
  threshold** for BSI in healthy adults - unlike, say, a T-score, this index has
  no reference population or diagnostic cut-point - so any dichotomy would be
  arbitrary, would throw away the information that gives this analysis its
  R-squared of `r fmt(s$r.squared, 2)`, and would produce an "outcome" that
  answers no question anyone asks. The dataset supports a **linear project
  only**, and `index.html` correctly says so.
- **MINOR: the two outcomes are correlated at
  `r fmt(cor(dat$bone_strength_index, dat$polar_strength_strain_index), 2)`**
  (Pearson), i.e. they share
  `r fmt(100 * cor(dat$bone_strength_index, dat$polar_strength_strain_index)^2, 0)`%
  of their variance. They are *not* redundant: they measure different bone
  compartments at different sites, the residual correlation after this model is
  only `r fmt(cor(resid(fit), resid(lm(polar_strength_strain_index ~ sex + age_band + body_mass_kg + peak_power_w, dat))), 2)`,
  and they give different answers - peak power dominates here while body mass
  dominates for the shaft. Offering both is justified.
- **Sample size is the binding constraint.** `r nrow(dat_raw)` participants and
  five candidate predictors, two of which cannot coexist, leaves essentially one
  sensible model. There is no room for interactions, splines or subgroup
  analyses, and students should be told that plainly rather than discovering it.
