---
title: "Cortical bone strength at the tibial shaft in healthy adults"
subtitle: "Dataset 12: Muscle power and tibial bone strength in healthy adults"
author: "RSR feasibility analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse)
library(gtsummary)
library(gt)
library(performance)
library(see)
library(broom)

theme_set(theme_minimal(base_size = 12))
theme_gtsummary_compact()
```

```{r data}
# knitr sets the working directory to this file's folder, so the data sits two
# levels up. here() is not used because the repo root is above this catalogue.
dat_raw <- read_csv(file.path("..", "..", "data", "tibial_bone_strength.csv"),
                    show_col_types = FALSE)

# Outcome and predictors are fixed here so every downstream chunk refers back to
# these two objects rather than repeating variable names. The predictor set is
# capped at five parameters because there are only 142 participants, and
# `relative_peak_power` is deliberately excluded because it is an exact function
# of `peak_power_w` and `body_mass_kg` (verified below).
outcome    <- "polar_strength_strain_index"
predictors <- c("sex", "age_band", "body_mass_kg", "peak_power_w")

lab <- function(x, l) { attr(x, "label") <- l; x }
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
pv  <- function(p) ifelse(p < 0.001, "<0.001", paste0("= ", fmt(p, 3)))

# --- Verifying the algebraic relationships between the shipped columns -------
# 1. relative_peak_power is peak_power_w / body_mass_kg, rounded to 2 dp.
rpp_err <- max(abs(dat_raw$relative_peak_power -
                     round(dat_raw$peak_power_w / dat_raw$body_mass_kg, 2)))

# 2. peak_power_w is the Sayers (1999) countermovement-jump equation,
#    PP = 51.9 * jump height (cm) + 48.9 * body mass (kg) - 2007.
#    Inverting it recovers the jump height each participant actually achieved.
jump_cm  <- (dat_raw$peak_power_w - 48.9 * dat_raw$body_mass_kg + 2007) / 51.9
pp_err   <- max(abs(dat_raw$peak_power_w -
                      (51.9 * round(jump_cm, 2) + 48.9 * dat_raw$body_mass_kg - 2007)))
# The recovered heights land on a half-inch grid, confirming the reconstruction.
halfinch_err <- max(abs(jump_cm / 2.54 * 2 - round(jump_cm / 2.54 * 2)))
n_jump_values <- n_distinct(round(jump_cm, 2))

dat <- dat_raw |>
  mutate(
    jump_cm = lab(jump_cm, "Recovered jump height (cm)"),
    # Deliberate reference levels: the largest or the clinically "unexposed" group.
    sex = lab(factor(sex, levels = c("female", "male"), labels = c("Female", "Male")),
              "Sex"),
    # Decades 3 to 6 hold 9, 6, 5 and 2 participants respectively, far too few to
    # estimate separately, so they are collapsed. Decade 2 is the reference
    # because it is by far the largest band.
    age_band = lab(factor(case_when(
      age_decade == 1 ~ "Decade 1",
      age_decade == 2 ~ "Decade 2",
      age_decade >= 3 ~ "Decades 3-6"),
      levels = c("Decade 2", "Decade 1", "Decades 3-6")),
      "Age band (10-year bins)"),
    body_mass_kg = lab(body_mass_kg, "Body mass (kg)"),
    peak_power_w = lab(peak_power_w, "Estimated peak jump power (W)"),
    relative_peak_power = lab(relative_peak_power, "Relative peak power (W/kg)"),
    polar_strength_strain_index = lab(polar_strength_strain_index,
                                      "Polar strength-strain index (mm3)")
  )

dat_model <- dat |> select(all_of(c(outcome, predictors))) |> drop_na()
n_dropped <- nrow(dat) - nrow(dat_model)
```

# Introduction

This is a cross-sectional laboratory study of `r nrow(dat_raw)` healthy adults
recruited as a convenience sample from the faculty, staff and students of
California State University East Bay, reported by Nickols-Richardson and
colleagues in the *Journal of Musculoskeletal and Neuronal Interactions* (2022).
Each participant performed a countermovement vertical jump, from which peak leg
power was estimated, and had the tibia scanned by peripheral quantitative
computed tomography (pQCT). The outcome here is the **polar strength-strain
index at the 50% tibial site**, a geometric index of the shaft's resistance to
bending and torsion; it is the standard pQCT summary of *cortical* bone strength
and it is the property that governs whether a long bone survives a bending load,
so it complements the trabecular index measured at the distal end.

**Aim.** Among these healthy adults, we examined which of sex, age band, body
mass and estimated peak jump power are independently associated with the polar
strength-strain index at the tibial shaft, mutually adjusting for the others.

This is a *Purpose 3* analysis: several predictors are of interest and the
question is which remain associated with the outcome after mutual adjustment.

# Methods

**Data source.** A convenience sample of `r nrow(dat_raw)` faculty, staff and
students at a mid-sized regional university in the United States, measured once
in a laboratory session. The published report gives a mean age of about 25 years
and describes the sample as healthy and free of conditions affecting bone. The
file shipped with this catalogue holds `r ncol(dat_raw)` columns and
`r nrow(dat_raw)` rows, and nothing is missing.

**Variables.** The outcome is `polar_strength_strain_index`, the polar
strength-strain index at the 50% tibial site in mm3, treated as continuous
(range `r fmt(min(dat$polar_strength_strain_index), 0)` to
`r fmt(max(dat$polar_strength_strain_index), 0)`). Predictors are sex (reference
female, the larger group), age band (reference Decade 2, the modal band), body
mass in kilograms, and estimated peak jump power in watts. Age is supplied only
as a 1-to-6 decade bin, not as exact years; the bins are consecutive ten-year
bands, and given the reported mean age of about 25 years, Decade 2 corresponds to
participants in their twenties. Because the file does not state the boundaries,
the bands are labelled by bin number rather than by age in years.

**Predictor budget.** With `r nrow(dat_raw)` participants and a conventional
allowance of 10 to 15 observations per estimated parameter, this dataset supports
roughly `r floor(nrow(dat_raw)/15)` to `r floor(nrow(dat_raw)/10)` parameters,
and the smaller end of that range is the honest target for a sample recruited by
convenience. The model below spends **`r length(predictors) + 1` parameters**
(sex 1, age band 2, body mass 1, peak power 1), i.e.
`r fmt(nrow(dat_raw) / (length(predictors) + 1), 0)` observations per parameter.
Entering `age_decade` as an unmodified six-level factor would cost five
parameters for age alone and take the model to eight; it is also not estimable in
any useful sense, because decades 3, 4, 5 and 6 contain
`r sum(dat$age_decade == 3)`, `r sum(dat$age_decade == 4)`,
`r sum(dat$age_decade == 5)` and `r sum(dat$age_decade == 6)` participants. Those
four bands are therefore collapsed into a single "Decades 3-6" level
(n = `r sum(dat$age_decade >= 3)`). The budget was fixed before any model was
fitted.

**Choosing between the two power variables.** `peak_power_w` and
`relative_peak_power` must never appear in the same model, because
`relative_peak_power` is exactly `peak_power_w / body_mass_kg` rounded to two
decimal places: across all `r nrow(dat_raw)` rows the largest discrepancy is
`r rpp_err`. The redundancy runs deeper than that. `peak_power_w` was not
measured; it was calculated from the Sayers (1999) equation, *peak power (W) =
51.9 x jump height (cm) + 48.9 x body mass (kg) - 2007*, so body mass is already
inside it with a fixed coefficient. Inverting the equation recovers each
participant's jump height, and the recovered values sit on a half-inch grid
(largest deviation `r fmt(halfinch_err, 5)` of a half-inch, with only
`r n_jump_values` distinct heights in the file), which confirms the
reconstruction; feeding those heights back through the equation reproduces
`peak_power_w` to within `r pp_err` W. Three parameterisations are therefore
available: body mass with absolute power, body mass with relative power, or
power alone. We use **body mass and absolute peak power**, for three reasons.
First, it is the parameterisation of the source paper, so the results here are
directly comparable with the published analysis, and it is the same
parameterisation used for the companion analysis of the distal-tibia trabecular
index, so the two outcomes can be compared coefficient for coefficient. Second,
because power is a fixed linear combination of body mass and jump height, a model
containing body mass and absolute power is *algebraically the same model* as one
containing body mass and jump height, with the power coefficient equal to the
jump-height coefficient divided by 51.9; this is demonstrated in the sensitivity
section and it gives the adjusted power coefficient a clean physical reading.
Third, `relative_peak_power` is a ratio whose denominator is another candidate
predictor, and entering a ratio alongside its own denominator produces
coefficients that answer no clear question. The choice matters more for this
outcome than for the trabecular one, because body mass is the dominant predictor
of cortical bone strength here, and the parameterisation determines how the mass
effect is split between the two terms. The cost is a correlation of
`r fmt(cor(dat$body_mass_kg, dat$peak_power_w), 2)` between body mass and peak
power, which inflates their standard errors; the resulting variance inflation
factors are reported with the diagnostics.

**Statistical analysis.** `lm()`, justified because the outcome is a continuous
measurement with no floor or ceiling. We fitted a univariable model for each
predictor first, then a single multivariable model containing all of them.
Predictors were chosen in advance on subject-matter grounds, not by significance.
Significance threshold 0.05; estimates are reported with 95% confidence
intervals. Assumptions were checked with `performance::check_model()`,
collinearity with `check_collinearity()`, and influence with Cook's distance;
because non-constant variance and non-normal residuals were flagged we
additionally report HC3 heteroscedasticity-consistent standard errors and a
log-transformed sensitivity model. Analysis used `r R.version.string` with
tidyverse, gtsummary and performance.

**Missing data.** Complete-case analysis: `r nrow(dat_model)` of `r nrow(dat)`
rows were analysed (`r n_dropped` dropped). Every variable in the file is 100%
complete, so listwise deletion costs nothing here and the complete-case N equals
the recruited N.

# Results

## Sample characteristics

```{r table-one}
dat_model |>
  tbl_summary(missing = "no") |>
  add_n() |>
  modify_caption("**Table 1. Sample characteristics.**")
```

The sample is young and female-weighted: `r sum(dat_model$sex == "Female")`
women and `r sum(dat_model$sex == "Male")` men, with
`r sum(dat_model$age_band == "Decade 2")` of `r nrow(dat_model)`
(`r fmt(100 * mean(dat_model$age_band == "Decade 2"), 0)`%) in the modal age
band and only `r sum(dat_model$age_band == "Decades 3-6")` participants in the
whole of decades 3 to 6. The polar strength-strain index has a median of
`r fmt(median(dat_model$polar_strength_strain_index), 0)` mm3 and a range from
`r fmt(min(dat_model$polar_strength_strain_index), 0)` to
`r fmt(max(dat_model$polar_strength_strain_index), 0)`, a fourfold spread that is
narrower in relative terms than the trabecular index measured on the same
participants.

## Exploratory figure

```{r explore, fig.cap="Figure 1. Polar strength-strain index against body mass and against estimated peak jump power, by sex."}
dat_model |>
  select(polar_strength_strain_index, sex, body_mass_kg, peak_power_w) |>
  pivot_longer(c(body_mass_kg, peak_power_w),
               names_to = "predictor", values_to = "value") |>
  mutate(predictor = factor(predictor,
                            levels = c("body_mass_kg", "peak_power_w"),
                            labels = c("Body mass (kg)",
                                       "Estimated peak jump power (W)"))) |>
  ggplot(aes(value, polar_strength_strain_index, colour = sex)) +
  geom_point(alpha = 0.7, size = 1.8) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.9) +
  facet_wrap(~ predictor, scales = "free_x") +
  scale_colour_manual(values = c(Female = "#C46B4A", Male = "#3F8F7A")) +
  labs(x = NULL, y = "Polar strength-strain index (mm3)", colour = "Sex")
```

Both body mass and estimated peak power track the polar strength-strain index
strongly and in the same direction in each sex, and unlike the trabecular index
the body-mass panel is at least as convincing as the power panel, which is the
first hint that the cortical shaft responds to carried load rather than to
explosive power alone.

## Univariable models

```{r univariable}
uv <- dat_model |>
  tbl_uvregression(
    y = all_of(outcome),
    method = lm,
    exponentiate = FALSE
  )
uv

utid <- map_dfr(predictors, function(v) {
  tidy(lm(reformulate(v, outcome), data = dat_model), conf.int = TRUE) |>
    filter(term != "(Intercept)")
})
ug <- function(term, d = 2) {
  r <- utid[utid$term == term, ]
  sprintf("%s mm3 (95%% CI %s to %s)", fmt(r$estimate, d),
          fmt(r$conf.low, d), fmt(r$conf.high, d))
}
up <- function(term) pv(utid$p.value[utid$term == term])
```

Body mass and peak power are both strong crude predictors. Each additional
kilogram of body mass is associated with `r ug("body_mass_kg", 1)` higher polar
strength-strain index (p `r up("body_mass_kg")`), and each additional watt of
estimated peak power with `r ug("peak_power_w", 3)`
(p `r up("peak_power_w")`), i.e.
`r fmt(1000 * utid$estimate[utid$term == "peak_power_w"], 0)` mm3 per 1000 W. Men
average `r ug("sexMale", 0)` higher than women (p `r up("sexMale")`).
Participants in decades 3 to 6 average `r ug("age_bandDecades 3-6", 0)` relative
to decade 2 (p `r up("age_bandDecades 3-6")`), while decade 1 is
indistinguishable from decade 2 in the crude analysis
(`r ug("age_bandDecade 1", 0)`, p `r up("age_bandDecade 1")`).

## Why adjustment is needed

**Body mass** confounds the crude peak-power association in a way that is
unusually explicit in this dataset: heavier participants have higher estimated
peak power partly because they jumped higher, but also partly by construction,
since body mass enters the Sayers equation with a coefficient of 48.9 W per
kilogram. Body mass is separately and powerfully associated with cortical bone
because the shaft's bending resistance adapts to the load it habitually carries,
which is essentially body weight during walking, so a crude power coefficient is
carrying a mass effect on both counts. **Sex** confounds the same association
from the other direction: men in this sample are heavier (median
`r fmt(median(dat_model$body_mass_kg[dat_model$sex == "Male"]), 1)` versus
`r fmt(median(dat_model$body_mass_kg[dat_model$sex == "Female"]), 1)` kg) and
more powerful (median
`r fmt(median(dat_model$peak_power_w[dat_model$sex == "Male"]), 0)` versus
`r fmt(median(dat_model$peak_power_w[dat_model$sex == "Female"]), 0)` W), and men
have longer, thicker tibiae for reasons of skeletal size that have nothing to do
with jumping, so the crude power slope is partly a comparison of men with women.
**Age band** confounds in the opposite direction and is worth adjusting for even
though its crude contrasts are unimpressive: the youngest participants are
*lighter* (mean body mass
`r fmt(mean(dat_model$body_mass_kg[dat_model$age_band == "Decade 1"]), 1)` kg in
decade 1 versus
`r fmt(mean(dat_model$body_mass_kg[dat_model$age_band == "Decade 2"]), 1)` kg in
decade 2) and more often female
(`r fmt(100 * mean(dat_model$sex[dat_model$age_band == "Decade 1"] == "Female"), 0)`%
versus
`r fmt(100 * mean(dat_model$sex[dat_model$age_band == "Decade 2"] == "Female"), 0)`%),
both of which push their cortical index down and mask whatever age pattern
exists. Mutual adjustment lets each coefficient be read as a comparison between
participants of the same sex, age band and body mass.

## Multivariable model

```{r multivariable}
fit <- lm(reformulate(predictors, outcome), data = dat_model)

mv <- tbl_regression(fit, exponentiate = FALSE)
mv

tid <- tidy(fit, conf.int = TRUE)
g <- function(term, d = 2) {
  r <- tid[tid$term == term, ]
  sprintf("%s mm3 (95%% CI %s to %s)", fmt(r$estimate, d),
          fmt(r$conf.low, d), fmt(r$conf.high, d))
}
gp <- function(term) pv(tid$p.value[tid$term == term])
```

```{r fit-stats}
s <- summary(fit)
tibble(
  Statistic = c("Complete-case N", "Parameters (excl. intercept)",
                "Observations per parameter",
                "R-squared", "Adjusted R-squared", "Residual SE (mm3)"),
  Value = c(as.character(nrow(model.frame(fit))),
            as.character(length(coef(fit)) - 1L),
            fmt(nrow(model.frame(fit)) / (length(coef(fit)) - 1L), 1),
            fmt(s$r.squared, 3), fmt(s$adj.r.squared, 3), fmt(s$sigma, 1))
) |>
  gt() |>
  tab_header(title = md("**Model fit: multivariable linear model for polar strength-strain index**"))
```

The model explains `r fmt(100 * s$r.squared, 1)`% of the variance in the polar
strength-strain index (adjusted R-squared `r fmt(s$adj.r.squared, 3)`), a
noticeably better fit than the same four predictors achieve for the trabecular
index at the distal tibia. **All four predictors are independently associated
with the outcome, and body mass is the strongest of them**, though not by a wide
margin: a one-standard-deviation increase in body mass corresponds to
`r fmt(tid$estimate[tid$term == "body_mass_kg"] * sd(dat_model$body_mass_kg), 0)`
mm3 against
`r fmt(tid$estimate[tid$term == "peak_power_w"] * sd(dat_model$peak_power_w), 0)`
mm3 for a standard deviation of peak power. Among participants of
the same sex, age band and estimated peak power, each additional kilogram of body
mass is associated with `r g("body_mass_kg", 1)` (p `r gp("body_mass_kg")`).
Among participants of the same sex, age band and body mass, each additional watt
of estimated peak power is associated with `r g("peak_power_w", 3)`
(p `r gp("peak_power_w")`), i.e.
`r fmt(1000 * tid$estimate[tid$term == "peak_power_w"], 0)` mm3 per 1000 W. Men
average `r g("sexMale", 0)` more than women alike on the other three
(p `r gp("sexMale")`). The decade 1 contrast, null in the crude analysis, becomes
clearly positive on adjustment: `r g("age_bandDecade 1", 0)`
(p `r gp("age_bandDecade 1")`), so at the same body mass, sex and power the
youngest participants have *stronger* cortical shafts than those in decade 2. The
decades 3-6 contrast remains null (`r g("age_bandDecades 3-6", 0)`,
p `r gp("age_bandDecades 3-6")`). These are associations among the participants
observed, not effects of intervening.

## Unadjusted versus adjusted

```{r side-by-side}
tbl_merge(list(uv, mv),
          tab_spanner = c("**Unadjusted**", "**Adjusted**")) |>
  modify_caption("**Table 2. Unadjusted and adjusted estimates.**")
```

**Peak power is the coefficient that moves most**, and it moves sharply toward the
null: from `r ug("peak_power_w", 3)` crude to `r g("peak_power_w", 3)` adjusted, a
loss of about
`r fmt(100 * (1 - tid$estimate[tid$term == "peak_power_w"] / utid$estimate[utid$term == "peak_power_w"]), 0)`%
of its magnitude, though it stays significant (p `r up("peak_power_w")` to
p `r gp("peak_power_w")`). This is confounding by body mass working exactly as
anticipated, compounded by the fact that body mass is literally a term inside the
power calculation. **Body mass** also attenuates substantially, from
`r ug("body_mass_kg", 1)` to `r g("body_mass_kg", 1)`, but it remains the
dominant predictor: body mass retains
`r fmt(100 * tid$estimate[tid$term == "body_mass_kg"] / utid$estimate[utid$term == "body_mass_kg"], 0)`%
of its crude estimate while peak power retains
`r fmt(100 * tid$estimate[tid$term == "peak_power_w"] / utid$estimate[utid$term == "peak_power_w"], 0)`%,
which is what happens when two heavily overlapping predictors are asked to
divide a shared association. **Sex** attenuates from
`r ug("sexMale", 0)` to `r g("sexMale", 0)` and stays significant, so more than
half of the crude male advantage in cortical strength is explained by men being
heavier and more powerful. The **decades 3-6** contrast is the one that changes
significance in the other direction, from `r ug("age_bandDecades 3-6", 0)`
(p `r up("age_bandDecades 3-6")`) to `r g("age_bandDecades 3-6", 0)`
(p `r gp("age_bandDecades 3-6")`): the older participants' apparent deficit is
entirely explained by their lower power, not by age. Most striking is **decade
1**, which moves *away* from the null, from `r ug("age_bandDecade 1", 0)`
(p `r up("age_bandDecade 1")`) to `r g("age_bandDecade 1", 0)`
(p `r gp("age_bandDecade 1")`) - a classic case of masking, since the youngest
participants are lighter and more often female, and once those are held constant
their cortical advantage appears. With
`r sum(dat_model$age_band == "Decade 1")` participants in that band it should be
treated as a hypothesis rather than a finding.

## Diagnostics

```{r diagnostics, fig.height=9}
check_model(fit)
```

```{r collin}
check_collinearity(fit)
```

```{r influence}
cd <- cooks.distance(fit)
i_max <- which.max(cd)
fit_drop <- lm(reformulate(predictors, outcome), data = dat_model[-i_max, ])
tibble(
  Term = names(coef(fit)),
  `Full sample` = fmt(coef(fit), 3),
  `Most influential dropped` = fmt(coef(fit_drop), 3),
  `Change (%)` = fmt(100 * (coef(fit_drop) - coef(fit)) / abs(coef(fit)), 1)
) |>
  filter(Term != "(Intercept)") |>
  gt() |>
  tab_header(title = md(paste0(
    "**Table 3. Coefficients with and without the single most influential ",
    "observation (Cook's D = ", fmt(max(cd), 3), ").**")))
```

```{r het-check}
het <- check_heteroscedasticity(fit)
print(het)
nor <- check_normality(fit)
print(nor)
```

```{r robust-se}
rob   <- lmtest::coeftest(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
robci <- lmtest::coefci(fit, vcov. = sandwich::vcovHC(fit, type = "HC3"))
tibble(
  Term = rownames(rob),
  Estimate = fmt(rob[, 1], 3),
  `SE (OLS)` = fmt(summary(fit)$coefficients[, 2], 3),
  `SE (HC3)` = fmt(rob[, 2], 3),
  `95% CI (HC3)` = paste0(fmt(robci[, 1], 2), " to ", fmt(robci[, 2], 2)),
  `p (HC3)` = fmt(rob[, 4], 3)
) |>
  filter(Term != "(Intercept)") |>
  gt() |>
  tab_header(title = md("**Table 4. HC3 heteroscedasticity-consistent standard errors.**"))
```

**Linearity and homogeneity of variance.** The residual-versus-fitted panel fans
out clearly: participants with high fitted values have far more variable
residuals than those with low ones, and `check_heteroscedasticity()` rejects
constant variance (p `r pv(as.numeric(het))`). Table 4 therefore reports HC3
robust standard errors. Inference is not overturned - body mass keeps
p `r pv(rob["body_mass_kg", 4])`, sex p `r pv(rob["sexMale", 4])`, and peak power
p `r pv(rob["peak_power_w", 4])` - but the HC3 standard error for peak power is
`r fmt(rob["peak_power_w", 2], 3)` against an OLS value of
`r fmt(summary(fit)$coefficients["peak_power_w", 2], 3)`, i.e. wider, so the
power coefficient is the least secure of the four.

**Normality of residuals.** The normal-QQ panel departs from the reference line
in the upper tail and `check_normality()` **rejects**
(p `r pv(as.numeric(nor))`). This is the diagnostic that fails hardest, and it is
driven by the same handful of very large tibiae that cause the variance to fan.
The posterior-predictive check shows the corresponding symptom: the fitted normal
predictive distribution is symmetric while the observed outcome is right-skewed.

**Collinearity.** The variance inflation factors are
`r fmt(check_collinearity(fit)$VIF[check_collinearity(fit)$Term == "peak_power_w"], 2)`
for peak power and
`r fmt(check_collinearity(fit)$VIF[check_collinearity(fit)$Term == "body_mass_kg"], 2)`
for body mass, with sex at
`r fmt(check_collinearity(fit)$VIF[check_collinearity(fit)$Term == "sex"], 2)` and
age band at
`r fmt(check_collinearity(fit)$VIF[check_collinearity(fit)$Term == "age_band"], 2)`.
These stay below the conventional threshold of 5 but they are the direct price of
the parameterisation chosen in Methods, and they are why the two dominant
predictors each absorb roughly half of the other's crude association.

**Influential observations.** This matters more here than for any other model in
this dataset. The largest Cook's distance is `r fmt(max(cd), 3)` - still far
below the conventional concern threshold of 0.5, and the largest leverage is
`r fmt(max(hatvalues(fit)), 3)`, with `r sum(cd > 4 / nrow(dat_model))`
observations exceeding the stricter 4/n rule. But Cook's distance understates the
problem for a single coefficient. Table 3 shows that refitting without that one
participant moves the **peak-power coefficient from
`r fmt(coef(fit)["peak_power_w"], 3)` to
`r fmt(coef(fit_drop)["peak_power_w"], 3)` mm3 per watt, a change of
`r fmt(abs(100 * (coef(fit_drop)["peak_power_w"] - coef(fit)["peak_power_w"]) / abs(coef(fit)["peak_power_w"])), 0)`%**,
while body mass moves only
`r fmt(abs(100 * (coef(fit_drop)["body_mass_kg"] - coef(fit)["body_mass_kg"]) / abs(coef(fit)["body_mass_kg"])), 1)`%
and sex only
`r fmt(abs(100 * (coef(fit_drop)["sexMale"] - coef(fit)["sexMale"]) / abs(coef(fit)["sexMale"])), 1)`%.
The participant in question has the **highest polar strength-strain index in the
sample** (`r fmt(dat_model[[outcome]][i_max], 0)` mm3, against a median of
`r fmt(median(dat_model[[outcome]]), 0)`) together with a peak power in the top
`r fmt(100 - 100 * mean(dat_model$peak_power_w < dat_model$peak_power_w[i_max]), 0)`%
of the sample (`r fmt(dat_model$peak_power_w[i_max], 0)` W); the model
under-predicts them by `r fmt(resid(fit)[i_max], 0)` mm3, and because they sit at
the top of the power range that residual levers the power slope upward. The
peak-power estimate remains
significant without them (p
`r pv(summary(fit_drop)$coefficients["peak_power_w", 4])`), so the conclusion
survives, but at n = `r nrow(dat_model)` a single participant moving a
coefficient by a fifth is exactly the fragility that a small sample produces and
it should be reported rather than hidden.

## Sensitivity: alternative parameterisations

```{r sensitivity}
alt <- list(
  `Chosen: body mass + absolute power`  = c("sex", "age_band", "body_mass_kg", "peak_power_w"),
  `Equivalent: body mass + jump height` = c("sex", "age_band", "body_mass_kg", "jump_cm"),
  `Alternative: body mass + relative power` = c("sex", "age_band", "body_mass_kg",
                                                "relative_peak_power"),
  `Absolute power only (no body mass)`  = c("sex", "age_band", "peak_power_w"),
  `WRONG: all three power/mass terms`   = c("sex", "age_band", "body_mass_kg",
                                            "peak_power_w", "relative_peak_power")
)
alt_fits <- map(alt, ~ lm(reformulate(.x, outcome), data = dat))

map2_dfr(names(alt_fits), alt_fits, function(nm, f) {
  tibble(
    Model = nm,
    `Parameters` = length(coef(f)) - 1L,
    `R-squared` = fmt(summary(f)$r.squared, 3),
    `Adjusted R-squared` = fmt(summary(f)$adj.r.squared, 3),
    `Residual SE` = fmt(summary(f)$sigma, 1),
    `Max VIF` = fmt(max(check_collinearity(f)$VIF), 1)
  )
}) |>
  gt() |>
  tab_header(title = md("**Table 5. The same data under five parameterisations of body mass and power.**"))
```

```{r reparam}
f_pp  <- alt_fits[["Chosen: body mass + absolute power"]]
f_jh  <- alt_fits[["Equivalent: body mass + jump height"]]
f_rpp <- alt_fits[["Alternative: body mass + relative power"]]
f_bad <- alt_fits[["WRONG: all three power/mass terms"]]
b_pp <- coef(f_pp)["peak_power_w"]
b_jh <- coef(f_jh)["jump_cm"]
```

Table 5 makes the argument of the Methods section concrete. The chosen model and
the jump-height model have **identical** R-squared
(`r fmt(summary(f_pp)$r.squared, 6)` versus
`r fmt(summary(f_jh)$r.squared, 6)`) and identical residual standard error,
because they are the same model written twice: multiplying the peak-power
coefficient by 51.9 gives `r fmt(51.9 * b_pp, 6)`, against a jump-height
coefficient of `r fmt(b_jh, 6)` mm3 per centimetre - the same number to six
decimal places, not merely a similar one. Read this way, the adjusted power
effect is a jump-height effect of
`r fmt(coef(f_jh)["jump_cm"], 1)` mm3 per centimetre jumped, and the body-mass
coefficient becomes `r fmt(coef(f_jh)["body_mass_kg"], 1)` mm3 per kilogram
(p `r pv(summary(f_jh)$coefficients["body_mass_kg", 4])`) rather than
`r fmt(coef(f_pp)["body_mass_kg"], 1)` - the same information, split differently
between the two terms, which is precisely why the parameterisation has to be
chosen deliberately and stated.

Substituting relative for absolute power costs a little fit
(R-squared `r fmt(summary(f_rpp)$r.squared, 3)` against
`r fmt(summary(f_pp)$r.squared, 3)`) and, more importantly, leaves the power term
non-significant (p
`r pv(summary(f_rpp)$coefficients["relative_peak_power", 4])`), so for this
outcome the two parameterisations do **not** lead to the same conclusion about
muscle power. Dropping body mass altogether costs more
(R-squared `r fmt(summary(alt_fits[[4]])$r.squared, 3)`). The last row is the
trap: entering body mass, absolute power *and* relative power together raises
R-squared to `r fmt(summary(f_bad)$r.squared, 3)` - which is exactly why a student
would be tempted - while pushing the maximum variance inflation factor to
`r fmt(max(check_collinearity(f_bad)$VIF), 0)`. In that model relative power
takes a coefficient of `r fmt(coef(f_bad)["relative_peak_power"], 1)` mm3 per
W/kg (p `r pv(summary(f_bad)$coefficients["relative_peak_power", 4])`) and body
mass flips to `r fmt(coef(f_bad)["body_mass_kg"], 1)`; both are significant or
near-significant and both are nonsense, since they say that being more powerful
for one's size and being heavier are each associated with *weaker* cortical bone.
That is what near-perfect collinearity looks like when it is significant, and it
is the single most dangerous thing a student can do with this dataset.

```{r log-sens}
fit_log <- lm(reformulate(predictors, paste0("log(", outcome, ")")), data = dat_model)
het_log <- check_heteroscedasticity(fit_log)
nor_log <- check_normality(fit_log)
```

Because both the variance and the normality checks failed, we also refitted on
the log scale. This resolves both problems - heteroscedasticity
p `r pv(as.numeric(het_log))`, normality p `r pv(as.numeric(nor_log))` - and
improves the fit (R-squared `r fmt(summary(fit_log)$r.squared, 3)` against
`r fmt(s$r.squared, 3)`). The conclusions are unchanged in direction and
significance: each kilogram of body mass is associated with
`r fmt(100 * (exp(coef(fit_log)["body_mass_kg"]) - 1), 2)`% higher polar
strength-strain index (p
`r pv(summary(fit_log)$coefficients["body_mass_kg", 4])`), each 1000 W of peak
power with `r fmt(100 * (exp(1000 * coef(fit_log)["peak_power_w"]) - 1), 1)`%
(p `r pv(summary(fit_log)$coefficients["peak_power_w", 4])`), and men with
`r fmt(100 * (exp(coef(fit_log)["sexMale"]) - 1), 1)`% higher
(p `r pv(summary(fit_log)$coefficients["sexMale", 4])`). On the evidence of the
diagnostics the log model is the better model for this outcome; the natural-scale
model is retained as primary only because its coefficients are in the units the
source paper uses, with HC3 standard errors doing the inferential work.

# Discussion

Among these `r nrow(dat_model)` healthy adults, **body mass is the characteristic
most strongly and independently associated with cortical bone strength at the
tibial shaft** - about `r fmt(tid$estimate[tid$term == "body_mass_kg"], 0)` mm3
per kilogram among participants of the same sex, age band and estimated jump
power - with estimated peak power making a smaller but still independent
contribution of about
`r fmt(1000 * tid$estimate[tid$term == "peak_power_w"], 0)` mm3 per 1000 W. This
is the mirror image of what the same four predictors show for the trabecular
index at the distal tibia, where power dominates and body mass falls away
entirely. The contrast is biologically coherent: the shaft's job is to resist the
bending moments generated by carrying body weight, whereas the distal end resists
the compressive impacts generated by fast, forceful movement, and each site
appears to track the loading it actually experiences. Practically, it cautions
against treating "bone strength" as one thing; an intervention that improved
jumping power would not necessarily do the same for both sites.

**Limitations.** The design is **cross-sectional**, so temporality is unknown and
reverse causation cannot be excluded - people with larger, stronger tibiae are
likely to have been more active for years, so the arrow may run from bone to
power rather than the other way. The relationship with body mass is also open to
a purely structural reading: bigger people have bigger bones, and a geometric
index of a larger bone will be larger without anything adaptive happening; tibial
length, which would allow this to be separated, is not in the file. The sample is
a **convenience sample of one university's staff and students**, young
(`r fmt(100 * mean(dat_model$age_band == "Decade 2"), 0)`% in a single decade
band) and healthy, so nothing here generalises to older adults or to anyone at
real fracture risk, and the decade 1 finding rests on
`r sum(dat_model$age_band == "Decade 1")` participants. **Peak power was
estimated, not measured**: it is the Sayers regression equation applied to jump
height and body mass, and the jump heights are recorded to the nearest half inch,
giving only `r n_jump_values` distinct values. Measurement error in a predictor
attenuates its coefficient toward the null, so the true association with
genuinely measured power is if anything stronger than reported; it also means the
power variable is partly a restatement of body mass, which is why the two compete
(VIF `r fmt(max(check_collinearity(fit)$VIF), 1)`) and why the split of the
association between them depends on the parameterisation. **Age is a decade bin,
not a number.** Residual variance is non-constant *and* residuals are non-normal;
HC3 addresses inference and the log model addresses both, but the natural-scale
model does not fit the upper tail. At n = `r nrow(dat_model)` the peak-power
coefficient moves by
`r fmt(abs(100 * (coef(fit_drop)["peak_power_w"] - coef(fit)["peak_power_w"]) / abs(coef(fit)["peak_power_w"])), 0)`%
when the single most influential participant is removed, so that estimate in
particular should be read as provisional. Physical activity, diet, tibial length,
limb dominance and hormonal history are **not measured**, so residual confounding
is certain. All estimates are associations, not causal effects.

**Future work.** A study with directly measured force-plate power and tibial
length would separate the geometric size effect from the loading effect, which is
the ambiguity this analysis cannot resolve.

# Feasibility notes

*NOT part of a student report.*

- **USABLE WITH WARNINGS.** The outcome is complete, the model is well specified
  at `r fmt(nrow(dat_model) / (length(coef(fit)) - 1L), 1)` observations per
  parameter and the fit is the best in the dataset
  (R-squared `r fmt(s$r.squared, 2)`), but **two diagnostics fail** and one
  coefficient is materially sensitive to a single participant. The warnings are
  all reportable rather than fatal.
- **MAJOR: both heteroscedasticity (p `r pv(as.numeric(het))`) and non-normality
  of residuals (p `r pv(as.numeric(nor))`) fail.** This is worse than the
  companion trabecular outcome, where normality passes. A log transform fixes
  both (p `r pv(as.numeric(het_log))` and p `r pv(as.numeric(nor_log))`) and
  raises R-squared from `r fmt(s$r.squared, 3)` to
  `r fmt(summary(fit_log)$r.squared, 3)`, so the honest answer is that this
  outcome should be modelled on the log scale. A student who reports the raw-scale
  model with no transformation and no robust standard errors has two failed
  checks on the record.
- **MAJOR: one participant moves the peak-power coefficient by
  `r fmt(abs(100 * (coef(fit_drop)["peak_power_w"] - coef(fit)["peak_power_w"]) / abs(coef(fit)["peak_power_w"])), 0)`%.**
  Dropping the observation with the largest Cook's distance
  (`r fmt(max(cd), 3)`, the participant with the highest outcome value in the
  file and a peak power in the top few per cent) changes the estimate from
  `r fmt(coef(fit)["peak_power_w"], 3)` to
  `r fmt(coef(fit_drop)["peak_power_w"], 3)`. Cook's distance alone does not flag
  it, because `r fmt(max(cd), 3)` looks reassuring next to the 0.5 rule of thumb;
  the leave-one-out refit is what exposes it. At n = `r nrow(dat_model)` this
  check should be mandatory, and students should be told to do it rather than to
  trust the Cook's distance panel.
- **MAJOR trap: `peak_power_w` and `relative_peak_power` are the same variable.**
  `relative_peak_power` is exactly `peak_power_w / body_mass_kg` rounded to 2 dp
  (largest discrepancy across all `r nrow(dat_raw)` rows: `r rpp_err`). This
  outcome is where the trap bites hardest, because putting all three of body mass,
  absolute power and relative power in one model **raises R-squared** from
  `r fmt(s$r.squared, 3)` to `r fmt(summary(f_bad)$r.squared, 3)` and returns
  *significant* coefficients (relative power
  `r fmt(coef(f_bad)["relative_peak_power"], 1)`,
  p `r pv(summary(f_bad)$coefficients["relative_peak_power", 4])`) with VIFs of
  `r fmt(max(check_collinearity(f_bad)$VIF), 0)`,
  `r fmt(sort(check_collinearity(f_bad)$VIF, decreasing = TRUE)[2], 0)` and
  `r fmt(sort(check_collinearity(f_bad)$VIF, decreasing = TRUE)[3], 0)`. A student
  chasing R-squared will build this model, get significant results, and write
  nonsense about heavier people having weaker cortical bone. The catalogue's own
  suggested predictor list contains **both** power variables, so this must be
  flagged in `meta.json`.
- **MAJOR trap: `peak_power_w` already contains `body_mass_kg`.** It is the
  Sayers (1999) equation, `PP = 51.9*jump_cm + 48.9*body_mass - 2007`, inverted;
  the recovered jump heights land on a half-inch grid to within
  `r fmt(halfinch_err, 5)` and reproduce `peak_power_w` to `r pp_err` W. Because
  of this, the model with body mass and absolute power is *algebraically the same*
  as the model with body mass and jump height, and the two terms split the
  association differently under each parameterisation
  (body mass `r fmt(coef(f_pp)["body_mass_kg"], 1)` versus
  `r fmt(coef(f_jh)["body_mass_kg"], 1)` mm3/kg). The parameterisation is a
  choice, not a detail, and it must be justified in Methods.
- **MAJOR: the choice of power variable changes the conclusion for this
  outcome.** With absolute power the power term is significant
  (p `r gp("peak_power_w")`); with relative power in its place it is not
  (p `r pv(summary(f_rpp)$coefficients["relative_peak_power", 4])`). Two students
  making different, individually defensible choices will write different
  discussions. This is unavoidable given the data, but markers should expect it.
- **MAJOR: `age_decade` cannot be used as a six-level factor.** Cell sizes are
  `r paste(as.integer(table(dat$age_decade)), collapse = ", ")` for decades 1
  through 6, so decade 6 holds `r sum(dat$age_decade == 6)` participants and
  there are no men at all in decades 4 or 6. As a six-level factor the largest
  standard error is
  `r fmt(max(summary(lm(polar_strength_strain_index ~ sex + factor(age_decade) + body_mass_kg + peak_power_w, dat))$coefficients[, 2]), 0)`
  mm3 and the extra three degrees of freedom buy nothing (F test against the
  collapsed model p =
  `r fmt(anova(fit, lm(polar_strength_strain_index ~ sex + factor(age_decade) + body_mass_kg + peak_power_w, dat))$"Pr(>F)"[2], 2)`).
  Collapsing to three bands is required, not optional.
- **MINOR: the decade 1 coefficient appears only after adjustment.** It goes from
  `r fmt(utid$estimate[utid$term == "age_bandDecade 1"], 0)` mm3
  (p `r up("age_bandDecade 1")`) crude to
  `r fmt(tid$estimate[tid$term == "age_bandDecade 1"], 0)` mm3
  (p `r gp("age_bandDecade 1")`) adjusted, on
  `r sum(dat_model$age_band == "Decade 1")` participants. It is a genuine masking
  effect (that band is lighter and more often female) but it is fragile and a
  student who leads on it is over-reaching.
- **MINOR: no binary outcome exists in this file, and none should be
  manufactured.** A student will be tempted to split the polar strength-strain
  index at its median or a percentile to run a logistic model. There is **no
  clinical threshold** for SSI in healthy adults - it has no diagnostic
  cut-point, no reference population and no fracture-risk calibration - so any
  dichotomy would be arbitrary, would discard the information that gives this
  model its R-squared of `r fmt(s$r.squared, 2)`, and would produce an "outcome"
  that answers no question anyone asks. The dataset supports a **linear project
  only**, and `index.html` correctly says so.
- **MINOR: the two outcomes are correlated at
  `r fmt(cor(dat$bone_strength_index, dat$polar_strength_strain_index), 2)`**
  (Pearson), i.e. they share
  `r fmt(100 * cor(dat$bone_strength_index, dat$polar_strength_strain_index)^2, 0)`%
  of their variance, and the residual correlation after this model is
  `r fmt(cor(resid(fit), resid(lm(bone_strength_index ~ sex + age_band + body_mass_kg + peak_power_w, dat))), 2)`.
  They are **not** redundant and both should stay in the catalogue: the two
  analyses reach genuinely different conclusions - body mass dominates the shaft
  (`r fmt(tid$estimate[tid$term == "body_mass_kg"], 1)` mm3/kg,
  p `r gp("body_mass_kg")`) while it is null at the distal site, and peak power
  dominates the distal site while it is the weaker of the two here.
- **Sample size is the binding constraint.** `r nrow(dat_raw)` participants and
  five candidate predictors, two of which cannot coexist, leaves essentially one
  sensible model. There is no room for interactions, splines or subgroup
  analyses, and students should be told that plainly rather than discovering it.
